### 2022-09-06 SticklesEDA.Rmd
Commiting Data and initial EDA

---
title: "Stickleback EDA"
author: "Akhil Ghosh"
date: "2022-09-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Loading Libraries
```{r}
library(tidyverse)
library(ggplot2)
library(readr)
library(dplyr)
library(DT)

```

```{r}
stickle <- read_csv('stickleback_fossil_vs_extant.csv')
summary(stickle)
#seems every column outside of ID has missing data to some extent
#Highest amount of NA's is cle.sc column
#least amount of NAs in mds column
```

```{r}
datatable(stickle)

#Should I split up univID?
#Maybe, maybe not
```
```{r}
stickles <- datatable(stickle)
grepl("M",stickle$univID, fixed=TRUE)
stickle$univID
stick <- stickle %>% 
  mutate(gender = ifelse(grepl("M",univID,fixed=TRUE),"M","F"))

table(stick$gender)
#Only 130 Males?????
#sheesh

```

### 2022-09-21 SticklesEDA.Rmd
lots of stickle changes

@@
 library(tidyverse)
-library(ggplot2)
-library(readr)
-library(dplyr)
 library(DT)
@@
 stickle$univID
+
 stick <- stickle %>% 
-  mutate(gender = ifelse(grepl("M",univID,fixed=TRUE),"M","F"))
+  mutate(gender = case_when(
+    grepl(".M.",univID,fixed=TRUE) ~ "Male",
+    grepl(".F.",univID,fixed=TRUE) ~ "Female",
+    grepl("Fos",univID,fixed=TRUE) ~ NA_character_
+    )
+  )
+           # ifelse(grepl("M",univID,fixed=TRUE),"M","F"))
 
 table(stick$gender)
-#Only 130 Males?????
-#sheesh
+#Only 284 extant fish
+
+summary(stick)
+
+summary(as.factor(stick$gender))
 
 ```
+```{r}
+install.packages("mice")
+library(mice)
+
+md.pattern(stickle)
+
+```
+
+```{r}
+# devtools::install_github("agnesdeng/mixgb")
+
+library(mixgb)
+
+stick_clean <- stick %>% 
+  select(.,-univID) %>% 
+  mutate(ips = as.factor(ips))
+
+stick_clean <- data_clean(stick_clean)
+stick_clean
+```
+```{r}
+# imputed_data <- mixgb(data = stick_clean, m = 5)
+# imputed_data
+
+cv.results <- mixgb_cv(data = stick_clean, verbose = FALSE)
+cv.results$response
+cv.results$best.nrounds
+
+cv.results2 <- mixgb_cv(data = stick_clean, nfold = 10, nrounds = 100,
+    early_stopping_rounds = 1, response = "gender", verbose = FALSE)
+cv.results2$best.nrounds
+
+imputed_data2 <- mixgb(data = stick_clean, m = 5, nrounds = 10)
+imputed_data3 <- mixgb(data = stick_clean, m = 5, nrounds = 30)
+imputed_data3
+
+```
+
+```{r}
+
+#Here I try to impute every variable in the data set and then use the imputed dataset to predict gender variable. Through this I can cross validate and avoid data cleaning issues.
+
+#Just a prelim thing I wanted to see the accuracy
+#For this, I will be focusing on the extant dataset.
+
+summary(stick_clean)
+
+non_gender <- stick_clean %>% 
+  select(-gender)
+
+cv.results_ <- mixgb_cv(data = non_gender, verbose=FALSE)
+cv.results_$best.nrounds
+cv.results_$response
+
+imputed_data_no_gender <- mixgb(data = no, m = 5, nrounds = 100)
+imputed_data_ng <- mixgb(data = non_gender, m=5, nrounds = 100)
+
+imputed_data_ng
+
+```
+
+```{r}
+best_rounds <- c()
+for(i in 1:50){
+cv.results_ <- mixgb_cv(data = non_gender, verbose=FALSE,nrounds=200)
+best_rounds <-append(best_rounds, cv.results_$best.nrounds)
+if(best_rounds[i]==max(best_rounds)){
+  maxresponse = cv.results_$response
+  maxRounds <- cv.results_$best.nrounds
+}
+
+}
+
+max(best_rounds)
+maxresponse
+
+cv.results_ <- mixgb_cv(data = non_gender,response="stl", verbose=FALSE,nrounds=200)
+cv.results_$best.nrounds
+
+parallel::detectCores()
+n.cores <- parallel::detectCores() - 1
+
+my.cluster <- parallel::makeCluster(
+  n.cores, 
+  type = "FORK"
+  )
+doParallel::registerDoParallel(cl = my.cluster)
+foreach::getDoParRegistered()
+foreach::getDoParWorkers()
+
+best_rounds<-
+x <- foreach(
+  i = 1:100, 
+  .combine = 'c'
+) %dopar% {
+  cv.results_ <- mixgb_cv(data = non_gender, verbose=FALSE,nrounds=200)
+  best_rounds <-append(best_rounds, cv.results_$best.nrounds)
+}  
+best_rounds
+max(x)
+```
+
+```{r}
+library(tidymodels)
+install.packages("janitor")
+stickelback_quang <- read_csv("https://www.dropbox.com/s/hpsn643fd8urph2/stickleback_fossil_vs_extant.csv?dl=1") %>% 
+  janitor::clean_names() %>% 
+  separate(univ_id, 
+           into = c("lake", "body_type", "gender", "id"), 
+           sep = "\\.",
+           remove = FALSE)
+
+stickelback_quang
+
+extant_quan <- stickelback_quang %>%
+  filter(lake != "Fos") %>% 
+  select(gender, stl, ips, mds, mdf, maf, mav, mcv, mpt, ips_prop, 
+         contains("_sc"), str_remove(names(stickelback_quang)[24:32], "_sc")) %>% 
+  mutate(across(where(is_character), as_factor))
+extant_quan
+
+```
+```{r}
+my_data <- imputed_data_ng[1]
+imputed_data_ng
+
+my_data_new <- cbind(my_data, stick_clean[c("gender")])
+
+my_data_extant <- my_data_new %>% 
+  filter(gender == "Female" | gender=="Male") %>% 
+  mutate(gender=as.factor(case_when(gender=="Female" ~ "F",
+                   gender=="Male" ~ "M"
+
+          )
+         )
+  )
+
+```
+
+```{r}
+all_cores <- parallel::detectCores(logical = FALSE)
+library(doParallel)
+
+cl <- makePSOCKcluster(all_cores)
+registerDoParallel(cl)
+```
+
+```{r}
+set.seed(100)
+
+split <- initial_split(my_data_extant, prop = 0.7, strata = gender)
+
+train <- training(split)
+test <- testing(split)
+
+train_fold <- vfold_cv(train, v = 10, strata = gender)
+train_fold$splits
+
+xg_spec <- boost_tree(trees = tune(),
+                      tree_depth = tune(), 
+                      min_n = tune(), 
+                      loss_reduction = tune(),
+                      sample_size = tune(), 
+                      mtry = tune(),
+                      learn_rate = tune()) %>% 
+  set_engine("xgboost") %>% 
+  set_mode("classification")
+
+xg_rec <- recipe(gender ~., data = train) %>% 
+  step_dummy(all_nominal_predictors())
+
+xg_rec$var_info
+
+xg_wf <- workflow() %>% 
+  add_model(xg_spec) %>% 
+  add_recipe(xg_rec)
+
+set.seed(105)
+xg_grid <- grid_latin_hypercube(trees(),
+                                tree_depth(),
+                                min_n(),
+                                loss_reduction(),
+                                sample_size = sample_prop(),
+                                finalize(mtry(), train),
+                                learn_rate(),
+                                size = 100)
+
+xg_tune <- tune_grid(xg_wf,
+                     resamples = train_fold,
+                     grid = xg_grid,
+                     control = control_grid(save_pred = TRUE))
+
+xg_tune$.predictions
+best_acc <- xg_tune %>% 
+  select_best("accuracy")
+
+best_acc
+
+xg_final <- finalize_workflow(
+  xg_wf,
+  best_acc
+)
+
+xg_final
+
+xg_last <- last_fit(xg_final, split)
+
+xg_last %>% 
+  collect_predictions() %>% 
+  roc_curve(gender, .pred_F) %>% 
+  ggplot(aes(1 - specificity, sensitivity)) +
+  geom_abline() +
+  geom_path() +
+  coord_fixed()
+
+xg_last %>% 
+  collect_predictions() %>% 
+  roc_auc(gender, .pred_F)
+
+xg_last %>% 
+  collect_predictions() %>% 
+  accuracy(gender, .pred_class)
+table(xg_last$.predictions[[1]]$.pred_class==xg_last$.predictions[[1]]$gender)
+
+```
+
+```{r}
+library(tidymodels)
+getImputedExtantData <- function(data,data2,num){
+  data_out <- data[num]
+  data_out <- cbind(data_out,data2[c("gender")])
+  data_out <- data_out %>% 
+    filter(gender == "Female" | gender=="Male") %>% 
+    mutate(gender=as.factor(case_when(gender=="Female" ~ "F",
+                   gender=="Male" ~ "M"
+
+          )
+         )
+  )
+  return(data_out)
+}
+
+getAccuracy  <- function(data){
+
+  split <- initial_split(data, prop = 0.7, strata = gender)
+  train <- training(split)
+  test <- testing(split)
+
+  train_fold <- vfold_cv(train, v = 10, strata = gender)
+
+  xg_spec <- boost_tree(trees = tune(),
+                        tree_depth = tune(), 
+                        min_n = tune(), 
+                        loss_reduction = tune(),
+                        sample_size = tune(), 
+                        mtry = tune(),
+                        learn_rate = tune()) %>% 
+    set_engine("xgboost") %>% 
+    set_mode("classification")
+
+  xg_rec <- recipe(gender ~., data = train) %>% 
+    step_dummy(all_nominal_predictors())
+
+  xg_wf <- workflow() %>% 
+    add_model(xg_spec) %>% 
+    add_recipe(xg_rec)
+  xg_grid <- grid_latin_hypercube(trees(),
+                                  tree_depth(),
+                                  min_n(),
+                                  loss_reduction(),
+                                  sample_size = sample_prop(),
+                                  finalize(mtry(), train),
+                                  learn_rate(),
+                                  size = 100)
+  xg_tune <- tune_grid(xg_wf,
+                       resamples = train_fold,
+                       grid = xg_grid,
+                       control = control_grid(save_pred = TRUE))
+
+  best_acc <- xg_tune %>% 
+    select_best("accuracy")
+
+  xg_final <- finalize_workflow(
+    xg_wf,
+    best_acc
+  )
+  xg_last <- last_fit(xg_final, split)
+
+  data_output <- xg_last %>% 
+    collect_predictions() %>% 
+    accuracy(gender, .pred_class)
+
+  return(data_output)
+
+}
+
+imputed_data_tuned <- mixgb(data = non_gender, m=5, nrounds = 140)
+
+new_list <- list()
+
+for(i in 1:5){
+  new_list[[i]] <- getAccuracy(getImputedExtantData(imputed_data_tuned,stick_clean,i))
+}
+
+new_list[[1]]
+c <- c()
+
+est/5
+imputed_data_tuned
+
+```

### 2023-01-23 stickle_imputation_error_bars.Rmd
pushing error bars plots for MICE imputation

---
title: "imputation_stickle"
author: "Akhil Ghosh"
output: html_document
---

```{r setup, include=FALSE}
library(mixgb)
library(mice)
library(tidyverse)
library(janitor)
library(missMethods)
library(ROCR)

```

DATA DICTIONARY
stl: standard length. A measure of size from tip of snout to end of tail. (mm)

ips: index pelvic score. a categorical measure of pelvic complexity based on the number of pelvic elements observed. 0-3 in fossils. 0-8 in extant fish.
ips.prop: ips/3 (fossils) or ips/8 (extant) to put ips on same scale from 0 to 1.
lps: length of a pelvic spine (if ips <3 (fossils) or ips < 8 (extant), then pelvic spines were considered evolutionarily lost, and were assigned lengths of 0)
tpg: pelvic girdle length (mm). (when missing (ips = 0), considered evolutionarily lost and assigned lengths of 0.)

mav: abdominal vertebrae number (count)
mcv: caudal vertebrae number (count)
mdf: dorsal fin ray number. number of bones in the dorsal fin (count)
maf: anal fin ray number (count)

mds: dorsal spine number (count)
ds1: length of the first dorsal spine (mm) (when missing, the first dorsal spine was considered evolutionarily lost, and assigned lengths of 0. when spines were broken assigned NA)
ds2: length of the second dorsal spine (mm) (when missing, the second dorsal spine was considered evolutionarily lost, and assigned lengths of 0. when spines were broken assigned NA)
ds3: length of the third dorsal spine (mm) (when missing, the third dorsal spine was considered evolutionarily lost, and assigned lengths of 0. when spines were broken assigned NA)

mpt: pterygiophore number (count). Number of armor bones on the "back". 
lpt: length of the pterygiophore anterior to the the third spine (mm)

ect: ectocoracoid length (mm)
cle: cleithrum length (mm)
pmx: premaxila length (mm)

```{r}
stickle <- read_csv('stickle_revamped.csv')

stick <- stickle %>% 
  mutate(gender = as.factor(case_when(
    grepl(".M.",univID,fixed=TRUE) ~ "Male",
    grepl(".F.",univID,fixed=TRUE) ~ "Female",
    grepl("Fos",univID,fixed=TRUE) ~ NA_character_
    )), ips = as.factor(ips)
  ) %>% 
  select(gender, everything())

stick

extant_stickle <- stick %>% filter(gender != "Fossil") %>% select(!univID)
extant_stickle
summary(extant_stickle)
```

```{r}

#Checking imputation accuracy/viability

M = 5
k = 5
stick_matrix <- matrix(NA, nrow=nrow(extant_stickle), ncol = M)
set.seed(12)
fold <- sample(1:k, size = nrow(extant_stickle), replace=TRUE)
for(i in 1:k){
extant_stickle_temp <- extant_stickle
extant_stickle_temp$gender[fold==i] <- NA
imputed_stick <- mice(extant_stickle_temp,m=M)
for(n in 1:M){

  complete_stick <- complete(imputed_stick, n)

  complete_stick$gender[fold==i]
  stick_matrix[fold==i, n] <- as.character(complete_stick$gender[fold==i])
}

}
pct_male <- apply(stick_matrix,1,function(x){
  mean(x == "Male")
})

imptest <- data_frame(truth = as.factor(extant_stickle$gender), pct_male)
roc_pred <- prediction(imptest$pct_male, imptest$truth)

auc_rocr <- performance(roc_pred, measure="auc")
plot(performance(roc_pred, measure="tpr", x.measure="fpr"))

auc_rocr@y.values[[1]]

```

```{r}
#Using MICE imputation to get imputed datasets
M = 5
stick_whole <- stick %>% select(-c(univID, ips))
summary(stick_whole)
imputed_stick_whole <- mice(stick_whole,m=5)
complete_stick_list <- list()
for(m in 1:M){
  complete_stick_list[[m]] <- complete(imputed_stick_whole, m)
}
```

```{r}
#getting time vector, adding it to imputed data
univ <- stick %>% 
  mutate(time = as.integer(sub(".*([0-9]{2})\\..*", "\\1",univID)))
times <- univ %>% drop_na(time) %>% pull(time)

times_vector <- c(rep(1,43)) %>% 
  append(21-times)
```

```{r}
#Initial grouping and summarizing
# temp_comp <- complete_stick_list[[1]][-c(1:367),]
# temp_comp$time <- times_vector
# 
# means_stl_time <- temp_comp %>% group_by(gender, time) %>% summarize(xbar = mean(stl), se = sqrt(var(stl)/n()), n())
# 
# 
# help(mice)
# temp_comp %>% select(gender, time, stl) %>% filter(time==17)
# 
# 
# with(temp_)
# 
# means_stl_time %>% ggplot(aes(x = time, y = xbar, col = gender))+ geom_point() +geom_line()
# 
# table(complete_stick_list[[1]][c(1:367),]$gender)
# 
# table(complete_stick_list[[1]][-c(1:367),]$gender)

```

```{r}
#doing the grouping and summarizing for all imputed datasets, looking specifically at mean standard length for each group

temp_comp_lists <- list()
est_list <- list()
M = 5
for(m in 1:M){

temp_comp_lists[[m]] <- complete_stick_list[[m]][-c(1:367),]
temp_comp_lists[[m]]$time <- times_vector

est_list[[m]] <- temp_comp_lists[[m]] %>% group_by(gender, time) %>% summarize(xbar = mean(stl), se = sqrt(var(stl)/n()), n())
est_list[[m]]$imp = m
}

# stacking the datasets
combined_est <- do.call(rbind,est_list)
combined_est
```

```{r}
#Calculating error bars of combined estimates

combined_est_errors <- combined_est %>% group_by(gender, time) %>% summarize(x_bar_bar  = mean(xbar), B_m = var(xbar), W_m = mean(se^2), se_xbar_bar = sqrt(W_m+(1+1/5)*(B_m)), r = (1+1/5)*B_m/W_m, U = (5-1)*(1+(1/r)^2), t_val = qt(0.05/2, df = U, lower.tail = FALSE), uci = x_bar_bar + t_val * se_xbar_bar, lci = x_bar_bar - t_val * se_xbar_bar, t_se = t_val * se_xbar_bar)

combined_est_errors %>% select(gender, time,x_bar_bar, lci, uci)

```

```{r}
#ggplot of the data for

ggplot(data= combined_est_errors, aes(x = time, y = x_bar_bar, col = gender))+geom_point()+geom_line() + geom_errorbar(aes(ymin = lci, ymax = uci, group = gender),
     width = 0.2)+ labs(x = "Time", y = "Mean of imputed mean standard length values (STL)")

```

```{r}
#Doing it for cle variable this time
temp_comp_lists_cle <- list()
est_list_cle <- list()
M = 5
for(m in 1:M){

temp_comp_lists_cle[[m]] <- complete_stick_list[[m]][-c(1:367),]
temp_comp_lists_cle[[m]]$time <- times_vector

est_list_cle[[m]] <- temp_comp_lists_cle[[m]] %>% group_by(gender, time) %>% summarize(xbar = mean(cle), se = sqrt(var(cle)/n()), n())
est_list_cle[[m]]$imp = m
}

est_list_cle

combined_est_cle <- do.call(rbind,est_list_cle)

#Calculating error bars of combined estimates

combined_est_errors_cle <- combined_est_cle %>% group_by(gender, time) %>% summarize(x_bar_bar  = mean(xbar), B_m = var(xbar), W_m = mean(se^2), se_xbar_bar = sqrt(W_m+(1+1/5)*(B_m)), r = (1+1/5)*B_m/W_m, U = (5-1)*(1+(1/r)^2), t_val = qt(0.05/2, df = U, lower.tail = FALSE), uci = x_bar_bar + t_val * se_xbar_bar, lci = x_bar_bar - t_val * se_xbar_bar, t_se = t_val * se_xbar_bar)

combined_est_errors_cle %>% select(gender, time,x_bar_bar, lci, uci)

ggplot(data= combined_est_errors_cle, aes(x = time, y = x_bar_bar, col = gender))+geom_point()+geom_line() + geom_errorbar(aes(ymin = lci, ymax = uci),
     width = 0.2) + labs(x = "Time", y = "Mean of imputed mean CLE")

```

### 2023-04-04 SticklesEDA.Rmd
updates including plots

--- 2022-09-06 SticklesEDA.Rmd
@@
 #Loading Libraries
+
 ```{r}
 library(tidyverse)
-library(ggplot2)
-library(readr)
-library(dplyr)
-library(DT)
+library(janitor)
+library(tidymodels)
+library(furrr)
+library(doParallel)
 
@@
 ```{r}
-stickle <- read_csv('stickleback_fossil_vs_extant.csv')
+stickle <- read_csv('stickle_revamped.csv')
+ggplot2(x=as.integer(stick %>% filter(!is.na(gender)) %>% pull(gender)),stick %>% filter(!(is.na(gender))) %>% pull(ips.prop))
+
+gens <- stick %>% filter(!is.na(gender)) %>% pull(gender)
+props <- stick %>% filter(!(is.na(gender))) %>% pull(ips.prop)
+ggplot(data=stick, aes(x = gens,y=props)) + geom_bar()
 summary(stickle)
@@
 ```{r}
-datatable(stickle)
+#generating gender for the dataset
 
-#Should I split up univID?
-#Maybe, maybe not
-```
-```{r}
-stickles <- datatable(stickle)
-grepl("M",stickle$univID, fixed=TRUE)
-stickle$univID
 stick <- stickle %>% 
-  mutate(gender = ifelse(grepl("M",univID,fixed=TRUE),"M","F"))
+  mutate(gender = as.factor(case_when(
+    grepl(".M.",univID,fixed=TRUE) ~ "Male",
+    grepl(".F.",univID,fixed=TRUE) ~ "Female",
+    grepl("Fos",univID,fixed=TRUE) ~ NA_character_
+    )), ips = as.factor(ips)
+  ) %>% 
+  select(gender, everything())
 
-table(stick$gender)
-#Only 130 Males?????
-#sheesh
+stick_lakes_bl <- stick %>% 
+  mutate( B_L = case_when( 
+    grepl("Emi.",univID,fixed=TRUE) ~ "Emily Lake",
+    grepl("LQx.",univID,fixed=TRUE) ~ "Little Quarry Lake",
+    grepl("Pax.",univID,fixed=TRUE) ~ "Pax Lake",
+    grepl("Pri.",univID,fixed=TRUE) ~ "Pri Lake",
+    grepl("Joe.",univID,fixed=TRUE) ~ "Joe Lake",
+    grepl("Nor.",univID,fixed=TRUE) ~ "Nor Lake",
+    grepl("Swa.",univID,fixed=TRUE) ~ "Swa Lake",
+    grepl("The.",univID,fixed=TRUE) ~ "Theimer Lake",
+    grepl("Vib.",univID,fixed=TRUE) ~ "Village Bay Lake",
+    grepl("Fos.",univID,fixed=TRUE) ~ "Fossil"
+  ), Ecotype = case_when(
+    grepl(".B.",univID,fixed=TRUE) ~ "Benthic",
+    grepl(".G.",univID,fixed=TRUE) ~ "Generalist",
+    grepl(".L.",univID,fixed=TRUE) ~ "Limnetic",
+    grepl(".K.",univID,fixed=TRUE) ~ NA_character_
+
+  )
+
+  )
+
+stick %>% summary(ips)
+range(stick$ips)
+stick %>% filter(is.na(ect))
+
+#ect = bottom of fish, belly
 
 ```
+
+```{r}
+
+extant_stickle <- stick %>% filter(gender != "Fossil")
+
+```
+
+Visualizing the missing variables found within dataset.
+
+```{r}
+# install.packages('visdat')
+# install.packages('UpSetR')
+# install.packages('naniar')
+library(naniar)
+library(visdat)
+library(UpSetR)
+```
+
+how many missing obs per variable
+
+```{r}
+gg_miss_var(stick_lakes_bl)
+vis_miss(stick_lakes_bl)
+```
+
+```{r}
+gg_miss_var(stick_no_gender)
+vis_miss(stick_no_gender)
+
+```
+
+Are they often missing in conjunction with one another?
+
+```{r}
+n_var_miss(stick) # gives all variables with missing data
+gg_miss_upset(stick) # gives plot that captures combination of missingness in data
+gg_miss_upset(stick,nsets = n_var_miss(stick))
+gg_miss_upset(stick,nsets=2) # gives plot that captures combination of missingness in data
+
+#cle - shoulder bone buried or not there in fossils.
+# NA only, needs a cle for a functional fish
+
+#jawbon - pmx. is always there for a functional fish.
+
+```
+
+How about when excluding gender from the dataset?
+
+```{r}
+
+stick_no_gender <- stick %>% 
+  select(-gender, -univID)
+gg_miss_upset(stick_no_gender)
+gg_miss_upset(stick_no_gender,nsets = n_var_miss(stick_no_gender))
+
+```
+
+```{r}
+ggplot(stick,
+       aes(x = cle.sc,
+           y = cle)) +
+ geom_miss_point() + 
+ facet_wrap(~gender)
+
+```
+
+Might be concerning. Very prelim look, but first thoughts are that the fossil group seems to have cle.sc values between 5-6 that are missing, rest are not. more too look into later.
+
+```{r}
+stick_no_id <- stick %>% select(!univID)
+plot_bar(imputation.list = stick_imputed, var.name = "gender",
+    original.data = stick_no_id)
+```
+
+```{r}
+plot_box(imputation.list = stick_imputed, var.name = "ect",
+    original.data = stick_no_id)
+```
+
+```{r}
+
+```
+
+```{r}
+# devtools::install_github("agnesdeng/mixgb") #installs the mixgb package which utilizes xgboost as imputation method for MI
+stick_clean <- stick %>% 
+  select(!univID)
+
+stick_clean <- data_clean(stick_clean)
+```
+
+```{r}
+#Finding the best rounds, very basic CV. Still need to tune other params outside of gender. confused about the best rounds
+
+library(mixgb)
+
+cv.results <- mixgb_cv(data = stick_clean, verbose = FALSE)
+
+cv.results$response
+
+cv.results$best.nrounds
+
+cv.results2 <- mixgb_cv(data = stick_clean, nfold = 5, nrounds = 100,
+    early_stopping_rounds = 1, response = "gender", verbose = FALSE)
+
+cv.results2$best.nrounds
+```
+
+```{r}
+#trying to tune parameters, bit lost at this
+
+best_rounds <- c()
+for(i in 1:50){
+cv.results_ <- mixgb_cv(data = non_gender, verbose=FALSE,nrounds=200)
+best_rounds <-append(best_rounds, cv.results_$best.nrounds)
+if(best_rounds[i]==max(best_rounds)){
+  maxresponse = cv.results_$response
+  maxRounds <- cv.results_$best.nrounds
+}
+
+}
+
+max(best_rounds)
+maxresponse
+
+cv.results_ <- mixgb_cv(data = non_gender,response="stl", verbose=FALSE,nrounds=200)
+cv.results_$best.nrounds
+
+parallel::detectCores()
+n.cores <- parallel::detectCores() - 1
+
+my.cluster <- parallel::makeCluster(
+  n.cores, 
+  type = "FORK"
+  )
+doParallel::registerDoParallel(cl = my.cluster)
+foreach::getDoParRegistered()
+foreach::getDoParWorkers()
+
+best_rounds <- c()
+x <- foreach(
+  i = 1:1000, 
+  .combine = 'c'
+) %dopar% {
+  cv.results_ <- mixgb_cv(data = stick_clean, verbose=FALSE,nrounds=200)
+  best_rounds <-append(best_rounds, cv.results_$best.nrounds)
+}  
+best_rounds
+max(x)
+
+```
+
+```{r}
+#imputed dataset
+stick_imputed <- mixgb(data = stick_clean, m = 15, nrounds = 30)
+
+# imputed_data_cvresp_else <- mixgb(data = stick_clean, m = 20, nrounds = 75)
+# stick_imputed_e <- imputed_data_cvresp_else
+
+```
+
+```{r}
+#making a function to grab means cause I'm lazy
+
+grab_means <- function(data, i){
+  data[[i]] %>% 
+    summarize(across(where(is.numeric),mean))
+  }
+
+#grab_means(stick_imputed,1)
+
+imputed_means <- c(1:10) %>% 
+  future_map_dfr(grab_means,data=stick_imputed)
+
+imputed_means
+
+```
+
+```{r}
+
+#Comparing results to complete case dataset and non completed, non imputed values
+
+mean_extant_complete <- stick %>% 
+  na.omit %>% 
+  summarise(across(where(is.numeric),mean))
+mean_extant_complete
+```
+
+```{r}
+all_cores <- parallel::detectCores(logical = FALSE)
+library(doParallel)
+
+cl <- makePSOCKcluster(all_cores)
+registerDoParallel(cl)
+```
+
+```{r}
+
+set.seed(100)
+extant <- stick %>% 
+  drop_na(gender) %>%
+  select(-c(univID, gender))
+extant_gender <- stick %>% 
+  drop_na(gender) %>% 
+  pull(gender)
+
+cv.results_ <- mixgb_cv(data = extant, verbose=FALSE, nfold=10,nrounds=100)
+cv.results_$best.nrounds
+cv.results_$response
+
+extant_imputed <- mixgb(data = extant, m = 10, nrounds = 100)
+
+extant %>% mutate(gender = c(extant_gender)) %>% select(gender, everything())
+add_gender <- function(data,i){
+  data[[i]] %>% 
+    mutate(gender = c(extant_gender)) %>% select(gender, everything())
+
+}
+imp_extant_with_gender <- c(1:10) %>% 
+  map(data=extant_imputed,add_gender)
+
+```
+
+```{r}
+library(tidyverse)
+library(tidymodels)
+library(rsample)
+
+```
+
+```{r}
+
+#general accuracy with imputed datasets(only including extant)
+#maybe figure out a way to combine the results? idk
+
+getAccuracy  <- function(data,i){
+
+  split <- initial_split(data[[i]], prop = 0.7, strata = gender)
+  train <- training(split)
+  test <- testing(split)
+
+  train_fold <- vfold_cv(train, v = 10, strata = gender)
+
+  xg_spec <- boost_tree(trees = tune(),
+                        tree_depth = tune(), 
+                        min_n = tune(), 
+                        loss_reduction = tune(),
+                        sample_size = tune(), 
+                        mtry = tune(),
+                        learn_rate = tune()) %>% 
+    set_engine("xgboost") %>% 
+    set_mode("classification")
+
+  xg_rec <- recipe(gender ~., data = train) %>% 
+    step_dummy(all_nominal_predictors())
+
+  xg_wf <- workflow() %>% 
+    add_model(xg_spec) %>% 
+    add_recipe(xg_rec)
+  xg_grid <- grid_latin_hypercube(trees(),
+                                  tree_depth(),
+                                  min_n(),
+                                  loss_reduction(),
+                                  sample_size = sample_prop(),
+                                  finalize(mtry(), train),
+                                  learn_rate(),
+                                  size = 100)
+  xg_tune <- tune_grid(xg_wf,
+                       resamples = train_fold,
+                       grid = xg_grid,
+                       control = control_grid(save_pred = TRUE))
+
+  best_acc <- xg_tune %>% 
+    select_best("accuracy")
+
+  xg_final <- finalize_workflow(
+    xg_wf,
+    best_acc
+  )
+  xg_last <- last_fit(xg_final, split)
+
+   data_output <- xg_last
+  #   collect_predictions() %>% 
+  #   accuracy(gender, .pred_class)
+
+  return(data_output)
+
+}
+
+imp_extant_with_gender[[1]]
+```
+
+```{r}
+library(furrr)
+
+# list <- c(1:10) %>% 
+#   map(getAccuracy, data=imp_extant_with_gender)
+list_of_acc <- list()
+for(i in 1:10){
+  list_of_acc[[i]] <- getAccuracy(data=imp_extant_with_gender, i)
+}
+
+list2 <- list()
+list2[[1]] <- getAccuracy(data=imp_extant_with_gender,1)
+
+imputed_data_tuned <- mixgb(data = non_gender, m=5, nrounds = 140)
+c(1:10)
+
+getacc <- function(data,i){
+  data[[i]] %>% 
+  collect_predictions() %>% accuracy(gender,.pred_class)
+}
+accuracy_per_imputed_data <- c(1:10) %>% 
+  map_dfr(getacc,data=list)
+accuracy_per_imputed_data
+```
+
+```{r}
+library(mice)
+# install.packages("mitools")
+# install.packages('miceadds')
+library(mitools)
+library(miceadds)
+with(stick_imputed, mean)
+as.mids(stick_imputed)
+help(with)
+stick_imputed
+stick %>% 
+  grepl()
+stick_imputed
+all_numeric <- function(data,i){ 
+  data[[i]] %>% 
+    select(!c("gender","ips"))
+}
+
+test4 <- with(test3,expr=mean)
+test[[1]]
+stick_test <- imputationList(stick_imputed)
+#getting time variable
+univ <- stick %>% 
+  mutate(time = as.integer(sub(".*([0-9]{2})\\..*", "\\1",univID)))
+
+times <- univ %>% drop_na(time) %>% pull(time)
+
+times_vector <- c(rep(1,43)) %>% 
+  append(21-times)
+
+means<- function(data){
+  data %>% 
+    summarise(across(where(is.numeric),mean))
+}
+abc <- with(stick_test, fun=means)
+MIcombine(abc)
+mm<- as.mira(abc)
+
+summary(MIcombine(abc))
+
+add_time<- function(data){
+data %>% 
+  slice(368:1181) %>%
+  mutate(time = times_vector) %>% 
+  select(time, everything())
+}
+
+means_by_time <- function(data){
+  data %>% group_by(time, gender) %>% 
+  summarize(across(where(is.numeric),mean))
+}
+
+```
+
+```{r}
+mean_diff <- function(data){
+  data %>% group_by(time, gender) %>% 
+  summarize(across(where(is.numeric),mean)) %>% 
+  summarize(across(where(is.numeric),diff))
+
+}
+
+mean_diff2 <- function(data){
+  data %>% 
+  summarize(across(where(is.numeric),diff))
+
+}
+
+stick_time[[3]] %>% group_by(time,gender) %>% summarize(across(where(is.numeric),mean))
+
+#this is an issue
+#basically, sometimes the imputation doesn't have values for both sexes at a given time point
+
+stick_imputed <- mixgb(data = stick_clean, m = 10, nrounds = 15)
+
+stick_time <- with(imputationList(stick_imputed), fun=add_time)
+
+stick_means <- with(imputationList(stick_time), fun=means_by_time)
+mean_differences <- with(imputationList(stick_means[-c(2,4,8,9)]),fun=mean_diff2)
+mean_differences
+
+```
+
+```{r}
+fin <- (mean_differences[[1]]+mean_differences[[2]]+mean_differences[[3]]+mean_differences[[4]]+mean_differences[[5]]+mean_differences[[6]])/6
+
+```
+
+```{r}
+fin %>% ungroup() %>%
+  select(where(is.numeric)&!time) %>% colnames() %>% 
+  map( ~ {ggplot(fin, aes(x = time,y=.data[[.]])) + geom_line()})
+
+```
+
+Mean difference has gone up by a decent bit over the course of t=0 to t=21
+
+```{r}
+fin %>% ungroup() %>%
+  select(where(is.numeric)&!time) %>% colnames() %>% 
+  map( ~ {ggplot(fin, aes(x = time,y=.data[[.]])) + geom_line()})
+
+```

### 2023-04-04 mixgb_errors.Rmd
updates including plots

---
title: "mixgb_imputation"
author: "Akhil Ghosh"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r setup, include=FALSE}
library(mice)
library(tidyverse)
library(janitor)
library(missMethods)
library(ROCR)

```

```{r}
stickle <- read_csv('stickle_revamped.csv')

stick <- stickle %>% 
  mutate(gender = as.factor(case_when(
    grepl(".M.",univID,fixed=TRUE) ~ "Male",
    grepl(".F.",univID,fixed=TRUE) ~ "Female",
    grepl("Fos",univID,fixed=TRUE) ~ NA_character_
    )), ips = as.factor(ips)
  ) %>% 
  select(gender, everything())

stick_whole <- stick %>% select(-c(univID, ips))

extant_stickle <- stick %>% filter(gender != "Fossil") %>% select(!univID)
extant_stickle <- extant_stickle %>% select(!ips)

summary(extant_stickle)
```

```{r}
k = 5
M = 5

#generate 30% MCAR for gender on extant to find best hyperparams to train mixgb
#this is most likely data leakage of some form, but I might as well try things and then have Greg tell me I'm an idiot later
extant_stickle_MAR <- delete_MCAR(extant_stickle, 1, "gender")

mixgb_cv(extant_stickle, response = "gender")
#seems like best rounds is rarely, if ever, above 15-20.
#tracks with my previous forays into doing this

#basically, the issue I'm having is how should I "train" the mixgb model without data leakage.
#maybe since its used for imputation, and not prediction it really doesn't matter
#need to consult greg
```

```{r}
library(xgboost)
library(caret)
library(caTools)
library(tidymodels)

#how about we do a grid search to train a xgboost model to predict gender of extants, and use those hyperparameters in the mixgb model for imputation?
#worth a try

extant_stickle

set.seed(2023) # For reproducibility
split <- initial_split(extant_stickle, prop = 0.7)
train <- training(split)
test <- testing(split)
```

```{r}
#recipe to pre-process data

set.seed(2023)
split <- initial_split(extant_stickle, prop = 0.7, strata = gender)
train <- training(split)
test <- testing(split)
library(x)
dtrain <- xgb.DMatrix(as.matrix(train[,-1]), label = (as.numeric(train$gender)-1))
dtest <- xgb.DMatrix(as.matrix(test[, -1]), label = (as.numeric(test$gender)-1))
```

```{r}

params <- list(
  booster = "gbtree",
  objective = "binary:logistic",
  eta = 0.3,
  max_depth = 6,
  min_child_weight = 1,
  subsample = 1,
  colsample_bytree = 1,
  eval_metric = "error",
  n_thread=3
)

# Train the xgboost model
xgb_model <- xgb.train(
  params = params,
  data = dtrain,
  nrounds = 100,
  watchlist = list(train = dtrain, test = dtest),
  print_every_n = 10
)

# Make predictions on the test set
pred <- predict(xgb_model, dtest)

# Calculate the accuracy
table(test$gender)
accuracy <- mean((pred > 0.7) == (as.numeric(test$gender)-1))
cat("Accuracy:", accuracy, "\n")
```

```{r}

folds <- vfold_cv(train, v = 5, strata=gender)
```

```{r}
xg_spec <- boost_tree(trees = tune(),
                      tree_depth = tune(), 
                      min_n = tune(), 
                      loss_reduction = tune(),
                      sample_size = tune(), 
                      mtry = tune(),
                      learn_rate = tune()) %>% 
  set_engine("xgboost") %>% 
  set_mode("classification")

xg_rec <- recipe(gender ~., data = train) %>% 
  step_dummy(all_nominal_predictors()) %>% 
  step_impute_median(all_numeric_predictors())

xg_wf <- workflow() %>% 
  add_model(xg_spec) %>% 
  add_recipe(xg_rec)

# tune
set.seed(105)
xg_grid <- grid_latin_hypercube(trees(),
                                tree_depth(),
                                min_n(),
                                loss_reduction(),
                                sample_size = sample_prop(),
                                finalize(mtry(), train),
                                learn_rate(),
                                size = 50)

xg_tune <- tune_grid(xg_wf,
                     resamples = folds,
                     grid = xg_grid,
                     control = control_grid(save_pred = TRUE))
best_acc <- xg_tune %>% 
  select_best("accuracy")
```

```{r}
library(future)

rec <- recipe(gender ~ ., data = train) %>%
  step_dummy(all_nominal_predictors()) %>% 
  step_impute_median(all_numeric()) %>% 
  step_scale(all_numeric())

xg_spec <- boost_tree(trees = tune(),
                      tree_depth = tune(), 
                      min_n = tune(), 
                      loss_reduction = tune(),
                      sample_size = tune(), 
                      mtry = tune(),
                      learn_rate = tune()) %>%   set_engine("xgboost") %>% set_mode("classification")

xg_wf <- workflow() %>% 
  add_model(xg_spec) %>% 
  add_recipe(rec)

#grid search
xg_grid <- grid_latin_hypercube(trees(),
                                tree_depth(),
                                min_n(),
                                max_n(),
                                loss_reduction(),
                                sample_size = sample_prop(),
                                finalize(mtry(), train),
                                learn_rate(),
                                size = 50)

best_model <- select_best(xg_tune, metric = "accuracy")

```

```{r}
library(doParallel)
cl <- makeCluster(8)
registerDoParallel(cl)

xg_tune <- tune_grid(xg_wf,
                     resamples = folds,
                     grid = xg_grid,
                     control = control_grid(save_pred = TRUE, parallel_over = "everything", allow_par = T))

best_acc <- select_best(xg_tune, metric = "accuracy")
```

```{r}
xg_final <- finalize_workflow(
  xg_wf,
  best_acc
)

xg_final

xg_last <- last_fit(xg_final, split)

xg_last %>% 
  collect_predictions() %>% 
  accuracy(gender, .pred_class)

xg_last %>% 
  collect_predictions() %>% 
  roc_curve(gender, .pred_Female) %>% 
  ggplot(aes(1 - specificity, sensitivity)) +
  geom_abline() +
  geom_path() +
  coord_fixed()

xg_last %>% 
  collect_predictions() %>% 
  roc_auc(gender, .pred_Male)

xg_last %>% 
  collect_predictions() %>% 
  accuracy(gender, .pred_class)

```

```{r}
stickle_wf <- workflow() %>%
  add_recipe(recipe) %>%
  add_model(xgbtree())

```

```{r}
library(mixgb)

```

```{r}
#okay, onto the real thing here
#Looking to validate how imputation is doing against reality
M=5
k = 5
stick_matrix <- matrix(NA, nrow=nrow(extant_stickle), ncol = M)

set.seed(2023) # seed used for 86% was 12

fold <- sample(1:k, size = nrow(extant_stickle), replace=TRUE)
for(i in 1:k){
extant_stickle_temp <- extant_stickle
extant_stickle_temp$gender[fold==i] <- NA

#imputed_stick <- mixgb(extant_stickle_temp,nrounds = 100, m = 5, xgb.params = list(max_depth = 12, eta = 0.014311225124087, gamma = 0,min_child_weight = 1,subsample = 1, colsample_bytree = 1, colsample_bylevel = 1, colsample_bynode = 1,tree_method = "auto", gpu_id = 0, predictor = "auto"))

imputed_stick <- mixgb(extant_stickle_temp, nrounds = 100, m=5)
#all that tuning for it to be a worse imputer....
#whatever

for(n in 1:M){

  complete_stick <- imputed_stick[[n]]

  complete_stick$gender[fold==i]
  stick_matrix[fold==i, n] <- as.character(complete_stick$gender[fold==i])
}

}

pct_male <- apply(stick_matrix,1,function(x){
  mean(x == "Male")
})

imptest <- data_frame(truth = as.factor(extant_stickle$gender), pct_male)
roc_pred <- prediction(imptest$pct_male, imptest$truth)

auc_rocr <- performance(roc_pred, measure="auc")
plot(performance(roc_pred, measure="tpr", x.measure="fpr"))

auc_rocr@y.values[[1]]

```

```{r}
#Using MixGB imputation to get imputed datasets
M = 5
stick_whole <- stick %>% select(-c(univID, ips))
imputed_stick_whole <- mixgb(stick_whole,m=5, nrounds=100)
imputed_stick_whole[[1]]
```

```{r}
univ <- stick %>% 
  mutate(time = as.integer(sub(".*([0-9]{2})\\..*", "\\1",univID)))
times <- univ %>% drop_na(time) %>% pull(time)

times_vector <- c(rep(1,43)) %>% 
  append(21-times)
```

```{r}
#again issues with not having enough males :/
#what to do...

temp_comp_lists <- list()
est_list <- list()
M = 5
for(m in 1:M){

temp_comp_lists[[m]] <- imputed_stick_whole[[m]][-c(1:367),]
temp_comp_lists[[m]]$time <- times_vector

est_list[[m]] <- temp_comp_lists[[m]] %>% group_by(gender, time) %>% summarize(xbar = mean(stl), se = sqrt(var(stl)/n()), n())
est_list[[m]]$imp = m
}

# stacking the datasets
combined_est <- do.call(rbind,est_list)
combined_est
```

```{r}
#Calculating error bars of combined estimates

combined_est_errors <- combined_est %>% group_by(gender, time) %>% summarize(x_bar_bar  = mean(xbar), B_m = var(xbar), W_m = mean(se^2), se_xbar_bar = sqrt(W_m+(1+1/5)*(B_m)), r = (1+1/5)*B_m/W_m, U = (5-1)*(1+(1/r)^2), t_val = qt(0.05/2, df = U, lower.tail = FALSE), uci = x_bar_bar + t_val * se_xbar_bar, lci = x_bar_bar - t_val * se_xbar_bar, t_se = t_val * se_xbar_bar)

combined_est_errors %>% select(gender, time,x_bar_bar, lci, uci)

```

```{r}
ggplot(data= combined_est_errors, aes(x = time, y = x_bar_bar, col = gender))+geom_point()+geom_line() + geom_errorbar(aes(ymin = lci, ymax = uci, group = gender),
     width = 0.2)+ labs(x = "Time", y = "Mean of imputed mean standard length values (STL)")

```

### 2023-04-04 stickle_plots/server.R

@@ -1,0 +1,239 @@
+library(tidyverse)
+library(mixgb)
+library(mice)
+library(janitor)
+library(missMethods)
+library(ROCR)
+library(shiny)
+library(missForest)
+
+stickle <- read_csv('stickle_revamped.csv')
+
+stick <- stickle %>%
+  mutate(gender = as.factor(case_when(
+    grepl(".M.",univID,fixed=TRUE) ~ "Male",
+    grepl(".F.",univID,fixed=TRUE) ~ "Female",
+    grepl("Fos",univID,fixed=TRUE) ~ NA_character_
+  )), ips = as.factor(ips)
+  ) %>%
+  select(gender, everything())
+
+stick_whole <- stick %>% select(-c(univID, ips))
+
+extant_stickle <- stick %>% filter(gender != "Fossil") %>% select(!univID)
+extant_stickle <- extant_stickle %>% select(!ips)
+
+univ <- stick %>%
+  mutate(time = as.integer(sub(".*([0-9]{2})\\..*", "\\1",univID)))
+
+times <- univ %>% drop_na(time) %>% pull(time)
+
+
+
+times_vector <- c(rep(1,43)) %>%
+  append(21-times)
+
+
+imputed_stick_whole <- mice(stick_whole,m=5)
+
+completed_stick_lists <- list()
+M=5
+
+for(m in 1:M){
+  completed_stick_lists[[m]] <- complete(imputed_stick_whole, m)
+}
+
+
+
+imputed_means <- function(M,target_variable, completed_stick_lists){
+  target_variable <- enquo(target_variable)
+
+  temp_comp_lists <- list()
+  est_list <- list()
+  for(m in 1:M){
+
+    temp_comp_lists[[m]] <- completed_stick_lists[[m]][-c(1:367),]
+    temp_comp_lists[[m]]$time <- times_vector
+
+
+    est_list[[m]] <- temp_comp_lists[[m]] %>% group_by(gender, time) %>% summarize(xbar = mean(!! target_variable), se = sqrt(var(!! target_variable)/n()), n())
+    est_list[[m]]$imp = m
+  }
+
+
+  # stacking the datasets
+  combined_est <- do.call(rbind,est_list)
+  return(combined_est)
+}
+
+
+combined_se_function <- function(M, variable, completed_stick_lists) {
+  variable <- enquo(variable)
+  combined_estimate <- imputed_means(5, !!variable, completed_stick_lists)
+
[…]

### 2023-04-04 stickle_plots/ui.R

@@ -1,0 +1,52 @@
+
+
+library(shiny)
+library(shinythemes)
+
+shinyUI(navbarPage("Imputation Plots",
+                   theme = shinytheme("cerulean"),
+                   tabPanel("Imputation Results",
+                            sidebarLayout(
+                              sidebarPanel(
+                                radioButtons("variable", "Variable:",
+                                             c("Standard Length" = "stl",
+                                               "Interpelvic Spine Proportion" = "ips.prop",
+                                               "Lower Pelvic Spine" = "lps",
+                                               "Total Pectoral Girdle" = "tpg",
+                                               "Maximum Ventral Angle" = "mav",
+                                               "Maximum Caudal Angle" = "mcv",
+                                               "Maximum Dorsal Fin" = "mdf",
+                                               "Maximum Anal Fin" = "maf",
+                                               "Minimum Dorsal Spine" = "mds",
+                                               "Dorsal Spine 1" = "ds1",
+                                               "Dorsal Spine 2" = "ds2",
+                                               "Dorsal Spine 3" = "ds3",
+                                               "Midpoint Pectoral Fin" = "mpt",
+                                               "Lower Pectoral Fin" = "lpt",
+                                               "Ectocoracoid Length" = "ect",
+                                               "Cleithrum Length" = "cle",
+                                               "Premaxillary Length" = "pmx"),
+                                             inline = TRUE),
+
+                              ),
+
+                              mainPanel(
+                                tabsetPanel(
+                                  tabPanel("MICE Imputation", plotOutput("plot_mice")),
+                                  tabPanel("mixgb Imputation", plotOutput("plot_mixgb"))
+                                )
+                              )
+                            )
+                   ),
+
+                   tabPanel("Accuracy Plots",
+                            mainPanel(
+                              plotOutput("accuracy_plot_mice"),
+                              verbatimTextOutput("auc_mice"),
+                              plotOutput("accuracy_plot_mixgb"),
+                              verbatimTextOutput("auc_mixgb"),
+                              plotOutput("accuracy_plot"),
+                              verbatimTextOutput("auc_missForest")
+                            )
+                   )
+))

### 2023-04-17 stuart_stickle_mle.R
Update stuart_stickle_mle.R

@@ -5,21 +5,24 @@ library(bbmle)
 dat <- read.csv("stuart_stickle_imp_raw.csv")
 M <- max(dat$imputation_number)
 
-minusll <- function(theta_m,theta_f,kappa,sigma,tau){
+minusll <- function(theta_m,theta_f,kappa,sigma,tau,rho){ #m0_m,m0_f,theta_m,theta_f,kappa,sigma,tau,rho){
   y <- dt$stl
   t <- dt$time
   sex <- model.matrix(~ 0 + gender,dt)
-  mm <- (sex %*% c(theta_f,theta_m))
+  # mm <- (sex %*% c(m0_f,m0_m))*exp(-kappa*t) +
+  #       (sex %*% c(theta_f,theta_m))*(1 - exp(-kappa*t))
+  mm <- sex %*% c(theta_f,theta_m)
   vars <- diag(sigma^2,nrow=nrow(dt))
+  rho_mat <- matrix(c(1,rho,rho,1),nrow=2)
   time_diff <- abs(matrix(t,nrow=nrow(dt),ncol=nrow(dt),byrow=TRUE) -
                    matrix(t,nrow=nrow(dt),ncol=nrow(dt)))
-  covs <- tau^2/(1 - kappa^2)*kappa^time_diff
+  # time_abs  <- abs(matrix(t,nrow=nrow(dt),ncol=nrow(dt),byrow=TRUE) +
+  #                  matrix(t,nrow=nrow(dt),ncol=nrow(dt)))
+  # covs <- tau^2/(2*kappa)*(sex %*% rho_mat %*% t(sex))*
+  #  (exp(-kappa*time_diff) - exp(-kappa*time_abs))
+  covs <- tau^2/(1 - kappa^2)*(sex %*% rho_mat %*% t(sex))*kappa^time_diff
   vv <- vars + covs
   return(-dmvnorm(y,mean=mm,sigma=vv,log=TRUE) %>% sum)
-}
-
-minus_ll <- function(theta_m,theta_f,kappa_m,kappa_f,sigma_m,sigma_f,tau_m,tau_f){
-
 }
 
 final <- NULL
@@ -28,11 +31,13 @@ for(m in 1:M){
   print(paste0("Running Imputation ",m))
   dt <- dat %>% filter(imputation_number == m)
   test <- mle2(minusll,
-               start=list(theta_m=56,theta_f=53,
-                          kappa=0.5,
+               start=list(# m0_m=70,m0_f=69,
+                          theta_m=53,theta_f=50,
[…]

### 2023-05-01 manuscript/paper.Rmd
Added a manuscript and a library folder.

---
title: |
  |  \Large Stickle!
author: | 
  | \large Matthew Stuart \vspace{-1.1mm}
  | \normalsize Department of Mathematics and Statistics \vspace{-1mm}
  | \normalsize Loyola University Chicago \vspace{-1mm}
  | \normalsize Chicago, IL 60660 \vspace{-1mm}
  | \normalsize [`mstuart@luc.edu`](mailto:mstuart@luc.edu) \vspace{-1mm}
  |
  | \large Akhil Ghosh \vspace{-1.1mm}
  | \normalsize Department of Mathematics and Statistics \vspace{-1mm}
  | \normalsize Loyola University Chicago \vspace{-1mm}
  | \normalsize Chicago, IL 60660 \vspace{-1mm}
  | \normalsize [`aghosh@luc.edu`](mailto:aghosh@luc.edu) \vspace{-1mm}
  |
  | \large Yoel Stuart \vspace{-1.1mm}
  | \normalsize Department of Biology \vspace{-1mm}
  | \normalsize Loyola University Chicago \vspace{-1mm}
  | \normalsize Chicago, IL 60660 \vspace{-1mm}
  | \normalsize [`ystuart@luc.edu`](mailto:ystuart@luc.edu) \vspace{-1mm}
  |
  | \large Gregory J. Matthews \vspace{-1.1mm}
  | \normalsize Department of Mathematics and Statistics \vspace{-1mm}
  | \normalsize Loyola University Chicago \vspace{-1mm}
  | \normalsize Chicago, IL 60660 \vspace{-1mm}
  | \normalsize [`gmatthews1@luc.edu`](mailto:gmatthews1@luc.edu) \vspace{-1mm}
abstract: |
  | Evewryone loves the stickle \vspace{2mm}
  | *Keywords*: Stickle
bibliography: references.bib
fontsize: 12pt
link-citations: true
linkcolor: cyan
urlcolor: cyan
output:
  pdf_document:
    df_print: kable
    number_sections: true
    keep_tex: true
header-includes:
 \usepackage{setspace}
 \setstretch{1.15}
 \usepackage{float}
 \floatplacement{figure}{t}
---

```{r setup, echo=FALSE, message=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	fig.align = "center",
	message = FALSE,
	warning = FALSE
)

gjm <- function(x, color = "red") {
  sprintf("\\textcolor{%s}{%s}", color, x)}
```

```{r pkgs}
library(tidyverse)
theme_set(theme_minimal())
library(Lahman)
```

\newpage

# Introduction {#sec:intro}

# Exploratory Analysis

# Models 
All models were built using @R2022language

# Results

# Future work and conclusions

# Acknowledgements {-}

Stickle!

# Supplementary Material {-}

All code for reproducing the analyses in this paper is publicly available at https://github.com/Akhil-Ghosh/SticklebackProject

# References

### 2023-05-01 manuscript/paper.tex
Added a manuscript and a library folder.

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  12pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{setspace} \setstretch{1.15} \usepackage{float} \floatplacement{figure}{t}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  colorlinks=true,
  linkcolor={cyan},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={cyan},
  pdfcreator={LaTeX via pandoc}}

\title{~\Large Stickle!}
\author{\large Matthew Stuart \vspace{-1.1mm}\\
\normalsize Department of Mathematics and Statistics \vspace{-1mm}\\
\normalsize Loyola University Chicago \vspace{-1mm}\\
\normalsize Chicago, IL 60660 \vspace{-1mm}\\
\normalsize \href{mailto:mstuart@luc.edu}{\texttt{mstuart@luc.edu}}
\vspace{-1mm}\\
\strut \\
\large Akhil Ghosh \vspace{-1.1mm}\\
\normalsize Department of Mathematics and Statistics \vspace{-1mm}\\
\normalsize Loyola University Chicago \vspace{-1mm}\\
\normalsize Chicago, IL 60660 \vspace{-1mm}\\
\normalsize \href{mailto:aghosh@luc.edu}{\texttt{aghosh@luc.edu}}
\vspace{-1mm}\\
\strut \\
\large Yoel Stuart \vspace{-1.1mm}\\
\normalsize Department of Biology \vspace{-1mm}\\
\normalsize Loyola University Chicago \vspace{-1mm}\\
\normalsize Chicago, IL 60660 \vspace{-1mm}\\
\normalsize \href{mailto:ystuart@luc.edu}{\texttt{ystuart@luc.edu}}
\vspace{-1mm}\\
\strut \\
\large Gregory J. Matthews \vspace{-1.1mm}\\
\normalsize Department of Mathematics and Statistics \vspace{-1mm}\\
\normalsize Loyola University Chicago \vspace{-1mm}\\
\normalsize Chicago, IL 60660 \vspace{-1mm}\\
\normalsize \href{mailto:gmatthews1@luc.edu}{\texttt{gmatthews1@luc.edu}}
\vspace{-1mm}}
\date{}

\begin{document}
\maketitle
\begin{abstract}
Evewryone loves the stickle \vspace{2mm}\\
\emph{Keywords}: Stickle
\end{abstract}

\newpage

\hypertarget{sec:intro}{%
\section{Introduction}\label{sec:intro}}

\hypertarget{models}{%
\section{Models}\label{models}}

All models were built using R Core Team
(\protect\hyperlink{ref-R2022language}{2022})

\hypertarget{results}{%
\section{Results}\label{results}}

\hypertarget{future-work-and-conclusions}{%
\section{Future work and
conclusions}\label{future-work-and-conclusions}}

\hypertarget{acknowledgements}{%
\section*{Acknowledgements}\label{acknowledgements}}
\addcontentsline{toc}{section}{Acknowledgements}

Stickle!

\hypertarget{supplementary-material}{%
\section*{Supplementary Material}\label{supplementary-material}}
\addcontentsline{toc}{section}{Supplementary Material}

All code for reproducing the analyses in this paper is publicly
available at \url{https://github.com/Akhil-Ghosh/SticklebackProject}

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-R2022language}{}}%
R Core Team. 2022. \emph{R: A Language and Environment for Statistical
Computing}. Vienna, Austria: R Foundation for Statistical Computing.
\url{https://www.R-project.org/}.

\end{CSLReferences}

\end{document}

### 2023-05-01 manuscript/paper.Rmd
I added the manuscript in .rmd format.

@@
 
-# Exploratory Analysis
+# Data
+
+```{r}
+stickle <- read_csv('stickle_revamped.csv')
+
+stick <- stickle %>% 
+  mutate(gender = as.factor(case_when(
+    grepl(".M.",univID,fixed=TRUE) ~ "Male",
+    grepl(".F.",univID,fixed=TRUE) ~ "Female",
+    grepl("Fos",univID,fixed=TRUE) ~ NA_character_
+    )), ips = as.factor(ips)
+  ) %>% 
+  select(gender, everything())
+
+#getting time vector, adding it to imputed data
+univ <- stick %>% 
+  mutate(time = as.integer(sub(".*([0-9]{2})\\..*", "\\1",univID)))
+times <- univ %>% drop_na(time) %>% pull(time)
+
+times_vector <- c(rep(1,43)) %>% 
+  append(21-times)
+
+#fossil n
+fortab <- data.frame(time = 1:18, count = as.character(as.numeric((table(times_vector)))))
+#knitr::kable(fortab)
+#print(xtable(fortab),include.rownames = FALSE)
+```
+
+The data used here consists of a total of `r sum(!is.na(stick$gender))` extant specimens with known sex.
+Of these, there are `r sum(stick$gender == "Female",na.rm = TRUE)` and `r sum(stick$gender == "Male",na.rm = TRUE)` female and male specimens, respectively.
+
+In addition there are `r sum(is.na(stick$gender))` fossil specimens with unknown sex over 18 time periods.
+Table 1 shows the sample size at each of the 18 time periods.
+There are at least 22 specimens at each time period with a high of 67 specimens in period 7.
+
+<!-- % latex table generated in R 4.2.2 by xtable 1.8-4 package -->
+<!-- % Mon May  1 11:20:28 2023 -->
+\begin{table}[ht]
+\centering
+\begin{tabular}{rl}
+  \hline
+time & count \\ 
+  \hline
+  1 & 43 \\ 
+    2 & 41 \\ 
+    3 & 51 \\ 
+    4 & 41 \\ 
+    5 & 46 \\ 
+    6 & 48 \\ 
+    7 & 67 \\ 
+    8 & 55 \\ 
+    9 & 42 \\ 
+   10 & 33 \\ 
+   11 & 37 \\ 
+   12 & 22 \\ 
+   13 & 41 \\ 
+   14 & 43 \\ 
+   15 & 46 \\ 
+   16 & 47 \\ 
+   17 & 56 \\ 
+   18 & 55 \\ 
+   \hline
+\end{tabular}
+<!-- \label{tab1} -->
+\caption{The number of stickleback fossils at each time point.  Sample size at each time point ranges from a low of 22 to a high of 67.}
+\end{table}
 

### 2023-05-01 R/stickle_imputation.Rmd
Add folders for R code and Data
Created two new folders.  R, for R code, and Data, for data.

--- 2023-01-23 stickle_imputation_error_bars.Rmd
@@
 library(ROCR)
-
 ```
@@
 #Checking imputation accuracy/viability
-
 M = 5

### 2023-05-01 manuscript/paper.Rmd
Added a manuscript.

@@
 # Introduction {#sec:intro}
+Sexual Dimorphism is interesting because......
 
-# Data
+Sexual Dimoprhism lit review:
+@SaittaEtAl2020
+
+Studying sexual dimorphism of phenotypes in modern species is a straightforward endeavor: measure a phenotype of interest and statistically test for differences between the sexes.
+However, examining sexual dimorphism in fossils is complicated substantially by the issue that sex may not be able to be directly observed from fossil specimens.
+For this study, we have two sources of data: 1) Modern stickleback specimens with observed sex and 2) fossil specimens with sex unobserved.
+We view the sex of the fossils as missing data and use multiple imputation (@little2002statistical) to impute the sex of the fossils using the modern stickleback fish with observed sex to model the relationship between sex and observed phenotypes.
+Once sex is imputed, an Ornstein–Uhlenbeck (OU) model is fit using a Bayesian framework to look for sexual dimorphism across a variety of stickleback phenotypes (e.g.
+length, vertebrae count?, etc).
+
+The remainder of this manuscript contains a description of the data in section \ref{sec:data} and a description of our models in \ref{sec:models}.
+Section \ref{sec:results} presents a summary of our results, and we end with our conclusion and future work in section \ref{sec:conclusions}.
+
+# Data {#sec:data}
 
@@
 
-The data used here consists of a total of `r sum(!is.na(stick$gender))` extant specimens with known sex.
+The data used here consists of a total of `r sum(!is.na(stick$gender))` extant specimens with known sex all collected in the last 30 years.
 Of these, there are `r sum(stick$gender == "Female",na.rm = TRUE)` and `r sum(stick$gender == "Male",na.rm = TRUE)` female and male specimens, respectively.
 
-In addition there are `r sum(is.na(stick$gender))` fossil specimens with unknown sex over 18 time periods.
+In addition there are `r sum(is.na(stick$gender))` fossil specimens from approximately 10.3 million years ago with unknown sex over 18 time periods spaced about 1000 years apart.
 Table 1 shows the sample size at each of the 18 time periods.
@@
 
-# Models 
+What covariates do we have in the data: length, what else,
+
+# Models  {#sec:models}
+## Imputation model 
+Let $W$ be sex and $X$ are covariates.  This data set up here is $n_{fossil} + n_{modern}$.  
+$W = (W_{obs},W_{mis})$ where $W_{obs}$ and $W_{mis}$ are the observed and missing parts of the covariate sex.
+In this setting, $W_{obs}$ perfectly corresponds to the sex of the modern specimens, and $W_{mis}$ perfectly corresponds to the fossil data.
+Missing values of sex are imputed by drawing from the posterior predictive distribution $P(W_{mis}|W_{obs}, X)$.
+Multiple imputation was implemented here using MICE (CITE) with predictive mean matching.
+The data were imputed M = 100 times.
+
+## OU model 
+After imputing sex, we fit an OU model for different phenotypes of interest. We only look at $n_{fossil}$ fossil observations.  
+WLOG we choose one of the columns of $X$ to be the target ophenotype and our analysis is repeated for all phenotypes of interest.
+
+Let $X_{gti}$ be a phenotype of interest for the $i$-th observation $i = 1\ldots,n_{tg}$, in sex $g = 1, 2$ and $t = 1, \ldots, 18$.
+
+$$
+X_{gti} \sim N(\mu_{gt}, \sigma^2)\\
+
+\mu_{gt} \sim \kappa \mu_{g(t-1)} + (1-\kappa)\theta_g + \nu_{gt}\\
+
+\nu_{gt} \sim N(0,\tau^2)\\
+$$
+Priors: 
+$$
+\sigma^2 \sim Scaled-Inv \chi^2(1)\\
+\tau^2 \sim Scaled-Inv \chi^2(1)\\
+
+\mu_{g0} \sim N(0,10000000000000), g = 1,2\\
+
+\kappa \sim N(0,10000000000)\\
+\theta \sim N(0,10000000000) \\
+
+$$
+
 All models were built using @R2022language
 
-# Results
+@Cornault2022 Bayesian OU model.
 
-# Future work and conclusions
+Bayesian Analysis after multiple imputation @ZhouReiter2009: They recommend using a large number of imputrations.  5 or 10 is too small.  We are using M = 100.
+
+# Results {#sec:results}
+
+# Future work and conclusions {#sec:conclusions}
 

### 2023-05-01 R/Stickle_plots.Rmd

---
title: "stickle_plots"
author: "Akhil Ghosh"
date: "2022-10-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(mice)
library(tidyverse)
library(tidymodels)

library(mitools)
library(miceadds)

library(janitor)
library(furrr)
library(doParallel)
library(mixgb)
```

```{r}
stickle <- read_csv('stickle_revamped.csv')

stick <- stickle %>% 
  mutate(gender = as.factor(case_when(
    grepl(".M.",univID,fixed=TRUE) ~ "Male",
    grepl(".F.",univID,fixed=TRUE) ~ "Female",
    grepl("Fos",univID,fixed=TRUE) ~ NA_character_
    )), ips = as.factor(ips)
  ) %>% 
  select(gender, everything())
```

```{r}
set.seed(205)

stick_clean <- stick %>% 
  select(!c(univID, ips))

stick_clean <- data_clean(stick_clean)
stick_imputed <- mixgb(data = stick_clean, m = 10, nrounds = 20)
stick_imputed
```

```{r}
mice(stick_clean, )
as <- mice(stick_clean, method = "pmm")
as <- mice(stick_clean, method = "rf")
as$imp

help(datalist2mids)
```

```{r}
#getting time variable
datalist2mids(stick_$imputations)
univ <- stick %>% 
  mutate(time = as.integer(sub(".*([0-9]{2})\\..*", "\\1",univID)))
times <- univ %>% drop_na(time) %>% pull(time)

times_vector <- c(rep(1,43)) %>% 
  append(21-times)

```

```{r}
add_time<- function(data){
  data %>% 
  slice(368:1181) %>%
  mutate(time = times_vector) %>% 
  select(time, everything())
}

means_by_time <- function(data){
  data %>% group_by(time, gender) %>% 
  summarize(across(where(is.numeric),mean))
}

mean_diff <- function(data){
  data %>% group_by(time, gender) %>% 
  summarize(across(where(is.numeric),mean)) %>% 
  summarize(across(where(is.numeric),diff))

}

mean_diff2 <- function(data){
  data %>% 
  summarize(across(where(is.numeric),diff))
}
```

```{r}
set.seed(205)
stick_ <- imputationList(stick_imputed)

final <- with(imputationList(with(imputationList(stick_imputed), fun=add_time)),fun=means_by_time)

mean_differences <- with(imputationList(final[-c(6:10)]),fun=mean_diff2)
mean_differences

fin <- (mean_differences[[1]]+mean_differences[[2]]+mean_differences[[3]]+mean_differences[[4]]+mean_differences[[5]])/5
fin

fin %>% ungroup() %>%
  select(where(is.numeric)&!time) %>% colnames() %>% 
  map( ~ {ggplot(fin, aes(x = time,y=.data[[.]])) + geom_line()})

```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

### 2023-05-01 R/rf_imputation.Rmd

---
title: "mice_rf"
author: "Akhil Ghosh"
output: html_document
---

```{r setup, include=FALSE}
library(mice)
library(tidyverse)
library(janitor)
library(missMethods)
library(ROCR)
```

```{r setup, include=FALSE}
stickle <- read_csv('stickle_revamped.csv')

stick <- stickle %>% 
  mutate(gender = as.factor(case_when(
    grepl(".M.",univID,fixed=TRUE) ~ "Male",
    grepl(".F.",univID,fixed=TRUE) ~ "Female",
    grepl("Fos",univID,fixed=TRUE) ~ NA_character_
    )), ips = as.factor(ips)
  ) %>% 
  select(gender, everything())

stick_whole <- stick %>% select(-c(univID, ips))

extant_stickle <- stick %>% filter(gender != "Fossil") %>% select(!univID)
extant_stickle <- extant_stickle %>% select(!ips)

mice(extant_stickle, method="rf")

```

```{r}
library(mice)

```

```{r}
library(missForest)
summary(extant_stickle)
test <- extant_stickle
missing_cols <- c(2:7,9:27)
extant_stickle_MAR <- delete_MCAR(extant_stickle, 0.3, "gender")
imputed_data <- missForest(extant_stickle_MAR)
test[, missing_cols] <- imputed_data$ximp[, missing_cols]
```

```{r}
extant_stickle_df <- as.data.frame(extant_stickle)
extant_stickle_impo <- missForest(extant_stickle_df)

new_data <- extant_stickle_impo$ximp
new_data

doParallel::registerDoParallel(cores = 8) # set based on number of CPU cores
doRNG::registerDoRNG(seed = 2023)
new_data <- missForest(extant_stickle_df, parallelize = 'forests')$ximp

```

```{r}
M = 5
k = 5
mice(extant_stickle, method = 'rf', rfpackage = "randomForest")

stick_matrix <- matrix(NA, nrow=nrow(extant_stickle), ncol = M)
set.seed(2023)
fold <- sample(1:k, size = nrow(extant_stickle), replace=TRUE)
for(i in 1:k){
extant_stickle_temp <- extant_stickle
extant_stickle_temp$gender[fold==i] <- NA
extant_stickle_temp_df <- as.data.frame(extant_stickle_temp)
doParallel::registerDoParallel(cores = 8) # set based on number of CPU cores
doRNG::registerDoRNG(seed = 2023)

imputed_stick <- missForest(extant_stickle_temp_df, parallelize = 'variables')$ximp

for(n in 1:M){

  imputed_stick$gender[fold==i]
  stick_matrix[fold==i, n] <- as.character(imputed_stick$gender[fold==i])
}

}

pct_male <- apply(stick_matrix,1,function(x){
  mean(x == "Male")
})

imptest <- data_frame(truth = as.factor(extant_stickle$gender), pct_male)
roc_pred <- prediction(imptest$pct_male, imptest$truth)

auc_rocr <- performance(roc_pred, measure="auc")
plot(performance(roc_pred, measure="tpr", x.measure="fpr"))

auc_rocr@y.values[[1]]
```

```{r}
mice(stick, method='rf')
```

### 2023-05-01 R/stickle_imputation_error_bars.Rmd

--- 2023-01-23 stickle_imputation_error_bars.Rmd
@@
 
-stick
+stick_whole <- stick %>% select(-c(univID, ips))
 
@@
 #Checking imputation accuracy/viability
-
-M = 5
-k = 5
-stick_matrix <- matrix(NA, nrow=nrow(extant_stickle), ncol = M)
-set.seed(12)
-fold <- sample(1:k, size = nrow(extant_stickle), replace=TRUE)
-for(i in 1:k){
-extant_stickle_temp <- extant_stickle
-extant_stickle_temp$gender[fold==i] <- NA
-imputed_stick <- mice(extant_stickle_temp,m=M)
-for(n in 1:M){
-
-  complete_stick <- complete(imputed_stick, n)
-
-  complete_stick$gender[fold==i]
-  stick_matrix[fold==i, n] <- as.character(complete_stick$gender[fold==i])
-}
-
-}
-pct_male <- apply(stick_matrix,1,function(x){
-  mean(x == "Male")
-})
-
-imptest <- data_frame(truth = as.factor(extant_stickle$gender), pct_male)
-roc_pred <- prediction(imptest$pct_male, imptest$truth)
-
-auc_rocr <- performance(roc_pred, measure="auc")
-plot(performance(roc_pred, measure="tpr", x.measure="fpr"))
-
-auc_rocr@y.values[[1]]
 
@@
   mutate(time = as.integer(sub(".*([0-9]{2})\\..*", "\\1",univID)))
-times <- univ %>% drop_na(time) %>% pull(time)
+times <- univ %>% drop_na(time) %>% cut(time, breaks = 1,3,5,7,9,11,13,15,17) pull(time)
 
-times_vector <- c(rep(1,43)) %>% 
-  append(21-times)
+cut(times, breaks = 8, labels = c(8,7,6,5,4,3,2,1))
+times_vector <-length(rep(c(1,2,3,4,5,6,7,8,9),each=86))
+cut(times_vector, breaks=2, labels)
+
+help(cut)
 ```
@@
 
-combined_est_errors <- combined_est %>% group_by(gender, time) %>% summarize(x_bar_bar  = mean(xbar), B_m = var(xbar), W_m = mean(se^2), se_xbar_bar = sqrt(W_m+(1+1/5)*(B_m)), r = (1+1/5)*B_m/W_m, U = (5-1)*(1+(1/r)^2), t_val = qt(0.05/2, df = U, lower.tail = FALSE), uci = x_bar_bar + t_val * se_xbar_bar, lci = x_bar_bar - t_val * se_xbar_bar, t_se = t_val * se_xbar_bar)
+combined_est_errors <- combined_est %>% 
+  group_by(gender, time) %>% 
+  summarize(x_bar_bar  = mean(xbar), 
+            B_m = var(xbar), 
+            W_m = mean(se^2), 
+            se_xbar_bar = sqrt(W_m+(1+1/5)*(B_m)), 
+            r = (1+1/5)*B_m/W_m, 
+            U = (5-1)*(1+(1/r)^2), 
+            t_val = qt(0.05/2, df = U, lower.tail = FALSE), 
+            uci = x_bar_bar + t_val * se_xbar_bar, 
+            lci = x_bar_bar - t_val * se_xbar_bar, 
+            t_se = t_val * se_xbar_bar)
 
@@
 ```
+
+```{r}
+#Using MICE imputation to get imputed datasets (function)
+
+impute_stickledata <- function(M, dataset){
+  imputed_stick_ <- mice(dataset,m=M)
+  complete_stick_list <- list()
+  for(m in 1:M){
+    complete_stick_list[[m]] <- complete(imputed_stick_, m)
+  }
+  return(complete_stick_list)
+}
+
+```
+```{r}
+# My debugging process. Found out about enquo() and !!, which allows for inputting variables within a function.
+# pretty, pretty cool
+
+# test_ <- function(M,variable,data){
+#   variable <- enquo(variable)
+#   temp_comp_lists <- list()
+#   est_list <- list()
+#   for(m in 1:M){
+#   temp_comp_lists[[m]] <- data[[m]][-c(1:367),]
+#   temp_comp_lists[[m]]$time <- times_vector
+#   
+#   
+#   est_list[[m]] <- temp_comp_lists[[m]] %>% group_by(gender, time) %>% summarize(xbar = mean(!! variable), se = sqrt(var(!! variable)/n()), n())
+#   est_list[[m]]$imp = m
+#   }
+#   return(est_list)
+# }
+# 
+# testing_ <- impute_stickledata(5,stick_whole)
+# test_(5, stl, testing_)
+
+```
+
+```{r}
+completed_stick_lists <- impute_stickledata(5,stick_whole)
+#doing the grouping and summarizing for all imputed datasets function
+imputed_means <- function(M,target_variable){
+target_variable <- enquo(target_variable)
+
+temp_comp_lists <- list()
+est_list <- list()
+
+# completed_stick_lists <- impute_stickledata(M,stick_whole). I could leave this in here but don't necessarily need it
+for(m in 1:M){
+
+temp_comp_lists[[m]] <- completed_stick_lists[[m]][-c(1:367),]
+temp_comp_lists[[m]]$time <- times_vector
+
+est_list[[m]] <- temp_comp_lists[[m]] %>% group_by(gender, time) %>% summarize(xbar = mean(!! target_variable), se = sqrt(var(!! target_variable)/n()), n())
+est_list[[m]]$imp = m
+}
+
+# stacking the datasets
+combined_est <- do.call(rbind,est_list)
+return(combined_est)
+}
+
+```
+
+```{r}
+#Function for combined SE
+combined_se_function <- function(M, variable){
+variable <- enquo(variable)
+combined_estimate <- imputed_means(5, !! variable)
+
+combined_est_errors <- combined_estimate %>% group_by(gender, time) %>% summarize(x_bar_bar  = mean(xbar), B_m = var(xbar), W_m = mean(se^2), se_xbar_bar = sqrt(W_m+(1+1/5)*(B_m)), r = (1+1/5)*B_m/W_m, U = (5-1)*(1+(1/r)^2), t_val = qt(0.05/2, df = U, lower.tail = FALSE), uci = x_bar_bar + t_val * se_xbar_bar, lci = x_bar_bar - t_val * se_xbar_bar, t_se = t_val * se_xbar_bar)
+
+return(combined_est_errors)
+}
+
+combined_se_function(5, stl)
+
+```
+
+```{r}
+#function for ggplot of desired var
+stickle_imputation_plotting <- function(M, var){
+  v <- as.character(substitute(var))
+  var =enquo(var)
+  string = paste("mean of ", v)
+  combined_errors <- combined_se_function(5, !! var)
+
+  bb<- ggplot(data= combined_errors, aes(x = time, y = x_bar_bar, col = gender))+geom_point()+geom_line() +   geom_errorbar(aes(ymin = lci, ymax = uci, group = gender),
+     width = 0.2)+ labs(x = "Time", y = string)
+
+  return(bb)
+}
+stickle_imputation_plotting(5,stl)
+
+```
+
+```{r}
+
+```

### 2023-05-01 manuscript/paper.Rmd
Added my affiliation. CDSC!

@@
   | \large Gregory J. Matthews \vspace{-1.1mm}
-  | \normalsize Department of Mathematics and Statistics \vspace{-1mm}
+  | \normalsize Center for Data Science and Consulting; Department of Mathematics and Statistics \vspace{-1mm}
   | \normalsize Loyola University Chicago \vspace{-1mm}
@@
 $$
-X_{gti} \sim N(\mu_{gt}, \sigma^2)\\
-
-\mu_{gt} \sim \kappa \mu_{g(t-1)} + (1-\kappa)\theta_g + \nu_{gt}\\
-
-\nu_{gt} \sim N(0,\tau^2)\\
+X_{gti} \sim N(\mu_{gt}, \sigma^2)
+$$
+$$
+\mu_{gt} \sim \kappa \mu_{g(t-1)} + (1-\kappa)\theta_{g} + \nu_{gt}
+$$
+$$
+\nu_{gt} \sim N(0,\tau^2)
 $$
@@
 $$
-\sigma^2 \sim Scaled-Inv \chi^2(1)\\
-\tau^2 \sim Scaled-Inv \chi^2(1)\\
-
+\sigma^2 \sim Scaled-Inv \chi^2(1)
+$$
+$$
+\tau^2 \sim Scaled-Inv \chi^2(1)
+$$
+$$
 \mu_{g0} \sim N(0,10000000000000), g = 1,2\\
-
+$$
+$$
 \kappa \sim N(0,10000000000)\\
+$$
+$$
 \theta \sim N(0,10000000000) \\
-
 $$

### 2023-05-03 R/stuart_stickle_bayes.R
Update Bayes Model and Add Posterior Samples

@@ -1,60 +1,126 @@
+rm(list = ls())
 library(tidyverse)
+library(rstan)
+library(ggridges)
 
-dat <- read.csv("stuart_stickle_imp_100_raw.csv")
-# M <- max(dat$imputation_number)
-# tmp <- dat %>% filter(imputation_number == 1)
-T <- max(tmp$time)
-N <- nrow(tmp)
+dat <- read.csv("../Data/stuart_stickle_imp_100_raw.csv")
+dt <- dat %>%
+  group_by(imputation_number,time,gender) %>%
+  summarise(mu = mean(stl),
+            n = length(stl)) %>%
+  ungroup %>%
+  complete(imputation_number,time,gender,fill = list(mu = 0,n = 0))
 
-new_dat <- dat %>%
-  group_by(gender,index) %>%
-  summarise(n = n()) %>%
-  pivot_wider(names_from="gender",values_from = "n") %>%
-  replace(is.na(.),0) %>%
-  mutate(prop = Female/M) %>%
-  select(index,prop) %>%
-  right_join(dat %>% filter(imputation_number == 1)) %>%
-  arrange(index) %>%
-  select(-c(imputation_number,gender))
+mod <- "
+data{
+  int<lower = 0> N;
+  int<lower = 0> T;
+  real<lower = 0> y[N];
+  int<lower = 0,upper = 1> gender[N];
+  int<lower = 1,upper = T> time[N];
+}
+parameters{
+  real<lower=0> theta_f;
+  real<lower=0> theta_m;
+  real u_f[T];
+  real u_m[T];
+  real<lower=0,upper=1> kappa;
+  real<lower=0> sigma;
+  real<lower=0> tau;
+}
+transformed parameters{
+  real mu_f[T];
+  real mu_m[T];
+  for (t in 1:T){
+    mu_f[t] = theta_f + u_f[t];
+    mu_m[t] = theta_m + u_m[t];
+  }
+}
+model{
+  for (i in 1:N){
+    y[i] ~ normal(gender[i]*(mu_f[time[i]]) +
+                  (1 - gender[i])*(mu_m[time[i]]),sigma);
+  }
+  for (t in 2:T){
+    u_f[t] ~ normal(kappa*u_f[t-1],tau);
+    u_m[t] ~ normal(kappa*u_m[t-1],tau);
+  }
+  u_f[1] ~ normal(0,tau/sqrt(1-kappa*kappa));
[…]

### 2023-05-04 manuscript/paper.Rmd
Update Paper and Bayes sampling

@@
 
-Let $X_{gti}$ be a phenotype of interest for the $i$-th observation $i = 1\ldots,n_{tg}$, in sex $g = 1, 2$ and $t = 1, \ldots, 18$.
+Let $X_{gti}$ be a phenotype of interest for the $i$-th observation $i = 1\ldots,n_{gt}$, in sex $g = 1, 2$ and $t = 1, \ldots, 18$.
 
 $$
-X_{gti} \sim N(\mu_{gt}, \sigma^2)
+X_{gti} = \theta_g + u_{gt} + \epsilon_{gti}
 $$
 $$
-\mu_{gt} \sim \kappa \mu_{g(t-1)} + (1-\kappa)\theta_{g} + \nu_{gt}
+u_{gt} = \kappa u_{g(t-1)} + \nu_{gt}
 $$
 $$
-\nu_{gt} \sim N(0,\tau^2)
+u_{g1} \sim \mathcal{N}\left(0,\frac{\tau^2}{1-\kappa^2}\right)
 $$
+
+$$
+\nu_{gt} \sim \mathcal{N}(0,\tau^2)
+$$
+
+$$
+\epsilon_{gti} \sim \mathcal{N}(0,\sigma^2)
+$$
+
 Priors: 
 $$
-\sigma^2 \sim Scaled-Inv \chi^2(1)
+\sigma \sim \mathcal{N}(0,100)I_{\{\sigma > 0\}}
 $$
 $$
-\tau^2 \sim Scaled-Inv \chi^2(1)
+\tau \sim \mathcal{N}(0,100)I_{\{\tau > 0\}}
+$$
+
+$$
+\kappa \sim \mathcal{N}(0,25)I_{\{-1 < \kappa < 1\}} \\
 $$
 $$
-\mu_{g0} \sim N(0,10000000000000), g = 1,2\\
-$$
-$$
-\kappa \sim N(0,10000000000)\\
-$$
-$$
-\theta \sim N(0,10000000000) \\
+\theta_g \sim \mathcal{N}(54,625)I_{\{\theta_g > 0\}} \\
 $$
@@
 
-Bayesian Analysis after multiple imputation @ZhouReiter2009: They recommend using a large number of imputrations.  5 or 10 is too small.  We are using M = 100.
+Bayesian Analysis after multiple imputation @ZhouReiter2009: They recommend using a large number of imputations.  5 or 10 is too small.  We are using M = 100.
 

### 2023-05-04 manuscript/paper.tex
Update Paper and Bayes sampling

@@
 \large Gregory J. Matthews \vspace{-1.1mm}\\
-\normalsize Department of Mathematics and Statistics \vspace{-1mm}\\
+\normalsize Center for Data Science and Consulting; Department of
+Mathematics and Statistics \vspace{-1mm}\\
 \normalsize Loyola University Chicago \vspace{-1mm}\\
@@
 
-\hypertarget{models}{%
-\section{Models}\label{models}}
+Sexual Dimorphism is interesting because\ldots\ldots{}
+
+Sexual Dimoprhism lit review: Saitta et al.
+(\protect\hyperlink{ref-SaittaEtAl2020}{2020})
+
+Studying sexual dimorphism of phenotypes in modern species is a
+straightforward endeavor: measure a phenotype of interest and
+statistically test for differences between the sexes. However, examining
+sexual dimorphism in fossils is complicated substantially by the issue
+that sex may not be able to be directly observed from fossil specimens.
+For this study, we have two sources of data: 1) Modern stickleback
+specimens with observed sex and 2) fossil specimens with sex unobserved.
+We view the sex of the fossils as missing data and use multiple
+imputation (Little and Rubin
+(\protect\hyperlink{ref-little2002statistical}{2002})) to impute the sex
+of the fossils using the modern stickleback fish with observed sex to
+model the relationship between sex and observed phenotypes. Once sex is
+imputed, an Ornstein--Uhlenbeck (OU) model is fit using a Bayesian
+framework to look for sexual dimorphism across a variety of stickleback
+phenotypes (e.g.~length, vertebrae count?, etc).
+
+The remainder of this manuscript contains a description of the data in
+section \ref{sec:data} and a description of our models in
+\ref{sec:models}. Section \ref{sec:results} presents a summary of our
+results, and we end with our conclusion and future work in section
+\ref{sec:conclusions}.
+
+\hypertarget{sec:data}{%
+\section{Data}\label{sec:data}}
+
+The data used here consists of a total of 367 extant specimens with
+known sex all collected in the last 30 years. Of these, there are 202
+and 165 female and male specimens, respectively.
+
+In addition there are 814 fossil specimens from approximately 10.3
+million years ago with unknown sex over 18 time periods spaced about
+1000 years apart. Table 1 shows the sample size at each of the 18 time
+periods. There are at least 22 specimens at each time period with a high
+of 67 specimens in period 7.
+
+\begin{table}[ht]
+\centering
+\begin{tabular}{rl}
+  \hline
+time & count \\ 
+  \hline
+  1 & 43 \\ 
+    2 & 41 \\ 
+    3 & 51 \\ 
+    4 & 41 \\ 
+    5 & 46 \\ 
+    6 & 48 \\ 
+    7 & 67 \\ 
+    8 & 55 \\ 
+    9 & 42 \\ 
+   10 & 33 \\ 
+   11 & 37 \\ 
+   12 & 22 \\ 
+   13 & 41 \\ 
+   14 & 43 \\ 
+   15 & 46 \\ 
+   16 & 47 \\ 
+   17 & 56 \\ 
+   18 & 55 \\ 
+   \hline
+\end{tabular}
+<!-- \label{tab1} -->
+\caption{The number of stickleback fossils at each time point.  Sample size at each time point ranges from a low of 22 to a high of 67.}
+\end{table}
+
+What covariates do we have in the data: length, what else,
+
+\hypertarget{sec:models}{%
+\section{Models}\label{sec:models}}
+
+\hypertarget{imputation-model}{%
+\subsection{Imputation model}\label{imputation-model}}
+
+Let \(W\) be sex and \(X\) are covariates. This data set up here is
+\(n_{fossil} + n_{modern}\).\\
+\(W = (W_{obs},W_{mis})\) where \(W_{obs}\) and \(W_{mis}\) are the
+observed and missing parts of the covariate sex. In this setting,
+\(W_{obs}\) perfectly corresponds to the sex of the modern specimens,
+and \(W_{mis}\) perfectly corresponds to the fossil data. Missing values
+of sex are imputed by drawing from the posterior predictive distribution
+\(P(W_{mis}|W_{obs}, X)\). Multiple imputation was implemented here
+using MICE (CITE) with predictive mean matching. The data were imputed M
+= 100 times.
+
+\hypertarget{ou-model}{%
+\subsection{OU model}\label{ou-model}}
+
+After imputing sex, we fit an OU model for different phenotypes of
+interest. We only look at \(n_{fossil}\) fossil observations.\\
+WLOG we choose one of the columns of \(X\) to be the target ophenotype
+and our analysis is repeated for all phenotypes of interest.
+
+Let \(X_{gti}\) be a phenotype of interest for the \(i\)-th observation
+\(i = 1\ldots,n_{gt}\), in sex \(g = 1, 2\) and \(t = 1, \ldots, 18\).
+
+\[
+X_{gti} = \theta_g + u_{gt} + \epsilon_{gti}
+\] \[
+u_{gt} = \kappa u_{g(t-1)} + \nu_{gt}
+\] \[
+u_{g1} \sim \mathcal{N}\left(0,\frac{\tau^2}{1-\kappa^2}\right)
+\]
+
+\[
+\nu_{gt} \sim \mathcal{N}(0,\tau^2)
+\]
+
+\[
+\epsilon_{gti} \sim \mathcal{N}(0,\sigma^2)
+\]
+
+Priors: \[
+\sigma \sim \mathcal{N}(0,2)I_{\sigma > 0}
+\] \[
+\tau \sim \mathcal{N}(0,2)I_{\tau > 0}
+\]
+
+\[
+\kappa \sim N(0.5,1)\\
+\] \[
+\theta_g \sim N(54,400) \\
+\]
 
@@
 
-\hypertarget{results}{%
-\section{Results}\label{results}}
+Cornuault (\protect\hyperlink{ref-Cornault2022}{2022}) Bayesian OU
+model.
 
-\hypertarget{future-work-and-conclusions}{%
-\section{Future work and
-conclusions}\label{future-work-and-conclusions}}
+Bayesian Analysis after multiple imputation Zhou and Reiter
+(\protect\hyperlink{ref-ZhouReiter2009}{2010}): They recommend using a
+large number of imputations. 5 or 10 is too small. We are using M = 100.
+
+\hypertarget{sec:results}{%
+\section{Results}\label{sec:results}}
+
+\hypertarget{sec:conclusions}{%
+\section{Future work and conclusions}\label{sec:conclusions}}
 
@@
 \begin{CSLReferences}{1}{0}
+\leavevmode\vadjust pre{\hypertarget{ref-Cornault2022}{}}%
+Cornuault, Josselin. 2022. {``{Bayesian Analyses of Comparative Data
+with the Ornstein--Uhlenbeck Model: Potential Pitfalls}.''}
+\emph{Systematic Biology} 71 (6): 1524--40.
+\url{https://doi.org/10.1093/sysbio/syac036}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-little2002statistical}{}}%
+Little, R. J. A., and D. B. Rubin. 2002. \emph{Statistical Analysis with
+Missing Data}. Wiley Series in Probability and Mathematical Statistics.
+Probability and Mathematical Statistics. Wiley.
+\url{http://books.google.com/books?id=aYPwAAAAMAAJ}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-R2022language}{}}%
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-SaittaEtAl2020}{}}%
+Saitta, Evan T, Maximilian T Stockdale, Nicholas R Longrich, Vincent
+Bonhomme, Michael J Benton, Innes C Cuthill, and Peter J Makovicky.
+2020. {``{An effect size statistical framework for investigating sexual
+dimorphism in non-avian dinosaurs and other extinct taxa}.''}
+\emph{Biological Journal of the Linnean Society} 131 (2): 231--73.
+\url{https://doi.org/10.1093/biolinnean/blaa105}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-ZhouReiter2009}{}}%
+Zhou, X., and J. Reiter. 2010. {``A Note on Bayesian Inference After
+Multiple Imputation.''} \emph{The American Statistician} 64 (2):
+159--63.
+
 \end{CSLReferences}

### 2023-05-04 R/stuart_stickle_bayes_kappa.R

@@ -18,7 +18,8 @@ parameters{
   real<lower=0> theta_m;
   real u_f[T];
   real u_m[T];
-  real<lower=-1,upper=1> kappa;
+  real<lower=-1,upper=1> kappa_f;
+  real<lower=-1,upper=1> kappa_m;
   real<lower=0> sigma;
   real<lower=0> tau;
 }
@@ -36,15 +37,16 @@ model{
                   (1 - gender[i])*(mu_m[time[i]]),sigma);
   }
   for (t in 2:T){
-    u_f[t] ~ normal(kappa*u_f[t-1],tau);
-    u_m[t] ~ normal(kappa*u_m[t-1],tau);
+    u_f[t] ~ normal(kappa_f*u_f[t-1],tau);
+    u_m[t] ~ normal(kappa_m*u_m[t-1],tau);
   }
-  u_f[1] ~ normal(0,tau/sqrt(1-kappa*kappa));
-  u_m[1] ~ normal(0,tau/sqrt(1-kappa*kappa));
+  u_f[1] ~ normal(0,tau/sqrt(1-kappa_f*kappa_f));
+  u_m[1] ~ normal(0,tau/sqrt(1-kappa_m*kappa_m));
 
   theta_f ~ normal(54,20);
   theta_m ~ normal(54,20);
-  kappa ~ normal(0.5,1);
+  kappa_f ~ normal(0.5,1);
+  kappa_m ~ normal(0.5,1);
   sigma ~ normal(0,10);
   tau ~ normal(0,10);
 }
@@ -64,7 +66,7 @@ for (m in 1:M){
   fit <- sampling(sm,
                   list(N=N,T=T,y=y,gender=gender,time=time),
                   iter=10000,chains=1,warmup=5000,thin=5,
-                  pars = c("theta_f","theta_m","mu_f","mu_m","kappa","sigma","tau"))
+                  pars = c("theta_f","theta_m","mu_f","mu_m","kappa_f","kappa_m","sigma","tau"))
 
   if (m == 1){
     samps <- as.data.frame(fit)
@@ -73,4 +75,4 @@ for (m in 1:M){
   }
 }
 
-saveRDS(samps,"posterior_samples.RDS")
+saveRDS(samps,"posterior_samples_kappa.RDS")

### 2023-05-04 R/stuart_stickle_bayes_plots.R

@@ -1,0 +1,55 @@
+samps <- readRDS("posterior_samples.RDS")
+
+samps %>%
+  rename(Female = theta_f) %>%
+  rename(Male = theta_m) %>%
+  select(Female,Male) %>%
+  gather("Gender","Value") %>%
+  ggplot() +
+  geom_density(aes(x=Value,color=Gender)) +
+  theme_bw() +
+  xlim(c(30,80)) +
+  xlab(expression(theta)) +
+  ylab(expression(paste("p(",theta,"|",x,")")))
+
+length(which(samps$theta_m > samps$theta_f))/nrow(samps)
+
+mu_f <- samps %>% select(contains("mu_f"))
+mu_m <- samps %>% select(contains("mu_m"))
+
+mu_diff <- mu_m - mu_f
+
+data.frame(time = 1:18,
+           med = apply(mu_diff,2,quantile,probs=0.5),
+           low = apply(mu_diff,2,quantile,probs=0.025),
+           upp = apply(mu_diff,2,quantile,probs=0.975)) %>%
+  ggplot(aes(x=time,y=med)) +
+  geom_point() +
+  geom_errorbar(aes(ymin=low,ymax=upp),color = "red") +
+  theme_bw() +
+  xlab(expression(t)) +
+  ylab(expression(paste(mu[m],"-",mu[f])))
+
+names(mu_diff) <- c(1:18)
+as.data.frame(mu_diff) %>%
+  gather("Time","Value") %>%
+  mutate(Time = as.numeric(Time)) %>%
+  ggplot(aes(x=Value,y=Time,group=Time)) +
+  geom_density_ridges(fill="blue",alpha=0.5) +
+  xlim(c(-15,20)) +
+  scale_y_reverse() +
+  geom_vline(aes(xintercept = 0),colour = "red",linetype=2) +
+  xlab(expression(paste(mu[mt],"-",mu[ft]))) +
+  ylab(expression(t)) +
+  theme_bw()
+
+post_probs <- apply(mu_diff,2,function(x){length(which(x > 0))})/nrow(mu_diff)
+
+data.frame(time=1:18,probs=post_probs) %>%
+  ggplot() +
+  geom_line(aes(x=time,y=probs)) +
+  xlab(expression(t)) +
+  ylab(expression(paste(P,"(",mu[mt] > mu[ft],"|",X,")"))) +
+  theme_bw()
+
+

### 2023-05-05 R/stuart_stickle_bayes_kappa_plots.R
Add plots files

@@ -1,4 +1,6 @@
-samps <- readRDS("posterior_samples.RDS")
+library(tidyverse)
+library(ggridges)
+samps <- readRDS("posterior_samples_kappa.RDS")
 
 samps %>%
   rename(Female = theta_f) %>%
@@ -13,6 +15,19 @@ samps %>%
   ylab(expression(paste("p(",theta[g],"|",bold(X),")")))
 
 length(which(samps$theta_m > samps$theta_f))/nrow(samps)
+
+samps %>%
+  rename(Female = kappa_f) %>%
+  rename(Male = kappa_m) %>%
+  select(Female,Male) %>%
+  gather("Gender","Value") %>%
+  ggplot() +
+  geom_density(aes(x=Value,color=Gender)) +
+  theme_bw() +
+  xlab(expression(kappa[g])) +
+  ylab(expression(paste("p(",kappa[g],"|",bold(X),")")))
+
+length(which(samps$kappa_m > samps$kappa_f))/nrow(samps)
 
 mu_f <- samps %>% select(contains("mu_f"))
 mu_m <- samps %>% select(contains("mu_m"))
@@ -50,7 +65,7 @@ data.frame(time=1:18,probs=post_probs) %>%
   geom_line(aes(x=time,y=probs)) +
   geom_hline(aes(yintercept=0.5),color="red",linetype=2)+
   xlab(expression(t)) +
-  ylab(expression(paste(P,"(",mu[mt] > mu[ft],"|",bold(X),")"))) +
+  ylab(expression(paste(P,"(",mu[mt] > mu[ft],"|",X,")"))) +
   theme_bw()
 
 

### 2023-05-05 R/stuart_stickle_bayes_kappa_plots.R
Add Figures
Upload figures discussed in the Cinco de Mayo meeting.

@@ -9,10 +9,12 @@ samps %>%
   gather("Gender","Value") %>%
   ggplot() +
   geom_density(aes(x=Value,color=Gender)) +
+  scale_color_manual(values = c("red","blue"))+
   theme_bw() +
   xlim(c(30,80)) +
   xlab(expression(theta[g])) +
   ylab(expression(paste("p(",theta[g],"|",bold(X),")")))
+ggsave("../Figures/DifferentKappa/theta.pdf",width=5,height=8)
 
 length(which(samps$theta_m > samps$theta_f))/nrow(samps)
 
@@ -23,9 +25,11 @@ samps %>%
   gather("Gender","Value") %>%
   ggplot() +
   geom_density(aes(x=Value,color=Gender)) +
+  scale_color_manual(values = c("red","blue"))+
   theme_bw() +
   xlab(expression(kappa[g])) +
   ylab(expression(paste("p(",kappa[g],"|",bold(X),")")))
+ggsave("../Figures/DifferentKappa/kappa.pdf",width=5,height=8)
 
 length(which(samps$kappa_m > samps$kappa_f))/nrow(samps)
 
@@ -33,17 +37,6 @@ mu_f <- samps %>% select(contains("mu_f"))
 mu_m <- samps %>% select(contains("mu_m"))
 
 mu_diff <- mu_m - mu_f
-
-data.frame(time = 1:18,
-           med = apply(mu_diff,2,quantile,probs=0.5),
-           low = apply(mu_diff,2,quantile,probs=0.025),
-           upp = apply(mu_diff,2,quantile,probs=0.975)) %>%
-  ggplot(aes(x=time,y=med)) +
-  geom_point() +
-  geom_errorbar(aes(ymin=low,ymax=upp),color = "red") +
-  theme_bw() +
-  xlab(expression(t)) +
-  ylab(expression(paste(mu[m],"-",mu[f])))
 
 names(mu_diff) <- c(1:18)
 as.data.frame(mu_diff) %>%
@@ -57,15 +50,40 @@ as.data.frame(mu_diff) %>%
   xlab(expression(paste(mu[mt],"-",mu[ft]))) +
   ylab(expression(t)) +
   theme_bw()
+ggsave("../Figures/DifferentKappa/mu_diff.pdf",width=5,height=8)
 
[…]

### 2023-05-05 R/stuart_stickle_bayes_plots.R

@@ -1,3 +1,5 @@
+library(tidyverse)
+library(ggridges)
 samps <- readRDS("posterior_samples.RDS")
 
 samps %>%
@@ -7,10 +9,12 @@ samps %>%
   gather("Gender","Value") %>%
   ggplot() +
   geom_density(aes(x=Value,color=Gender)) +
+  scale_color_manual(values = c("red","blue"))+
   theme_bw() +
   xlim(c(30,80)) +
   xlab(expression(theta[g])) +
   ylab(expression(paste("p(",theta[g],"|",bold(X),")")))
+ggsave("../Figures/SameKappa/theta.pdf",width=5,height=8)
 
 length(which(samps$theta_m > samps$theta_f))/nrow(samps)
 
@@ -18,17 +22,6 @@ mu_f <- samps %>% select(contains("mu_f"))
 mu_m <- samps %>% select(contains("mu_m"))
 
 mu_diff <- mu_m - mu_f
-
-data.frame(time = 1:18,
-           med = apply(mu_diff,2,quantile,probs=0.5),
-           low = apply(mu_diff,2,quantile,probs=0.025),
-           upp = apply(mu_diff,2,quantile,probs=0.975)) %>%
-  ggplot(aes(x=time,y=med)) +
-  geom_point() +
-  geom_errorbar(aes(ymin=low,ymax=upp),color = "red") +
-  theme_bw() +
-  xlab(expression(t)) +
-  ylab(expression(paste(mu[m],"-",mu[f])))
 
 names(mu_diff) <- c(1:18)
 as.data.frame(mu_diff) %>%
@@ -42,15 +35,40 @@ as.data.frame(mu_diff) %>%
   xlab(expression(paste(mu[mt],"-",mu[ft]))) +
   ylab(expression(t)) +
   theme_bw()
+ggsave("../Figures/SameKappa/mu_diff.pdf",width=5,height=8)
 
-post_probs <- apply(mu_diff,2,function(x){length(which(x > 0))})/nrow(mu_diff)
-
-data.frame(time=1:18,probs=post_probs) %>%
-  ggplot() +
-  geom_line(aes(x=time,y=probs)) +
-  geom_hline(aes(yintercept=0.5),color="red",linetype=2)+
-  xlab(expression(t)) +
-  ylab(expression(paste(P,"(",mu[mt] > mu[ft],"|",bold(X),")"))) +
+names(mu_f) <- c(1:18)
[…]

### 2023-05-08 manuscript/paper.Rmd
Update Manuscript
Added some more description to our model specification.

@@
 ```
+
+\newcommand{\iid}{\overset{iid}{\sim}}
 
@@
 \end{tabular}
-<!-- \label{tab1} -->
+\label{tab1}
 \caption{The number of stickleback fossils at each time point.  Sample size at each time point ranges from a low of 22 to a high of 67.}
@@
 ## Imputation model 
-Let $W$ be sex and $X$ are covariates.  This data set up here is $n_{fossil} + n_{modern}$.  
-$W = (W_{obs},W_{mis})$ where $W_{obs}$ and $W_{mis}$ are the observed and missing parts of the covariate sex.
-In this setting, $W_{obs}$ perfectly corresponds to the sex of the modern specimens, and $W_{mis}$ perfectly corresponds to the fossil data.
-Missing values of sex are imputed by drawing from the posterior predictive distribution $P(W_{mis}|W_{obs}, X)$.
-Multiple imputation was implemented here using MICE (CITE) with predictive mean matching.
-The data were imputed M = 100 times.
+Let $\boldsymbol{W}$ be an $(n_{extant} + n_{fossil}) \times 1$ vector of the covariate gender of the stickleback fish and $\boldsymbol{X}$ be an $(n_{extant} + n_{fossil}) \times p$ matrix of the $p$ phenotypes of interest of the fish.
+Because $\boldsymbol{W}$ contains missing data, we further define $\boldsymbol{W} = (\boldsymbol{W}_{obs}^T,\boldsymbol{W}_{mis}^T)^T$ where $\boldsymbol{W}_{obs}$ and $\boldsymbol{W}_{mis}$ are the observed and missing data of the covariate gender, respectively.
+In this setting, $\boldsymbol{W}_{obs}$ is a $n_{extant} \times 1$ vector and $\boldsymbol{W}_{obs}$ is a $n_{fossil} \times 1$ vector, perfectly corresponding to the observable gender of the modern extants and the unobservable gender of the fossil data, respectively.
+
+We impute the missing gender for the fossil data by sampling from the posterior predictive distribution $P(\boldsymbol{W}_{mis}|\boldsymbol{W}_{obs}, \boldsymbol{X})$ using the multiple imputation with chained equations (MICE) algorithm (@MICE) with predictive mean matching.
+The imputation algorithm is run to obtain a total of $M = 100$ completed datasets.
 
@@
 After imputing sex, we fit an OU model for different phenotypes of interest. We only look at $n_{fossil}$ fossil observations.  
-WLOG we choose one of the columns of $X$ to be the target ophenotype and our analysis is repeated for all phenotypes of interest.
+WLOG we choose one of the columns of $X$ to be the target phenotype and our analysis is repeated for all phenotypes of interest.
 
-Let $X_{gti}$ be a phenotype of interest for the $i$-th observation $i = 1\ldots,n_{gt}$, in sex $g = 1, 2$ and $t = 1, \ldots, 18$.
+Let $W_{ti}$ and $X_{ti}$ be the gender and phenotype of interest for the $i$-th observation in year $t$ for $t = 1, \ldots, 18$ and  $i = 1\ldots,n_{t}$.  We assume that
 
 $$
-X_{gti} = \theta_g + u_{gt} + \epsilon_{gti}
+\begin{aligned}
+X_{ti} \iid \left\{\begin{array}{ll} \mathcal{N}(\mu_{ft},\sigma^2) & W_{ti} = \text{Female} \\ \mathcal{N}(\mu_{mt},\sigma^2) & W_{ti} = \text{Male} \end{array}\right.
+\end{aligned}
 $$
-$$
-u_{gt} = \kappa u_{g(t-1)} + \nu_{gt}
-$$
-$$
-u_{g1} \sim \mathcal{N}\left(0,\frac{\tau^2}{1-\kappa^2}\right)
-$$
+We then fit the time and gender dependent means, $\mu_{ft}$ and $\mu_{mt}$, to an autoregressive process of lag 1, which is the discretization of an OU process.
+For $t = 2,\ldots,T$ and $g \in \{f,m\}$, the AR1 model for the mean structure defined as is
 
 $$
-\nu_{gt} \sim \mathcal{N}(0,\tau^2)
+\begin{aligned}
+\mu_{gt} \iid \mathcal{N}(\kappa \mu_{g,t-1} + (1 - \kappa)\theta_g, \tau^2).
+\end{aligned}
 $$
+In this setting, $\theta_g, g \in \{f,m\}$ is the defined as the overall long-run mean of the phenotype of interest for each gender, and $\kappa$ represents how quickly over time $\mu_{gt}$ reverts back to $\theta_g$; the larger the value of $\kappa$, the slower $\mu_{gt}$ reverts back to $\theta_g$.
+
+We also set
 
 $$
-\epsilon_{gti} \sim \mathcal{N}(0,\sigma^2)
+\begin{aligned}
+\mu_{g1} \sim \mathcal{N}\left(0,\frac{\tau^2}{1-\kappa^2}\right).
+\end{aligned}
 $$
+This model choice is to preserve stationarity, i.e. $p(\mu_{gs}) = p(\mu_{gt})$ for $s,t \in \{1,\dots,T\}$.
 
@@
 $$
-\kappa \sim \mathcal{N}(0,25)I_{\{-1 < \kappa < 1\}} \\
+\kappa \sim \mathcal{N}(0,25)I_{\{0 < \kappa < 1\}} \\
 $$

### 2023-05-08 manuscript/paper.tex
Add Figures
Upload figures discussed in the Cinco de Mayo meeting.
Update Manuscript
Added some more description to our model specification.

@@
 
+\newcommand{\iid}{\overset{iid}{\sim}}
+
 \newpage
@@
 
-Let \(W\) be sex and \(X\) are covariates. This data set up here is
-\(n_{fossil} + n_{modern}\).\\
-\(W = (W_{obs},W_{mis})\) where \(W_{obs}\) and \(W_{mis}\) are the
-observed and missing parts of the covariate sex. In this setting,
-\(W_{obs}\) perfectly corresponds to the sex of the modern specimens,
-and \(W_{mis}\) perfectly corresponds to the fossil data. Missing values
-of sex are imputed by drawing from the posterior predictive distribution
-\(P(W_{mis}|W_{obs}, X)\). Multiple imputation was implemented here
-using MICE (CITE) with predictive mean matching. The data were imputed M
-= 100 times.
+Let \(\boldsymbol{W}\) be an \((n_{extant} + n_{fossil}) \times 1\)
+vector of the covariate gender of the stickleback fish and
+\(\boldsymbol{X}\) be an \((n_{extant} + n_{fossil}) \times p\) matrix
+of the \(p\) phenotypes of interest of the fish. Because
+\(\boldsymbol{W}\) contains missing data, we further define
+\(\boldsymbol{W} = (\boldsymbol{W}_{obs}^T,\boldsymbol{W}_{mis}^T)^T\)
+where \(\boldsymbol{W}_{obs}\) and \(\boldsymbol{W}_{mis}\) are the
+observed and missing data of the covariate gender, respectively. In this
+setting, \(\boldsymbol{W}_{obs}\) is a \(n_{extant} \times 1\) vector
+and \(\boldsymbol{W}_{obs}\) is a \(n_{fossil} \times 1\) vector,
+perfectly corresponding to the observable gender of the modern extants
+and the unobservable gender of the fossil, respectively.
+
+We impute the missing gender for the fossil data by sampling from the
+posterior predictive distribution
+\(P(\boldsymbol{W}_{mis}|\boldsymbol{W}_{obs}, \boldsymbol{X})\) using
+the multiple imputation with chained equations (MICE) algorithm (Buuren
+and Groothuis-Oudshoorn (\protect\hyperlink{ref-MICE}{2011})) with
+predictive mean matching. The imputation algorithm is run to obtain a
+total of \(M = 100\) completed datasets.
 
@@
 interest. We only look at \(n_{fossil}\) fossil observations.\\
-WLOG we choose one of the columns of \(X\) to be the target ophenotype
+WLOG we choose one of the columns of \(X\) to be the target phenotype
 and our analysis is repeated for all phenotypes of interest.
 
-Let \(X_{gti}\) be a phenotype of interest for the \(i\)-th observation
-\(i = 1\ldots,n_{gt}\), in sex \(g = 1, 2\) and \(t = 1, \ldots, 18\).
+Let \(W_{ti}\) and \(X_{ti}\) be the gender and phenotype of interest
+for the \(i\)-th observation in year \(t\) for \(t = 1, \ldots, 18\) and
+\(i = 1\ldots,n_{t}\). We assume that
 
 \[
-X_{gti} = \theta_g + u_{gt} + \epsilon_{gti}
+\begin{aligned}
+X_{ti} \overset{iid}{\sim}\left\{\begin{array}{ll} \mathcal{N}(\mu_{ft},\sigma^2) & W_{ti} = \text{Female} \\ \mathcal{N}(\mu_{mt},\sigma^2) & W_{ti} = \text{Male} \end{array}\right.
+\end{aligned}
+\] We then fit the time and gender dependent means, \(\mu_{ft}\) and
+\(\mu_{mt}\), to an autoregressive process of lag 1, which is the
+discretization of an OU process. For \(t = 2,\ldots,T\) and
+\(g \in \{f,m\}\), the AR1 model for the mean structure defined as is
+
+\[
+\begin{aligned}
+\mu_{gt} \overset{iid}{\sim}\mathcal{N}(\kappa \mu_{g,t-1} + (1 - \kappa)\theta_g, \tau^2).
+\end{aligned}
+\] In this setting, \(\theta_g, g \in \{f,m\}\) is the defined as the
+overall long-run mean of the phenotype of interest for each gender, and
+\(\kappa\) represents how quickly over time \(\mu_{gt}\) reverts back to
+\(\theta_g\); the larger the value of \(\kappa\), the slower
+\(\mu_{gt}\) reverts back to \(\theta_g\).
+
+We also set
+
+\[
+\begin{aligned}
+\mu_{g1} \sim \mathcal{N}\left(0,\frac{\tau^2}{1-\kappa^2}\right).
+\end{aligned}
+\] This model choice is to preserve stationarity,
+i.e.~\(p(\mu_{gs}) = p(\mu_{gt})\) for \(s,t \in \{1,\dots,T\}\).
+
+Priors: \[
+\sigma \sim \mathcal{N}(0,100)I_{\{\sigma > 0\}}
 \] \[
-u_{gt} = \kappa u_{g(t-1)} + \nu_{gt}
-\] \[
-u_{g1} \sim \mathcal{N}\left(0,\frac{\tau^2}{1-\kappa^2}\right)
+\tau \sim \mathcal{N}(0,100)I_{\{\tau > 0\}}
 \]
@@
 \[
-\nu_{gt} \sim \mathcal{N}(0,\tau^2)
-\]
-
-\[
-\epsilon_{gti} \sim \mathcal{N}(0,\sigma^2)
-\]
-
-Priors: \[
-\sigma \sim \mathcal{N}(0,2)I_{\sigma > 0}
+\kappa \sim \mathcal{N}(0,25)I_{\{0 < \kappa < 1\}} \\
 \] \[
-\tau \sim \mathcal{N}(0,2)I_{\tau > 0}
-\]
-
-\[
-\kappa \sim N(0.5,1)\\
-\] \[
-\theta_g \sim N(54,400) \\
+\theta_g \sim \mathcal{N}(54,625)I_{\{\theta_g > 0\}} \\
 \]
@@
 \begin{CSLReferences}{1}{0}
+\leavevmode\vadjust pre{\hypertarget{ref-MICE}{}}%
+Buuren, Stef van, and Karin Groothuis-Oudshoorn. 2011. {``Mice:
+Multivariate Imputation by Chained Equations in r.''} \emph{Journal of
+Statistical Software} 45 (3): 1--67.
+\url{https://doi.org/10.18637/jss.v045.i03}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-Cornault2022}{}}%

### 2023-05-10 manuscript/paper.Rmd
Edited Paper

@@
   | \large Matthew Stuart \vspace{-1.1mm}
-  | \normalsize Department of Mathematics and Statistics \vspace{-1mm}
-  | \normalsize Loyola University Chicago \vspace{-1mm}
-  | \normalsize Chicago, IL 60660 \vspace{-1mm}
-  | \normalsize [`mstuart@luc.edu`](mailto:mstuart@luc.edu) \vspace{-1mm}
+  | \normalsize Department of Mathematics and Statistics \vspace{-1.1mm}
+  | \normalsize Loyola University Chicago \vspace{-1.1mm}
+  | \normalsize Chicago, IL 60660 \vspace{-1.1mm}
+  | \normalsize [`mstuart@luc.edu`](mailto:mstuart@luc.edu) \vspace{-1.1mm}
   |
   | \large Akhil Ghosh \vspace{-1.1mm}
-  | \normalsize Department of Mathematics and Statistics \vspace{-1mm}
-  | \normalsize Loyola University Chicago \vspace{-1mm}
-  | \normalsize Chicago, IL 60660 \vspace{-1mm}
-  | \normalsize [`aghosh@luc.edu`](mailto:aghosh@luc.edu) \vspace{-1mm}
+  | \normalsize Department of Mathematics and Statistics \vspace{-1.1mm}
+  | \normalsize Loyola University Chicago \vspace{-1.1mm}
+  | \normalsize Chicago, IL 60660 \vspace{-1.1mm}
+  | \normalsize [`aghosh@luc.edu`](mailto:aghosh@luc.edu) \vspace{-1.1mm}
   |
   | \large Yoel Stuart \vspace{-1.1mm}
-  | \normalsize Department of Biology \vspace{-1mm}
-  | \normalsize Loyola University Chicago \vspace{-1mm}
-  | \normalsize Chicago, IL 60660 \vspace{-1mm}
-  | \normalsize [`ystuart@luc.edu`](mailto:ystuart@luc.edu) \vspace{-1mm}
+  | \normalsize Department of Biology \vspace{-1.1mm}
+  | \normalsize Loyola University Chicago \vspace{-1.1mm}
+  | \normalsize Chicago, IL 60660 \vspace{-1.1mm}
+  | \normalsize [`ystuart@luc.edu`](mailto:ystuart@luc.edu) \vspace{-1.1mm}
   |
   | \large Gregory J. Matthews \vspace{-1.1mm}
-  | \normalsize Center for Data Science and Consulting; Department of Mathematics and Statistics \vspace{-1mm}
-  | \normalsize Loyola University Chicago \vspace{-1mm}
-  | \normalsize Chicago, IL 60660 \vspace{-1mm}
-  | \normalsize [`gmatthews1@luc.edu`](mailto:gmatthews1@luc.edu) \vspace{-1mm}
+  | \normalsize Center for Data Science and Consulting; Department of Mathematics and Statistics \vspace{-1.1mm}
+  | \normalsize Loyola University Chicago \vspace{-1.1mm}
+  | \normalsize Chicago, IL 60660 \vspace{-1.1mm}
+  | \normalsize [`gmatthews1@luc.edu`](mailto:gmatthews1@luc.edu) \vspace{-1.1mm}
 abstract: |
@@
 ## Imputation model 
-Let $\boldsymbol{W}$ be an $(n_{extant} + n_{fossil}) \times 1$ vector of the covariate gender of the stickleback fish and $\boldsymbol{X}$ be an $(n_{extant} + n_{fossil}) \times p$ matrix of the $p$ phenotypes of interest of the fish.
+Let $\boldsymbol{W}$ be an $(n_{extant} + n_{fossil}) \times 1$ vector of the covariate gender of the stickleback fish and $\boldsymbol{X}$ be an $(n_{extant} + n_{fossil}) \times k$ matrix of the $k$ phenotypes of interest of the fish.
 Because $\boldsymbol{W}$ contains missing data, we further define $\boldsymbol{W} = (\boldsymbol{W}_{obs}^T,\boldsymbol{W}_{mis}^T)^T$ where $\boldsymbol{W}_{obs}$ and $\boldsymbol{W}_{mis}$ are the observed and missing data of the covariate gender, respectively.
@@
 ## OU model 
-After imputing sex, we fit an OU model for different phenotypes of interest. We only look at $n_{fossil}$ fossil observations.  
+After imputing sex, we fit an OU model for different phenotypes of interest.
 WLOG we choose one of the columns of $X$ to be the target phenotype and our analysis is repeated for all phenotypes of interest.
@@
 \begin{aligned}
-X_{ti} \iid \left\{\begin{array}{ll} \mathcal{N}(\mu_{ft},\sigma^2) & W_{ti} = \text{Female} \\ \mathcal{N}(\mu_{mt},\sigma^2) & W_{ti} = \text{Male} \end{array}\right.
+X_{ti} \iid \left\{\begin{array}{ll} \mathcal{N}(\theta_f + u_{ft},\sigma^2), & W_{ti} = \text{Female} \\ \mathcal{N}(\theta_m + u_{mt},\sigma^2), & W_{ti} = \text{Male} \end{array}\right.
 \end{aligned}
 $$
-We then fit the time and gender dependent means, $\mu_{ft}$ and $\mu_{mt}$, to an autoregressive process of lag 1, which is the discretization of an OU process.
-For $t = 2,\ldots,T$ and $g \in \{f,m\}$, the AR1 model for the mean structure defined as is
+In this setting, for $g \in \{f,m\}$, $\theta_g$, is the defined as the overall mean of the phenotype of interest for each gender, and $u_{gt}$ are the gender specific random effects, measuring the difference between the gender and time-specific mean and the overall gender-specific mean, which we fit to an autoregressive process of lag 1, which is the discretization of an OU process.
+For $t = 2,\ldots,T$ and $g \in \{f,m\}$, the AR1 model for is defined as
 
@@
 \begin{aligned}
-\mu_{gt} \iid \mathcal{N}(\kappa \mu_{g,t-1} + (1 - \kappa)\theta_g, \tau^2).
+u_{g,t} \iid \mathcal{N}(\kappa_g u_{g,t-1} , \tau^2).
 \end{aligned}
 $$
-In this setting, $\theta_g, g \in \{f,m\}$ is the defined as the overall long-run mean of the phenotype of interest for each gender, and $\kappa$ represents how quickly over time $\mu_{gt}$ reverts back to $\theta_g$; the larger the value of $\kappa$, the slower $\mu_{gt}$ reverts back to $\theta_g$.
+$\kappa_g$ represents the gender-specific correlation between the previous and current time periods' random effects.
+In addition, $\text{cor}(u_{g,t},u_{g,t+h}) = (\kappa_g)^h$ for $h \in \mathbb{N}$.
 
@@
 \begin{aligned}
-\mu_{g1} \sim \mathcal{N}\left(0,\frac{\tau^2}{1-\kappa^2}\right).
+u_{g,1} \iid \mathcal{N}\left(0,\frac{\tau^2}{1-\kappa_g^2}\right).
 \end{aligned}
 $$
-This model choice is to preserve stationarity, i.e. $p(\mu_{gs}) = p(\mu_{gt})$ for $s,t \in \{1,\dots,T\}$.
+This model choice is to make the distribution of the random effects stationary, i.e. $p(\mu_{g,s}) = p(\mu_{g,t})$ for $s,t \in \{1,\dots,T\}$.
 
@@
 $$
-\sigma \sim \mathcal{N}(0,100)I_{\{\sigma > 0\}}
-$$
-$$
-\tau \sim \mathcal{N}(0,100)I_{\{\tau > 0\}}
-$$
-
-$$
-\kappa \sim \mathcal{N}(0,25)I_{\{0 < \kappa < 1\}} \\
-$$
-$$
-\theta_g \sim \mathcal{N}(54,625)I_{\{\theta_g > 0\}} \\
+\begin{aligned}
+\sigma & \sim \mathcal{N}(0,100)I_{\{\sigma > 0\}} \\
+\tau & \sim \mathcal{N}(0,100)I_{\{\tau > 0\}} \\
+\kappa & \iid \mathcal{N}(0.5,1)I_{\{-1 < \kappa_g < 1\}} \\
+\theta_g & \iid \mathcal{N}(54,400)I_{\{\theta_g > 0\}}
+\end{aligned}
 $$

### 2023-05-10 manuscript/paper.tex
Edited Paper

@@
 \author{\large Matthew Stuart \vspace{-1.1mm}\\
-\normalsize Department of Mathematics and Statistics \vspace{-1mm}\\
-\normalsize Loyola University Chicago \vspace{-1mm}\\
-\normalsize Chicago, IL 60660 \vspace{-1mm}\\
+\normalsize Department of Mathematics and Statistics \vspace{-1.1mm}\\
+\normalsize Loyola University Chicago \vspace{-1.1mm}\\
+\normalsize Chicago, IL 60660 \vspace{-1.1mm}\\
 \normalsize \href{mailto:mstuart@luc.edu}{\texttt{mstuart@luc.edu}}
-\vspace{-1mm}\\
+\vspace{-1.1mm}\\
 \strut \\
 \large Akhil Ghosh \vspace{-1.1mm}\\
-\normalsize Department of Mathematics and Statistics \vspace{-1mm}\\
-\normalsize Loyola University Chicago \vspace{-1mm}\\
-\normalsize Chicago, IL 60660 \vspace{-1mm}\\
+\normalsize Department of Mathematics and Statistics \vspace{-1.1mm}\\
+\normalsize Loyola University Chicago \vspace{-1.1mm}\\
+\normalsize Chicago, IL 60660 \vspace{-1.1mm}\\
 \normalsize \href{mailto:aghosh@luc.edu}{\texttt{aghosh@luc.edu}}
-\vspace{-1mm}\\
+\vspace{-1.1mm}\\
 \strut \\
 \large Yoel Stuart \vspace{-1.1mm}\\
-\normalsize Department of Biology \vspace{-1mm}\\
-\normalsize Loyola University Chicago \vspace{-1mm}\\
-\normalsize Chicago, IL 60660 \vspace{-1mm}\\
+\normalsize Department of Biology \vspace{-1.1mm}\\
+\normalsize Loyola University Chicago \vspace{-1.1mm}\\
+\normalsize Chicago, IL 60660 \vspace{-1.1mm}\\
 \normalsize \href{mailto:ystuart@luc.edu}{\texttt{ystuart@luc.edu}}
-\vspace{-1mm}\\
+\vspace{-1.1mm}\\
 \strut \\
@@
 \normalsize Center for Data Science and Consulting; Department of
-Mathematics and Statistics \vspace{-1mm}\\
-\normalsize Loyola University Chicago \vspace{-1mm}\\
-\normalsize Chicago, IL 60660 \vspace{-1mm}\\
+Mathematics and Statistics \vspace{-1.1mm}\\
+\normalsize Loyola University Chicago \vspace{-1.1mm}\\
+\normalsize Chicago, IL 60660 \vspace{-1.1mm}\\
 \normalsize \href{mailto:gmatthews1@luc.edu}{\texttt{gmatthews1@luc.edu}}
-\vspace{-1mm}}
+\vspace{-1.1mm}}
 \date{}
@@
 vector of the covariate gender of the stickleback fish and
-\(\boldsymbol{X}\) be an \((n_{extant} + n_{fossil}) \times p\) matrix
-of the \(p\) phenotypes of interest of the fish. Because
+\(\boldsymbol{X}\) be an \((n_{extant} + n_{fossil}) \times k\) matrix
+of the \(k\) phenotypes of interest of the fish. Because
 \(\boldsymbol{W}\) contains missing data, we further define
@@
 perfectly corresponding to the observable gender of the modern extants
-and the unobservable gender of the fossil, respectively.
+and the unobservable gender of the fossil data, respectively.
 
@@
 After imputing sex, we fit an OU model for different phenotypes of
-interest. We only look at \(n_{fossil}\) fossil observations.\\
-WLOG we choose one of the columns of \(X\) to be the target phenotype
-and our analysis is repeated for all phenotypes of interest.
+interest. WLOG we choose one of the columns of \(X\) to be the target
+phenotype and our analysis is repeated for all phenotypes of interest.
 
@@
 \begin{aligned}
-X_{ti} \overset{iid}{\sim}\left\{\begin{array}{ll} \mathcal{N}(\mu_{ft},\sigma^2) & W_{ti} = \text{Female} \\ \mathcal{N}(\mu_{mt},\sigma^2) & W_{ti} = \text{Male} \end{array}\right.
+X_{ti} \overset{iid}{\sim}\left\{\begin{array}{ll} \mathcal{N}(\theta_f + u_{ft},\sigma^2), & W_{ti} = \text{Female} \\ \mathcal{N}(\theta_m + u_{mt},\sigma^2), & W_{ti} = \text{Male} \end{array}\right.
 \end{aligned}
-\] We then fit the time and gender dependent means, \(\mu_{ft}\) and
-\(\mu_{mt}\), to an autoregressive process of lag 1, which is the
-discretization of an OU process. For \(t = 2,\ldots,T\) and
-\(g \in \{f,m\}\), the AR1 model for the mean structure defined as is
+\] In this setting, for \(g \in \{f,m\}\), \(\theta_g\), is the defined
+as the overall mean of the phenotype of interest for each gender, and
+\(u_{gt}\) are the gender specific random effects, measuring the
+difference between the gender and time-specific mean and the overall
+gender-specific mean, which we fit to an autoregressive process of lag
+1, which is the discretization of an OU process. For \(t = 2,\ldots,T\)
+and \(g \in \{f,m\}\), the AR1 model for is defined as
 
@@
 \begin{aligned}
-\mu_{gt} \overset{iid}{\sim}\mathcal{N}(\kappa \mu_{g,t-1} + (1 - \kappa)\theta_g, \tau^2).
+u_{g,t} \overset{iid}{\sim}\mathcal{N}(\kappa_g u_{g,t-1} , \tau^2).
 \end{aligned}
-\] In this setting, \(\theta_g, g \in \{f,m\}\) is the defined as the
-overall long-run mean of the phenotype of interest for each gender, and
-\(\kappa\) represents how quickly over time \(\mu_{gt}\) reverts back to
-\(\theta_g\); the larger the value of \(\kappa\), the slower
-\(\mu_{gt}\) reverts back to \(\theta_g\).
+\] \(\kappa_g\) represents the gender-specific correlation between the
+previous and current time periods' random effects. In addition,
+\(\text{cor}(u_{g,t},u_{g,t+h}) = (\kappa_g)^h\) for
+\(h \in \mathbb{N}\).
 
@@
 \begin{aligned}
-\mu_{g1} \sim \mathcal{N}\left(0,\frac{\tau^2}{1-\kappa^2}\right).
+u_{g,1} \overset{iid}{\sim}\mathcal{N}\left(0,\frac{\tau^2}{1-\kappa_g^2}\right).
 \end{aligned}
-\] This model choice is to preserve stationarity,
-i.e.~\(p(\mu_{gs}) = p(\mu_{gt})\) for \(s,t \in \{1,\dots,T\}\).
+\] This model choice is to make the distribution of the random effects
+stationary, i.e.~\(p(\mu_{g,s}) = p(\mu_{g,t})\) for
+\(s,t \in \{1,\dots,T\}\).
 
 Priors: \[
-\sigma \sim \mathcal{N}(0,100)I_{\{\sigma > 0\}}
-\] \[
-\tau \sim \mathcal{N}(0,100)I_{\{\tau > 0\}}
-\]
-
-\[
-\kappa \sim \mathcal{N}(0,25)I_{\{0 < \kappa < 1\}} \\
-\] \[
-\theta_g \sim \mathcal{N}(54,625)I_{\{\theta_g > 0\}} \\
+\begin{aligned}
+\sigma & \sim \mathcal{N}(0,100)I_{\{\sigma > 0\}} \\
+\tau & \sim \mathcal{N}(0,100)I_{\{\tau > 0\}} \\
+\kappa & \overset{iid}{\sim}\mathcal{N}(0.5,1)I_{\{-1 < \kappa_g < 1\}} \\
+\theta_g & \overset{iid}{\sim}\mathcal{N}(54,400)I_{\{\theta_g > 0\}}
+\end{aligned}
 \]

### 2023-05-10 manuscript/paper.Rmd
Update Paper

@@
   | \normalsize Chicago, IL 60660 \vspace{-1.1mm}
-  | \normalsize [`mstuart@luc.edu`](mailto:mstuart@luc.edu) \vspace{-1.1mm}
+  | \normalsize [`mstuart1@luc.edu`](mailto:mstuart1@luc.edu) \vspace{-1.1mm}
   |

### 2023-05-10 manuscript/paper.Rmd
added middle initial

--- 2023-05-10 manuscript/paper.Rmd
@@
   |
-  | \large Yoel Stuart \vspace{-1.1mm}
+  | \large Yoel E. Stuart \vspace{-1.1mm}
   | \normalsize Department of Biology \vspace{-1.1mm}

### 2023-05-10 R/stuart_stickle_bayes.R
Update Bayes analysis to add additional phenotypes

@@ -3,7 +3,7 @@ library(tidyverse)
 library(rstan)
 library(ggridges)
 
-dat <- read.csv("Data/stuart_stickle_imp_100_raw.csv")
+dat <- read.csv("Data/stuart_stickle_100imputations_all.csv")
 
 mod <- "
 data{
@@ -50,27 +50,47 @@ model{
 }
 "
 sm <- stan_model(model_code = mod, verbose = TRUE)
-M <- max(dat$imputation_number)
+M <- max(dat$imp)
 
 for (m in 1:M){
-  tmp <- dat %>% filter(imputation_number == m)
-  y <- tmp$stl
+  tmp <- dat %>% filter(imp == m)
+  stl <- tmp$stl
+  cle <- tmp$cle
+  pmx <- tmp$pmx
   gender <- ifelse(tmp$gender == "Female",1,0)
   time <- tmp$time
 
-  N <- length(y)
+  N <- length(stl)
   T <- max(time)
 
-  fit <- sampling(sm,
-                  list(N=N,T=T,y=y,gender=gender,time=time),
-                  iter=10000,chains=1,warmup=5000,thin=5,
-                  pars = c("theta_f","theta_m","mu_f","mu_m","kappa","sigma","tau"))
+  print(paste("Running Imputation",m,"for stl"))
+  fit.stl <- sampling(sm,
+                      list(N=N,T=T,y=stl,gender=gender,time=time),
+                      iter=10000,chains=1,warmup=5000,thin=5,
+                      pars = c("theta_f","theta_m","mu_f","mu_m","kappa","sigma","tau"))
+  print(paste("Running Imputation",m,"for cle"))
+  fit.cle <- sampling(sm,
+                      list(N=N,T=T,y=cle,gender=gender,time=time),
+                      iter=10000,chains=1,warmup=5000,thin=5,
+                      pars = c("theta_f","theta_m","mu_f","mu_m","kappa","sigma","tau"))
+  print(paste("Running Imputation",m,"for pmx"))
+  fit.pmx <- sampling(sm,
+                      list(N=N,T=T,y=pmx,gender=gender,time=time),
+                      iter=10000,chains=1,warmup=5000,thin=5,
+                      pars = c("theta_f","theta_m","mu_f","mu_m","kappa","sigma","tau"))
 
   if (m == 1){
-    samps <- as.data.frame(fit)
[…]

### 2023-05-11 R/stuart_stickle_bayes_plots.R
Adding plots from the new phenotypes discussed in our 5/10 meeting.

@@ -1,74 +1,85 @@
 library(tidyverse)
 library(ggridges)
-samps <- readRDS("posterior_samples.RDS")
 
-samps %>%
-  rename(Female = theta_f) %>%
-  rename(Male = theta_m) %>%
-  select(Female,Male) %>%
-  gather("Gender","Value") %>%
-  ggplot() +
-  geom_density(aes(x=Value,color=Gender)) +
-  scale_color_manual(values = c("red","blue"))+
-  theme_bw() +
-  xlim(c(30,80)) +
-  xlab(expression(theta[g])) +
-  ylab(expression(paste("p(",theta[g],"|",bold(X),")")))
-ggsave("../Figures/SameKappa/theta.pdf",width=5,height=8)
+vars <- c("stl","cle","pmx")
 
-length(which(samps$theta_m > samps$theta_f))/nrow(samps)
+for (x in vars){
+  samps <- readRDS(paste0("posterior_samples_",x,".RDS"))
 
-mu_f <- samps %>% select(contains("mu_f"))
-mu_m <- samps %>% select(contains("mu_m"))
+  xlim <- if (x == "stl"){
+    xlim(30,80)
+  } else if (x == "cle"){
+    xlim(5,20)
+  } else {
+    xlim(25,85)
+  }
 
-mu_diff <- mu_m - mu_f
+  samps %>%
+    rename(Female = theta_f) %>%
+    rename(Male = theta_m) %>%
+    select(Female,Male) %>%
+    gather("Gender","Value") %>%
+    ggplot() +
+    geom_density(aes(x=Value,color=Gender)) +
+    scale_color_manual(values = c("red","blue"))+
+    theme_bw() +
+    xlim +
+    xlab(expression(theta[g])) +
+    ylab(expression(paste("p(",theta[g],"|",bold(X),")")))
+  ggsave(paste0("Figures/",x,"/theta.pdf"),width=5,height=8)
 
-names(mu_diff) <- c(1:18)
-as.data.frame(mu_diff) %>%
-  gather("Time","Value") %>%
-  mutate(Time = as.numeric(Time)) %>%
-  ggplot(aes(x=Value,y=Time,group=Time)) +
-  geom_density_ridges(fill="blue",alpha=0.5) +
-  xlim(c(-15,20)) +
-  scale_y_reverse() +
[…]

### 2023-05-11 manuscript/paper.Rmd
If you just use align without the $$ you get numbering.

@@
 
-$$
-\begin{aligned}
+\begin{align}
 X_{ti} \iid \left\{\begin{array}{ll} \mathcal{N}(\theta_f + u_{ft},\sigma^2), & W_{ti} = \text{Female} \\ \mathcal{N}(\theta_m + u_{mt},\sigma^2), & W_{ti} = \text{Male} \end{array}\right. 
-\end{aligned}
-$$
+\end{align}
+
 In this setting, for $g \in \{f,m\}$, $\theta_g$, is the defined as the overall mean of the phenotype of interest for each gender, and $u_{gt}$ are the gender specific random effects, measuring the difference between the gender and time-specific mean and the overall gender-specific mean, which we fit to an autoregressive process of lag 1, which is the discretization of an OU process.

### 2023-05-12 R/stuart_stickle_bayes.R
Rerun model for size corrected cle and pmx.

@@ -55,8 +55,8 @@ M <- max(dat$imp)
 for (m in 1:M){
   tmp <- dat %>% filter(imp == m)
   stl <- tmp$stl
-  cle <- tmp$cle
-  pmx <- tmp$pmx
+  cle <- tmp$cle.sc
+  pmx <- tmp$pmx.sc
   gender <- ifelse(tmp$gender == "Female",1,0)
   time <- tmp$time
 

### 2023-05-12 R/stuart_stickle_bayes_plots.R

@@ -9,7 +9,7 @@ for (x in vars){
   xlim <- if (x == "stl"){
     xlim(30,80)
   } else if (x == "cle"){
-    xlim(5,20)
+    xlim(5,12)
   } else {
     xlim(25,85)
   }

### 2023-05-12 manuscript/paper.Rmd
numbering
Update paper.

@@
 # Models  {#sec:models}
-## Imputation model 
+## Imputation Model 
 Let $\boldsymbol{W}$ be an $(n_{extant} + n_{fossil}) \times 1$ vector of the covariate gender of the stickleback fish and $\boldsymbol{X}$ be an $(n_{extant} + n_{fossil}) \times k$ matrix of the $k$ phenotypes of interest of the fish.
@@
 
-## OU model 
-After imputing sex, we fit an OU model for different phenotypes of interest.
-WLOG we choose one of the columns of $X$ to be the target phenotype and our analysis is repeated for all phenotypes of interest.
+## Completed Data Model 
+For each of the $M = 100$ imputed datasets, we fit an OU model for different phenotypes of interest.
+We choose one of the columns of $X$ to be the target phenotype and our analysis is repeated for all phenotypes of interest.
 
@@
 X_{ti} \iid \left\{\begin{array}{ll} \mathcal{N}(\theta_f + u_{ft},\sigma^2), & W_{ti} = \text{Female} \\ \mathcal{N}(\theta_m + u_{mt},\sigma^2), & W_{ti} = \text{Male} \end{array}\right.
+\label{eq:x}
 \end{align}
 
-In this setting, for $g \in \{f,m\}$, $\theta_g$, is the defined as the overall mean of the phenotype of interest for each gender, and $u_{gt}$ are the gender specific random effects, measuring the difference between the gender and time-specific mean and the overall gender-specific mean, which we fit to an autoregressive process of lag 1, which is the discretization of an OU process.
+Look at \ref{eq:x}
+
+For $g \in \{f,m\}$, $\theta_g$, is the defined as the overall mean of the phenotype of interest for each gender, and $u_{gt}$ are the gender specific random effects, measuring the difference between the gender and time-specific mean and the overall gender-specific mean, which we fit to an autoregressive process of lag 1, which is the discretization of an OU process.
 For $t = 2,\ldots,T$ and $g \in \{f,m\}$, the AR1 model for is defined as
 
-$$
-\begin{aligned}
+\begin{align}
 u_{g,t} \iid \mathcal{N}(\kappa_g u_{g,t-1} , \tau^2).
-\end{aligned}
-$$
+\label{eq:ar1}
+\end{align}
+
 $\kappa_g$ represents the gender-specific correlation between the previous and current time periods' random effects.
@@
 
-$$
-\begin{aligned}
+\begin{align}
 u_{g,1} \iid \mathcal{N}\left(0,\frac{\tau^2}{1-\kappa_g^2}\right).
-\end{aligned}
-$$
+\label{eq:u1}
+\end{align}
+
 This model choice is to make the distribution of the random effects stationary, i.e. $p(\mu_{g,s}) = p(\mu_{g,t})$ for $s,t \in \{1,\dots,T\}$.
@@
 Priors:
-$$
-\begin{aligned}
-\sigma & \sim \mathcal{N}(0,100)I_{\{\sigma > 0\}} \\
-\tau & \sim \mathcal{N}(0,100)I_{\{\tau > 0\}} \\
-\kappa & \iid \mathcal{N}(0.5,1)I_{\{-1 < \kappa_g < 1\}} \\
+
+\begin{align}
+\sigma & \sim \mathcal{N}(0,100)I_{\{\sigma > 0\}} \nonumber \\
+\tau & \sim \mathcal{N}(0,100)I_{\{\tau > 0\}} \nonumber \\
+\kappa & \iid \mathcal{N}(0.5,1)I_{\{-1 < \kappa_g < 1\}} \nonumber \\
 \theta_g & \iid \mathcal{N}(54,400)I_{\{\theta_g > 0\}}
-\end{aligned}
-$$
+\label{eq:priors}
+\end{align}
 

### 2023-05-12 manuscript/paper.tex
Update Paper
Update paper.

@@
 \normalsize Chicago, IL 60660 \vspace{-1.1mm}\\
-\normalsize \href{mailto:mstuart@luc.edu}{\texttt{mstuart@luc.edu}}
+\normalsize \href{mailto:mstuart1@luc.edu}{\texttt{mstuart1@luc.edu}}
 \vspace{-1.1mm}\\
@@
 \strut \\
-\large Yoel Stuart \vspace{-1.1mm}\\
+\large Yoel E. Stuart \vspace{-1.1mm}\\
 \normalsize Department of Biology \vspace{-1.1mm}\\
@@
 
-\[
-\begin{aligned}
+\begin{align}
 X_{ti} \overset{iid}{\sim}\left\{\begin{array}{ll} \mathcal{N}(\theta_f + u_{ft},\sigma^2), & W_{ti} = \text{Female} \\ \mathcal{N}(\theta_m + u_{mt},\sigma^2), & W_{ti} = \text{Male} \end{array}\right.
-\end{aligned}
-\] In this setting, for \(g \in \{f,m\}\), \(\theta_g\), is the defined
-as the overall mean of the phenotype of interest for each gender, and
-\(u_{gt}\) are the gender specific random effects, measuring the
-difference between the gender and time-specific mean and the overall
-gender-specific mean, which we fit to an autoregressive process of lag
-1, which is the discretization of an OU process. For \(t = 2,\ldots,T\)
-and \(g \in \{f,m\}\), the AR1 model for is defined as
+\label{eq:x}
+\end{align}
 
-\[
-\begin{aligned}
+Look at \ref{eq:x}
+
+For \(g \in \{f,m\}\), \(\theta_g\), is the defined as the overall mean
+of the phenotype of interest for each gender, and \(u_{gt}\) are the
+gender specific random effects, measuring the difference between the
+gender and time-specific mean and the overall gender-specific mean,
+which we fit to an autoregressive process of lag 1, which is the
+discretization of an OU process. For \(t = 2,\ldots,T\) and
+\(g \in \{f,m\}\), the AR1 model for is defined as
+
+\begin{align}
 u_{g,t} \overset{iid}{\sim}\mathcal{N}(\kappa_g u_{g,t-1} , \tau^2).
-\end{aligned}
-\] \(\kappa_g\) represents the gender-specific correlation between the
+\label{eq:ar1}
+\end{align}
+
+\(\kappa_g\) represents the gender-specific correlation between the
 previous and current time periods' random effects. In addition,
@@
 
-\[
-\begin{aligned}
+\begin{align}
 u_{g,1} \overset{iid}{\sim}\mathcal{N}\left(0,\frac{\tau^2}{1-\kappa_g^2}\right).
-\end{aligned}
-\] This model choice is to make the distribution of the random effects
+\label{eq:u1}
+\end{align}
+
+This model choice is to make the distribution of the random effects
 stationary, i.e.~\(p(\mu_{g,s}) = p(\mu_{g,t})\) for
@@
 
-Priors: \[
-\begin{aligned}
-\sigma & \sim \mathcal{N}(0,100)I_{\{\sigma > 0\}} \\
-\tau & \sim \mathcal{N}(0,100)I_{\{\tau > 0\}} \\
-\kappa & \overset{iid}{\sim}\mathcal{N}(0.5,1)I_{\{-1 < \kappa_g < 1\}} \\
+Priors:
+
+\begin{align}
+\sigma & \sim \mathcal{N}(0,100)I_{\{\sigma > 0\}} \nonumber \\
+\tau & \sim \mathcal{N}(0,100)I_{\{\tau > 0\}} \nonumber \\
+\kappa & \overset{iid}{\sim}\mathcal{N}(0.5,1)I_{\{-1 < \kappa_g < 1\}} \nonumber \\
 \theta_g & \overset{iid}{\sim}\mathcal{N}(54,400)I_{\{\theta_g > 0\}}
-\end{aligned}
-\]
+\label{eq:priors}
+\end{align}
 

### 2023-05-15 R/stuart_stickle_bayes.R
Update Bayesian Model to add all phenotypes.

@@ -4,47 +4,177 @@ library(rstan)
 library(ggridges)
 
 dat <- read.csv("Data/stuart_stickle_100imputations_all.csv")
+dat$group <- factor(paste(dat$gender,letters[dat$time]))
 
 mod <- "
 data{
-  int<lower = 0> N;
-  int<lower = 0> T;
-  real<lower = 0> y[N];
-  int<lower = 0,upper = 1> gender[N];
-  int<lower = 1,upper = T> time[N];
+  int<lower=0> N;
+  int<lower=0> T;
+  matrix[N,2*T] X;
+  vector[N] stl;
+  vector[N] lps;
+  vector[N] ect;
+  vector[N] tpg;
+  vector[N] cle;
+  vector[N] pmx;
+  vector[N] ds1;
+  vector[N] ds2;
+  vector[N] ds3;
+  vector[N] lpt;
+  int<lower=0> mds[N];
+  int<lower=0> mdf[N];
+  int<lower=0> mav[N];
+  int<lower=0> maf[N];
+  int<lower=0> mcf[N];
+  int<lower=0> mpt[N];
 }
 parameters{
-  real<lower=0> theta_f;
-  real<lower=0> theta_m;
-  real u_f[T];
-  real u_m[T];
-  real<lower=-1,upper=1> kappa;
-  real<lower=0> sigma;
-  real<lower=0> tau;
+  real<lower=0> theta_f[16];
+  real<lower=0> theta_m[16];
+  real<lower=-1,upper=1> kappa[16];
+  real<lower=0> sigma[10];
+  real<lower=0> tau[16];
+  vector[2*T] u_stl;
+  vector[2*T] u_lps;
+  vector[2*T] u_ect;
+  vector[2*T] u_tpg;
+  vector[2*T] u_cle;
+  vector[2*T] u_pmx;
+  vector[2*T] u_ds1;
+  vector[2*T] u_ds2;
+  vector[2*T] u_ds3;
+  vector[2*T] u_lpt;
+  vector[2*T] u_mds;
+  vector[2*T] u_mdf;
+  vector[2*T] u_mav;
[…]

### 2023-05-15 R/stuart_stickle_bayes.R
Update Bayes

@@ -69,40 +69,38 @@ transformed parameters{
   vector[2*T] mu_mcf;
   vector[2*T] mu_mpt;
 
-  for (t in 1:T){
-    mu_stl[t] = theta_f[1] + u_stl[t];
-    mu_stl[T+t] = theta_m[1] + u_stl[T+t];
-    mu_lps[t] = theta_f[2] + u_lps[t];
-    mu_lps[T+t] = theta_m[2] + u_lps[T+t];
-    mu_ect[t] = theta_f[3] + u_ect[t];
-    mu_ect[T+t] = theta_m[3] + u_ect[T+t];
-    mu_tpg[t] = theta_f[4] + u_tpg[t];
-    mu_tpg[T+t] = theta_m[4] + u_tpg[T+t];
-    mu_cle[t] = theta_f[5] + u_cle[t];
-    mu_cle[T+t] = theta_m[5] + u_cle[T+t];
-    mu_pmx[t] = theta_f[6] + u_pmx[t];
-    mu_pmx[T+t] = theta_m[6] + u_pmx[T+t];
-    mu_ds1[t] = theta_f[7] + u_ds1[t];
-    mu_ds1[T+t] = theta_m[7] + u_ds1[T+t];
-    mu_ds2[t] = theta_f[8] + u_ds2[t];
-    mu_ds2[T+t] = theta_m[8] + u_ds2[T+t];
-    mu_ds3[t] = theta_f[9] + u_ds3[t];
-    mu_ds3[T+t] = theta_m[9] + u_ds3[T+t];
-    mu_lpt[t] = theta_f[10] + u_lpt[t];
-    mu_lpt[T+t] = theta_m[10] + u_lpt[T+t];
-    mu_mds[t] = theta_f[11] + u_mds[t];
-    mu_mds[T+t] = theta_m[11] + u_mds[T+t];
-    mu_mdf[t] = theta_f[12] + u_mdf[t];
-    mu_mdf[T+t] = theta_m[12] + u_mdf[T+t];
-    mu_mav[t] = theta_f[13] + u_mav[t];
-    mu_mav[T+t] = theta_m[13] + u_mav[T+t];
-    mu_maf[t] = theta_f[14] + u_maf[t];
[…]

### 2023-05-17 R/stuart_stickle_bayes.R
Update STAN model print message

@@ -205,7 +205,7 @@ for (m in 1:M){
   N <- length(stl)
   T <- max(time)
 
-  print(paste("Running Imputation",m,"for stl"))
+  print(paste("Running Imputation",m))
   fit <- sampling(sm,
                   list(N=N,T=T,X=X,stl=stl,
                        lpt=lpt,mds=mds,mdf=mdf,

### 2023-05-19 manuscript/paper.Rmd
Update Manuscript

@@
 ## Imputation Model 
-Let $\boldsymbol{W}$ be an $(n_{extant} + n_{fossil}) \times 1$ vector of the covariate gender of the stickleback fish and $\boldsymbol{X}$ be an $(n_{extant} + n_{fossil}) \times k$ matrix of the $k$ phenotypes of interest of the fish.
-Because $\boldsymbol{W}$ contains missing data, we further define $\boldsymbol{W} = (\boldsymbol{W}_{obs}^T,\boldsymbol{W}_{mis}^T)^T$ where $\boldsymbol{W}_{obs}$ and $\boldsymbol{W}_{mis}$ are the observed and missing data of the covariate gender, respectively.
-In this setting, $\boldsymbol{W}_{obs}$ is a $n_{extant} \times 1$ vector and $\boldsymbol{W}_{obs}$ is a $n_{fossil} \times 1$ vector, perfectly corresponding to the observable gender of the modern extants and the unobservable gender of the fossil data, respectively.
+Let $\boldsymbol{W}$ be an $(n_{extant} + n_{fossil}) \times 1$ vector of the covariate gender of the stickleback fish, $\boldsymbol{X}$ be an $(n_{extant} + n_{fossil}) \times K_{cont}$ matrix of the $K_{cont}$ continuous phenotypes of interest, and $\boldsymbol{Y}$ be an $(n_{extant} + n_{fossil}) \times K_{disc}$ matrix of the $K_{disc}$ discrete phenotypes of interest.
+Because the gender of the fossilized stickleback fish is unobservable, we further define $\boldsymbol{W} = (\boldsymbol{W}_{extant}^T,\boldsymbol{W}_{fossil}^T)^T$ where $\boldsymbol{W}_{extant}$ and $\boldsymbol{W}_{fossil}$ are the $n_{extant} \times 1$ and $n_{fossil} \times 1$ vectors of the observed extant gender and missing fossil gender, respectively.
 
-We impute the missing gender for the fossil data by sampling from the posterior predictive distribution $P(\boldsymbol{W}_{mis}|\boldsymbol{W}_{obs}, \boldsymbol{X})$ using the multiple imputation with chained equations (MICE) algorithm (@MICE) with predictive mean matching.
+We impute the missing gender for the fossil data by sampling from the posterior predictive distribution $P(\boldsymbol{W}_{fossil}|\boldsymbol{W}_{extant}, \boldsymbol{X}, \boldsymbol{Y})$ using the multiple imputation with chained equations (MICE) algorithm (@MICE) with predictive mean matching.
 The imputation algorithm is run to obtain a total of $M = 100$ completed datasets.
@@
 
-Let $W_{ti}$ and $X_{ti}$ be the gender and phenotype of interest for the $i$-th observation in year $t$ for $t = 1, \ldots, 18$ and  $i = 1\ldots,n_{t}$.  We assume that
+Let $W_{ti}$, $\boldsymbol{X}_{ti}$, and $\boldsymbol{Y}_{ti}$ be the imputed gender, $K_{cont} \times 1$ vector of continuous phenotypes, and $K_{disc} \times 1$ vector of discrete phenotypes, respectively, for the $i$-th observed fossil in year $t$ for $t = 1, \ldots, 18$ and  $i = 1\ldots,n_{t}$.
+In addition,
 

### 2023-05-21 manuscript/paper.Rmd
Update Manuscript

@@
 For each of the $M = 100$ imputed datasets, we fit an OU model for different phenotypes of interest.
-We choose one of the columns of $X$ to be the target phenotype and our analysis is repeated for all phenotypes of interest.
-
-Let $W_{ti}$, $\boldsymbol{X}_{ti}$, and $\boldsymbol{Y}_{ti}$ be the imputed gender, $K_{cont} \times 1$ vector of continuous phenotypes, and $K_{disc} \times 1$ vector of discrete phenotypes, respectively, for the $i$-th observed fossil in year $t$ for $t = 1, \ldots, 18$ and  $i = 1\ldots,n_{t}$.
+For the $i$-th observed fossil in year $t$ ($t = 1, \ldots, 18$ and  $i = 1\ldots,n_{t}$), let $W_{ti}$ be the imputed gender, $\boldsymbol{X}_{ti}$ be the $K_{cont} \times 1$ vector of continuous phenotypes, and $\boldsymbol{Y}_{ti}$ be the $K_{disc} \times 1$ vector of discrete phenotypes for.
 In addition,

### 2023-05-22 R/stuart_stickle_bayes.R
Running model without the non-integer count until Yoel gets back to us.

@@ -203,7 +203,7 @@ for (m in 1:M){
   X <- model.matrix(~0+group,tmp)
 
   N <- length(stl)
-  T <- max(time)
+  T <- max(tmp$time)
 
   print(paste("Running Imputation",m))
   fit <- sampling(sm,

### 2023-05-22 R/stuart_stickle_bayes.R
Update Bayes model

@@ -21,7 +21,6 @@ data{
   vector[N] ds2;
   vector[N] ds3;
   vector[N] lpt;
-  int<lower=0> mds[N];
   int<lower=0> mdf[N];
   int<lower=0> mav[N];
   int<lower=0> maf[N];
@@ -29,11 +28,11 @@ data{
   int<lower=0> mpt[N];
 }
 parameters{
-  real<lower=0> theta_f[16];
-  real<lower=0> theta_m[16];
-  real<lower=-1,upper=1> kappa[16];
+  real<lower=0> theta_f[15];
+  real<lower=0> theta_m[15];
+  real<lower=-1,upper=1> kappa[15];
   real<lower=0> sigma[10];
-  real<lower=0> tau[16];
+  real<lower=0> tau[15];
   real beta[9];
   vector[2*T] u_stl;
   vector[2*T] u_lps;
@@ -45,7 +44,6 @@ parameters{
   vector[2*T] u_ds2;
   vector[2*T] u_ds3;
   vector[2*T] u_lpt;
-  vector[2*T] u_mds;
   vector[2*T] u_mdf;
   vector[2*T] u_mav;
   vector[2*T] u_maf;
@@ -63,7 +61,6 @@ transformed parameters{
   vector[2*T] mu_ds2;
   vector[2*T] mu_ds3;
   vector[2*T] mu_lpt;
-  vector[2*T] mu_mds;
   vector[2*T] mu_mdf;
   vector[2*T] mu_mav;
   vector[2*T] mu_maf;
@@ -90,18 +87,16 @@ transformed parameters{
   mu_ds3[(T+1):(2*T)] = theta_m[9] + u_ds3[(T+1):(2*T)];
   mu_lpt[1:T] = theta_f[10] + u_lpt[1:T];
   mu_lpt[(T+1):(2*T)] = theta_m[10] + u_lpt[(T+1):(2*T)];
-  mu_mds[1:T] = theta_f[11] + u_mds[1:T];
-  mu_mds[(T+1):(2*T)] = theta_m[11] + u_mds[(T+1):(2*T)];
-  mu_mdf[1:T] = theta_f[12] + u_mdf[1:T];
-  mu_mdf[(T+1):(2*T)] = theta_m[12] + u_mdf[(T+1):(2*T)];
[…]

### 2023-05-23 manuscript/paper.Rmd
Update post-imputation model specifications.

@@
 ## Imputation Model 
-Let $\boldsymbol{W}$ be an $(n_{extant} + n_{fossil}) \times 1$ vector of the covariate gender of the stickleback fish, $\boldsymbol{X}$ be an $(n_{extant} + n_{fossil}) \times K_{cont}$ matrix of the $K_{cont}$ continuous phenotypes of interest, and $\boldsymbol{Y}$ be an $(n_{extant} + n_{fossil}) \times K_{disc}$ matrix of the $K_{disc}$ discrete phenotypes of interest.
+Let $\boldsymbol{W}$ be an $(n_{extant} + n_{fossil}) \times 1$ vector of the covariate gender of the stickleback fish, $\boldsymbol{X}$ be an $(n_{extant} + n_{fossil}) \times K$ matrix of the $K$ continuous phenotypes of interest, and $\boldsymbol{Y}$ be an $(n_{extant} + n_{fossil}) \times L$ matrix of the $L$ discrete phenotypes of interest.
 Because the gender of the fossilized stickleback fish is unobservable, we further define $\boldsymbol{W} = (\boldsymbol{W}_{extant}^T,\boldsymbol{W}_{fossil}^T)^T$ where $\boldsymbol{W}_{extant}$ and $\boldsymbol{W}_{fossil}$ are the $n_{extant} \times 1$ and $n_{fossil} \times 1$ vectors of the observed extant gender and missing fossil gender, respectively.
@@
 ## Completed Data Model 
-For each of the $M = 100$ imputed datasets, we fit an OU model for different phenotypes of interest.
-For the $i$-th observed fossil in year $t$ ($t = 1, \ldots, 18$ and  $i = 1\ldots,n_{t}$), let $W_{ti}$ be the imputed gender, $\boldsymbol{X}_{ti}$ be the $K_{cont} \times 1$ vector of continuous phenotypes, and $\boldsymbol{Y}_{ti}$ be the $K_{disc} \times 1$ vector of discrete phenotypes for.
-In addition,
+For a given imputed dataset, let $W_{ti}$ be the imputed gender, $\boldsymbol{X}_{ti}$ be the $K \times 1$ vector of continuous phenotypes, and $\boldsymbol{Y}_{ti}$ be the $L \times 1$ vector of discrete phenotypes for $t = 1, \ldots, T$ and  $i = 1\ldots,n_{t}$).
+
+Define the last element of $\boldsymbol{X}_{ti}$ ($X_{K,ti}$) as the length of the stickleback.  For $k = 1,\cdots,K - 1$, we will assume
+\begin{align}
+{X}_{k,ti} & \iid \left\{\begin{array}{llll} \mathcal{N}(\mu_{k,ft} & + \beta_k(X_{K,ti} - \mu_{K,ft}),&\sigma_k^2), & W_{ti} = \text{Female} \\ \mathcal{N}(\mu_{k,mt} & + \beta_k(X_{K,ti} - \mu_{K,mt}),&\sigma_k^2), & W_{ti} = \text{Male} \end{array}\right.
+\\
+{X}_{K,ti} & \iid \left\{\begin{array}{lll} \mathcal{N}(\mu_{K,ft},&\sigma_{K}^2), & W_{ti} = \text{Female} \\ \mathcal{N}(\mu_{K,mt},&\sigma_{K}^2), & W_{ti} = \text{Male} \end{array}\right.
+\label{eq:X}
+\end{align}
+where $\mu_{k,ft}$ and $\mu_{k,mt}$ represent the time-$t$ specific mean of phenotype $k$ for female and male stickleback fish, respectively, and $\beta_k$ is an additional parameter to account for the biological phenomenon that other continuous phenotypes for a particular animal are positively correlated with their length \textbf{CITATION NEEDED}.
+We further set \begin{align}
+\mu_{k,gt} = \theta_{k,g} + u_{k,gt},
+\label{eq:mu}
+\end{align}
+for $g \in \{f,m\}$ where $\theta_{k,g}$ is the overall mean of phenotype $X_k$ for each gender, and $u_{k,gt}$ measures the difference between $\mu_{k,gt}$ and $\theta_{k,g}$.
+We then fit $\boldsymbol{u}_{k,g} = \{u_{k,g1},\cdots,u_{k,gT}\}$ to an Ornstein-Uhlenback (OU) process (@OUprocess) where we assume $u_{k,gT}$ have a marginal mean of 0.
+Because each time period is ~1000 years, we will discretize the OU process without loss of generality.
+For $t = 2,\ldots,T$, we assume
+\begin{align}
+u_{k,gt} \iid \mathcal{N}(\kappa_{k} u_{k,g(t-1)} , \tau_k^2).
+\label{eq:u_ar1}
+\end{align}
+
+$\kappa_{k,g}$ represents the correlation between $\mu_{k,gt}$ and $\mu_{k,g(t+1)}$, and it also follows that $\text{cor}(u_{g,t},u_{g,t+h}) = \kappa^h$ for $h \in \mathbb{N}$.  In addition, we assume
+\begin{align}
+u_{k,g1} \iid \mathcal{N}\left(0,\frac{\tau_k^2}{1-\kappa_{k}^2}\right).
+\label{eq:u1}
+\end{align}
+This model choice is to preserve stationarity; i.e. $p(u_{k,gs}) = p(u_{k,gt})$ for $s \neq t$.
+
+For the discrete phenotypes, we will assume
+\begin{align}
+{Y}_{l,ti} & \sim \left\{\begin{array}{ll} Poisson(\lambda_{l,ft}), & W_{ti} = \text{Female} \\ Poisson(\lambda_{l,mt}), & W_{ti} = \text{Male} \end{array}\right.,
+\label{eq:Y}
+\end{align}
+for $l = 1,\cdots,L$.  Similar to the continuous phenotypes, we set
+\begin{align}
+\log(\lambda_{l,gt}) = \gamma_{l,g} + v_{l,gt}.
+\label{eq:lambda}
+\end{align}
+We use the log function to allow $v_{l,gt}$ to take all values on the real number line so we can properly fit an OU process to these effects.  We further assume
+\begin{align}
+v_{l,gt} \iid \mathcal{N}(\phi_{l} v_{l,g(t-1)} , \omega_l^2),
+\label{eq:v_ar1}
+\end{align}
+and
+\begin{align}
+v_{l,g1} \iid \mathcal{N}\left(0,\frac{\omega_l^2}{1 - \phi_{l}^2}\right)..
+\label{eq:v1}
+\end{align}
+
+Because we are fitting a dataset with a stochastic structure on the means of the phenotypes, we analyze the data via a Bayesian analysis.
+Bayesian data analysis is also more naturally used when we have to impute data (\textbf{CITATION}).
+
+Priors: For $k = 1,\cdots,K$ and $l = 1,\cdots,L$,
 
 \begin{align}
-X_{ti} \iid \left\{\begin{array}{ll} \mathcal{N}(\theta_f + u_{ft},\sigma^2), & W_{ti} = \text{Female} \\ \mathcal{N}(\theta_m + u_{mt},\sigma^2), & W_{ti} = \text{Male} \end{array}\right.
-\label{eq:x}
-\end{align}
-
-Look at \ref{eq:x}
-
-For $g \in \{f,m\}$, $\theta_g$, is the defined as the overall mean of the phenotype of interest for each gender, and $u_{gt}$ are the gender specific random effects, measuring the difference between the gender and time-specific mean and the overall gender-specific mean, which we fit to an autoregressive process of lag 1, which is the discretization of an OU process.
-For $t = 2,\ldots,T$ and $g \in \{f,m\}$, the AR1 model for is defined as
-
-\begin{align}
-u_{g,t} \iid \mathcal{N}(\kappa_g u_{g,t-1} , \tau^2).
-\label{eq:ar1}
-\end{align}
-
-$\kappa_g$ represents the gender-specific correlation between the previous and current time periods' random effects.
-In addition, $\text{cor}(u_{g,t},u_{g,t+h}) = (\kappa_g)^h$ for $h \in \mathbb{N}$.
-
-We also set
-
-\begin{align}
-u_{g,1} \iid \mathcal{N}\left(0,\frac{\tau^2}{1-\kappa_g^2}\right).
-\label{eq:u1}
-\end{align}
-
-This model choice is to make the distribution of the random effects stationary, i.e. $p(\mu_{g,s}) = p(\mu_{g,t})$ for $s,t \in \{1,\dots,T\}$.
-
-Priors:
-
-\begin{align}
-\sigma & \sim \mathcal{N}(0,100)I_{\{\sigma > 0\}} \nonumber \\
-\tau & \sim \mathcal{N}(0,100)I_{\{\tau > 0\}} \nonumber \\
-\kappa & \iid \mathcal{N}(0.5,1)I_{\{-1 < \kappa_g < 1\}} \nonumber \\
-\theta_g & \iid \mathcal{N}(54,400)I_{\{\theta_g > 0\}}
+\sigma_k & \iid \mathcal{N}(0,100)I_{\{\sigma > 0\}} \nonumber \\
+\tau_k & \iid \mathcal{N}(0,100)I_{\{\tau > 0\}} \nonumber \\
+\kappa_k & \iid \mathcal{N}(0,1)I_{\{-1 < \kappa_g < 1\}} \nonumber \\
+\beta_k & \iid \mathcal{N}(0,5) \\
+\theta_{k,g} & \iid \mathcal{N}(0,10000)
+\omega_l & \iid \mathcal{N}(0,100)I_{\{\tau > 0\}} \nonumber \\
+\phi_l & \iid \mathcal{N}(0,1)I_{\{-1 < \kappa_g < 1\}} \nonumber \\
+\gamma_{l,g} & \iid \mathcal{N}(0,10000)
 \label{eq:priors}

### 2023-05-23 manuscript/paper.tex
Update post-imputation model specifications.

@@
 \hypertarget{imputation-model}{%
-\subsection{Imputation model}\label{imputation-model}}
+\subsection{Imputation Model}\label{imputation-model}}
 
 Let \(\boldsymbol{W}\) be an \((n_{extant} + n_{fossil}) \times 1\)
-vector of the covariate gender of the stickleback fish and
-\(\boldsymbol{X}\) be an \((n_{extant} + n_{fossil}) \times k\) matrix
-of the \(k\) phenotypes of interest of the fish. Because
-\(\boldsymbol{W}\) contains missing data, we further define
-\(\boldsymbol{W} = (\boldsymbol{W}_{obs}^T,\boldsymbol{W}_{mis}^T)^T\)
-where \(\boldsymbol{W}_{obs}\) and \(\boldsymbol{W}_{mis}\) are the
-observed and missing data of the covariate gender, respectively. In this
-setting, \(\boldsymbol{W}_{obs}\) is a \(n_{extant} \times 1\) vector
-and \(\boldsymbol{W}_{obs}\) is a \(n_{fossil} \times 1\) vector,
-perfectly corresponding to the observable gender of the modern extants
-and the unobservable gender of the fossil data, respectively.
+vector of the covariate gender of the stickleback fish,
+\(\boldsymbol{X}\) be an \((n_{extant} + n_{fossil}) \times K\) matrix
+of the \(K\) continuous phenotypes of interest, and \(\boldsymbol{Y}\)
+be an \((n_{extant} + n_{fossil}) \times L\) matrix of the \(L\)
+discrete phenotypes of interest. Because the gender of the fossilized
+stickleback fish is unobservable, we further define
+\(\boldsymbol{W} = (\boldsymbol{W}_{extant}^T,\boldsymbol{W}_{fossil}^T)^T\)
+where \(\boldsymbol{W}_{extant}\) and \(\boldsymbol{W}_{fossil}\) are
+the \(n_{extant} \times 1\) and \(n_{fossil} \times 1\) vectors of the
+observed extant gender and missing fossil gender, respectively.
 
@@
 posterior predictive distribution
-\(P(\boldsymbol{W}_{mis}|\boldsymbol{W}_{obs}, \boldsymbol{X})\) using
-the multiple imputation with chained equations (MICE) algorithm (Buuren
-and Groothuis-Oudshoorn (\protect\hyperlink{ref-MICE}{2011})) with
-predictive mean matching. The imputation algorithm is run to obtain a
-total of \(M = 100\) completed datasets.
+\(P(\boldsymbol{W}_{fossil}|\boldsymbol{W}_{extant}, \boldsymbol{X}, \boldsymbol{Y})\)
+using the multiple imputation with chained equations (MICE) algorithm
+(Buuren and Groothuis-Oudshoorn (\protect\hyperlink{ref-MICE}{2011}))
+with predictive mean matching. The imputation algorithm is run to obtain
+a total of \(M = 100\) completed datasets.
 
-\hypertarget{ou-model}{%
-\subsection{OU model}\label{ou-model}}
+\hypertarget{completed-data-model}{%
+\subsection{Completed Data Model}\label{completed-data-model}}
 
-After imputing sex, we fit an OU model for different phenotypes of
-interest. WLOG we choose one of the columns of \(X\) to be the target
-phenotype and our analysis is repeated for all phenotypes of interest.
+For a given imputed dataset, let \(W_{ti}\) be the imputed gender,
+\(\boldsymbol{X}_{ti}\) be the \(K \times 1\) vector of continuous
+phenotypes, and \(\boldsymbol{Y}_{ti}\) be the \(L \times 1\) vector of
+discrete phenotypes for \(t = 1, \ldots, T\) and \(i = 1\ldots,n_{t}\)).
 
-Let \(W_{ti}\) and \(X_{ti}\) be the gender and phenotype of interest
-for the \(i\)-th observation in year \(t\) for \(t = 1, \ldots, 18\) and
-\(i = 1\ldots,n_{t}\). We assume that
+Define the last element of \(\boldsymbol{X}_{ti}\) (\(X_{K,ti}\)) as the
+length of the stickleback. For \(k = 1,\cdots,K - 1\), we will assume
+\begin{align}
+{X}_{k,ti} & \overset{iid}{\sim}\left\{\begin{array}{llll} \mathcal{N}(\mu_{k,ft} & + \beta_k(X_{K,ti} - \mu_{K,ft}),&\sigma_k^2), & W_{ti} = \text{Female} \\ \mathcal{N}(\mu_{k,mt} & + \beta_k(X_{K,ti} - \mu_{K,mt}),&\sigma_k^2), & W_{ti} = \text{Male} \end{array}\right.
+\\
+{X}_{K,ti} & \overset{iid}{\sim}\left\{\begin{array}{lll} \mathcal{N}(\mu_{K,ft},&\sigma_{K}^2), & W_{ti} = \text{Female} \\ \mathcal{N}(\mu_{K,mt},&\sigma_{K}^2), & W_{ti} = \text{Male} \end{array}\right.
+\label{eq:X}
+\end{align} where \(\mu_{k,ft}\) and \(\mu_{k,mt}\) represent the
+time-\(t\) specific mean of phenotype \(k\) for female and male
+stickleback fish, respectively, and \(\beta_k\) is an additional
+parameter to account for the biological phenomenon that other continuous
+phenotypes for a particular animal are positively correlated with their
+length \textbf{CITATION NEEDED}. We further set \begin{align}
+\mu_{k,gt} = \theta_{k,g} + u_{k,gt},
+\label{eq:mu}
+\end{align} for \(g \in \{f,m\}\) where \(\theta_{k,g}\) is the overall
+mean of phenotype \(X_k\) for each gender, and \(u_{k,gt}\) measures the
+difference between \(\mu_{k,gt}\) and \(\theta_{k,g}\). We then fit
+\(\boldsymbol{u}_{k,g} = \{u_{k,g1},\cdots,u_{k,gT}\}\) to an
+Ornstein-Uhlenback (OU) process
+((\protect\hyperlink{ref-OUprocess}{\textbf{OUprocess?}})) where we
+assume \(u_{k,gT}\) have a marginal mean of 0. Because each time period
+is \textasciitilde1000 years, we will discretize the OU process without
+loss of generality. For \(t = 2,\ldots,T\), we assume \begin{align}
+u_{k,gt} \overset{iid}{\sim}\mathcal{N}(\kappa_{k} u_{k,g(t-1)} , \tau_k^2).
+\label{eq:u_ar1}
+\end{align}
+
+\(\kappa_{k,g}\) represents the correlation between \(\mu_{k,gt}\) and
+\(\mu_{k,g(t+1)}\), and it also follows that
+\(\text{cor}(u_{g,t},u_{g,t+h}) = \kappa^h\) for \(h \in \mathbb{N}\).
+In addition, we assume \begin{align}
+u_{k,g1} \overset{iid}{\sim}\mathcal{N}\left(0,\frac{\tau_k^2}{1-\kappa_{k}^2}\right).
+\label{eq:u1}
+\end{align} This model choice is to preserve stationarity;
+i.e.~\(p(u_{k,gs}) = p(u_{k,gt})\) for \(s \neq t\).
+
+For the discrete phenotypes, we will assume \begin{align}
+{Y}_{l,ti} & \sim \left\{\begin{array}{ll} Poisson(\lambda_{l,ft}), & W_{ti} = \text{Female} \\ Poisson(\lambda_{l,mt}), & W_{ti} = \text{Male} \end{array}\right.,
+\label{eq:Y}
+\end{align} for \(l = 1,\cdots,L\). Similar to the continuous
+phenotypes, we set \begin{align}
+\log(\lambda_{l,gt}) = \gamma_{l,g} + v_{l,gt}.
+\label{eq:lambda}
+\end{align} We use the log function to allow \(v_{l,gt}\) to take all
+values on the real number line so we can properly fit an OU process to
+these effects. We further assume \begin{align}
+v_{l,gt} \overset{iid}{\sim}\mathcal{N}(\phi_{l} v_{l,g(t-1)} , \omega_l^2),
+\label{eq:v_ar1}
+\end{align} and \begin{align}
+v_{l,g1} \overset{iid}{\sim}\mathcal{N}\left(0,\frac{\omega_l^2}{1 - \phi_{l}^2}\right)..
+\label{eq:v1}
+\end{align}
+
+Because we are fitting a dataset with a stochastic structure on the
+means of the phenotypes, we analyze the data via a Bayesian analysis.
+Bayesian data analysis is also more naturally used when we have to
+impute data (\textbf{CITATION}).
+
+Priors: For \(k = 1,\cdots,K\) and \(l = 1,\cdots,L\),
 
 \begin{align}
-X_{ti} \overset{iid}{\sim}\left\{\begin{array}{ll} \mathcal{N}(\theta_f + u_{ft},\sigma^2), & W_{ti} = \text{Female} \\ \mathcal{N}(\theta_m + u_{mt},\sigma^2), & W_{ti} = \text{Male} \end{array}\right.
-\label{eq:x}
-\end{align}
-
-Look at \ref{eq:x}
-
-For \(g \in \{f,m\}\), \(\theta_g\), is the defined as the overall mean
-of the phenotype of interest for each gender, and \(u_{gt}\) are the
-gender specific random effects, measuring the difference between the
-gender and time-specific mean and the overall gender-specific mean,
-which we fit to an autoregressive process of lag 1, which is the
-discretization of an OU process. For \(t = 2,\ldots,T\) and
-\(g \in \{f,m\}\), the AR1 model for is defined as
-
-\begin{align}
-u_{g,t} \overset{iid}{\sim}\mathcal{N}(\kappa_g u_{g,t-1} , \tau^2).
-\label{eq:ar1}
-\end{align}
-
-\(\kappa_g\) represents the gender-specific correlation between the
-previous and current time periods' random effects. In addition,
-\(\text{cor}(u_{g,t},u_{g,t+h}) = (\kappa_g)^h\) for
-\(h \in \mathbb{N}\).
-
-We also set
-
-\begin{align}
-u_{g,1} \overset{iid}{\sim}\mathcal{N}\left(0,\frac{\tau^2}{1-\kappa_g^2}\right).
-\label{eq:u1}
-\end{align}
-
-This model choice is to make the distribution of the random effects
-stationary, i.e.~\(p(\mu_{g,s}) = p(\mu_{g,t})\) for
-\(s,t \in \{1,\dots,T\}\).
-
-Priors:
-
-\begin{align}
-\sigma & \sim \mathcal{N}(0,100)I_{\{\sigma > 0\}} \nonumber \\
-\tau & \sim \mathcal{N}(0,100)I_{\{\tau > 0\}} \nonumber \\
-\kappa & \overset{iid}{\sim}\mathcal{N}(0.5,1)I_{\{-1 < \kappa_g < 1\}} \nonumber \\
-\theta_g & \overset{iid}{\sim}\mathcal{N}(54,400)I_{\{\theta_g > 0\}}
+\sigma_k & \overset{iid}{\sim}\mathcal{N}(0,100)I_{\{\sigma > 0\}} \nonumber \\
+\tau_k & \overset{iid}{\sim}\mathcal{N}(0,100)I_{\{\tau > 0\}} \nonumber \\
+\kappa_k & \overset{iid}{\sim}\mathcal{N}(0,1)I_{\{-1 < \kappa_g < 1\}} \nonumber \\
+\beta_k & \overset{iid}{\sim}\mathcal{N}(0,5) \\
+\theta_{k,g} & \overset{iid}{\sim}\mathcal{N}(0,10000)
+\omega_l & \overset{iid}{\sim}\mathcal{N}(0,100)I_{\{\tau > 0\}} \nonumber \\
+\phi_l & \overset{iid}{\sim}\mathcal{N}(0,1)I_{\{-1 < \kappa_g < 1\}} \nonumber \\
+\gamma_{l,g} & \overset{iid}{\sim}\mathcal{N}(0,10000)
 \label{eq:priors}

### 2023-05-25 R/stuart_stickle_bayes.R
Add posterior samples and plots sorted by phenotype.

@@ -4,6 +4,7 @@ library(rstan)
 library(ggridges)
 
 dat <- read.csv("Data/stuart_stickle_100imputations_all.csv")
+dat$mds[dat$mds==1.2] <- 1
 dat$group <- factor(paste(dat$gender,letters[dat$time]))
 
 mod <- "
@@ -24,15 +25,16 @@ data{
   int<lower=0> mdf[N];
   int<lower=0> mav[N];
   int<lower=0> maf[N];
-  int<lower=0> mcf[N];
+  int<lower=0> mcv[N];
+  int<lower=0> mds[N];
   int<lower=0> mpt[N];
 }
 parameters{
-  real<lower=0> theta_f[15];
-  real<lower=0> theta_m[15];
-  real<lower=-1,upper=1> kappa[15];
+  real<lower=0> theta_f[16];
+  real<lower=0> theta_m[16];
+  real<lower=-1,upper=1> kappa[16];
   real<lower=0> sigma[10];
-  real<lower=0> tau[15];
+  real<lower=0> tau[16];
   real beta[9];
   vector[2*T] u_stl;
   vector[2*T] u_lps;
@@ -47,7 +49,8 @@ parameters{
   vector[2*T] u_mdf;
   vector[2*T] u_mav;
   vector[2*T] u_maf;
-  vector[2*T] u_mcf;
+  vector[2*T] u_mcv;
+  vector[2*T] u_mds;
   vector[2*T] u_mpt;
 }
 transformed parameters{
@@ -64,7 +67,8 @@ transformed parameters{
   vector[2*T] mu_mdf;
   vector[2*T] mu_mav;
   vector[2*T] mu_maf;
-  vector[2*T] mu_mcf;
+  vector[2*T] mu_mcv;
+  vector[2*T] mu_mds;
   vector[2*T] mu_mpt;
 
   mu_stl[1:T] = theta_f[1] + u_stl[1:T];
@@ -93,10 +97,12 @@ transformed parameters{
   mu_mav[(T+1):(2*T)] = theta_m[12] + u_mav[(T+1):(2*T)];
   mu_maf[1:T] = theta_f[13] + u_maf[1:T];
   mu_maf[(T+1):(2*T)] = theta_m[13] + u_maf[(T+1):(2*T)];
[…]

### 2023-05-25 R/stuart_stickle_bayes_plots.R

@@ -1,46 +1,60 @@
 library(tidyverse)
 library(ggridges)
 
-vars <- c("stl","cle","pmx")
+vars <- c("stl","lps","ect","tpg",
+          "cle","pmx","ds1","ds2",
+          "ds3","lpt","mdf","mav",
+          "maf","mcv","mds","mpt")
 
-for (x in vars){
-  samps <- readRDS(paste0("posterior_samples_",x,".RDS"))
+for (i in 1:16){
+  phenotype = vars[i]
 
-  xlim <- if (x == "stl"){
-    xlim(30,80)
-  } else if (x == "cle"){
-    xlim(5,12)
+  thetalab <- if (i <= 10){
+    xlab(expression(theta[g]))
   } else {
-    xlim(25,85)
+    xlab(expression(gamma[g]))
   }
 
   samps %>%
-    rename(Female = theta_f) %>%
-    rename(Male = theta_m) %>%
+    rename(Female = paste0("theta_f[",i,"]")) %>%
+    rename(Male = paste0("theta_m[",i,"]")) %>%
     select(Female,Male) %>%
     gather("Gender","Value") %>%
     ggplot() +
     geom_density(aes(x=Value,color=Gender)) +
     scale_color_manual(values = c("red","blue"))+
     theme_bw() +
-    xlim +
-    xlab(expression(theta[g])) +
+    thetalab +
     ylab(expression(paste("p(",theta[g],"|",bold(X),")")))
-  ggsave(paste0("Figures/",x,"/theta.pdf"),width=5,height=8)
+  if (i <= 10){
+    ggsave(paste0("Figures/",phenotype,"/theta.pdf"),width=5,height=8)
+  } else {
+    ggsave(paste0("Figures/",phenotype,"/gamma.pdf"),width=5,height=8)
+  }
 
-  length(which(samps$theta_m > samps$theta_f))/nrow(samps)
-
-  mu_f <- samps %>% select(contains("mu_f"))
-  mu_m <- samps %>% select(contains("mu_m"))
+  mu <- samps %>% select(contains(paste0("mu_",phenotype)))
+  if (i >= 11) {mu <- exp(mu)}
+  mu_f <- mu[1:T]
+  mu_m <- mu[(T+1):(2*T)]
 
   mu_diff <- mu_m - mu_f
 
-  xlim <- if (x == "stl"){
-    xlim(-15,25)
[…]

### 2023-05-31 manuscript/paper.Rmd
Updated a bunch of bib engtries and added Joel's text to the document.

@@
 title: |
-  |  \Large Stickle!
+  |  \Large Increased sexual dimorphism evolves in a fossil stickleback following ecological release from fish piscivores
 author: | 
-  | \large Matthew Stuart \vspace{-1.1mm}
-  | \normalsize Department of Mathematics and Statistics \vspace{-1.1mm}
-  | \normalsize Loyola University Chicago \vspace{-1.1mm}
-  | \normalsize Chicago, IL 60660 \vspace{-1.1mm}
-  | \normalsize [`mstuart1@luc.edu`](mailto:mstuart1@luc.edu) \vspace{-1.1mm}
+  | \large Allison Ozark$^1$, Akhil Ghosh$^1$, Raheyma Siddiqui$^1$, Matthew Stuart$^1$,
+  | \large  Samantha Swank$^1$, Michael A. Bell$^2$, Gregory J. Matthews$^1$, and Yoel E. Stuart$^1$ \vspace{-1.1mm}
   |
-  | \large Akhil Ghosh \vspace{-1.1mm}
-  | \normalsize Department of Mathematics and Statistics \vspace{-1.1mm}
-  | \normalsize Loyola University Chicago \vspace{-1.1mm}
-  | \normalsize Chicago, IL 60660 \vspace{-1.1mm}
-  | \normalsize [`aghosh@luc.edu`](mailto:aghosh@luc.edu) \vspace{-1.1mm}
+  | \large $^1$ Loyola University Chicago, Chicago, IL, USA \vspace{-1.1mm}
   |
-  | \large Yoel E. Stuart \vspace{-1.1mm}
-  | \normalsize Department of Biology \vspace{-1.1mm}
-  | \normalsize Loyola University Chicago \vspace{-1.1mm}
-  | \normalsize Chicago, IL 60660 \vspace{-1.1mm}
-  | \normalsize [`ystuart@luc.edu`](mailto:ystuart@luc.edu) \vspace{-1.1mm}
+  | \large $^2$ University of California Museum of Paleontology, Berkeley, CA, USA \vspace{-1.1mm}
   |
-  | \large Gregory J. Matthews \vspace{-1.1mm}
-  | \normalsize Center for Data Science and Consulting; Department of Mathematics and Statistics \vspace{-1.1mm}
-  | \normalsize Loyola University Chicago \vspace{-1.1mm}
-  | \normalsize Chicago, IL 60660 \vspace{-1.1mm}
-  | \normalsize [`gmatthews1@luc.edu`](mailto:gmatthews1@luc.edu) \vspace{-1.1mm}
 abstract: |
-  | Evewryone loves the stickle \vspace{2mm}
+  | Everyone loves the stickle \vspace{2mm}
   | *Keywords*: Stickle
@@
 # Introduction {#sec:intro}
+
+Ecological release theory suggests that a prey population’s niche should expand when predation is relaxed or removed (reviewed in @Herrmann2021).
+Niche changes in habitat and resource use should then drive phenotypic evolution as the population adapts to new conditions (@Herrmann2021).
+For example, resource use expansion should result in trophic trait evolution (cite; e.g.,).
+Habitat expansion should result in locomotor change (cite; e.g.,).
+And release from enemies should result in reduction of costly defense traits (cite; e.g.,).
+
+Niche expansion could manifest through a combination of increasing within- and between-individual niche widths (@Bolnick2010; @Herrmann2021), including divergence between the sexes (@Bolnick2003; @Cooper2011).
+Theory and empirical data suggest that a population presented with novel ecological opportunity might experience disruptive selection on males and females stemming from intraspecific competition over newly accessible resources, resulting in intersexual divergence in habitat use and associated phenotypes (Schoener 1968; @Shine1989; @Bolnick2003; @Butler2007; @Bolnick2008; @Cooper2011; but see @Stuart2021; @Blain2022).
+Such character displacement between the sexes is therefore one explanation for sexual dimorphism (de Lisle papers).
+
+We tested this link between release from predators and the evolution of sexual dimorphism in a well-preserved, high resolution lineage of the fossil stickleback fish (Gasterosteus doryssus).
+The sex of the individual fossils, which is unobserved in practice, was imputed using multiple imputation by chained equations (MICE) (@MICE) with observed data used to train the imputation model based on extant stickleback with known sex.
+Then we tracked the multivariate evolution of sexual dimorphism over ~16,000 years for traits related to swimming, feeding and defense.
+The fossil sequence captures a bout of adaptive evolution following ecological release from predators, as follows.
+The fossil lineage appeared in the depositional environment as a fully armored form, with complete pelvic girdles, two pelvic spines, and three dorsal spines, on average (@Bell2006; @Stuart2020).
+However, this habitat appears to have been missing fish piscivores (e.g., trout and other salmonids known to prey on modern stickleback; @Bell2009), relaxing putative selection for armor (Reimchen papers, Bell papers, @Bowne1994; Roesti et al.
+2023).
+The lineage thus adaptively evolved armor loss and reduction in several other traits as well (@Hunt2008; @Stuart2020; Siddiqui et al.
+in prep), and tooth wear data suggest that the population also began eating more planktonic prey, expanding to an open water lifestyle (Purnell et al.
+2007).
+
 Sexual Dimorphism is interesting because......
@@
 
-Studying sexual dimorphism of phenotypes in modern species is a straightforward endeavor: measure a phenotype of interest and statistically test for differences between the sexes.
-However, examining sexual dimorphism in fossils is complicated substantially by the issue that sex may not be able to be directly observed from fossil specimens.
-For this study, we have two sources of data: 1) Modern stickleback specimens with observed sex and 2) fossil specimens with sex unobserved.
-We view the sex of the fossils as missing data and use multiple imputation (@little2002statistical) to impute the sex of the fossils using the modern stickleback fish with observed sex to model the relationship between sex and observed phenotypes.
-Once sex is imputed, an Ornstein–Uhlenbeck (OU) model is fit using a Bayesian framework to look for sexual dimorphism across a variety of stickleback phenotypes (e.g.
-length, vertebrae count?, etc).
-
 The remainder of this manuscript contains a description of the data in section \ref{sec:data} and a description of our models in \ref{sec:models}.
@@
 
+# Methods and Materials 
+<!-- Studying sexual dimorphism of phenotypes in modern species is a straightforward endeavor: measure a phenotype of interest and statistically test for differences between the sexes.
+However, examining sexual dimorphism in fossils is complicated substantially by the issue that sex may not be able to be directly observed from fossil specimens.
+For this study, we have two sources of data: 1) Modern stickleback specimens with observed sex and 2) fossil specimens with sex unobserved.
+-->
+
+A major challenge in this work is inferring the sex of fossil specimens.
+This typically cannot be done directly, except for lineages whose sexes are distinguished by the presence or absence of sex-specific characters (@Hone2017; @SaittaEtAl2020).
+Instead, paleobiologists resort to statistical detection of sex and sexual dimorphism, including tests for bimodality in trait distributions (e.g., @Hone2017) and divergence in growth curves (e.g., @SaittaEtAl2020).
+
+We used a third approach to infer fossil sex and study the evolution of sexual dimorphism by taking advantage of the fact that G.
+doryssus is part of the extant threespine stickleback species complex (Gastersteus aculeatus).
+Along with the fossil data collected for this study, we also have data from extant populations of the Threespine Stickleback, Gasterosteus aculeatus, measuring the same traits for individuals whose sex could be determined directly.
+We then combined these sets of data and treat the sex of the fossil specimens as missing and  used multiple imputation (@little2002statistical) to impute the sex of the fossils many times.
+Imputations were performed using the MICE (@MICE) algorithm and implemented in R (@R2022language) with M = 100 imputations performed.
+
+After imputing the sex of the fossil data, each completed data set is then used to fit a modified Ornstein–Uhlenbeck (OU) (@OUProcess) model using a Bayesian framework to look for sexual dimorphism across a variety of stickleback phenotypes
+
 # Data {#sec:data}
-
 ```{r}
@@
 
-The data used here consists of a total of `r sum(!is.na(stick$gender))` extant specimens with known sex all collected in the last 30 years.
-Of these, there are `r sum(stick$gender == "Female",na.rm = TRUE)` and `r sum(stick$gender == "Male",na.rm = TRUE)` female and male specimens, respectively.
+## Fossil Specimen Data
+We used Gasterosteus doryssus data that were previously reported by @Stuart2020, @Voje2022, and Siddiqui et al.
+(in review).
+Briefly, the data were collected from fossil Series K from Quarry D (Cerasoni et al., in review), dug from an open pit diatomite mine at 9.526° N, 119.094° W, near Fernley, Nevada, USA.
+Series K consisted of 18 samples taken at ~1000-year intervals, and mean sample times span ~16,363 years.
+Fish from series K were measured for 16 ecomorphological traits related to armor, swimming, and feeding (Table 1).
+Series K started at a previously documented horizon when a low armored lineage of stickleback with zero to one dorsal spines, zero pelvic spines, and highly reduced pelvises was completely replaced by a high armored lineage of stickleback with three dorsal spines, two pelvic spines, and complete (Bell et al.
+2006; Bell 2009; Stuart et al.
+2020).
+This lineage subsequently evolved reduction in armor, body size, and traits related to swimming and feeding (Bell et al.
+2006; Stuart et al.
+2020; Siddiqui et al.
+in review).
+The tempo and mode of armor reduction during this sequence suggests adaptive evolution by natural selection (Hunt et al.
+2008), and we focus on the multivariate evolution of sexual dimorphism by this second lineage.
 
-In addition there are `r sum(is.na(stick$gender))` fossil specimens from approximately 10.3 million years ago with unknown sex over 18 time periods spaced about 1000 years apart.
-Table 1 shows the sample size at each of the 18 time periods.
+The fossil data consists of `r sum(is.na(stick$gender))` specimens with unknown sex over 18 time periods spaced  ~1000 years apart.
+Table 2 shows the sample size at each of the 18 time periods.
 There are at least 22 specimens at each time period with a high of 67 specimens in period 7.
+
+| Trait Name           | Trait Code | Trait Description           |
+|----------------------|------------|-----------------------------|
+|Standard Length|	sl|	Distance from anterior tip of premaxilla to posterior end of last vertebra (hypural plate)|
+|Dorsal Spine | mds |	Number of dorsal spines from 0 to 3|
+|Dorsal Fin Ray|mdf|	Number of bones in the dorsal fin posterior to the third dorsal spine (i.e., soft dorsal fin rays)
+|Anal Fin Ray |maf|	Number of bones in the anal fin posterior to the anal spine (i.e., soft dorsal fin rays)|
+|Abdominal Vertebra|mav|	Number of vertebrae anterior to the first vertebra contacting an anal fin pterygiophore (Aguirre et al. 2014)|
+|Caudal Vertebra|mcv	| Number of vertebrae posterior to and including the first vertebra contacting an anal fin pterygiophore (Aguirre et al. 2014)|
+|Pterygiophore number|mpt|	Number of pterygiophores anterior to but excluding the pterygiophore under the third dorsal spine, which is immediately anterior to and contiguous with the dorsal fin|
+|Pelvic Spine length|lps.sc	|Length from the base of one pelvic spine above its articulation with the pelvic girdle to its distal tip|
+|Ectocoracoid|ect.sc|	Length between the anterior and posterior tips of the shoulder girdle base (i.e., ectocoracoid)|
+|Pelvic Girdle|tpg.sc|	Length between the anterior to posterior tips along midline. If vestigial, the sum of longest anterior-posterior axis for the vestiges|
+|Cleithrum length|cle.sc|	Length from free dorsal tip to ventral tip of the cleithrum on the anterior margin of the shoulder girdle (i.e., cleithrum)|
+|Premaxilla|pmx.sc|	Length from the anterior tip of the premaxilla to the distal tip of the ascending process of the premaxilla |
+|Dorsal Spine|Ds#.sc# = 1,2,or 3|	Length from the base of a dorsal spine above the pterygiophore to its distal tip along the anterior edge|
+|Pterygiophore|lpt.sc|	Distance between the anterior to posterior tips of the pterygiophore immediately preceding the 3rd dorsal spine (when present)|
+
+Table: Traits and trait descriptions. ‘sc’ denotes size correction of trait against standard length. Names of bones follow @Bowne1994 unless otherwise noted.
 
@@
 \centering
+\caption{The number of stickleback fossils at each time point.  Sample size at each time point ranges from a low of 22 to a high of 67.}
 \begin{tabular}{rl}
@@
 \end{tabular}
-\label{tab1}
-\caption{The number of stickleback fossils at each time point.  Sample size at each time point ranges from a low of 22 to a high of 67.}
+\label{tabn}
+
 \end{table}
+
+## Extant Specimen data
+To span the gamut of stickleback diversity for our predictive model, we sampled modern stickleback from lakes containing generalist stickleback populations (@Hendry2009; @Bolnick2011) and from lakes containing benthic-limnetic species pairs (Baumgartner et al.
+1988; @Schluter1992) (Table 2 (What is this?)).
+The generalist populations were collected by YES in 2013 and previously described in @Stuart2017.
+These samples were fixed in formalin, then stained for bone with Alizarin Red in 2013.
+Benthic and limnetic specimens were kindly loaned by D.
+Schluter and his lab at University of British Columbia.
+They collected benthic and limnetic individuals from Enos Lake in 1988 and from Emily Lake, Little Quarry Lake, Paxton Lake, and Priest Lake in 2018.
+The Enos specimens had been fixed whole in formalin and stored in 40% isopropanol.
+The specimens from the other lakes were initially preserved whole in 95% ethanol in the field before being gradually transferred to water then formalin in the lab and ultimately stored in 40% isopropanol.
+In 2019, we stained these specimens for bone using Alizarin Red.
+
+We next replicated fossil data collection (Table 1) on these extant specimens.
+Standard length as well as pelvic-spine length on each side were measured with calipers.
+We used a dissection microscope to count dorsal spines, pelvic spines, dorsal-fin rays, and anal-fin rays.
+Right and left-side pelvic girdle lengths and ectocoracoid lengths were measured from ventral photographs taken using a Canon EOS Rebel T7 with a Tamron 16-300 mm MACRO lens mounted on a leveled Kaiser RS1 copy stand.
+Specimens were held in place for ventral photographs using a small tabletop vise with an attached scale bar.
+Lateral X-rays were used to measure dorsal spine length, number of pterygiophores anterior to the pterygiophore holding the third spine, length of the pterygiophore just anterior to the third spine, cleithrum length, and pre-maxilla ascending branch length.
+We also counted vertebrae from the X-rays: abdominal vertebrae were counted anterior to the first vertebra with a haemal spine contacting an anal fin pterygiophore.
+Caudal vertebrae were posterior, including the first vertebra with the haemal spine contacting the anal fin pterygiophore (following @Aguirre2014.
+X-rays were taken with an AXR Hot Shot X-ray Machine (Associated X-ray Corporation) at the Field Museum of Natural History.
+Specimens were exposed at 35kV and 4mA.
+Small fish were exposed for 7s, medium fish for 8s, and large fish for 10s.
+We developed the film and scanned individual images of each fish using the B&W Negatives setting on an Epson Perfection 4990 Photo flatbed at 2400 dpi.
+Measurements from photographs and X-rays were taken with FIJI (Schindelin et al.
+2012) and its plugin ObjectJ (https://sils.fnwi.uva.nl/bcb/objectj/).
+All photographs, X-rays, and ObjectJ files have been uploaded to Morphosource.org (accession # TBD).
+We dissected individuals from the generalist populations to determine sex from the gonads.
+Individuals from the species-pair lakes were sexed by Schluter and his lab, using a genotyping protocol (confirm and cite).
+
+The extant data used here consists of a total of `r sum(!is.na(stick$gender))` specimens all with known sex.
+Of these, there are `r sum(stick$gender == "Female",na.rm = TRUE)` and `r sum(stick$gender == "Male",na.rm = TRUE)` female and male specimens, respectively.
+
+## Outlier analysis. 
+To check for outliers, we calculated within-group means and standard deviations for each trait separately for K series fossil specimens (pooled across samples) and for extant specimens (pooled across lakes).
+We noted trait values greater than 3.5 standard deviations from the mean as potential outliers.
+We checked whether these potential outliers were a result of data entry and collection error and corrected them if they were.
+We turned the remaining outlier trait values to NAs.
+(confirm)  (Wait, what?
+Why?
+Anything that was 3.5 SD above the mean was recorded as missing?
+How do you justify this?
+)
+
+Missing data imputation, including fossil sex
+
+Quantification and evolution of sexual dimorphism
 
@@
 
-We impute the missing gender for the fossil data by sampling from the posterior predictive distribution $P(\boldsymbol{W}_{fossil}|\boldsymbol{W}_{extant}, \boldsymbol{X}, \boldsymbol{Y})$ using the multiple imputation with chained equations (MICE) algorithm (@MICE) with predictive mean matching.
-The imputation algorithm is run to obtain a total of $M = 100$ completed datasets.
+We impute the missing gender for the fossil data by sampling from the posterior predictive distribution $P(\boldsymbol{W}_{fossil}|\boldsymbol{W}_{extant}, \boldsymbol{X}, \boldsymbol{Y})$ using the multiple imputation by chained equations (MICE) algorithm (@MICE) with predictive mean matching.
+Traditionally, the choice for the number of completed dats sets is a relatively small number such as $M = 5$ or $M = 10$.
+However, @ZhouReiter2009 recommend a larger number of imputed data sets if the data users intend on performing Bayesians analysis after imputation, which in this case, we do.
+Therefore, the imputation algorithm is run to obtain a total of $M = 100$ completed datasets.
+In addition to this, @ZhouReiter2009 suggests rather than using Rubin's combining rules to combine across imputed data sets, instead pool all of the draws from the posterior distributions across all of the imputed data sets to estimate the posterior distributions of parameters of interest.
+We proceed with our Bayesian analysis in this manner.
+
+\textcolor{red}{Akhil's stuff about validating the imputation model goes here.}
+
+<!-- Bayesian Analysis after multiple imputation @ZhouReiter2009: They recommend using a large number of imputations.  5 or 10 is too small.  We are using M = 100. -->
 
 ## Completed Data Model 
+\textcolor{red}{We should note somewhere that we are only using the $W_{fossil}$ in the modeling part.
+We drop the $W_{extant}$.
+So just note that $W_{ti}$ is really $W_{fossil,ti}$.
+Not sure how to say this, but we need to make it clear that we are only using the fossil data for the OU modeling.
+}
+
+### Continuous Phenotypes
 For a given imputed dataset, let $W_{ti}$ be the imputed gender, $\boldsymbol{X}_{ti}$ be the $K \times 1$ vector of continuous phenotypes, and $\boldsymbol{Y}_{ti}$ be the $L \times 1$ vector of discrete phenotypes for $t = 1, \ldots, T$ and  $i = 1\ldots,n_{t}$).
@@
 for $g \in \{f,m\}$ where $\theta_{k,g}$ is the overall mean of phenotype $X_k$ for each gender, and $u_{k,gt}$ measures the difference between $\mu_{k,gt}$ and $\theta_{k,g}$.
-We then fit $\boldsymbol{u}_{k,g} = \{u_{k,g1},\cdots,u_{k,gT}\}$ to an Ornstein-Uhlenback (OU) process (@OUprocess) where we assume $u_{k,gT}$ have a marginal mean of 0.
+We then fit $\boldsymbol{u}_{k,g} = \{u_{k,g1},\cdots,u_{k,gT}\}$ to an Ornstein-Uhlenback (OU) process (@OUProcess) where we assume $u_{k,gT}$ have a marginal mean of 0.
 Because each time period is ~1000 years, we will discretize the OU process without loss of generality.
@@
 
+### Discrete Phenotypes
 For the discrete phenotypes, we will assume
@@
 
-# Future work and conclusions {#sec:conclusions}
+# Discsusson, Future work and conclusions {#sec:conclusions}
+We predicted that release from predators would result in niche expansion and increased sexual dimorphism, based on several studies of modern stickleback.
+For example, in lakes where sculpin competitors are absent and stickleback (Roesti et al.
+2023)
+See Spoljaric and Reimchen 2008, page 512 right column for references and discussion of differences between benthic males and limnetic females.
+Male stickleback are benthic and littoral (Wootton 1976)….
+Reimchen papers in general good for this section.
 
@@
 
-Stickle!
+We thank O.
+Abughoush, S.
+Blaine, A.
+Chaudhary, M.
+Islam, F.
+Joaquin, C.
+Lawson-Weinert, R.
+Sullivan, J.
+Tien, M.P.
+Travis, and W.
+Shim for help with data collection.
+We thank D.
+Schluter and S.
+Blain for loaning specimens and sharing data.
+We thank K.
+Swagel and C.
+McMahan of the Field Museum for assistance with specimen x-rays.
+This research was supported by NSF grants BSR-8111013, EAR-9870337, and DEB-0322818, the Center for Field Research (Earthwatch), and the National Geographic Society (2869-84) to MAB.
+It was also supported by NSF grants DEB-1456462 and EAR-2145830 to YES.
+And NSF DMS-2015374 (GJM)
 

### 2023-06-01 manuscript/paper.tex
Fix notation mistake in priors.

@@
 \usepackage[margin=1in]{geometry}
+\usepackage{longtable,booktabs,array}
+\usepackage{calc} % for calculating minipage widths
+% Correct order of tables after \paragraph or \subparagraph
+\usepackage{etoolbox}
+\makeatletter
+\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
+\makeatother
+% Allow footnotes in longtable head/foot
+\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
+\makesavenoteenv{longtable}
 \usepackage{graphicx}
@@
 
-\title{~\Large Stickle!}
-\author{\large Matthew Stuart \vspace{-1.1mm}\\
-\normalsize Department of Mathematics and Statistics \vspace{-1.1mm}\\
-\normalsize Loyola University Chicago \vspace{-1.1mm}\\
-\normalsize Chicago, IL 60660 \vspace{-1.1mm}\\
-\normalsize \href{mailto:mstuart1@luc.edu}{\texttt{mstuart1@luc.edu}}
+\title{~\Large Increased sexual dimorphism evolves in a fossil
+stickleback following ecological release from fish piscivores}
+\author{\large Allison Ozark\(^1\), Akhil Ghosh\(^1\), Raheyma
+Siddiqui\(^1\), Matthew Stuart\(^1\),\\
+\large  Samantha Swank\(^1\), Michael A. Bell\(^2\), Gregory J.
+Matthews\(^1\), and Yoel E. Stuart\(^1\) \vspace{-1.1mm}\\
+\strut \\
+\large \(^1\) Loyola University Chicago, Chicago, IL, USA
 \vspace{-1.1mm}\\
 \strut \\
-\large Akhil Ghosh \vspace{-1.1mm}\\
-\normalsize Department of Mathematics and Statistics \vspace{-1.1mm}\\
-\normalsize Loyola University Chicago \vspace{-1.1mm}\\
-\normalsize Chicago, IL 60660 \vspace{-1.1mm}\\
-\normalsize \href{mailto:aghosh@luc.edu}{\texttt{aghosh@luc.edu}}
-\vspace{-1.1mm}\\
-\strut \\
-\large Yoel E. Stuart \vspace{-1.1mm}\\
-\normalsize Department of Biology \vspace{-1.1mm}\\
-\normalsize Loyola University Chicago \vspace{-1.1mm}\\
-\normalsize Chicago, IL 60660 \vspace{-1.1mm}\\
-\normalsize \href{mailto:ystuart@luc.edu}{\texttt{ystuart@luc.edu}}
-\vspace{-1.1mm}\\
-\strut \\
-\large Gregory J. Matthews \vspace{-1.1mm}\\
-\normalsize Center for Data Science and Consulting; Department of
-Mathematics and Statistics \vspace{-1.1mm}\\
-\normalsize Loyola University Chicago \vspace{-1.1mm}\\
-\normalsize Chicago, IL 60660 \vspace{-1.1mm}\\
-\normalsize \href{mailto:gmatthews1@luc.edu}{\texttt{gmatthews1@luc.edu}}
-\vspace{-1.1mm}}
+\large \(^2\) University of California Museum of Paleontology, Berkeley,
+CA, USA \vspace{-1.1mm}\\}
 \date{}
@@
 \begin{abstract}
-Evewryone loves the stickle \vspace{2mm}\\
+Everyone loves the stickle \vspace{2mm}\\
 \emph{Keywords}: Stickle
@@
 
+Ecological release theory suggests that a prey population's niche should
+expand when predation is relaxed or removed (reviewed in Herrmann,
+Stroud, and Losos (\protect\hyperlink{ref-Herrmann2021}{2021})). Niche
+changes in habitat and resource use should then drive phenotypic
+evolution as the population adapts to new conditions (Herrmann, Stroud,
+and Losos (\protect\hyperlink{ref-Herrmann2021}{2021})). For example,
+resource use expansion should result in trophic trait evolution (cite;
+e.g.,). Habitat expansion should result in locomotor change (cite;
+e.g.,). And release from enemies should result in reduction of costly
+defense traits (cite; e.g.,).
+
+Niche expansion could manifest through a combination of increasing
+within- and between-individual niche widths (Bolnick et al.
+(\protect\hyperlink{ref-Bolnick2010}{2010}); Herrmann, Stroud, and Losos
+(\protect\hyperlink{ref-Herrmann2021}{2021})), including divergence
+between the sexes (Bolnick and Doebeli
+(\protect\hyperlink{ref-Bolnick2003}{2003}); Cooper, Gilman, and
+Boughman (\protect\hyperlink{ref-Cooper2011}{2011})). Theory and
+empirical data suggest that a population presented with novel ecological
+opportunity might experience disruptive selection on males and females
+stemming from intraspecific competition over newly accessible resources,
+resulting in intersexual divergence in habitat use and associated
+phenotypes (Schoener 1968; Shine
+(\protect\hyperlink{ref-Shine1989}{1989}); Bolnick and Doebeli
+(\protect\hyperlink{ref-Bolnick2003}{2003}); Butler, Sawyer, and Losos
+(\protect\hyperlink{ref-Butler2007}{2007}); Bolnick and Lau
+(\protect\hyperlink{ref-Bolnick2008}{2008}); Cooper, Gilman, and
+Boughman (\protect\hyperlink{ref-Cooper2011}{2011}); but see Stuart et
+al. (\protect\hyperlink{ref-Stuart2021}{2021}); Blain
+(\protect\hyperlink{ref-Blain2022}{2022})). Such character displacement
+between the sexes is therefore one explanation for sexual dimorphism (de
+Lisle papers).
+
+We tested this link between release from predators and the evolution of
+sexual dimorphism in a well-preserved, high resolution lineage of the
+fossil stickleback fish (Gasterosteus doryssus). The sex of the
+individual fossils, which is unobserved in practice, was imputed using
+multiple imputation by chained equations (MICE) (Buuren and
+Groothuis-Oudshoorn (\protect\hyperlink{ref-MICE}{2011})) with observed
+data used to train the imputation model based on extant stickleback with
+known sex. Then we tracked the multivariate evolution of sexual
+dimorphism over \textasciitilde16,000 years for traits related to
+swimming, feeding and defense. The fossil sequence captures a bout of
+adaptive evolution following ecological release from predators, as
+follows. The fossil lineage appeared in the depositional environment as
+a fully armored form, with complete pelvic girdles, two pelvic spines,
+and three dorsal spines, on average (Michael A. Bell, Travis, and Blouw
+(\protect\hyperlink{ref-Bell2006}{2006}); Stuart, Travis, and Bell
+(\protect\hyperlink{ref-Stuart2020}{2020})). However, this habitat
+appears to have been missing fish piscivores (e.g., trout and other
+salmonids known to prey on modern stickleback; M. A. Bell
+(\protect\hyperlink{ref-Bell2009}{2009})), relaxing putative selection
+for armor (Reimchen papers, Bell papers, Bowne
+(\protect\hyperlink{ref-Bowne1994}{1994}); Roesti et al.~2023). The
+lineage thus adaptively evolved armor loss and reduction in several
+other traits as well (Hunt, Bell, and Travis
+(\protect\hyperlink{ref-Hunt2008}{2008}); Stuart, Travis, and Bell
+(\protect\hyperlink{ref-Stuart2020}{2020}); Siddiqui et al.~in prep),
+and tooth wear data suggest that the population also began eating more
+planktonic prey, expanding to an open water lifestyle (Purnell et
+al.~2007).
+
 Sexual Dimorphism is interesting because\ldots\ldots{}
@@
 (\protect\hyperlink{ref-SaittaEtAl2020}{2020})
-
-Studying sexual dimorphism of phenotypes in modern species is a
-straightforward endeavor: measure a phenotype of interest and
-statistically test for differences between the sexes. However, examining
-sexual dimorphism in fossils is complicated substantially by the issue
-that sex may not be able to be directly observed from fossil specimens.
-For this study, we have two sources of data: 1) Modern stickleback
-specimens with observed sex and 2) fossil specimens with sex unobserved.
-We view the sex of the fossils as missing data and use multiple
-imputation (Little and Rubin
-(\protect\hyperlink{ref-little2002statistical}{2002})) to impute the sex
-of the fossils using the modern stickleback fish with observed sex to
-model the relationship between sex and observed phenotypes. Once sex is
-imputed, an Ornstein--Uhlenbeck (OU) model is fit using a Bayesian
-framework to look for sexual dimorphism across a variety of stickleback
-phenotypes (e.g.~length, vertebrae count?, etc).
 
@@
 
+\hypertarget{methods-and-materials}{%
+\section{Methods and Materials}\label{methods-and-materials}}
+
+A major challenge in this work is inferring the sex of fossil specimens.
+This typically cannot be done directly, except for lineages whose sexes
+are distinguished by the presence or absence of sex-specific characters
+(Hone and Mallon (\protect\hyperlink{ref-Hone2017}{2017}); Saitta et al.
+(\protect\hyperlink{ref-SaittaEtAl2020}{2020})). Instead,
+paleobiologists resort to statistical detection of sex and sexual
+dimorphism, including tests for bimodality in trait distributions (e.g.,
+Hone and Mallon (\protect\hyperlink{ref-Hone2017}{2017})) and divergence
+in growth curves (e.g., Saitta et al.
+(\protect\hyperlink{ref-SaittaEtAl2020}{2020})).
+
+We used a third approach to infer fossil sex and study the evolution of
+sexual dimorphism by taking advantage of the fact that G. doryssus is
+part of the extant threespine stickleback species complex (Gastersteus
+aculeatus). Along with the fossil data collected for this study, we also
+have data from extant populations of the Threespine Stickleback,
+Gasterosteus aculeatus, measuring the same traits for individuals whose
+sex could be determined directly. We then combined these sets of data
+and treat the sex of the fossil specimens as missing and used multiple
+imputation (Little and Rubin
+(\protect\hyperlink{ref-little2002statistical}{2002})) to impute the sex
+of the fossils many times. Imputations were performed using the MICE
+(Buuren and Groothuis-Oudshoorn (\protect\hyperlink{ref-MICE}{2011}))
+algorithm and implemented in R (R Core Team
+(\protect\hyperlink{ref-R2022language}{2022})) with M = 100 imputations
+performed.
+
+After imputing the sex of the fossil data, each completed data set is
+then used to fit a modified Ornstein--Uhlenbeck (OU) (Uhlenbeck and
+Ornstein (\protect\hyperlink{ref-OUProcess}{1930})) model using a
+Bayesian framework to look for sexual dimorphism across a variety of
+stickleback phenotypes
+
 \hypertarget{sec:data}{%
@@
 
-The data used here consists of a total of 367 extant specimens with
-known sex all collected in the last 30 years. Of these, there are 202
-and 165 female and male specimens, respectively.
+\hypertarget{fossil-specimen-data}{%
+\subsection{Fossil Specimen Data}\label{fossil-specimen-data}}
 
-In addition there are 814 fossil specimens from approximately 10.3
-million years ago with unknown sex over 18 time periods spaced about
-1000 years apart. Table 1 shows the sample size at each of the 18 time
-periods. There are at least 22 specimens at each time period with a high
-of 67 specimens in period 7.
+We used Gasterosteus doryssus data that were previously reported by
+Stuart, Travis, and Bell (\protect\hyperlink{ref-Stuart2020}{2020}),
+Voje, Bell, and Stuart (\protect\hyperlink{ref-Voje2022}{2022}), and
+Siddiqui et al.~(in review). Briefly, the data were collected from
+fossil Series K from Quarry D (Cerasoni et al., in review), dug from an
+open pit diatomite mine at 9.526° N, 119.094° W, near Fernley, Nevada,
+USA. Series K consisted of 18 samples taken at \textasciitilde1000-year
+intervals, and mean sample times span \textasciitilde16,363 years. Fish
+from series K were measured for 16 ecomorphological traits related to
+armor, swimming, and feeding (Table 1). Series K started at a previously
+documented horizon when a low armored lineage of stickleback with zero
+to one dorsal spines, zero pelvic spines, and highly reduced pelvises
+was completely replaced by a high armored lineage of stickleback with
+three dorsal spines, two pelvic spines, and complete (Bell et al.~2006;
+Bell 2009; Stuart et al.~2020). This lineage subsequently evolved
+reduction in armor, body size, and traits related to swimming and
+feeding (Bell et al.~2006; Stuart et al.~2020; Siddiqui et al.~in
+review). The tempo and mode of armor reduction during this sequence
+suggests adaptive evolution by natural selection (Hunt et al.~2008), and
+we focus on the multivariate evolution of sexual dimorphism by this
+second lineage.
+
+The fossil data consists of 814 specimens with unknown sex over 18 time
+periods spaced \textasciitilde1000 years apart. Table 2 shows the sample
+size at each of the 18 time periods. There are at least 22 specimens at
+each time period with a high of 67 specimens in period 7.
+
+\begin{longtable}[]{@{}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.3492}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.1905}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4603}}@{}}
+\caption{Traits and trait descriptions. `sc' denotes size correction of
+trait against standard length. Names of bones follow Bowne
+(\protect\hyperlink{ref-Bowne1994}{1994}) unless otherwise
+noted.}\tabularnewline
+\toprule()
+\begin{minipage}[b]{\linewidth}\raggedright
+Trait Name
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Trait Code
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Trait Description
+\end{minipage} \\
+\midrule()
+\endfirsthead
+\toprule()
+\begin{minipage}[b]{\linewidth}\raggedright
+Trait Name
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Trait Code
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Trait Description
+\end{minipage} \\
+\midrule()
+\endhead
+Standard Length & sl & Distance from anterior tip of premaxilla to
+posterior end of last vertebra (hypural plate) \\
+Dorsal Spine & mds & Number of dorsal spines from 0 to 3 \\
+Dorsal Fin Ray & mdf & Number of bones in the dorsal fin posterior to
+the third dorsal spine (i.e., soft dorsal fin rays) \\
+Anal Fin Ray & maf & Number of bones in the anal fin posterior to the
+anal spine (i.e., soft dorsal fin rays) \\
+Abdominal Vertebra & mav & Number of vertebrae anterior to the first
+vertebra contacting an anal fin pterygiophore (Aguirre et al.~2014) \\
+Caudal Vertebra & mcv & Number of vertebrae posterior to and including
+the first vertebra contacting an anal fin pterygiophore (Aguirre et
+al.~2014) \\
+Pterygiophore number & mpt & Number of pterygiophores anterior to but
+excluding the pterygiophore under the third dorsal spine, which is
+immediately anterior to and contiguous with the dorsal fin \\
+Pelvic Spine length & lps.sc & Length from the base of one pelvic spine
+above its articulation with the pelvic girdle to its distal tip \\
+Ectocoracoid & ect.sc & Length between the anterior and posterior tips
+of the shoulder girdle base (i.e., ectocoracoid) \\
+Pelvic Girdle & tpg.sc & Length between the anterior to posterior tips
+along midline. If vestigial, the sum of longest anterior-posterior axis
+for the vestiges \\
+Cleithrum length & cle.sc & Length from free dorsal tip to ventral tip
+of the cleithrum on the anterior margin of the shoulder girdle (i.e.,
+cleithrum) \\
+Premaxilla & pmx.sc & Length from the anterior tip of the premaxilla to
+the distal tip of the ascending process of the premaxilla \\
+Dorsal Spine & Ds\#.sc\# = 1,2,or 3 & Length from the base of a dorsal
+spine above the pterygiophore to its distal tip along the anterior
+edge \\
+Pterygiophore & lpt.sc & Distance between the anterior to posterior tips
+of the pterygiophore immediately preceding the 3rd dorsal spine (when
+present) \\
+\bottomrule()
+\end{longtable}
 
@@
 \end{tabular}
-\label{tab1}
-\caption{The number of stickleback fossils at each time point.  Sample size at each time point ranges from a low of 22 to a high of 67.}
+\label{tabn}
+
 \end{table}
+
+\hypertarget{extant-specimen-data}{%
+\subsection{Extant Specimen data}\label{extant-specimen-data}}
+
+To span the gamut of stickleback diversity for our predictive model, we
+sampled modern stickleback from lakes containing generalist stickleback
+populations (Hendry et al. (\protect\hyperlink{ref-Hendry2009}{2009});
+Bolnick (\protect\hyperlink{ref-Bolnick2011}{2011})) and from lakes
+containing benthic-limnetic species pairs (Baumgartner et al.~1988;
+Schluter and McPhail (\protect\hyperlink{ref-Schluter1992}{1992}))
+(Table 2 (What is this?)). The generalist populations were collected by
+YES in 2013 and previously described in Stuart et al.
+(\protect\hyperlink{ref-Stuart2017}{2017}). These samples were fixed in
+formalin, then stained for bone with Alizarin Red in 2013. Benthic and
+limnetic specimens were kindly loaned by D. Schluter and his lab at
+University of British Columbia. They collected benthic and limnetic
+individuals from Enos Lake in 1988 and from Emily Lake, Little Quarry
+Lake, Paxton Lake, and Priest Lake in 2018. The Enos specimens had been
+fixed whole in formalin and stored in 40\% isopropanol. The specimens
+from the other lakes were initially preserved whole in 95\% ethanol in
+the field before being gradually transferred to water then formalin in
+the lab and ultimately stored in 40\% isopropanol. In 2019, we stained
+these specimens for bone using Alizarin Red.
+
+We next replicated fossil data collection (Table 1) on these extant
+specimens. Standard length as well as pelvic-spine length on each side
+were measured with calipers. We used a dissection microscope to count
+dorsal spines, pelvic spines, dorsal-fin rays, and anal-fin rays. Right
+and left-side pelvic girdle lengths and ectocoracoid lengths were
+measured from ventral photographs taken using a Canon EOS Rebel T7 with
+a Tamron 16-300 mm MACRO lens mounted on a leveled Kaiser RS1 copy
+stand. Specimens were held in place for ventral photographs using a
+small tabletop vise with an attached scale bar. Lateral X-rays were used
+to measure dorsal spine length, number of pterygiophores anterior to the
+pterygiophore holding the third spine, length of the pterygiophore just
+anterior to the third spine, cleithrum length, and pre-maxilla ascending
+branch length. We also counted vertebrae from the X-rays: abdominal
+vertebrae were counted anterior to the first vertebra with a haemal
+spine contacting an anal fin pterygiophore. Caudal vertebrae were
+posterior, including the first vertebra with the haemal spine contacting
+the anal fin pterygiophore (following Aguirre, Walker, and Gideon
+(\protect\hyperlink{ref-Aguirre2014}{2014}). X-rays were taken with an
+AXR Hot Shot X-ray Machine (Associated X-ray Corporation) at the Field
+Museum of Natural History. Specimens were exposed at 35kV and 4mA. Small
+fish were exposed for 7s, medium fish for 8s, and large fish for 10s. We
+developed the film and scanned individual images of each fish using the
+B\&W Negatives setting on an Epson Perfection 4990 Photo flatbed at 2400
+dpi. Measurements from photographs and X-rays were taken with FIJI
+(Schindelin et al.~2012) and its plugin ObjectJ
+(\url{https://sils.fnwi.uva.nl/bcb/objectj/}). All photographs, X-rays,
+and ObjectJ files have been uploaded to Morphosource.org (accession \#
+TBD). We dissected individuals from the generalist populations to
+determine sex from the gonads. Individuals from the species-pair lakes
+were sexed by Schluter and his lab, using a genotyping protocol (confirm
+and cite).
+
+The extant data used here consists of a total of 367 specimens all with
+known sex. Of these, there are 202 and 165 female and male specimens,
+respectively.
+
+\hypertarget{outlier-analysis.}{%
+\subsection{Outlier analysis.}\label{outlier-analysis.}}
+
+To check for outliers, we calculated within-group means and standard
+deviations for each trait separately for K series fossil specimens
+(pooled across samples) and for extant specimens (pooled across lakes).
+We noted trait values greater than 3.5 standard deviations from the mean
+as potential outliers. We checked whether these potential outliers were
+a result of data entry and collection error and corrected them if they
+were. We turned the remaining outlier trait values to NAs. (confirm)
+(Wait, what? Why? Anything that was 3.5 SD above the mean was recorded
+as missing? How do you justify this? )
+
+Missing data imputation, including fossil sex
+
+Quantification and evolution of sexual dimorphism
 
@@
 \(P(\boldsymbol{W}_{fossil}|\boldsymbol{W}_{extant}, \boldsymbol{X}, \boldsymbol{Y})\)
-using the multiple imputation with chained equations (MICE) algorithm
+using the multiple imputation by chained equations (MICE) algorithm
 (Buuren and Groothuis-Oudshoorn (\protect\hyperlink{ref-MICE}{2011}))
-with predictive mean matching. The imputation algorithm is run to obtain
-a total of \(M = 100\) completed datasets.
+with predictive mean matching. Traditionally, the choice for the number
+of completed dats sets is a relatively small number such as \(M = 5\) or
+\(M = 10\). However, Zhou and Reiter
+(\protect\hyperlink{ref-ZhouReiter2009}{2010}) recommend a larger number
+of imputed data sets if the data users intend on performing Bayesians
+analysis after imputation, which in this case, we do. Therefore, the
+imputation algorithm is run to obtain a total of \(M = 100\) completed
+datasets. In addition to this, Zhou and Reiter
+(\protect\hyperlink{ref-ZhouReiter2009}{2010}) suggests rather than
+using Rubin's combining rules to combine across imputed data sets,
+instead pool all of the draws from the posterior distributions across
+all of the imputed data sets to estimate the posterior distributions of
+parameters of interest. We proceed with our Bayesian analysis in this
+manner.
+
+\textcolor{red}{Akhil's stuff about validating the imputation model goes here.}
 
@@
 \subsection{Completed Data Model}\label{completed-data-model}}
+
+\textcolor{red}{We should note somewhere that we are only using the $W_{fossil}$ in the modeling part.
+We drop the $W_{extant}$.
+So just note that $W_{ti}$ is really $W_{fossil,ti}$.
+Not sure how to say this, but we need to make it clear that we are only using the fossil data for the OU modeling.
+}
+
+\hypertarget{continuous-phenotypes}{%
+\subsubsection{Continuous Phenotypes}\label{continuous-phenotypes}}
 
@@
 \(\boldsymbol{u}_{k,g} = \{u_{k,g1},\cdots,u_{k,gT}\}\) to an
-Ornstein-Uhlenback (OU) process
-((\protect\hyperlink{ref-OUprocess}{\textbf{OUprocess?}})) where we
-assume \(u_{k,gT}\) have a marginal mean of 0. Because each time period
-is \textasciitilde1000 years, we will discretize the OU process without
+Ornstein-Uhlenback (OU) process (Uhlenbeck and Ornstein
+(\protect\hyperlink{ref-OUProcess}{1930})) where we assume \(u_{k,gT}\)
+have a marginal mean of 0. Because each time period is
+\textasciitilde1000 years, we will discretize the OU process without
 loss of generality. For \(t = 2,\ldots,T\), we assume \begin{align}
@@
 i.e.~\(p(u_{k,gs}) = p(u_{k,gt})\) for \(s \neq t\).
+
+\hypertarget{discrete-phenotypes}{%
+\subsubsection{Discrete Phenotypes}\label{discrete-phenotypes}}
 
@@
 \hypertarget{sec:conclusions}{%
-\section{Future work and conclusions}\label{sec:conclusions}}
+\section{Discsusson, Future work and
+conclusions}\label{sec:conclusions}}
+
+We predicted that release from predators would result in niche expansion
+and increased sexual dimorphism, based on several studies of modern
+stickleback. For example, in lakes where sculpin competitors are absent
+and stickleback (Roesti et al.~2023) See Spoljaric and Reimchen 2008,
+page 512 right column for references and discussion of differences
+between benthic males and limnetic females. Male stickleback are benthic
+and littoral (Wootton 1976)\ldots. Reimchen papers in general good for
+this section.
 
@@
 
-Stickle!
+We thank O. Abughoush, S. Blaine, A. Chaudhary, M. Islam, F. Joaquin, C.
+Lawson-Weinert, R. Sullivan, J. Tien, M.P. Travis, and W. Shim for help
+with data collection. We thank D. Schluter and S. Blain for loaning
+specimens and sharing data. We thank K. Swagel and C. McMahan of the
+Field Museum for assistance with specimen x-rays. This research was
+supported by NSF grants BSR-8111013, EAR-9870337, and DEB-0322818, the
+Center for Field Research (Earthwatch), and the National Geographic
+Society (2869-84) to MAB. It was also supported by NSF grants
+DEB-1456462 and EAR-2145830 to YES. And NSF DMS-2015374 (GJM)
 
@@
 \begin{CSLReferences}{1}{0}
+\leavevmode\vadjust pre{\hypertarget{ref-Aguirre2014}{}}%
+Aguirre, Windsor E, Kendal Walker, and Shawn Gideon. 2014. {``Tinkering
+with the Axial Skeleton: Vertebral Number Variation in Ecologically
+Divergent Threespine Stickleback Populations.''} \emph{Biol. J. Linn.
+Soc. Lond.} 113 (1): 204--19.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Bell2009}{}}%
+Bell, M A. 2009. {``Implications of a Fossil Stickleback Assemblage for
+Darwinian Gradualism.''} \emph{J. Fish Biol.} 75 (8): 1977--99.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Bell2006}{}}%
+Bell, Michael A, Matthew P Travis, and D Max Blouw. 2006. {``Inferring
+Natural Selection in a Fossil Threespine Stickleback.''}
+\emph{Paleobiology} 32 (4): 562--77.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Blain2022}{}}%
+Blain, S A. 2022. \emph{Evolutionary Outcomes of Interactions Among
+Phenotypes in Post-Glacial Lakes}. University of British Columbia,
+Canada.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Bolnick2011}{}}%
+Bolnick, Daniel I. 2011. {``Sympatric Speciation in Threespine
+Stickleback: Why Not?''} \emph{Int. J. Ecol.} 2011: 1--15.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Bolnick2003}{}}%
+Bolnick, Daniel I, and Michael Doebeli. 2003. {``Sexual Dimorphism and
+Adaptive Speciation: Two Sides of the Same Ecological Coin.''}
+\emph{Evolution} 57 (11): 2433--49.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Bolnick2010}{}}%
+Bolnick, Daniel I, Travis Ingram, William E Stutz, Lisa K Snowberg, On
+Lee Lau, and Jeff S Paull. 2010. {``Ecological Release from
+Interspecific Competition Leads to Decoupled Changes in Population and
+Individual Niche Width.''} \emph{Proc. Biol. Sci.} 277 (1689): 1789--97.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Bolnick2008}{}}%
+Bolnick, Daniel I, and On Lee Lau. 2008. {``Predictable Patterns of
+Disruptive Selection in Stickleback in Postglacial Lakes.''} \emph{Am.
+Nat.} 172 (1): 1--11.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Bowne1994}{}}%
+Bowne, P S. 1994. {``Systematics and Morphology of the
+Gasterosteiformes.''} In \emph{The Evolutionary Biology of the
+Threespine Stickleback}, edited by M A Bell and S A Foster. Oxford, UK:
+Oxford University Press.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Butler2007}{}}%
+Butler, Marguerite A, Stanley A Sawyer, and Jonathan B Losos. 2007.
+{``Sexual Dimorphism and Adaptive Radiation in Anolis Lizards.''}
+\emph{Nature} 447 (7141): 202--5.
+
 \leavevmode\vadjust pre{\hypertarget{ref-MICE}{}}%
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-Cooper2011}{}}%
+Cooper, Idelle A, R Tucker Gilman, and Janette Wenrick Boughman. 2011.
+{``Sexual Dimorphism and Speciation on Two Ecological Coins: Patterns
+from Nature and Theoretical Predictions.''} \emph{Evolution} 65 (9):
+2553--71.
+
 \leavevmode\vadjust pre{\hypertarget{ref-Cornault2022}{}}%
@@
 \url{https://doi.org/10.1093/sysbio/syac036}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Hendry2009}{}}%
+Hendry, A P, D I Bolnick, D Berner, and C L Peichel. 2009. {``Along the
+Speciation Continuum in Sticklebacks.''} \emph{J. Fish Biol.} 75 (8):
+2000--2036.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Herrmann2021}{}}%
+Herrmann, Nicholas C, James T Stroud, and Jonathan B Losos. 2021. {``The
+Evolution of 'Ecological Release' into the 21st Century.''} \emph{Trends
+Ecol. Evol.} 36 (3): 206--15.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Hone2017}{}}%
+Hone, David W E, and Jordan C Mallon. 2017. {``Protracted Growth Impedes
+the Detection of Sexual Dimorphism in Non-Avian Dinosaurs.''}
+\emph{Palaeontology} 60 (4): 535--45.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Hunt2008}{}}%
+Hunt, Gene, Michael A Bell, and Matthew P Travis. 2008. {``Evolution
+Toward a New Adaptive Optimum: Phenotypic Evolution in a Fossil
+Stickleback Lineage.''} \emph{Evolution} 62 (3): 700--710.
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-Schluter1992}{}}%
+Schluter, D, and J D McPhail. 1992. {``Ecological Character Displacement
+and Speciation in Sticklebacks.''} \emph{Am. Nat.} 140 (1): 85--108.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Shine1989}{}}%
+Shine, R. 1989. {``Ecological Causes for the Evolution of Sexual
+Dimorphism: A Review of the Evidence.''} \emph{Q. Rev. Biol.} 64 (4):
+419--61.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Stuart2021}{}}%
+Stuart, Yoel E, J William Sherwin, Ambika Kamath, and Thor Veen. 2021.
+{``Male and Female Anolis Carolinensis Maintain Their Dimorphism Despite
+the Presence of Novel Interspecific Competition.''} \emph{Evolution} 75
+(11): 2708--16.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Stuart2020}{}}%
+Stuart, Yoel E, Matthew P Travis, and Michael A Bell. 2020. {``Inferred
+Genetic Architecture Underlying Evolution in a Fossil Stickleback
+Lineage.''} \emph{Nat. Ecol. Evol.} 4 (11): 1549--57.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Stuart2017}{}}%
+Stuart, Yoel E, Thor Veen, Jesse N Weber, Dieta Hanson, Mark Ravinet,
+Brian K Lohman, Cole J Thompson, et al. 2017. {``Contrasting Effects of
+Environment and Genetics Generate a Continuum of Parallel Evolution.''}
+\emph{Nat. Ecol. Evol.} 1 (6): 158.
+
+\leavevmode\vadjust pre{\hypertarget{ref-OUProcess}{}}%
+Uhlenbeck, G. E., and L. S. Ornstein. 1930. {``On the Theory of the
+Brownian Motion.''} \emph{Phys. Rev.} 36 (September): 823--41.
+\url{https://doi.org/10.1103/PhysRev.36.823}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Voje2022}{}}%
+Voje, Kjetil L, Michael A Bell, and Yoel E Stuart. 2022. {``Evolution of
+Static Allometry and Constraint on Evolutionary Allometry in a Fossil
+Stickleback.''} \emph{J. Evol. Biol.} 35 (3): 423--38.
+
 \leavevmode\vadjust pre{\hypertarget{ref-ZhouReiter2009}{}}%

### 2023-06-01 R/stuart_stickle_bayes.R
Update Model to add correlation between mav and stl as well as mcv and stl.

@@ -35,7 +35,7 @@ parameters{
   real<lower=-1,upper=1> kappa[16];
   real<lower=0> sigma[10];
   real<lower=0> tau[16];
-  real beta[9];
+  real beta[11];
   vector[2*T] u_stl;
   vector[2*T] u_lps;
   vector[2*T] u_ect;
@@ -116,9 +116,9 @@ model{
   ds3 ~ normal(X * mu_ds3 + beta[8] * (stl - X * mu_stl),sigma[9]);
   lpt ~ normal(X * mu_lpt + beta[9] * (stl - X * mu_stl),sigma[10]);
   mdf ~ poisson(exp(X * mu_mdf));
-  mav ~ poisson(exp(X * mu_mav));
+  mav ~ poisson(exp(X * mu_mav + beta[10] * (stl - X * mu_stl) - pow(beta[10]*sigma[1],2)/2));
   maf ~ poisson(exp(X * mu_maf));
-  mcv ~ poisson(exp(X * mu_mcv));
+  mcv ~ poisson(exp(X * mu_mcv + beta[11] * (stl - X * mu_stl) - pow(beta[11]*sigma[1],2)/2));
   mds ~ poisson(exp(X * mu_mds));
   mpt ~ poisson(exp(X * mu_mpt));
 

### 2023-06-01 R/stuart_stickle_bayes_plots.R
Update plot file

@@ -117,6 +117,31 @@ for (i in 1:16){
       ylab(expression(paste("p(",beta[g],"|",bold(X),")")))
     ggsave(paste0("Figures/",phenotype,"/beta.pdf"),width=5,height=8)
   }
+  if (phenotype == "mav"){
+    samps %>%
+      rename(beta = paste0("beta[",10,"]"))%>%
+      select(beta) %>%
+      ggplot() +
+      geom_density(aes(x=beta)) +
+      geom_vline(aes(xintercept = 0),colour = "red",linetype=2) +
+      theme_bw() +
+      xlab(expression(beta[g])) +
+      ylab(expression(paste("p(",beta[g],"|",bold(X),")")))
+    ggsave(paste0("Figures/",phenotype,"/beta.pdf"),width=5,height=8)
+  }
+
+  if (phenotype == "mcv"){
+    samps %>%
+      rename(beta = paste0("beta[",11,"]"))%>%
+      select(beta) %>%
+      ggplot() +
+      geom_density(aes(x=beta)) +
+      geom_vline(aes(xintercept = 0),colour = "red",linetype=2) +
+      theme_bw() +
+      xlab(expression(beta[g])) +
+      ylab(expression(paste("p(",beta[g],"|",bold(X),")")))
+    ggsave(paste0("Figures/",phenotype,"/beta.pdf"),width=5,height=8)
+  }
 }
 
 

### 2023-06-01 R/stuart_stickle_bayes.R
Change name of save file for new posterior samples.

@@ -227,4 +227,5 @@ for (m in 1:M){
   }
 }
 
-saveRDS(samps,"posterior_samples.RDS")
+# saveRDS(samps,"posterior_samples.RDS")
+saveRDS(samps,"posterior_samples_mav_adj.RDS")

### 2023-06-02 manuscript/paper.Rmd
Fix notation mistake in priors.
Updated affiliations

@@
 author: | 
-  | \large Allison Ozark$^1$, Akhil Ghosh$^1$, Raheyma Siddiqui$^1$, Matthew Stuart$^1$,
-  | \large  Samantha Swank$^1$, Michael A. Bell$^2$, Gregory J. Matthews$^1$, and Yoel E. Stuart$^1$ \vspace{-1.1mm}
+  | \large Allison Ozark$^1$, Akhil Ghosh$^2$, Raheyma Siddiqui$^1$, Matthew Stuart$^{2,3}$,
+  | \large  Samantha Swank$^1$, Michael A. Bell$^4$, Gregory J. Matthews$^{2,3}$, and Yoel E. Stuart$^1$ \vspace{-1.1mm}
   |
-  | \large $^1$ Loyola University Chicago, Chicago, IL, USA \vspace{-1.1mm}
+  | \large $^1$ Department of Biology, Loyola University Chicago, Chicago, IL, USA \vspace{-1.1mm}
   |
-  | \large $^2$ University of California Museum of Paleontology, Berkeley, CA, USA \vspace{-1.1mm}
+  | \large $^2$ Department of Mathematics and Statistics, Loyola University Chicago, Chicago, IL, USA \vspace{-1.1mm}
+  |  
+  | \large $^3$ Center for Data Science and Consulting, Loyola University Chicago, Chicago, IL, USA \vspace{-1.1mm}
+  |  
+  | \large $^4$ University of California Museum of Paleontology, Berkeley, CA, USA \vspace{-1.1mm}
   |
@@
 \kappa_k & \iid \mathcal{N}(0,1)I_{\{-1 < \kappa_g < 1\}} \nonumber \\
-\beta_k & \iid \mathcal{N}(0,5) \\
-\theta_{k,g} & \iid \mathcal{N}(0,10000)
+\beta_k & \iid \mathcal{N}(0,5) \nonumber \\
+\theta_{k,g} & \iid \mathcal{N}(0,10000) \nonumber \\
 \omega_l & \iid \mathcal{N}(0,100)I_{\{\tau > 0\}} \nonumber \\

### 2023-06-02 manuscript/paper.Rmd
Added plot of fossil sampel size per time period.

@@
 Along with the fossil data collected for this study, we also have data from extant populations of the Threespine Stickleback, Gasterosteus aculeatus, measuring the same traits for individuals whose sex could be determined directly.
+
 We then combined these sets of data and treat the sex of the fossil specimens as missing and  used multiple imputation (@little2002statistical) to impute the sex of the fossils many times.
@@
 ```{r}
+library(tidyverse)
 stickle <- read_csv('stickle_revamped.csv')
@@
 #fossil n
-fortab <- data.frame(time = 1:18, count = as.character(as.numeric((table(times_vector)))))
+#fortab <- data.frame(time = 1:18, count = as.character(as.numeric((table(times_vector)))))
+fortab <- data.frame(time = 1:18, count = as.numeric((table(times_vector))))
 #knitr::kable(fortab)
@@
 The fossil data consists of `r sum(is.na(stick$gender))` specimens with unknown sex over 18 time periods spaced  ~1000 years apart.
-Table 2 shows the sample size at each of the 18 time periods.
+Figure 1 shows the sample size at each of the 18 time periods.
 There are at least 22 specimens at each time period with a high of 67 specimens in period 7.
+
+```{r fig.cap = "Fossil sampel size for each time period"}
+ggplot(aes(x = time, y = count), data = fortab) + geom_col() + xlab("Time Period")  + scale_x_continuous(breaks = 1:18, labels = 1:18)
+```
 
@@
 
+<!-- I removed the table and replaced it with a figure -->
 <!-- % latex table generated in R 4.2.2 by xtable 1.8-4 package -->
 <!-- % Mon May  1 11:20:28 2023 -->
-\begin{table}[ht]
-\centering
-\caption{The number of stickleback fossils at each time point.  Sample size at each time point ranges from a low of 22 to a high of 67.}
-\begin{tabular}{rl}
-  \hline
-time & count \\ 
-  \hline
-  1 & 43 \\ 
-    2 & 41 \\ 
-    3 & 51 \\ 
-    4 & 41 \\ 
-    5 & 46 \\ 
-    6 & 48 \\ 
-    7 & 67 \\ 
-    8 & 55 \\ 
-    9 & 42 \\ 
-   10 & 33 \\ 
-   11 & 37 \\ 
-   12 & 22 \\ 
-   13 & 41 \\ 
-   14 & 43 \\ 
-   15 & 46 \\ 
-   16 & 47 \\ 
-   17 & 56 \\ 
-   18 & 55 \\ 
-   \hline
-\end{tabular}
-\label{tabn}
+<!-- \begin{table}[ht] -->
+<!-- \centering -->
+<!-- \caption{The number of stickleback fossils at each time point.  Sample size at each time point ranges from a low of 22 to a high of 67.} -->
+<!-- \begin{tabular}{rl} -->
+<!--   \hline -->
+<!-- time & count \\  -->
+<!--   \hline -->
+<!--   1 & 43 \\  -->
+<!--     2 & 41 \\  -->
+<!--     3 & 51 \\  -->
+<!--     4 & 41 \\  -->
+<!--     5 & 46 \\  -->
+<!--     6 & 48 \\  -->
+<!--     7 & 67 \\  -->
+<!--     8 & 55 \\  -->
+<!--     9 & 42 \\  -->
+<!--    10 & 33 \\  -->
+<!--    11 & 37 \\  -->
+<!--    12 & 22 \\  -->
+<!--    13 & 41 \\  -->
+<!--    14 & 43 \\  -->
+<!--    15 & 46 \\  -->
+<!--    16 & 47 \\  -->
+<!--    17 & 56 \\  -->
+<!--    18 & 55 \\  -->
+<!--    \hline -->
+<!-- \end{tabular} -->
+<!-- \label{tabn} -->
 
-\end{table}
+<!-- \end{table} -->
 

### 2023-06-07 manuscript/paper.Rmd
Changing author order and testing my commit and push

@@
 author: | 
-  | \large Allison Ozark$^1$, Akhil Ghosh$^2$, Raheyma Siddiqui$^1$, Matthew Stuart$^{2,3}$,
-  | \large  Samantha Swank$^1$, Michael A. Bell$^4$, Gregory J. Matthews$^{2,3}$, and Yoel E. Stuart$^1$ \vspace{-1.1mm}
+  | \large Allison Ozark$^{1,*}$, Matthew Stuart$^{2,3,*}$, Raheyma Siddiqui$^1$, Akhil Ghosh$^2$, 
+  | \large  Samantha Swank$^1$, Michael A. Bell$^4$, Gregory J. Matthews$^{2,3}$, and Yoel E. Stuart${^1,+$} \vspace{-1.1mm}
   |
@@
   | \large $^4$ University of California Museum of Paleontology, Berkeley, CA, USA \vspace{-1.1mm}
+  |
+  | \large $^*$ Equal contribution \vspace{-1.1mm}
+  |
+  | \large $^+$ Corresponding: ystuart@luc.edu \vspace{-1.1mm}
   |

### 2023-06-07 manuscript/paper.Rmd
trying again?
updated bibliography and some intro methods text

@@
 
-Ecological release theory suggests that a prey population’s niche should expand when predation is relaxed or removed (reviewed in @Herrmann2021).
-Niche changes in habitat and resource use should then drive phenotypic evolution as the population adapts to new conditions (@Herrmann2021).
-For example, resource use expansion should result in trophic trait evolution (cite; e.g.,).
-Habitat expansion should result in locomotor change (cite; e.g.,).
-And release from enemies should result in reduction of costly defense traits (cite; e.g.,).
+Ecological release theory suggests that a prey population’s niche should shift and expand when predation is relaxed or removed (reviewed in @Herrmann2021). 
+Niche expansion could manifest through a combination of increasing within- and between-individual niche widths (@Bolnick2010; @Herrmann2021), including divergence between the sexes (@Bolnick2003; @Cooper2011).
+Theory and empirical data suggest that a population presented with novel ecological opportunity might experience disruptive selection on males and females stemming from intraspecific competition over newly accessible resources, resulting in intersexual divergence in habitat use (Schoener 1968; @Shine1989; @Bolnick2003; @Butler2007; @Bolnick2008; @Cooper2011; but see @Stuart2021; @Blain2022).
+Accordingly, sexes should also diverge in morphological phenotypes associated with habitat use; such ecologically-driven "character displacement between the sexes" is therefore one explanation for sexual dimorphism (de Lisle papers).
 
-Niche expansion could manifest through a combination of increasing within- and between-individual niche widths (@Bolnick2010; @Herrmann2021), including divergence between the sexes (@Bolnick2003; @Cooper2011).
-Theory and empirical data suggest that a population presented with novel ecological opportunity might experience disruptive selection on males and females stemming from intraspecific competition over newly accessible resources, resulting in intersexual divergence in habitat use and associated phenotypes (Schoener 1968; @Shine1989; @Bolnick2003; @Butler2007; @Bolnick2008; @Cooper2011; but see @Stuart2021; @Blain2022).
-Such character displacement between the sexes is therefore one explanation for sexual dimorphism (de Lisle papers).
+We tested this link between release from predators and the evolution of sexual dimorphism in a well-preserved, high resolution, 16,000 year-long sequence of the fossil stickleback fish (<i>Gasterosteus doryssus</i>).
+The fossil lineage appeared in the depositional environment as a fully armored form, with complete pelvic girdles, two pelvic spines, and three dorsal spines, on average (@Bell2006; @Stuart2020).
+However, this habitat appears to have been missing fish piscivores (e.g., trout and other salmonids known to prey on modern stickleback; @Bell2009), relaxing putative selection for armor (Reimchen papers, Bell papers, @Bowne1994; @Roestietal2023).
+The lineage adaptively evolved armor loss but also reduction in the means of several other traits as well (@Hunt2008; @Stuart2020; Siddiqui et al.
+in prep).
+In particular, tooth wear data suggest that the population also began eating more planktonic prey, indicative of increased open water habitat use (Purnell et al.
+2007).
+If this divergence in habitat use is partitioned between sexes (@ReimchenNosil2004; @SpoljaricReimchen2008;@Reimchen2016), then we should expect to observe sexual dimorphism evolution for traits related to habitat use in <i>G.
+doryssus</i>.
 
-We tested this link between release from predators and the evolution of sexual dimorphism in a well-preserved, high resolution lineage of the fossil stickleback fish (Gasterosteus doryssus).
-The sex of the individual fossils, which is unobserved in practice, was imputed using multiple imputation by chained equations (MICE) (@MICE) with observed data used to train the imputation model based on extant stickleback with known sex.
-Then we tracked the multivariate evolution of sexual dimorphism over ~16,000 years for traits related to swimming, feeding and defense.
-The fossil sequence captures a bout of adaptive evolution following ecological release from predators, as follows.
-The fossil lineage appeared in the depositional environment as a fully armored form, with complete pelvic girdles, two pelvic spines, and three dorsal spines, on average (@Bell2006; @Stuart2020).
-However, this habitat appears to have been missing fish piscivores (e.g., trout and other salmonids known to prey on modern stickleback; @Bell2009), relaxing putative selection for armor (Reimchen papers, Bell papers, @Bowne1994; Roesti et al.
-2023).
-The lineage thus adaptively evolved armor loss and reduction in several other traits as well (@Hunt2008; @Stuart2020; Siddiqui et al.
-in prep), and tooth wear data suggest that the population also began eating more planktonic prey, expanding to an open water lifestyle (Purnell et al.
-2007).
-
-Sexual Dimorphism is interesting because......
-
-Sexual Dimoprhism lit review:
-@SaittaEtAl2020
-
-The remainder of this manuscript contains a description of the data in section \ref{sec:data} and a description of our models in \ref{sec:models}.
-Section \ref{sec:results} presents a summary of our results, and we end with our conclusion and future work in section \ref{sec:conclusions}.
+We imputed the sex of individual fossils, which is unobserved in practice, with Multiple Imputation by Chained Equations (MICE) (@MICE), using observed data from extant stickleback of known sex to train the imputation model.
+Then we tracked the multivariate evolution of sexual dimorphism over ~16,000 years for traits related to swimming, feeding and defense to test whether increased sexual dimorphism is associated with ecological release from predators.
 
@@
 Instead, paleobiologists resort to statistical detection of sex and sexual dimorphism, including tests for bimodality in trait distributions (e.g., @Hone2017) and divergence in growth curves (e.g., @SaittaEtAl2020).
-
 We used a third approach to infer fossil sex and study the evolution of sexual dimorphism by taking advantage of the fact that G.
-doryssus is part of the extant threespine stickleback species complex (Gastersteus aculeatus).
+doryssus is part of the extant threespine stickleback species complex (Gasterosteus aculeatus).
 Along with the fossil data collected for this study, we also have data from extant populations of the Threespine Stickleback, Gasterosteus aculeatus, measuring the same traits for individuals whose sex could be determined directly.

### 2023-06-08 manuscript/paper.Rmd
adding refs

@@
 Niche expansion could manifest through a combination of increasing within- and between-individual niche widths (@Bolnick2010; @Herrmann2021), including divergence between the sexes (@Bolnick2003; @Cooper2011).
-Theory and empirical data suggest that a population presented with novel ecological opportunity might experience disruptive selection on males and females stemming from intraspecific competition over newly accessible resources, resulting in intersexual divergence in habitat use (Schoener 1968; @Shine1989; @Bolnick2003; @Butler2007; @Bolnick2008; @Cooper2011; but see @Stuart2021; @Blain2022).
-Accordingly, sexes should also diverge in morphological phenotypes associated with habitat use; such ecologically-driven "character displacement between the sexes" is therefore one explanation for sexual dimorphism (de Lisle papers).
+Theory and empirical data suggest that a population presented with ecological opportunity following ecological release might experience disruptive selection on males and females stemming from intraspecific competition over newly accessible resources, resulting in intersexual divergence in habitat use (@Schoener1968; @Shine1989; @Bolnick2003; @Butler2007; @Bolnick2008; @Cooper2011; but see @Stuart2021; @Blain2022).
+Accordingly, sexes should also diverge in morphological traits associated with habitat use; such ecologically-driven "character displacement between the sexes" is therefore one explanation for sexual dimorphism (@deLisleRowe2015; @deLisleNco2018).
 

### 2023-06-08 manuscript/paper.Rmd
More text and references added.

@@
 
-Ecological release theory suggests that a prey population’s niche should shift and expand when predation is relaxed or removed (reviewed in @Herrmann2021). 
+Ecological release theory suggests that a prey population’s niche should expand when predation is relaxed or removed, as prey take advantage of new enemy free space (reviewed in @Herrmann2021).
 Niche expansion could manifest through a combination of increasing within- and between-individual niche widths (@Bolnick2010; @Herrmann2021), including divergence between the sexes (@Bolnick2003; @Cooper2011).
 Theory and empirical data suggest that a population presented with ecological opportunity following ecological release might experience disruptive selection on males and females stemming from intraspecific competition over newly accessible resources, resulting in intersexual divergence in habitat use (@Schoener1968; @Shine1989; @Bolnick2003; @Butler2007; @Bolnick2008; @Cooper2011; but see @Stuart2021; @Blain2022).
-Accordingly, sexes should also diverge in morphological traits associated with habitat use; such ecologically-driven "character displacement between the sexes" is therefore one explanation for sexual dimorphism (@deLisleRowe2015; @deLisleNco2018).
+Accordingly, sexes should also diverge in morphological traits associated with habitat use; such ecologically-driven "character displacement between the sexes" is therefore one explanation for phenotypic sexual dimorphism (@deLisleRowe2015; @deLisleNco2018).
 
-We tested this link between release from predators and the evolution of sexual dimorphism in a well-preserved, high resolution, 16,000 year-long sequence of the fossil stickleback fish (<i>Gasterosteus doryssus</i>).
+We tested this link between release from predators and the evolution of sexual dimorphism in a well-preserved, finely-resolved, ~16,000 year-long sequence of the fossil stickleback fish (<i>Gasterosteus doryssus</i>).
 The fossil lineage appeared in the depositional environment as a fully armored form, with complete pelvic girdles, two pelvic spines, and three dorsal spines, on average (@Bell2006; @Stuart2020).
-However, this habitat appears to have been missing fish piscivores (e.g., trout and other salmonids known to prey on modern stickleback; @Bell2009), relaxing putative selection for armor (Reimchen papers, Bell papers, @Bowne1994; @Roestietal2023).
-The lineage adaptively evolved armor loss but also reduction in the means of several other traits as well (@Hunt2008; @Stuart2020; Siddiqui et al.
+However, this habitat appears to have been missing fish piscivores (e.g., trout and other salmonids known to prey on modern stickleback; @Bell2009), relaxing putative selection for armor (@Reichen1980, @Reimchen1994, @BellEtAl1993; @Bowne1994; @Roestietal2023).
+The lineage adaptively evolved armor loss but also reduction in the means of several other traits  (@Hunt2008; @Stuart2020; Siddiqui et al.
 in prep).
-In particular, tooth wear data suggest that the population also began eating more planktonic prey, indicative of increased open water habitat use (Purnell et al.
+In particular, tooth wear data suggest that the population shifted to eat more planktonic prey through time, indicative of increased open water habitat use (Purnell et al.
 2007).
-If this divergence in habitat use is partitioned between sexes (@ReimchenNosil2004; @SpoljaricReimchen2008;@Reimchen2016), then we should expect to observe sexual dimorphism evolution for traits related to habitat use in <i>G.
+If this divergence in habitat use is partitioned between sexes (@ReimchenNosil2004; @SpoljaricReimchen2008;@Reimchen2016), then we should expect sexual dimorphism to evolve for traits related to habitat use in <i>G.
 doryssus</i>.
 
-We imputed the sex of individual fossils, which is unobserved in practice, with Multiple Imputation by Chained Equations (MICE) (@MICE), using observed data from extant stickleback of known sex to train the imputation model.
-Then we tracked the multivariate evolution of sexual dimorphism over ~16,000 years for traits related to swimming, feeding and defense to test whether increased sexual dimorphism is associated with ecological release from predators.
+To test this prediction, we imputed the sex of individual fossils, which is unobserved in practice, with Multiple Imputation by Chained Equations (MICE) (@MICE), using observed data from extant stickleback of known sex to train the imputation model.
+Then we tracked the multivariate evolution of sexual dimorphism over ~16,000 years for 16 traits related to swimming, feeding and defense.
 
@@
 
-A major challenge in this work is inferring the sex of fossil specimens.
-This typically cannot be done directly, except for lineages whose sexes are distinguished by the presence or absence of sex-specific characters (@Hone2017; @SaittaEtAl2020).
+A major challenge in this work is inferring the sex of fossil stickleback specimens.
+Indeed, for most species, sex cannot be detected directly, except for lineages whose sexes are distinguished by the presence or absence of sex-specific characters (@Hone2017; @SaittaEtAl2020).
 Instead, paleobiologists resort to statistical detection of sex and sexual dimorphism, including tests for bimodality in trait distributions (e.g., @Hone2017) and divergence in growth curves (e.g., @SaittaEtAl2020).
 We used a third approach to infer fossil sex and study the evolution of sexual dimorphism by taking advantage of the fact that G.
-doryssus is part of the extant threespine stickleback species complex (Gasterosteus aculeatus).
-Along with the fossil data collected for this study, we also have data from extant populations of the Threespine Stickleback, Gasterosteus aculeatus, measuring the same traits for individuals whose sex could be determined directly.
-
-We then combined these sets of data and treat the sex of the fossil specimens as missing and  used multiple imputation (@little2002statistical) to impute the sex of the fossils many times.
-Imputations were performed using the MICE (@MICE) algorithm and implemented in R (@R2022language) with M = 100 imputations performed.
-
-After imputing the sex of the fossil data, each completed data set is then used to fit a modified Ornstein–Uhlenbeck (OU) (@OUProcess) model using a Bayesian framework to look for sexual dimorphism across a variety of stickleback phenotypes
+doryssus is part of the extant threespine stickleback species complex, Gasterosteus aculeatus.
+We measured the same trait set from a phenotypically diverse set of extant G.
+aculeatus populations, for whom individual sex was known.
+We then combined fossil and extant data sets, treated fossil sex as missing data, and used the MICE (@MICE) multiple imputation (@little2002statistical) algorithm, and implemented in R (@R2022language), to impute individual sex of the fossil data set 100 times.
+We then fit a modified Ornstein–Uhlenbeck (OU) (@OUProcess) model using a Bayesian framework to the data to test for the evolution of sexual dimorphism in each trait.
 

### 2023-06-08 R/stuart_stickle_bayes.R
Update Stan Code to be more efficient.

@@ -4,174 +4,84 @@ library(rstan)
 library(ggridges)
 
 dat <- read.csv("Data/stuart_stickle_100imputations_all.csv")
-dat$mds[dat$mds==1.2] <- 1
 dat$group <- factor(paste(dat$gender,letters[dat$time]))
 
-mod <- "
+cont_ind <- "
 data{
   int<lower=0> N;
   int<lower=0> T;
   matrix[N,2*T] X;
-  vector[N] stl;
-  vector[N] lps;
-  vector[N] ect;
-  vector[N] tpg;
-  vector[N] cle;
-  vector[N] pmx;
-  vector[N] ds1;
-  vector[N] ds2;
-  vector[N] ds3;
-  vector[N] lpt;
-  int<lower=0> mdf[N];
-  int<lower=0> mav[N];
-  int<lower=0> maf[N];
-  int<lower=0> mcv[N];
-  int<lower=0> mds[N];
-  int<lower=0> mpt[N];
+  vector[N] y;
 }
 parameters{
-  real<lower=0> theta_f[16];
-  real<lower=0> theta_m[16];
-  real<lower=-1,upper=1> kappa[16];
-  real<lower=0> sigma[10];
-  real<lower=0> tau[16];
-  real beta[11];
-  vector[2*T] u_stl;
-  vector[2*T] u_lps;
-  vector[2*T] u_ect;
-  vector[2*T] u_tpg;
-  vector[2*T] u_cle;
-  vector[2*T] u_pmx;
-  vector[2*T] u_ds1;
-  vector[2*T] u_ds2;
-  vector[2*T] u_ds3;
-  vector[2*T] u_lpt;
-  vector[2*T] u_mdf;
-  vector[2*T] u_mav;
-  vector[2*T] u_maf;
-  vector[2*T] u_mcv;
-  vector[2*T] u_mds;
-  vector[2*T] u_mpt;
+  real<lower=0> theta_f;
+  real<lower=0> theta_m;
+  real<lower=-1,upper=1> kappa;
+  real<lower=0> sigma;
+  real<lower=0> tau;
+  vector[2*T] u;
 }
 transformed parameters{
-  vector[2*T] mu_stl;
[…]

### 2023-06-22 RawData_ReadOnly/FvE.AO.Final_forMICE_Modeling.R
Data for analysis, as of June 22, 2023.
Fixed size correction function.

Fixed NA vs. 0 discrepancies.

Generated an R script that goes from raw data to a cleaned data set ready for MICE and modeling.

@@ -1,0 +1,225 @@
+# === Description ==================================================================
+# This script (and accompanying data) should be sufficient to recreate the analyses in Siddiqui et al. (under review, Proceedings B)
+# Built 13 June 2023
+
+# === 0) Preparation, Housekeeping, etc. ======
+## empty environment
+rm(list = ls())
+
+## Set working directory. Hint, open this R script in the same folder as the data.
+working.dir <- getwd()
+setwd(working.dir)
+
+## libraries
+library(lme4)
+
+
+##Global Variables
+#variables
+clrs <- c("#000000", "#FF0000") # c("black", "red")
+options(max.print=1000000)
+#Lake abbreviations in the study
+LS_lake_IDs <- c("Com", "Fre", "Joe", "Nor", "Swa", "The", "Vib")
+SP_lake_IDs <- c("Eno", "LQx", "Pax", "Pri", "Emi")
+SP_lake_names <- c("Enos", "Little Quarry", "Paxton", "Priest", "Emily")
+time.since.colonization <- read.csv("a_raw.data.readonly/KSampleMeanTimes.csv", header = TRUE, stringsAsFactors = FALSE)[, c(1,3)]
+#function for later
+f.sample.size.noNAs <- function(x) length(x) - sum(is.na(x))
+#end 0)
+# **********
+
+# === 1) Data clean up and and Size Correction ======
+## Read in data
+#Using the scripts 2a.raw.data.manip.FvE.R and 2b.outliers.size.correction.FvE.R, and the data called in those scripts, we previously checked data for outliers greater than 3sd from the mean. Outliers were checked. If they were part of a set of outliers contained in a single indvidual, they were left alone. If they were obvious data collection or entry errors, they were fixed. Any outliers that remained were set to NA.
+data.for.sc <- read.csv("b_cleaned.data/AO.sexualdimorphismPruned.for.sizecorrection.csv", header = TRUE, stringsAsFactors = FALSE)
+
+#checking lake names
+unique(substr(data.for.sc$univID, 1, 3))
+
[…]

### 2023-06-22 R/stuart_stickle_bayes.R
Update STAN model

@@ -6,47 +6,6 @@ library(ggridges)
 dat <- read.csv("Data/stuart_stickle_100imputations_all.csv")
 dat$group <- factor(paste(dat$gender,letters[dat$time]))
 
-cont_ind <- "
-data{
-  int<lower=0> N;
-  int<lower=0> T;
-  matrix[N,2*T] X;
-  vector[N] y;
-}
-parameters{
-  real<lower=0> theta_f;
-  real<lower=0> theta_m;
-  real<lower=-1,upper=1> kappa;
-  real<lower=0> sigma;
-  real<lower=0> tau;
-  vector[2*T] u;
-}
-transformed parameters{
-  vector[2*T] mu;
-  mu[1:T] = theta_f + u[1:T];
-  mu[(T+1):(2*T)] = theta_m + u[(T+1):(2*T)];
-}
-
-model{
-  // Data Model
-  y ~ normal(X * mu,sigma);
-
-  // OU Process for Gender and Time Specific Means
-  u[2:T] ~ normal(kappa*u[1:(T-1)],tau);
-  u[(T+2):(2*T)] ~ normal(kappa*u[(T+1):(2*T-1)],tau);
-
-  // Initial Gender and Time Mean Model
-  u[{1,T+1}] ~ normal(0,tau/sqrt(1-kappa*kappa));
-
-  // Priors
-  theta_f ~ normal(0,100);
-  theta_m ~ normal(0,100);
-  kappa ~ normal(0,1);
-  sigma ~ normal(0,10);
-  tau ~ normal(0,10);
-}
-"
-
 cont_dep <- "
 data{
   int<lower=0> N;
@@ -58,28 +17,41 @@ data{
 parameters{
   real<lower=0> theta_f;
   real<lower=0> theta_m;
+  real<lower=0> theta_f_stl;
+  real<lower=0> theta_m_stl;
   real<lower=-1,upper=1> kappa;
+  real<lower=-1,upper=1> kappa_stl;
   real<lower=0> sigma;
+  real<lower=0> sigma_stl;
   real<lower=0> tau;
+  real<lower=0> tau_stl;
   real beta;
   vector[2*T] u;
+  vector[2*T] u_stl;
 }
 transformed parameters{
   vector[2*T] mu;
   mu[1:T] = theta_f + u[1:T];
[…]

### 2023-06-23 R/StickleImputationCode.Rmd
commiting R code for stickle imputation process and committing the updated imputations

---
title: "StickleImputationCode"
author: "Akhil Ghosh"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(mice)
```

```{r}
setwd('/Users/akhilghosh/Desktop/RStudioProjects2022Fall/SticklebackProject/RawData_ReadOnly')
stickle <- read_csv("AO.sexualdimorphismPruned.sc.forAnalysis_forMICE_Modeling_20June2023.csv")

summary(stickle)
stick <- stickle %>% 
  mutate(gender = as.factor(case_when(
    grepl(".M.",univID,fixed=TRUE) ~ "Male",
    grepl(".F.",univID,fixed=TRUE) ~ "Female",
    grepl("Fos",univID,fixed=TRUE) ~ NA_character_
    ))
  ) %>% 
  select(gender, everything())

```

```{r}
set.seed(0314)
imputations = 100
stick_whole <- stick %>% select(-c(univID))
summary(stick_whole)
imputed_stick_whole <- mice(stick_whole,m=imputations)
complete_stick_list <- list()
for(m in 1:imputations){
  complete_stick_list[[m]] <- complete(imputed_stick_whole, m)
}
```

```{r}
univ <- stick %>% 
  mutate(time = as.integer(sub(".*([0-9]{2})\\..*", "\\1",univID)))
times <- univ %>% drop_na(time) %>% pull(time)

times_vector <- c(rep(1,43)) %>% 
  append(21-times)
```

```{r}
csv_list <-list()
for(i in 1:imputations){
csv_list[[i]] <- complete_stick_list[[i]][-c(1:367),] %>% mutate(imp = rep(i, 814), time = times_vector)
}

combined_csv<- do.call(rbind,csv_list) %>% select(imp, time,gender, everything())
```

### 2023-06-28 R/stuart_stickle_bayes.R
One more slight change to the Stan code.

@@ -3,7 +3,7 @@ library(tidyverse)
 library(rstan)
 library(ggridges)
 
-dat <- read.csv("Data/stuart_stickle_100imputations_all.csv")
+dat <- read.csv("Data/updated_stickle_imputations_100.csv")
 dat$group <- factor(paste(dat$gender,letters[dat$time]))
 
 cont_dep <- "
@@ -73,7 +73,7 @@ data{
   int<lower=0> N;
   int<lower=0> T;
   matrix[N,2*T] X;
-  int<lower=0>[N] y;
+  int<lower=0> y[N];
 }
 parameters{
   real<lower=0> theta_f;
@@ -112,7 +112,7 @@ data{
   int<lower=0> N;
   int<lower=0> T;
   matrix[N,2*T] X;
-  int<lower=0>[N] y;
+  int<lower=0> y[N];
   vector[N] stl;
 }
 parameters{
@@ -170,54 +170,69 @@ stan_disc_ind <- stan_model(model_code = disc_ind)
 stan_disc_dep <- stan_model(model_code = disc_dep)
 M <- max(dat$imp)
 
-vars <- c("stl","lps","ect","tpg",
+vars <- c("lps","ect","tpg",
           "cle","pmx","ds1","ds2",
           "ds3","lpt","mdf","mav",
           "maf","mcv","mds","mpt")
 
 for (m in 1:M){
   tmp <- dat %>% filter(imp == m)
-  stl <- tmp$stl
-  mdf <- tmp$mdf
-  maf <- tmp$maf
-  lps <- tmp$lps
-  ect <- tmp$ect
-  tpg <- tmp$tpg
-  mav <- tmp$mav
-  mcv <- tmp$mcv
-  mds <- tmp$mds
-  mpt <- tmp$mpt
-  cle <- tmp$cle
-  pmx <- tmp$pmx
-  ds1 <- tmp$ds1
-  ds2 <- tmp$ds2
-  ds3 <- tmp$ds3
-  lpt <- tmp$lpt
   X <- model.matrix(~0+group,tmp)
 
-  N <- length(stl)
+  N <- nrow(tmp)
   T <- max(tmp$time)
 
-  print(paste("Running Imputation",m))
-  fit <- sampling(sm,
-                  list(N=N,T=T,X=X,mds=mds,
-                       stl=stl,lpt=lpt,mdf=mdf,
-                       maf=maf,lps=lps,ect=ect,
-                       tpg=tpg,mav=mav,mcv=mcv,
[…]

### 2023-06-29 R/stuart_stickle_bayes.R
Update stan model

@@ -86,6 +86,7 @@ transformed parameters{
   vector[2*T] mu;
   mu[1:T] = theta_f + u[1:T];
   mu[(T+1):(2*T)] = theta_m + u[(T+1):(2*T)];
+  vector[2*T] lambda = exp(mu);
 }
 
 model{
@@ -136,6 +137,7 @@ transformed parameters{
   vector[2*T] mu_stl;
   mu_stl[1:T] = theta_f_stl + u_stl[1:T];
   mu_stl[(T+1):(2*T)] = theta_m_stl + u_stl[(T+1):(2*T)];
+  vector[2*T] lambda = exp(mu);
 }
 model{
   // Data Model
@@ -189,22 +191,22 @@ for (m in 1:M){
     stl <- tmp$stl
 
     if(var == "mav" | var == "mcv"){
-      fit <- sampling(disc_dep,
+      fit <- sampling(stan_disc_dep,
                       list(N=N,T=T,X=X,y=y,stl=stl),
                       iter=1500,chains=1,warmup=1000,thin=5,
                       pars = c("theta_f","theta_m",
                                "kappa","tau",
-                               "beta","mu"))
+                               "beta","lambda"))
     } else if (substr(var,1,1) == "m") {
-      fit <- sampling(disc_ind,
+      fit <- sampling(stan_disc_ind,
                       list(N=N,T=T,X=X,y=y),
                       iter=1500,chains=1,warmup=1000,thin=5,
                       pars = c("theta_f","theta_m",
                                "kappa","tau",
-                               "beta","mu"))
+                               "lambda"))
     } else if (var == "lps") {
-      fit <- sampling(cont_dep,
-                      list(N=N,T=T,X=X,y=y),
+      fit <- sampling(stan_cont_dep,
+                      list(N=N,T=T,X=X,y=y,stl=stl),
                       iter=1500,chains=1,warmup=1000,thin=5,
                       pars = c("theta_f","theta_m",
                                "kappa","sigma",
@@ -213,8 +215,8 @@ for (m in 1:M){
                                "kappa_stl","sigma_stl",
                                "tau_stl","mu_stl"))
     } else {
-      fit <- sampling(cont_dep,
-                      list(N=N,T=T,X=X,y=y),
+      fit <- sampling(stan_cont_dep,
+                      list(N=N,T=T,X=X,y=y,stl=stl),
                       iter=1500,chains=1,warmup=1000,thin=5,
                       pars = c("theta_f","theta_m",
                                "kappa","sigma",

### 2023-07-01 R/stuart_stickle_bayes.R
Update STAN Model

@@ -225,8 +225,25 @@ for (m in 1:M){
 
     if (var == "lps"){
       imp_fit <- as.data.frame(fit)
+      imp_fit <- imp_fit %>% select(!'lp__')
+      names(imp_fit)[names(imp_fit) == "theta_f"] <- paste0("theta_f_",var)
+      names(imp_fit)[names(imp_fit) == "theta_m"] <- paste0("theta_m_",var)
+      names(imp_fit)[names(imp_fit) == "kappa"] <- paste0("kappa_",var)
+      names(imp_fit)[names(imp_fit) == "sigma"] <- paste0("sigma_",var)
+      names(imp_fit)[names(imp_fit) == "tau"] <- paste0("tau_",var)
+      names(imp_fit)[names(imp_fit) == "beta"] <- paste0("beta_",var)
+      names(imp_fit)[grep("mu\\[",names(imp_fit))] <- paste0("mu_",var,"[",1:36,"]")
+      names(imp_fit)[grep("lambda\\[",names(imp_fit))] <- paste0("lambda_",var,"[",1:36,"]")
     } else {
       imp_fit <- cbind(imp_fit,as.data.frame(fit))
+      names(imp_fit)[names(imp_fit) == "theta_f"] <- paste0("theta_f_",var)
+      names(imp_fit)[names(imp_fit) == "theta_m"] <- paste0("theta_m_",var)
+      names(imp_fit)[names(imp_fit) == "kappa"] <- paste0("kappa_",var)
+      names(imp_fit)[names(imp_fit) == "sigma"] <- paste0("sigma_",var)
+      names(imp_fit)[names(imp_fit) == "tau"] <- paste0("tau_",var)
+      names(imp_fit)[names(imp_fit) == "beta"] <- paste0("beta_",var)
+      names(imp_fit)[grep("mu\\[",names(imp_fit))] <- paste0("mu_",var,"[",1:36,"]")
+      names(imp_fit)[grep("lambda\\[",names(imp_fit))] <- paste0("lambda_",var,"[",1:36,"]")
     }
   }
 

### 2023-07-25 manuscript/paper.Rmd
Update STAN code and manuscript (Note: I, Matt, cannot compile the pdf)

@@
 ## Imputation Model 
-Let $\boldsymbol{W}$ be an $(n_{extant} + n_{fossil}) \times 1$ vector of the covariate gender of the stickleback fish, $\boldsymbol{X}$ be an $(n_{extant} + n_{fossil}) \times K$ matrix of the $K$ continuous phenotypes of interest, and $\boldsymbol{Y}$ be an $(n_{extant} + n_{fossil}) \times L$ matrix of the $L$ discrete phenotypes of interest.
+Let $\boldsymbol{W}$ be an $(n_{extant} + n_{fossil}) \times 1$ vector of the covariate gender of the stickleback fish and $\boldsymbol{Y}$ be an $(n_{extant} + n_{fossil}) \times K$ matrix of the $K$ phenotypes of interest.
 Because the gender of the fossilized stickleback fish is unobservable, we further define $\boldsymbol{W} = (\boldsymbol{W}_{extant}^T,\boldsymbol{W}_{fossil}^T)^T$ where $\boldsymbol{W}_{extant}$ and $\boldsymbol{W}_{fossil}$ are the $n_{extant} \times 1$ and $n_{fossil} \times 1$ vectors of the observed extant gender and missing fossil gender, respectively.
 
-We impute the missing gender for the fossil data by sampling from the posterior predictive distribution $P(\boldsymbol{W}_{fossil}|\boldsymbol{W}_{extant}, \boldsymbol{X}, \boldsymbol{Y})$ using the multiple imputation by chained equations (MICE) algorithm (@MICE) with predictive mean matching.
+We impute the missing gender for the fossil data by sampling from the posterior predictive distribution $P(\boldsymbol{W}_{fossil}|\boldsymbol{W}_{extant},\boldsymbol{Y})$ using the multiple imputation by chained equations (MICE) algorithm (@MICE) with predictive mean matching.
 Traditionally, the choice for the number of completed dats sets is a relatively small number such as $M = 5$ or $M = 10$.
@@
 So just note that $W_{ti}$ is really $W_{fossil,ti}$.
-Not sure how to say this, but we need to make it clear that we are only using the fossil data for the OU modeling.
-}
+Not sure how to say this, but we need to make it clear that we are only using the fossil data for the OU modeling.}
 
-### Continuous Phenotypes
-For a given imputed dataset, let $W_{ti}$ be the imputed gender, $\boldsymbol{X}_{ti}$ be the $K \times 1$ vector of continuous phenotypes, and $\boldsymbol{Y}_{ti}$ be the $L \times 1$ vector of discrete phenotypes for $t = 1, \ldots, T$ and  $i = 1\ldots,n_{t}$).
-
-Define the last element of $\boldsymbol{X}_{ti}$ ($X_{K,ti}$) as the length of the stickleback.  For $k = 1,\cdots,K - 1$, we will assume
+For a given imputed dataset, let $W_{ij}$ be the imputed gender and $\boldsymbol{Y}_{ij}$ be the $K \times 1$ vector of phenotypes for stickleback fish $j$ at time $t_i$ where $i = 1, \ldots, T$ and $j = 1,\ldots,n_{t}$).
+In addition, we denote $Y_{K,ij}$, the last variable in $\boldsymbol{Y}_{ij}$, to be the standard length of the fish.
+We assume that $Y_{K,ij}$ follows
 \begin{align}
-{X}_{k,ti} & \iid \left\{\begin{array}{llll} \mathcal{N}(\mu_{k,ft} & + \beta_k(X_{K,ti} - \mu_{K,ft}),&\sigma_k^2), & W_{ti} = \text{Female} \\ \mathcal{N}(\mu_{k,mt} & + \beta_k(X_{K,ti} - \mu_{K,mt}),&\sigma_k^2), & W_{ti} = \text{Male} \end{array}\right.
-\\
-{X}_{K,ti} & \iid \left\{\begin{array}{lll} \mathcal{N}(\mu_{K,ft},&\sigma_{K}^2), & W_{ti} = \text{Female} \\ \mathcal{N}(\mu_{K,mt},&\sigma_{K}^2), & W_{ti} = \text{Male} \end{array}\right.
+{Y}_{K,ij} & \iid \left\{\begin{array}{lll} \mathcal{N}(\mu_{K,ft_i},&\sigma_{K}^2), & W_{ij} = \text{Female} \\ \mathcal{N}(\mu_{K,mt_i},&\sigma_{K}^2), & W_{ij} = \text{Male} \end{array}\right.
 \label{eq:X}
 \end{align}
-where $\mu_{k,ft}$ and $\mu_{k,mt}$ represent the time-$t$ specific mean of phenotype $k$ for female and male stickleback fish, respectively, and $\beta_k$ is an additional parameter to account for the biological phenomenon that other continuous phenotypes for a particular animal are positively correlated with their length \textbf{CITATION NEEDED}.
-We further set \begin{align}
-\mu_{k,gt} = \theta_{k,g} + u_{k,gt},
-\label{eq:mu}
-\end{align}
-for $g \in \{f,m\}$ where $\theta_{k,g}$ is the overall mean of phenotype $X_k$ for each gender, and $u_{k,gt}$ measures the difference between $\mu_{k,gt}$ and $\theta_{k,g}$.
+
+<!-- ### Continuous Phenotypes -->
+<!-- For a given imputed dataset, let $W_{ti}$ be the imputed gender, $\boldsymbol{X}_{ti}$ be the $K \times 1$ vector of continuous phenotypes, and $\boldsymbol{Y}_{ti}$ be the $L \times 1$ vector of discrete phenotypes for $t = 1, \ldots, T$ and  $i = 1\ldots,n_{t}$).
+-->
+
+<!-- Define the last element of $\boldsymbol{X}_{ti}$ ($X_{K,ti}$) as the length of the stickleback.  For $k = 1,\cdots,K - 1$, we will assume -->
+<!-- \begin{align} -->
+<!-- {X}_{k,ti} & \iid \left\{\begin{array}{llll} \mathcal{N}(\mu_{k,ft} & + \beta_k(X_{K,ti} - \mu_{K,ft}),&\sigma_k^2), & W_{ti} = \text{Female} \\ \mathcal{N}(\mu_{k,mt} & + \beta_k(X_{K,ti} - \mu_{K,mt}),&\sigma_k^2), & W_{ti} = \text{Male} \end{array}\right.
+\\ -->
+<!-- {X}_{K,ti} & \iid \left\{\begin{array}{lll} \mathcal{N}(\mu_{K,ft},&\sigma_{K}^2), & W_{ti} = \text{Female} \\ \mathcal{N}(\mu_{K,mt},&\sigma_{K}^2), & W_{ti} = \text{Male} \end{array}\right. -->
+<!-- \label{eq:X} -->
+<!-- \end{align} -->
+<!-- where $\mu_{k,ft}$ and $\mu_{k,mt}$ represent the time-$t$ specific mean of phenotype $k$ for female and male stickleback fish, respectively, and $\beta_k$ is an additional parameter to account for the biological phenomenon that other continuous phenotypes for a particular animal are positively correlated with their length \textbf{CITATION NEEDED}.
+We further set \begin{align} -->
+<!-- \mu_{k,gt} = \theta_{k,g} + u_{k,gt}, -->
+<!-- \label{eq:mu} -->
+<!-- \end{align} -->
+<!-- for $g \in \{f,m\}$ where $\theta_{k,g}$ is the overall mean of phenotype $X_k$ for each gender, and $u_{k,gt}$ measures the difference between $\mu_{k,gt}$ and $\theta_{k,g}$.
 We then fit $\boldsymbol{u}_{k,g} = \{u_{k,g1},\cdots,u_{k,gT}\}$ to an Ornstein-Uhlenback (OU) process (@OUProcess) where we assume $u_{k,gT}$ have a marginal mean of 0.
 Because each time period is ~1000 years, we will discretize the OU process without loss of generality.
-For $t = 2,\ldots,T$, we assume
-\begin{align}
-u_{k,gt} \iid \mathcal{N}(\kappa_{k} u_{k,g(t-1)} , \tau_k^2).
-\label{eq:u_ar1}
-\end{align}
+For $t = 2,\ldots,T$, we assume -->
+<!-- \begin{align} -->
+<!-- u_{k,gt} \iid \mathcal{N}(\kappa_{k} u_{k,g(t-1)} , \tau_k^2). -->
+<!-- \label{eq:u_ar1} -->
+<!-- \end{align} -->
 
-$\kappa_{k,g}$ represents the correlation between $\mu_{k,gt}$ and $\mu_{k,g(t+1)}$, and it also follows that $\text{cor}(u_{g,t},u_{g,t+h}) = \kappa^h$ for $h \in \mathbb{N}$.  In addition, we assume
-\begin{align}
-u_{k,g1} \iid \mathcal{N}\left(0,\frac{\tau_k^2}{1-\kappa_{k}^2}\right).
-\label{eq:u1}
-\end{align}
-This model choice is to preserve stationarity; i.e. $p(u_{k,gs}) = p(u_{k,gt})$ for $s \neq t$.
+<!-- $\kappa_{k,g}$ represents the correlation between $\mu_{k,gt}$ and $\mu_{k,g(t+1)}$, and it also follows that $\text{cor}(u_{g,t},u_{g,t+h}) = \kappa^h$ for $h \in \mathbb{N}$.
+In addition, we assume -->
+<!-- \begin{align} -->
+<!-- u_{k,g1} \iid \mathcal{N}\left(0,\frac{\tau_k^2}{1-\kappa_{k}^2}\right). -->
+<!-- \label{eq:u1} -->
+<!-- \end{align} -->
+<!-- This model choice is to preserve stationarity; i.e. $p(u_{k,gs}) = p(u_{k,gt})$ for $s \neq t$. -->
 
-### Discrete Phenotypes
-For the discrete phenotypes, we will assume
-\begin{align}
-{Y}_{l,ti} & \sim \left\{\begin{array}{ll} Poisson(\lambda_{l,ft}), & W_{ti} = \text{Female} \\ Poisson(\lambda_{l,mt}), & W_{ti} = \text{Male} \end{array}\right.,
-\label{eq:Y}
-\end{align}
-for $l = 1,\cdots,L$.  Similar to the continuous phenotypes, we set
-\begin{align}
-\log(\lambda_{l,gt}) = \gamma_{l,g} + v_{l,gt}.
-\label{eq:lambda}
-\end{align}
-We use the log function to allow $v_{l,gt}$ to take all values on the real number line so we can properly fit an OU process to these effects.  We further assume
-\begin{align}
-v_{l,gt} \iid \mathcal{N}(\phi_{l} v_{l,g(t-1)} , \omega_l^2),
-\label{eq:v_ar1}
-\end{align}
-and
-\begin{align}
-v_{l,g1} \iid \mathcal{N}\left(0,\frac{\omega_l^2}{1 - \phi_{l}^2}\right)..
-\label{eq:v1}
-\end{align}
+<!-- ### Discrete Phenotypes -->
+<!-- For the discrete phenotypes, we will assume -->
+<!-- \begin{align} -->
+<!-- {Y}_{l,ti} & \sim \left\{\begin{array}{ll} Poisson(\lambda_{l,ft}), & W_{ti} = \text{Female} \\ Poisson(\lambda_{l,mt}), & W_{ti} = \text{Male} \end{array}\right., -->
+<!-- \label{eq:Y} -->
+<!-- \end{align} -->
+<!-- for $l = 1,\cdots,L$.  Similar to the continuous phenotypes, we set -->
+<!-- \begin{align} -->
+<!-- \log(\lambda_{l,gt}) = \gamma_{l,g} + v_{l,gt}. -->
+<!-- \label{eq:lambda} -->
+<!-- \end{align} -->
+<!-- We use the log function to allow $v_{l,gt}$ to take all values on the real number line so we can properly fit an OU process to these effects.  We further assume -->
+<!-- \begin{align} -->
+<!-- v_{l,gt} \iid \mathcal{N}(\phi_{l} v_{l,g(t-1)} , \omega_l^2), -->
+<!-- \label{eq:v_ar1} -->
+<!-- \end{align} -->
+<!-- and -->
+<!-- \begin{align} -->
+<!-- v_{l,g1} \iid \mathcal{N}\left(0,\frac{\omega_l^2}{1 - \phi_{l}^2}\right).. -->
+<!-- \label{eq:v1} -->
+<!-- \end{align} -->
 

### 2023-07-25 R/stuart_stickle_bayes.R

@@ -11,15 +11,13 @@ cont <- "
 data{
   int<lower=0> N;
   int<lower=0> T;
-  matrix[N,2*T] X;
   vector[N] y;
   vector[N] stl;
+  matrix[N,2*T] X;
   vector[T] time;
   vector[T-1] Delta;
-  int<lower=0,upper=1> OU;
-  int<lower=0,upper=1> trend;
-  int<lower=0,upper=1> OU_stl;
-  int<lower=0,upper=1> trend_stl;
+  int<lower=0,upper=3> mod;
+  int<lower=0,upper=3> mod_stl;
 }
 parameters{
   real beta0_f;
@@ -45,24 +43,40 @@ parameters{
 transformed parameters{
   vector[2*T] mu;
   vector[2*T] mu_stl;
+
+  if (mod == 0){
+    mu[1:T] = beta0_f + beta1_f*time + u[1:T];
+    mu[(T+1):(2*T)] = beta0_m + beta1_m*time + u[(T+1):(2*T)];
+  } else if (mod == 1) {
+    mu[1:T] = beta0_f + beta1_f*time;
+    mu[(T+1):(2*T)] = beta0_m + beta1_m*time;
+  } else if (mod == 2){
+    mu[1:T] = beta0_f + u[1:T];
+    mu[(T+1):(2*T)] = beta0_m + u[(T+1):(2*T)];
+  } else {
+    mu[1:T] = beta0_f + 0*time;
+    mu[(T+1):(2*T)] = beta0_m + 0*time;
+  }
+
+  if (mod_stl==0){
+    mu_stl[1:T] = beta0_f + beta1_f*time + u_stl[1:T];
+    mu_stl[(T+1):(2*T)] = beta0_m + beta1_m*time + u_stl[(T+1):(2*T)];
+  } else if (mod_stl == 1) {
+    mu_stl[1:T] = beta0_f + beta1_f*time;
+    mu_stl[(T+1):(2*T)] = beta0_m + beta1_m*time;
+  } else if (mod_stl == 2){
+    mu_stl[1:T] = beta0_f + u_stl[1:T];
+    mu_stl[(T+1):(2*T)] = beta0_m + u_stl[(T+1):(2*T)];
+  } else {
+    mu_stl[1:T] = beta0_f + 0*time;
+    mu_stl[(T+1):(2*T)] = beta0_m + 0*time;
+  }
+
   vector[T-1] OU_mean;
[…]

### 2023-08-03 R/stuart_stickle_bayes.R
Add posterior samples and DIC Check code.

@@ -5,6 +5,9 @@ library(ggridges)
 
 dat <- read.csv("Data/updated_stickle_imputations_100.csv")
 times <- read.csv("RawData_ReadOnly/KSampleMeanTimes.csv")
+times$time <- c(1:18)
+dat <- dat %>% left_join(times %>% select(time,Inverted.Year))
+dat$Inverted.Year <- dat$Inverted.Year/1000
 dat$group <- factor(paste(dat$gender,letters[dat$time]))
 
 cont <- "
@@ -17,33 +20,22 @@ data{
   vector[T] time;
   vector[T-1] Delta;
   int<lower=0,upper=3> mod;
-  int<lower=0,upper=3> mod_stl;
+  int<lower=0,upper=1> ind;
 }
 parameters{
   real beta0_f;
   real beta0_m;
-  real beta0_f_stl;
-  real beta0_m_stl;
   real beta1_f;
   real beta1_m;
-  real beta1_f_stl;
-  real beta1_m_stl;
   real<lower=0> kappa;
-  real<lower=0> kappa_stl;
   real<lower=0> sigma;
-  real<lower=0> sigma_stl;
   real<lower=0> tau;
-  real<lower=0> tau_stl;
   real<lower=0> tau_0;
-  real<lower=0> tau_0_stl;
   real gamma;
   vector[2*T] u;
-  vector[2*T] u_stl;
 }
 transformed parameters{
   vector[2*T] mu;
-  vector[2*T] mu_stl;
-
   if (mod == 0){
     mu[1:T] = beta0_f + beta1_f*time + u[1:T];
     mu[(T+1):(2*T)] = beta0_m + beta1_m*time + u[(T+1):(2*T)];
@@ -58,35 +50,20 @@ transformed parameters{
     mu[(T+1):(2*T)] = beta0_m + 0*time;
   }
 
-  if (mod_stl==0){
-    mu_stl[1:T] = beta0_f + beta1_f*time + u_stl[1:T];
-    mu_stl[(T+1):(2*T)] = beta0_m + beta1_m*time + u_stl[(T+1):(2*T)];
-  } else if (mod_stl == 1) {
-    mu_stl[1:T] = beta0_f + beta1_f*time;
-    mu_stl[(T+1):(2*T)] = beta0_m + beta1_m*time;
-  } else if (mod_stl == 2){
[…]

### 2023-08-03 R/stuart_stickle_bayes_check.R

@@ -1,0 +1,107 @@
+rm(list = ls())
+library(tidyverse)
+library(rstan)
+library(ggridges)
+
+DIC_cont_ind <- function(y,X,mu,sigma){
+  dnorm(y,as.vector(X %*% mu),sigma,log=TRUE) %>% sum
+}
+
+DIC_cont_dep <- function(y,X,mu,stl,gamma,sigma){
+  dnorm(y,as.vector(X %*% mu) + gamma*stl,sigma,log=TRUE) %>% sum
+}
+
+DIC_disc_ind <- function(y,X,mu){
+  dpois(y,exp(as.vector(X %*% mu)),log=TRUE) %>% sum
+}
+
+DIC_disc_dep <- function(y,X,mu,stl,gamma){
+  dpois(y,exp(as.vector(X %*% mu) + gamma*stl),log=TRUE) %>% sum
+}
+
+dat <- read.csv("Data/updated_stickle_imputations_100.csv")
+times <- read.csv("RawData_ReadOnly/KSampleMeanTimes.csv")
+times$time <- c(1:18)
+dat <- dat %>% left_join(times %>% select(time,Inverted.Year))
+dat$Inverted.Year <- dat$Inverted.Year/1000
+dat$group <- factor(paste(dat$gender,letters[dat$time]))
+
+M <- max(dat$imp)
+vars <- c("stl","lps","ect","tpg",
+          "cle","pmx","ds1","ds2",
+          "ds3","lpt","mdf","mav",
+          "maf","mcv","mds","mpt")
+stl <- matrix(dat$stl,ncol=M)
+
+DICs <- NULL
+
+apply(samps,)
+
+for (var in vars){
+  y <- matrix(dat[,var],ncol=M)
+  for (mod in 0:3){
+    samps <- readRDS(paste0("Figures/",var,"/posterior_samples_",var,"_",mod,".RDS"))
+
+    N <- nrow(samps)
+
+    mu <- samps %>% select(contains("mu"))
+    if (substr(var,1,1) != "m"){
+      sigma <- samps %>% select(sigma)
+    }
+    gamma <- samps %>% select(gamma)
+
+    if (substr(var,1,1) != "m"){
+      params <- cbind(mu,gamma,sigma)
+    } else {
+      params <- cbind(mu,gamma)
+    }
+
+    apply(params,1,function(pars){
+      sapply(1:M,function(m){
+        X <- model.matrix(~0+group,dat %>% filter(imp==m))
[…]

### 2023-08-04 manuscript/paper.Rmd
Manuscript now compiles.  There were some latex errors.

@@
 author: | 
-  | \large Allison Ozark$^{1,*}$, Matthew Stuart$^{2,3,*}$, Raheyma Siddiqui$^1$, Akhil Ghosh$^2$, 
-  | \large  Samantha Swank$^1$, Michael A. Bell$^4$, Gregory J. Matthews$^{2,3}$, and Yoel E. Stuart${^1,+$} \vspace{-1.1mm}
-  |
+  | \large Allison Ozark$^{1,*}$, Matthew Stuart$^{2,3,*}$, Raheyma Siddiqui$^1$, Akhil Ghosh$^2$
+  | \large Samantha Swank$^1$, Michael A. Bell$^4$, Gregory J. Matthews$^{2,3}$, and Yoel E. Stuart$^{1,+}$ 
+  | \vspace{-1.1mm}
   | \large $^1$ Department of Biology, Loyola University Chicago, Chicago, IL, USA \vspace{-1.1mm}
-  |
   | \large $^2$ Department of Mathematics and Statistics, Loyola University Chicago, Chicago, IL, USA \vspace{-1.1mm}
-  |  
   | \large $^3$ Center for Data Science and Consulting, Loyola University Chicago, Chicago, IL, USA \vspace{-1.1mm}
-  |  
   | \large $^4$ University of California Museum of Paleontology, Berkeley, CA, USA \vspace{-1.1mm}
-  |
   | \large $^*$ Equal contribution \vspace{-1.1mm}
-  |
   | \large $^+$ Corresponding: ystuart@luc.edu \vspace{-1.1mm}
-  |
 abstract: |
@@
 The fossil lineage appeared in the depositional environment as a fully armored form, with complete pelvic girdles, two pelvic spines, and three dorsal spines, on average (@Bell2006; @Stuart2020).
-However, this habitat appears to have been missing fish piscivores (e.g., trout and other salmonids known to prey on modern stickleback; @Bell2009), relaxing putative selection for armor (@Reichen1980, @Reimchen1994, @BellEtAl1993; @Bowne1994; @Roestietal2023).
+However, this habitat appears to have been missing fish piscivores (e.g., trout and other salmonids known to prey on modern stickleback; @Bell2009), relaxing putative selection for armor (@Reimchen1980, @Reimchen1994, @BellEtAl1993; @Bowne1994; @Roestietal2023).
 The lineage adaptively evolved armor loss but also reduction in the means of several other traits  (@Hunt2008; @Stuart2020; Siddiqui et al.

### 2023-08-08 manuscript/paper.Rmd
Update manuscript

@@
 |----------------------|------------|-----------------------------|
-|Standard Length|	sl|	Distance from anterior tip of premaxilla to posterior end of last vertebra (hypural plate)|
+|Standard Length|	stl|	Distance from anterior tip of premaxilla to posterior end of last vertebra (hypural plate)|
 |Dorsal Spine | mds |	Number of dorsal spines from 0 to 3|
@@
 So just note that $W_{ti}$ is really $W_{fossil,ti}$.
-Not sure how to say this, but we need to make it clear that we are only using the fossil data for the OU modeling.}
+Not sure how to say this, but we need to make it clear that we are only using the fossil data for the OU modeling.
+Greg, check and see if this makes sense.}
 
-For a given imputed dataset, let $W_{ij}$ be the imputed gender and $\boldsymbol{Y}_{ij}$ be the $K \times 1$ vector of phenotypes for stickleback fish $j$ at time $t_i$ where $i = 1, \ldots, T$ and $j = 1,\ldots,n_{t}$).
-In addition, we denote $Y_{K,ij}$, the last variable in $\boldsymbol{Y}_{ij}$, to be the standard length of the fish.
-We assume that $Y_{K,ij}$ follows
+For a given imputed dataset, let $W_{ij}$ be the imputed gender and $\boldsymbol{Y}_{ij}$ be the $K \times 1$ vector of phenotypes for stickleback fossil $j$ at time $t_i$ where $i = 1, \ldots, T$ and $j = 1,\ldots,n_{t}$.
+Note, we are inputting only the fossil data into our model; connecting to the previous section, $W_{ij}$ and $\boldsymbol{Y}_{ij}$ can be interpretted as $W_{fossil,ij}$ and $\boldsymbol{Y}_{fossil,ij}$.
+In addition, we denote $Y_{K,ij}$, the last variable in $\boldsymbol{Y}_{ij}$, to be the standard length of the fish,
 \begin{align}
-{Y}_{K,ij} & \iid \left\{\begin{array}{lll} \mathcal{N}(\mu_{K,ft_i},&\sigma_{K}^2), & W_{ij} = \text{Female} \\ \mathcal{N}(\mu_{K,mt_i},&\sigma_{K}^2), & W_{ij} = \text{Male} \end{array}\right.
-\label{eq:X}
+{Y}_{K,ij} & \iid \left\{\begin{array}{lll} \mathcal{N}(\mu_{K,ft_i},&\sigma_{K}^2), & W_{ij} = \text{Female} \\ \mathcal{N}(\mu_{K,mt_i},&\sigma_{K}^2), & W_{ij} = \text{Male} \end{array}\right..
+\label{eq:stl}
 \end{align}
 
-<!-- ### Continuous Phenotypes -->
-<!-- For a given imputed dataset, let $W_{ti}$ be the imputed gender, $\boldsymbol{X}_{ti}$ be the $K \times 1$ vector of continuous phenotypes, and $\boldsymbol{Y}_{ti}$ be the $L \times 1$ vector of discrete phenotypes for $t = 1, \ldots, T$ and  $i = 1\ldots,n_{t}$).
--->
+It is reasonable to assume that the other continuous traits of stickleback fish will have some correlation with its standard length \textbf{(CITATION)}.
+We account for this by adding an additional parameter, $\gamma_k$, into our model.
+More specifically, if $Y_{k,ij}$ is a continuous trait, then
+\begin{align}
+{Y}_{k,ij} & \iid \left\{\begin{array}{llll} \mathcal{N}(\mu_{k,ft_i} + \gamma_kY_{K,ij},\sigma_k^2), & W_{ij} = \text{Female} \\ \mathcal{N}(\mu_{k,mt_i} + \gamma_kY_{K,ij},\sigma_k^2), & W_{ij} = \text{Male} \end{array}\right..
+\label{eq:cont}
+\end{align}
 
-<!-- Define the last element of $\boldsymbol{X}_{ti}$ ($X_{K,ti}$) as the length of the stickleback.  For $k = 1,\cdots,K - 1$, we will assume -->
-<!-- \begin{align} -->
-<!-- {X}_{k,ti} & \iid \left\{\begin{array}{llll} \mathcal{N}(\mu_{k,ft} & + \beta_k(X_{K,ti} - \mu_{K,ft}),&\sigma_k^2), & W_{ti} = \text{Female} \\ \mathcal{N}(\mu_{k,mt} & + \beta_k(X_{K,ti} - \mu_{K,mt}),&\sigma_k^2), & W_{ti} = \text{Male} \end{array}\right.
-\\ -->
-<!-- {X}_{K,ti} & \iid \left\{\begin{array}{lll} \mathcal{N}(\mu_{K,ft},&\sigma_{K}^2), & W_{ti} = \text{Female} \\ \mathcal{N}(\mu_{K,mt},&\sigma_{K}^2), & W_{ti} = \text{Male} \end{array}\right. -->
-<!-- \label{eq:X} -->
-<!-- \end{align} -->
-<!-- where $\mu_{k,ft}$ and $\mu_{k,mt}$ represent the time-$t$ specific mean of phenotype $k$ for female and male stickleback fish, respectively, and $\beta_k$ is an additional parameter to account for the biological phenomenon that other continuous phenotypes for a particular animal are positively correlated with their length \textbf{CITATION NEEDED}.
-We further set \begin{align} -->
-<!-- \mu_{k,gt} = \theta_{k,g} + u_{k,gt}, -->
-<!-- \label{eq:mu} -->
-<!-- \end{align} -->
-<!-- for $g \in \{f,m\}$ where $\theta_{k,g}$ is the overall mean of phenotype $X_k$ for each gender, and $u_{k,gt}$ measures the difference between $\mu_{k,gt}$ and $\theta_{k,g}$.
-We then fit $\boldsymbol{u}_{k,g} = \{u_{k,g1},\cdots,u_{k,gT}\}$ to an Ornstein-Uhlenback (OU) process (@OUProcess) where we assume $u_{k,gT}$ have a marginal mean of 0.
-Because each time period is ~1000 years, we will discretize the OU process without loss of generality.
-For $t = 2,\ldots,T$, we assume -->
-<!-- \begin{align} -->
-<!-- u_{k,gt} \iid \mathcal{N}(\kappa_{k} u_{k,g(t-1)} , \tau_k^2). -->
-<!-- \label{eq:u_ar1} -->
-<!-- \end{align} -->
+In addition, we assume that the discrete traits abdominal vertebrae (mav) and caudal vertebrae (mcv) also have correlation with the standard length of the fish (\textbf{CITATION}).
+If $Y_{k,ij}$ is one of the above traits, then
+\begin{align}
+{Y}_{k,ij} & \sim \left\{\begin{array}{ll} Poisson(\exp\{\mu_{k,ft_i} + \gamma_kY_{K,ij}\}), & W_{ij} = \text{Female} \\ Poisson(\exp\{\mu_{k,mt_i} + \gamma_kY_{K,ij}\}), & W_{ij} = \text{Male} \end{array}\right..
+\label{eq:disc_corr}
+\end{align}
 
-<!-- $\kappa_{k,g}$ represents the correlation between $\mu_{k,gt}$ and $\mu_{k,g(t+1)}$, and it also follows that $\text{cor}(u_{g,t},u_{g,t+h}) = \kappa^h$ for $h \in \mathbb{N}$.
-In addition, we assume -->
-<!-- \begin{align} -->
-<!-- u_{k,g1} \iid \mathcal{N}\left(0,\frac{\tau_k^2}{1-\kappa_{k}^2}\right). -->
-<!-- \label{eq:u1} -->
-<!-- \end{align} -->
-<!-- This model choice is to preserve stationarity; i.e. $p(u_{k,gs}) = p(u_{k,gt})$ for $s \neq t$. -->
+For the other discrete traits, we also assume the above model except we set $\gamma_k = 0$ because of the assumption of no correlation between the standard length and these traits.
+More specifically, we assume
+\begin{align}
+{Y}_{k,ij} & \sim \left\{\begin{array}{ll} Poisson(\exp\{\mu_{k,ft_i}\}), & W_{ij} = \text{Female} \\ Poisson(\exp\{\mu_{k,mt_i}\}), & W_{ij} = \text{Male} \end{array}\right..
+\label{eq:disc_ind}
+\end{align}
 
-<!-- ### Discrete Phenotypes -->
-<!-- For the discrete phenotypes, we will assume -->
-<!-- \begin{align} -->
-<!-- {Y}_{l,ti} & \sim \left\{\begin{array}{ll} Poisson(\lambda_{l,ft}), & W_{ti} = \text{Female} \\ Poisson(\lambda_{l,mt}), & W_{ti} = \text{Male} \end{array}\right., -->
-<!-- \label{eq:Y} -->
-<!-- \end{align} -->
-<!-- for $l = 1,\cdots,L$.  Similar to the continuous phenotypes, we set -->
-<!-- \begin{align} -->
-<!-- \log(\lambda_{l,gt}) = \gamma_{l,g} + v_{l,gt}. -->
-<!-- \label{eq:lambda} -->
-<!-- \end{align} -->
-<!-- We use the log function to allow $v_{l,gt}$ to take all values on the real number line so we can properly fit an OU process to these effects.  We further assume -->
-<!-- \begin{align} -->
-<!-- v_{l,gt} \iid \mathcal{N}(\phi_{l} v_{l,g(t-1)} , \omega_l^2), -->
-<!-- \label{eq:v_ar1} -->
-<!-- \end{align} -->
-<!-- and -->
-<!-- \begin{align} -->
-<!-- v_{l,g1} \iid \mathcal{N}\left(0,\frac{\omega_l^2}{1 - \phi_{l}^2}\right).. -->
-<!-- \label{eq:v1} -->
-<!-- \end{align} -->
+In the above model descriptions, $\mu_{k,ft_i}$ and $\mu_{k,mt_i}$ model the time-$t_i$ specific mean of phenotype $k$ for female and male stickleback fish, respectively.
+We point out that, for the discrete traits, the means are represented by $\exp\{\mu_{k,ft_i}\}$ and $\exp\{\mu_{k,mt_i}\}$ for ease of use in our modeling technique.
+We further set
+\begin{align}
+\mu_{k,gt_i} = \beta_{0,kg} + \beta_{1,kg}t_i + u_{k,gt_i},
+\label{eq:mu}
+\end{align}
+for $g \in \{f,m\}$ where $\beta_{0,kg}$ and $\beta_{1,kg}$ are regression parameters of phenotype $Y_k$ for each gender, accounting for the possibility of a time-dependent trend in the mean structure, and $u_{k,gt_i}$ is the corresponding residual.
+To account for potential correlations between the residuals for a given trait $k$ and gender $g$, we fit an Ornstein-Uhlenback (OU) process (@OUProcess).
+More specifically, define $du_{k,gt} = u_{k,g(t + dt)} - u_{k,gt}$, the change in $u_k,gt$ for a given trait $k$ and gender $g$ over a miniscule time period $dt$.
+The OU process is defined as
+\begin{align}
+du_{k,gt} = -\kappa_k u_{k,gt} dt + \tau_k dW_t,
+\label{eq:OU_cont}
+\end{align}
+where $\kappa_k$ is a parameter associated with the correlation between $u_{k,gt}$ and $u_{k,g(t+dt)}$, $\tau_k$ is the standard deviation of the OU process, and $W_t$ is a standard Brownian motion.
+As shown in (@OUProcess), the closed form solution for the SDE in (\ref{eq:OU_cont}) is
+\begin{align}
+u_{k,gt_i} \iid \mathcal{N}\left(u_{k,gt_{i-1}}\exp\{-\kappa_k(t_{i} - t_{i-1})\} , \frac{\tau_k^2(1 - \exp\{-2\kappa_k(t_{i} - t_{i-1}))\}}{2\kappa_k}\right)
+\label{eq:OU_sol}
+\end{align}
+for $i = 2,\cdots,T$. In a traditional OU process, the initial value $u_{k,gt_1}$ is assumed to be a (potentially unknown) constant, and we discuss the procedure for estimating this value below.
+
+In our empirical study, we analyze the following four nested models for the mean process outlined in (\ref{eq:mu}):
+\begin{itemize}
+\item OU with Trend: The mean process for each phenotype $k=1,\cdots,K$ and $g \in \{f,m\}$ as previously defined.
+\item OU with No Trend: No linear trend on the mean processes and the overall mean of the phenotype is constant for each gender.
+This is achieved by setting $\beta_{1,kg} = 0$ in (\ref{eq:mu}) for $k = 1,\cdots,K$ and $g \in \{f,m\}$
+\item No OU with Trend: No random fluctuations on the mean processes, i.e.
+the OU process assumption is not needed.
+This is achieved by setting $u_{k,gt_i} = 0$ in (\ref{eq:mu}) for $k = 1,\cdots,K$, $g \in \{f,m\}$, and $i = 1,\cdots,T$.
+\item No OU with No Trend: The mean for a specific trait and specific gender at each time point is a constant value.
+This is achieved by setting $\beta_{1,kg} = 0$ and $u_{k,gt_i} = 0$ in (\ref{eq:mu}) for $k = 1,\cdots,K$, $g \in \{f,m\}$, and $i = 1,\cdots,T$.
+\end{itemize}
 
@@
 
-Priors: For $k = 1,\cdots,K$ and $l = 1,\cdots,L$,
+Priors: For $k = 1,\cdots,K$,
 
 \begin{align}
-\sigma_k & \iid \mathcal{N}(0,100)I_{\{\sigma > 0\}} \nonumber \\
-\tau_k & \iid \mathcal{N}(0,100)I_{\{\tau > 0\}} \nonumber \\
-\kappa_k & \iid \mathcal{N}(0,1)I_{\{-1 < \kappa_g < 1\}} \nonumber \\
-\beta_k & \iid \mathcal{N}(0,5) \nonumber \\
-\theta_{k,g} & \iid \mathcal{N}(0,10000) \nonumber \\
-\omega_l & \iid \mathcal{N}(0,100)I_{\{\tau > 0\}} \nonumber \\
-\phi_l & \iid \mathcal{N}(0,1)I_{\{-1 < \kappa_g < 1\}} \nonumber \\
-\gamma_{l,g} & \iid \mathcal{N}(0,10000)
+u_{k,gt_1} & \iid \mathcal{N}(0,\tau_{0,k}) \nonumber \\
+\sigma_k & \iid \mathcal{N}(0,10)I_{\{\sigma > 0\}} \nonumber \\
+\tau_k & \iid \mathcal{N}(0,10)I_{\{\tau > 0\}} \nonumber \\
+\tau_{0,k} & \iid \mathcal{N}(0,20)I_{\{\tau > 0\}} \nonumber \\
+\kappa_k & \iid \mathcal{N}(0,1)I_{\{\kappa_g > 0\}} \nonumber \\
+\gamma_{k} & \iid \mathcal{N}(0,5) \nonumber \\
+\beta_{0,kg} & \iid \mathcal{N}(0,100) \nonumber \\
+\beta_{1,kg} & \iid \mathcal{N}(0,3) \nonumber \\
 \label{eq:priors}
 \end{align}
+We also note that, for the discrete phenotypes in equations (\ref{eq:disc_corr}) and (\ref{eq:disc_ind}), there is no $\sigma_k$, and thus is not sampled for those phenotypes.
 
-All models were built using @R2022language
+All models were built using @R2022language and (\textbf STAN)
 

### 2023-08-08 manuscript/paper.tex
Moved imputed data set to data folder, and updated STAN code appropriately to run faster.
Update manuscript

@@
 stickleback following ecological release from fish piscivores}
-\author{\large Allison Ozark\(^1\), Akhil Ghosh\(^1\), Raheyma
-Siddiqui\(^1\), Matthew Stuart\(^1\),\\
-\large  Samantha Swank\(^1\), Michael A. Bell\(^2\), Gregory J.
-Matthews\(^1\), and Yoel E. Stuart\(^1\) \vspace{-1.1mm}\\
-\strut \\
-\large \(^1\) Loyola University Chicago, Chicago, IL, USA
+\author{\large Allison Ozark\(^{1,*}\), Matthew Stuart\(^{2,3,*}\),
+Raheyma Siddiqui\(^1\), Akhil Ghosh\(^2\)\\
+\large Samantha Swank\(^1\), Michael A. Bell\(^4\), Gregory J.
+Matthews\(^{2,3}\), and Yoel E. Stuart\(^{1,+}\)\\
 \vspace{-1.1mm}\\
-\strut \\
-\large \(^2\) University of California Museum of Paleontology, Berkeley,
-CA, USA \vspace{-1.1mm}\\}
+\large \(^1\) Department of Biology, Loyola University Chicago, Chicago,
+IL, USA \vspace{-1.1mm}\\
+\large \(^2\) Department of Mathematics and Statistics, Loyola
+University Chicago, Chicago, IL, USA \vspace{-1.1mm}\\
+\large \(^3\) Center for Data Science and Consulting, Loyola University
+Chicago, Chicago, IL, USA \vspace{-1.1mm}\\
+\large \(^4\) University of California Museum of Paleontology, Berkeley,
+CA, USA \vspace{-1.1mm}\\
+\large \(^*\) Equal contribution \vspace{-1.1mm}\\
+\large \(^+\) Corresponding:
+\href{mailto:ystuart@luc.edu}{\nolinkurl{ystuart@luc.edu}}
+\vspace{-1.1mm}}
 \date{}
@@
 Ecological release theory suggests that a prey population's niche should
-expand when predation is relaxed or removed (reviewed in Herrmann,
-Stroud, and Losos (\protect\hyperlink{ref-Herrmann2021}{2021})). Niche
-changes in habitat and resource use should then drive phenotypic
-evolution as the population adapts to new conditions (Herrmann, Stroud,
-and Losos (\protect\hyperlink{ref-Herrmann2021}{2021})). For example,
-resource use expansion should result in trophic trait evolution (cite;
-e.g.,). Habitat expansion should result in locomotor change (cite;
-e.g.,). And release from enemies should result in reduction of costly
-defense traits (cite; e.g.,).
-
-Niche expansion could manifest through a combination of increasing
-within- and between-individual niche widths (Bolnick et al.
+expand when predation is relaxed or removed, as prey take advantage of
+new enemy free space (reviewed in Herrmann, Stroud, and Losos
+(\protect\hyperlink{ref-Herrmann2021}{2021})). Niche expansion could
+manifest through a combination of increasing within- and
+between-individual niche widths (Bolnick et al.
 (\protect\hyperlink{ref-Bolnick2010}{2010}); Herrmann, Stroud, and Losos
@@
 Boughman (\protect\hyperlink{ref-Cooper2011}{2011})). Theory and
-empirical data suggest that a population presented with novel ecological
-opportunity might experience disruptive selection on males and females
-stemming from intraspecific competition over newly accessible resources,
-resulting in intersexual divergence in habitat use and associated
-phenotypes (Schoener 1968; Shine
+empirical data suggest that a population presented with ecological
+opportunity following ecological release might experience disruptive
+selection on males and females stemming from intraspecific competition
+over newly accessible resources, resulting in intersexual divergence in
+habitat use (W. (\protect\hyperlink{ref-Schoener1968}{1968}); Shine
 (\protect\hyperlink{ref-Shine1989}{1989}); Bolnick and Doebeli
@@
 al. (\protect\hyperlink{ref-Stuart2021}{2021}); Blain
-(\protect\hyperlink{ref-Blain2022}{2022})). Such character displacement
-between the sexes is therefore one explanation for sexual dimorphism (de
-Lisle papers).
+(\protect\hyperlink{ref-Blain2022}{2022})). Accordingly, sexes should
+also diverge in morphological traits associated with habitat use; such
+ecologically-driven ``character displacement between the sexes'' is
+therefore one explanation for phenotypic sexual dimorphism (S. P. De
+Lisle and Rowe (\protect\hyperlink{ref-deLisleRowe2015}{2015}); S. Paiva
+De Lisle S. P. and Rowe (\protect\hyperlink{ref-deLisleNco2018}{2018})).
 
 We tested this link between release from predators and the evolution of
-sexual dimorphism in a well-preserved, high resolution lineage of the
-fossil stickleback fish (Gasterosteus doryssus). The sex of the
-individual fossils, which is unobserved in practice, was imputed using
-multiple imputation by chained equations (MICE) (Buuren and
-Groothuis-Oudshoorn (\protect\hyperlink{ref-MICE}{2011})) with observed
-data used to train the imputation model based on extant stickleback with
-known sex. Then we tracked the multivariate evolution of sexual
-dimorphism over \textasciitilde16,000 years for traits related to
-swimming, feeding and defense. The fossil sequence captures a bout of
-adaptive evolution following ecological release from predators, as
-follows. The fossil lineage appeared in the depositional environment as
-a fully armored form, with complete pelvic girdles, two pelvic spines,
-and three dorsal spines, on average (Michael A. Bell, Travis, and Blouw
-(\protect\hyperlink{ref-Bell2006}{2006}); Stuart, Travis, and Bell
-(\protect\hyperlink{ref-Stuart2020}{2020})). However, this habitat
-appears to have been missing fish piscivores (e.g., trout and other
-salmonids known to prey on modern stickleback; M. A. Bell
+sexual dimorphism in a well-preserved, finely-resolved,
+\textasciitilde16,000 year-long sequence of the fossil stickleback fish
+(Gasterosteus doryssus). The fossil lineage appeared in the depositional
+environment as a fully armored form, with complete pelvic girdles, two
+pelvic spines, and three dorsal spines, on average (Michael A. Bell,
+Travis, and Blouw (\protect\hyperlink{ref-Bell2006}{2006}); Stuart,
+Travis, and Bell (\protect\hyperlink{ref-Stuart2020}{2020})). However,
+this habitat appears to have been missing fish piscivores (e.g., trout
+and other salmonids known to prey on modern stickleback; M. A. Bell
 (\protect\hyperlink{ref-Bell2009}{2009})), relaxing putative selection
-for armor (Reimchen papers, Bell papers, Bowne
-(\protect\hyperlink{ref-Bowne1994}{1994}); Roesti et al.~2023). The
-lineage thus adaptively evolved armor loss and reduction in several
-other traits as well (Hunt, Bell, and Travis
-(\protect\hyperlink{ref-Hunt2008}{2008}); Stuart, Travis, and Bell
-(\protect\hyperlink{ref-Stuart2020}{2020}); Siddiqui et al.~in prep),
-and tooth wear data suggest that the population also began eating more
-planktonic prey, expanding to an open water lifestyle (Purnell et
-al.~2007).
+for armor (T. E. Reimchen (\protect\hyperlink{ref-Reimchen1980}{1980}),
+T. E. Reimchen (\protect\hyperlink{ref-Reimchen1994}{1994}), O. Bell M.
+A. and Koenings (\protect\hyperlink{ref-BellEtAl1993}{1993}); Bowne
+(\protect\hyperlink{ref-Bowne1994}{1994}); Roesti and Schluter
+(\protect\hyperlink{ref-Roestietal2023}{2023})). The lineage adaptively
+evolved armor loss but also reduction in the means of several other
+traits (Hunt, Bell, and Travis (\protect\hyperlink{ref-Hunt2008}{2008});
+Stuart, Travis, and Bell (\protect\hyperlink{ref-Stuart2020}{2020});
+Siddiqui et al.~in prep). In particular, tooth wear data suggest that
+the population shifted to eat more planktonic prey through time,
+indicative of increased open water habitat use (Purnell et al.~2007). If
+this divergence in habitat use is partitioned between sexes (Reimchen T.
+E. and Nosil (\protect\hyperlink{ref-ReimchenNosil2004}{2004});
+Spoljaric and Reimchen
+(\protect\hyperlink{ref-SpoljaricReimchen2008}{2008});D. S. Reimchen T.
+E. (\protect\hyperlink{ref-Reimchen2016}{2016})), then we should expect
+sexual dimorphism to evolve for traits related to habitat use in G.
+doryssus.
 
-Sexual Dimorphism is interesting because\ldots\ldots{}
-
-Sexual Dimoprhism lit review: Saitta et al.
-(\protect\hyperlink{ref-SaittaEtAl2020}{2020})
-
-The remainder of this manuscript contains a description of the data in
-section \ref{sec:data} and a description of our models in
-\ref{sec:models}. Section \ref{sec:results} presents a summary of our
-results, and we end with our conclusion and future work in section
-\ref{sec:conclusions}.
+To test this prediction, we imputed the sex of individual fossils, which
+is unobserved in practice, with Multiple Imputation by Chained Equations
+(MICE) (Buuren and Groothuis-Oudshoorn
+(\protect\hyperlink{ref-MICE}{2011})), using observed data from extant
+stickleback of known sex to train the imputation model. Then we tracked
+the multivariate evolution of sexual dimorphism over
+\textasciitilde16,000 years for 16 traits related to swimming, feeding
+and defense.
 
@@
 
-A major challenge in this work is inferring the sex of fossil specimens.
-This typically cannot be done directly, except for lineages whose sexes
-are distinguished by the presence or absence of sex-specific characters
-(Hone and Mallon (\protect\hyperlink{ref-Hone2017}{2017}); Saitta et al.
+A major challenge in this work is inferring the sex of fossil
+stickleback specimens. Indeed, for most species, sex cannot be detected
+directly, except for lineages whose sexes are distinguished by the
+presence or absence of sex-specific characters (Hone and Mallon
+(\protect\hyperlink{ref-Hone2017}{2017}); Saitta et al.
 (\protect\hyperlink{ref-SaittaEtAl2020}{2020})). Instead,
@@
 in growth curves (e.g., Saitta et al.
-(\protect\hyperlink{ref-SaittaEtAl2020}{2020})).
-
-We used a third approach to infer fossil sex and study the evolution of
-sexual dimorphism by taking advantage of the fact that G. doryssus is
-part of the extant threespine stickleback species complex (Gastersteus
-aculeatus). Along with the fossil data collected for this study, we also
-have data from extant populations of the Threespine Stickleback,
-Gasterosteus aculeatus, measuring the same traits for individuals whose
-sex could be determined directly. We then combined these sets of data
-and treat the sex of the fossil specimens as missing and used multiple
-imputation (Little and Rubin
-(\protect\hyperlink{ref-little2002statistical}{2002})) to impute the sex
-of the fossils many times. Imputations were performed using the MICE
-(Buuren and Groothuis-Oudshoorn (\protect\hyperlink{ref-MICE}{2011}))
-algorithm and implemented in R (R Core Team
-(\protect\hyperlink{ref-R2022language}{2022})) with M = 100 imputations
-performed.
-
-After imputing the sex of the fossil data, each completed data set is
-then used to fit a modified Ornstein--Uhlenbeck (OU) (Uhlenbeck and
-Ornstein (\protect\hyperlink{ref-OUProcess}{1930})) model using a
-Bayesian framework to look for sexual dimorphism across a variety of
-stickleback phenotypes
+(\protect\hyperlink{ref-SaittaEtAl2020}{2020})). We used a third
+approach to infer fossil sex and study the evolution of sexual
+dimorphism by taking advantage of the fact that G. doryssus is part of
+the extant threespine stickleback species complex, Gasterosteus
+aculeatus. We measured the same trait set from a phenotypically diverse
+set of extant G. aculeatus populations, for whom individual sex was
+known. We then combined fossil and extant data sets, treated fossil sex
+as missing data, and used the MICE (Buuren and Groothuis-Oudshoorn
+(\protect\hyperlink{ref-MICE}{2011})) multiple imputation (Little and
+Rubin (\protect\hyperlink{ref-little2002statistical}{2002})) algorithm,
+and implemented in R (Team
+(\protect\hyperlink{ref-R2022language}{2022})), to impute individual sex
+of the fossil data set 100 times. We then fit a modified
+Ornstein--Uhlenbeck (OU) (Uhlenbeck and Ornstein
+(\protect\hyperlink{ref-OUProcess}{1930})) model using a Bayesian
+framework to the data to test for the evolution of sexual dimorphism in
+each trait.
 
@@
 The fossil data consists of 814 specimens with unknown sex over 18 time
-periods spaced \textasciitilde1000 years apart. Table 2 shows the sample
-size at each of the 18 time periods. There are at least 22 specimens at
-each time period with a high of 67 specimens in period 7.
+periods spaced \textasciitilde1000 years apart. Figure 1 shows the
+sample size at each of the 18 time periods. There are at least 22
+specimens at each time period with a high of 67 specimens in period 7.
+
+\begin{figure}
+
+{\centering \includegraphics{paper_files/figure-latex/unnamed-chunk-2-1}
+
+}
+
+\caption{Fossil sampel size for each time period}\label{fig:unnamed-chunk-2}
+\end{figure}
 
@@
 \endhead
-Standard Length & sl & Distance from anterior tip of premaxilla to
+Standard Length & stl & Distance from anterior tip of premaxilla to
 posterior end of last vertebra (hypural plate) \\
@@
 \end{longtable}
-
-\begin{table}[ht]
-\centering
-\caption{The number of stickleback fossils at each time point.  Sample size at each time point ranges from a low of 22 to a high of 67.}
-\begin{tabular}{rl}
-  \hline
-time & count \\ 
-  \hline
-  1 & 43 \\ 
-    2 & 41 \\ 
-    3 & 51 \\ 
-    4 & 41 \\ 
-    5 & 46 \\ 
-    6 & 48 \\ 
-    7 & 67 \\ 
-    8 & 55 \\ 
-    9 & 42 \\ 
-   10 & 33 \\ 
-   11 & 37 \\ 
-   12 & 22 \\ 
-   13 & 41 \\ 
-   14 & 43 \\ 
-   15 & 46 \\ 
-   16 & 47 \\ 
-   17 & 56 \\ 
-   18 & 55 \\ 
-   \hline
-\end{tabular}
-\label{tabn}
-
-\end{table}
 
@@
 Let \(\boldsymbol{W}\) be an \((n_{extant} + n_{fossil}) \times 1\)
-vector of the covariate gender of the stickleback fish,
-\(\boldsymbol{X}\) be an \((n_{extant} + n_{fossil}) \times K\) matrix
-of the \(K\) continuous phenotypes of interest, and \(\boldsymbol{Y}\)
-be an \((n_{extant} + n_{fossil}) \times L\) matrix of the \(L\)
-discrete phenotypes of interest. Because the gender of the fossilized
-stickleback fish is unobservable, we further define
+vector of the covariate gender of the stickleback fish and
+\(\boldsymbol{Y}\) be an \((n_{extant} + n_{fossil}) \times K\) matrix
+of the \(K\) phenotypes of interest. Because the gender of the
+fossilized stickleback fish is unobservable, we further define
 \(\boldsymbol{W} = (\boldsymbol{W}_{extant}^T,\boldsymbol{W}_{fossil}^T)^T\)
@@
 posterior predictive distribution
-\(P(\boldsymbol{W}_{fossil}|\boldsymbol{W}_{extant}, \boldsymbol{X}, \boldsymbol{Y})\)
+\(P(\boldsymbol{W}_{fossil}|\boldsymbol{W}_{extant},\boldsymbol{Y})\)
 using the multiple imputation by chained equations (MICE) algorithm
@@
 Not sure how to say this, but we need to make it clear that we are only using the fossil data for the OU modeling.
-}
+Greg, check and see if this makes sense.}
 
-\hypertarget{continuous-phenotypes}{%
-\subsubsection{Continuous Phenotypes}\label{continuous-phenotypes}}
-
-For a given imputed dataset, let \(W_{ti}\) be the imputed gender,
-\(\boldsymbol{X}_{ti}\) be the \(K \times 1\) vector of continuous
-phenotypes, and \(\boldsymbol{Y}_{ti}\) be the \(L \times 1\) vector of
-discrete phenotypes for \(t = 1, \ldots, T\) and \(i = 1\ldots,n_{t}\)).
-
-Define the last element of \(\boldsymbol{X}_{ti}\) (\(X_{K,ti}\)) as the
-length of the stickleback. For \(k = 1,\cdots,K - 1\), we will assume
-\begin{align}
-{X}_{k,ti} & \overset{iid}{\sim}\left\{\begin{array}{llll} \mathcal{N}(\mu_{k,ft} & + \beta_k(X_{K,ti} - \mu_{K,ft}),&\sigma_k^2), & W_{ti} = \text{Female} \\ \mathcal{N}(\mu_{k,mt} & + \beta_k(X_{K,ti} - \mu_{K,mt}),&\sigma_k^2), & W_{ti} = \text{Male} \end{array}\right.
-\\
-{X}_{K,ti} & \overset{iid}{\sim}\left\{\begin{array}{lll} \mathcal{N}(\mu_{K,ft},&\sigma_{K}^2), & W_{ti} = \text{Female} \\ \mathcal{N}(\mu_{K,mt},&\sigma_{K}^2), & W_{ti} = \text{Male} \end{array}\right.
-\label{eq:X}
-\end{align} where \(\mu_{k,ft}\) and \(\mu_{k,mt}\) represent the
-time-\(t\) specific mean of phenotype \(k\) for female and male
-stickleback fish, respectively, and \(\beta_k\) is an additional
-parameter to account for the biological phenomenon that other continuous
-phenotypes for a particular animal are positively correlated with their
-length \textbf{CITATION NEEDED}. We further set \begin{align}
-\mu_{k,gt} = \theta_{k,g} + u_{k,gt},
-\label{eq:mu}
-\end{align} for \(g \in \{f,m\}\) where \(\theta_{k,g}\) is the overall
-mean of phenotype \(X_k\) for each gender, and \(u_{k,gt}\) measures the
-difference between \(\mu_{k,gt}\) and \(\theta_{k,g}\). We then fit
-\(\boldsymbol{u}_{k,g} = \{u_{k,g1},\cdots,u_{k,gT}\}\) to an
-Ornstein-Uhlenback (OU) process (Uhlenbeck and Ornstein
-(\protect\hyperlink{ref-OUProcess}{1930})) where we assume \(u_{k,gT}\)
-have a marginal mean of 0. Because each time period is
-\textasciitilde1000 years, we will discretize the OU process without
-loss of generality. For \(t = 2,\ldots,T\), we assume \begin{align}
-u_{k,gt} \overset{iid}{\sim}\mathcal{N}(\kappa_{k} u_{k,g(t-1)} , \tau_k^2).
-\label{eq:u_ar1}
+For a given imputed dataset, let \(W_{ij}\) be the imputed gender and
+\(\boldsymbol{Y}_{ij}\) be the \(K \times 1\) vector of phenotypes for
+stickleback fossil \(j\) at time \(t_i\) where \(i = 1, \ldots, T\) and
+\(j = 1,\ldots,n_{t}\). Note, we are inputting only the fossil data into
+our model; connecting to the previous section, \(W_{ij}\) and
+\(\boldsymbol{Y}_{ij}\) can be interpretted as \(W_{fossil,ij}\) and
+\(\boldsymbol{Y}_{fossil,ij}\). In addition, we denote \(Y_{K,ij}\), the
+last variable in \(\boldsymbol{Y}_{ij}\), to be the standard length of
+the fish, \begin{align}
+{Y}_{K,ij} & \overset{iid}{\sim}\left\{\begin{array}{lll} \mathcal{N}(\mu_{K,ft_i},&\sigma_{K}^2), & W_{ij} = \text{Female} \\ \mathcal{N}(\mu_{K,mt_i},&\sigma_{K}^2), & W_{ij} = \text{Male} \end{array}\right..
+\label{eq:stl}
 \end{align}
 
-\(\kappa_{k,g}\) represents the correlation between \(\mu_{k,gt}\) and
-\(\mu_{k,g(t+1)}\), and it also follows that
-\(\text{cor}(u_{g,t},u_{g,t+h}) = \kappa^h\) for \(h \in \mathbb{N}\).
-In addition, we assume \begin{align}
-u_{k,g1} \overset{iid}{\sim}\mathcal{N}\left(0,\frac{\tau_k^2}{1-\kappa_{k}^2}\right).
-\label{eq:u1}
-\end{align} This model choice is to preserve stationarity;
-i.e.~\(p(u_{k,gs}) = p(u_{k,gt})\) for \(s \neq t\).
+It is reasonable to assume that the other continuous traits of
+stickleback fish will have some correlation with its standard length
+\textbf{(CITATION)}. We account for this by adding an additional
+parameter, \(\gamma_k\), into our model. More specifically, if
+\(Y_{k,ij}\) is a continuous trait, then \begin{align}
+{Y}_{k,ij} & \overset{iid}{\sim}\left\{\begin{array}{llll} \mathcal{N}(\mu_{k,ft_i} + \gamma_kY_{K,ij},\sigma_k^2), & W_{ij} = \text{Female} \\ \mathcal{N}(\mu_{k,mt_i} + \gamma_kY_{K,ij},\sigma_k^2), & W_{ij} = \text{Male} \end{array}\right..
+\label{eq:cont}
+\end{align}
 
-\hypertarget{discrete-phenotypes}{%
-\subsubsection{Discrete Phenotypes}\label{discrete-phenotypes}}
+In addition, we assume that the discrete traits abdominal vertebrae
+(mav) and caudal vertebrae (mcv) also have correlation with the standard
+length of the fish (\textbf{CITATION}). If \(Y_{k,ij}\) is one of the
+above traits, then \begin{align}
+{Y}_{k,ij} & \sim \left\{\begin{array}{ll} Poisson(\exp\{\mu_{k,ft_i} + \gamma_kY_{K,ij}\}), & W_{ij} = \text{Female} \\ Poisson(\exp\{\mu_{k,mt_i} + \gamma_kY_{K,ij}\}), & W_{ij} = \text{Male} \end{array}\right..
+\label{eq:disc_corr}
+\end{align}
 
-For the discrete phenotypes, we will assume \begin{align}
-{Y}_{l,ti} & \sim \left\{\begin{array}{ll} Poisson(\lambda_{l,ft}), & W_{ti} = \text{Female} \\ Poisson(\lambda_{l,mt}), & W_{ti} = \text{Male} \end{array}\right.,
-\label{eq:Y}
-\end{align} for \(l = 1,\cdots,L\). Similar to the continuous
-phenotypes, we set \begin{align}
-\log(\lambda_{l,gt}) = \gamma_{l,g} + v_{l,gt}.
-\label{eq:lambda}
-\end{align} We use the log function to allow \(v_{l,gt}\) to take all
-values on the real number line so we can properly fit an OU process to
-these effects. We further assume \begin{align}
-v_{l,gt} \overset{iid}{\sim}\mathcal{N}(\phi_{l} v_{l,g(t-1)} , \omega_l^2),
-\label{eq:v_ar1}
-\end{align} and \begin{align}
-v_{l,g1} \overset{iid}{\sim}\mathcal{N}\left(0,\frac{\omega_l^2}{1 - \phi_{l}^2}\right)..
-\label{eq:v1}
+For the other discrete traits, we also assume the above model except we
+set \(\gamma_k = 0\) because of the assumption of no correlation between
+the standard length and these traits. More specifically, we assume
+\begin{align}
+{Y}_{k,ij} & \sim \left\{\begin{array}{ll} Poisson(\exp\{\mu_{k,ft_i}\}), & W_{ij} = \text{Female} \\ Poisson(\exp\{\mu_{k,mt_i}\}), & W_{ij} = \text{Male} \end{array}\right..
+\label{eq:disc_ind}
 \end{align}
+
+In the above model descriptions, \(\mu_{k,ft_i}\) and \(\mu_{k,mt_i}\)
+model the time-\(t_i\) specific mean of phenotype \(k\) for female and
+male stickleback fish, respectively. We point out that, for the discrete
+traits, the means are represented by \(\exp\{\mu_{k,ft_i}\}\) and
+\(\exp\{\mu_{k,mt_i}\}\) for ease of use in our modeling technique. We
+further set \begin{align}
+\mu_{k,gt_i} = \beta_{0,kg} + \beta_{1,kg}t_i + u_{k,gt_i},
+\label{eq:mu}
+\end{align} for \(g \in \{f,m\}\) where \(\beta_{0,kg}\) and
+\(\beta_{1,kg}\) are regression parameters of phenotype \(Y_k\) for each
+gender, accounting for the possibility of a time-dependent trend in the
+mean structure, and \(u_{k,gt_i}\) is the corresponding residual. To
+account for potential correlations between the residuals for a given
+trait \(k\) and gender \(g\), we fit an Ornstein-Uhlenback (OU) process
+(Uhlenbeck and Ornstein (\protect\hyperlink{ref-OUProcess}{1930})). More
+specifically, define \(du_{k,gt} = u_{k,g(t + dt)} - u_{k,gt}\), the
+change in \(u_k,gt\) for a given trait \(k\) and gender \(g\) over a
+miniscule time period \(dt\). The OU process is defined as \begin{align}
+du_{k,gt} = -\kappa_k u_{k,gt} dt + \tau_k dW_t,
+\label{eq:OU_cont}
+\end{align} where \(\kappa_k\) is a parameter associated with the
+correlation between \(u_{k,gt}\) and \(u_{k,g(t+dt)}\), \(\tau_k\) is
+the standard deviation of the OU process, and \(W_t\) is a standard
+Brownian motion. As shown in (Uhlenbeck and Ornstein
+(\protect\hyperlink{ref-OUProcess}{1930})), the closed form solution for
+the SDE in (\ref{eq:OU_cont}) is \begin{align}
+u_{k,gt_i} \overset{iid}{\sim}\mathcal{N}\left(u_{k,gt_{i-1}}\exp\{-\kappa_k(t_{i} - t_{i-1})\} , \frac{\tau_k^2(1 - \exp\{-2\kappa_k(t_{i} - t_{i-1}))\}}{2\kappa_k}\right)
+\label{eq:OU_sol}
+\end{align} for \(i = 2,\cdots,T\). In a traditional OU process, the
+initial value \(u_{k,gt_1}\) is assumed to be a (potentially unknown)
+constant, and we discuss the procedure for estimating this value below.
+
+In our empirical study, we analyze the following four nested models for
+the mean process outlined in (\ref{eq:mu}):
+
+\begin{itemize}
+\item OU with Trend: The mean process for each phenotype $k=1,\cdots,K$ and $g \in \{f,m\}$ as previously defined.
+\item OU with No Trend: No linear trend on the mean processes and the overall mean of the phenotype is constant for each gender.
+This is achieved by setting $\beta_{1,kg} = 0$ in (\ref{eq:mu}) for $k = 1,\cdots,K$ and $g \in \{f,m\}$
+\item No OU with Trend: No random fluctuations on the mean processes, i.e.
+the OU process assumption is not needed.
+This is achieved by setting $u_{k,gt_i} = 0$ in (\ref{eq:mu}) for $k = 1,\cdots,K$, $g \in \{f,m\}$, and $i = 1,\cdots,T$.
+\item No OU with No Trend: The mean for a specific trait and specific gender at each time point is a constant value.
+This is achieved by setting $\beta_{1,kg} = 0$ and $u_{k,gt_i} = 0$ in (\ref{eq:mu}) for $k = 1,\cdots,K$, $g \in \{f,m\}$, and $i = 1,\cdots,T$.
+\end{itemize}
 
@@
 
-Priors: For \(k = 1,\cdots,K\) and \(l = 1,\cdots,L\),
+Priors: For \(k = 1,\cdots,K\),
 
 \begin{align}
-\sigma_k & \overset{iid}{\sim}\mathcal{N}(0,100)I_{\{\sigma > 0\}} \nonumber \\
-\tau_k & \overset{iid}{\sim}\mathcal{N}(0,100)I_{\{\tau > 0\}} \nonumber \\
-\kappa_k & \overset{iid}{\sim}\mathcal{N}(0,1)I_{\{-1 < \kappa_g < 1\}} \nonumber \\
-\beta_k & \overset{iid}{\sim}\mathcal{N}(0,5) \\
-\theta_{k,g} & \overset{iid}{\sim}\mathcal{N}(0,10000)
-\omega_l & \overset{iid}{\sim}\mathcal{N}(0,100)I_{\{\tau > 0\}} \nonumber \\
-\phi_l & \overset{iid}{\sim}\mathcal{N}(0,1)I_{\{-1 < \kappa_g < 1\}} \nonumber \\
-\gamma_{l,g} & \overset{iid}{\sim}\mathcal{N}(0,10000)
+u_{k,gt_1} & \overset{iid}{\sim}\mathcal{N}(0,\tau_{0,k}) \nonumber \\
+\sigma_k & \overset{iid}{\sim}\mathcal{N}(0,10)I_{\{\sigma > 0\}} \nonumber \\
+\tau_k & \overset{iid}{\sim}\mathcal{N}(0,10)I_{\{\tau > 0\}} \nonumber \\
+\tau_{0,k} & \overset{iid}{\sim}\mathcal{N}(0,20)I_{\{\tau > 0\}} \nonumber \\
+\kappa_k & \overset{iid}{\sim}\mathcal{N}(0,1)I_{\{\kappa_g > 0\}} \nonumber \\
+\gamma_{k} & \overset{iid}{\sim}\mathcal{N}(0,5) \nonumber \\
+\beta_{0,kg} & \overset{iid}{\sim}\mathcal{N}(0,100) \nonumber \\
+\beta_{1,kg} & \overset{iid}{\sim}\mathcal{N}(0,3) \nonumber \\
 \label{eq:priors}
-\end{align}
+\end{align} We also note that, for the discrete phenotypes in equations
+(\ref{eq:disc_corr}) and (\ref{eq:disc_ind}), there is no \(\sigma_k\),
+and thus is not sampled for those phenotypes.
 
-All models were built using R Core Team
-(\protect\hyperlink{ref-R2022language}{2022})
+All models were built using Team
+(\protect\hyperlink{ref-R2022language}{2022}) and (\textbf STAN)
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-BellEtAl1993}{}}%
+Bell, Orti, M. A., and J. P. Koenings. 1993. {``Evolution of Pelvic
+Reduction in Threespine Sticklebacks: A Test of Competing Hypotheses.''}
+\emph{Evolution} 47: 906--14.
+
 \leavevmode\vadjust pre{\hypertarget{ref-Blain2022}{}}%
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Cornault2022}{}}%
-Cornuault, Josselin. 2022. {``{Bayesian Analyses of Comparative Data
-with the Ornstein--Uhlenbeck Model: Potential Pitfalls}.''}
-\emph{Systematic Biology} 71 (6): 1524--40.
-\url{https://doi.org/10.1093/sysbio/syac036}.
+Cornuault, Josselin. 2022. {``Bayesian Analyses of Comparative Data with
+the Ornstein--Uhlenbeck Model: Potential Pitfalls.''} \emph{Systematic
+Biology} 71 (6): 1524--40. \url{https://doi.org/10.1093/sysbio/syac036}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-deLisleNco2018}{}}%
+De Lisle, S. Paiva, S. P., and L. Rowe. 2018. {``Habitat Partitioning
+During Character Displacement Between the Sexes.''} \emph{Biology
+Letters} 14: 20180124.
+
+\leavevmode\vadjust pre{\hypertarget{ref-deLisleRowe2015}{}}%
+De Lisle, S. P., and L. Rowe. 2015. {``Ecological Character Displacement
+Between the Sexes.''} \emph{The American Naturalist} 186: 693--707.
 
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-R2022language}{}}%
-R Core Team. 2022. \emph{R: A Language and Environment for Statistical
-Computing}. Vienna, Austria: R Foundation for Statistical Computing.
-\url{https://www.R-project.org/}.
+\leavevmode\vadjust pre{\hypertarget{ref-Reimchen2016}{}}%
+Reimchen, D. Steeves, T. E. 2016. {``Sex Matters for Defence and Trophic
+Traits of Threespine Stickleback.''} \emph{Evolutionary Ecology
+Research} 17: 459-485-.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Reimchen1980}{}}%
+Reimchen, T. E. 1980. {``Spine Deficiency and Polymorphism in a
+Poulation of Gasterosteus Aculeatus: An Adaptation to Predators?''}
+\emph{Canadian Journal of Zoology} 58: 1232--44.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Reimchen1994}{}}%
+---------. 1994. {``Predators and Morphological Evolution in Threespine
+Stickleback.''} In \emph{The Evolutionary Biology of the Threespine
+Stickleback}, edited by M A Bell and S A Foster. Oxford, UK: Oxford
+University Press.
+
+\leavevmode\vadjust pre{\hypertarget{ref-ReimchenNosil2004}{}}%
+Reimchen, T. E., and P. Nosil. 2004. {``Variable Predation Regimes
+Predict the Evolution of Sexual Dimorphism in a Population of Threespine
+Stickleback.''} \emph{Evolution} 58 (6): 1274--81.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Roestietal2023}{}}%
+Roesti, J. S. Groh, M, and D. Schluter. 2023. {``Species Divergence
+Under Competition and Shared Predation.''} \emph{Ecology Letters} 26:
+111--23.
 
@@
 Bonhomme, Michael J Benton, Innes C Cuthill, and Peter J Makovicky.
-2020. {``{An effect size statistical framework for investigating sexual
-dimorphism in non-avian dinosaurs and other extinct taxa}.''}
+2020. {``An Effect Size Statistical Framework for Investigating Sexual
+Dimorphism in Non-Avian Dinosaurs and Other Extinct Taxa.''}
 \emph{Biological Journal of the Linnean Society} 131 (2): 231--73.
@@
 419--61.
+
+\leavevmode\vadjust pre{\hypertarget{ref-SpoljaricReimchen2008}{}}%
+Spoljaric, M. A., and T. E. Reimchen. 2008. {``Habitat-Dependent
+Reduction of Sexual Dimorphism in Geometric Body Shape of Haida Gwaii
+Threespine Stickleback.''} \emph{Biological Journal of the Linnean
+Society} 95: 505--16.
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-R2022language}{}}%
+Team, R Core. 2022. \emph{R: A Language and Environment for Statistical
+Computing}. Vienna, Austria: R Foundation for Statistical Computing.
+\url{https://www.R-project.org/}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-OUProcess}{}}%
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-Schoener1968}{}}%
+W., Schoener T. 1968. {``The Anolis Lizards of Bimini: Resource
+Partitioning in a Complex Fauna.''} \emph{Ecology} 49: 704--26.
+
 \leavevmode\vadjust pre{\hypertarget{ref-ZhouReiter2009}{}}%

### 2023-08-08 R/stuart_stickle_bayes_check.R
Add DIC and posterior predictive files

@@ -4,19 +4,19 @@ library(rstan)
 library(ggridges)
 
 DIC_cont_ind <- function(y,X,mu,sigma){
-  dnorm(y,as.vector(X %*% mu),sigma,log=TRUE) %>% sum
+  (dnorm(y,as.vector(X %*% mu),sigma,log=TRUE) %>% sum)*-2
 }
 
 DIC_cont_dep <- function(y,X,mu,stl,gamma,sigma){
-  dnorm(y,as.vector(X %*% mu) + gamma*stl,sigma,log=TRUE) %>% sum
+  (dnorm(y,as.vector(X %*% mu) + gamma*stl,sigma,log=TRUE) %>% sum)*-2
 }
 
 DIC_disc_ind <- function(y,X,mu){
-  dpois(y,exp(as.vector(X %*% mu)),log=TRUE) %>% sum
+  (dpois(y,exp(as.vector(X %*% mu)),log=TRUE) %>% sum)*-2
 }
 
 DIC_disc_dep <- function(y,X,mu,stl,gamma){
-  dpois(y,exp(as.vector(X %*% mu) + gamma*stl),log=TRUE) %>% sum
+  (dpois(y,exp(as.vector(X %*% mu) + gamma*stl),log=TRUE) %>% sum)*-2
 }
 
 dat <- read.csv("Data/updated_stickle_imputations_100.csv")
@@ -32,17 +32,15 @@ vars <- c("stl","lps","ect","tpg",
           "ds3","lpt","mdf","mav",
           "maf","mcv","mds","mpt")
 stl <- matrix(dat$stl,ncol=M)
+N <- nrow(stl)
 
 DICs <- NULL
-
-apply(samps,)
 
 for (var in vars){
   y <- matrix(dat[,var],ncol=M)
   for (mod in 0:3){
+    probs <- NULL
     samps <- readRDS(paste0("Figures/",var,"/posterior_samples_",var,"_",mod,".RDS"))
-
-    N <- nrow(samps)
 
     mu <- samps %>% select(contains("mu"))
     if (substr(var,1,1) != "m"){
@@ -50,58 +48,83 @@ for (var in vars){
     }
     gamma <- samps %>% select(gamma)
 
-    if (substr(var,1,1) != "m"){
-      params <- cbind(mu,gamma,sigma)
-    } else {
-      params <- cbind(mu,gamma)
-    }
+    print(paste0("Calculating DIC for ",var," for model ",mod))
 
-    apply(params,1,function(pars){
-      sapply(1:M,function(m){
[…]

### 2023-08-08 manuscript/paper.Rmd
Update Paper

@@
+<<<<<<< HEAD
 ---
@@
 # References
+=======
+---
+title: |
+  |  \Large Increased sexual dimorphism evolves in a fossil stickleback following ecological release from fish piscivores
+author: | 
+  | \large Allison Ozark$^{1,*}$, Matthew Stuart$^{2,3,*}$, Raheyma Siddiqui$^1$, Akhil Ghosh$^2$
+  | \large Samantha Swank$^1$, Michael A. Bell$^4$, Gregory J. Matthews$^{2,3}$, and Yoel E. Stuart$^{1,+}$ 
+  | \vspace{-1.1mm}
+  | \large $^1$ Department of Biology, Loyola University Chicago, Chicago, IL, USA \vspace{-1.1mm}
+  | \large $^2$ Department of Mathematics and Statistics, Loyola University Chicago, Chicago, IL, USA \vspace{-1.1mm}
+  | \large $^3$ Center for Data Science and Consulting, Loyola University Chicago, Chicago, IL, USA \vspace{-1.1mm}
+  | \large $^4$ University of California Museum of Paleontology, Berkeley, CA, USA \vspace{-1.1mm}
+  | \large $^*$ Equal contribution \vspace{-1.1mm}
+  | \large $^+$ Corresponding: ystuart@luc.edu \vspace{-1.1mm}
+abstract: |
+  | Everyone loves the stickle \vspace{2mm}
+  | *Keywords*: Stickle
+bibliography: references.bib
+fontsize: 12pt
+link-citations: true
+linkcolor: cyan
+urlcolor: cyan
+output:
+  pdf_document:
+    df_print: kable
+    number_sections: true
+    keep_tex: true
+header-includes:
+ \usepackage{setspace}
+ \setstretch{1.15}
+ \usepackage{float}
+ \floatplacement{figure}{t}
+---
+
+```{r setup, echo=FALSE, message=FALSE}
+knitr::opts_chunk$set(
+	echo = FALSE,
+	fig.align = "center",
+	message = FALSE,
+	warning = FALSE
+)
+
+gjm <- function(x, color = "red") {
+  sprintf("\\textcolor{%s}{%s}", color, x)}
+```
+
+\newcommand{\iid}{\overset{iid}{\sim}}
+
+```{r pkgs}
+library(tidyverse)
+theme_set(theme_minimal())
+library(Lahman)
+```
+
+\newpage
+
+# Introduction {#sec:intro}
+
+Ecological release theory suggests that a prey population’s niche should expand when predation is relaxed or removed, as prey take advantage of new enemy free space (reviewed in @Herrmann2021).
+Niche expansion could manifest through a combination of increasing within- and between-individual niche widths (@Bolnick2010; @Herrmann2021), including divergence between the sexes (@Bolnick2003; @Cooper2011).
+Theory and empirical data suggest that a population presented with ecological opportunity following ecological release might experience disruptive selection on males and females stemming from intraspecific competition over newly accessible resources, resulting in intersexual divergence in habitat use (@Schoener1968; @Shine1989; @Bolnick2003; @Butler2007; @Bolnick2008; @Cooper2011; but see @Stuart2021; @Blain2022).
+Accordingly, sexes should also diverge in morphological traits associated with habitat use; such ecologically-driven "character displacement between the sexes" is therefore one explanation for phenotypic sexual dimorphism (@deLisleRowe2015; @deLisleNco2018).
+
+We tested this link between release from predators and the evolution of sexual dimorphism in a well-preserved, finely-resolved, ~16,000 year-long sequence of the fossil stickleback fish (<i>Gasterosteus doryssus</i>).
+The fossil lineage appeared in the depositional environment as a fully armored form, with complete pelvic girdles, two pelvic spines, and three dorsal spines, on average (@Bell2006; @Stuart2020).
+However, this habitat appears to have been missing fish piscivores (e.g., trout and other salmonids known to prey on modern stickleback; @Bell2009), relaxing putative selection for armor (@Reimchen1980, @Reimchen1994, @BellEtAl1993; @Bowne1994; @Roestietal2023).
+The lineage adaptively evolved armor loss but also reduction in the means of several other traits  (@Hunt2008; @Stuart2020; Siddiqui et al.
+in prep).
+In particular, tooth wear data suggest that the population shifted to eat more planktonic prey through time, indicative of increased open water habitat use (Purnell et al.
+2007).
+If this divergence in habitat use is partitioned between sexes (@ReimchenNosil2004; @SpoljaricReimchen2008;@Reimchen2016), then we should expect sexual dimorphism to evolve for traits related to habitat use in <i>G.
+doryssus</i>.
+
+To test this prediction, we imputed the sex of individual fossils, which is unobserved in practice, with Multiple Imputation by Chained Equations (MICE) (@MICE), using observed data from extant stickleback of known sex to train the imputation model.
+Then we tracked the multivariate evolution of sexual dimorphism over ~16,000 years for 16 traits related to swimming, feeding and defense.
+
+# Methods and Materials 
+<!-- Studying sexual dimorphism of phenotypes in modern species is a straightforward endeavor: measure a phenotype of interest and statistically test for differences between the sexes.
+However, examining sexual dimorphism in fossils is complicated substantially by the issue that sex may not be able to be directly observed from fossil specimens.
+For this study, we have two sources of data: 1) Modern stickleback specimens with observed sex and 2) fossil specimens with sex unobserved.
+-->
+
+A major challenge in this work is inferring the sex of fossil stickleback specimens.
+Indeed, for most species, sex cannot be detected directly, except for lineages whose sexes are distinguished by the presence or absence of sex-specific characters (@Hone2017; @SaittaEtAl2020).
+Instead, paleobiologists resort to statistical detection of sex and sexual dimorphism, including tests for bimodality in trait distributions (e.g., @Hone2017) and divergence in growth curves (e.g., @SaittaEtAl2020).
+We used a third approach to infer fossil sex and study the evolution of sexual dimorphism by taking advantage of the fact that G.
+doryssus is part of the extant threespine stickleback species complex, Gasterosteus aculeatus.
+We measured the same trait set from a phenotypically diverse set of extant G.
+aculeatus populations, for whom individual sex was known.
+We then combined fossil and extant data sets, treated fossil sex as missing data, and used the MICE (@MICE) multiple imputation (@little2002statistical) algorithm, and implemented in R (@R2022language), to impute individual sex of the fossil data set 100 times.
+We then fit a modified Ornstein–Uhlenbeck (OU) (@OUProcess) model using a Bayesian framework to the data to test for the evolution of sexual dimorphism in each trait.
+
+# Data {#sec:data}
+```{r}
+library(tidyverse)
+stickle <- read_csv('stickle_revamped.csv')
+
+stick <- stickle %>% 
+  mutate(gender = as.factor(case_when(
+    grepl(".M.",univID,fixed=TRUE) ~ "Male",
+    grepl(".F.",univID,fixed=TRUE) ~ "Female",
+    grepl("Fos",univID,fixed=TRUE) ~ NA_character_
+    )), ips = as.factor(ips)
+  ) %>% 
+  select(gender, everything())
+
+#getting time vector, adding it to imputed data
+univ <- stick %>% 
+  mutate(time = as.integer(sub(".*([0-9]{2})\\..*", "\\1",univID)))
+times <- univ %>% drop_na(time) %>% pull(time)
+
+times_vector <- c(rep(1,43)) %>% 
+  append(21-times)
+
+#fossil n
+#fortab <- data.frame(time = 1:18, count = as.character(as.numeric((table(times_vector)))))
+fortab <- data.frame(time = 1:18, count = as.numeric((table(times_vector))))
+#knitr::kable(fortab)
+#print(xtable(fortab),include.rownames = FALSE)
+```
+
+## Fossil Specimen Data
+We used Gasterosteus doryssus data that were previously reported by @Stuart2020, @Voje2022, and Siddiqui et al.
+(in review).
+Briefly, the data were collected from fossil Series K from Quarry D (Cerasoni et al., in review), dug from an open pit diatomite mine at 9.526° N, 119.094° W, near Fernley, Nevada, USA.
+Series K consisted of 18 samples taken at ~1000-year intervals, and mean sample times span ~16,363 years.
+Fish from series K were measured for 16 ecomorphological traits related to armor, swimming, and feeding (Table 1).
+Series K started at a previously documented horizon when a low armored lineage of stickleback with zero to one dorsal spines, zero pelvic spines, and highly reduced pelvises was completely replaced by a high armored lineage of stickleback with three dorsal spines, two pelvic spines, and complete (Bell et al.
+2006; Bell 2009; Stuart et al.
+2020).
+This lineage subsequently evolved reduction in armor, body size, and traits related to swimming and feeding (Bell et al.
+2006; Stuart et al.
+2020; Siddiqui et al.
+in review).
+The tempo and mode of armor reduction during this sequence suggests adaptive evolution by natural selection (Hunt et al.
+2008), and we focus on the multivariate evolution of sexual dimorphism by this second lineage.
+
+The fossil data consists of `r sum(is.na(stick$gender))` specimens with unknown sex over 18 time periods spaced  ~1000 years apart.
+Figure 1 shows the sample size at each of the 18 time periods.
+There are at least 22 specimens at each time period with a high of 67 specimens in period 7.
+
+```{r fig.cap = "Fossil sampel size for each time period"}
+ggplot(aes(x = time, y = count), data = fortab) + geom_col() + xlab("Time Period")  + scale_x_continuous(breaks = 1:18, labels = 1:18)
+```
+
+| Trait Name           | Trait Code | Trait Description           |
+|----------------------|------------|-----------------------------|
+|Standard Length|	stl|	Distance from anterior tip of premaxilla to posterior end of last vertebra (hypural plate)|
+|Dorsal Spine | mds |	Number of dorsal spines from 0 to 3|
+|Dorsal Fin Ray|mdf|	Number of bones in the dorsal fin posterior to the third dorsal spine (i.e., soft dorsal fin rays)
+|Anal Fin Ray |maf|	Number of bones in the anal fin posterior to the anal spine (i.e., soft dorsal fin rays)|
+|Abdominal Vertebra|mav|	Number of vertebrae anterior to the first vertebra contacting an anal fin pterygiophore (Aguirre et al. 2014)|
+|Caudal Vertebra|mcv	| Number of vertebrae posterior to and including the first vertebra contacting an anal fin pterygiophore (Aguirre et al. 2014)|
+|Pterygiophore number|mpt|	Number of pterygiophores anterior to but excluding the pterygiophore under the third dorsal spine, which is immediately anterior to and contiguous with the dorsal fin|
+|Pelvic Spine length|lps.sc	|Length from the base of one pelvic spine above its articulation with the pelvic girdle to its distal tip|
+|Ectocoracoid|ect.sc|	Length between the anterior and posterior tips of the shoulder girdle base (i.e., ectocoracoid)|
+|Pelvic Girdle|tpg.sc|	Length between the anterior to posterior tips along midline. If vestigial, the sum of longest anterior-posterior axis for the vestiges|
+|Cleithrum length|cle.sc|	Length from free dorsal tip to ventral tip of the cleithrum on the anterior margin of the shoulder girdle (i.e., cleithrum)|
+|Premaxilla|pmx.sc|	Length from the anterior tip of the premaxilla to the distal tip of the ascending process of the premaxilla |
+|Dorsal Spine|Ds#.sc# = 1,2,or 3|	Length from the base of a dorsal spine above the pterygiophore to its distal tip along the anterior edge|
+|Pterygiophore|lpt.sc|	Distance between the anterior to posterior tips of the pterygiophore immediately preceding the 3rd dorsal spine (when present)|
+
+Table: Traits and trait descriptions. ‘sc’ denotes size correction of trait against standard length. Names of bones follow @Bowne1994 unless otherwise noted.
+
+<!-- I removed the table and replaced it with a figure -->
+<!-- % latex table generated in R 4.2.2 by xtable 1.8-4 package -->
+<!-- % Mon May  1 11:20:28 2023 -->
+<!-- \begin{table}[ht] -->
+<!-- \centering -->
+<!-- \caption{The number of stickleback fossils at each time point.  Sample size at each time point ranges from a low of 22 to a high of 67.} -->
+<!-- \begin{tabular}{rl} -->
+<!--   \hline -->
+<!-- time & count \\  -->
+<!--   \hline -->
+<!--   1 & 43 \\  -->
+<!--     2 & 41 \\  -->
+<!--     3 & 51 \\  -->
+<!--     4 & 41 \\  -->
+<!--     5 & 46 \\  -->
+<!--     6 & 48 \\  -->
+<!--     7 & 67 \\  -->
+<!--     8 & 55 \\  -->
+<!--     9 & 42 \\  -->
+<!--    10 & 33 \\  -->
+<!--    11 & 37 \\  -->
+<!--    12 & 22 \\  -->
+<!--    13 & 41 \\  -->
+<!--    14 & 43 \\  -->
+<!--    15 & 46 \\  -->
+<!--    16 & 47 \\  -->
+<!--    17 & 56 \\  -->
+<!--    18 & 55 \\  -->
+<!--    \hline -->
+<!-- \end{tabular} -->
+<!-- \label{tabn} -->
+
+<!-- \end{table} -->
+
+## Extant Specimen data
+To span the gamut of stickleback diversity for our predictive model, we sampled modern stickleback from lakes containing generalist stickleback populations (@Hendry2009; @Bolnick2011) and from lakes containing benthic-limnetic species pairs (Baumgartner et al.
+1988; @Schluter1992) (Table 2 (What is this?)).
+The generalist populations were collected by YES in 2013 and previously described in @Stuart2017.
+These samples were fixed in formalin, then stained for bone with Alizarin Red in 2013.
+Benthic and limnetic specimens were kindly loaned by D.
+Schluter and his lab at University of British Columbia.
+They collected benthic and limnetic individuals from Enos Lake in 1988 and from Emily Lake, Little Quarry Lake, Paxton Lake, and Priest Lake in 2018.
+The Enos specimens had been fixed whole in formalin and stored in 40% isopropanol.
+The specimens from the other lakes were initially preserved whole in 95% ethanol in the field before being gradually transferred to water then formalin in the lab and ultimately stored in 40% isopropanol.
+In 2019, we stained these specimens for bone using Alizarin Red.
+
+We next replicated fossil data collection (Table 1) on these extant specimens.
+Standard length as well as pelvic-spine length on each side were measured with calipers.
+We used a dissection microscope to count dorsal spines, pelvic spines, dorsal-fin rays, and anal-fin rays.
+Right and left-side pelvic girdle lengths and ectocoracoid lengths were measured from ventral photographs taken using a Canon EOS Rebel T7 with a Tamron 16-300 mm MACRO lens mounted on a leveled Kaiser RS1 copy stand.
+Specimens were held in place for ventral photographs using a small tabletop vise with an attached scale bar.
+Lateral X-rays were used to measure dorsal spine length, number of pterygiophores anterior to the pterygiophore holding the third spine, length of the pterygiophore just anterior to the third spine, cleithrum length, and pre-maxilla ascending branch length.
+We also counted vertebrae from the X-rays: abdominal vertebrae were counted anterior to the first vertebra with a haemal spine contacting an anal fin pterygiophore.
+Caudal vertebrae were posterior, including the first vertebra with the haemal spine contacting the anal fin pterygiophore (following @Aguirre2014.
+X-rays were taken with an AXR Hot Shot X-ray Machine (Associated X-ray Corporation) at the Field Museum of Natural History.
+Specimens were exposed at 35kV and 4mA.
+Small fish were exposed for 7s, medium fish for 8s, and large fish for 10s.
+We developed the film and scanned individual images of each fish using the B&W Negatives setting on an Epson Perfection 4990 Photo flatbed at 2400 dpi.
+Measurements from photographs and X-rays were taken with FIJI (Schindelin et al.
+2012) and its plugin ObjectJ (https://sils.fnwi.uva.nl/bcb/objectj/).
+All photographs, X-rays, and ObjectJ files have been uploaded to Morphosource.org (accession # TBD).
+We dissected individuals from the generalist populations to determine sex from the gonads.
+Individuals from the species-pair lakes were sexed by Schluter and his lab, using a genotyping protocol (confirm and cite).
+
+The extant data used here consists of a total of `r sum(!is.na(stick$gender))` specimens all with known sex.
+Of these, there are `r sum(stick$gender == "Female",na.rm = TRUE)` and `r sum(stick$gender == "Male",na.rm = TRUE)` female and male specimens, respectively.
+
+## Outlier analysis. 
+To check for outliers, we calculated within-group means and standard deviations for each trait separately for K series fossil specimens (pooled across samples) and for extant specimens (pooled across lakes).
+We noted trait values greater than 3.5 standard deviations from the mean as potential outliers.
+We checked whether these potential outliers were a result of data entry and collection error and corrected them if they were.
+We turned the remaining outlier trait values to NAs.
+(confirm)  (Wait, what?
+Why?
+Anything that was 3.5 SD above the mean was recorded as missing?
+How do you justify this?
+)
+
+Missing data imputation, including fossil sex
+
+Quantification and evolution of sexual dimorphism
+
+What covariates do we have in the data: length, what else,
+
+# Models  {#sec:models}
+## Imputation Model 
+Let $\boldsymbol{W}$ be an $(n_{extant} + n_{fossil}) \times 1$ vector of the covariate gender of the stickleback fish and $\boldsymbol{Y}$ be an $(n_{extant} + n_{fossil}) \times K$ matrix of the $K$ phenotypes of interest.
+Because the gender of the fossilized stickleback fish is unobservable, we further define $\boldsymbol{W} = (\boldsymbol{W}_{extant}^T,\boldsymbol{W}_{fossil}^T)^T$ where $\boldsymbol{W}_{extant}$ and $\boldsymbol{W}_{fossil}$ are the $n_{extant} \times 1$ and $n_{fossil} \times 1$ vectors of the observed extant gender and missing fossil gender, respectively.
+
+We impute the missing gender for the fossil data by sampling from the posterior predictive distribution $P(\boldsymbol{W}_{fossil}|\boldsymbol{W}_{extant},\boldsymbol{Y})$ using the multiple imputation by chained equations (MICE) algorithm (@MICE) with predictive mean matching.
+Traditionally, the choice for the number of completed dats sets is a relatively small number such as $M = 5$ or $M = 10$.
+However, @ZhouReiter2009 recommend a larger number of imputed data sets if the data users intend on performing Bayesians analysis after imputation, which in this case, we do.
+Therefore, the imputation algorithm is run to obtain a total of $M = 100$ completed datasets.
+In addition to this, @ZhouReiter2009 suggests rather than using Rubin's combining rules to combine across imputed data sets, instead pool all of the draws from the posterior distributions across all of the imputed data sets to estimate the posterior distributions of parameters of interest.
+We proceed with our Bayesian analysis in this manner.
+
+\textcolor{red}{Akhil's stuff about validating the imputation model goes here.}
+
+<!-- Bayesian Analysis after multiple imputation @ZhouReiter2009: They recommend using a large number of imputations.  5 or 10 is too small.  We are using M = 100. -->
+
+## Completed Data Model 
+\textcolor{red}{We should note somewhere that we are only using the $W_{fossil}$ in the modeling part.
+We drop the $W_{extant}$.
+So just note that $W_{ti}$ is really $W_{fossil,ti}$.
+Not sure how to say this, but we need to make it clear that we are only using the fossil data for the OU modeling.
+Greg, check and see if this makes sense.}
+
+For a given imputed dataset, let $W_{ij}$ be the imputed gender and $\boldsymbol{Y}_{ij}$ be the $K \times 1$ vector of phenotypes for stickleback fossil $j$ at time $t_i$ where $i = 1, \ldots, T$ and $j = 1,\ldots,n_{t}$.
+Note, we are inputting only the fossil data into our model; connecting to the previous section, $W_{ij}$ and $\boldsymbol{Y}_{ij}$ can be interpretted as $W_{fossil,ij}$ and $\boldsymbol{Y}_{fossil,ij}$.
+In addition, we denote $Y_{K,ij}$, the last variable in $\boldsymbol{Y}_{ij}$, to be the standard length of the fish,
+\begin{align}
+{Y}_{K,ij} & \iid \left\{\begin{array}{lll} \mathcal{N}(\mu_{K,ft_i},&\sigma_{K}^2), & W_{ij} = \text{Female} \\ \mathcal{N}(\mu_{K,mt_i},&\sigma_{K}^2), & W_{ij} = \text{Male} \end{array}\right..
+\label{eq:stl}
+\end{align}
+
+It is reasonable to assume that the other continuous traits of stickleback fish will have some correlation with its standard length \textbf{(CITATION)}.
+We account for this by adding an additional parameter, $\gamma_k$, into our model.
+More specifically, if $Y_{k,ij}$ is a continuous trait, then
+\begin{align}
+{Y}_{k,ij} & \iid \left\{\begin{array}{llll} \mathcal{N}(\mu_{k,ft_i} + \gamma_kY_{K,ij},\sigma_k^2), & W_{ij} = \text{Female} \\ \mathcal{N}(\mu_{k,mt_i} + \gamma_kY_{K,ij},\sigma_k^2), & W_{ij} = \text{Male} \end{array}\right..
+\label{eq:cont}
+\end{align}
+
+In addition, we assume that the discrete traits abdominal vertebrae (mav) and caudal vertebrae (mcv) also have correlation with the standard length of the fish (\textbf{CITATION}).
+If $Y_{k,ij}$ is one of the above traits, then
+\begin{align}
+{Y}_{k,ij} & \sim \left\{\begin{array}{ll} Poisson(\exp\{\mu_{k,ft_i} + \gamma_kY_{K,ij}\}), & W_{ij} = \text{Female} \\ Poisson(\exp\{\mu_{k,mt_i} + \gamma_kY_{K,ij}\}), & W_{ij} = \text{Male} \end{array}\right..
+\label{eq:disc_corr}
+\end{align}
+
+For the other discrete traits, we also assume the above model except we set $\gamma_k = 0$ because of the assumption of no correlation between the standard length and these traits.
+More specifically, we assume
+\begin{align}
+{Y}_{k,ij} & \sim \left\{\begin{array}{ll} Poisson(\exp\{\mu_{k,ft_i}\}), & W_{ij} = \text{Female} \\ Poisson(\exp\{\mu_{k,mt_i}\}), & W_{ij} = \text{Male} \end{array}\right..
+\label{eq:disc_ind}
+\end{align}
+
+In the above model descriptions, $\mu_{k,ft_i}$ and $\mu_{k,mt_i}$ model the time-$t_i$ specific mean of phenotype $k$ for female and male stickleback fish, respectively.
+We point out that, for the discrete traits, the means are represented by $\exp\{\mu_{k,ft_i}\}$ and $\exp\{\mu_{k,mt_i}\}$ for ease of use in our modeling technique.
+We further set
+\begin{align}
+\mu_{k,gt_i} = \beta_{0,kg} + \beta_{1,kg}t_i + u_{k,gt_i},
+\label{eq:mu}
+\end{align}
+for $g \in \{f,m\}$ where $\beta_{0,kg}$ and $\beta_{1,kg}$ are regression parameters of phenotype $Y_k$ for each gender, accounting for the possibility of a time-dependent trend in the mean structure, and $u_{k,gt_i}$ is the corresponding residual.
+To account for potential correlations between the residuals for a given trait $k$ and gender $g$, we fit an Ornstein-Uhlenback (OU) process (@OUProcess).
+More specifically, define $du_{k,gt} = u_{k,g(t + dt)} - u_{k,gt}$, the change in $u_k,gt$ for a given trait $k$ and gender $g$ over a miniscule time period $dt$.
+The OU process is defined as
+\begin{align}
+du_{k,gt} = -\kappa_k u_{k,gt} dt + \tau_k dW_t,
+\label{eq:OU_cont}
+\end{align}
+where $\kappa_k$ is a parameter associated with the correlation between $u_{k,gt}$ and $u_{k,g(t+dt)}$, $\tau_k$ is the standard deviation of the OU process, and $W_t$ is a standard Brownian motion.
+As shown in (@OUProcess), the closed form solution for the SDE in (\ref{eq:OU_cont}) is
+\begin{align}
+u_{k,gt_i} \iid \mathcal{N}\left(u_{k,gt_{i-1}}\exp\{-\kappa_k(t_{i} - t_{i-1})\} , \frac{\tau_k^2(1 - \exp\{-2\kappa_k(t_{i} - t_{i-1}))\}}{2\kappa_k}\right)
+\label{eq:OU_sol}
+\end{align}
+for $i = 2,\cdots,T$. In a traditional OU process, the initial value $u_{k,gt_1}$ is assumed to be a (potentially unknown) constant, and we discuss the procedure for estimating this value below.
+
+In our empirical study, we analyze the following four nested models for the mean process outlined in (\ref{eq:mu}):
+\begin{itemize}
+\item OU with Trend: The mean process for each phenotype $k=1,\cdots,K$ and $g \in \{f,m\}$ as previously defined.
+\item OU with No Trend: No linear trend on the mean processes and the overall mean of the phenotype is constant for each gender.
+This is achieved by setting $\beta_{1,kg} = 0$ in (\ref{eq:mu}) for $k = 1,\cdots,K$ and $g \in \{f,m\}$
+\item No OU with Trend: No random fluctuations on the mean processes, i.e.
+the OU process assumption is not needed.
+This is achieved by setting $u_{k,gt_i} = 0$ in (\ref{eq:mu}) for $k = 1,\cdots,K$, $g \in \{f,m\}$, and $i = 1,\cdots,T$.
+\item No OU with No Trend: The mean for a specific trait and specific gender at each time point is a constant value.
+This is achieved by setting $\beta_{1,kg} = 0$ and $u_{k,gt_i} = 0$ in (\ref{eq:mu}) for $k = 1,\cdots,K$, $g \in \{f,m\}$, and $i = 1,\cdots,T$.
+\end{itemize}
+
+Because we are fitting a dataset with a stochastic structure on the means of the phenotypes, we analyze the data via a Bayesian analysis.
+Bayesian data analysis is also more naturally used when we have to impute data (\textbf{CITATION}).
+
+Priors: For $k = 1,\cdots,K$,
+
+\begin{align}
+u_{k,gt_1} & \iid \mathcal{N}(0,\tau_{0,k}) \nonumber \\
+\sigma_k & \iid \mathcal{N}(0,10)I_{\{\sigma > 0\}} \nonumber \\
+\tau_k & \iid \mathcal{N}(0,10)I_{\{\tau > 0\}} \nonumber \\
+\tau_{0,k} & \iid \mathcal{N}(0,20)I_{\{\tau > 0\}} \nonumber \\
+\kappa_k & \iid \mathcal{N}(0,1)I_{\{\kappa_g > 0\}} \nonumber \\
+\gamma_{k} & \iid \mathcal{N}(0,5) \nonumber \\
+\beta_{0,kg} & \iid \mathcal{N}(0,100) \nonumber \\
+\beta_{1,kg} & \iid \mathcal{N}(0,3) \nonumber \\
+\label{eq:priors}
+\end{align}
+We also note that, for the discrete phenotypes in equations (\ref{eq:disc_corr}) and (\ref{eq:disc_ind}), there is no $\sigma_k$, and thus is not sampled for those phenotypes.
+
+All models were built using @R2022language and (\textbf STAN)
+
+@Cornault2022 Bayesian OU model.
+
+Bayesian Analysis after multiple imputation @ZhouReiter2009: They recommend using a large number of imputations.  5 or 10 is too small.  We are using M = 100.
+
+# Results {#sec:results}
+
+# Discsusson, Future work and conclusions {#sec:conclusions}
+We predicted that release from predators would result in niche expansion and increased sexual dimorphism, based on several studies of modern stickleback.
+For example, in lakes where sculpin competitors are absent and stickleback (Roesti et al.
+2023)
+See Spoljaric and Reimchen 2008, page 512 right column for references and discussion of differences between benthic males and limnetic females.
+Male stickleback are benthic and littoral (Wootton 1976)….
+Reimchen papers in general good for this section.
+
+# Acknowledgements {-}
+
+We thank O.
+Abughoush, S.
+Blaine, A.
+Chaudhary, M.
+Islam, F.
+Joaquin, C.
+Lawson-Weinert, R.
+Sullivan, J.
+Tien, M.P.
+Travis, and W.
+Shim for help with data collection.
+We thank D.
+Schluter and S.
+Blain for loaning specimens and sharing data.
+We thank K.
+Swagel and C.
+McMahan of the Field Museum for assistance with specimen x-rays.
+This research was supported by NSF grants BSR-8111013, EAR-9870337, and DEB-0322818, the Center for Field Research (Earthwatch), and the National Geographic Society (2869-84) to MAB.
+It was also supported by NSF grants DEB-1456462 and EAR-2145830 to YES.
+And NSF DMS-2015374 (GJM)
+
+# Supplementary Material {-}
+
+All code for reproducing the analyses in this paper is publicly available at https://github.com/Akhil-Ghosh/SticklebackProject
+
+# References
+>>>>>>> 471fb0b0ba3b93e818ccc4987a5a22f419617cec

### 2023-08-11 R/stuart_stickle_bayes.R
Update Checks and add Posterior predictive plots

@@ -2,6 +2,7 @@ rm(list = ls())
 library(tidyverse)
 library(rstan)
 library(ggridges)
+library(bayesplot)
 
 dat <- read.csv("Data/updated_stickle_imputations_100.csv")
 times <- read.csv("RawData_ReadOnly/KSampleMeanTimes.csv")

### 2023-08-11 R/stuart_stickle_bayes_check.R

@@ -2,6 +2,7 @@ rm(list = ls())
 library(tidyverse)
 library(rstan)
 library(ggridges)
+library(bayesplot)
 
 DIC_cont_ind <- function(y,X,mu,sigma){
   (dnorm(y,as.vector(X %*% mu),sigma,log=TRUE) %>% sum)*-2
@@ -69,23 +70,34 @@ for (var in vars){
 }
 
 DICList <- as.data.frame(matrix(DICs,nrow=16,byrow=TRUE))
-rownames(DICList) <- vars
 colnames(DICList) <- c("OU-Trend","No OU-Trend","OU-No Trend","No OU-No Trend")
+DICList$var <- vars
+DICList <- DICList %>% pivot_longer(cols=-var,names_to="model",values_to="DIC")
 
 num_dens <- 100
 
-for (var in vars){
+for (j in 1:length(vars)){
+  var <- vars[j]
   y <- matrix(dat[,var],ncol=M)
-  specific_x_values <- seq(min(density(y)$x),max(density(y)$x),length.out=100)
-
-  dens <- rep(0,100)
-  for (m in 1:M){
-    kde <- density(y[,m])
-    dens <- dens + 1/M*approx(kde$x,kde$y,xout = specific_x_values,method = "linear")$y
+  if (substr(var,1,1) == "m"){
+    specific_x_values <- as.vector(y) %>% unique() %>% sort()
+  } else {
+    specific_x_values <- seq(min(density(y)$x),max(density(y)$x),length.out=100)
   }
-  tmp <- data.frame(x=specific_x_values,y=dens)
+
+  dens <- rep(0,length(specific_x_values))
+  for (m in 1:M){
+    if(substr(var,1,1) == "m"){
+      dens <- dens + 1/M*sapply(specific_x_values,function(x){length(which(y[,1] == x))})/N
+    } else {
+      kde <- density(y[,m])
+      dens <- dens + 1/M*approx(kde$x,kde$y,xout = specific_x_values,method = "linear")$y
+    }
+  }
+  dens <- data.frame(x=specific_x_values,y=dens)
 
   for (mod in 0:3){
+    tmp <- data.frame(x=specific_x_values)
     samps <- readRDS(paste0("Figures/",var,"/posterior_samples_",var,"_",mod,".RDS"))
 
     n_samps <- nrow(samps)
@@ -98,33 +110,87 @@ for (var in vars){
 
     for (i in 1:num_dens){
[…]

### 2023-09-05 manuscript/paper.Rmd
Add description of generalized Poisson to paper

--- 2023-08-08 manuscript/paper.Rmd
@@
 
+If $Y_{k,ij}$ is a discrete trait, the conventional method of modelling this data is by fitting a Poisson distribution.
+However, the Poisson distribution assumes that the mean and variance of $Y_{k,ij}$ are equal ($E(Y_{k,ij}) = \text{Var}(Y_{k,ij})$), while the empirical fossil data may have variances that are smaller than their respective means.
+To combat this issue, we propose to fit the discrete traits to a generalized Poisson model as defined in @GeneralizedPoisson.
+Specifically, if $X \sim GP(\lambda,\alpha)$, then
+\begin{align}
+P(X = x) = \left\{\begin{array}{cc} \frac{(1 - \alpha)\lambda[(1 - \alpha)\lambda + \alpha x]^{x - 1} \exp\left\{-((1 - \alpha)\lambda  + \alpha x)\right\}}{x!} & (1 - \alpha)\lambda  + \alpha x \geq 0  \\ 0 & (1 - \alpha)\lambda  + \alpha x < 0 \end{array}\right.,
+\label{eq:GP_pmf}
+\end{align}
+where $E(Y_{k,ij}) = \lambda$ and $Var(Y_{k,ij}) = \frac{\lambda}{(1 - \alpha)^2}$.
+If $\alpha > 0$, then the variance is greater than the mean, called overdispersion; if $\alpha < 0$, then the variance is greater than the mean, called underdispersion; if $\alpha = 0$, then the model degenerates to a Poisson distribution.
+
+We will use the generalized Poisson to model all of the discrete traits in our dataset.
 In addition, we assume that the discrete traits abdominal vertebrae (mav) and caudal vertebrae (mcv) also have correlation with the standard length of the fish (\textbf{CITATION}).
@@
 \begin{align}
-{Y}_{k,ij} & \sim \left\{\begin{array}{ll} Poisson(\exp\{\mu_{k,ft_i} + \gamma_kY_{K,ij}\}), & W_{ij} = \text{Female} \\ Poisson(\exp\{\mu_{k,mt_i} + \gamma_kY_{K,ij}\}), & W_{ij} = \text{Male} \end{array}\right..
+{Y}_{k,ij} & \sim \left\{\begin{array}{ll} GP(\exp\{\mu_{k,ft_i} + \gamma_kY_{K,ij}\},\alpha_k), & W_{ij} = \text{Female} \\ GP(\exp\{\mu_{k,mt_i} + \gamma_kY_{K,ij}\},\alpha_k), & W_{ij} = \text{Male} \end{array}\right..
 \label{eq:disc_corr}
@@
 \begin{align}
-{Y}_{k,ij} & \sim \left\{\begin{array}{ll} Poisson(\exp\{\mu_{k,ft_i}\}), & W_{ij} = \text{Female} \\ Poisson(\exp\{\mu_{k,mt_i}\}), & W_{ij} = \text{Male} \end{array}\right..
+{Y}_{k,ij} & \sim \left\{\begin{array}{ll} GP(\exp\{\mu_{k,ft_i}\},\alpha_k), & W_{ij} = \text{Female} \\ GP(\exp\{\mu_{k,mt_i}\},\alpha_k), & W_{ij} = \text{Male} \end{array}\right..
 \label{eq:disc_ind}
@@
 Bayesian data analysis is also more naturally used when we have to impute data (\textbf{CITATION}).
+To aid in the sampling procedure for the discrete phenotypes, we declare
 
-Priors: For $k = 1,\cdots,K$,
+Priors: To aid in the sampling procedure for the discrete phenotypes, we declare $\phi_k = \log\left(\frac{\alpha_k - \max_{i,j}(-\lambda_{k,ij}/y_{k,ij})}{1 - \alpha_k}\right)$ where $\lambda_{k,ij} = \left\{\begin{array}{cc} \exp(\mu_{k,ft_i} + \gamma_k Y_{K,ij}) & W_{ij} = \text{Female} \\ \exp(\mu_{k,mt_i} + \gamma_k Y_{K,ij}) & W_{ij} = \text{Male} \end{array}\right.$.
+This transformation is performed to ensure our algorithm can properly sample from the posterior distribution of interest.
+
+For $k = 1,\cdots,K$,
 
@@
 \sigma_k & \iid \mathcal{N}(0,10)I_{\{\sigma > 0\}} \nonumber \\
+\sigma_k & \iid \mathcal{N}(0,10) \nonumber \\
 \tau_k & \iid \mathcal{N}(0,10)I_{\{\tau > 0\}} \nonumber \\
@@
 \end{align}
-We also note that, for the discrete phenotypes in equations (\ref{eq:disc_corr}) and (\ref{eq:disc_ind}), there is no $\sigma_k$, and thus is not sampled for those phenotypes.
+We also note that, for the discrete phenotypes in equations (\ref{eq:disc_corr}) and (\ref{eq:disc_ind}), there is no $\sigma_k$, and is not sampled, and for the continuous phenotypes in equation (\ref{eq:cont}), there is no $\phi_k$ and is not sampled.
 

### 2023-09-05 R/stuart_stickle_bayes_v2.R

@@ -155,10 +155,15 @@ model{
 "
 
 disc <- "
+function{
+  real genpoiss_lpmf(int y, real lambda_1, real lambda_2){
+
+  }
+}
 data{
   int<lower=0> N;
   int<lower=0> T;
-  int y[N];
+  vector[N] y;
   vector[N] stl;
   matrix[N,2*T] X;
   vector[T] time;
@@ -174,7 +179,7 @@ parameters{
   real beta1_f_stl;
   real beta1_m_stl;
   real<lower=0> kappa;
-  real<upper=1> alpha;
+  real phi;
   real<lower=0> tau;
   real<lower=0> tau_0;
   real gamma;
@@ -209,18 +214,16 @@ transformed parameters{
   } else {
     lambda = exp(X * mu + gamma * stl);
   }
+
+  real alpha = -(exp(phi) + max(-lambda ./ y)) / (1 + exp(phi));
 }
 
 model{
 // Data Model
   for (i in 1:N){
-    if (alpha == 0){
-      target += poisson_lpmf(y[i] | lambda[i]);
-    } else {
-      target += log(lambda[i]) +
-                (y[i] - 1) * log(lambda[i] + alpha * y[i]) -
-                (lambda[i] + alpha * y[i]);
-    }
+    target += log(lambda[i]) +
+              (y[i] - 1) * log(lambda[i] + alpha * y[i]) -
+              (lambda[i] + alpha * y[i]);
   }
 
   // OU Process for Gender and Time Specific Means
@@ -236,7 +239,7 @@ model{
   beta1_f ~ normal(0,3);
   beta1_m ~ normal(0,3);
   kappa ~ normal(0,1);
-  alpha ~ normal(0,10);
+  phi ~ normal(0,10);
   tau ~ normal(0,10);
   tau_0 ~ normal(0,20);
   gamma ~ normal(0,5);

### 2024-05-06 manuscript/paper.Rmd
Small changes and some notes in the results section.

@@
 author: | 
-  | \large Allison Ozark$^{1,*}$, Matthew Stuart$^{2,3,*}$, Raheyma Siddiqui$^1$, Akhil Ghosh$^2$
+  | \large Allison Ozark$^{1,*}$, Matthew Stuart$^{2,3,*}$, Raheyma Siddiqui$^1$, Akhil Ghosh$^{2,3}$
   | \large Samantha Swank$^1$, Michael A. Bell$^4$, Gregory J. Matthews$^{2,3}$, and Yoel E. Stuart$^{1,+}$ 
@@
 |----------------------|------------|-----------------------------|
-|Standard Length|	stl|	Distance from anterior tip of premaxilla to posterior end of last vertebra (hypural plate)|
-|Dorsal Spine | mds |	Number of dorsal spines from 0 to 3|
-|Dorsal Fin Ray|mdf|	Number of bones in the dorsal fin posterior to the third dorsal spine (i.e., soft dorsal fin rays)
-|Anal Fin Ray |maf|	Number of bones in the anal fin posterior to the anal spine (i.e., soft dorsal fin rays)|
+|Standard Length|	`stl` |	Distance from anterior tip of premaxilla to posterior end of last vertebra (hypural plate)|
+|Dorsal Spine | `mds` |	Number of dorsal spines from 0 to 3|
+|Dorsal Fin Ray|`mdf`|	Number of bones in the dorsal fin posterior to the third dorsal spine (i.e., soft dorsal fin rays)
+|Anal Fin Ray |`maf`|	Number of bones in the anal fin posterior to the anal spine (i.e., soft dorsal fin rays)|
 |Abdominal Vertebra|mav|	Number of vertebrae anterior to the first vertebra contacting an anal fin pterygiophore (Aguirre et al. 2014)|
@@
 Traditionally, the choice for the number of completed dats sets is a relatively small number such as $M = 5$ or $M = 10$.
-However, @ZhouReiter2009 recommend a larger number of imputed data sets if the data users intend on performing Bayesians analysis after imputation, which in this case, we do.
+However, @ZhouReiter2010 recommend a larger number of imputed data sets if the data users intend on performing Bayesians analysis after imputation, which in this case, we do.
 Therefore, the imputation algorithm is run to obtain a total of $M = 100$ completed datasets.
-In addition to this, @ZhouReiter2009 suggests rather than using Rubin's combining rules to combine across imputed data sets, instead pool all of the draws from the posterior distributions across all of the imputed data sets to estimate the posterior distributions of parameters of interest.
+In addition to this, @ZhouReiter2010 suggests rather than using Rubin's combining rules to combine across imputed data sets, instead pool all of the draws from the posterior distributions across all of the imputed data sets to estimate the posterior distributions of parameters of interest.
 We proceed with our Bayesian analysis in this manner.
@@
 
-<!-- Bayesian Analysis after multiple imputation @ZhouReiter2009: They recommend using a large number of imputations.  5 or 10 is too small.  We are using M = 100. -->
+<!-- Bayesian Analysis after multiple imputation @ZhouReiter2010: They recommend using a large number of imputations.  5 or 10 is too small.  We are using M = 100. -->
 
@@
 Greg, check and see if this makes sense.}
+
+\textcolor{red}{I think we need to move a lot of these details to the appendix.  }
 
@@
 \end{align}
-where $E(Y_{k,ij}) = \lambda$ and $Var(Y_{k,ij}) = \frac{\lambda}{(1 - \alpha)^2}$.
+where $E(X) = \lambda$ and $Var(X) = \frac{\lambda}{(1 - \alpha)^2}$.
 If $\alpha > 0$, then the variance is greater than the mean, called overdispersion; if $\alpha < 0$, then the variance is greater than the mean, called underdispersion; if $\alpha = 0$, then the model degenerates to a Poisson distribution.
@@
 
-Bayesian Analysis after multiple imputation @ZhouReiter2009: They recommend using a large number of imputations.  5 or 10 is too small.  We are using M = 100.
-
 # Results {#sec:results}
+Greg notes: 
+- mdf and maf have interesting trends.  
+- ds3 flips over time.  Females have larger values to start and over the course of the time frame wer are looking at males end up with larger mean values.  
+- Most of the other traits have mean differences that are relatively constant over time.
 

### 2024-05-15 manuscript/paper.Rmd
Manuscript updates.

@@
 
+- FINAL SUMMARY TABLES 
+ 1. posterior probabilities of mu diff greter than 0. (sexual dimorphism)
+ 2. Posterior probabilities of the difference in the distribution at time 0 and time 19.  (trends)
+
+```{r echo = FALSE}
+ library(tidyverse)
+library(ROCR)
+model_list <- list()
+#Indicate which model was used for each trai
+model_list[["cle"]] <- 3
+model_list[["ds1"]] <- 3
+model_list[["ds2"]] <- 1
+model_list[["ds3"]] <- 1
+model_list[["ect"]] <- 1
+model_list[["lps"]] <- 3
+model_list[["lpt"]] <- 2
+model_list[["maf"]] <- 1
+model_list[["mav"]] <- 2
+model_list[["mcv"]] <- 0
+model_list[["mdf"]] <- 1
+#model_list[["mds"]] <- 3
+#model_list[["mpt"]] <- 0
+model_list[["pmx"]] <- 3
+model_list[["stl"]] <- 2
+#model_list[["tpg"]] <- 1
+library(here)
+samps_stl <- readRDS(here("Figures/stl","posterior_samples_stl_2.RDS"))
+mu_stl <- samps_stl %>% select(contains("mu"))
+ maxtime <- 18
+ results_prob_mu_diff_gt_mu_diff0 <- results_prob_mu_diff_gt_0 <- results_auc_mu_diff <-  data.frame()
+for (phenotype in names(model_list)){
+  samps <- readRDS(here(paste0("Figures/",phenotype),paste0("posterior_samples_",phenotype,"_",model_list[[phenotype]],".RDS")))
+  gamma <- samps %>% select(gamma)
+  mu <- samps %>% select(contains("mu"))
+  if (phenotype == "stl") {
+    mu_f <-
+      mu[, 1:maxtime]
+    mu_m <-
+      mu[, (maxtime + 1):(2 * maxtime)]
+  } else {
+
+    if (phenotype %in% c("mdf","maf","mav","mcv")){
+      mu_f <-exp(mu[, 1:maxtime])
+      mu_m <- exp(mu[, (maxtime + 1):(2 * maxtime)])
+
+    } else {
+      mu_f <-
+        mu[, 1:maxtime] + as.data.frame(rep(gamma, 18)) * mu_stl[, 1:maxtime]
+      mu_m <-
+        mu[, (maxtime + 1):(2 * maxtime)] + as.data.frame(rep(gamma, 18)) * mu_stl[, (maxtime +
+                                                                                        1):(2 * maxtime)]
+    }
+
+    }
+
+  mu_diff <- mu_m - mu_f
+
+  #Probability males are greater than females
+  results_prob_mu_diff_gt_0 <- rbind(results_prob_mu_diff_gt_0,data.frame(
+    phenotype = phenotype,
+    time = 1:18,
+    prob_mu_diff_gt_0 = apply(mu_diff, 2, function(x) {
+      mean(x > 0)
+    })
+  ))
+
+  #Differences in the distributions from time point 1 to time point 18
+  auc_mu_diff <- c()
+  for (j in 1:18){
+  pred <- prediction(c(mu_diff[,1], mu_diff[,j]), rep(c(1,0), each = nrow(mu_diff)))
+  auc_mu_diff[j] <- perf <- performance(pred, "auc")@y.values[[1]]
+  }
+
+  results_auc_mu_diff <- rbind(results_auc_mu_diff,data.frame(
+    phenotype = phenotype,
+    time = 1:18,
+    auc_mu_diff = auc_mu_diff
+    ))
+
+  prob_mu_diff_gt_mu_diff0 <- c()
+  for (j in 1:18){
+    prob_mu_diff_gt_mu_diff0[j] <- mean(sample(mu_diff[,1],10000, replace = TRUE) >= sample(mu_diff[,j],10000, replace = TRUE))
+
+  }
+
+  results_prob_mu_diff_gt_mu_diff0 <- rbind(results_prob_mu_diff_gt_mu_diff0,data.frame(
+    phenotype = phenotype,
+    time = 1:18,
+    prob_mu_diff_gt_mu_diff0 = prob_mu_diff_gt_mu_diff0
+  ))
+
+  # plot(perf,
+  #      avg= "threshold",
+  #      colorize=TRUE,
+  #      lwd= 3,
+  #      main= "With ROCR you can produce standard plots\nlike ROC curves ...")
+  # plot(perf,
+  #      lty=3,
+  #      col="black",
+  #      add=TRUE)
+  #
+
+}
+
+# ggplot(aes(x = time, y = auc_mu_diff), data = results_auc_mu_diff) + facet_wrap(~phenotype) + geom_path() +  geom_path() + geom_hline(yintercept = 0.5, lty = 3) + ylim(0,1)
+# ggplot(aes(x = time, y = prob_mu_diff_gt_mu_diff0), data = results_prob_mu_diff_gt_mu_diff0) + facet_wrap(~phenotype) + geom_path() +  geom_path() + geom_hline(yintercept = 0.5, lty = 3)
+#ggplot(aes(x = time, y = prob_mu_diff_gt_0), data = results_prob_mu_diff_gt_0) + facet_wrap(~phenotype) + geom_path() + geom_hline(yintercept = 0.5, lty = 3) + ylim(0,1)
+
+#Now make some tables.
+#Rows are traits.
+
+```
+
+```{r echo = FALSE, tab.cap= "Posterior probability that the mean for male sis larger than the mean for females"}
+#Which model was used
+mod_df <- data.frame(phenotype = names(unlist(model_list)), mod = unlist(model_list))
+mod_code <- data.frame(mod = 0:3, mod_name = c("OU-Trend","No OU-Trend",
+                                               "OU-No Trend","NO OU-No Trend"))
+results_prob_mu_diff_gt_0_table <- results_prob_mu_diff_gt_0 %>%  mutate(prob_mu_diff_gt_0 = round(prob_mu_diff_gt_0,3)) %>% pivot_wider(names_from = "phenotype", values_from = "prob_mu_diff_gt_0")
+library(xtable)
+xtable(results_prob_mu_diff_gt_0_table)
+```
+
+```{r echo = FALSE, tab.cap= "AUC comparing time x versus time 0."}
+results_auc_mu_diff_table <- results_auc_mu_diff %>%  mutate(auc_mu_diff = round(auc_mu_diff,3)) %>% pivot_wider(names_from = "phenotype", values_from = "auc_mu_diff")
+library(xtable)
+xtable(results_auc_mu_diff_table)
+```
+
+```{r echo = FALSE, fig.cap = "Posterior probabilities that mean for males is greater than mean for females"}
+ggplot(aes(x = time, y = prob_mu_diff_gt_0), data = results_prob_mu_diff_gt_0) + facet_wrap(~phenotype) + geom_path() + geom_hline(yintercept = 0.5, lty = 3) + ylim(0,1) + ylab("Posterior Probability")
+```
+
+```{r, echo = FALSE, fig.cap= "AUC comparing time x versus time 0."}
+ggplot(aes(x = time, y = auc_mu_diff), data = results_auc_mu_diff) + facet_wrap(~phenotype) + geom_path() +  geom_path() + geom_hline(yintercept = 0.5, lty = 3) + ylim(0,1) + ylab("Posterior Probability")
+```
+
 # Discsusson, Future work and conclusions {#sec:conclusions}

### 2024-05-15 R/matthews_table_results_summary.R

@@ -1,0 +1,124 @@
+library(tidyverse)
+library(ROCR)
+model_list <- list()
+#Indicate which model was used for each trai
+model_list[["cle"]] <- 3
+model_list[["ds1"]] <- 3
+model_list[["ds2"]] <- 1
+model_list[["ds3"]] <- 1
+model_list[["ect"]] <- 1
+model_list[["lps"]] <- 3
+model_list[["lpt"]] <- 2
+model_list[["maf"]] <- 1
+model_list[["mav"]] <- 2
+model_list[["mcv"]] <- 0
+model_list[["mdf"]] <- 1
+#model_list[["mds"]] <- 3
+#model_list[["mpt"]] <- 0
+model_list[["pmx"]] <- 3
+model_list[["stl"]] <- 2
+#model_list[["tpg"]] <- 1
+
+library(here)
+samps_stl <- readRDS(here("Figures/stl","posterior_samples_stl_2.RDS"))
+mu_stl <- samps_stl %>% select(contains("mu"))
+maxtime <- 18
+results_prob_mu_diff_gt_mu_diff0 <- results_prob_mu_diff_gt_0 <- results_auc_mu_diff <-  data.frame()
+for (phenotype in names(model_list)){
+  samps <- readRDS(here(paste0("Figures/",phenotype),paste0("posterior_samples_",phenotype,"_",model_list[[phenotype]],".RDS")))
+  gamma <- samps %>% select(gamma)
+  mu <- samps %>% select(contains("mu"))
+  if (phenotype == "stl") {
+    mu_f <-
+      mu[, 1:maxtime]
+    mu_m <-
+      mu[, (maxtime + 1):(2 * maxtime)]
+  } else {
+
+    if (phenotype %in% c("mdf","maf","mav","mcv")){
+      mu_f <-exp(mu[, 1:maxtime])
+      mu_m <- exp(mu[, (maxtime + 1):(2 * maxtime)])
+
+    } else {
+      mu_f <-
+        mu[, 1:maxtime] + as.data.frame(rep(gamma, 18)) * mu_stl[, 1:maxtime]
+      mu_m <-
+        mu[, (maxtime + 1):(2 * maxtime)] + as.data.frame(rep(gamma, 18)) * mu_stl[, (maxtime +
+                                                                                        1):(2 * maxtime)]
+    }
+
+
+
+    }
+
+  mu_diff <- mu_m - mu_f
+
+  #Probability males are greater than females
[…]

### 2024-05-15 manuscript/paper.Rmd
First draft of results with table and plots.  Added posterior mean differences over time in supplementary materials.

@@
 |Abdominal Vertebra|mav|	Number of vertebrae anterior to the first vertebra contacting an anal fin pterygiophore (Aguirre et al. 2014)|
-|Caudal Vertebra|mcv	| Number of vertebrae posterior to and including the first vertebra contacting an anal fin pterygiophore (Aguirre et al. 2014)|
-|Pterygiophore number|mpt|	Number of pterygiophores anterior to but excluding the pterygiophore under the third dorsal spine, which is immediately anterior to and contiguous with the dorsal fin|
-|Pelvic Spine length|lps.sc	|Length from the base of one pelvic spine above its articulation with the pelvic girdle to its distal tip|
-|Ectocoracoid|ect.sc|	Length between the anterior and posterior tips of the shoulder girdle base (i.e., ectocoracoid)|
-|Pelvic Girdle|tpg.sc|	Length between the anterior to posterior tips along midline. If vestigial, the sum of longest anterior-posterior axis for the vestiges|
-|Cleithrum length|cle.sc|	Length from free dorsal tip to ventral tip of the cleithrum on the anterior margin of the shoulder girdle (i.e., cleithrum)|
-|Premaxilla|pmx.sc|	Length from the anterior tip of the premaxilla to the distal tip of the ascending process of the premaxilla |
-|Dorsal Spine|Ds#.sc# = 1,2,or 3|	Length from the base of a dorsal spine above the pterygiophore to its distal tip along the anterior edge|
-|Pterygiophore|lpt.sc|	Distance between the anterior to posterior tips of the pterygiophore immediately preceding the 3rd dorsal spine (when present)|
+|Caudal Vertebra|`mcv`	| Number of vertebrae posterior to and including the first vertebra contacting an anal fin pterygiophore (Aguirre et al. 2014)|
+|Pterygiophore number|`mpt`|	Number of pterygiophores anterior to but excluding the pterygiophore under the third dorsal spine, which is immediately anterior to and contiguous with the dorsal fin|
+|Pelvic Spine length|`lps.sc`	|Length from the base of one pelvic spine above its articulation with the pelvic girdle to its distal tip|
+|Ectocoracoid|`ect.sc`|	Length between the anterior and posterior tips of the shoulder girdle base (i.e., ectocoracoid)|
+|Pelvic Girdle|`tpg.sc`|	Length between the anterior to posterior tips along midline. If vestigial, the sum of longest anterior-posterior axis for the vestiges|
+|Cleithrum length|`cle.sc`|	Length from free dorsal tip to ventral tip of the cleithrum on the anterior margin of the shoulder girdle (i.e., cleithrum)|
+|Premaxilla|`pmx.sc`|	Length from the anterior tip of the premaxilla to the distal tip of the ascending process of the premaxilla |
+|Dorsal Spine|`Ds#.sc# = 1,2,or 3`|	Length from the base of a dorsal spine above the pterygiophore to its distal tip along the anterior edge|
+|Pterygiophore|`lpt.sc`|	Distance between the anterior to posterior tips of the pterygiophore immediately preceding the 3rd dorsal spine (when present)|
 
@@
 # Results {#sec:results}
-Greg notes: 
-- mdf and maf have interesting trends.  
-- ds3 flips over time.  Females have larger values to start and over the course of the time frame wer are looking at males end up with larger mean values.  
-- Most of the other traits have mean differences that are relatively constant over time.
+As previously described, four models for the mean process were considered for each trait.
+Within each trait, the model with the lowest DIC was chosen for that train.
+All subsequent results presented here use the model with the lowest DIC individually by trait.
+Table \ref{whichmodel} shows the model that was selected for each trait.
+About half of the traits had models with a trend, and most models did not include an OU component.
+The traits that did include an OU component were caudal vertebra (mcv), pelvic spine lentgh (lpt), abdominal vertebra (mav), and standard length (stl) with caudal vertebra (mcv) being the onyl model with both a trend and an OU component.
+NOTE FOR MATT: You should mention something about MCV being the only one with a trend and an OU.
+Explain why this might be and what it means.
 
-- FINAL SUMMARY TABLES 
- 1. posterior probabilities of mu diff greter than 0. (sexual dimorphism)
- 2. Posterior probabilities of the difference in the distribution at time 0 and time 19.  (trends)
+## Best fitting models 
+\begin{table}[ht]
+\centering
+\begin{tabular}{|c||c|c|}
+  \hline
+ & OU & No OU  \\ 
+  \hline
+  \hline
+Trend  &  mcv  & ds2, ds3, ect, maf, mdf \\ 
+  \hline
+  No Trend & lpt, mav, stl  &  cle, ds1, lps, pmx\\
+   \hline
+\end{tabular}
+\label{whichmodel}
+\end{table}
+
+<!-- Trends terms included in were observed in ds2, ds3, ect, maf, mdf, mcv,  -->
+<!-- OU term included in model: mcv, lpt, stl -->
+<!-- model_list[["cle"]] <- 3 -->
+<!-- model_list[["ds1"]] <- 3 -->
+<!-- model_list[["ds2"]] <- 1 -->
+<!-- model_list[["ds3"]] <- 1 -->
+<!-- model_list[["ect"]] <- 1 -->
+<!-- model_list[["lps"]] <- 3 -->
+<!-- model_list[["lpt"]] <- 2 -->
+<!-- model_list[["maf"]] <- 1 -->
+<!-- model_list[["mav"]] <- 2 -->
+<!-- model_list[["mcv"]] <- 0 -->
+<!-- model_list[["mdf"]] <- 1 -->
+<!-- #model_list[["mds"]] <- 3 -->
+<!-- #model_list[["mpt"]] <- 0 -->
+<!-- model_list[["pmx"]] <- 3 -->
+<!-- model_list[["stl"]] <- 2 -->
+<!-- #model_list[["tpg"]] <- 1 -->
+<!-- Greg notes:  -->
+<!-- - mdf and maf have interesting trends.   -->
+<!-- - ds3 flips over time.  Females have larger values to start and over the course of the time frame wer are looking at males end up with larger mean values.   -->
+<!-- - Most of the other traits have mean differences that are relatively constant over time.   -->
+
+## Sexual Dimorphism 
+Table \ref{prob0-table} shows the posterior probability of the differences in the means between male and female specimens at each time point.
+Probabilities near 0.5 indicate very little difference in the means whereas probabilities far from 0.5 indicate sexual dimorphism with values of 0 and 1 indicating larger mean values for females and males, respectively.
+These results are shown graphically in figure \ref{fig:prob0-plot}.
+
+Traits such as standard length (stl), anal fin ray (maf), dorsal fin ray (mdf), premaxilla (pmx), and dorsal spine 1 (ds1) demonstrate constant sexual dimorphism with all but the last having learger mean values in male sepcimen.
+Other traits such as abdominal vertebra (mav), caudal vertebra (mcv), and pelvic spine length (lps) don't appear to have any discernable sexual dimorphism in the means.
+Dorsal spine 3 (ds3) exhibits sexual dimorphism that changes over time.
+The mean number of dorsal spine 3 begins larger for females for the earliest observed time periods.
+Over time, however, this switches and the male specimens exhibit larger mean values for the latest observed time periods.
+
+## Changes over time
+Table \ref{auc-table} shows the probability that the posterior distribution of the differences in the mean is greater than the posterior distribution of the differences at the first time point.
+Values near 0.5 indicate very little difference between the distributions whereas values closer to 0 and 1 indicate changes in the distributions with values of 1 meaning the difference in means has shifted towards males having a larger mean and values of 0 indicating the difference in means has shifted towards larger means in females.
+These results are shown graphically in figure \ref{fig:AUC-plot}.
+
+The amount of sexual dimporphism over time does not change substantially for a majority of the trait observed here (i.e.
+cle, ds1, ds2, ect, lps, lpt, mav, mcv, pmx, stl).
+Traits that trend away from the mean difference at the first observed time point include dorsal spine 3 (ds3), anal fin ray (maf), and dorsal fin ray (mdf).
+
+<!-- #ds3 -->
+<!-- apply(mu_m,2,mean)[1] #2.291846 -->
+<!-- apply(mu_f,2,mean)[1] #2.448115  -->
+<!-- apply(mu_m,2,mean)[18] #1.264728  -->
+<!--   apply(mu_f,2,mean)[18] #1.184534  -->
+<!-- apply(mu_m,2,mean)[18] - apply(mu_m,2,mean)[1] #-1.027118  -->
+<!-- apply(mu_f,2,mean)[18] - apply(mu_f,2,mean)[1] #-1.263581  -->
+The posterior mean of the mean number of dorsal spine rays at the first observed time point is 2.29 for males and for females it is 2.45.
+At the last observed time point, the average number for males and female both dropped to 1.265 and 1.185 for males and females, respectively.
+This resulted in a mean number of dorsal spine 3 lost by males of 1.027 and for females it was 1.264.
+This resulted in a change of the mean difference from time point 1 to time point 18 of 0.236.
+Basically, males and females both lowered the mean number of dorsal spine 3, but females lowered their mean more than the males.
+
+<!-- #maf -->
+  <!-- apply(mu_m,2,mean)[1] #8.517534 -->
+  <!-- apply(mu_f,2,mean)[1] #8.290749  -->
+  <!-- apply(mu_m,2,mean)[18] #8.549155  -->
+  <!-- apply(mu_f,2,mean)[18] #8.170195  -->
+  <!-- apply(mu_m,2,mean)[18] - apply(mu_m,2,mean)[1] #0.03162047  -->
+  <!-- apply(mu_f,2,mean)[18] - apply(mu_f,2,mean)[1] #-0.1205537 -->
+  <!-- (apply(mu_m,2,mean) - apply(mu_f,2,mean))[18] - (apply(mu_m,2,mean) - apply(mu_f,2,mean))[1] #0.1521742  -->
+<!-- MAF: Number of bones in the anal fin posterior to the anal spine (i.e., softdorsal fin rays) -->
+
+We also observed a change in the differences between males and females for the number of bones int he anal fin ray (maf) and the number of bones in the dorsal fin ray.
+The difference in the means of males versus females for anal fin ray bones changed by 0.152 over the course of the observed time periods.
+This change was driven mostly by the change in females.
+At the first observed time period the had on average 8.291 bones dropping to 8.170 over the 18 times periods observed.
+Whereas for males the change in means rose only from  8.518 to 8.549, about 0.0316 bones on average.
+
+A similar change in the difference of the means over time was observed for the number of bones in the dorsal fin (mdf).
+However, unlike maf, this change in sexual dimporhism was driven entirely by the males.
+The mean change for females over the 18 observed time periods was only 0.0055 with the average number of dorsal fin bones for females starting at 9.287 and the final observed mean of 9.281 resulting in almost no change in the mean.
+For males, the average number of dorsal fin bones at the first time point was about 9.568 dropping to 9.734 over the course of the observed time periods for a gain of 0.166 bones on average.
+
+<!-- #mdf -->
+<!--   apply(mu_m,2,mean)[1] #9.567811  -->
+<!--   apply(mu_f,2,mean)[1] #9.286521 -->
+<!--   apply(mu_m,2,mean)[18] #9.733923 -->
+<!--   apply(mu_f,2,mean)[18] #9.281005 -->
+<!--   apply(mu_m,2,mean)[18] - apply(mu_m,2,mean)[1] #0.1661123  -->
+<!--   apply(mu_f,2,mean)[18] - apply(mu_f,2,mean)[1] #-0.005516126  -->
+<!--   (apply(mu_m,2,mean) - apply(mu_f,2,mean))[18] - (apply(mu_m,2,mean) - apply(mu_f,2,mean))[1] #0.1716284 -->
+<!-- MDF: Dorsal fin ray: Number of bones in the dorsal finposterior to the third dorsal spine (i.e.,soft dorsal fin rays. -->
+
+```{r echo = FALSE}
+library(tidyverse)
+library(ROCR)
+model_list <- list()
+#Indicate which model was used for each trai
+model_list[["cle"]] <- 3
+model_list[["ds1"]] <- 3
+model_list[["ds2"]] <- 1
+model_list[["ds3"]] <- 1
+model_list[["ect"]] <- 1
+model_list[["lps"]] <- 3
+model_list[["lpt"]] <- 2
+model_list[["maf"]] <- 1
+model_list[["mav"]] <- 2
+model_list[["mcv"]] <- 0
+model_list[["mdf"]] <- 1
+#model_list[["mds"]] <- 3
+#model_list[["mpt"]] <- 0
+model_list[["pmx"]] <- 3
+model_list[["stl"]] <- 2
+#model_list[["tpg"]] <- 1
+library(here)
+samps_stl <- readRDS(here("Figures/stl","posterior_samples_stl_2.RDS"))
+mu_stl <- samps_stl %>% select(contains("mu"))
+ maxtime <- 18
+ results_prob_mu_diff_gt_mu_diff0 <- results_prob_mu_diff_gt_0 <- results_auc_mu_diff <-  data.frame()
+phenotype <- "ds3"
+  samps <- readRDS(here(paste0("Figures/",phenotype),paste0("posterior_samples_",phenotype,"_",model_list[[phenotype]],".RDS")))
+  gamma <- samps %>% select(gamma)
+  mu <- samps %>% select(contains("mu"))
+  if (phenotype == "stl") {
+    mu_f <-
+      mu[, 1:maxtime]
+    mu_m <-
+      mu[, (maxtime + 1):(2 * maxtime)]
+  } else {
+
+    if (phenotype %in% c("mdf","maf","mav","mcv")){
+      mu_f <-exp(mu[, 1:maxtime])
+      mu_m <- exp(mu[, (maxtime + 1):(2 * maxtime)])
+
+    } else {
+      mu_f <-
+        mu[, 1:maxtime] + as.data.frame(rep(gamma, 18)) * mu_stl[, 1:maxtime]
+      mu_m <-
+        mu[, (maxtime + 1):(2 * maxtime)] + as.data.frame(rep(gamma, 18)) * mu_stl[, (maxtime +
+                                                                                        1):(2 * maxtime)]
+    }
+
+  }
+
+  #ds3 
+  # apply(mu_m,2,mean)[1] #2.291846
+  # apply(mu_f,2,mean)[1] #2.448115 
+  # 
+  # apply(mu_m,2,mean)[18] #1.264728 
+  # apply(mu_f,2,mean)[18] #1.184534 
+  # 
+  # apply(mu_m,2,mean)[18] - apply(mu_m,2,mean)[1] #-1.027118 
+  # apply(mu_f,2,mean)[18] - apply(mu_f,2,mean)[1] #-1.263581 
+  # 
+  # (apply(mu_m,2,mean) - apply(mu_f,2,mean))[18] - (apply(mu_m,2,mean) - apply(mu_f,2,mean))[1] #0.2364628
+```
+
+```{r echo = FALSE}
+library(tidyverse)
+library(ROCR)
+model_list <- list()
+#Indicate which model was used for each trai
+model_list[["cle"]] <- 3
+model_list[["ds1"]] <- 3
+model_list[["ds2"]] <- 1
+model_list[["ds3"]] <- 1
+model_list[["ect"]] <- 1
+model_list[["lps"]] <- 3
+model_list[["lpt"]] <- 2
+model_list[["maf"]] <- 1
+model_list[["mav"]] <- 2
+model_list[["mcv"]] <- 0
+model_list[["mdf"]] <- 1
+#model_list[["mds"]] <- 3
+#model_list[["mpt"]] <- 0
+model_list[["pmx"]] <- 3
+model_list[["stl"]] <- 2
+#model_list[["tpg"]] <- 1
+library(here)
+samps_stl <- readRDS(here("Figures/stl","posterior_samples_stl_2.RDS"))
+mu_stl <- samps_stl %>% select(contains("mu"))
+ maxtime <- 18
+ results_prob_mu_diff_gt_mu_diff0 <- results_prob_mu_diff_gt_0 <- results_auc_mu_diff <-  data.frame()
+phenotype <- "maf"
+  samps <- readRDS(here(paste0("Figures/",phenotype),paste0("posterior_samples_",phenotype,"_",model_list[[phenotype]],".RDS")))
+  gamma <- samps %>% select(gamma)
+  mu <- samps %>% select(contains("mu"))
+  if (phenotype == "stl") {
+    mu_f <-
+      mu[, 1:maxtime]
+    mu_m <-
+      mu[, (maxtime + 1):(2 * maxtime)]
+  } else {
+
+    if (phenotype %in% c("mdf","maf","mav","mcv")){
+      mu_f <-exp(mu[, 1:maxtime])
+      mu_m <- exp(mu[, (maxtime + 1):(2 * maxtime)])
+
+    } else {
+      mu_f <-
+        mu[, 1:maxtime] + as.data.frame(rep(gamma, 18)) * mu_stl[, 1:maxtime]
+      mu_m <-
+        mu[, (maxtime + 1):(2 * maxtime)] + as.data.frame(rep(gamma, 18)) * mu_stl[, (maxtime +
+                                                                                        1):(2 * maxtime)]
+    }
+
+  }
+
+  # #maf
+  # apply(mu_m,2,mean)[1] #8.517534
+  # apply(mu_f,2,mean)[1] #8.290749 
+  # 
+  # apply(mu_m,2,mean)[18] #8.549155 
+  # apply(mu_f,2,mean)[18] #8.170195 
+  # 
+  # apply(mu_m,2,mean)[18] - apply(mu_m,2,mean)[1] #0.03162047 
+  # apply(mu_f,2,mean)[18] - apply(mu_f,2,mean)[1] #-0.1205537
+  # 
+  # (apply(mu_m,2,mean) - apply(mu_f,2,mean))[18] - (apply(mu_m,2,mean) - apply(mu_f,2,mean))[1] #0.1521742 
+```
+
+```{r echo = FALSE}
+library(tidyverse)
+library(ROCR)
+model_list <- list()
+#Indicate which model was used for each trai
+model_list[["cle"]] <- 3
+model_list[["ds1"]] <- 3
+model_list[["ds2"]] <- 1
+model_list[["ds3"]] <- 1
+model_list[["ect"]] <- 1
+model_list[["lps"]] <- 3
+model_list[["lpt"]] <- 2
+model_list[["maf"]] <- 1
+model_list[["mav"]] <- 2
+model_list[["mcv"]] <- 0
+model_list[["mdf"]] <- 1
+#model_list[["mds"]] <- 3
+#model_list[["mpt"]] <- 0
+model_list[["pmx"]] <- 3
+model_list[["stl"]] <- 2
+#model_list[["tpg"]] <- 1
+library(here)
+samps_stl <- readRDS(here("Figures/stl","posterior_samples_stl_2.RDS"))
+mu_stl <- samps_stl %>% select(contains("mu"))
+ maxtime <- 18
+ results_prob_mu_diff_gt_mu_diff0 <- results_prob_mu_diff_gt_0 <- results_auc_mu_diff <-  data.frame()
+phenotype <- "mdf"
+  samps <- readRDS(here(paste0("Figures/",phenotype),paste0("posterior_samples_",phenotype,"_",model_list[[phenotype]],".RDS")))
+  gamma <- samps %>% select(gamma)
+  mu <- samps %>% select(contains("mu"))
+  if (phenotype == "stl") {
+    mu_f <-
+      mu[, 1:maxtime]
+    mu_m <-
+      mu[, (maxtime + 1):(2 * maxtime)]
+  } else {
+
+    if (phenotype %in% c("mdf","maf","mav","mcv")){
+      mu_f <-exp(mu[, 1:maxtime])
+      mu_m <- exp(mu[, (maxtime + 1):(2 * maxtime)])
+
+    } else {
+      mu_f <-
+        mu[, 1:maxtime] + as.data.frame(rep(gamma, 18)) * mu_stl[, 1:maxtime]
+      mu_m <-
+        mu[, (maxtime + 1):(2 * maxtime)] + as.data.frame(rep(gamma, 18)) * mu_stl[, (maxtime +
+                                                                                        1):(2 * maxtime)]
+    }
+
+  }
+
+  #mdf
+  # apply(mu_m,2,mean)[1] #9.567811 
+  # apply(mu_f,2,mean)[1] #9.286521
+  # 
+  # apply(mu_m,2,mean)[18] #9.733923
+  # apply(mu_f,2,mean)[18] #9.281005
+  # 
+  # apply(mu_m,2,mean)[18] - apply(mu_m,2,mean)[1] #0.1661123 
+  # apply(mu_f,2,mean)[18] - apply(mu_f,2,mean)[1] #-0.005516126 
+  # 
+  # (apply(mu_m,2,mean) - apply(mu_f,2,mean))[18] - (apply(mu_m,2,mean) - apply(mu_f,2,mean))[1] #0.1716284
+```
+
+<!-- %- FINAL SUMMARY TABLES  -->
+<!--  1. posterior probabilities of mu diff greter than 0. (sexual dimorphism) -->
+<!--  2. Posterior probabilities of the difference in the distribution at time 0 and time 19.  (trends) -->
 
@@
 
-```{r echo = FALSE, tab.cap= "Posterior probability that the mean for male sis larger than the mean for females"}
+```{r prob0-table, echo = FALSE, tab.cap= "Posterior probability that the mean for males is larger than the mean for females. \\label{prob0-table}"}
 #Which model was used
@@
 
-```{r echo = FALSE, tab.cap= "AUC comparing time x versus time 0."}
+```{r auc-table, echo = FALSE, tab.cap= "AUC comparing time x versus time 0. \\label{auc-table}"}
 results_auc_mu_diff_table <- results_auc_mu_diff %>%  mutate(auc_mu_diff = round(auc_mu_diff,3)) %>% pivot_wider(names_from = "phenotype", values_from = "auc_mu_diff")
-library(xtable)
-xtable(results_auc_mu_diff_table)
+#library(xtable)
+xtable(results_auc_mu_diff_table[1,])
+#knitr::kable(results_auc_mu_diff_table, caption = "The mtcars data.")
 ```
 
-```{r echo = FALSE, fig.cap = "Posterior probabilities that mean for males is greater than mean for females"}
+```{r prob0-plot, echo = FALSE, fig.cap = "Posterior probabilities that mean for males is greater than mean for females"}
 ggplot(aes(x = time, y = prob_mu_diff_gt_0), data = results_prob_mu_diff_gt_0) + facet_wrap(~phenotype) + geom_path() + geom_hline(yintercept = 0.5, lty = 3) + ylim(0,1) + ylab("Posterior Probability")
@@
 
-```{r, echo = FALSE, fig.cap= "AUC comparing time x versus time 0."}
+```{r AUC-plot, echo = FALSE, fig.cap= "AUC comparing time x versus time 0."}
 ggplot(aes(x = time, y = auc_mu_diff), data = results_auc_mu_diff) + facet_wrap(~phenotype) + geom_path() +  geom_path() + geom_hline(yintercept = 0.5, lty = 3) + ylim(0,1) + ylab("Posterior Probability")
@@
 
+DIC results here.
+
+## Figures
+```{r, out.width="0.9\\linewidth", include=TRUE, fig.align="center", fig.cap=c("cle"), echo=FALSE}
+knitr::include_graphics(here("Figures/cle","mu_diff.pdf"))
+```
+```{r, out.width="0.9\\linewidth", include=TRUE, fig.align="center", fig.cap=c("ds1"), echo=FALSE}
+knitr::include_graphics(here("Figures/ds1","mu_diff.pdf"))
+```
+```{r, out.width="0.9\\linewidth", include=TRUE, fig.align="center", fig.cap=c("ds2"), echo=FALSE}
+knitr::include_graphics(here("Figures/ds2","mu_diff.pdf"))
+```
+```{r, out.width="0.9\\linewidth", include=TRUE, fig.align="center", fig.cap=c("ds3"), echo=FALSE}
+knitr::include_graphics(here("Figures/ds3","mu_diff.pdf"))
+```
+```{r, out.width="0.9\\linewidth", include=TRUE, fig.align="center", fig.cap=c("ect"), echo=FALSE}
+knitr::include_graphics(here("Figures/ect","mu_diff.pdf"))
+```
+```{r, out.width="0.9\\linewidth", include=TRUE, fig.align="center", fig.cap=c("lps"), echo=FALSE}
+knitr::include_graphics(here("Figures/lps","mu_diff.pdf"))
+```
+```{r, out.width="0.9\\linewidth", include=TRUE, fig.align="center", fig.cap=c("lpt"), echo=FALSE}
+knitr::include_graphics(here("Figures/lpt","mu_diff.pdf"))
+```
+```{r, out.width="0.9\\linewidth", include=TRUE, fig.align="center", fig.cap=c("maf"), echo=FALSE}
+knitr::include_graphics(here("Figures/maf","lambda_diff.pdf"))
+```
+```{r, out.width="0.9\\linewidth", include=TRUE, fig.align="center", fig.cap=c("mav"), echo=FALSE}
+knitr::include_graphics(here("Figures/mav","lambda_diff.pdf"))
+```
+```{r, out.width="0.9\\linewidth", include=TRUE, fig.align="center", fig.cap=c("mcv"), echo=FALSE}
+knitr::include_graphics(here("Figures/mcv","lambda_diff.pdf"))
+```
+```{r, out.width="0.9\\linewidth", include=TRUE, fig.align="center", fig.cap=c("mdf"), echo=FALSE}
+knitr::include_graphics(here("Figures/mdf","lambda_diff.pdf"))
+```
+```{r, out.width="0.9\\linewidth", include=TRUE, fig.align="center", fig.cap=c("mds"), echo=FALSE}
+knitr::include_graphics(here("Figures/mds","lambda_diff.pdf"))
+```
+```{r, out.width="0.9\\linewidth", include=TRUE, fig.align="center", fig.cap=c("mpt"), echo=FALSE}
+knitr::include_graphics(here("Figures/mpt","lambda_diff.pdf"))
+```
+```{r, out.width="0.9\\linewidth", include=TRUE, fig.align="center", fig.cap=c("pmx"), echo=FALSE}
+knitr::include_graphics(here("Figures/pmx","mu_diff.pdf"))
+```
+
+```{r, out.width="0.9\\linewidth", include=TRUE, fig.align="center", fig.cap=c("stl"), echo=FALSE}
+knitr::include_graphics(here("Figures/stl","mu_diff.pdf"))
+```
+
 # References

### 2024-05-28 manuscript/paper.Rmd
Updated Manuscript.

@@
 
+Figure \ref{fig:mu-diff-plot} shows the posterior mean difference for males versus females (values above 0 indicate means that were greater for males versus females).
+The full poosterior distributions over time for all traits are shown in the appendix.
+
 Traits such as standard length (stl), anal fin ray (maf), dorsal fin ray (mdf), premaxilla (pmx), and dorsal spine 1 (ds1) demonstrate constant sexual dimorphism with all but the last having learger mean values in male sepcimen.
@@
  maxtime <- 18
- results_prob_mu_diff_gt_mu_diff0 <- results_prob_mu_diff_gt_0 <- results_auc_mu_diff <-  data.frame()
+ results_posterior_mean_diff <- results_prob_mu_diff_gt_mu_diff0 <- results_prob_mu_diff_gt_0 <- results_auc_mu_diff <-  data.frame()
+
 for (phenotype in names(model_list)){
@@
 
+  results_posterior_mean_diff <- rbind( results_posterior_mean_diff,data.frame(
+    phenotype = phenotype,
+    time = 1:18,
+    mu_diff = apply(mu_diff,2,mean)
+  ))
+
   # plot(perf,
@@
 #library(xtable)
-xtable(results_auc_mu_diff_table[1,])
+xtable(results_auc_mu_diff_table)
 #knitr::kable(results_auc_mu_diff_table, caption = "The mtcars data.")
@@
 
+```{r mu-diff-plot, echo = FALSE, fig.cap = "Posterior mean difference between males and females"}
+ggplot(aes(x = time, y = mu_diff), data = results_posterior_mean_diff) + facet_wrap(~phenotype, scale = "free_y") + geom_path() + geom_hline(yintercept = 0, lty = 3) + ylab("Posterior Mean")
+```
+
 ```{r prob0-plot, echo = FALSE, fig.cap = "Posterior probabilities that mean for males is greater than mean for females"}
-ggplot(aes(x = time, y = prob_mu_diff_gt_0), data = results_prob_mu_diff_gt_0) + facet_wrap(~phenotype) + geom_path() + geom_hline(yintercept = 0.5, lty = 3) + ylim(0,1) + ylab("Posterior Probability")
+ggplot(aes(x = time, y = prob_mu_diff_gt_0), data = results_prob_mu_diff_gt_0) + facet_wrap(~phenotype) + geom_path() + geom_hline(yintercept = 0, lty = 3) + ylim(0,1) + ylab("Posterior Probability")
 ```

### 2024-06-06 manuscript/paper.Rmd
Add minor changes to manuscript

@@
 library(Lahman)
+library(bookdown)
 ```
@@
 where $\kappa_k$ is a parameter associated with the correlation between $u_{k,gt}$ and $u_{k,g(t+dt)}$, $\tau_k$ is the standard deviation of the OU process, and $W_t$ is a standard Brownian motion.
-As shown in (@OUProcess), the closed form solution for the SDE in (\ref{eq:OU_cont}) is
+As shown in @OUProcess, the closed form solution for the SDE in (\ref{eq:OU_cont}) is
 \begin{align}
@@
 
-Because we are fitting a dataset with a stochastic structure on the means of the phenotypes, we analyze the data via a Bayesian analysis.
+Because we are fitting a dataset with a stochastic structure on the means of the phenotypes, and we are interested in the distribution of the overall structure of these means, we analyze the data via a Bayesian analysis.
 Bayesian data analysis is also more naturally used when we have to impute data (\textbf{CITATION}).
@@
 About half of the traits had models with a trend, and most models did not include an OU component.
-The traits that did include an OU component were caudal vertebra (mcv), pelvic spine lentgh (lpt), abdominal vertebra (mav), and standard length (stl) with caudal vertebra (mcv) being the onyl model with both a trend and an OU component.
-NOTE FOR MATT: You should mention something about MCV being the only one with a trend and an OU.
+The traits that did include an OU component were caudal vertebra (mcv), pelvic spine lentgh (lpt), abdominal vertebra (mav), and standard length (stl) with caudal vertebra (mcv) being the only model with both a trend and an OU component.
+This implies that not only does caudal vertebra size have a linear trend across gender, but the average size at one time period is highly correlated with the average size at the next time period.<!-- NOTE FOR MATT: You should mention something about MCV being the only one with a trend and an OU.
 Explain why this might be and what it means.
+-->
 
@@
 ## Sexual Dimorphism 
-Table \ref{prob0-table} shows the posterior probability of the differences in the means between male and female specimens at each time point.
+Table \ref{tab:prob0-table} shows the posterior probability of the differences in the means between male and female specimens at each time point.
 Probabilities near 0.5 indicate very little difference in the means whereas probabilities far from 0.5 indicate sexual dimorphism with values of 0 and 1 indicating larger mean values for females and males, respectively.
@@
 ## Changes over time
-Table \ref{auc-table} shows the probability that the posterior distribution of the differences in the mean is greater than the posterior distribution of the differences at the first time point.
+Table \ref{tab:auc-table} shows the probability that the posterior distribution of the differences in the mean is greater than the posterior distribution of the differences at the first time point.
 Values near 0.5 indicate very little difference between the distributions whereas values closer to 0 and 1 indicate changes in the distributions with values of 1 meaning the difference in means has shifted towards males having a larger mean and values of 0 indicating the difference in means has shifted towards larger means in females.
@@
 
-```{r prob0-table, echo = FALSE, tab.cap= "Posterior probability that the mean for males is larger than the mean for females. \\label{prob0-table}"}
+```{r prob0-table, echo = FALSE, tab.cap= "Posterior probability that the mean for males is larger than the mean for females"}
 #Which model was used
@@
 
-```{r auc-table, echo = FALSE, tab.cap= "AUC comparing time x versus time 0. \\label{auc-table}"}
+```{r auc-table, echo = FALSE, tab.cap= "AUC comparing time x versus time 0."}
 results_auc_mu_diff_table <- results_auc_mu_diff %>%  mutate(auc_mu_diff = round(auc_mu_diff,3)) %>% pivot_wider(names_from = "phenotype", values_from = "auc_mu_diff")
@@
 ```{r prob0-plot, echo = FALSE, fig.cap = "Posterior probabilities that mean for males is greater than mean for females"}
-ggplot(aes(x = time, y = prob_mu_diff_gt_0), data = results_prob_mu_diff_gt_0) + facet_wrap(~phenotype) + geom_path() + geom_hline(yintercept = 0, lty = 3) + ylim(0,1) + ylab("Posterior Probability")
+ggplot(aes(x = time, y = prob_mu_diff_gt_0), data = results_prob_mu_diff_gt_0) + facet_wrap(~phenotype) + geom_path() + #geom_hline(yintercept = 0, lty = 3) + ylim(0,1) + ylab("Posterior Probability") Why yintercept in 0 and not 0.5?
+geom_hline(yintercept = 0.5, lty = 3) + ylim(0,1) + ylab("Posterior Probability")
 ```

### 2024-07-24 manuscript/paper.Rmd
Yoel adding text to introduction, updating and formatting references, etc.

@@
 
-Ecological release theory suggests that a prey population’s niche should expand when predation is relaxed or removed, as prey take advantage of new enemy free space (reviewed in @Herrmann2021).
-Niche expansion could manifest through a combination of increasing within- and between-individual niche widths (@Bolnick2010; @Herrmann2021), including divergence between the sexes (@Bolnick2003; @Cooper2011).
-Theory and empirical data suggest that a population presented with ecological opportunity following ecological release might experience disruptive selection on males and females stemming from intraspecific competition over newly accessible resources, resulting in intersexual divergence in habitat use (@Schoener1968; @Shine1989; @Bolnick2003; @Butler2007; @Bolnick2008; @Cooper2011; but see @Stuart2021; @Blain2022).
-Accordingly, sexes should also diverge in morphological traits associated with habitat use; such ecologically-driven "character displacement between the sexes" is therefore one explanation for phenotypic sexual dimorphism (@deLisleRowe2015; @deLisleNco2018).
+Ecological release theory suggests that a population’s niche should change when important species interactions (e.g., resource competition) are relaxed or removed (reviewed in @Herrmannetal2021).
+Removal is posited to create ecological opportunities—i.e., aspects of the niche become newly accessible, and the focal population shifts and/or expands its resource use to the new opportunity (@ParentandCrespi2009; @Herrmannetal2021).
+Ecological release may then be followed by adaptive morphological evolution (@ParentandCrespi2009; @Herrmannetal2021).
+For example, a population undergoing ecological release can experience disruptive selection on males and females stemming from intraspecific competition over newly accessible resources (@BolnickandDoebeli2003; @BolnickandLau2008; @Cooperetal2011).
+This is predicted to result in intersexual divergence in habitat use and associated phenotypes (@Schoener1968; @Shine1989; @BolnickandDoebeli2003; @Butleretal2007; @BolnickandLau2008; @Cooperetal2011; but see @Stuartetal2021; @Blain2022).
+Such competition-driven “intraspecific character displacement” between the sexes is therefore one explanation for the evolution of sexual dimorphism (@PfennigandPfennig2012; @DeLisleandRowe2015; @DeLisleandRowe2017, @De Lisleetal2018).
 
-We tested this link between release from predators and the evolution of sexual dimorphism in a well-preserved, finely-resolved, ~16,000 year-long sequence of the fossil stickleback fish (<i>Gasterosteus doryssus</i>).
-The fossil lineage appeared in the depositional environment as a fully armored form, with complete pelvic girdles, two pelvic spines, and three dorsal spines, on average (@Bell2006; @Stuart2020).
-However, this habitat appears to have been missing fish piscivores (e.g., trout and other salmonids known to prey on modern stickleback; @Bell2009), relaxing putative selection for armor (@Reimchen1980, @Reimchen1994, @BellEtAl1993; @Bowne1994; @Roestietal2023).
-The lineage adaptively evolved armor loss but also reduction in the means of several other traits  (@Hunt2008; @Stuart2020; Siddiqui et al.
-in prep).
-In particular, tooth wear data suggest that the population shifted to eat more planktonic prey through time, indicative of increased open water habitat use (Purnell et al.
+Much of the theory and empirical data for character displacement between the sexes is based on release from resource competition specifically (e.g., @BolnickandDoebeli2003; @Cooperetal2011; @PfennigandPfennig2012; @DeLisleandRowe2015; @DeLisleetal2018).
+However, release from predation should also result in ecological release followed by the evolution of increased sexual dimorphism because the absence of predators should generate ecological opportunity (@ReimchenandNosil2004; @ParentandCrespi2009; @Herrmannetal2021).
+
+Here, we tested the prediction that a release from predation should result in the evolution of increased sexual dimorphism using a well-preserved, finely-resolved sequence of a fossil threespine stickleback fish (Gasterosteus doryssus).
+The sequence in this depositional environment is comprised of two lineages (@Bell2009; @Cerasonietal2024).
+Lineage I was a low-armor form with zero to one dorsal spines and highly reduced pelvises, on average.
+It lasted for at least 93,000 years before it was replaced suddenly by a second lineage (@Belletal1985).
+Lineage II appeared in the depositional environment fully armored, with complete pelvic girdles, two pelvic spines, and three dorsal spines, on average (@Belletal2006; @Stuartetal2020).
+It then immediately began evolving adaptive reduction in its armor traits (@Huntetal2008; @Stuartetal2020) until it reached the same low-armor state previously held by lineage I.
+The observation of 93,000 years of low-armor stasis in lineage I (@Belletal1985) and the observation of rapid evolution of armor loss by lineage II, both suggest that this depositional environment lacked piscivorous fish (e.g., trout and other salmonids known to prey on modern threespine stickleback).
+Armor maintenance in extant threespine stickleback (Gasterosteus aculeatus) correlates strongly the presence of vertebrate piscivores; populations with less predation pressure typically have less armor (@Reimchen1994, @ReimchenandNosil2004; @Belletal1993; @Roestietal2023).
+Consistently, only three fossil trout have been recorded from this paleolake basin, sampled from the same rock that has revealed >15,000 threespine stickleback fossils as well as occasional killifish (Fundulus nevadensis) (@Bell2009; @Cerasonietal2024).
+Finally, tooth wear data from the lineage II sequence suggest that individuals in this lineage began eating more planktonic prey over time, expanding toward an open-water niche from the benthic, bottom-feeding niche they started with (Purnell et al.
 2007).
-If this divergence in habitat use is partitioned between sexes (@ReimchenNosil2004; @SpoljaricReimchen2008;@Reimchen2016), then we should expect sexual dimorphism to evolve for traits related to habitat use in <i>G.
-doryssus</i>.
+That this expansion into open water by lineage II coincided with armor reduction is further indicative of a limnetic system with fewer salmonid piscivores (@SchluterandMcPhail1992; @VamosiandSchluter2004; @Roestietal2023).
+In sum, we infer that lineage II likely started in a nearby paleolake basin that had predators since it was armored when it arrived (@Bell2009; @Cerasonietal2024).
+Lineage II then experienced release from predators and dietary expansion in the focal paleolake basin, generating a scenario consistent with the starting conditions for models predicting the evolution of increased sexual dimorphism following ecological release.
 
-To test this prediction, we imputed the sex of individual fossils, which is unobserved in practice, with Multiple Imputation by Chained Equations (MICE) (@MICE), using observed data from extant stickleback of known sex to train the imputation model.
-Then we tracked the multivariate evolution of sexual dimorphism over ~16,000 years for 16 traits related to swimming, feeding and defense.
+A major challenge in the study of fossilized sexual dimorphism is assigning sex to individual specimens in the first place (@HoneandMallon2017; @Mallon2017; @Saittaetal2020).
+This typically cannot be done directly, except for taxa whose sexes are distinguished by the presence or absence of sex-specific characters that preserve well.
+Instead, paleobiologists often resort to statistical detection of sex and sexual dimorphism, including tests for normality and bimodality in trait distributions (e.g., @Mallon2017), mixture modeling (e.g., @Mallon2017), divergence in growth curves (e.g., @Saittaetal2020), and emphasis on effect size statistics rather than significance testing (e.g., @Saittaetal2020).
+However, dimorphic signal can be masked by noise introduced by factors both biological and artifactual, including extended growth during ontogeny, age-based bias in survivorship, small sample sizes, time averaging, and taphonomic bias (@Godfreyetal1993; @KoscinskiandPietraszewski2004; @HoneandMallon2017; reviewed in @Mallon2017; reviewed in @Saittaetal2020).
+Thus, the best approach to sex classification is likely one of total evidence (@Saittaetal2020), including comparison to closely-related, extant species of known sex (e.g., @HoneandMallon2017; @Saittaetal2020).
+
+For our study, we inferred sex in G.
+doryssus fossils by comparing multivariate morphological trait data from lineage II samples to the same multivariate trait set collected from multiple populations in the closely-related, extant Threespine Stickleback species complex (Gasterosteus aculeatus).
+Crucially, we determined G.
+aculeatus sex directly via dissection and/or PCR genotyping.
+This enabled us to use Multiple Imputation by Chained Equations (MICE; @MICE) to build a predictive imputation model based on the multivariate morphology of G.
+aculeatus individuals of known sex.
+We applied this model to the fossil data to impute individual sex based on morphology, treating fossil sex as a missing variable.
+With sex assigned to fossil specimens (and accounting for uncertainty in those assignments), we tracked the multivariate evolution of sexual dimorphism over ~16,000 years of lineage II.
+We hypothesized that sexual dimorphism should increase following ecological release from predators.
 
@@
 ## Fossil Specimen Data
-We used Gasterosteus doryssus data that were previously reported by @Stuart2020, @Voje2022, and Siddiqui et al.
-(in review).
-Briefly, the data were collected from fossil Series K from Quarry D (Cerasoni et al., in review), dug from an open pit diatomite mine at 9.526° N, 119.094° W, near Fernley, Nevada, USA.
+We used Gasterosteus doryssus data that were previously reported by @Stuartet al2020, @Voje etal2022, and @Siddiquietal2024.
+Briefly, the data were collected from fossil Series K from Quarry D (@Cerasonietal2024), dug from an open pit diatomite mine at 9.526° N, 119.094° W, near Fernley, Nevada, USA.
 Series K consisted of 18 samples taken at ~1000-year intervals, and mean sample times span ~16,363 years.
 Fish from series K were measured for 16 ecomorphological traits related to armor, swimming, and feeding (Table 1).
-Series K started at a previously documented horizon when a low armored lineage of stickleback with zero to one dorsal spines, zero pelvic spines, and highly reduced pelvises was completely replaced by a high armored lineage of stickleback with three dorsal spines, two pelvic spines, and complete (Bell et al.
-2006; Bell 2009; Stuart et al.
-2020).
-This lineage subsequently evolved reduction in armor, body size, and traits related to swimming and feeding (Bell et al.
-2006; Stuart et al.
-2020; Siddiqui et al.
-in review).
-The tempo and mode of armor reduction during this sequence suggests adaptive evolution by natural selection (Hunt et al.
-2008), and we focus on the multivariate evolution of sexual dimorphism by this second lineage.
+Series K started at a previously documented horizon when a low armored lineage of stickleback with zero to one dorsal spines, zero pelvic spines, and highly reduced pelvises (i.e., lineage I described above) was completely replaced by a high armored lineage of stickleback with three dorsal spines, two pelvic spines, and complete pelvises (i.e., lineage II described above) (@Belletal2006; @Bell2009; @Stuartetal2020).
+Lineage II subsequently evolved reduction in armor, body size, and traits related to swimming and feeding (@Belletal2006; @Stuartetal2020; @Siddiquietal2024).
+The tempo and mode of armor reduction during this sequence suggests adaptive evolution by natural selection (@Huntetal2008), and we focus on the multivariate evolution of sexual dimorphism by this second lineage.
 
-The fossil data consists of `r sum(is.na(stick$gender))` specimens with unknown sex over 18 time periods spaced  ~1000 years apart.
-Figure 1 shows the sample size at each of the 18 time periods.
-There are at least 22 specimens at each time period with a high of 67 specimens in period 7.
+The lineage II fossil data consist of `r sum(is.na(stick$gender))` specimens of unknown sex sampled across 18 time periods spaced  ~1000 years apart.
+Figure 1 shows the sample size for each of the 18 samples.
+There are at least 22 specimens in each sample with a high of 67 specimens in sample 7.
 
-```{r fig.cap = "Fossil sampel size for each time period"}
-ggplot(aes(x = time, y = count), data = fortab) + geom_col() + xlab("Time Period")  + scale_x_continuous(breaks = 1:18, labels = 1:18)
+```{r fig.cap = "Fossil sample size for each time period"}
+ggplot(aes(x = time, y = count), data = fortab) + geom_col() + xlab("Sample")  + scale_x_continuous(breaks = 1:18, labels = 1:18)
 ```
@@
 ## Extant Specimen data
-To span the gamut of stickleback diversity for our predictive model, we sampled modern stickleback from lakes containing generalist stickleback populations (@Hendry2009; @Bolnick2011) and from lakes containing benthic-limnetic species pairs (Baumgartner et al.
-1988; @Schluter1992) (Table 2 (What is this?)).
-The generalist populations were collected by YES in 2013 and previously described in @Stuart2017.
+To span the gamut of stickleback diversity for our predictive imputation model, we sampled modern stickleback from lakes containing generalist stickleback populations (@Hendryetal2009; @Bolnick2011) and from lakes containing benthic-limnetic species pairs (@Baumgartneretal1988; @SchluterandMcPhail1992).
+The generalist populations used here were the remainder of collections made by YES in 2013 and were previously described in @Stuartetal2017 (Table S1).
 These samples were fixed in formalin, then stained for bone with Alizarin Red in 2013.
@@
 Schluter and his lab at University of British Columbia.
-They collected benthic and limnetic individuals from Enos Lake in 1988 and from Emily Lake, Little Quarry Lake, Paxton Lake, and Priest Lake in 2018.
+They collected benthic and limnetic individuals from Enos Lake in 1988 and from Emily Lake, Little Quarry Lake, Paxton Lake, and Priest Lake in 2018 (Table S1).
 The Enos specimens had been fixed whole in formalin and stored in 40% isopropanol.
@@
 In 2019, we stained these specimens for bone using Alizarin Red.
-
 We next replicated fossil data collection (Table 1) on these extant specimens.
@@
 We also counted vertebrae from the X-rays: abdominal vertebrae were counted anterior to the first vertebra with a haemal spine contacting an anal fin pterygiophore.
-Caudal vertebrae were posterior, including the first vertebra with the haemal spine contacting the anal fin pterygiophore (following @Aguirre2014.
+Caudal vertebrae were posterior, including the first vertebra with the haemal spine contacting the anal fin pterygiophore (following @Aguirreetal2014).
 X-rays were taken with an AXR Hot Shot X-ray Machine (Associated X-ray Corporation) at the Field Museum of Natural History.
@@
 We developed the film and scanned individual images of each fish using the B&W Negatives setting on an Epson Perfection 4990 Photo flatbed at 2400 dpi.
-Measurements from photographs and X-rays were taken with FIJI (Schindelin et al.
-2012) and its plugin ObjectJ (https://sils.fnwi.uva.nl/bcb/objectj/).
-All photographs, X-rays, and ObjectJ files have been uploaded to Morphosource.org (accession # TBD).
-We dissected individuals from the generalist populations to determine sex from the gonads.
-Individuals from the species-pair lakes were sexed by Schluter and his lab, using a genotyping protocol (confirm and cite).
+Measurements from photographs and X-rays were taken with FIJI (@Schindelinetal2012) and its plugin ObjectJ (https://sils.fnwi.uva.nl/bcb/objectj/).
 
-The extant data used here consists of a total of `r sum(!is.na(stick$gender))` specimens all with known sex.
+We dissected individuals from the generalist populations (Table S1) to determine sex from the gonads.
+Individuals from the species-pair lakes (TableS1) were previously sexed by the Schluter lab, using either dissection or a genotyping protocol (Whom, personal communication).
+
+The extant data used here consists of a total of `r sum(!is.na(stick$gender))` specimens all with known sex (Table S1).
 Of these, there are `r sum(stick$gender == "Female",na.rm = TRUE)` and `r sum(stick$gender == "Male",na.rm = TRUE)` female and male specimens, respectively.
 
-## Outlier analysis. 
-To check for outliers, we calculated within-group means and standard deviations for each trait separately for K series fossil specimens (pooled across samples) and for extant specimens (pooled across lakes).
-We noted trait values greater than 3.5 standard deviations from the mean as potential outliers.
+## Outlier analysis and size correction.
+
+To check for outliers, we calculated within-group means and standard deviations for each trait separately for K series fossil specimens (pooled across samples) and for extant specimens (within generalist, benthic, or limnetic categories).
+We noted trait values greater than 3.0 standard deviations from the mean as potential outliers.
+We deemed 3.0 s.d.
+to be a reasonable threshold for detecting errors without excluding biologically relevant values.
 We checked whether these potential outliers were a result of data entry and collection error and corrected them if they were.
 We turned the remaining outlier trait values to NAs.
-(confirm)  (Wait, what?
-Why?
-Anything that was 3.5 SD above the mean was recorded as missing?
-How do you justify this?
-)
+We size-corrected continuous traits only, as they varied with size, unlike count traits that are fixed during early development.
+We regressed each trait on standard length using a mixed-model regression, pooling all specimens, following @Stuartetal2017.
+We appended the size corrected data to the uncorrected trait data frame and used all of our data to build the imputation model described next.
 
-Missing data imputation, including fossil sex
-
-Quantification and evolution of sexual dimorphism
-
-What covariates do we have in the data: length, what else,
+## Missing data imputation, including fossil sex
+We used multiple imputation (@little2002statistical) to impute the sex of the fossils many times.
+Imputations were performed using the MICE (@MICE) algorithm and implemented in R (@R2022language) with M = 100 imputations performed.
+After imputing the sex of the fossil data, each completed data set is then used to fit a modified Ornstein–Uhlenbeck (OU) (@OUProcess) model using a Bayesian framework to look for sexual dimorphism across a variety of stickleback phenotypes.
 

### 2024-07-24 manuscript/paper.tex
Update Paper
Yoel adding text to introduction, updating and formatting references, etc.

@@
 \author{\large Allison Ozark\(^{1,*}\), Matthew Stuart\(^{2,3,*}\),
-Raheyma Siddiqui\(^1\), Akhil Ghosh\(^2\)\\
+Raheyma Siddiqui\(^1\), Akhil Ghosh\(^{2,3}\)\\
 \large Samantha Swank\(^1\), Michael A. Bell\(^4\), Gregory J.
@@
 \endhead
-Standard Length & stl & Distance from anterior tip of premaxilla to
-posterior end of last vertebra (hypural plate) \\
-Dorsal Spine & mds & Number of dorsal spines from 0 to 3 \\
-Dorsal Fin Ray & mdf & Number of bones in the dorsal fin posterior to
-the third dorsal spine (i.e., soft dorsal fin rays) \\
-Anal Fin Ray & maf & Number of bones in the anal fin posterior to the
-anal spine (i.e., soft dorsal fin rays) \\
+Standard Length & \texttt{stl} & Distance from anterior tip of
+premaxilla to posterior end of last vertebra (hypural plate) \\
+Dorsal Spine & \texttt{mds} & Number of dorsal spines from 0 to 3 \\
+Dorsal Fin Ray & \texttt{mdf} & Number of bones in the dorsal fin
+posterior to the third dorsal spine (i.e., soft dorsal fin rays) \\
+Anal Fin Ray & \texttt{maf} & Number of bones in the anal fin posterior
+to the anal spine (i.e., soft dorsal fin rays) \\
 Abdominal Vertebra & mav & Number of vertebrae anterior to the first
 vertebra contacting an anal fin pterygiophore (Aguirre et al.~2014) \\
-Caudal Vertebra & mcv & Number of vertebrae posterior to and including
-the first vertebra contacting an anal fin pterygiophore (Aguirre et
-al.~2014) \\
-Pterygiophore number & mpt & Number of pterygiophores anterior to but
-excluding the pterygiophore under the third dorsal spine, which is
-immediately anterior to and contiguous with the dorsal fin \\
-Pelvic Spine length & lps.sc & Length from the base of one pelvic spine
-above its articulation with the pelvic girdle to its distal tip \\
-Ectocoracoid & ect.sc & Length between the anterior and posterior tips
-of the shoulder girdle base (i.e., ectocoracoid) \\
-Pelvic Girdle & tpg.sc & Length between the anterior to posterior tips
-along midline. If vestigial, the sum of longest anterior-posterior axis
-for the vestiges \\
-Cleithrum length & cle.sc & Length from free dorsal tip to ventral tip
-of the cleithrum on the anterior margin of the shoulder girdle (i.e.,
-cleithrum) \\
-Premaxilla & pmx.sc & Length from the anterior tip of the premaxilla to
-the distal tip of the ascending process of the premaxilla \\
-Dorsal Spine & Ds\#.sc\# = 1,2,or 3 & Length from the base of a dorsal
-spine above the pterygiophore to its distal tip along the anterior
-edge \\
-Pterygiophore & lpt.sc & Distance between the anterior to posterior tips
-of the pterygiophore immediately preceding the 3rd dorsal spine (when
-present) \\
+Caudal Vertebra & \texttt{mcv} & Number of vertebrae posterior to and
+including the first vertebra contacting an anal fin pterygiophore
+(Aguirre et al.~2014) \\
+Pterygiophore number & \texttt{mpt} & Number of pterygiophores anterior
+to but excluding the pterygiophore under the third dorsal spine, which
+is immediately anterior to and contiguous with the dorsal fin \\
+Pelvic Spine length & \texttt{lps.sc} & Length from the base of one
+pelvic spine above its articulation with the pelvic girdle to its distal
+tip \\
+Ectocoracoid & \texttt{ect.sc} & Length between the anterior and
+posterior tips of the shoulder girdle base (i.e., ectocoracoid) \\
+Pelvic Girdle & \texttt{tpg.sc} & Length between the anterior to
+posterior tips along midline. If vestigial, the sum of longest
+anterior-posterior axis for the vestiges \\
+Cleithrum length & \texttt{cle.sc} & Length from free dorsal tip to
+ventral tip of the cleithrum on the anterior margin of the shoulder
+girdle (i.e., cleithrum) \\
+Premaxilla & \texttt{pmx.sc} & Length from the anterior tip of the
+premaxilla to the distal tip of the ascending process of the
+premaxilla \\
+Dorsal Spine & \texttt{Ds\#.sc\#\ =\ 1,2,or\ 3} & Length from the base
+of a dorsal spine above the pterygiophore to its distal tip along the
+anterior edge \\
+Pterygiophore & \texttt{lpt.sc} & Distance between the anterior to
+posterior tips of the pterygiophore immediately preceding the 3rd dorsal
+spine (when present) \\
 \bottomrule()
@@
 \(M = 10\). However, Zhou and Reiter
-(\protect\hyperlink{ref-ZhouReiter2009}{2010}) recommend a larger number
+(\protect\hyperlink{ref-ZhouReiter2010}{2010}) recommend a larger number
 of imputed data sets if the data users intend on performing Bayesians
@@
 datasets. In addition to this, Zhou and Reiter
-(\protect\hyperlink{ref-ZhouReiter2009}{2010}) suggests rather than
+(\protect\hyperlink{ref-ZhouReiter2010}{2010}) suggests rather than
 using Rubin's combining rules to combine across imputed data sets,
@@
 
-In addition, we assume that the discrete traits abdominal vertebrae
-(mav) and caudal vertebrae (mcv) also have correlation with the standard
-length of the fish (\textbf{CITATION}). If \(Y_{k,ij}\) is one of the
-above traits, then \begin{align}
-{Y}_{k,ij} & \sim \left\{\begin{array}{ll} Poisson(\exp\{\mu_{k,ft_i} + \gamma_kY_{K,ij}\}), & W_{ij} = \text{Female} \\ Poisson(\exp\{\mu_{k,mt_i} + \gamma_kY_{K,ij}\}), & W_{ij} = \text{Male} \end{array}\right..
+If \(Y_{k,ij}\) is a discrete trait, the conventional method of
+modelling this data is by fitting a Poisson distribution. However, the
+Poisson distribution assumes that the mean and variance of \(Y_{k,ij}\)
+are equal (\(E(Y_{k,ij}) = \text{Var}(Y_{k,ij})\)), while the empirical
+fossil data may have variances that are smaller than their respective
+means. To combat this issue, we propose to fit the discrete traits to a
+generalized Poisson model as defined in
+(\protect\hyperlink{ref-GeneralizedPoisson}{\textbf{GeneralizedPoisson?}}).
+Specifically, if \(X \sim GP(\lambda,\alpha)\), then \begin{align}
+P(X = x) = \left\{\begin{array}{cc} \frac{(1 - \alpha)\lambda[(1 - \alpha)\lambda + \alpha x]^{x - 1} \exp\left\{-((1 - \alpha)\lambda  + \alpha x)\right\}}{x!} & (1 - \alpha)\lambda  + \alpha x \geq 0  \\ 0 & (1 - \alpha)\lambda  + \alpha x < 0 \end{array}\right.,
+\label{eq:GP_pmf}
+\end{align} where \(E(X) = \lambda\) and
+\(Var(X) = \frac{\lambda}{(1 - \alpha)^2}\). If \(\alpha > 0\), then the
+variance is greater than the mean, called overdispersion; if
+\(\alpha < 0\), then the variance is greater than the mean, called
+underdispersion; if \(\alpha = 0\), then the model degenerates to a
+Poisson distribution.
+
+We will use the generalized Poisson to model all of the discrete traits
+in our dataset. In addition, we assume that the discrete traits
+abdominal vertebrae (mav) and caudal vertebrae (mcv) also have
+correlation with the standard length of the fish (\textbf{CITATION}). If
+\(Y_{k,ij}\) is one of the above traits, then \begin{align}
+{Y}_{k,ij} & \sim \left\{\begin{array}{ll} GP(\exp\{\mu_{k,ft_i} + \gamma_kY_{K,ij}\},\alpha_k), & W_{ij} = \text{Female} \\ GP(\exp\{\mu_{k,mt_i} + \gamma_kY_{K,ij}\},\alpha_k), & W_{ij} = \text{Male} \end{array}\right..
 \label{eq:disc_corr}
@@
 the standard deviation of the OU process, and \(W_t\) is a standard
-Brownian motion. As shown in (Uhlenbeck and Ornstein
-(\protect\hyperlink{ref-OUProcess}{1930})), the closed form solution for
+Brownian motion. As shown in Uhlenbeck and Ornstein
+(\protect\hyperlink{ref-OUProcess}{1930}), the closed form solution for
 the SDE in (\ref{eq:OU_cont}) is \begin{align}
@@
 Because we are fitting a dataset with a stochastic structure on the
-means of the phenotypes, we analyze the data via a Bayesian analysis.
-Bayesian data analysis is also more naturally used when we have to
-impute data (\textbf{CITATION}).
-
-Priors: For \(k = 1,\cdots,K\),
+means of the phenotypes, and we are interested in the distribution of
+the overall structure of these means, we analyze the data via a Bayesian
+analysis. Bayesian data analysis is also more naturally used when we
+have to impute data (\textbf{CITATION}). To aid in the sampling
+procedure for the discrete phenotypes, we declare
+
+Priors: To aid in the sampling procedure for the discrete phenotypes, we
+declare
+\(\phi_k = \log\left(\frac{\alpha_k - \max_{i,j}(-\lambda_{k,ij}/y_{k,ij})}{1 - \alpha_k}\right)\)
+where
+\(\lambda_{k,ij} = \left\{\begin{array}{cc} \exp(\mu_{k,ft_i} + \gamma_k Y_{K,ij}) & W_{ij} = \text{Female} \\ \exp(\mu_{k,mt_i} + \gamma_k Y_{K,ij}) & W_{ij} = \text{Male} \end{array}\right.\).
+This transformation is performed to ensure our algorithm can properly
+sample from the posterior distribution of interest.
+
+For \(k = 1,\cdots,K\),
 
@@
 \sigma_k & \overset{iid}{\sim}\mathcal{N}(0,10)I_{\{\sigma > 0\}} \nonumber \\
+\sigma_k & \overset{iid}{\sim}\mathcal{N}(0,10) \nonumber \\
 \tau_k & \overset{iid}{\sim}\mathcal{N}(0,10)I_{\{\tau > 0\}} \nonumber \\
@@
 (\ref{eq:disc_corr}) and (\ref{eq:disc_ind}), there is no \(\sigma_k\),
-and thus is not sampled for those phenotypes.
+and is not sampled, and for the continuous phenotypes in equation
+(\ref{eq:cont}), there is no \(\phi_k\) and is not sampled.
 
@@
 
-Bayesian Analysis after multiple imputation Zhou and Reiter
-(\protect\hyperlink{ref-ZhouReiter2009}{2010}): They recommend using a
-large number of imputations. 5 or 10 is too small. We are using M = 100.
-
 \hypertarget{sec:results}{%
 \section{Results}\label{sec:results}}
+
+As previously described, four models for the mean process were
+considered for each trait. Within each trait, the model with the lowest
+DIC was chosen for that train. All subsequent results presented here use
+the model with the lowest DIC individually by trait. Table
+\ref{whichmodel} shows the model that was selected for each trait. About
+half of the traits had models with a trend, and most models did not
+include an OU component. The traits that did include an OU component
+were caudal vertebra (mcv), pelvic spine lentgh (lpt), abdominal
+vertebra (mav), and standard length (stl) with caudal vertebra (mcv)
+being the only model with both a trend and an OU component. This implies
+that not only does caudal vertebra size have a linear trend across
+gender, but the average size at one time period is highly correlated
+with the average size at the next time period.
+
+\hypertarget{best-fitting-models}{%
+\subsection{Best fitting models}\label{best-fitting-models}}
+
+\begin{table}[ht]
+\centering
+\begin{tabular}{|c||c|c|}
+  \hline
+ & OU & No OU  \\ 
+  \hline
+  \hline
+Trend  &  mcv  & ds2, ds3, ect, maf, mdf \\ 
+  \hline
+  No Trend & lpt, mav, stl  &  cle, ds1, lps, pmx\\
+   \hline
+\end{tabular}
+\label{whichmodel}
+\end{table}
+
+\hypertarget{sexual-dimorphism}{%
+\subsection{Sexual Dimorphism}\label{sexual-dimorphism}}
+
+Table \ref{tab:prob0-table} shows the posterior probability of the
+differences in the means between male and female specimens at each time
+point. Probabilities near 0.5 indicate very little difference in the
+means whereas probabilities far from 0.5 indicate sexual dimorphism with
+values of 0 and 1 indicating larger mean values for females and males,
+respectively. These results are shown graphically in figure
+\ref{fig:prob0-plot}.
+
+Figure \ref{fig:mu-diff-plot} shows the posterior mean difference for
+males versus females (values above 0 indicate means that were greater
+for males versus females). The full poosterior distributions over time
+for all traits are shown in the appendix.
+
+Traits such as standard length (stl), anal fin ray (maf), dorsal fin ray
+(mdf), premaxilla (pmx), and dorsal spine 1 (ds1) demonstrate constant
+sexual dimorphism with all but the last having learger mean values in
+male sepcimen. Other traits such as abdominal vertebra (mav), caudal
+vertebra (mcv), and pelvic spine length (lps) don't appear to have any
+discernable sexual dimorphism in the means. Dorsal spine 3 (ds3)
+exhibits sexual dimorphism that changes over time. The mean number of
+dorsal spine 3 begins larger for females for the earliest observed time
+periods. Over time, however, this switches and the male specimens
+exhibit larger mean values for the latest observed time periods.
+
+\hypertarget{changes-over-time}{%
+\subsection{Changes over time}\label{changes-over-time}}
+
+Table \ref{tab:auc-table} shows the probability that the posterior
+distribution of the differences in the mean is greater than the
+posterior distribution of the differences at the first time point.
+Values near 0.5 indicate very little difference between the
+distributions whereas values closer to 0 and 1 indicate changes in the
+distributions with values of 1 meaning the difference in means has
+shifted towards males having a larger mean and values of 0 indicating
+the difference in means has shifted towards larger means in females.
+These results are shown graphically in figure \ref{fig:AUC-plot}.
+
+The amount of sexual dimporphism over time does not change substantially
+for a majority of the trait observed here (i.e.~cle, ds1, ds2, ect, lps,
+lpt, mav, mcv, pmx, stl). Traits that trend away from the mean
+difference at the first observed time point include dorsal spine 3
+(ds3), anal fin ray (maf), and dorsal fin ray (mdf).
+
+The posterior mean of the mean number of dorsal spine rays at the first
+observed time point is 2.29 for males and for females it is 2.45. At the
+last observed time point, the average number for males and female both
+dropped to 1.265 and 1.185 for males and females, respectively. This
+resulted in a mean number of dorsal spine 3 lost by males of 1.027 and
+for females it was 1.264. This resulted in a change of the mean
+difference from time point 1 to time point 18 of 0.236. Basically, males
+and females both lowered the mean number of dorsal spine 3, but females
+lowered their mean more than the males.
+
+We also observed a change in the differences between males and females
+for the number of bones int he anal fin ray (maf) and the number of
+bones in the dorsal fin ray. The difference in the means of males versus
+females for anal fin ray bones changed by 0.152 over the course of the
+observed time periods. This change was driven mostly by the change in
+females. At the first observed time period the had on average 8.291
+bones dropping to 8.170 over the 18 times periods observed. Whereas for
+males the change in means rose only from 8.518 to 8.549, about 0.0316
+bones on average.
+
+A similar change in the difference of the means over time was observed
+for the number of bones in the dorsal fin (mdf). However, unlike maf,
+this change in sexual dimporhism was driven entirely by the males. The
+mean change for females over the 18 observed time periods was only
+0.0055 with the average number of dorsal fin bones for females starting
+at 9.287 and the final observed mean of 9.281 resulting in almost no
+change in the mean. For males, the average number of dorsal fin bones at
+the first time point was about 9.568 dropping to 9.734 over the course
+of the observed time periods for a gain of 0.166 bones on average.
+
+\begin{longtable}[]{@{}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 26\tabcolsep) * \real{0.0602}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 26\tabcolsep) * \real{0.0723}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 26\tabcolsep) * \real{0.0723}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 26\tabcolsep) * \real{0.0723}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 26\tabcolsep) * \real{0.0723}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 26\tabcolsep) * \real{0.0723}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 26\tabcolsep) * \real{0.0723}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 26\tabcolsep) * \real{0.0723}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 26\tabcolsep) * \real{0.0723}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 26\tabcolsep) * \real{0.0723}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 26\tabcolsep) * \real{0.0723}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 26\tabcolsep) * \real{0.0723}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 26\tabcolsep) * \real{0.0723}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 26\tabcolsep) * \real{0.0723}}@{}}
+\toprule()
+\begin{minipage}[b]{\linewidth}\raggedleft
+time
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+cle
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+ds1
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+ds2
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+ds3
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+ect
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+lps
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+lpt
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+maf
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+mav
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+mcv
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+mdf
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+pmx
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+stl
+\end{minipage} \\
+\midrule()
+\endhead
+1 & 0.913 & 0.215 & 0.359 & 0.203 & 0.865 & 0.428 & 0.853 & 0.944 &
+0.391 & 0.507 & 0.959 & 0.948 & 0.883 \\
+2 & 0.942 & 0.212 & 0.325 & 0.180 & 0.891 & 0.404 & 0.853 & 0.948 &
+0.520 & 0.520 & 0.963 & 0.967 & 0.908 \\
+3 & 0.843 & 0.207 & 0.249 & 0.153 & 0.770 & 0.279 & 0.803 & 0.965 &
+0.289 & 0.525 & 0.977 & 0.912 & 0.777 \\
+4 & 0.890 & 0.210 & 0.249 & 0.201 & 0.830 & 0.372 & 0.741 & 0.977 &
+0.435 & 0.589 & 0.986 & 0.935 & 0.831 \\
+5 & 0.954 & 0.211 & 0.211 & 0.195 & 0.909 & 0.377 & 0.756 & 0.985 &
+0.533 & 0.622 & 0.992 & 0.975 & 0.921 \\
+6 & 0.885 & 0.211 & 0.134 & 0.171 & 0.820 & 0.291 & 0.686 & 0.991 &
+0.645 & 0.456 & 0.996 & 0.949 & 0.811 \\
+7 & 0.812 & 0.211 & 0.124 & 0.239 & 0.760 & 0.340 & 0.617 & 0.994 &
+0.224 & 0.717 & 0.999 & 0.879 & 0.747 \\
+8 & 0.928 & 0.210 & 0.135 & 0.258 & 0.885 & 0.387 & 0.706 & 0.994 &
+0.342 & 0.730 & 1.000 & 0.969 & 0.872 \\
+9 & 0.899 & 0.213 & 0.145 & 0.304 & 0.857 & 0.406 & 0.715 & 0.995 &
+0.521 & 0.756 & 1.000 & 0.940 & 0.845 \\
+10 & 0.954 & 0.217 & 0.235 & 0.555 & 0.940 & 0.571 & 0.810 & 0.997 &
+0.454 & 0.719 & 1.000 & 0.976 & 0.935 \\
+11 & 0.913 & 0.211 & 0.162 & 0.476 & 0.875 & 0.388 & 0.703 & 0.997 &
+0.363 & 0.657 & 1.000 & 0.957 & 0.864 \\
+12 & 0.949 & 0.221 & 0.262 & 0.726 & 0.929 & 0.585 & 0.803 & 0.997 &
+0.462 & 0.595 & 1.000 & 0.968 & 0.926 \\
+13 & 0.904 & 0.220 & 0.260 & 0.789 & 0.885 & 0.598 & 0.795 & 0.998 &
+0.470 & 0.370 & 1.000 & 0.929 & 0.872 \\
+14 & 0.925 & 0.220 & 0.280 & 0.883 & 0.906 & 0.659 & 0.854 & 0.997 &
+0.526 & 0.472 & 1.000 & 0.958 & 0.898 \\
+15 & 0.880 & 0.225 & 0.283 & 0.839 & 0.849 & 0.617 & 0.796 & 0.997 &
+0.490 & 0.398 & 1.000 & 0.919 & 0.834 \\
+16 & 0.953 & 0.216 & 0.276 & 0.931 & 0.931 & 0.614 & 0.831 & 0.997 &
+0.414 & 0.541 & 1.000 & 0.975 & 0.925 \\
+17 & 0.803 & 0.219 & 0.232 & 0.832 & 0.765 & 0.383 & 0.655 & 0.996 &
+0.569 & 0.758 & 1.000 & 0.854 & 0.751 \\
+18 & 0.780 & 0.212 & 0.225 & 0.817 & 0.726 & 0.297 & 0.558 & 0.995 &
+0.705 & 0.722 & 0.999 & 0.859 & 0.717 \\
+\bottomrule()
+\end{longtable}
+
+\begin{longtable}[]{@{}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 26\tabcolsep) * \real{0.0602}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 26\tabcolsep) * \real{0.0723}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 26\tabcolsep) * \real{0.0723}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 26\tabcolsep) * \real{0.0723}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 26\tabcolsep) * \real{0.0723}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 26\tabcolsep) * \real{0.0723}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 26\tabcolsep) * \real{0.0723}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 26\tabcolsep) * \real{0.0723}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 26\tabcolsep) * \real{0.0723}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 26\tabcolsep) * \real{0.0723}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 26\tabcolsep) * \real{0.0723}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 26\tabcolsep) * \real{0.0723}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 26\tabcolsep) * \real{0.0723}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 26\tabcolsep) * \real{0.0723}}@{}}
+\toprule()
+\begin{minipage}[b]{\linewidth}\raggedleft
+time
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+cle
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+ds1
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+ds2
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+ds3
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+ect
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+lps
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+lpt
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+maf
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+mav
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+mcv
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+mdf
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+pmx
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+stl
+\end{minipage} \\
+\midrule()
+\endhead
+1 & 0.500 & 0.500 & 0.500 & 0.500 & 0.500 & 0.500 & 0.500 & 0.500 &
+0.500 & 0.500 & 0.500 & 0.500 & 0.500 \\
+2 & 0.514 & 0.504 & 0.509 & 0.495 & 0.512 & 0.508 & 0.517 & 0.495 &
+0.396 & 0.490 & 0.495 & 0.511 & 0.513 \\
+3 & 0.636 & 0.514 & 0.559 & 0.516 & 0.628 & 0.601 & 0.602 & 0.475 &
+0.523 & 0.480 & 0.476 & 0.626 & 0.641 \\
+4 & 0.563 & 0.509 & 0.544 & 0.459 & 0.558 & 0.539 & 0.672 & 0.456 &
+0.436 & 0.450 & 0.458 & 0.556 & 0.562 \\
+5 & 0.543 & 0.507 & 0.551 & 0.417 & 0.539 & 0.522 & 0.687 & 0.436 &
+0.389 & 0.425 & 0.439 & 0.537 & 0.544 \\
+6 & 0.618 & 0.512 & 0.589 & 0.415 & 0.613 & 0.579 & 0.716 & 0.415 &
+0.317 & 0.522 & 0.419 & 0.612 & 0.620 \\
+7 & 0.591 & 0.510 & 0.601 & 0.364 & 0.589 & 0.564 & 0.712 & 0.379 &
+0.546 & 0.382 & 0.383 & 0.588 & 0.592 \\
+8 & 0.539 & 0.507 & 0.579 & 0.324 & 0.536 & 0.514 & 0.683 & 0.373 &
+0.478 & 0.368 & 0.378 & 0.537 & 0.540 \\
+9 & 0.524 & 0.506 & 0.586 & 0.299 & 0.521 & 0.510 & 0.636 & 0.355 &
+0.408 & 0.340 & 0.359 & 0.523 & 0.522 \\
+10 & 0.379 & 0.497 & 0.544 & 0.203 & 0.377 & 0.400 & 0.578 & 0.321 &
+0.438 & 0.359 & 0.324 & 0.385 & 0.373 \\
+11 & 0.531 & 0.508 & 0.605 & 0.226 & 0.528 & 0.517 & 0.685 & 0.301 &
+0.528 & 0.403 & 0.303 & 0.530 & 0.533 \\
+12 & 0.363 & 0.492 & 0.547 & 0.150 & 0.360 & 0.386 & 0.566 & 0.284 &
+0.439 & 0.436 & 0.284 & 0.369 & 0.358 \\
+13 & 0.373 & 0.495 & 0.555 & 0.133 & 0.372 & 0.386 & 0.561 & 0.269 &
+0.432 & 0.575 & 0.268 & 0.378 & 0.371 \\
+14 & 0.326 & 0.496 & 0.548 & 0.102 & 0.326 & 0.349 & 0.514 & 0.255 &
+0.399 & 0.514 & 0.253 & 0.334 & 0.323 \\
+15 & 0.365 & 0.492 & 0.552 & 0.107 & 0.365 & 0.373 & 0.502 & 0.250 &
+0.433 & 0.554 & 0.247 & 0.368 & 0.367 \\
+16 & 0.361 & 0.497 & 0.574 & 0.084 & 0.360 & 0.381 & 0.570 & 0.234 &
+0.462 & 0.463 & 0.229 & 0.369 & 0.356 \\
+17 & 0.561 & 0.504 & 0.629 & 0.116 & 0.559 & 0.542 & 0.669 & 0.226 &
+0.364 & 0.313 & 0.219 & 0.559 & 0.565 \\
+18 & 0.625 & 0.512 & 0.648 & 0.121 & 0.617 & 0.596 & 0.735 & 0.220 &
+0.281 & 0.356 & 0.212 & 0.622 & 0.627 \\
+\bottomrule()
+\end{longtable}
+
+\begin{figure}
+
+{\centering \includegraphics{paper_files/figure-latex/mu-diff-plot-1}
+
+}
+
+\caption{Posterior mean difference between males and females}\label{fig:mu-diff-plot}
+\end{figure}
+
+\begin{figure}
+
+{\centering \includegraphics{paper_files/figure-latex/prob0-plot-1}
+
+}
+
+\caption{Posterior probabilities that mean for males is greater than mean for females}\label{fig:prob0-plot}
+\end{figure}
+
+\begin{figure}
+
+{\centering \includegraphics{paper_files/figure-latex/AUC-plot-1}
+
+}
+
+\caption{AUC comparing time x versus time 0.}\label{fig:AUC-plot}
+\end{figure}
 
@@
 
+DIC results here.
+
+\hypertarget{figures}{%
+\subsection{Figures}\label{figures}}
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.9\linewidth]{../Figures/cle/mu_diff}
+
+}
+
+\caption{cle}\label{fig:unnamed-chunk-7}
+\end{figure}
+\begin{figure}
+
+{\centering \includegraphics[width=0.9\linewidth]{../Figures/ds1/mu_diff}
+
+}
+
+\caption{ds1}\label{fig:unnamed-chunk-8}
+\end{figure}
+\begin{figure}
+
+{\centering \includegraphics[width=0.9\linewidth]{../Figures/ds2/mu_diff}
+
+}
+
+\caption{ds2}\label{fig:unnamed-chunk-9}
+\end{figure}
+\begin{figure}
+
+{\centering \includegraphics[width=0.9\linewidth]{../Figures/ds3/mu_diff}
+
+}
+
+\caption{ds3}\label{fig:unnamed-chunk-10}
+\end{figure}
+\begin{figure}
+
+{\centering \includegraphics[width=0.9\linewidth]{../Figures/ect/mu_diff}
+
+}
+
+\caption{ect}\label{fig:unnamed-chunk-11}
+\end{figure}
+\begin{figure}
+
+{\centering \includegraphics[width=0.9\linewidth]{../Figures/lps/mu_diff}
+
+}
+
+\caption{lps}\label{fig:unnamed-chunk-12}
+\end{figure}
+\begin{figure}
+
+{\centering \includegraphics[width=0.9\linewidth]{../Figures/lpt/mu_diff}
+
+}
+
+\caption{lpt}\label{fig:unnamed-chunk-13}
+\end{figure}
+\begin{figure}
+
+{\centering \includegraphics[width=0.9\linewidth]{../Figures/maf/lambda_diff}
+
+}
+
+\caption{maf}\label{fig:unnamed-chunk-14}
+\end{figure}
+\begin{figure}
+
+{\centering \includegraphics[width=0.9\linewidth]{../Figures/mav/lambda_diff}
+
+}
+
+\caption{mav}\label{fig:unnamed-chunk-15}
+\end{figure}
+\begin{figure}
+
+{\centering \includegraphics[width=0.9\linewidth]{../Figures/mcv/lambda_diff}
+
+}
+
+\caption{mcv}\label{fig:unnamed-chunk-16}
+\end{figure}
+\begin{figure}
+
+{\centering \includegraphics[width=0.9\linewidth]{../Figures/mdf/lambda_diff}
+
+}
+
+\caption{mdf}\label{fig:unnamed-chunk-17}
+\end{figure}
+\begin{figure}
+
+{\centering \includegraphics[width=0.9\linewidth]{../Figures/mds/lambda_diff}
+
+}
+
+\caption{mds}\label{fig:unnamed-chunk-18}
+\end{figure}
+\begin{figure}
+
+{\centering \includegraphics[width=0.9\linewidth]{../Figures/mpt/lambda_diff}
+
+}
+
+\caption{mpt}\label{fig:unnamed-chunk-19}
+\end{figure}
+\begin{figure}
+
+{\centering \includegraphics[width=0.9\linewidth]{../Figures/pmx/mu_diff}
+
+}
+
+\caption{pmx}\label{fig:unnamed-chunk-20}
+\end{figure}
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.9\linewidth]{../Figures/stl/mu_diff}
+
+}
+
+\caption{stl}\label{fig:unnamed-chunk-21}
+\end{figure}
+
 \hypertarget{references}{%
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-ZhouReiter2009}{}}%
+\leavevmode\vadjust pre{\hypertarget{ref-ZhouReiter2010}{}}%
 Zhou, X., and J. Reiter. 2010. {``A Note on Bayesian Inference After

### 2024-07-25 manuscript/paper.Rmd
manuscript additions

@@
 This is predicted to result in intersexual divergence in habitat use and associated phenotypes (@Schoener1968; @Shine1989; @BolnickandDoebeli2003; @Butleretal2007; @BolnickandLau2008; @Cooperetal2011; but see @Stuartetal2021; @Blain2022).
-Such competition-driven “intraspecific character displacement” between the sexes is therefore one explanation for the evolution of sexual dimorphism (@PfennigandPfennig2012; @DeLisleandRowe2015; @DeLisleandRowe2017, @De Lisleetal2018).
+Such competition-driven “intraspecific character displacement” between the sexes is therefore one explanation for the evolution of sexual dimorphism (@PfennigandPfennig2012; @DeLisleandRowe2015; @DeLisleandRowe2017, @DeLisleetal2018).
 
@@
 aculeatus sex directly via dissection and/or PCR genotyping.
-This enabled us to use Multiple Imputation by Chained Equations (MICE; @MICE) to build a predictive imputation model based on the multivariate morphology of G.
+This enabled us to use Multiple Imputation by Chained Equations (MICE; @MICE) to build a predictive multiple imputation algorithm (@little2002statistical) based on the multivariate morphology of G.
 aculeatus individuals of known sex.
-We applied this model to the fossil data to impute individual sex based on morphology, treating fossil sex as a missing variable.
-With sex assigned to fossil specimens (and accounting for uncertainty in those assignments), we tracked the multivariate evolution of sexual dimorphism over ~16,000 years of lineage II.
+We applied this algorithm to the fossil data to impute individual sex based on morphology, treating fossil sex as a missing variable.
+With sex assigned to fossil specimens (and accounting for uncertainty in those assignments), we fit a modified Ornstein-Uhlenbeck (OU) (@OUProcess) model using a Bayesian framework to test for evolution of sexual dimorphism in each trait over ~16,000 years of lineage II.
 We hypothesized that sexual dimorphism should increase following ecological release from predators.
-
-# Methods and Materials 
-<!-- Studying sexual dimorphism of phenotypes in modern species is a straightforward endeavor: measure a phenotype of interest and statistically test for differences between the sexes.
-However, examining sexual dimorphism in fossils is complicated substantially by the issue that sex may not be able to be directly observed from fossil specimens.
-For this study, we have two sources of data: 1) Modern stickleback specimens with observed sex and 2) fossil specimens with sex unobserved.
--->
-
-A major challenge in this work is inferring the sex of fossil stickleback specimens.
-Indeed, for most species, sex cannot be detected directly, except for lineages whose sexes are distinguished by the presence or absence of sex-specific characters (@Hone2017; @SaittaEtAl2020).
-Instead, paleobiologists resort to statistical detection of sex and sexual dimorphism, including tests for bimodality in trait distributions (e.g., @Hone2017) and divergence in growth curves (e.g., @SaittaEtAl2020).
-We used a third approach to infer fossil sex and study the evolution of sexual dimorphism by taking advantage of the fact that G.
-doryssus is part of the extant threespine stickleback species complex, Gasterosteus aculeatus.
-We measured the same trait set from a phenotypically diverse set of extant G.
-aculeatus populations, for whom individual sex was known.
-We then combined fossil and extant data sets, treated fossil sex as missing data, and used the MICE (@MICE) multiple imputation (@little2002statistical) algorithm, and implemented in R (@R2022language), to impute individual sex of the fossil data set 100 times.
-We then fit a modified Ornstein–Uhlenbeck (OU) (@OUProcess) model using a Bayesian framework to the data to test for the evolution of sexual dimorphism in each trait.
 
@@
 ## Fossil Specimen Data
-We used Gasterosteus doryssus data that were previously reported by @Stuartet al2020, @Voje etal2022, and @Siddiquietal2024.
+We used Gasterosteus doryssus data that were previously reported by @Stuartetal2020, @Vojeetal2022, and @Siddiquietal2024.
 Briefly, the data were collected from fossil Series K from Quarry D (@Cerasonietal2024), dug from an open pit diatomite mine at 9.526° N, 119.094° W, near Fernley, Nevada, USA.
@@
 ## Imputation Model 
-Let $\boldsymbol{W}$ be an $(n_{extant} + n_{fossil}) \times 1$ vector of the covariate gender of the stickleback fish and $\boldsymbol{Y}$ be an $(n_{extant} + n_{fossil}) \times K$ matrix of the $K$ phenotypes of interest.
-Because the gender of the fossilized stickleback fish is unobservable, we further define $\boldsymbol{W} = (\boldsymbol{W}_{extant}^T,\boldsymbol{W}_{fossil}^T)^T$ where $\boldsymbol{W}_{extant}$ and $\boldsymbol{W}_{fossil}$ are the $n_{extant} \times 1$ and $n_{fossil} \times 1$ vectors of the observed extant gender and missing fossil gender, respectively.
+Let $\boldsymbol{W}$ be an $(n_{extant} + n_{fossil}) \times 1$ vector of the covariate sex of the stickleback fish and $\boldsymbol{Y}$ be an $(n_{extant} + n_{fossil}) \times K$ matrix of the $K$ phenotypes of interest.
+Because the sex of the fossilized stickleback fish is unobservable, we further define $\boldsymbol{W} = (\boldsymbol{W}_{extant}^T,\boldsymbol{W}_{fossil}^T)^T$ where $\boldsymbol{W}_{extant}$ and $\boldsymbol{W}_{fossil}$ are the $n_{extant} \times 1$ and $n_{fossil} \times 1$ vectors of the observed extant sex and missing fossil sex, respectively.
 
-We impute the missing gender for the fossil data by sampling from the posterior predictive distribution $P(\boldsymbol{W}_{fossil}|\boldsymbol{W}_{extant},\boldsymbol{Y})$ using the multiple imputation by chained equations (MICE) algorithm (@MICE) with predictive mean matching.
+We impute the missing sex for the fossil data by sampling from the posterior predictive distribution $P(\boldsymbol{W}_{fossil}|\boldsymbol{W}_{extant},\boldsymbol{Y})$ using the multiple imputation by chained equations (MICE) algorithm (@MICE) with predictive mean matching.
 Traditionally, the choice for the number of completed dats sets is a relatively small number such as $M = 5$ or $M = 10$.
@@
 
-For a given imputed dataset, let $W_{ij}$ be the imputed gender and $\boldsymbol{Y}_{ij}$ be the $K \times 1$ vector of phenotypes for stickleback fossil $j$ at time $t_i$ where $i = 1, \ldots, T$ and $j = 1,\ldots,n_{t}$.
+For a given imputed dataset, let $W_{ij}$ be the imputed sex and $\boldsymbol{Y}_{ij}$ be the $K \times 1$ vector of phenotypes for stickleback fossil $j$ at time $t_i$ where $i = 1, \ldots, T$ and $j = 1,\ldots,n_{t}$.
 Note, we are inputting only the fossil data into our model; connecting to the previous section, $W_{ij}$ and $\boldsymbol{Y}_{ij}$ can be interpretted as $W_{fossil,ij}$ and $\boldsymbol{Y}_{fossil,ij}$.
@@
 \end{align}
-for $g \in \{f,m\}$ where $\beta_{0,kg}$ and $\beta_{1,kg}$ are regression parameters of phenotype $Y_k$ for each gender, accounting for the possibility of a time-dependent trend in the mean structure, and $u_{k,gt_i}$ is the corresponding residual.
-To account for potential correlations between the residuals for a given trait $k$ and gender $g$, we fit an Ornstein-Uhlenback (OU) process (@OUProcess).
-More specifically, define $du_{k,gt} = u_{k,g(t + dt)} - u_{k,gt}$, the change in $u_k,gt$ for a given trait $k$ and gender $g$ over a miniscule time period $dt$.
+for $g \in \{f,m\}$ where $\beta_{0,kg}$ and $\beta_{1,kg}$ are regression parameters of phenotype $Y_k$ for each sex, accounting for the possibility of a time-dependent trend in the mean structure, and $u_{k,gt_i}$ is the corresponding residual.
+To account for potential correlations between the residuals for a given trait $k$ and sex $g$, we fit an Ornstein-Uhlenback (OU) process (@OUProcess).
+More specifically, define $du_{k,gt} = u_{k,g(t + dt)} - u_{k,gt}$, the change in $u_k,gt$ for a given trait $k$ and sex $g$ over a miniscule time period $dt$.
 The OU process is defined as
@@
 \item OU with Trend: The mean process for each phenotype $k=1,\cdots,K$ and $g \in \{f,m\}$ as previously defined.
-\item OU with No Trend: No linear trend on the mean processes and the overall mean of the phenotype is constant for each gender.
+\item OU with No Trend: No linear trend on the mean processes and the overall mean of the phenotype is constant for each sex.
 This is achieved by setting $\beta_{1,kg} = 0$ in (\ref{eq:mu}) for $k = 1,\cdots,K$ and $g \in \{f,m\}$
@@
 This is achieved by setting $u_{k,gt_i} = 0$ in (\ref{eq:mu}) for $k = 1,\cdots,K$, $g \in \{f,m\}$, and $i = 1,\cdots,T$.
-\item No OU with No Trend: The mean for a specific trait and specific gender at each time point is a constant value.
+\item No OU with No Trend: The mean for a specific trait and specific sex at each time point is a constant value.
 This is achieved by setting $\beta_{1,kg} = 0$ and $u_{k,gt_i} = 0$ in (\ref{eq:mu}) for $k = 1,\cdots,K$, $g \in \{f,m\}$, and $i = 1,\cdots,T$.
@@
 The traits that did include an OU component were caudal vertebra (mcv), pelvic spine lentgh (lpt), abdominal vertebra (mav), and standard length (stl) with caudal vertebra (mcv) being the only model with both a trend and an OU component.
-This implies that not only does caudal vertebra size have a linear trend across gender, but the average size at one time period is highly correlated with the average size at the next time period.<!-- NOTE FOR MATT: You should mention something about MCV being the only one with a trend and an OU.
+This implies that not only does caudal vertebra size have a linear trend across sex, but the average size at one time period is highly correlated with the average size at the next time period.<!-- NOTE FOR MATT: You should mention something about MCV being the only one with a trend and an OU.
 Explain why this might be and what it means.
 -->
+
+\textcolor{red} {
+Yoel's interpretation of the above paragraph.
+Taking the model with the lowest DIC score as the best model for each trait, we found that cle, ds1, lps, and pmx did not fit either the trend model or the OU model for male-female trait mean differences.
+In contrast, we found a trend through time in male-female trait mean differences (i.e., sexual dimorphism) for mcv, ds2, ds3, ect, maf, and mdf.
+This suggests that these traits changed their dimorphism from one time period to the next, with an overall tendency toward ???GREATER/LESS???
+dimorphism.
+}
+\textcolor{red} {
+We found evidence for an OU process but no trend for lpt, mav, and stl, suggesting that these traits varied through time, about a fixed value for dimorphism.
+Finally, we found evidence that change in mcv fitted both a trend an an OU process, meaning ???
+WHAT BIOLOGICALLY ???.
+}
+
+\textcolor{red} {HOW DO WE RECONCILE THE PRESENCE OF TRENDS IN THESE MODELS WITH AN ABSENCE OF VISIBLE TRENDS IN FIGURES 2 THROUGH 4?
+I WONDER IF THIS RESULT PARAGRAPH BELONGS BEST AT THE END OF THE RESULTS, RATHER THAN THE BEGINNING.
+CONSIDER THE STATISTICAL EVIDENCE FOR TRENDS AND OU AFTER WE'VE WALKED THE READERS THROUGH VISUAL EVIDENCE FOR CHANGE.
+}
 
@@
 Figure \ref{fig:mu-diff-plot} shows the posterior mean difference for males versus females (values above 0 indicate means that were greater for males versus females).
-The full poosterior distributions over time for all traits are shown in the appendix.
+The full posterior distributions over time for all traits are shown in the appendix.
 
@@
 Over time, however, this switches and the male specimens exhibit larger mean values for the latest observed time periods.
+
+\textcolor{red} {
+Yoel's interpretation of the three paragraphs above.
+Male and female G.
+doryssus differ for some traits, indicating detectable sexual dimorphism.
+This is evidenced by traits with substantial differences in male and female posterior means (i.e., deviation from 0 in Figure \ref{fig:mu-diff-plot}), supported by posterior probabilities that deviate from 0.5 (Figure \ref{fig:prob0-plot}).
+For example, males are larger than females by 3-8mm on average through time (stl panel in Figure \ref{fig:mu-diff-plot}), with a posterior probability for that mean difference above 70 percent through time (Figure \ref{fig:prob0-plot}).
+In contrast, the number of abdominal vertebrae (mav) fluctuates about 0 (Figure \ref{fig:mu-diff-plot}), with posterior probabilities fluctuating about 50 percent through time (Figure \ref{fig:prob0-plot}).
+Table \ref{tab:prob0-table} in the Supplementary Material summarizes the posterior probabilities that male means differ from female means by trait and sample.
+}
+
+\textcolor{red}{can we move mu-diff-plot and prob0-plot here?}
 
@@
 <!-- MDF: Dorsal fin ray: Number of bones in the dorsal finposterior to the third dorsal spine (i.e.,soft dorsal fin rays. -->
+
+\textcolor{red} {
+Yoel's interpretation of the five paragraphs above. 
+Figure \ref{fig:AUC-plot} shows the posterior probability that the mean male-female difference at time t is different from the male-female difference at time zero.
+In other words, Figure \ref{fig:AUC-plot} plots whether the sexual dimorphism changes in subsequent samples, always relative to the first sample.
+The amount of sexual dimporphism over time does not change substantially for a majority of the traits (i.e.
+cle, ds1, ds2, ect, lps, lpt, mav, mcv, pmx, stl; Figure \ref{fig:AUC-plot}).
+Traits that change in dimorphism are anal fin ray (maf), and dorsal fin ray (mdf), and dorsal spine 3 (ds3).}
+
+\textcolor{red} {
+For example, females at time zero averaged 8.29 rays in the anal fin, dropping to 8.17 rays in the final sample.
+Males averaged 8.52 anal fin rays in the first sample and 8.55 in the last sample.
+Thus, the difference between males and females started at 0.23 and ended at 0.38, an increase in sexual dimorphism.
+In contrast, the sign on dimorphism flipped for ds3.
+The posterior mean length of the third dorsal spine is 2.29mm for males and 2.45mm for females.
+By the end of the sequence, mean ds3 for males was 1.27 for males and 1.19 for females.
+Both sexes evolved reduction of ds3, but females lowered their mean more than the males.
+This resulted in a flip in dimorphism, from larger females to larger males, though both sexes evolved trait reduction.
+BUT DIMORPHISM SHRUNK THEN, AS THE DIFFERENCE BETWEEN MALES AND FEMALES IS NOW SMALLER??
+}
 
@@
 
-```{r mu-diff-plot, echo = FALSE, fig.cap = "Posterior mean difference between males and females"}
-ggplot(aes(x = time, y = mu_diff), data = results_posterior_mean_diff) + facet_wrap(~phenotype, scale = "free_y") + geom_path() + geom_hline(yintercept = 0, lty = 3) + ylab("Posterior Mean")
+posterior mean difference for males versus females (values above 0 indicate means that were greater for males versus females)
+
+```{r mu-diff-plot, echo = FALSE, fig.cap = "Posterior mean difference between males and females, in UNITS?. Values above 0 indicate larger male means"}
+ggplot(aes(x = time, y = mu_diff), data = results_posterior_mean_diff) + facet_wrap(~phenotype, scale = "free_y") + geom_path() + geom_hline(yintercept = 0, lty = 3) + ylab("Posterior Mean Difference")
 ```
 
-```{r prob0-plot, echo = FALSE, fig.cap = "Posterior probabilities that mean for males is greater than mean for females"}
+```{r prob0-plot, echo = FALSE, fig.cap = "Posterior probabilities that the male posterior mean is larger than the female posterior mean"}
 ggplot(aes(x = time, y = prob_mu_diff_gt_0), data = results_prob_mu_diff_gt_0) + facet_wrap(~phenotype) + geom_path() + #geom_hline(yintercept = 0, lty = 3) + ylim(0,1) + ylab("Posterior Probability") Why yintercept in 0 and not 0.5?
@@
 
-```{r AUC-plot, echo = FALSE, fig.cap= "AUC comparing time x versus time 0."}
+```{r AUC-plot, echo = FALSE, fig.cap= "The Posterior Probability that male-female differences in subsequent samples are different from male-female differenceds in the initial sample.
+Values near 0.5 indicate little difference between the distributions.
+Values closer to 0 indicate that the difference in means has shifted towards larger means in females.
+Values closer to 1 indicate that the difference in means has shifted toward larger means in males."}
 ggplot(aes(x = time, y = auc_mu_diff), data = results_auc_mu_diff) + facet_wrap(~phenotype) + geom_path() +  geom_path() + geom_hline(yintercept = 0.5, lty = 3) + ylim(0,1) + ylab("Posterior Probability")
@@
 
-# Discsusson, Future work and conclusions {#sec:conclusions}
+# Discussion, Future work and conclusions {#sec:conclusions}
 We predicted that release from predators would result in niche expansion and increased sexual dimorphism, based on several studies of modern stickleback.
@@
 Abughoush, S.
-Blaine, A.
+Blain, A.
 Chaudhary, M.

### 2024-07-25 manuscript/paper.tex
manuscript additions

@@
 
-Ecological release theory suggests that a prey population's niche should
-expand when predation is relaxed or removed, as prey take advantage of
-new enemy free space (reviewed in Herrmann, Stroud, and Losos
-(\protect\hyperlink{ref-Herrmann2021}{2021})). Niche expansion could
-manifest through a combination of increasing within- and
-between-individual niche widths (Bolnick et al.
-(\protect\hyperlink{ref-Bolnick2010}{2010}); Herrmann, Stroud, and Losos
-(\protect\hyperlink{ref-Herrmann2021}{2021})), including divergence
-between the sexes (Bolnick and Doebeli
-(\protect\hyperlink{ref-Bolnick2003}{2003}); Cooper, Gilman, and
-Boughman (\protect\hyperlink{ref-Cooper2011}{2011})). Theory and
-empirical data suggest that a population presented with ecological
-opportunity following ecological release might experience disruptive
+Ecological release theory suggests that a population's niche should
+change when important species interactions (e.g., resource competition)
+are relaxed or removed (reviewed in Herrmann, Stroud, and Losos
+(\protect\hyperlink{ref-Herrmannetal2021}{2021})). Removal is posited to
+create ecological opportunities---i.e., aspects of the niche become
+newly accessible, and the focal population shifts and/or expands its
+resource use to the new opportunity (Parent and Crespi
+(\protect\hyperlink{ref-ParentandCrespi2009}{2009}); Herrmann, Stroud,
+and Losos (\protect\hyperlink{ref-Herrmannetal2021}{2021})). Ecological
+release may then be followed by adaptive morphological evolution (Parent
+and Crespi (\protect\hyperlink{ref-ParentandCrespi2009}{2009});
+Herrmann, Stroud, and Losos
+(\protect\hyperlink{ref-Herrmannetal2021}{2021})). For example, a
+population undergoing ecological release can experience disruptive
 selection on males and females stemming from intraspecific competition
-over newly accessible resources, resulting in intersexual divergence in
-habitat use (W. (\protect\hyperlink{ref-Schoener1968}{1968}); Shine
+over newly accessible resources (Bolnick and Doebeli
+(\protect\hyperlink{ref-BolnickandDoebeli2003}{2003}); Bolnick and Lau
+(\protect\hyperlink{ref-BolnickandLau2008}{2008}); Cooper, Gilman, and
+Boughman (\protect\hyperlink{ref-Cooperetal2011}{2011})). This is
+predicted to result in intersexual divergence in habitat use and
+associated phenotypes (Schoener
+(\protect\hyperlink{ref-Schoener1968}{1968}); Shine
 (\protect\hyperlink{ref-Shine1989}{1989}); Bolnick and Doebeli
-(\protect\hyperlink{ref-Bolnick2003}{2003}); Butler, Sawyer, and Losos
-(\protect\hyperlink{ref-Butler2007}{2007}); Bolnick and Lau
-(\protect\hyperlink{ref-Bolnick2008}{2008}); Cooper, Gilman, and
-Boughman (\protect\hyperlink{ref-Cooper2011}{2011}); but see Stuart et
-al. (\protect\hyperlink{ref-Stuart2021}{2021}); Blain
-(\protect\hyperlink{ref-Blain2022}{2022})). Accordingly, sexes should
-also diverge in morphological traits associated with habitat use; such
-ecologically-driven ``character displacement between the sexes'' is
-therefore one explanation for phenotypic sexual dimorphism (S. P. De
-Lisle and Rowe (\protect\hyperlink{ref-deLisleRowe2015}{2015}); S. Paiva
-De Lisle S. P. and Rowe (\protect\hyperlink{ref-deLisleNco2018}{2018})).
-
-We tested this link between release from predators and the evolution of
-sexual dimorphism in a well-preserved, finely-resolved,
-\textasciitilde16,000 year-long sequence of the fossil stickleback fish
-(Gasterosteus doryssus). The fossil lineage appeared in the depositional
-environment as a fully armored form, with complete pelvic girdles, two
-pelvic spines, and three dorsal spines, on average (Michael A. Bell,
-Travis, and Blouw (\protect\hyperlink{ref-Bell2006}{2006}); Stuart,
-Travis, and Bell (\protect\hyperlink{ref-Stuart2020}{2020})). However,
-this habitat appears to have been missing fish piscivores (e.g., trout
-and other salmonids known to prey on modern stickleback; M. A. Bell
-(\protect\hyperlink{ref-Bell2009}{2009})), relaxing putative selection
-for armor (T. E. Reimchen (\protect\hyperlink{ref-Reimchen1980}{1980}),
-T. E. Reimchen (\protect\hyperlink{ref-Reimchen1994}{1994}), O. Bell M.
-A. and Koenings (\protect\hyperlink{ref-BellEtAl1993}{1993}); Bowne
-(\protect\hyperlink{ref-Bowne1994}{1994}); Roesti and Schluter
-(\protect\hyperlink{ref-Roestietal2023}{2023})). The lineage adaptively
-evolved armor loss but also reduction in the means of several other
-traits (Hunt, Bell, and Travis (\protect\hyperlink{ref-Hunt2008}{2008});
-Stuart, Travis, and Bell (\protect\hyperlink{ref-Stuart2020}{2020});
-Siddiqui et al.~in prep). In particular, tooth wear data suggest that
-the population shifted to eat more planktonic prey through time,
-indicative of increased open water habitat use (Purnell et al.~2007). If
-this divergence in habitat use is partitioned between sexes (Reimchen T.
-E. and Nosil (\protect\hyperlink{ref-ReimchenNosil2004}{2004});
-Spoljaric and Reimchen
-(\protect\hyperlink{ref-SpoljaricReimchen2008}{2008});D. S. Reimchen T.
-E. (\protect\hyperlink{ref-Reimchen2016}{2016})), then we should expect
-sexual dimorphism to evolve for traits related to habitat use in G.
-doryssus.
-
-To test this prediction, we imputed the sex of individual fossils, which
-is unobserved in practice, with Multiple Imputation by Chained Equations
-(MICE) (Buuren and Groothuis-Oudshoorn
-(\protect\hyperlink{ref-MICE}{2011})), using observed data from extant
-stickleback of known sex to train the imputation model. Then we tracked
-the multivariate evolution of sexual dimorphism over
-\textasciitilde16,000 years for 16 traits related to swimming, feeding
-and defense.
-
-\hypertarget{methods-and-materials}{%
-\section{Methods and Materials}\label{methods-and-materials}}
-
-A major challenge in this work is inferring the sex of fossil
-stickleback specimens. Indeed, for most species, sex cannot be detected
-directly, except for lineages whose sexes are distinguished by the
-presence or absence of sex-specific characters (Hone and Mallon
-(\protect\hyperlink{ref-Hone2017}{2017}); Saitta et al.
-(\protect\hyperlink{ref-SaittaEtAl2020}{2020})). Instead,
-paleobiologists resort to statistical detection of sex and sexual
-dimorphism, including tests for bimodality in trait distributions (e.g.,
-Hone and Mallon (\protect\hyperlink{ref-Hone2017}{2017})) and divergence
-in growth curves (e.g., Saitta et al.
-(\protect\hyperlink{ref-SaittaEtAl2020}{2020})). We used a third
-approach to infer fossil sex and study the evolution of sexual
-dimorphism by taking advantage of the fact that G. doryssus is part of
-the extant threespine stickleback species complex, Gasterosteus
-aculeatus. We measured the same trait set from a phenotypically diverse
-set of extant G. aculeatus populations, for whom individual sex was
-known. We then combined fossil and extant data sets, treated fossil sex
-as missing data, and used the MICE (Buuren and Groothuis-Oudshoorn
-(\protect\hyperlink{ref-MICE}{2011})) multiple imputation (Little and
-Rubin (\protect\hyperlink{ref-little2002statistical}{2002})) algorithm,
-and implemented in R (Team
-(\protect\hyperlink{ref-R2022language}{2022})), to impute individual sex
-of the fossil data set 100 times. We then fit a modified
-Ornstein--Uhlenbeck (OU) (Uhlenbeck and Ornstein
-(\protect\hyperlink{ref-OUProcess}{1930})) model using a Bayesian
-framework to the data to test for the evolution of sexual dimorphism in
-each trait.
+(\protect\hyperlink{ref-BolnickandDoebeli2003}{2003}); Butler and Losos
+(\protect\hyperlink{ref-Butleretal2007}{2007}); Bolnick and Lau
+(\protect\hyperlink{ref-BolnickandLau2008}{2008}); Cooper, Gilman, and
+Boughman (\protect\hyperlink{ref-Cooperetal2011}{2011}); but see Stuart
+et al. (\protect\hyperlink{ref-Stuartetal2021}{2021}); Blain
+(\protect\hyperlink{ref-Blain2022}{2022})). Such competition-driven
+``intraspecific character displacement'' between the sexes is therefore
+one explanation for the evolution of sexual dimorphism (Pfennig and
+Pfennig (\protect\hyperlink{ref-PfennigandPfennig2012}{2012}); De Lisle
+and Rowe (\protect\hyperlink{ref-DeLisleandRowe2015}{2015}); De Lisle
+and Rowe (\protect\hyperlink{ref-DeLisleandRowe2017}{2017}), De Lisle,
+Paiva, and Rowe (\protect\hyperlink{ref-DeLisleetal2018}{2018})).
+
+Much of the theory and empirical data for character displacement between
+the sexes is based on release from resource competition specifically
+(e.g., Bolnick and Doebeli
+(\protect\hyperlink{ref-BolnickandDoebeli2003}{2003}); Cooper, Gilman,
+and Boughman (\protect\hyperlink{ref-Cooperetal2011}{2011}); Pfennig and
+Pfennig (\protect\hyperlink{ref-PfennigandPfennig2012}{2012}); De Lisle
+and Rowe (\protect\hyperlink{ref-DeLisleandRowe2015}{2015}); De Lisle,
+Paiva, and Rowe (\protect\hyperlink{ref-DeLisleetal2018}{2018})).
+However, release from predation should also result in ecological release
+followed by the evolution of increased sexual dimorphism because the
+absence of predators should generate ecological opportunity (Reimchen
+and Nosil (\protect\hyperlink{ref-ReimchenandNosil2004}{2004}); Parent
+and Crespi (\protect\hyperlink{ref-ParentandCrespi2009}{2009});
+Herrmann, Stroud, and Losos
+(\protect\hyperlink{ref-Herrmannetal2021}{2021})).
+
+Here, we tested the prediction that a release from predation should
+result in the evolution of increased sexual dimorphism using a
+well-preserved, finely-resolved sequence of a fossil threespine
+stickleback fish (Gasterosteus doryssus). The sequence in this
+depositional environment is comprised of two lineages (Bell
+(\protect\hyperlink{ref-Bell2009}{2009}); Cerasoni, Bell, and Stuart
+(\protect\hyperlink{ref-Cerasonietal2024}{2024})). Lineage I was a
+low-armor form with zero to one dorsal spines and highly reduced
+pelvises, on average. It lasted for at least 93,000 years before it was
+replaced suddenly by a second lineage (Bell, Baumgartner, and Oslon
+(\protect\hyperlink{ref-Belletal1985}{1985})). Lineage II appeared in
+the depositional environment fully armored, with complete pelvic
+girdles, two pelvic spines, and three dorsal spines, on average (Bell,
+Travis, and Blouw (\protect\hyperlink{ref-Belletal2006}{2006}); Stuart,
+Travis, and Bell (\protect\hyperlink{ref-Stuartetal2020}{2020})). It
+then immediately began evolving adaptive reduction in its armor traits
+(Hunt, Bell, and Travis (\protect\hyperlink{ref-Huntetal2008}{2008});
+Stuart, Travis, and Bell (\protect\hyperlink{ref-Stuartetal2020}{2020}))
+until it reached the same low-armor state previously held by lineage I.
+The observation of 93,000 years of low-armor stasis in lineage I (Bell,
+Baumgartner, and Oslon (\protect\hyperlink{ref-Belletal1985}{1985})) and
+the observation of rapid evolution of armor loss by lineage II, both
+suggest that this depositional environment lacked piscivorous fish
+(e.g., trout and other salmonids known to prey on modern threespine
+stickleback). Armor maintenance in extant threespine stickleback
+(Gasterosteus aculeatus) correlates strongly the presence of vertebrate
+piscivores; populations with less predation pressure typically have less
+armor (Reimchen (\protect\hyperlink{ref-Reimchen1994}{1994}), Reimchen
+and Nosil (\protect\hyperlink{ref-ReimchenandNosil2004}{2004}); Bell et
+al. (\protect\hyperlink{ref-Belletal1993}{1993}); Roesti et al.
+(\protect\hyperlink{ref-Roestietal2023}{2023})). Consistently, only
+three fossil trout have been recorded from this paleolake basin, sampled
+from the same rock that has revealed \textgreater15,000 threespine
+stickleback fossils as well as occasional killifish (Fundulus
+nevadensis) (Bell (\protect\hyperlink{ref-Bell2009}{2009}); Cerasoni,
+Bell, and Stuart (\protect\hyperlink{ref-Cerasonietal2024}{2024})).
+Finally, tooth wear data from the lineage II sequence suggest that
+individuals in this lineage began eating more planktonic prey over time,
+expanding toward an open-water niche from the benthic, bottom-feeding
+niche they started with (Purnell et al.~2007). That this expansion into
+open water by lineage II coincided with armor reduction is further
+indicative of a limnetic system with fewer salmonid piscivores (Schluter
+and McPhail (\protect\hyperlink{ref-SchluterandMcPhail1992}{1992});
+Vamosi and Schluter
+(\protect\hyperlink{ref-VamosiandSchluter2004}{2004}); Roesti et al.
+(\protect\hyperlink{ref-Roestietal2023}{2023})). In sum, we infer that
+lineage II likely started in a nearby paleolake basin that had predators
+since it was armored when it arrived (Bell
+(\protect\hyperlink{ref-Bell2009}{2009}); Cerasoni, Bell, and Stuart
+(\protect\hyperlink{ref-Cerasonietal2024}{2024})). Lineage II then
+experienced release from predators and dietary expansion in the focal
+paleolake basin, generating a scenario consistent with the starting
+conditions for models predicting the evolution of increased sexual
+dimorphism following ecological release.
+
+A major challenge in the study of fossilized sexual dimorphism is
+assigning sex to individual specimens in the first place (Hone and
+Mallon (\protect\hyperlink{ref-HoneandMallon2017}{2017}); Mallon
+(\protect\hyperlink{ref-Mallon2017}{2017}); Saitta et al.
+(\protect\hyperlink{ref-Saittaetal2020}{2020})). This typically cannot
+be done directly, except for taxa whose sexes are distinguished by the
+presence or absence of sex-specific characters that preserve well.
+Instead, paleobiologists often resort to statistical detection of sex
+and sexual dimorphism, including tests for normality and bimodality in
+trait distributions (e.g., Mallon
+(\protect\hyperlink{ref-Mallon2017}{2017})), mixture modeling (e.g.,
+Mallon (\protect\hyperlink{ref-Mallon2017}{2017})), divergence in growth
+curves (e.g., Saitta et al.
+(\protect\hyperlink{ref-Saittaetal2020}{2020})), and emphasis on effect
+size statistics rather than significance testing (e.g., Saitta et al.
+(\protect\hyperlink{ref-Saittaetal2020}{2020})). However, dimorphic
+signal can be masked by noise introduced by factors both biological and
+artifactual, including extended growth during ontogeny, age-based bias
+in survivorship, small sample sizes, time averaging, and taphonomic bias
+(Godfrey, Lyon, and Sutherland
+(\protect\hyperlink{ref-Godfreyetal1993}{1993}); Koscinski and
+Pietraszewski
+(\protect\hyperlink{ref-KoscinskiandPietraszewski2004}{2004}); Hone and
+Mallon (\protect\hyperlink{ref-HoneandMallon2017}{2017}); reviewed in
+Mallon (\protect\hyperlink{ref-Mallon2017}{2017}); reviewed in Saitta et
+al. (\protect\hyperlink{ref-Saittaetal2020}{2020})). Thus, the best
+approach to sex classification is likely one of total evidence (Saitta
+et al. (\protect\hyperlink{ref-Saittaetal2020}{2020})), including
+comparison to closely-related, extant species of known sex (e.g., Hone
+and Mallon (\protect\hyperlink{ref-HoneandMallon2017}{2017}); Saitta et
+al. (\protect\hyperlink{ref-Saittaetal2020}{2020})).
+
+For our study, we inferred sex in G. doryssus fossils by comparing
+multivariate morphological trait data from lineage II samples to the
+same multivariate trait set collected from multiple populations in the
+closely-related, extant Threespine Stickleback species complex
+(Gasterosteus aculeatus). Crucially, we determined G. aculeatus sex
+directly via dissection and/or PCR genotyping. This enabled us to use
+Multiple Imputation by Chained Equations (MICE; Buuren and
+Groothuis-Oudshoorn (\protect\hyperlink{ref-MICE}{2011})) to build a
+predictive multiple imputation algorithm (Little and Rubin
+(\protect\hyperlink{ref-little2002statistical}{2002})) based on the
+multivariate morphology of G. aculeatus individuals of known sex. We
+applied this algorithm to the fossil data to impute individual sex based
+on morphology, treating fossil sex as a missing variable. With sex
+assigned to fossil specimens (and accounting for uncertainty in those
+assignments), we fit a modified Ornstein-Uhlenbeck (OU) (Uhlenbeck and
+Ornstein (\protect\hyperlink{ref-OUProcess}{1930})) model using a
+Bayesian framework to test for evolution of sexual dimorphism in each
+trait over \textasciitilde16,000 years of lineage II. We hypothesized
+that sexual dimorphism should increase following ecological release from
+predators.
 
@@
 We used Gasterosteus doryssus data that were previously reported by
-Stuart, Travis, and Bell (\protect\hyperlink{ref-Stuart2020}{2020}),
-Voje, Bell, and Stuart (\protect\hyperlink{ref-Voje2022}{2022}), and
-Siddiqui et al.~(in review). Briefly, the data were collected from
-fossil Series K from Quarry D (Cerasoni et al., in review), dug from an
-open pit diatomite mine at 9.526° N, 119.094° W, near Fernley, Nevada,
-USA. Series K consisted of 18 samples taken at \textasciitilde1000-year
+Stuart, Travis, and Bell (\protect\hyperlink{ref-Stuartetal2020}{2020}),
+Voje, Bell, and Stuart (\protect\hyperlink{ref-Vojeetal2022}{2022}), and
+Siddiqui et al. (\protect\hyperlink{ref-Siddiquietal2024}{2024}).
+Briefly, the data were collected from fossil Series K from Quarry D
+(Cerasoni, Bell, and Stuart
+(\protect\hyperlink{ref-Cerasonietal2024}{2024})), dug from an open pit
+diatomite mine at 9.526° N, 119.094° W, near Fernley, Nevada, USA.
+Series K consisted of 18 samples taken at \textasciitilde1000-year
 intervals, and mean sample times span \textasciitilde16,363 years. Fish
@@
 to one dorsal spines, zero pelvic spines, and highly reduced pelvises
-was completely replaced by a high armored lineage of stickleback with
-three dorsal spines, two pelvic spines, and complete (Bell et al.~2006;
-Bell 2009; Stuart et al.~2020). This lineage subsequently evolved
-reduction in armor, body size, and traits related to swimming and
-feeding (Bell et al.~2006; Stuart et al.~2020; Siddiqui et al.~in
-review). The tempo and mode of armor reduction during this sequence
-suggests adaptive evolution by natural selection (Hunt et al.~2008), and
-we focus on the multivariate evolution of sexual dimorphism by this
-second lineage.
-
-The fossil data consists of 814 specimens with unknown sex over 18 time
-periods spaced \textasciitilde1000 years apart. Figure 1 shows the
-sample size at each of the 18 time periods. There are at least 22
-specimens at each time period with a high of 67 specimens in period 7.
+(i.e., lineage I described above) was completely replaced by a high
+armored lineage of stickleback with three dorsal spines, two pelvic
+spines, and complete pelvises (i.e., lineage II described above) (Bell,
+Travis, and Blouw (\protect\hyperlink{ref-Belletal2006}{2006}); Bell
+(\protect\hyperlink{ref-Bell2009}{2009}); Stuart, Travis, and Bell
+(\protect\hyperlink{ref-Stuartetal2020}{2020})). Lineage II subsequently
+evolved reduction in armor, body size, and traits related to swimming
+and feeding (Bell, Travis, and Blouw
+(\protect\hyperlink{ref-Belletal2006}{2006}); Stuart, Travis, and Bell
+(\protect\hyperlink{ref-Stuartetal2020}{2020}); Siddiqui et al.
+(\protect\hyperlink{ref-Siddiquietal2024}{2024})). The tempo and mode of
+armor reduction during this sequence suggests adaptive evolution by
+natural selection (Hunt, Bell, and Travis
+(\protect\hyperlink{ref-Huntetal2008}{2008})), and we focus on the
+multivariate evolution of sexual dimorphism by this second lineage.
+
+The lineage II fossil data consist of 814 specimens of unknown sex
+sampled across 18 time periods spaced \textasciitilde1000 years apart.
+Figure 1 shows the sample size for each of the 18 samples. There are at
+least 22 specimens in each sample with a high of 67 specimens in sample
+7.
 
@@
 
-\caption{Fossil sampel size for each time period}\label{fig:unnamed-chunk-2}
+\caption{Fossil sample size for each time period}\label{fig:unnamed-chunk-2}
 \end{figure}
@@
 
-To span the gamut of stickleback diversity for our predictive model, we
-sampled modern stickleback from lakes containing generalist stickleback
-populations (Hendry et al. (\protect\hyperlink{ref-Hendry2009}{2009});
-Bolnick (\protect\hyperlink{ref-Bolnick2011}{2011})) and from lakes
-containing benthic-limnetic species pairs (Baumgartner et al.~1988;
-Schluter and McPhail (\protect\hyperlink{ref-Schluter1992}{1992}))
-(Table 2 (What is this?)). The generalist populations were collected by
-YES in 2013 and previously described in Stuart et al.
-(\protect\hyperlink{ref-Stuart2017}{2017}). These samples were fixed in
-formalin, then stained for bone with Alizarin Red in 2013. Benthic and
-limnetic specimens were kindly loaned by D. Schluter and his lab at
-University of British Columbia. They collected benthic and limnetic
-individuals from Enos Lake in 1988 and from Emily Lake, Little Quarry
-Lake, Paxton Lake, and Priest Lake in 2018. The Enos specimens had been
-fixed whole in formalin and stored in 40\% isopropanol. The specimens
-from the other lakes were initially preserved whole in 95\% ethanol in
-the field before being gradually transferred to water then formalin in
-the lab and ultimately stored in 40\% isopropanol. In 2019, we stained
-these specimens for bone using Alizarin Red.
-
-We next replicated fossil data collection (Table 1) on these extant
+To span the gamut of stickleback diversity for our predictive imputation
+model, we sampled modern stickleback from lakes containing generalist
+stickleback populations (Hendry et al.
+(\protect\hyperlink{ref-Hendryetal2009}{2009}); Bolnick
+(\protect\hyperlink{ref-Bolnick2011}{2011})) and from lakes containing
+benthic-limnetic species pairs (Baumgartner, Bell, and Weinberg
+(\protect\hyperlink{ref-Baumgartneretal1988}{1988}); Schluter and
+McPhail (\protect\hyperlink{ref-SchluterandMcPhail1992}{1992})). The
+generalist populations used here were the remainder of collections made
+by YES in 2013 and were previously described in Stuart et al.
+(\protect\hyperlink{ref-Stuartetal2017}{2017}) (Table S1). These samples
+were fixed in formalin, then stained for bone with Alizarin Red in 2013.
+Benthic and limnetic specimens were kindly loaned by D. Schluter and his
+lab at University of British Columbia. They collected benthic and
+limnetic individuals from Enos Lake in 1988 and from Emily Lake, Little
+Quarry Lake, Paxton Lake, and Priest Lake in 2018 (Table S1). The Enos
+specimens had been fixed whole in formalin and stored in 40\%
+isopropanol. The specimens from the other lakes were initially preserved
+whole in 95\% ethanol in the field before being gradually transferred to
+water then formalin in the lab and ultimately stored in 40\%
+isopropanol. In 2019, we stained these specimens for bone using Alizarin
+Red. We next replicated fossil data collection (Table 1) on these extant
 specimens. Standard length as well as pelvic-spine length on each side
@@
 the anal fin pterygiophore (following Aguirre, Walker, and Gideon
-(\protect\hyperlink{ref-Aguirre2014}{2014}). X-rays were taken with an
-AXR Hot Shot X-ray Machine (Associated X-ray Corporation) at the Field
-Museum of Natural History. Specimens were exposed at 35kV and 4mA. Small
-fish were exposed for 7s, medium fish for 8s, and large fish for 10s. We
-developed the film and scanned individual images of each fish using the
-B\&W Negatives setting on an Epson Perfection 4990 Photo flatbed at 2400
-dpi. Measurements from photographs and X-rays were taken with FIJI
-(Schindelin et al.~2012) and its plugin ObjectJ
-(\url{https://sils.fnwi.uva.nl/bcb/objectj/}). All photographs, X-rays,
-and ObjectJ files have been uploaded to Morphosource.org (accession \#
-TBD). We dissected individuals from the generalist populations to
+(\protect\hyperlink{ref-Aguirreetal2014}{2014})). X-rays were taken with
+an AXR Hot Shot X-ray Machine (Associated X-ray Corporation) at the
+Field Museum of Natural History. Specimens were exposed at 35kV and 4mA.
+Small fish were exposed for 7s, medium fish for 8s, and large fish for
+10s. We developed the film and scanned individual images of each fish
+using the B\&W Negatives setting on an Epson Perfection 4990 Photo
+flatbed at 2400 dpi. Measurements from photographs and X-rays were taken
+with FIJI (Schindelin et al.
+(\protect\hyperlink{ref-Schindelinetal2012}{2012})) and its plugin
+ObjectJ (\url{https://sils.fnwi.uva.nl/bcb/objectj/}).
+
+We dissected individuals from the generalist populations (Table S1) to
 determine sex from the gonads. Individuals from the species-pair lakes
-were sexed by Schluter and his lab, using a genotyping protocol (confirm
-and cite).
+(TableS1) were previously sexed by the Schluter lab, using either
+dissection or a genotyping protocol (Whom, personal communication).
 
 The extant data used here consists of a total of 367 specimens all with
-known sex. Of these, there are 202 and 165 female and male specimens,
-respectively.
-
-\hypertarget{outlier-analysis.}{%
-\subsection{Outlier analysis.}\label{outlier-analysis.}}
+known sex (Table S1). Of these, there are 202 and 165 female and male
+specimens, respectively.
+
+\hypertarget{outlier-analysis-and-size-correction.}{%
+\subsection{Outlier analysis and size
+correction.}\label{outlier-analysis-and-size-correction.}}
 
@@
 deviations for each trait separately for K series fossil specimens
-(pooled across samples) and for extant specimens (pooled across lakes).
-We noted trait values greater than 3.5 standard deviations from the mean
-as potential outliers. We checked whether these potential outliers were
-a result of data entry and collection error and corrected them if they
-were. We turned the remaining outlier trait values to NAs. (confirm)
-(Wait, what? Why? Anything that was 3.5 SD above the mean was recorded
-as missing? How do you justify this? )
-
-Missing data imputation, including fossil sex
-
-Quantification and evolution of sexual dimorphism
-
-What covariates do we have in the data: length, what else,
+(pooled across samples) and for extant specimens (within generalist,
+benthic, or limnetic categories). We noted trait values greater than 3.0
+standard deviations from the mean as potential outliers. We deemed 3.0
+s.d. to be a reasonable threshold for detecting errors without excluding
+biologically relevant values. We checked whether these potential
+outliers were a result of data entry and collection error and corrected
+them if they were. We turned the remaining outlier trait values to NAs.
+We size-corrected continuous traits only, as they varied with size,
+unlike count traits that are fixed during early development. We
+regressed each trait on standard length using a mixed-model regression,
+pooling all specimens, following Stuart et al.
+(\protect\hyperlink{ref-Stuartetal2017}{2017}). We appended the size
+corrected data to the uncorrected trait data frame and used all of our
+data to build the imputation model described next.
+
+\hypertarget{missing-data-imputation-including-fossil-sex}{%
+\subsection{Missing data imputation, including fossil
+sex}\label{missing-data-imputation-including-fossil-sex}}
+
+We used multiple imputation (Little and Rubin
+(\protect\hyperlink{ref-little2002statistical}{2002})) to impute the sex
+of the fossils many times. Imputations were performed using the MICE
+(Buuren and Groothuis-Oudshoorn (\protect\hyperlink{ref-MICE}{2011}))
+algorithm and implemented in R (Team
+(\protect\hyperlink{ref-R2022language}{2022})) with M = 100 imputations
+performed. After imputing the sex of the fossil data, each completed
+data set is then used to fit a modified Ornstein--Uhlenbeck (OU)
+(Uhlenbeck and Ornstein (\protect\hyperlink{ref-OUProcess}{1930})) model
+using a Bayesian framework to look for sexual dimorphism across a
+variety of stickleback phenotypes.
 
@@
 Let \(\boldsymbol{W}\) be an \((n_{extant} + n_{fossil}) \times 1\)
-vector of the covariate gender of the stickleback fish and
+vector of the covariate sex of the stickleback fish and
 \(\boldsymbol{Y}\) be an \((n_{extant} + n_{fossil}) \times K\) matrix
-of the \(K\) phenotypes of interest. Because the gender of the
-fossilized stickleback fish is unobservable, we further define
+of the \(K\) phenotypes of interest. Because the sex of the fossilized
+stickleback fish is unobservable, we further define
 \(\boldsymbol{W} = (\boldsymbol{W}_{extant}^T,\boldsymbol{W}_{fossil}^T)^T\)
@@
 the \(n_{extant} \times 1\) and \(n_{fossil} \times 1\) vectors of the
-observed extant gender and missing fossil gender, respectively.
-
-We impute the missing gender for the fossil data by sampling from the
+observed extant sex and missing fossil sex, respectively.
+
+We impute the missing sex for the fossil data by sampling from the
 posterior predictive distribution
@@
 
-For a given imputed dataset, let \(W_{ij}\) be the imputed gender and
+For a given imputed dataset, let \(W_{ij}\) be the imputed sex and
 \(\boldsymbol{Y}_{ij}\) be the \(K \times 1\) vector of phenotypes for
@@
 \(\beta_{1,kg}\) are regression parameters of phenotype \(Y_k\) for each
-gender, accounting for the possibility of a time-dependent trend in the
+sex, accounting for the possibility of a time-dependent trend in the
 mean structure, and \(u_{k,gt_i}\) is the corresponding residual. To
 account for potential correlations between the residuals for a given
-trait \(k\) and gender \(g\), we fit an Ornstein-Uhlenback (OU) process
+trait \(k\) and sex \(g\), we fit an Ornstein-Uhlenback (OU) process
 (Uhlenbeck and Ornstein (\protect\hyperlink{ref-OUProcess}{1930})). More
 specifically, define \(du_{k,gt} = u_{k,g(t + dt)} - u_{k,gt}\), the
-change in \(u_k,gt\) for a given trait \(k\) and gender \(g\) over a
+change in \(u_k,gt\) for a given trait \(k\) and sex \(g\) over a
 miniscule time period \(dt\). The OU process is defined as \begin{align}
@@
 being the only model with both a trend and an OU component. This implies
-that not only does caudal vertebra size have a linear trend across
-gender, but the average size at one time period is highly correlated
-with the average size at the next time period.
+that not only does caudal vertebra size have a linear trend across sex,
+but the average size at one time period is highly correlated with the
+average size at the next time period.
+
+\textcolor{red} {
+Yoel's interpretation of the above paragraph.
+Taking the model with the lowest DIC score as the best model for each trait, we found that cle, ds1, lps, and pmx did not fit either the trend model or the OU model for male-female trait mean differences.
+In contrast, we found a trend through time in male-female trait mean differences (i.e., sexual dimorphism) for mcv, ds2, ds3, ect, maf, and mdf.
+This suggests that these traits changed their dimorphism from one time period to the next, with an overall tendency toward ???GREATER/LESS???
+dimorphism.
+} \textcolor{red} {
+We found evidence for an OU process but no trend for lpt, mav, and stl, suggesting that these traits varied through time, about a fixed value for dimorphism.
+Finally, we found evidence that change in mcv fitted both a trend an an OU process, meaning ???
+WHAT BIOLOGICALLY ???.
+}
+
+\textcolor{red} {HOW DO WE RECONCILE THE PRESENCE OF TRENDS IN THESE MODELS WITH AN ABSENCE OF VISIBLE TRENDS IN FIGURES 2 THROUGH 4?
+I WONDER IF THIS RESULT PARAGRAPH BELONGS BEST AT THE END OF THE RESULTS, RATHER THAN THE BEGINNING.
+CONSIDER THE STATISTICAL EVIDENCE FOR TRENDS AND OU AFTER WE'VE WALKED THE READERS THROUGH VISUAL EVIDENCE FOR CHANGE.
+}
 
@@
 males versus females (values above 0 indicate means that were greater
-for males versus females). The full poosterior distributions over time
+for males versus females). The full posterior distributions over time
 for all traits are shown in the appendix.
@@
 
+\textcolor{red} {
+Yoel's interpretation of the three paragraphs above.
+Male and female G.
+doryssus differ for some traits, indicating detectable sexual dimorphism.
+This is evidenced by traits with substantial differences in male and female posterior means (i.e., deviation from 0 in Figure \ref{fig:mu-diff-plot}), supported by posterior probabilities that deviate from 0.5 (Figure \ref{fig:prob0-plot}).
+For example, males are larger than females by 3-8mm on average through time (stl panel in Figure \ref{fig:mu-diff-plot}), with a posterior probability for that mean difference above 70 percent through time (Figure \ref{fig:prob0-plot}).
+In contrast, the number of abdominal vertebrae (mav) fluctuates about 0 (Figure \ref{fig:mu-diff-plot}), with posterior probabilities fluctuating about 50 percent through time (Figure \ref{fig:prob0-plot}).
+Table \ref{tab:prob0-table} in the Supplementary Material summarizes the posterior probabilities that male means differ from female means by trait and sample.
+}
+
+\textcolor{red}{can we move mu-diff-plot and prob0-plot here?}
+
 \hypertarget{changes-over-time}{%
@@
 of the observed time periods for a gain of 0.166 bones on average.
+
+\textcolor{red} {
+Yoel's interpretation of the five paragraphs above. 
+Figure \ref{fig:AUC-plot} shows the posterior probability that the mean male-female difference at time t is different from the male-female difference at time zero.
+In other words, Figure \ref{fig:AUC-plot} plots whether the sexual dimorphism changes in subsequent samples, always relative to the first sample.
+The amount of sexual dimporphism over time does not change substantially for a majority of the traits (i.e.
+cle, ds1, ds2, ect, lps, lpt, mav, mcv, pmx, stl; Figure \ref{fig:AUC-plot}).
+Traits that change in dimorphism are anal fin ray (maf), and dorsal fin ray (mdf), and dorsal spine 3 (ds3).}
+
+\textcolor{red} {
+For example, females at time zero averaged 8.29 rays in the anal fin, dropping to 8.17 rays in the final sample.
+Males averaged 8.52 anal fin rays in the first sample and 8.55 in the last sample.
+Thus, the difference between males and females started at 0.23 and ended at 0.38, an increase in sexual dimorphism.
+In contrast, the sign on dimorphism flipped for ds3.
+The posterior mean length of the third dorsal spine is 2.29mm for males and 2.45mm for females.
+By the end of the sequence, mean ds3 for males was 1.27 for males and 1.19 for females.
+Both sexes evolved reduction of ds3, but females lowered their mean more than the males.
+This resulted in a flip in dimorphism, from larger females to larger males, though both sexes evolved trait reduction.
+BUT DIMORPHISM SHRUNK THEN, AS THE DIFFERENCE BETWEEN MALES AND FEMALES IS NOW SMALLER??
+}
 
@@
 
+posterior mean difference for males versus females (values above 0
+indicate means that were greater for males versus females)
+
 \begin{figure}
@@
 
-\caption{Posterior mean difference between males and females}\label{fig:mu-diff-plot}
+\caption{Posterior mean difference between males and females, in UNITS?. Values above 0 indicate larger male means}\label{fig:mu-diff-plot}
 \end{figure}
@@
 
-\caption{Posterior probabilities that mean for males is greater than mean for females}\label{fig:prob0-plot}
+\caption{Posterior probabilities that the male posterior mean is larger than the female posterior mean}\label{fig:prob0-plot}
 \end{figure}
@@
 
-\caption{AUC comparing time x versus time 0.}\label{fig:AUC-plot}
+\caption{The Posterior Probability that male-female differences in subsequent samples are different from male-female differenceds in the initial sample.
+Values near 0.5 indicate little difference between the distributions.
+Values closer to 0 indicate that the difference in means has shifted towards larger means in females.
+Values closer to 1 indicate that the difference in means has shifted toward larger means in males.}\label{fig:AUC-plot}
 \end{figure}
@@
 \hypertarget{sec:conclusions}{%
-\section{Discsusson, Future work and
+\section{Discussion, Future work and
 conclusions}\label{sec:conclusions}}
@@
 
-We thank O. Abughoush, S. Blaine, A. Chaudhary, M. Islam, F. Joaquin, C.
+We thank O. Abughoush, S. Blain, A. Chaudhary, M. Islam, F. Joaquin, C.
 Lawson-Weinert, R. Sullivan, J. Tien, M.P. Travis, and W. Shim for help
@@
 \begin{CSLReferences}{1}{0}
-\leavevmode\vadjust pre{\hypertarget{ref-Aguirre2014}{}}%
-Aguirre, Windsor E, Kendal Walker, and Shawn Gideon. 2014. {``Tinkering
-with the Axial Skeleton: Vertebral Number Variation in Ecologically
-Divergent Threespine Stickleback Populations.''} \emph{Biol. J. Linn.
-Soc. Lond.} 113 (1): 204--19.
+\leavevmode\vadjust pre{\hypertarget{ref-Aguirreetal2014}{}}%
+Aguirre, W E, K Walker, and S Gideon. 2014. {``Tinkering with the Axial
+Skeleton: Vertebral Number Variation in Ecologically Divergent
+Threespine Stickleback Populations.''} \emph{Biological Journal of the
+Linnean Society} 113 (1): 204--19.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Baumgartneretal1988}{}}%
+Baumgartner, J V, M A Bell, and P H Weinberg. 1988. {``Body Form
+Differences Between the Enos Lake Species Pair of Threespine
+Sticklebacks (Gasterosteus Aculeatus Complex).''} \emph{Canadian Journal
+of Zoology} 66 (2): 467--74.
 
@@
 Bell, M A. 2009. {``Implications of a Fossil Stickleback Assemblage for
-Darwinian Gradualism.''} \emph{J. Fish Biol.} 75 (8): 1977--99.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Bell2006}{}}%
-Bell, Michael A, Matthew P Travis, and D Max Blouw. 2006. {``Inferring
-Natural Selection in a Fossil Threespine Stickleback.''}
-\emph{Paleobiology} 32 (4): 562--77.
-
-\leavevmode\vadjust pre{\hypertarget{ref-BellEtAl1993}{}}%
-Bell, Orti, M. A., and J. P. Koenings. 1993. {``Evolution of Pelvic
-Reduction in Threespine Sticklebacks: A Test of Competing Hypotheses.''}
-\emph{Evolution} 47: 906--14.
+Darwinian Gradualism.''} \emph{Journal of Fish Biology} 75 (8):
+1977--99.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Belletal1985}{}}%
+Bell, M A, J V Baumgartner, and E C Oslon. 1985. {``Patterns of Temporal
+Change in Single Morphological Characters of a Miocene Stickleback
+Fish.''} \emph{Paleobiology} 11: 258--71.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Belletal1993}{}}%
+Bell, M A, G Orti, J A Walker, and J P Koenings. 1993. {``Evolution of
+Pelvic Reduction in Threespine Sticklebacks: A Test of Competing
+Hypotheses.''} \emph{Evolution} 47: 906--14.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Belletal2006}{}}%
+Bell, M A, M P Travis, and D M Blouw. 2006. {``Inferring Natural
+Selection in a Fossil Threespine Stickleback.''} \emph{Paleobiology} 32
+(4): 562--77.
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Bolnick2011}{}}%
-Bolnick, Daniel I. 2011. {``Sympatric Speciation in Threespine
-Stickleback: Why Not?''} \emph{Int. J. Ecol.} 2011: 1--15.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Bolnick2003}{}}%
-Bolnick, Daniel I, and Michael Doebeli. 2003. {``Sexual Dimorphism and
-Adaptive Speciation: Two Sides of the Same Ecological Coin.''}
-\emph{Evolution} 57 (11): 2433--49.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Bolnick2010}{}}%
-Bolnick, Daniel I, Travis Ingram, William E Stutz, Lisa K Snowberg, On
-Lee Lau, and Jeff S Paull. 2010. {``Ecological Release from
-Interspecific Competition Leads to Decoupled Changes in Population and
-Individual Niche Width.''} \emph{Proc. Biol. Sci.} 277 (1689): 1789--97.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Bolnick2008}{}}%
-Bolnick, Daniel I, and On Lee Lau. 2008. {``Predictable Patterns of
-Disruptive Selection in Stickleback in Postglacial Lakes.''} \emph{Am.
-Nat.} 172 (1): 1--11.
+Bolnick, D I. 2011. {``Sympatric Speciation in Threespine Stickleback:
+Why Not?''} \emph{International Journal of Ecology} 2011: 1--15.
+
+\leavevmode\vadjust pre{\hypertarget{ref-BolnickandDoebeli2003}{}}%
+Bolnick, D I, and M Doebeli. 2003. {``Sexual Dimorphism and Adaptive
+Speciation: Two Sides of the Same Ecological Coin.''} \emph{Evolution}
+57 (11): 2433--49.
+
+\leavevmode\vadjust pre{\hypertarget{ref-BolnickandLau2008}{}}%
+Bolnick, D I, and O L Lau. 2008. {``Predictable Patterns of Disruptive
+Selection in Stickleback in Postglacial Lakes.''} \emph{The American
+Naturalist} 172 (1): 1--11.
 
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Butler2007}{}}%
-Butler, Marguerite A, Stanley A Sawyer, and Jonathan B Losos. 2007.
-{``Sexual Dimorphism and Adaptive Radiation in Anolis Lizards.''}
-\emph{Nature} 447 (7141): 202--5.
+\leavevmode\vadjust pre{\hypertarget{ref-Butleretal2007}{}}%
+Butler, Sawyer, M A, and J B Losos. 2007. {``Sexual Dimorphism and
+Adaptive Radiation in Anolis Lizards.''} \emph{Nature} 447 (7141):
+202--5.
 
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Cooper2011}{}}%
-Cooper, Idelle A, R Tucker Gilman, and Janette Wenrick Boughman. 2011.
-{``Sexual Dimorphism and Speciation on Two Ecological Coins: Patterns
-from Nature and Theoretical Predictions.''} \emph{Evolution} 65 (9):
-2553--71.
+\leavevmode\vadjust pre{\hypertarget{ref-Cerasonietal2024}{}}%
+Cerasoni, J N, M A Bell, and Y E Stuart. 2024. {``Geology,
+Microstratigraphy, and Paleontology of Truckee Formation Lacustrine
+Diatomite Deposits Near Hazen, Nevada, USA, with Emphasis on Fossil
+Stickleback Fish.''} \emph{PaleoBios} 41: 1--15.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Cooperetal2011}{}}%
+Cooper, I A, R T Gilman, and J W Boughman. 2011. {``Sexual Dimorphism
+and Speciation on Two Ecological Coins: Patterns from Nature and
+Theoretical Predictions.''} \emph{Evolution} 65 (9): 2553--71.
 
 \leavevmode\vadjust pre{\hypertarget{ref-Cornault2022}{}}%
-Cornuault, Josselin. 2022. {``Bayesian Analyses of Comparative Data with
-the Ornstein--Uhlenbeck Model: Potential Pitfalls.''} \emph{Systematic
+Cornuault, J. 2022. {``Bayesian Analyses of Comparative Data with the
+Ornstein--Uhlenbeck Model: Potential Pitfalls.''} \emph{Systematic
 Biology} 71 (6): 1524--40. \url{https://doi.org/10.1093/sysbio/syac036}.
 
-\leavevmode\vadjust pre{\hypertarget{ref-deLisleNco2018}{}}%
-De Lisle, S. Paiva, S. P., and L. Rowe. 2018. {``Habitat Partitioning
-During Character Displacement Between the Sexes.''} \emph{Biology
-Letters} 14: 20180124.
-
-\leavevmode\vadjust pre{\hypertarget{ref-deLisleRowe2015}{}}%
-De Lisle, S. P., and L. Rowe. 2015. {``Ecological Character Displacement
+\leavevmode\vadjust pre{\hypertarget{ref-DeLisleetal2018}{}}%
+De Lisle, S P, S Paiva, and L Rowe. 2018. {``Habitat Partitioning During
+Character Displacement Between the Sexes.''} \emph{Biology Letters} 14:
+20180124.
+
+\leavevmode\vadjust pre{\hypertarget{ref-DeLisleandRowe2015}{}}%
+De Lisle, S P, and L Rowe. 2015. {``Ecological Character Displacement
 Between the Sexes.''} \emph{The American Naturalist} 186: 693--707.
 
-\leavevmode\vadjust pre{\hypertarget{ref-Hendry2009}{}}%
+\leavevmode\vadjust pre{\hypertarget{ref-DeLisleandRowe2017}{}}%
+---------. 2017. {``Disruptive Natural Selection Predicts Divergence
+Between the Sexes During Adaptive Radiation.''} \emph{Ecology and
+Evolution} 7: 3590--3601.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Godfreyetal1993}{}}%
+Godfrey, L R, S K Lyon, and M R Sutherland. 1993. {``Sexual Dimorphism
+in Large-Bodied Primates: The Case of the Subfossil Lemurs.''}
+\emph{American Journal of Physical Anthropology} 90: 315--34.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Hendryetal2009}{}}%
 Hendry, A P, D I Bolnick, D Berner, and C L Peichel. 2009. {``Along the
-Speciation Continuum in Sticklebacks.''} \emph{J. Fish Biol.} 75 (8):
-2000--2036.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Herrmann2021}{}}%
-Herrmann, Nicholas C, James T Stroud, and Jonathan B Losos. 2021. {``The
-Evolution of 'Ecological Release' into the 21st Century.''} \emph{Trends
-Ecol. Evol.} 36 (3): 206--15.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Hone2017}{}}%
-Hone, David W E, and Jordan C Mallon. 2017. {``Protracted Growth Impedes
-the Detection of Sexual Dimorphism in Non-Avian Dinosaurs.''}
+Speciation Continuum in Sticklebacks.''} \emph{Journal of Fish Biology}
+75 (8): 2000--2036.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Herrmannetal2021}{}}%
+Herrmann, N C, J T Stroud, and J B Losos. 2021. {``The Evolution of
+'Ecological Release' into the 21st Century.''} \emph{Trends in Ecology
+and Evolution} 36 (3): 206--15.
+
+\leavevmode\vadjust pre{\hypertarget{ref-HoneandMallon2017}{}}%
+Hone, D W E, and J C Mallon. 2017. {``Protracted Growth Impedes the
+Detection of Sexual Dimorphism in Non-Avian Dinosaurs.''}
 \emph{Palaeontology} 60 (4): 535--45.
 
-\leavevmode\vadjust pre{\hypertarget{ref-Hunt2008}{}}%
-Hunt, Gene, Michael A Bell, and Matthew P Travis. 2008. {``Evolution
-Toward a New Adaptive Optimum: Phenotypic Evolution in a Fossil
-Stickleback Lineage.''} \emph{Evolution} 62 (3): 700--710.
+\leavevmode\vadjust pre{\hypertarget{ref-Huntetal2008}{}}%
+Hunt, G, M A Bell, and M P Travis. 2008. {``Evolution Toward a New
+Adaptive Optimum: Phenotypic Evolution in a Fossil Stickleback
+Lineage.''} \emph{Evolution} 62 (3): 700--710.
+
+\leavevmode\vadjust pre{\hypertarget{ref-KoscinskiandPietraszewski2004}{}}%
+Koscinski, K, and S Pietraszewski. 2004. {``Methods to Estimate Sexual
+Dimorphism from Unsexed Samples: A Test with Computer Generated
+Samples.''} \emph{Przeglad Antropologiczny - Anthropological Review} 67:
+33--55.
 
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Reimchen2016}{}}%
-Reimchen, D. Steeves, T. E. 2016. {``Sex Matters for Defence and Trophic
-Traits of Threespine Stickleback.''} \emph{Evolutionary Ecology
-Research} 17: 459-485-.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Reimchen1980}{}}%
-Reimchen, T. E. 1980. {``Spine Deficiency and Polymorphism in a
-Poulation of Gasterosteus Aculeatus: An Adaptation to Predators?''}
-\emph{Canadian Journal of Zoology} 58: 1232--44.
+\leavevmode\vadjust pre{\hypertarget{ref-Mallon2017}{}}%
+Mallon, J C. 2017. {``Recognizing Sexual Dimorphism in the Fossil
+Record: Lessons from Nonavian Dinosaurs.''} \emph{Paleobiology} 43 (3):
+495--507.
+
+\leavevmode\vadjust pre{\hypertarget{ref-ParentandCrespi2009}{}}%
+Parent, C E, and B J Crespi. 2009. {``Ecological Opportunity in Adaptive
+Radiation of Galapagos Endemic Land Snails.''} \emph{The American
+Naturalist} 174: 898--905.
+
+\leavevmode\vadjust pre{\hypertarget{ref-PfennigandPfennig2012}{}}%
+Pfennig, D W, and K S Pfennig. 2012. \emph{\href{}{Evolution's Wedge:
+Competition and the Origins of Diversity}}. University of California
+Press, Berkeley, USA.
 
 \leavevmode\vadjust pre{\hypertarget{ref-Reimchen1994}{}}%
----------. 1994. {``Predators and Morphological Evolution in Threespine
-Stickleback.''} In \emph{The Evolutionary Biology of the Threespine
-Stickleback}, edited by M A Bell and S A Foster. Oxford, UK: Oxford
-University Press.
-
-\leavevmode\vadjust pre{\hypertarget{ref-ReimchenNosil2004}{}}%
-Reimchen, T. E., and P. Nosil. 2004. {``Variable Predation Regimes
-Predict the Evolution of Sexual Dimorphism in a Population of Threespine
+Reimchen, T E. 1994. {``Predators and Morphological Evolution in
+Threespine Stickleback.''} In \emph{The Evolutionary Biology of the
+Threespine Stickleback}, edited by M A Bell and S A Foster. Oxford, UK:
+Oxford University Press.
+
+\leavevmode\vadjust pre{\hypertarget{ref-ReimchenandNosil2004}{}}%
+Reimchen, T E, and P Nosil. 2004. {``Variable Predation Regimes Predict
+the Evolution of Sexual Dimorphism in a Population of Threespine
 Stickleback.''} \emph{Evolution} 58 (6): 1274--81.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Roestietal2023}{}}%
-Roesti, J. S. Groh, M, and D. Schluter. 2023. {``Species Divergence
-Under Competition and Shared Predation.''} \emph{Ecology Letters} 26:
-111--23.
-
-\leavevmode\vadjust pre{\hypertarget{ref-SaittaEtAl2020}{}}%
-Saitta, Evan T, Maximilian T Stockdale, Nicholas R Longrich, Vincent
-Bonhomme, Michael J Benton, Innes C Cuthill, and Peter J Makovicky.
-2020. {``An Effect Size Statistical Framework for Investigating Sexual
-Dimorphism in Non-Avian Dinosaurs and Other Extinct Taxa.''}
-\emph{Biological Journal of the Linnean Society} 131 (2): 231--73.
-\url{https://doi.org/10.1093/biolinnean/blaa105}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Schluter1992}{}}%
+Roesti, M, J S Groh, S A Blain, M Huss, P Rassias, D I Bolnick, Y E
+Stuart, C L Peichel, and D Schluter. 2023. {``Species Divergence Under
+Competition and Shared Predation.''} \emph{Ecology Letters} 26: 111--23.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Saittaetal2020}{}}%
+Saitta, T, M T Stockdale, N R Longrich, V Bonhomme, M J Benton, I C
+Cuthill, and P J Makovicky. 2020. {``An Effect Size Statistical
+Framework for Investigating Sexual Dimorphism in Non-Avian Dinosaurs and
+Other Extinct Taxa.''} \emph{Biological Journal of the Linnean Society}
+131 (2): 231--73. \url{https://doi.org/10.1093/biolinnean/blaa105}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Schindelinetal2012}{}}%
+Schindelin, J, I Arganda-Carreras, E Frise, V Kaynig, M Longair, T
+Pietzsch, S Preibisch, C Rueden, S Saalfeld, and B Schmid. 2012.
+{``Fiji: An Open-Source Platform for Biological-Image Analysis.''}
+\emph{Nature Methods} 9: 976--682.
+
+\leavevmode\vadjust pre{\hypertarget{ref-SchluterandMcPhail1992}{}}%
 Schluter, D, and J D McPhail. 1992. {``Ecological Character Displacement
-and Speciation in Sticklebacks.''} \emph{Am. Nat.} 140 (1): 85--108.
+and Speciation in Sticklebacks.''} \emph{The American Naturalist} 140
+(1): 85--108.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Schoener1968}{}}%
+Schoener, T W. 1968. {``The Anolis Lizards of Bimini: Resource
+Partitioning in a Complex Fauna.''} \emph{Ecology} 29: 704--26.
 
@@
 Shine, R. 1989. {``Ecological Causes for the Evolution of Sexual
-Dimorphism: A Review of the Evidence.''} \emph{Q. Rev. Biol.} 64 (4):
-419--61.
-
-\leavevmode\vadjust pre{\hypertarget{ref-SpoljaricReimchen2008}{}}%
-Spoljaric, M. A., and T. E. Reimchen. 2008. {``Habitat-Dependent
-Reduction of Sexual Dimorphism in Geometric Body Shape of Haida Gwaii
-Threespine Stickleback.''} \emph{Biological Journal of the Linnean
-Society} 95: 505--16.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Stuart2021}{}}%
-Stuart, Yoel E, J William Sherwin, Ambika Kamath, and Thor Veen. 2021.
-{``Male and Female Anolis Carolinensis Maintain Their Dimorphism Despite
-the Presence of Novel Interspecific Competition.''} \emph{Evolution} 75
-(11): 2708--16.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Stuart2020}{}}%
-Stuart, Yoel E, Matthew P Travis, and Michael A Bell. 2020. {``Inferred
-Genetic Architecture Underlying Evolution in a Fossil Stickleback
-Lineage.''} \emph{Nat. Ecol. Evol.} 4 (11): 1549--57.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Stuart2017}{}}%
-Stuart, Yoel E, Thor Veen, Jesse N Weber, Dieta Hanson, Mark Ravinet,
-Brian K Lohman, Cole J Thompson, et al. 2017. {``Contrasting Effects of
-Environment and Genetics Generate a Continuum of Parallel Evolution.''}
-\emph{Nat. Ecol. Evol.} 1 (6): 158.
+Dimorphism: A Review of the Evidence.''} \emph{The Quarterly Review of
+Biology} 64 (4): 419--61.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Siddiquietal2024}{}}%
+Siddiqui, R, S Swank, A Ozark, F Joaquin, M P Travis, C D McMahan, M A
+Bell, and Y E Stuart. 2024. {``Inferring the Evolution of Reproductive
+Isolation in a Lineage of Fossil Threespine Stickleback, Gasterosteus
+Doryssus.''} \emph{Proceedings of the Royal Society B} 291: 20240337.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Stuartetal2021}{}}%
+Stuart, Y E, J W Sherwin, A Kamath, and T Veen. 2021. {``Male and Female
+Anolis Carolinensis Maintain Their Dimorphism Despite the Presence of
+Novel Interspecific Competition.''} \emph{Evolution} 75 (11): 2708--16.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Stuartetal2020}{}}%
+Stuart, Y E, M P Travis, and M A Bell. 2020. {``Inferred Genetic
+Architecture Underlying Evolution in a Fossil Stickleback Lineage.''}
+\emph{Nature Ecology and Evolution} 4 (11): 1549--57.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Stuartetal2017}{}}%
+Stuart, Y E, T Veen, J N Weber, D Hanson, M Ravinet, B K Lohman, C J
+Thompson, et al. 2017. {``Contrasting Effects of Environment and
+Genetics Generate a Continuum of Parallel Evolution.''} \emph{Nature
+Ecology and Evolution} 1 (6): 158.
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-OUProcess}{}}%
-Uhlenbeck, G. E., and L. S. Ornstein. 1930. {``On the Theory of the
-Brownian Motion.''} \emph{Phys. Rev.} 36 (September): 823--41.
+Uhlenbeck, G E, and L S Ornstein. 1930. {``On the Theory of the Brownian
+Motion.''} \emph{Phys. Rev.} 36 (September): 823--41.
 \url{https://doi.org/10.1103/PhysRev.36.823}.
 
-\leavevmode\vadjust pre{\hypertarget{ref-Voje2022}{}}%
-Voje, Kjetil L, Michael A Bell, and Yoel E Stuart. 2022. {``Evolution of
-Static Allometry and Constraint on Evolutionary Allometry in a Fossil
-Stickleback.''} \emph{J. Evol. Biol.} 35 (3): 423--38.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Schoener1968}{}}%
-W., Schoener T. 1968. {``The Anolis Lizards of Bimini: Resource
-Partitioning in a Complex Fauna.''} \emph{Ecology} 49: 704--26.
+\leavevmode\vadjust pre{\hypertarget{ref-VamosiandSchluter2004}{}}%
+Vamosi, S M, and D Schluter. 2004. {``Character Shifts in the Defensive
+Armor of Sympatric Sticklebacks.''} \emph{Evolution} 58: 376--85.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Vojeetal2022}{}}%
+Voje, K L, M A Bell, and Y E Stuart. 2022. {``Evolution of Static
+Allometry and Constraint on Evolutionary Allometry in a Fossil
+Stickleback.''} \emph{Journal of Evolutionary Biology} 35 (3): 423--38.
 
 \leavevmode\vadjust pre{\hypertarget{ref-ZhouReiter2010}{}}%
-Zhou, X., and J. Reiter. 2010. {``A Note on Bayesian Inference After
+Zhou, X, and J Reiter. 2010. {``A Note on Bayesian Inference After
 Multiple Imputation.''} \emph{The American Statistician} 64 (2):

### 2024-07-30 manuscript/paper.Rmd
rearranging author order

@@
 author: | 
-  | \large Allison Ozark$^{1,*}$, Matthew Stuart$^{2,3,*}$, Raheyma Siddiqui$^1$, Akhil Ghosh$^{2,3}$
-  | \large Samantha Swank$^1$, Michael A. Bell$^4$, Gregory J. Matthews$^{2,3}$, and Yoel E. Stuart$^{1,+}$ 
+  | \large Matthew Stuart$^{1,2}$, Allison Ozark$^{3}$, Raheyma Siddiqui$^3$, Akhil Ghosh$^{1,2}$
+  | \large Samantha Swank$^3$, Michael A. Bell$^4$, Gregory J. Matthews$^{1,2}$, and Yoel E. Stuart$^{3,+}$ 
   | \vspace{-1.1mm}
-  | \large $^1$ Department of Biology, Loyola University Chicago, Chicago, IL, USA \vspace{-1.1mm}
-  | \large $^2$ Department of Mathematics and Statistics, Loyola University Chicago, Chicago, IL, USA \vspace{-1.1mm}
-  | \large $^3$ Center for Data Science and Consulting, Loyola University Chicago, Chicago, IL, USA \vspace{-1.1mm}
+  | \large $^1$ Department of Mathematics and Statistics, Loyola University Chicago, Chicago, IL, USA \vspace{-1.1mm}
+  | \large $^2$ Center for Data Science and Consulting, Loyola University Chicago, Chicago, IL, USA \vspace{-1.1mm}
+  | \large $^3$ Department of Biology, Loyola University Chicago, Chicago, IL, USA \vspace{-1.1mm}
   | \large $^4$ University of California Museum of Paleontology, Berkeley, CA, USA \vspace{-1.1mm}
-  | \large $^*$ Equal contribution \vspace{-1.1mm}
   | \large $^+$ Corresponding: ystuart@luc.edu \vspace{-1.1mm}

### 2024-09-18 manuscript/paper.tex
some manuscript work

@@
 stickleback following ecological release from fish piscivores}
-\author{\large Allison Ozark\(^{1,*}\), Matthew Stuart\(^{2,3,*}\),
-Raheyma Siddiqui\(^1\), Akhil Ghosh\(^{2,3}\)\\
-\large Samantha Swank\(^1\), Michael A. Bell\(^4\), Gregory J.
-Matthews\(^{2,3}\), and Yoel E. Stuart\(^{1,+}\)\\
+\author{\large Matthew Stuart\(^{1,2}\), Allison Ozark\(^{3}\), Raheyma
+Siddiqui\(^3\), Akhil Ghosh\(^{1,2}\)\\
+\large Samantha Swank\(^3\), Michael A. Bell\(^4\), Gregory J.
+Matthews\(^{1,2}\), and Yoel E. Stuart\(^{3,+}\)\\
 \vspace{-1.1mm}\\
-\large \(^1\) Department of Biology, Loyola University Chicago, Chicago,
+\large \(^1\) Department of Mathematics and Statistics, Loyola
+University Chicago, Chicago, IL, USA \vspace{-1.1mm}\\
+\large \(^2\) Center for Data Science and Consulting, Loyola University
+Chicago, Chicago, IL, USA \vspace{-1.1mm}\\
+\large \(^3\) Department of Biology, Loyola University Chicago, Chicago,
 IL, USA \vspace{-1.1mm}\\
-\large \(^2\) Department of Mathematics and Statistics, Loyola
-University Chicago, Chicago, IL, USA \vspace{-1.1mm}\\
-\large \(^3\) Center for Data Science and Consulting, Loyola University
-Chicago, Chicago, IL, USA \vspace{-1.1mm}\\
 \large \(^4\) University of California Museum of Paleontology, Berkeley,
 CA, USA \vspace{-1.1mm}\\
-\large \(^*\) Equal contribution \vspace{-1.1mm}\\
 \large \(^+\) Corresponding:

### 2024-09-19 manuscript/paper.Rmd
Edit to Introduction, which is mostly done. Edits to methods and results which need joint writing.

@@
 
-Ecological release theory suggests that a population’s niche should change when important species interactions (e.g., resource competition) are relaxed or removed (reviewed in @Herrmannetal2021).
-Removal is posited to create ecological opportunities—i.e., aspects of the niche become newly accessible, and the focal population shifts and/or expands its resource use to the new opportunity (@ParentandCrespi2009; @Herrmannetal2021).
-Ecological release may then be followed by adaptive morphological evolution (@ParentandCrespi2009; @Herrmannetal2021).
+Ecological release theory suggests that a population’s niche should change when important species interactions like resource competition or predation are relaxed or removed (reviewed in @Herrmannetal2021).
+Removal is posited to create ecological opportunity—i.e., aspects of the niche become newly accessible, and the focal population shifts and/or expands its resource use new resources (@ParentandCrespi2009; @Herrmannetal2021).
+Ecological release may then be followed by adaptive morphological evolution as traits change to reflect the new niche (@ParentandCrespi2009; @Herrmannetal2021).
+
 For example, a population undergoing ecological release can experience disruptive selection on males and females stemming from intraspecific competition over newly accessible resources (@BolnickandDoebeli2003; @BolnickandLau2008; @Cooperetal2011).
@@
 Much of the theory and empirical data for character displacement between the sexes is based on release from resource competition specifically (e.g., @BolnickandDoebeli2003; @Cooperetal2011; @PfennigandPfennig2012; @DeLisleandRowe2015; @DeLisleetal2018).
-However, release from predation should also result in ecological release followed by the evolution of increased sexual dimorphism because the absence of predators should generate ecological opportunity (@ReimchenandNosil2004; @ParentandCrespi2009; @Herrmannetal2021).
+However, release from predation might also drive the evolution of increased sexual dimorphism because an absence of predators should generate ecological opportunity (@ReimchenandNosil2004; @ParentandCrespi2009; @Herrmannetal2021).
 
-Here, we tested the prediction that a release from predation should result in the evolution of increased sexual dimorphism using a well-preserved, finely-resolved sequence of a fossil threespine stickleback fish (Gasterosteus doryssus).
+Here, we tested the prediction that a release from predation results in the evolution of increased sexual dimorphism using a well-preserved, finely-resolved sequence of a fossil threespine stickleback fish (Gasterosteus doryssus).
 The sequence in this depositional environment is comprised of two lineages (@Bell2009; @Cerasonietal2024).
 Lineage I was a low-armor form with zero to one dorsal spines and highly reduced pelvises, on average.
-It lasted for at least 93,000 years before it was replaced suddenly by a second lineage (@Belletal1985).
-Lineage II appeared in the depositional environment fully armored, with complete pelvic girdles, two pelvic spines, and three dorsal spines, on average (@Belletal2006; @Stuartetal2020).
+Lineage I lasted for at least 93,000 years before it was replaced by a second lineage, suddenly, likely on the order of years (@Belletal1985; Stuart et al.
+unpublished data).
+Lineage II appeared in the depositional environment fully armored, with complete pelvic girdles, two pelvic spines, and three dorsal spines (@Belletal2006; @Stuartetal2020).
 It then immediately began evolving adaptive reduction in its armor traits (@Huntetal2008; @Stuartetal2020) until it reached the same low-armor state previously held by lineage I.
-The observation of 93,000 years of low-armor stasis in lineage I (@Belletal1985) and the observation of rapid evolution of armor loss by lineage II, both suggest that this depositional environment lacked piscivorous fish (e.g., trout and other salmonids known to prey on modern threespine stickleback).
-Armor maintenance in extant threespine stickleback (Gasterosteus aculeatus) correlates strongly the presence of vertebrate piscivores; populations with less predation pressure typically have less armor (@Reimchen1994, @ReimchenandNosil2004; @Belletal1993; @Roestietal2023).
-Consistently, only three fossil trout have been recorded from this paleolake basin, sampled from the same rock that has revealed >15,000 threespine stickleback fossils as well as occasional killifish (Fundulus nevadensis) (@Bell2009; @Cerasonietal2024).
-Finally, tooth wear data from the lineage II sequence suggest that individuals in this lineage began eating more planktonic prey over time, expanding toward an open-water niche from the benthic, bottom-feeding niche they started with (Purnell et al.
-2007).
-That this expansion into open water by lineage II coincided with armor reduction is further indicative of a limnetic system with fewer salmonid piscivores (@SchluterandMcPhail1992; @VamosiandSchluter2004; @Roestietal2023).
-In sum, we infer that lineage II likely started in a nearby paleolake basin that had predators since it was armored when it arrived (@Bell2009; @Cerasonietal2024).
-Lineage II then experienced release from predators and dietary expansion in the focal paleolake basin, generating a scenario consistent with the starting conditions for models predicting the evolution of increased sexual dimorphism following ecological release.
+
+The observation of 93,000 years of low-armor stasis in lineage I (@Belletal1985) and the observation of rapid evolution of armor loss by lineage II, both suggest that this depositional environment lacked piscivorous fish like trout and other salmonids known to prey on modern threespine stickleback.
+Armor presence in extant threespine stickleback (Gasterosteus aculeatus) correlates strongly with the presence of vertebrate piscivores; populations with less predation pressure typically have less armor (@Reimchen1994, @ReimchenandNosil2004; @Belletal1993; @Roestietal2023).
+In our paleolake basin, only three fossil trout have been found in the same section of rock that has revealed >20,000 threespine stickleback fossils as well as occasional killifish (Fundulus nevadensis) (@Bell2009; @Cerasonietal2024).
+Thus, we reconstruct an evolutionar history in which it is likely that lineage II migrated from a nearby paleolake basin that had predators since it was armored when it arrived (@Bell2009; @Cerasonietal2024).
+Lineage II then experienced release from predators in the focal paleolake basin, generating the initial conditions of evolutionary models predicting the evolution of increased sexual dimorphism following ecological release.
+We test this prediction here.
 
@@
 We applied this algorithm to the fossil data to impute individual sex based on morphology, treating fossil sex as a missing variable.
-With sex assigned to fossil specimens (and accounting for uncertainty in those assignments), we fit a modified Ornstein-Uhlenbeck (OU) (@OUProcess) model using a Bayesian framework to test for evolution of sexual dimorphism in each trait over ~16,000 years of lineage II.
-We hypothesized that sexual dimorphism should increase following ecological release from predators.
+With sex assigned to fossil specimens, we fit a modified Ornstein-Uhlenbeck (OU) (@OUProcess) model using a Bayesian framework that accounts for uncertainty in sex classification to test for evolution of sexual dimorphism in each trait over ~16,000 years of lineage II.
 
@@
 We used Gasterosteus doryssus data that were previously reported by @Stuartetal2020, @Vojeetal2022, and @Siddiquietal2024.
-Briefly, the data were collected from fossil Series K from Quarry D (@Cerasonietal2024), dug from an open pit diatomite mine at 9.526° N, 119.094° W, near Fernley, Nevada, USA.
+Briefly, the data were collected from fossil Series K from Quarry D (@Cerasonietal2024), dug from an open pit diatomite mine at 9.526° N, 119.094° W, near Hazen, Nevada, USA.
 Series K consisted of 18 samples taken at ~1000-year intervals, and mean sample times span ~16,363 years.
 Fish from series K were measured for 16 ecomorphological traits related to armor, swimming, and feeding (Table 1).
-Series K started at a previously documented horizon when a low armored lineage of stickleback with zero to one dorsal spines, zero pelvic spines, and highly reduced pelvises (i.e., lineage I described above) was completely replaced by a high armored lineage of stickleback with three dorsal spines, two pelvic spines, and complete pelvises (i.e., lineage II described above) (@Belletal2006; @Bell2009; @Stuartetal2020).
-Lineage II subsequently evolved reduction in armor, body size, and traits related to swimming and feeding (@Belletal2006; @Stuartetal2020; @Siddiquietal2024).
-The tempo and mode of armor reduction during this sequence suggests adaptive evolution by natural selection (@Huntetal2008), and we focus on the multivariate evolution of sexual dimorphism by this second lineage.
+Series K started at the previously documented horizon when lineage I was replaced by lineage II.
+The tempo and mode of lineage II armor reduction during this sequence suggests adaptive evolution by natural selection (@Huntetal2008), and we focus on the multivariate evolution of sexual dimorphism by this second lineage.
 
@@
 To span the gamut of stickleback diversity for our predictive imputation model, we sampled modern stickleback from lakes containing generalist stickleback populations (@Hendryetal2009; @Bolnick2011) and from lakes containing benthic-limnetic species pairs (@Baumgartneretal1988; @SchluterandMcPhail1992).
-The generalist populations used here were the remainder of collections made by YES in 2013 and were previously described in @Stuartetal2017 (Table S1).
+The generalist specimens used here were collected by YES in 2013 and were previously described in @Stuartetal2017 (Table S1).
 These samples were fixed in formalin, then stained for bone with Alizarin Red in 2013.
 Benthic and limnetic specimens were kindly loaned by D.
-Schluter and his lab at University of British Columbia.
-They collected benthic and limnetic individuals from Enos Lake in 1988 and from Emily Lake, Little Quarry Lake, Paxton Lake, and Priest Lake in 2018 (Table S1).
+Schluter and S.
+Blain at the University of British Columbia.
+The Schluter lab collected benthic and limnetic individuals from Enos Lake in 1988 and from Emily Lake, Little Quarry Lake, Paxton Lake, and Priest Lake in 2018 (Table S1).
 The Enos specimens had been fixed whole in formalin and stored in 40% isopropanol.
@@
 We regressed each trait on standard length using a mixed-model regression, pooling all specimens, following @Stuartetal2017.
-We appended the size corrected data to the uncorrected trait data frame and used all of our data to build the imputation model described next.
+We appended the size corrected data to the uncorrected trait data frame and used all of our data (raw and size-corrected) to build the imputation model described next.
 
 ## Missing data imputation, including fossil sex
-We used multiple imputation (@little2002statistical) to impute the sex of the fossils many times.
-Imputations were performed using the MICE (@MICE) algorithm and implemented in R (@R2022language) with M = 100 imputations performed.
-After imputing the sex of the fossil data, each completed data set is then used to fit a modified Ornstein–Uhlenbeck (OU) (@OUProcess) model using a Bayesian framework to look for sexual dimorphism across a variety of stickleback phenotypes.
+Briefly, We used multiple imputation (@little2002statistical) to impute the sex of the fossils.
+Imputations were performed using the MICE (@MICE) algorithm and implemented in R (@R2022language) with M = 100 imputations.
+After imputing the sex, we fit a modified Orenstein-Uhlenbeck (OU; @OUProcess) model to each imputed data set using a Bayesian framework to {THIS NEEDS MORE}.
+We then quantified sexual dimorphism as the {THIS NEEDS MORE; MAYBE WE CAN WRITE TOGETHER}.
+We describe the details of the imputation model and OU modeling next.
 
@@
 
-We impute the missing sex for the fossil data by sampling from the posterior predictive distribution $P(\boldsymbol{W}_{fossil}|\boldsymbol{W}_{extant},\boldsymbol{Y})$ using the multiple imputation by chained equations (MICE) algorithm (@MICE) with predictive mean matching.
+\textcolor{red}{AKHIL'S VALIDATION OF THE IMPUTATION MODEL GOES HERE}
+
+We impute missing sex for the fossil data by sampling from the posterior predictive distribution $P(\boldsymbol{W}_{fossil}|\boldsymbol{W}_{extant},\boldsymbol{Y})$ using the multiple imputation by chained equations (MICE) algorithm (@MICE) with predictive mean matching.
 Traditionally, the choice for the number of completed dats sets is a relatively small number such as $M = 5$ or $M = 10$.
-However, @ZhouReiter2010 recommend a larger number of imputed data sets if the data users intend on performing Bayesians analysis after imputation, which in this case, we do.
-Therefore, the imputation algorithm is run to obtain a total of $M = 100$ completed datasets.
-In addition to this, @ZhouReiter2010 suggests rather than using Rubin's combining rules to combine across imputed data sets, instead pool all of the draws from the posterior distributions across all of the imputed data sets to estimate the posterior distributions of parameters of interest.
-We proceed with our Bayesian analysis in this manner.
+However, @ZhouReiter2010 recommend a larger number of imputed data sets if users intend on performing Bayesian analysis after imputation.
+Therefore, we imputed $M = 100$ complet datasets.
 
-\textcolor{red}{Akhil's stuff about validating the imputation model goes here.}
+\textcolor{red}{presumably this also imputes missing data in the other traits? We should probably say that explicitly. -YS}
+
+We then pooled all of the draws from the posterior distribution across all the imputed data sets to estimate the posterior distributions of our parameters of interest (i.e., the trait values), rather than usingRubin's combining rules, following @ZhouReiter2010.
+We used these posterior distributions in our Bayesian analysis.
 
@@
 
-## Completed Data Model 
-\textcolor{red}{We should note somewhere that we are only using the $W_{fossil}$ in the modeling part.
-We drop the $W_{extant}$.
-So just note that $W_{ti}$ is really $W_{fossil,ti}$.
-Not sure how to say this, but we need to make it clear that we are only using the fossil data for the OU modeling.
-Greg, check and see if this makes sense.}
+## Bayesian OU modeling
+For Bayesian modeling, we dropped the extant data and focused on modeling the evolution of sexual dimorphism through time in the fossils.
 
-\textcolor{red}{I think we need to move a lot of these details to the appendix.  }
-
+\textcolor{red}{DOES THE NEXT SENTENCE CONTRADICT THE PREVIOUS ONE? BELOW, WE SAY THAT WE USED EACH IMPUTED DATA SET. ABOVE, WE SAY THAT WE POOLED ACROSS IMPUTED DATA SETS.-YS}
 For a given imputed dataset, let $W_{ij}$ be the imputed sex and $\boldsymbol{Y}_{ij}$ be the $K \times 1$ vector of phenotypes for stickleback fossil $j$ at time $t_i$ where $i = 1, \ldots, T$ and $j = 1,\ldots,n_{t}$.
-Note, we are inputting only the fossil data into our model; connecting to the previous section, $W_{ij}$ and $\boldsymbol{Y}_{ij}$ can be interpretted as $W_{fossil,ij}$ and $\boldsymbol{Y}_{fossil,ij}$.
+Note, we are inputting only the fossil data into our model; connecting to the previous section, $W_{ij}$ and $\boldsymbol{Y}_{ij}$ can be interpreted as $W_{fossil,ij}$ and $\boldsymbol{Y}_{fossil,ij}$.
 In addition, we denote $Y_{K,ij}$, the last variable in $\boldsymbol{Y}_{ij}$, to be the standard length of the fish,
@@
 
-Because we are fitting a dataset with a stochastic structure on the means of the phenotypes, and we are interested in the distribution of the overall structure of these means, we analyze the data via a Bayesian analysis.
+Because we are fitting a dataset with a stochastic structure on phenotype means, and we are interested in the distribution of the overall structure of these means, we analyze the data via a Bayesian analysis <THIS LAST CLAUSE DOESN'T FOLLOW FROM THE BECAUSE CLAUSE.
+BETTER EXPLAIN WHY BAYES IS GOOD HERE>.
 Bayesian data analysis is also more naturally used when we have to impute data (\textbf{CITATION}).
@@
 # Results {#sec:results}
-As previously described, four models for the mean process were considered for each trait.
-Within each trait, the model with the lowest DIC was chosen for that train.
-All subsequent results presented here use the model with the lowest DIC individually by trait.
-Table \ref{whichmodel} shows the model that was selected for each trait.
-About half of the traits had models with a trend, and most models did not include an OU component.
-The traits that did include an OU component were caudal vertebra (mcv), pelvic spine lentgh (lpt), abdominal vertebra (mav), and standard length (stl) with caudal vertebra (mcv) being the only model with both a trend and an OU component.
-This implies that not only does caudal vertebra size have a linear trend across sex, but the average size at one time period is highly correlated with the average size at the next time period.<!-- NOTE FOR MATT: You should mention something about MCV being the only one with a trend and an OU.
-Explain why this might be and what it means.
--->
+<WE NEED RESULTS SECTION DESCRIBING THE RESULTS OF AKHIL'S MULTIPLE IMPUTATION, BOTH THE TESTS OF HOW WELL THE MODEL WORKED CLASSIFYING MODEL SAMPLES, AND SUMMARY STATISTICS OF FOSSIL IMPUTATION RESULTS -YS>
+
+Within each trait, the model with the lowest DIC was chosen for that trait <THIS IS METHODS--WHERE IS THIS DESCRIBED IN THE METHODS>.
+All subsequent results presented here use the model with the lowest DIC individually by trait <THIS IS METHODS--WHERE IS THIS DESCRIBED IN THE METHODS>.
+
+Table \ref{whichmodel} reports the results of model fitting.
+X of the Y traits were best fit by a trend model.
+Best models included an OU component for caudal and abdominal vertebra number (mcv, mav), pelvic spine length (lpt <LPT IS PTERYGIOPHORE LENGTH>) and standard length (stl).
+Caudal vertebra number was the only trait best fit by a model including both a trend and an OU component.
+This implies that not only does caudal vertebra size have a linear trend across sex <WHAT DO YOU MEAN BY LINEAR TREND ACROSS SEX?>, but the average size at one time period is highly correlated with the average size at the next time period <I'VE ALWAYS FOUND THAT THIS AID TO MODEL INTEPRETATION IS BETTER IN THE METHODS, JUST AFTER THE MODELS ARE DESCRIBED>.
 
@@
 dimorphism.
+IS THIS CORRECT?
+- YS
 }
@@
 Reimchen papers in general good for this section.
+Moreover, tooth wear data from the lineage II sequence suggest that individuals in this lineage began eating more planktonic prey over time, expanding toward an open-water niche from the benthic, bottom-feeding niche they started with (Purnell et al.
+2007).
+That this expansion into open water by lineage II coincided with armor reduction is further indicative of a limnetic system with fewer salmonid piscivores (@SchluterandMcPhail1992; @VamosiandSchluter2004; @Roestietal2023).
 

### 2024-09-26 manuscript/paper.tex
changes caused by knitting the Rmd

@@
 Ecological release theory suggests that a population's niche should
-change when important species interactions (e.g., resource competition)
-are relaxed or removed (reviewed in Herrmann, Stroud, and Losos
-(\protect\hyperlink{ref-Herrmannetal2021}{2021})). Removal is posited to
-create ecological opportunities---i.e., aspects of the niche become
-newly accessible, and the focal population shifts and/or expands its
-resource use to the new opportunity (Parent and Crespi
+change when important species interactions like resource competition or
+predation are relaxed or removed (reviewed in Herrmann, Stroud, and
+Losos (\protect\hyperlink{ref-Herrmannetal2021}{2021})). Removal is
+posited to create ecological opportunity---i.e., aspects of the niche
+become newly accessible, and the focal population shifts and/or expands
+its resource use new resources (Parent and Crespi
 (\protect\hyperlink{ref-ParentandCrespi2009}{2009}); Herrmann, Stroud,
 and Losos (\protect\hyperlink{ref-Herrmannetal2021}{2021})). Ecological
-release may then be followed by adaptive morphological evolution (Parent
-and Crespi (\protect\hyperlink{ref-ParentandCrespi2009}{2009});
-Herrmann, Stroud, and Losos
-(\protect\hyperlink{ref-Herrmannetal2021}{2021})). For example, a
-population undergoing ecological release can experience disruptive
-selection on males and females stemming from intraspecific competition
-over newly accessible resources (Bolnick and Doebeli
+release may then be followed by adaptive morphological evolution as
+traits change to reflect the new niche (Parent and Crespi
+(\protect\hyperlink{ref-ParentandCrespi2009}{2009}); Herrmann, Stroud,
+and Losos (\protect\hyperlink{ref-Herrmannetal2021}{2021})).
+
+For example, a population undergoing ecological release can experience
+disruptive selection on males and females stemming from intraspecific
+competition over newly accessible resources (Bolnick and Doebeli
 (\protect\hyperlink{ref-BolnickandDoebeli2003}{2003}); Bolnick and Lau
@@
 Paiva, and Rowe (\protect\hyperlink{ref-DeLisleetal2018}{2018})).
-However, release from predation should also result in ecological release
-followed by the evolution of increased sexual dimorphism because the
-absence of predators should generate ecological opportunity (Reimchen
-and Nosil (\protect\hyperlink{ref-ReimchenandNosil2004}{2004}); Parent
-and Crespi (\protect\hyperlink{ref-ParentandCrespi2009}{2009});
-Herrmann, Stroud, and Losos
-(\protect\hyperlink{ref-Herrmannetal2021}{2021})).
-
-Here, we tested the prediction that a release from predation should
-result in the evolution of increased sexual dimorphism using a
-well-preserved, finely-resolved sequence of a fossil threespine
-stickleback fish (Gasterosteus doryssus). The sequence in this
-depositional environment is comprised of two lineages (Bell
+However, release from predation might also drive the evolution of
+increased sexual dimorphism because an absence of predators should
+generate ecological opportunity (Reimchen and Nosil
+(\protect\hyperlink{ref-ReimchenandNosil2004}{2004}); Parent and Crespi
+(\protect\hyperlink{ref-ParentandCrespi2009}{2009}); Herrmann, Stroud,
+and Losos (\protect\hyperlink{ref-Herrmannetal2021}{2021})).
+
+Here, we tested the prediction that a release from predation results in
+the evolution of increased sexual dimorphism using a well-preserved,
+finely-resolved sequence of a fossil threespine stickleback fish
+(Gasterosteus doryssus). The sequence in this depositional environment
+is comprised of two lineages (Bell
 (\protect\hyperlink{ref-Bell2009}{2009}); Cerasoni, Bell, and Stuart
@@
 low-armor form with zero to one dorsal spines and highly reduced
-pelvises, on average. It lasted for at least 93,000 years before it was
-replaced suddenly by a second lineage (Bell, Baumgartner, and Oslon
-(\protect\hyperlink{ref-Belletal1985}{1985})). Lineage II appeared in
-the depositional environment fully armored, with complete pelvic
-girdles, two pelvic spines, and three dorsal spines, on average (Bell,
-Travis, and Blouw (\protect\hyperlink{ref-Belletal2006}{2006}); Stuart,
-Travis, and Bell (\protect\hyperlink{ref-Stuartetal2020}{2020})). It
-then immediately began evolving adaptive reduction in its armor traits
-(Hunt, Bell, and Travis (\protect\hyperlink{ref-Huntetal2008}{2008});
-Stuart, Travis, and Bell (\protect\hyperlink{ref-Stuartetal2020}{2020}))
-until it reached the same low-armor state previously held by lineage I.
+pelvises, on average. Lineage I lasted for at least 93,000 years before
+it was replaced by a second lineage, suddenly, likely on the order of
+years (Bell, Baumgartner, and Oslon
+(\protect\hyperlink{ref-Belletal1985}{1985}); Stuart et al.~unpublished
+data). Lineage II appeared in the depositional environment fully
+armored, with complete pelvic girdles, two pelvic spines, and three
+dorsal spines (Bell, Travis, and Blouw
+(\protect\hyperlink{ref-Belletal2006}{2006}); Stuart, Travis, and Bell
+(\protect\hyperlink{ref-Stuartetal2020}{2020})). It then immediately
+began evolving adaptive reduction in its armor traits (Hunt, Bell, and
+Travis (\protect\hyperlink{ref-Huntetal2008}{2008}); Stuart, Travis, and
+Bell (\protect\hyperlink{ref-Stuartetal2020}{2020})) until it reached
+the same low-armor state previously held by lineage I.
+
 The observation of 93,000 years of low-armor stasis in lineage I (Bell,
@@
 the observation of rapid evolution of armor loss by lineage II, both
-suggest that this depositional environment lacked piscivorous fish
-(e.g., trout and other salmonids known to prey on modern threespine
-stickleback). Armor maintenance in extant threespine stickleback
-(Gasterosteus aculeatus) correlates strongly the presence of vertebrate
-piscivores; populations with less predation pressure typically have less
-armor (Reimchen (\protect\hyperlink{ref-Reimchen1994}{1994}), Reimchen
-and Nosil (\protect\hyperlink{ref-ReimchenandNosil2004}{2004}); Bell et
-al. (\protect\hyperlink{ref-Belletal1993}{1993}); Roesti et al.
-(\protect\hyperlink{ref-Roestietal2023}{2023})). Consistently, only
-three fossil trout have been recorded from this paleolake basin, sampled
-from the same rock that has revealed \textgreater15,000 threespine
-stickleback fossils as well as occasional killifish (Fundulus
-nevadensis) (Bell (\protect\hyperlink{ref-Bell2009}{2009}); Cerasoni,
-Bell, and Stuart (\protect\hyperlink{ref-Cerasonietal2024}{2024})).
-Finally, tooth wear data from the lineage II sequence suggest that
-individuals in this lineage began eating more planktonic prey over time,
-expanding toward an open-water niche from the benthic, bottom-feeding
-niche they started with (Purnell et al.~2007). That this expansion into
-open water by lineage II coincided with armor reduction is further
-indicative of a limnetic system with fewer salmonid piscivores (Schluter
-and McPhail (\protect\hyperlink{ref-SchluterandMcPhail1992}{1992});
-Vamosi and Schluter
-(\protect\hyperlink{ref-VamosiandSchluter2004}{2004}); Roesti et al.
-(\protect\hyperlink{ref-Roestietal2023}{2023})). In sum, we infer that
-lineage II likely started in a nearby paleolake basin that had predators
-since it was armored when it arrived (Bell
+suggest that this depositional environment lacked piscivorous fish like
+trout and other salmonids known to prey on modern threespine
+stickleback. Armor presence in extant threespine stickleback
+(Gasterosteus aculeatus) correlates strongly with the presence of
+vertebrate piscivores; populations with less predation pressure
+typically have less armor (Reimchen
+(\protect\hyperlink{ref-Reimchen1994}{1994}), Reimchen and Nosil
+(\protect\hyperlink{ref-ReimchenandNosil2004}{2004}); Bell et al.
+(\protect\hyperlink{ref-Belletal1993}{1993}); Roesti et al.
+(\protect\hyperlink{ref-Roestietal2023}{2023})). In our paleolake basin,
+only three fossil trout have been found in the same section of rock that
+has revealed \textgreater20,000 threespine stickleback fossils as well
+as occasional killifish (Fundulus nevadensis) (Bell
 (\protect\hyperlink{ref-Bell2009}{2009}); Cerasoni, Bell, and Stuart
+(\protect\hyperlink{ref-Cerasonietal2024}{2024})). Thus, we reconstruct
+an evolutionar history in which it is likely that lineage II migrated
+from a nearby paleolake basin that had predators since it was armored
+when it arrived (Bell (\protect\hyperlink{ref-Bell2009}{2009});
+Cerasoni, Bell, and Stuart
 (\protect\hyperlink{ref-Cerasonietal2024}{2024})). Lineage II then
-experienced release from predators and dietary expansion in the focal
-paleolake basin, generating a scenario consistent with the starting
-conditions for models predicting the evolution of increased sexual
-dimorphism following ecological release.
+experienced release from predators in the focal paleolake basin,
+generating the initial conditions of evolutionary models predicting the
+evolution of increased sexual dimorphism following ecological release.
+We test this prediction here.
 
@@
 on morphology, treating fossil sex as a missing variable. With sex
-assigned to fossil specimens (and accounting for uncertainty in those
-assignments), we fit a modified Ornstein-Uhlenbeck (OU) (Uhlenbeck and
-Ornstein (\protect\hyperlink{ref-OUProcess}{1930})) model using a
-Bayesian framework to test for evolution of sexual dimorphism in each
-trait over \textasciitilde16,000 years of lineage II. We hypothesized
-that sexual dimorphism should increase following ecological release from
-predators.
+assigned to fossil specimens, we fit a modified Ornstein-Uhlenbeck (OU)
+(Uhlenbeck and Ornstein (\protect\hyperlink{ref-OUProcess}{1930})) model
+using a Bayesian framework that accounts for uncertainty in sex
+classification to test for evolution of sexual dimorphism in each trait
+over \textasciitilde16,000 years of lineage II.
 
@@
 (\protect\hyperlink{ref-Cerasonietal2024}{2024})), dug from an open pit
-diatomite mine at 9.526° N, 119.094° W, near Fernley, Nevada, USA.
-Series K consisted of 18 samples taken at \textasciitilde1000-year
-intervals, and mean sample times span \textasciitilde16,363 years. Fish
-from series K were measured for 16 ecomorphological traits related to
-armor, swimming, and feeding (Table 1). Series K started at a previously
-documented horizon when a low armored lineage of stickleback with zero
-to one dorsal spines, zero pelvic spines, and highly reduced pelvises
-(i.e., lineage I described above) was completely replaced by a high
-armored lineage of stickleback with three dorsal spines, two pelvic
-spines, and complete pelvises (i.e., lineage II described above) (Bell,
-Travis, and Blouw (\protect\hyperlink{ref-Belletal2006}{2006}); Bell
-(\protect\hyperlink{ref-Bell2009}{2009}); Stuart, Travis, and Bell
-(\protect\hyperlink{ref-Stuartetal2020}{2020})). Lineage II subsequently
-evolved reduction in armor, body size, and traits related to swimming
-and feeding (Bell, Travis, and Blouw
-(\protect\hyperlink{ref-Belletal2006}{2006}); Stuart, Travis, and Bell
-(\protect\hyperlink{ref-Stuartetal2020}{2020}); Siddiqui et al.
-(\protect\hyperlink{ref-Siddiquietal2024}{2024})). The tempo and mode of
-armor reduction during this sequence suggests adaptive evolution by
-natural selection (Hunt, Bell, and Travis
+diatomite mine at 9.526° N, 119.094° W, near Hazen, Nevada, USA. Series
+K consisted of 18 samples taken at \textasciitilde1000-year intervals,
+and mean sample times span \textasciitilde16,363 years. Fish from series
+K were measured for 16 ecomorphological traits related to armor,
+swimming, and feeding (Table 1). Series K started at the previously
+documented horizon when lineage I was replaced by lineage II. The tempo
+and mode of lineage II armor reduction during this sequence suggests
+adaptive evolution by natural selection (Hunt, Bell, and Travis
 (\protect\hyperlink{ref-Huntetal2008}{2008})), and we focus on the
@@
 McPhail (\protect\hyperlink{ref-SchluterandMcPhail1992}{1992})). The
-generalist populations used here were the remainder of collections made
-by YES in 2013 and were previously described in Stuart et al.
+generalist specimens used here were collected by YES in 2013 and were
+previously described in Stuart et al.
 (\protect\hyperlink{ref-Stuartetal2017}{2017}) (Table S1). These samples
 were fixed in formalin, then stained for bone with Alizarin Red in 2013.
-Benthic and limnetic specimens were kindly loaned by D. Schluter and his
-lab at University of British Columbia. They collected benthic and
-limnetic individuals from Enos Lake in 1988 and from Emily Lake, Little
-Quarry Lake, Paxton Lake, and Priest Lake in 2018 (Table S1). The Enos
-specimens had been fixed whole in formalin and stored in 40\%
-isopropanol. The specimens from the other lakes were initially preserved
-whole in 95\% ethanol in the field before being gradually transferred to
-water then formalin in the lab and ultimately stored in 40\%
-isopropanol. In 2019, we stained these specimens for bone using Alizarin
-Red. We next replicated fossil data collection (Table 1) on these extant
-specimens. Standard length as well as pelvic-spine length on each side
-were measured with calipers. We used a dissection microscope to count
-dorsal spines, pelvic spines, dorsal-fin rays, and anal-fin rays. Right
-and left-side pelvic girdle lengths and ectocoracoid lengths were
-measured from ventral photographs taken using a Canon EOS Rebel T7 with
-a Tamron 16-300 mm MACRO lens mounted on a leveled Kaiser RS1 copy
-stand. Specimens were held in place for ventral photographs using a
-small tabletop vise with an attached scale bar. Lateral X-rays were used
-to measure dorsal spine length, number of pterygiophores anterior to the
-pterygiophore holding the third spine, length of the pterygiophore just
-anterior to the third spine, cleithrum length, and pre-maxilla ascending
-branch length. We also counted vertebrae from the X-rays: abdominal
-vertebrae were counted anterior to the first vertebra with a haemal
-spine contacting an anal fin pterygiophore. Caudal vertebrae were
-posterior, including the first vertebra with the haemal spine contacting
-the anal fin pterygiophore (following Aguirre, Walker, and Gideon
+Benthic and limnetic specimens were kindly loaned by D. Schluter and S.
+Blain at the University of British Columbia. The Schluter lab collected
+benthic and limnetic individuals from Enos Lake in 1988 and from Emily
+Lake, Little Quarry Lake, Paxton Lake, and Priest Lake in 2018 (Table
+S1). The Enos specimens had been fixed whole in formalin and stored in
+40\% isopropanol. The specimens from the other lakes were initially
+preserved whole in 95\% ethanol in the field before being gradually
+transferred to water then formalin in the lab and ultimately stored in
+40\% isopropanol. In 2019, we stained these specimens for bone using
+Alizarin Red. We next replicated fossil data collection (Table 1) on
+these extant specimens. Standard length as well as pelvic-spine length
+on each side were measured with calipers. We used a dissection
+microscope to count dorsal spines, pelvic spines, dorsal-fin rays, and
+anal-fin rays. Right and left-side pelvic girdle lengths and
+ectocoracoid lengths were measured from ventral photographs taken using
+a Canon EOS Rebel T7 with a Tamron 16-300 mm MACRO lens mounted on a
+leveled Kaiser RS1 copy stand. Specimens were held in place for ventral
+photographs using a small tabletop vise with an attached scale bar.
+Lateral X-rays were used to measure dorsal spine length, number of
+pterygiophores anterior to the pterygiophore holding the third spine,
+length of the pterygiophore just anterior to the third spine, cleithrum
+length, and pre-maxilla ascending branch length. We also counted
+vertebrae from the X-rays: abdominal vertebrae were counted anterior to
+the first vertebra with a haemal spine contacting an anal fin
+pterygiophore. Caudal vertebrae were posterior, including the first
+vertebra with the haemal spine contacting the anal fin pterygiophore
+(following Aguirre, Walker, and Gideon
 (\protect\hyperlink{ref-Aguirreetal2014}{2014})). X-rays were taken with
@@
 corrected data to the uncorrected trait data frame and used all of our
-data to build the imputation model described next.
+data (raw and size-corrected) to build the imputation model described
+next.
 
@@
 
-We used multiple imputation (Little and Rubin
+Briefly, We used multiple imputation (Little and Rubin
 (\protect\hyperlink{ref-little2002statistical}{2002})) to impute the sex
-of the fossils many times. Imputations were performed using the MICE
-(Buuren and Groothuis-Oudshoorn (\protect\hyperlink{ref-MICE}{2011}))
-algorithm and implemented in R (Team
-(\protect\hyperlink{ref-R2022language}{2022})) with M = 100 imputations
-performed. After imputing the sex of the fossil data, each completed
-data set is then used to fit a modified Ornstein--Uhlenbeck (OU)
-(Uhlenbeck and Ornstein (\protect\hyperlink{ref-OUProcess}{1930})) model
-using a Bayesian framework to look for sexual dimorphism across a
-variety of stickleback phenotypes.
+of the fossils. Imputations were performed using the MICE (Buuren and
+Groothuis-Oudshoorn (\protect\hyperlink{ref-MICE}{2011})) algorithm and
+implemented in R (Team (\protect\hyperlink{ref-R2022language}{2022}))
+with M = 100 imputations. After imputing the sex, we fit a modified
+Orenstein-Uhlenbeck (OU; Uhlenbeck and Ornstein
+(\protect\hyperlink{ref-OUProcess}{1930})) model to each imputed data
+set using a Bayesian framework to \{THIS NEEDS MORE\}. We then
+quantified sexual dimorphism as the \{THIS NEEDS MORE; MAYBE WE CAN
+WRITE TOGETHER\}. We describe the details of the imputation model and OU
+modeling next.
 
@@
 
-We impute the missing sex for the fossil data by sampling from the
-posterior predictive distribution
+\textcolor{red}{AKHIL'S VALIDATION OF THE IMPUTATION MODEL GOES HERE}
+
+We impute missing sex for the fossil data by sampling from the posterior
+predictive distribution
 \(P(\boldsymbol{W}_{fossil}|\boldsymbol{W}_{extant},\boldsymbol{Y})\)
@@
 (\protect\hyperlink{ref-ZhouReiter2010}{2010}) recommend a larger number
-of imputed data sets if the data users intend on performing Bayesians
-analysis after imputation, which in this case, we do. Therefore, the
-imputation algorithm is run to obtain a total of \(M = 100\) completed
-datasets. In addition to this, Zhou and Reiter
-(\protect\hyperlink{ref-ZhouReiter2010}{2010}) suggests rather than
-using Rubin's combining rules to combine across imputed data sets,
-instead pool all of the draws from the posterior distributions across
-all of the imputed data sets to estimate the posterior distributions of
-parameters of interest. We proceed with our Bayesian analysis in this
-manner.
-
-\textcolor{red}{Akhil's stuff about validating the imputation model goes here.}
-
-\hypertarget{completed-data-model}{%
-\subsection{Completed Data Model}\label{completed-data-model}}
-
-\textcolor{red}{We should note somewhere that we are only using the $W_{fossil}$ in the modeling part.
-We drop the $W_{extant}$.
-So just note that $W_{ti}$ is really $W_{fossil,ti}$.
-Not sure how to say this, but we need to make it clear that we are only using the fossil data for the OU modeling.
-Greg, check and see if this makes sense.}
-
-\textcolor{red}{I think we need to move a lot of these details to the appendix.  }
-
+of imputed data sets if users intend on performing Bayesian analysis
+after imputation. Therefore, we imputed \(M = 100\) complet datasets.
+
+\textcolor{red}{presumably this also imputes missing data in the other traits? We should probably say that explicitly. -YS}
+
+We then pooled all of the draws from the posterior distribution across
+all the imputed data sets to estimate the posterior distributions of our
+parameters of interest (i.e., the trait values), rather than
+usingRubin's combining rules, following Zhou and Reiter
+(\protect\hyperlink{ref-ZhouReiter2010}{2010}). We used these posterior
+distributions in our Bayesian analysis.
+
+\hypertarget{bayesian-ou-modeling}{%
+\subsection{Bayesian OU modeling}\label{bayesian-ou-modeling}}
+
+For Bayesian modeling, we dropped the extant data and focused on
+modeling the evolution of sexual dimorphism through time in the fossils.
+
+\textcolor{red}{DOES THE NEXT SENTENCE CONTRADICT THE PREVIOUS ONE? BELOW, WE SAY THAT WE USED EACH IMPUTED DATA SET. ABOVE, WE SAY THAT WE POOLED ACROSS IMPUTED DATA SETS.-YS}
 For a given imputed dataset, let \(W_{ij}\) be the imputed sex and
@@
 our model; connecting to the previous section, \(W_{ij}\) and
-\(\boldsymbol{Y}_{ij}\) can be interpretted as \(W_{fossil,ij}\) and
+\(\boldsymbol{Y}_{ij}\) can be interpreted as \(W_{fossil,ij}\) and
 \(\boldsymbol{Y}_{fossil,ij}\). In addition, we denote \(Y_{K,ij}\), the
@@
 
-Because we are fitting a dataset with a stochastic structure on the
-means of the phenotypes, and we are interested in the distribution of
-the overall structure of these means, we analyze the data via a Bayesian
-analysis. Bayesian data analysis is also more naturally used when we
-have to impute data (\textbf{CITATION}). To aid in the sampling
-procedure for the discrete phenotypes, we declare
+Because we are fitting a dataset with a stochastic structure on
+phenotype means, and we are interested in the distribution of the
+overall structure of these means, we analyze the data via a Bayesian
+analysis \textless THIS LAST CLAUSE DOESN'T FOLLOW FROM THE BECAUSE
+CLAUSE. BETTER EXPLAIN WHY BAYES IS GOOD HERE\textgreater. Bayesian data
+analysis is also more naturally used when we have to impute data
+(\textbf{CITATION}). To aid in the sampling procedure for the discrete
+phenotypes, we declare
 
@@
 
-As previously described, four models for the mean process were
-considered for each trait. Within each trait, the model with the lowest
-DIC was chosen for that train. All subsequent results presented here use
-the model with the lowest DIC individually by trait. Table
-\ref{whichmodel} shows the model that was selected for each trait. About
-half of the traits had models with a trend, and most models did not
-include an OU component. The traits that did include an OU component
-were caudal vertebra (mcv), pelvic spine lentgh (lpt), abdominal
-vertebra (mav), and standard length (stl) with caudal vertebra (mcv)
-being the only model with both a trend and an OU component. This implies
-that not only does caudal vertebra size have a linear trend across sex,
-but the average size at one time period is highly correlated with the
-average size at the next time period.
+\textless WE NEED RESULTS SECTION DESCRIBING THE RESULTS OF AKHIL'S
+MULTIPLE IMPUTATION, BOTH THE TESTS OF HOW WELL THE MODEL WORKED
+CLASSIFYING MODEL SAMPLES, AND SUMMARY STATISTICS OF FOSSIL IMPUTATION
+RESULTS -YS\textgreater{}
+
+Within each trait, the model with the lowest DIC was chosen for that
+trait . All subsequent results presented here use the model with the
+lowest DIC individually by trait .
+
+Table \ref{whichmodel} reports the results of model fitting. X of the Y
+traits were best fit by a trend model. Best models included an OU
+component for caudal and abdominal vertebra number (mcv, mav), pelvic
+spine length (lpt ) and standard length (stl). Caudal vertebra number
+was the only trait best fit by a model including both a trend and an OU
+component. This implies that not only does caudal vertebra size have a
+linear trend across sex \textless WHAT DO YOU MEAN BY LINEAR TREND
+ACROSS SEX?\textgreater, but the average size at one time period is
+highly correlated with the average size at the next time period
+\textless I'VE ALWAYS FOUND THAT THIS AID TO MODEL INTEPRETATION IS
+BETTER IN THE METHODS, JUST AFTER THE MODELS ARE DESCRIBED\textgreater.
 
@@
 dimorphism.
+IS THIS CORRECT?
+- YS
 } \textcolor{red} {
@@
 and littoral (Wootton 1976)\ldots. Reimchen papers in general good for
-this section.
+this section. Moreover, tooth wear data from the lineage II sequence
+suggest that individuals in this lineage began eating more planktonic
+prey over time, expanding toward an open-water niche from the benthic,
+bottom-feeding niche they started with (Purnell et al.~2007). That this
+expansion into open water by lineage II coincided with armor reduction
+is further indicative of a limnetic system with fewer salmonid
+piscivores (Schluter and McPhail
+(\protect\hyperlink{ref-SchluterandMcPhail1992}{1992}); Vamosi and
+Schluter (\protect\hyperlink{ref-VamosiandSchluter2004}{2004}); Roesti
+et al. (\protect\hyperlink{ref-Roestietal2023}{2023})).
 

### 2024-09-26 manuscript/paper.Rmd
joint writing updates. Waiting for some methods

@@
 
-\textcolor{red}{AKHIL'S VALIDATION OF THE IMPUTATION MODEL GOES HERE}
+\textcolor{red}{AKHIL'S METHODS FOR VALIDATION OF THE IMPUTATION MODEL GOES HERE}
 
@@
 However, @ZhouReiter2010 recommend a larger number of imputed data sets if users intend on performing Bayesian analysis after imputation.
-Therefore, we imputed $M = 100$ complet datasets.
+Therefore, we imputed $M = 100$ complete datasets.
 
@@
 
-We then pooled all of the draws from the posterior distribution across all the imputed data sets to estimate the posterior distributions of our parameters of interest (i.e., the trait values), rather than usingRubin's combining rules, following @ZhouReiter2010.
+We then pooled all of the draws from the posterior distribution across all the imputed data sets to estimate the posterior distributions of our parameters of interest (i.e., the trait values), rather than using Rubin's combining rules, following @ZhouReiter2010.
 We used these posterior distributions in our Bayesian analysis.
+\textcolor{red}{This was done, was it used?
+If Matt ran his models on each imputed data set?.-YS}
 
@@
 ## Bayesian OU modeling
+latent variable is the u in equation 8.
+likelihood doesn't work as well because it doesn't allow for incorporation of prior information and uncertainty when estimating a latent variable.
+Knowing u is helpful because it's the error in equation 6.
+Can estimate the means in eqn 6 that the crux of the research.
+better able to approximate the latent variables within the model.
 For Bayesian modeling, we dropped the extant data and focused on modeling the evolution of sexual dimorphism through time in the fossils.
@@
 If $Y_{k,ij}$ is a discrete trait, the conventional method of modelling this data is by fitting a Poisson distribution.
-However, the Poisson distribution assumes that the mean and variance of $Y_{k,ij}$ are equal ($E(Y_{k,ij}) = \text{Var}(Y_{k,ij})$), while the empirical fossil data may have variances that are smaller than their respective means.
+However, the Poisson distribution assumes that the mean and variance of $Y_{k,ij}$ are equal ($E(Y_{k,ij}) = \text{Var}(Y_{k,ij})$), while the empirical fossil data may have variances that are not equal to their respective means.
 To combat this issue, we propose to fit the discrete traits to a generalized Poisson model as defined in @GeneralizedPoisson.
@@
 \end{align}
-where $E(X) = \lambda$ and $Var(X) = \frac{\lambda}{(1 - \alpha)^2}$.
-If $\alpha > 0$, then the variance is greater than the mean, called overdispersion; if $\alpha < 0$, then the variance is greater than the mean, called underdispersion; if $\alpha = 0$, then the model degenerates to a Poisson distribution.
+where GP denotes a generalized Poisson distribution with $E(X) = \lambda$ and $Var(X) = \frac{\lambda}{(1 - \alpha)^2}$.
 
@@
 where $\kappa_k$ is a parameter associated with the correlation between $u_{k,gt}$ and $u_{k,g(t+dt)}$, $\tau_k$ is the standard deviation of the OU process, and $W_t$ is a standard Brownian motion.
-As shown in @OUProcess, the closed form solution for the SDE in (\ref{eq:OU_cont}) is
+As shown in @OUProcess, the closed form solution for the stochastic differential equation in (\ref{eq:OU_cont}) is
 \begin{align}
@@
 
-In our empirical study, we analyze the following four nested models for the mean process outlined in (\ref{eq:mu}):
-\begin{itemize}
-\item OU with Trend: The mean process for each phenotype $k=1,\cdots,K$ and $g \in \{f,m\}$ as previously defined.
-\item OU with No Trend: No linear trend on the mean processes and the overall mean of the phenotype is constant for each sex.
-This is achieved by setting $\beta_{1,kg} = 0$ in (\ref{eq:mu}) for $k = 1,\cdots,K$ and $g \in \{f,m\}$
-\item No OU with Trend: No random fluctuations on the mean processes, i.e.
+<!-- In our empirical study, we analyze the following four nested models for the mean process outlined in (\ref{eq:mu}): -->
+<!-- \begin{itemize} -->
+<!-- \item OU with Trend: The mean process for each phenotype $k=1,\cdots,K$ and $g \in \{f,m\}$ as previously defined. -->
+<!-- \item OU with No Trend: No linear trend on the mean processes and the overall mean of the phenotype is constant for each sex.
+This is achieved by setting $\beta_{1,kg} = 0$ in (\ref{eq:mu}) for $k = 1,\cdots,K$ and $g \in \{f,m\}$ -->
+<!-- \item No OU with Trend: No random fluctuations on the mean processes, i.e.
 the OU process assumption is not needed.
 This is achieved by setting $u_{k,gt_i} = 0$ in (\ref{eq:mu}) for $k = 1,\cdots,K$, $g \in \{f,m\}$, and $i = 1,\cdots,T$.
-\item No OU with No Trend: The mean for a specific trait and specific sex at each time point is a constant value.
+-->
+<!-- \item No OU with No Trend: The mean for a specific trait and specific sex at each time point is a constant value.
 This is achieved by setting $\beta_{1,kg} = 0$ and $u_{k,gt_i} = 0$ in (\ref{eq:mu}) for $k = 1,\cdots,K$, $g \in \{f,m\}$, and $i = 1,\cdots,T$.
-\end{itemize}
+-->
+<!-- \end{itemize} -->
 
-Because we are fitting a dataset with a stochastic structure on phenotype means, and we are interested in the distribution of the overall structure of these means, we analyze the data via a Bayesian analysis <THIS LAST CLAUSE DOESN'T FOLLOW FROM THE BECAUSE CLAUSE.
-BETTER EXPLAIN WHY BAYES IS GOOD HERE>.
-Bayesian data analysis is also more naturally used when we have to impute data (\textbf{CITATION}).
+Because we are fitting a dataset with a stochastic structure with latent variables to phenotype means (i.e., eqn.
+7 is a stochastic structure), we analyze the data via a Bayesian analysis <THIS LAST CLAUSE DOESN'T FOLLOW FROM THE BECAUSE CLAUSE.
+Add the language about latent variables here>.
+Bayesian data analysis is also more naturally used when we have to impute data because imputation is sampling from a posterior predictive distribution of the missing data given the observed data (\textbf{CITATION}).
 To aid in the sampling procedure for the discrete phenotypes, we declare
 
-Priors: To aid in the sampling procedure for the discrete phenotypes, we declare $\phi_k = \log\left(\frac{\alpha_k - \max_{i,j}(-\lambda_{k,ij}/y_{k,ij})}{1 - \alpha_k}\right)$ where $\lambda_{k,ij} = \left\{\begin{array}{cc} \exp(\mu_{k,ft_i} + \gamma_k Y_{K,ij}) & W_{ij} = \text{Female} \\ \exp(\mu_{k,mt_i} + \gamma_k Y_{K,ij}) & W_{ij} = \text{Male} \end{array}\right.$.
-This transformation is performed to ensure our algorithm can properly sample from the posterior distribution of interest.
+Priors: To aid in the sampling procedure for the discrete phenotypes, we first have to transform the parameters of the generalized Poisson distribution onto the real number line in order to ensure our algorithm can properly sample from the posterior distribution of interest using STAN software.
+This was done by declaring $\phi_k = \log\left(\frac{\alpha_k - \max_{i,j}(-\lambda_{k,ij}/y_{k,ij})}{1 - \alpha_k}\right)$ where $\lambda_{k,ij} = \left\{\begin{array}{cc} \exp(\mu_{k,ft_i} + \gamma_k Y_{K,ij}) & W_{ij} = \text{Female} \\ \exp(\mu_{k,mt_i} + \gamma_k Y_{K,ij}) & W_{ij} = \text{Male} \end{array}\right.$.
 
@@
 \end{align}
+These priors were chosen to be weakly informative such that the posterior distributions were heavily influenced by the data.
+
 We also note that, for the discrete phenotypes in equations (\ref{eq:disc_corr}) and (\ref{eq:disc_ind}), there is no $\sigma_k$, and is not sampled, and for the continuous phenotypes in equation (\ref{eq:cont}), there is no $\phi_k$ and is not sampled.
@@
 
-@Cornault2022 Bayesian OU model.
+After running these models, we have posterior distributions of all the model parameters of interest.
+From these, we can obtain samples from the posterior distribution of mean trait values for each trait for males and females, using equation 6.
+We use these values for the following analyses.
+
+# Section whatever it happens to be
+Description of methods used to calculate sexual dimorphism (um-uf)
+
+# Section whatever ithappens to be plus 1
+Description of Tau analyses, to ask if SD calculated in the section above changes with time.
 
@@
 
-Within each trait, the model with the lowest DIC was chosen for that trait <THIS IS METHODS--WHERE IS THIS DESCRIBED IN THE METHODS>.
+<!-- Within each trait, the model with the lowest DIC was chosen for that trait <THIS IS METHODS--WHERE IS THIS DESCRIBED IN THE METHODS>.
 All subsequent results presented here use the model with the lowest DIC individually by trait <THIS IS METHODS--WHERE IS THIS DESCRIBED IN THE METHODS>.
+-->
 
-Table \ref{whichmodel} reports the results of model fitting.
+<!-- Table \ref{whichmodel} reports the results of model fitting.
 X of the Y traits were best fit by a trend model.
@@
 This implies that not only does caudal vertebra size have a linear trend across sex <WHAT DO YOU MEAN BY LINEAR TREND ACROSS SEX?>, but the average size at one time period is highly correlated with the average size at the next time period <I'VE ALWAYS FOUND THAT THIS AID TO MODEL INTEPRETATION IS BETTER IN THE METHODS, JUST AFTER THE MODELS ARE DESCRIBED>.
+-->
 
-\textcolor{red} {
-Yoel's interpretation of the above paragraph.
+<!-- \textcolor{red} { -->
+<!-- Yoel's interpretation of the above paragraph.
 Taking the model with the lowest DIC score as the best model for each trait, we found that cle, ds1, lps, and pmx did not fit either the trend model or the OU model for male-female trait mean differences.
@@
 IS THIS CORRECT?
-- YS
-}
-\textcolor{red} {
-We found evidence for an OU process but no trend for lpt, mav, and stl, suggesting that these traits varied through time, about a fixed value for dimorphism.
+- YS -->
+<!-- } -->
+<!-- \textcolor{red} { -->
+<!-- We found evidence for an OU process but no trend for lpt, mav, and stl, suggesting that these traits varied through time, about a fixed value for dimorphism.
 Finally, we found evidence that change in mcv fitted both a trend an an OU process, meaning ???
 WHAT BIOLOGICALLY ???.
-}
+-->
+<!-- } -->
 
-\textcolor{red} {HOW DO WE RECONCILE THE PRESENCE OF TRENDS IN THESE MODELS WITH AN ABSENCE OF VISIBLE TRENDS IN FIGURES 2 THROUGH 4?
+<!-- \textcolor{red} {HOW DO WE RECONCILE THE PRESENCE OF TRENDS IN THESE MODELS WITH AN ABSENCE OF VISIBLE TRENDS IN FIGURES 2 THROUGH 4?
 I WONDER IF THIS RESULT PARAGRAPH BELONGS BEST AT THE END OF THE RESULTS, RATHER THAN THE BEGINNING.
 CONSIDER THE STATISTICAL EVIDENCE FOR TRENDS AND OU AFTER WE'VE WALKED THE READERS THROUGH VISUAL EVIDENCE FOR CHANGE.
-}
+-->
+<!-- } -->
 
-## Best fitting models 
-\begin{table}[ht]
-\centering
-\begin{tabular}{|c||c|c|}
-  \hline
- & OU & No OU  \\ 
-  \hline
-  \hline
-Trend  &  mcv  & ds2, ds3, ect, maf, mdf \\ 
-  \hline
-  No Trend & lpt, mav, stl  &  cle, ds1, lps, pmx\\
-   \hline
-\end{tabular}
-\label{whichmodel}
-\end{table}
+<!-- ## Best fitting models  -->
+<!-- \begin{table}[ht] -->
+<!-- \centering -->
+<!-- \begin{tabular}{|c||c|c|} -->
+<!--   \hline -->
+<!--  & OU & No OU  \\  -->
+<!--   \hline -->
+<!--   \hline -->
+<!-- Trend  &  mcv  & ds2, ds3, ect, maf, mdf \\  -->
+<!--   \hline -->
+<!--   No Trend & lpt, mav, stl  &  cle, ds1, lps, pmx\\ -->
+<!--    \hline -->
+<!-- \end{tabular} -->
+<!-- \label{whichmodel} -->
+<!-- \end{table} -->
 

### 2024-09-26 manuscript/paper.Rmd
made edits per Matt and my joint-writing on 26 September. Made clearer the to-do items and questions for Greg and Matt to fill in.

@@
 
-The lineage II fossil data consist of `r sum(is.na(stick$gender))` specimens of unknown sex sampled across 18 time periods spaced  ~1000 years apart.
+The lineage II fossil data consist of `r sum(is.na(stick$gender))` specimens of unknown sex sampled across the 18 K series samples.
 Figure 1 shows the sample size for each of the 18 samples.
@@
 
+THIS FIGURE DOESN'T SHOW MUCH. LET'S CUT FOR SPACE. 
 ```{r fig.cap = "Fossil sample size for each time period"}
@@
 Briefly, We used multiple imputation (@little2002statistical) to impute the sex of the fossils.
-Imputations were performed using the MICE (@MICE) algorithm and implemented in R (@R2022language) with M = 100 imputations.
-After imputing the sex, we fit a modified Orenstein-Uhlenbeck (OU; @OUProcess) model to each imputed data set using a Bayesian framework to {THIS NEEDS MORE}.
-We then quantified sexual dimorphism as the {THIS NEEDS MORE; MAYBE WE CAN WRITE TOGETHER}.
-We describe the details of the imputation model and OU modeling next.
-
-# Models  {#sec:models}
-## Imputation Model 
 Let $\boldsymbol{W}$ be an $(n_{extant} + n_{fossil}) \times 1$ vector of the covariate sex of the stickleback fish and $\boldsymbol{Y}$ be an $(n_{extant} + n_{fossil}) \times K$ matrix of the $K$ phenotypes of interest.
 Because the sex of the fossilized stickleback fish is unobservable, we further define $\boldsymbol{W} = (\boldsymbol{W}_{extant}^T,\boldsymbol{W}_{fossil}^T)^T$ where $\boldsymbol{W}_{extant}$ and $\boldsymbol{W}_{fossil}$ are the $n_{extant} \times 1$ and $n_{fossil} \times 1$ vectors of the observed extant sex and missing fossil sex, respectively.
-
-\textcolor{red}{AKHIL'S METHODS FOR VALIDATION OF THE IMPUTATION MODEL GOES HERE}
-
-We impute missing sex for the fossil data by sampling from the posterior predictive distribution $P(\boldsymbol{W}_{fossil}|\boldsymbol{W}_{extant},\boldsymbol{Y})$ using the multiple imputation by chained equations (MICE) algorithm (@MICE) with predictive mean matching.
+We imputeD missing sex for the fossil data by sampling from the posterior predictive distribution $P(\boldsymbol{W}_{fossil}|\boldsymbol{W}_{extant},\boldsymbol{Y})$ using the multiple imputation by chained equations (MICE) algorithm (@MICE) with predictive mean matching, implemented in R (CITATION).
 Traditionally, the choice for the number of completed dats sets is a relatively small number such as $M = 5$ or $M = 10$.
@@
 \textcolor{red}{presumably this also imputes missing data in the other traits? We should probably say that explicitly. -YS}
-
+\textcolor{red}{AKHIL'S METHODS FOR VALIDATION OF THE IMPUTATION MODEL GOES HERE}
+\textcolor{red}{THE POOLING DESCRIBED IN THE NEXT SENTENCE. WAS IT DONE, AND IF SO, HOW WAS IT USED IN DOWNSTREAM ANALYSES. MATT SAYS THAT HE RAN IS MODELS ON EACH IMPUTED DATA SET. -YS}
 We then pooled all of the draws from the posterior distribution across all the imputed data sets to estimate the posterior distributions of our parameters of interest (i.e., the trait values), rather than using Rubin's combining rules, following @ZhouReiter2010.
 We used these posterior distributions in our Bayesian analysis.
-\textcolor{red}{This was done, was it used?
-If Matt ran his models on each imputed data set?.-YS}
 
-<!-- Bayesian Analysis after multiple imputation @ZhouReiter2010: They recommend using a large number of imputations.  5 or 10 is too small.  We are using M = 100. -->
-
-## Bayesian OU modeling
-latent variable is the u in equation 8.
-likelihood doesn't work as well because it doesn't allow for incorporation of prior information and uncertainty when estimating a latent variable.
-Knowing u is helpful because it's the error in equation 6.
-Can estimate the means in eqn 6 that the crux of the research.
-better able to approximate the latent variables within the model.
-For Bayesian modeling, we dropped the extant data and focused on modeling the evolution of sexual dimorphism through time in the fossils.
+## Modeling trait means through time for males and females
+The goal of our analysis was to estimate fossil trait means through time, by sex, while incorporating uncertainty in the imputed data, especially uncertainty sex classification.
+We took a Bayesian approach because imputation is the process of sampling from a posterior predictive distribution of missing data, given observed data.
+\textcolor{red}{THIS NEEDS MORE WORK TO JUSTIFY THE BAYESIAN APPROACH.
+-YS}
 
 \textcolor{red}{DOES THE NEXT SENTENCE CONTRADICT THE PREVIOUS ONE? BELOW, WE SAY THAT WE USED EACH IMPUTED DATA SET. ABOVE, WE SAY THAT WE POOLED ACROSS IMPUTED DATA SETS.-YS}
-For a given imputed dataset, let $W_{ij}$ be the imputed sex and $\boldsymbol{Y}_{ij}$ be the $K \times 1$ vector of phenotypes for stickleback fossil $j$ at time $t_i$ where $i = 1, \ldots, T$ and $j = 1,\ldots,n_{t}$.
-Note, we are inputting only the fossil data into our model; connecting to the previous section, $W_{ij}$ and $\boldsymbol{Y}_{ij}$ can be interpreted as $W_{fossil,ij}$ and $\boldsymbol{Y}_{fossil,ij}$.
-In addition, we denote $Y_{K,ij}$, the last variable in $\boldsymbol{Y}_{ij}$, to be the standard length of the fish,
+To estimate fossil sample means through time, we first dropped the extant individuals from our imputed data sets.
+Then, for a given imputed dataset, we let $W_{ij}$ be the imputed sex and $\boldsymbol{Y}_{ij}$ be the $K \times 1$ vector of phenotypes for stickleback fossil $j$ at time $t_i$ where $i = 1, \ldots, T$ and $j = 1,\ldots,n_{t}$.
+In addition, we denote $Y_{K,ij}$, the last variable in $\boldsymbol{Y}_{ij}$, to be the standard length of each fish individual, a measure of body size.
 \begin{align}
@@
 
-It is reasonable to assume that the other continuous traits of stickleback fish will have some correlation with its standard length \textbf{(CITATION)}.
-We account for this by adding an additional parameter, $\gamma_k$, into our model.
-More specifically, if $Y_{k,ij}$ is a continuous trait, then
+It is reasonable to assume that the other continuous traits will have some correlation with standard length \textbf{(CITATION)}.
+We account for this by adding an additional parameter, $\gamma_k$, If $Y_{k,ij}$ is a continuous trait, then
 \begin{align}
@@
 
-If $Y_{k,ij}$ is a discrete trait, the conventional method of modelling this data is by fitting a Poisson distribution.
-However, the Poisson distribution assumes that the mean and variance of $Y_{k,ij}$ are equal ($E(Y_{k,ij}) = \text{Var}(Y_{k,ij})$), while the empirical fossil data may have variances that are not equal to their respective means.
-To combat this issue, we propose to fit the discrete traits to a generalized Poisson model as defined in @GeneralizedPoisson.
+If $Y_{k,ij}$ is a discrete trait, the conventional modelling method is to fit a Poisson distribution.
+However, because the empirical fossil data violate the assumption that the Poisson variance is equal to its mean, we fit discrete traits to a generalized Poisson model as defined in @GeneralizedPoisson.
 Specifically, if $X \sim GP(\lambda,\alpha)$, then
@@
 
-We will use the generalized Poisson to model all of the discrete traits in our dataset.
-In addition, we assume that the discrete traits abdominal vertebrae (mav) and caudal vertebrae (mcv) also have correlation with the standard length of the fish (\textbf{CITATION}).
+In addition, we allowed that two discrete traits, abdominal vertebrae number (mav) and caudal vertebrae number (mcv) could also with  standard length (\textbf{CITATION}).
+This is because these are vertebral counts that run in parallel with the standard length measurement, along the anterior-posterior access of the fish; these vertebrae thus contribute to the length of the fish directly, and may evolve in a correlated fashion.
 If $Y_{k,ij}$ is one of the above traits, then
@@
 
-For the other discrete traits, we also assume the above model except we set $\gamma_k = 0$ because of the assumption of no correlation between the standard length and these traits.
-More specifically, we assume
+For the other discrete traits, we set $\gamma_k = 0$ because we expect that trait counts are set during early ontogeny, and should not change with size in adult fish. Thus,
 \begin{align}
@@
 \end{align}
+We point out that, for the discrete traits, the means are represented by $\exp\{\mu_{k,ft_i}\}$ and $\exp\{\mu_{k,mt_i}\}$ for ease of modeling.
 
-In the above model descriptions, $\mu_{k,ft_i}$ and $\mu_{k,mt_i}$ model the time-$t_i$ specific mean of phenotype $k$ for female and male stickleback fish, respectively.
-We point out that, for the discrete traits, the means are represented by $\exp\{\mu_{k,ft_i}\}$ and $\exp\{\mu_{k,mt_i}\}$ for ease of use in our modeling technique.
-We further set
+In the above model descriptions, $\mu_{k,ft_i}$ and $\mu_{k,mt_i}$ model the time-$t_i$, sample-specific trait mean $k$ for females and males, respectively.
+To account for the possibility that there is a time -dependent trend in $\mu_{k,ft_i}$ and $\mu_{k,mt_i}$, we further set
 \begin{align}
@@
 \end{align}
-for $g \in \{f,m\}$ where $\beta_{0,kg}$ and $\beta_{1,kg}$ are regression parameters of phenotype $Y_k$ for each sex, accounting for the possibility of a time-dependent trend in the mean structure, and $u_{k,gt_i}$ is the corresponding residual.
+for $g \in \{f,m\}$, where $\beta_{0,kg}$ and $\beta_{1,kg}$ are regression parameters of phenotype $Y_k$ for each sex, and $u_{k,gt_i}$ is the corresponding residual.
 To account for potential correlations between the residuals for a given trait $k$ and sex $g$, we fit an Ornstein-Uhlenback (OU) process (@OUProcess).
-More specifically, define $du_{k,gt} = u_{k,g(t + dt)} - u_{k,gt}$, the change in $u_k,gt$ for a given trait $k$ and sex $g$ over a miniscule time period $dt$.
+More specifically, we defined $du_{k,gt} = u_{k,g(t + dt)} - u_{k,gt}$, the change in $u_k,gt$ for a given trait $k$ and sex $g$ over a miniscule time period $dt$.
 The OU process is defined as
@@
 \end{align}
-for $i = 2,\cdots,T$. In a traditional OU process, the initial value $u_{k,gt_1}$ is assumed to be a (potentially unknown) constant, and we discuss the procedure for estimating this value below.
+for $i = 2,\cdots,T$.
+In a traditional OU process, the initial value $u_{k,gt_1}$ is assumed to be a (potentially unknown) constant.
+This unknown constant can therefore be considered a latent variable, further justifying our Bayesian approach, which better incorporates prior information and uncertainty from missing data imputation in the estimate of the variable.
+
+We discuss the procedure for estimating this value below. \textcolor{red}{DID THIS GET DISCUSSED?-YS}
 
@@
 
-Because we are fitting a dataset with a stochastic structure with latent variables to phenotype means (i.e., eqn.
-7 is a stochastic structure), we analyze the data via a Bayesian analysis <THIS LAST CLAUSE DOESN'T FOLLOW FROM THE BECAUSE CLAUSE.
-Add the language about latent variables here>.
-Bayesian data analysis is also more naturally used when we have to impute data because imputation is sampling from a posterior predictive distribution of the missing data given the observed data (\textbf{CITATION}).
-To aid in the sampling procedure for the discrete phenotypes, we declare
-
-Priors: To aid in the sampling procedure for the discrete phenotypes, we first have to transform the parameters of the generalized Poisson distribution onto the real number line in order to ensure our algorithm can properly sample from the posterior distribution of interest using STAN software.
+Priors: To aid in the sampling procedure for the discrete phenotypes, we first had to transform the parameters of the generalized Poisson distribution onto the real number line in order to ensure our algorithm can properly sample from the posterior distribution of interest using STAN software.
 This was done by declaring $\phi_k = \log\left(\frac{\alpha_k - \max_{i,j}(-\lambda_{k,ij}/y_{k,ij})}{1 - \alpha_k}\right)$ where $\lambda_{k,ij} = \left\{\begin{array}{cc} \exp(\mu_{k,ft_i} + \gamma_k Y_{K,ij}) & W_{ij} = \text{Female} \\ \exp(\mu_{k,mt_i} + \gamma_k Y_{K,ij}) & W_{ij} = \text{Male} \end{array}\right.$.
 
-For $k = 1,\cdots,K$,
+We chose the following priors, designing them to be weakly informative such that the posterior distributions could be influenced by the data, if warranted. For $k = 1,\cdots,K$,
 
@@
 \end{align}
-These priors were chosen to be weakly informative such that the posterior distributions were heavily influenced by the data.
 
-We also note that, for the discrete phenotypes in equations (\ref{eq:disc_corr}) and (\ref{eq:disc_ind}), there is no $\sigma_k$, and is not sampled, and for the continuous phenotypes in equation (\ref{eq:cont}), there is no $\phi_k$ and is not sampled.
+We also note that, for the discrete phenotypes in equations (\ref{eq:disc_corr}) and (\ref{eq:disc_ind}), there is no $\sigma_k$, and it is not sampled, and for the continuous phenotypes in equation (\ref{eq:cont}), there is no $\phi_k$ and it is not sampled.
 
@@
 
-# Section whatever it happens to be
-Description of methods used to calculate sexual dimorphism (um-uf)
+## Quantifying sexual dimorphism, and testing for increased sexual dimorphism through time.
+We took the mean trait values for each sex at each temporal sample and quantified sexual dimorphism by \textcolor{red}{FILL IN THE METHODS HERE}
 
-# Section whatever ithappens to be plus 1
-Description of Tau analyses, to ask if SD calculated in the section above changes with time.
+Recall that we predicted that sexual dimorphism should increase through time in Lineage II, following ecological release from predators.
+To test this prediction, we \textcolor{red}{FILL IN THE METHODS REGARDING TAU ANALYSES HERE}
 
 # Results {#sec:results}
-<WE NEED RESULTS SECTION DESCRIBING THE RESULTS OF AKHIL'S MULTIPLE IMPUTATION, BOTH THE TESTS OF HOW WELL THE MODEL WORKED CLASSIFYING MODEL SAMPLES, AND SUMMARY STATISTICS OF FOSSIL IMPUTATION RESULTS -YS>
+\textcolor{red}{WE NEED RESULTS SECTION DESCRIBING THE RESULTS OF AKHIL'S MULTIPLE IMPUTATION, BOTH THE TESTS OF HOW WELL THE MODEL WORKED CLASSIFYING MODEL SAMPLES, AND SUMMARY STATISTICS OF FOSSIL IMPUTATION RESULTS -YS}
 

### 2024-09-30 manuscript/paper.Rmd
Update manuscript with posterior and Kendall's Tau

@@
 \sigma_k & \iid \mathcal{N}(0,10)I_{\{\sigma > 0\}} \nonumber \\
-\sigma_k & \iid \mathcal{N}(0,10) \nonumber \\
 \tau_k & \iid \mathcal{N}(0,10)I_{\{\tau > 0\}} \nonumber \\
@@
 
+We follow the method outlined in @ZhouandReiter to obtain $R$ draws from the posterior distribution of the outlined model parameters given our observed data and imputed values.
+More specifically, for the $j^{th}$ imputed data set where $j = 1,\cdots,M$, we obtain $\frac{R}{M}$ draws from the posterior distribution using an MCMC algorithm with a sufficient burnin period, and then we combine these $M$ MCMC draws to say that we have a total of $R$ draws from the overall posterior distribution of interest, which can then be used to assess the presence of sexual dimorphism for a given trait.
+
 All models were built using @R2022language and (\textbf STAN)
@@
 ## Quantifying sexual dimorphism, and testing for increased sexual dimorphism through time.
-We took the mean trait values for each sex at each temporal sample and quantified sexual dimorphism by \textcolor{red}{FILL IN THE METHODS HERE}
+We identify sexual dimorphism first by calculating the difference in mean trait size for males and females at each time for phenotype $k$ ($\mu_{k,m,t_i} - \mu_{k,m,t_i}$) for $i = 1,\cdots,T$ where, for the continuous phenotypes, the mean trait sizes for each sex $\mu_{k,m,t_i}$ and $\mu_{k,f,t_i}$ are calculated using (\ref{eq:mu}).
+For the discrete phenotypes, we note that in (\ref{eq:disc_corr}) and (\ref{eq:disc_ind}) the mean trait size is $\lambda_{k,m,t_i} = \exp(\mu_{k,m,t_i})$ and $\lambda_{k,f,t_i} = \exp(\mu_{k,f,t_i})$ for males and females, respectively, which is the value we use in our analysis.
+For simplicity, we will use $\mu_{k,m,t_i}$ and $\mu_{k,f,t_i}$ as the notation for mean size for both discrete and continuous traits for the rest of this discussion.
+From our MCMC algorithm, we obtain $M$ samples from the posterior distribution of $\mu_{k,m,t_i} - \mu_{k,f,t_i}$ to use in our analysis.
 
 Recall that we predicted that sexual dimorphism should increase through time in Lineage II, following ecological release from predators.
-To test this prediction, we \textcolor{red}{FILL IN THE METHODS REGARDING TAU ANALYSES HERE}
+However, this change in dimorphism make be increasing at a non-constant rate, as pointed out by the inclusion of the OU process.
+We assess for dimorphism by calculating a Kendall's $\tau$ coefficient (@KendallsTau).
+Kendall's $\tau$ assesses the ordinal association between a set of bivariate observations, where $\tau > 0$ indicates concordance between the observations (i.e., as $x$ increases, $y$ also tends to increase.)
+
+For our analysis, we calculate $\boldsymbol{\tau}_k = \{\tau_{k,r}: r = 1,\cdots,R\}$ for trait $k$ where $\tau_r$ is the estimate of the Kendall's $\tau$ correlation for $\mu_{k,m,t_i} - \mu_{k,f,t_i}$ versus time $t_i$ for the $r^{th}$ MCMC sample from the posterior distribution.
+The posterior probability of sexual dimorphism for trait $k$ can be estimated as the proportion of values in $\boldsymbol{\tau}_k$ that are greater than 0 ($\frac{1}{R} \sum_{r=1}^R I(\tau_{k,r} > 0)$).
+We can also assess the strength of the dimorphism by calculating the proportion of posterior samples that are greater than a prespecified threshold \textcolor{red}{(0.5??, idk what this could be.) -MS}
 

### 2024-09-30 manuscript/paper.tex
Update manuscript with posterior and Kendall's Tau

@@
 \usepackage{amsmath,amssymb}
-\usepackage{lmodern}
 \usepackage{iftex}
@@
 \else % if luatex or xetex
-  \usepackage{unicode-math}
+  \usepackage{unicode-math} % this also loads fontspec
   \defaultfontfeatures{Scale=MatchLowercase}
   \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
+\fi
+\usepackage{lmodern}
+\ifPDFTeX\else
+  % xetex/luatex font selection
 \fi
@@
 \IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\urlstyle{same} % disable monospaced font for URLs
+\urlstyle{same}
 \hypersetup{
@@
 The lineage II fossil data consist of 814 specimens of unknown sex
-sampled across 18 time periods spaced \textasciitilde1000 years apart.
-Figure 1 shows the sample size for each of the 18 samples. There are at
-least 22 specimens in each sample with a high of 67 specimens in sample
-7.
+sampled across the 18 K series samples. Figure 1 shows the sample size
+for each of the 18 samples. There are at least 22 specimens in each
+sample with a high of 67 specimens in sample 7.
+
+THIS FIGURE DOESN'T SHOW MUCH. LET'S CUT FOR SPACE.
 
@@
 noted.}\tabularnewline
-\toprule()
+\toprule\noalign{}
 \begin{minipage}[b]{\linewidth}\raggedright
@@
 \end{minipage} \\
-\midrule()
+\midrule\noalign{}
 \endfirsthead
-\toprule()
+\toprule\noalign{}
 \begin{minipage}[b]{\linewidth}\raggedright
@@
 \end{minipage} \\
-\midrule()
+\midrule\noalign{}
 \endhead
+\bottomrule\noalign{}
+\endlastfoot
 Standard Length & \texttt{stl} & Distance from anterior tip of
@@
 spine (when present) \\
-\bottomrule()
 \end{longtable}
@@
 (\protect\hyperlink{ref-little2002statistical}{2002})) to impute the sex
-of the fossils. Imputations were performed using the MICE (Buuren and
-Groothuis-Oudshoorn (\protect\hyperlink{ref-MICE}{2011})) algorithm and
-implemented in R (Team (\protect\hyperlink{ref-R2022language}{2022}))
-with M = 100 imputations. After imputing the sex, we fit a modified
-Orenstein-Uhlenbeck (OU; Uhlenbeck and Ornstein
-(\protect\hyperlink{ref-OUProcess}{1930})) model to each imputed data
-set using a Bayesian framework to \{THIS NEEDS MORE\}. We then
-quantified sexual dimorphism as the \{THIS NEEDS MORE; MAYBE WE CAN
-WRITE TOGETHER\}. We describe the details of the imputation model and OU
-modeling next.
-
-\hypertarget{sec:models}{%
-\section{Models}\label{sec:models}}
-
-\hypertarget{imputation-model}{%
-\subsection{Imputation Model}\label{imputation-model}}
-
-Let \(\boldsymbol{W}\) be an \((n_{extant} + n_{fossil}) \times 1\)
-vector of the covariate sex of the stickleback fish and
-\(\boldsymbol{Y}\) be an \((n_{extant} + n_{fossil}) \times K\) matrix
-of the \(K\) phenotypes of interest. Because the sex of the fossilized
-stickleback fish is unobservable, we further define
+of the fossils. Let \(\boldsymbol{W}\) be an
+\((n_{extant} + n_{fossil}) \times 1\) vector of the covariate sex of
+the stickleback fish and \(\boldsymbol{Y}\) be an
+\((n_{extant} + n_{fossil}) \times K\) matrix of the \(K\) phenotypes of
+interest. Because the sex of the fossilized stickleback fish is
+unobservable, we further define
 \(\boldsymbol{W} = (\boldsymbol{W}_{extant}^T,\boldsymbol{W}_{fossil}^T)^T\)
@@
 the \(n_{extant} \times 1\) and \(n_{fossil} \times 1\) vectors of the
-observed extant sex and missing fossil sex, respectively.
-
-\textcolor{red}{AKHIL'S VALIDATION OF THE IMPUTATION MODEL GOES HERE}
-
-We impute missing sex for the fossil data by sampling from the posterior
+observed extant sex and missing fossil sex, respectively. We imputeD
+missing sex for the fossil data by sampling from the posterior
 predictive distribution
@@
 (Buuren and Groothuis-Oudshoorn (\protect\hyperlink{ref-MICE}{2011}))
-with predictive mean matching. Traditionally, the choice for the number
-of completed dats sets is a relatively small number such as \(M = 5\) or
-\(M = 10\). However, Zhou and Reiter
-(\protect\hyperlink{ref-ZhouReiter2010}{2010}) recommend a larger number
-of imputed data sets if users intend on performing Bayesian analysis
-after imputation. Therefore, we imputed \(M = 100\) complet datasets.
+with predictive mean matching, implemented in R (CITATION).
+Traditionally, the choice for the number of completed dats sets is a
+relatively small number such as \(M = 5\) or \(M = 10\). However, Zhou
+and Reiter (\protect\hyperlink{ref-ZhouReiter2010}{2010}) recommend a
+larger number of imputed data sets if users intend on performing
+Bayesian analysis after imputation. Therefore, we imputed \(M = 100\)
+complete datasets.
 
 \textcolor{red}{presumably this also imputes missing data in the other traits? We should probably say that explicitly. -YS}
-
+\textcolor{red}{AKHIL'S METHODS FOR VALIDATION OF THE IMPUTATION MODEL GOES HERE}
+\textcolor{red}{THE POOLING DESCRIBED IN THE NEXT SENTENCE. WAS IT DONE, AND IF SO, HOW WAS IT USED IN DOWNSTREAM ANALYSES. MATT SAYS THAT HE RAN IS MODELS ON EACH IMPUTED DATA SET. -YS}
 We then pooled all of the draws from the posterior distribution across
 all the imputed data sets to estimate the posterior distributions of our
-parameters of interest (i.e., the trait values), rather than
-usingRubin's combining rules, following Zhou and Reiter
+parameters of interest (i.e., the trait values), rather than using
+Rubin's combining rules, following Zhou and Reiter
 (\protect\hyperlink{ref-ZhouReiter2010}{2010}). We used these posterior
@@
 
-\hypertarget{bayesian-ou-modeling}{%
-\subsection{Bayesian OU modeling}\label{bayesian-ou-modeling}}
-
-For Bayesian modeling, we dropped the extant data and focused on
-modeling the evolution of sexual dimorphism through time in the fossils.
+\hypertarget{modeling-trait-means-through-time-for-males-and-females}{%
+\subsection{Modeling trait means through time for males and
+females}\label{modeling-trait-means-through-time-for-males-and-females}}
+
+The goal of our analysis was to estimate fossil trait means through
+time, by sex, while incorporating uncertainty in the imputed data,
+especially uncertainty sex classification. We took a Bayesian approach
+because imputation is the process of sampling from a posterior
+predictive distribution of missing data, given observed data.
+\textcolor{red}{THIS NEEDS MORE WORK TO JUSTIFY THE BAYESIAN APPROACH. -YS}
 
 \textcolor{red}{DOES THE NEXT SENTENCE CONTRADICT THE PREVIOUS ONE? BELOW, WE SAY THAT WE USED EACH IMPUTED DATA SET. ABOVE, WE SAY THAT WE POOLED ACROSS IMPUTED DATA SETS.-YS}
-For a given imputed dataset, let \(W_{ij}\) be the imputed sex and
+To estimate fossil sample means through time, we first dropped the
+extant individuals from our imputed data sets. Then, for a given imputed
+dataset, we let \(W_{ij}\) be the imputed sex and
 \(\boldsymbol{Y}_{ij}\) be the \(K \times 1\) vector of phenotypes for
 stickleback fossil \(j\) at time \(t_i\) where \(i = 1, \ldots, T\) and
-\(j = 1,\ldots,n_{t}\). Note, we are inputting only the fossil data into
-our model; connecting to the previous section, \(W_{ij}\) and
-\(\boldsymbol{Y}_{ij}\) can be interpreted as \(W_{fossil,ij}\) and
-\(\boldsymbol{Y}_{fossil,ij}\). In addition, we denote \(Y_{K,ij}\), the
-last variable in \(\boldsymbol{Y}_{ij}\), to be the standard length of
-the fish, \begin{align}
+\(j = 1,\ldots,n_{t}\). In addition, we denote \(Y_{K,ij}\), the last
+variable in \(\boldsymbol{Y}_{ij}\), to be the standard length of each
+fish individual, a measure of body size. \begin{align}
 {Y}_{K,ij} & \overset{iid}{\sim}\left\{\begin{array}{lll} \mathcal{N}(\mu_{K,ft_i},&\sigma_{K}^2), & W_{ij} = \text{Female} \\ \mathcal{N}(\mu_{K,mt_i},&\sigma_{K}^2), & W_{ij} = \text{Male} \end{array}\right..
@@
 
-It is reasonable to assume that the other continuous traits of
-stickleback fish will have some correlation with its standard length
-\textbf{(CITATION)}. We account for this by adding an additional
-parameter, \(\gamma_k\), into our model. More specifically, if
+It is reasonable to assume that the other continuous traits will have
+some correlation with standard length \textbf{(CITATION)}. We account
+for this by adding an additional parameter, \(\gamma_k\), If
 \(Y_{k,ij}\) is a continuous trait, then \begin{align}
@@
 
-If \(Y_{k,ij}\) is a discrete trait, the conventional method of
-modelling this data is by fitting a Poisson distribution. However, the
-Poisson distribution assumes that the mean and variance of \(Y_{k,ij}\)
-are equal (\(E(Y_{k,ij}) = \text{Var}(Y_{k,ij})\)), while the empirical
-fossil data may have variances that are smaller than their respective
-means. To combat this issue, we propose to fit the discrete traits to a
-generalized Poisson model as defined in
+If \(Y_{k,ij}\) is a discrete trait, the conventional modelling method
+is to fit a Poisson distribution. However, because the empirical fossil
+data violate the assumption that the Poisson variance is equal to its
+mean, we fit discrete traits to a generalized Poisson model as defined
+in
 (\protect\hyperlink{ref-GeneralizedPoisson}{\textbf{GeneralizedPoisson?}}).
@@
 \label{eq:GP_pmf}
-\end{align} where \(E(X) = \lambda\) and
-\(Var(X) = \frac{\lambda}{(1 - \alpha)^2}\). If \(\alpha > 0\), then the
-variance is greater than the mean, called overdispersion; if
-\(\alpha < 0\), then the variance is greater than the mean, called
-underdispersion; if \(\alpha = 0\), then the model degenerates to a
-Poisson distribution.
-
-We will use the generalized Poisson to model all of the discrete traits
-in our dataset. In addition, we assume that the discrete traits
-abdominal vertebrae (mav) and caudal vertebrae (mcv) also have
-correlation with the standard length of the fish (\textbf{CITATION}). If
-\(Y_{k,ij}\) is one of the above traits, then \begin{align}
+\end{align} where GP denotes a generalized Poisson distribution with
+\(E(X) = \lambda\) and \(Var(X) = \frac{\lambda}{(1 - \alpha)^2}\).
+
+In addition, we allowed that two discrete traits, abdominal vertebrae
+number (mav) and caudal vertebrae number (mcv) could also with standard
+length (\textbf{CITATION}). This is because these are vertebral counts
+that run in parallel with the standard length measurement, along the
+anterior-posterior access of the fish; these vertebrae thus contribute
+to the length of the fish directly, and may evolve in a correlated
+fashion. If \(Y_{k,ij}\) is one of the above traits, then \begin{align}
 {Y}_{k,ij} & \sim \left\{\begin{array}{ll} GP(\exp\{\mu_{k,ft_i} + \gamma_kY_{K,ij}\},\alpha_k), & W_{ij} = \text{Female} \\ GP(\exp\{\mu_{k,mt_i} + \gamma_kY_{K,ij}\},\alpha_k), & W_{ij} = \text{Male} \end{array}\right..
@@
 
-For the other discrete traits, we also assume the above model except we
-set \(\gamma_k = 0\) because of the assumption of no correlation between
-the standard length and these traits. More specifically, we assume
-\begin{align}
+For the other discrete traits, we set \(\gamma_k = 0\) because we expect
+that trait counts are set during early ontogeny, and should not change
+with size in adult fish. Thus, \begin{align}
 {Y}_{k,ij} & \sim \left\{\begin{array}{ll} GP(\exp\{\mu_{k,ft_i}\},\alpha_k), & W_{ij} = \text{Female} \\ GP(\exp\{\mu_{k,mt_i}\},\alpha_k), & W_{ij} = \text{Male} \end{array}\right..
 \label{eq:disc_ind}
-\end{align}
+\end{align} We point out that, for the discrete traits, the means are
+represented by \(\exp\{\mu_{k,ft_i}\}\) and \(\exp\{\mu_{k,mt_i}\}\) for
+ease of modeling.
 
 In the above model descriptions, \(\mu_{k,ft_i}\) and \(\mu_{k,mt_i}\)
-model the time-\(t_i\) specific mean of phenotype \(k\) for female and
-male stickleback fish, respectively. We point out that, for the discrete
-traits, the means are represented by \(\exp\{\mu_{k,ft_i}\}\) and
-\(\exp\{\mu_{k,mt_i}\}\) for ease of use in our modeling technique. We
-further set \begin{align}
+model the time-\(t_i\), sample-specific trait mean \(k\) for females and
+males, respectively. To account for the possibility that there is a time
+-dependent trend in \(\mu_{k,ft_i}\) and \(\mu_{k,mt_i}\), we further
+set \begin{align}
 \mu_{k,gt_i} = \beta_{0,kg} + \beta_{1,kg}t_i + u_{k,gt_i},
 \label{eq:mu}
-\end{align} for \(g \in \{f,m\}\) where \(\beta_{0,kg}\) and
+\end{align} for \(g \in \{f,m\}\), where \(\beta_{0,kg}\) and
 \(\beta_{1,kg}\) are regression parameters of phenotype \(Y_k\) for each
-sex, accounting for the possibility of a time-dependent trend in the
-mean structure, and \(u_{k,gt_i}\) is the corresponding residual. To
-account for potential correlations between the residuals for a given
-trait \(k\) and sex \(g\), we fit an Ornstein-Uhlenback (OU) process
-(Uhlenbeck and Ornstein (\protect\hyperlink{ref-OUProcess}{1930})). More
-specifically, define \(du_{k,gt} = u_{k,g(t + dt)} - u_{k,gt}\), the
-change in \(u_k,gt\) for a given trait \(k\) and sex \(g\) over a
-miniscule time period \(dt\). The OU process is defined as \begin{align}
+sex, and \(u_{k,gt_i}\) is the corresponding residual. To account for
+potential correlations between the residuals for a given trait \(k\) and
+sex \(g\), we fit an Ornstein-Uhlenback (OU) process (Uhlenbeck and
+Ornstein (\protect\hyperlink{ref-OUProcess}{1930})). More specifically,
+we defined \(du_{k,gt} = u_{k,g(t + dt)} - u_{k,gt}\), the change in
+\(u_k,gt\) for a given trait \(k\) and sex \(g\) over a miniscule time
+period \(dt\). The OU process is defined as \begin{align}
 du_{k,gt} = -\kappa_k u_{k,gt} dt + \tau_k dW_t,
@@
 (\protect\hyperlink{ref-OUProcess}{1930}), the closed form solution for
-the SDE in (\ref{eq:OU_cont}) is \begin{align}
+the stochastic differential equation in (\ref{eq:OU_cont}) is
+\begin{align}
 u_{k,gt_i} \overset{iid}{\sim}\mathcal{N}\left(u_{k,gt_{i-1}}\exp\{-\kappa_k(t_{i} - t_{i-1})\} , \frac{\tau_k^2(1 - \exp\{-2\kappa_k(t_{i} - t_{i-1}))\}}{2\kappa_k}\right)
@@
 initial value \(u_{k,gt_1}\) is assumed to be a (potentially unknown)
-constant, and we discuss the procedure for estimating this value below.
-
-In our empirical study, we analyze the following four nested models for
-the mean process outlined in (\ref{eq:mu}):
-
-\begin{itemize}
-\item OU with Trend: The mean process for each phenotype $k=1,\cdots,K$ and $g \in \{f,m\}$ as previously defined.
-\item OU with No Trend: No linear trend on the mean processes and the overall mean of the phenotype is constant for each sex.
-This is achieved by setting $\beta_{1,kg} = 0$ in (\ref{eq:mu}) for $k = 1,\cdots,K$ and $g \in \{f,m\}$
-\item No OU with Trend: No random fluctuations on the mean processes, i.e.
-the OU process assumption is not needed.
-This is achieved by setting $u_{k,gt_i} = 0$ in (\ref{eq:mu}) for $k = 1,\cdots,K$, $g \in \{f,m\}$, and $i = 1,\cdots,T$.
-\item No OU with No Trend: The mean for a specific trait and specific sex at each time point is a constant value.
-This is achieved by setting $\beta_{1,kg} = 0$ and $u_{k,gt_i} = 0$ in (\ref{eq:mu}) for $k = 1,\cdots,K$, $g \in \{f,m\}$, and $i = 1,\cdots,T$.
-\end{itemize}
-
-Because we are fitting a dataset with a stochastic structure on
-phenotype means, and we are interested in the distribution of the
-overall structure of these means, we analyze the data via a Bayesian
-analysis \textless THIS LAST CLAUSE DOESN'T FOLLOW FROM THE BECAUSE
-CLAUSE. BETTER EXPLAIN WHY BAYES IS GOOD HERE\textgreater. Bayesian data
-analysis is also more naturally used when we have to impute data
-(\textbf{CITATION}). To aid in the sampling procedure for the discrete
-phenotypes, we declare
+constant. This unknown constant can therefore be considered a latent
+variable, further justifying our Bayesian approach, which better
+incorporates prior information and uncertainty from missing data
+imputation in the estimate of the variable.
+
+We discuss the procedure for estimating this value below.
+\textcolor{red}{DID THIS GET DISCUSSED?-YS}
 
 Priors: To aid in the sampling procedure for the discrete phenotypes, we
-declare
+first had to transform the parameters of the generalized Poisson
+distribution onto the real number line in order to ensure our algorithm
+can properly sample from the posterior distribution of interest using
+STAN software. This was done by declaring
 \(\phi_k = \log\left(\frac{\alpha_k - \max_{i,j}(-\lambda_{k,ij}/y_{k,ij})}{1 - \alpha_k}\right)\)
@@
 \(\lambda_{k,ij} = \left\{\begin{array}{cc} \exp(\mu_{k,ft_i} + \gamma_k Y_{K,ij}) & W_{ij} = \text{Female} \\ \exp(\mu_{k,mt_i} + \gamma_k Y_{K,ij}) & W_{ij} = \text{Male} \end{array}\right.\).
-This transformation is performed to ensure our algorithm can properly
-sample from the posterior distribution of interest.
-
-For \(k = 1,\cdots,K\),
+
+We chose the following priors, designing them to be weakly informative
+such that the posterior distributions could be influenced by the data,
+if warranted. For \(k = 1,\cdots,K\),
 
@@
 \sigma_k & \overset{iid}{\sim}\mathcal{N}(0,10)I_{\{\sigma > 0\}} \nonumber \\
-\sigma_k & \overset{iid}{\sim}\mathcal{N}(0,10) \nonumber \\
 \tau_k & \overset{iid}{\sim}\mathcal{N}(0,10)I_{\{\tau > 0\}} \nonumber \\
@@
 \label{eq:priors}
-\end{align} We also note that, for the discrete phenotypes in equations
+\end{align}
+
+We also note that, for the discrete phenotypes in equations
 (\ref{eq:disc_corr}) and (\ref{eq:disc_ind}), there is no \(\sigma_k\),
-and is not sampled, and for the continuous phenotypes in equation
-(\ref{eq:cont}), there is no \(\phi_k\) and is not sampled.
+and it is not sampled, and for the continuous phenotypes in equation
+(\ref{eq:cont}), there is no \(\phi_k\) and it is not sampled.
+
+We follow the method outlined in
+(\protect\hyperlink{ref-ZhouandReiter}{\textbf{ZhouandReiter?}}) to
+obtain \(R\) draws from the posterior distribution of the outlined model
+parameters given our observed data and imputed values. More
+specifically, for the \(j^{th}\) imputed data set where
+\(j = 1,\cdots,M\), we obtain \(\frac{R}{M}\) draws from the posterior
+distribution using an MCMC algorithm with a sufficient burnin period,
+and then we combine these \(M\) MCMC draws to say that we have a total
+of \(R\) draws from the overall posterior distribution of interest,
+which can then be used to assess the presence of sexual dimorphism for a
+given trait.
 
@@
 
-Cornuault (\protect\hyperlink{ref-Cornault2022}{2022}) Bayesian OU
-model.
+After running these models, we have posterior distributions of all the
+model parameters of interest. From these, we can obtain samples from the
+posterior distribution of mean trait values for each trait for males and
+females, using equation 6. We use these values for the following
+analyses.
+
+\hypertarget{quantifying-sexual-dimorphism-and-testing-for-increased-sexual-dimorphism-through-time.}{%
+\subsection{Quantifying sexual dimorphism, and testing for increased
+sexual dimorphism through
+time.}\label{quantifying-sexual-dimorphism-and-testing-for-increased-sexual-dimorphism-through-time.}}
+
+We identify sexual dimorphism first by calculating the difference in
+mean trait size for males and females at each time for phenotype \(k\)
+(\(\mu_{k,m,t_i} - \mu_{k,m,t_i}\)) for \(i = 1,\cdots,T\) where, for
+the continuous phenotypes, the mean trait sizes for each sex
+\(\mu_{k,m,t_i}\) and \(\mu_{k,f,t_i}\) are calculated using
+(\ref{eq:mu}). For the discrete phenotypes, we note that in
+(\ref{eq:disc_corr}) and (\ref{eq:disc_ind}) the mean trait size is
+\(\lambda_{k,m,t_i} = \exp(\mu_{k,m,t_i})\) and
+\(\lambda_{k,f,t_i} = \exp(\mu_{k,f,t_i})\) for males and females,
+respectively, which is the value we use in our analysis. For simplicity,
+we will use \(\mu_{k,m,t_i}\) and \(\mu_{k,f,t_i}\) as the notation for
+mean size for both discrete and continuous traits for the rest of this
+discussion. From our MCMC algorithm, we obtain \(M\) samples from the
+posterior distribution of \(\mu_{k,m,t_i} - \mu_{k,f,t_i}\) to use in
+our analysis.
+
+Recall that we predicted that sexual dimorphism should increase through
+time in Lineage II, following ecological release from predators.
+However, this change in dimorphism make be increasing at a non-constant
+rate, as pointed out by the inclusion of the OU process. We assess for
+dimorphism by calculating a Kendall's \(\tau\) coefficient (KENDALL
+(\protect\hyperlink{ref-KendallsTau}{1938})). Kendall's \(\tau\)
+assesses the ordinal association between a set of bivariate
+observations, where \(\tau > 0\) indicates concordance between the
+observations (i.e., as \(x\) increases, \(y\) also tends to increase.)
+
+For our analysis, we calculate
+\(\boldsymbol{\tau}_k = \{\tau_{k,r}: r = 1,\cdots,R\}\) for trait \(k\)
+where \(\tau_r\) is the estimate of the Kendall's \(\tau\) correlation
+for \(\mu_{k,m,t_i} - \mu_{k,f,t_i}\) versus time \(t_i\) for the
+\(r^{th}\) MCMC sample from the posterior distribution. The posterior
+probability of sexual dimorphism for trait \(k\) can be estimated as the
+proportion of values in \(\boldsymbol{\tau}_k\) that are greater than 0
+(\(\frac{1}{R} \sum_{r=1}^R I(\tau_{k,r} > 0)\)). We can also assess the
+strength of the dimorphism by calculating the proportion of posterior
+samples that are greater than a prespecified threshold
+\textcolor{red}{(0.5??, idk what this could be.) -MS}
 
@@
 
-\textless WE NEED RESULTS SECTION DESCRIBING THE RESULTS OF AKHIL'S
-MULTIPLE IMPUTATION, BOTH THE TESTS OF HOW WELL THE MODEL WORKED
-CLASSIFYING MODEL SAMPLES, AND SUMMARY STATISTICS OF FOSSIL IMPUTATION
-RESULTS -YS\textgreater{}
-
-Within each trait, the model with the lowest DIC was chosen for that
-trait . All subsequent results presented here use the model with the
-lowest DIC individually by trait .
-
-Table \ref{whichmodel} reports the results of model fitting. X of the Y
-traits were best fit by a trend model. Best models included an OU
-component for caudal and abdominal vertebra number (mcv, mav), pelvic
-spine length (lpt ) and standard length (stl). Caudal vertebra number
-was the only trait best fit by a model including both a trend and an OU
-component. This implies that not only does caudal vertebra size have a
-linear trend across sex \textless WHAT DO YOU MEAN BY LINEAR TREND
-ACROSS SEX?\textgreater, but the average size at one time period is
-highly correlated with the average size at the next time period
-\textless I'VE ALWAYS FOUND THAT THIS AID TO MODEL INTEPRETATION IS
-BETTER IN THE METHODS, JUST AFTER THE MODELS ARE DESCRIBED\textgreater.
-
-\textcolor{red} {
-Yoel's interpretation of the above paragraph.
-Taking the model with the lowest DIC score as the best model for each trait, we found that cle, ds1, lps, and pmx did not fit either the trend model or the OU model for male-female trait mean differences.
-In contrast, we found a trend through time in male-female trait mean differences (i.e., sexual dimorphism) for mcv, ds2, ds3, ect, maf, and mdf.
-This suggests that these traits changed their dimorphism from one time period to the next, with an overall tendency toward ???GREATER/LESS???
-dimorphism.
-IS THIS CORRECT?
-- YS
-} \textcolor{red} {
-We found evidence for an OU process but no trend for lpt, mav, and stl, suggesting that these traits varied through time, about a fixed value for dimorphism.
-Finally, we found evidence that change in mcv fitted both a trend an an OU process, meaning ???
-WHAT BIOLOGICALLY ???.
-}
-
-\textcolor{red} {HOW DO WE RECONCILE THE PRESENCE OF TRENDS IN THESE MODELS WITH AN ABSENCE OF VISIBLE TRENDS IN FIGURES 2 THROUGH 4?
-I WONDER IF THIS RESULT PARAGRAPH BELONGS BEST AT THE END OF THE RESULTS, RATHER THAN THE BEGINNING.
-CONSIDER THE STATISTICAL EVIDENCE FOR TRENDS AND OU AFTER WE'VE WALKED THE READERS THROUGH VISUAL EVIDENCE FOR CHANGE.
-}
-
-\hypertarget{best-fitting-models}{%
-\subsection{Best fitting models}\label{best-fitting-models}}
-
-\begin{table}[ht]
-\centering
-\begin{tabular}{|c||c|c|}
-  \hline
- & OU & No OU  \\ 
-  \hline
-  \hline
-Trend  &  mcv  & ds2, ds3, ect, maf, mdf \\ 
-  \hline
-  No Trend & lpt, mav, stl  &  cle, ds1, lps, pmx\\
-   \hline
-\end{tabular}
-\label{whichmodel}
-\end{table}
+\textcolor{red}{WE NEED RESULTS SECTION DESCRIBING THE RESULTS OF AKHIL'S MULTIPLE IMPUTATION, BOTH THE TESTS OF HOW WELL THE MODEL WORKED CLASSIFYING MODEL SAMPLES, AND SUMMARY STATISTICS OF FOSSIL IMPUTATION RESULTS -YS}
 
@@
   >{\raggedleft\arraybackslash}p{(\columnwidth - 26\tabcolsep) * \real{0.0723}}@{}}
-\toprule()
+\toprule\noalign{}
 \begin{minipage}[b]{\linewidth}\raggedleft
@@
 \end{minipage} \\
-\midrule()
+\midrule\noalign{}
 \endhead
+\bottomrule\noalign{}
+\endlastfoot
 1 & 0.913 & 0.215 & 0.359 & 0.203 & 0.865 & 0.428 & 0.853 & 0.944 &
@@
 0.705 & 0.722 & 0.999 & 0.859 & 0.717 \\
-\bottomrule()
 \end{longtable}
@@
 \end{minipage} \\
-\midrule()
+\midrule\noalign{}
 \endhead
+\bottomrule\noalign{}
+\endlastfoot
 1 & 0.500 & 0.500 & 0.500 & 0.500 & 0.500 & 0.500 & 0.500 & 0.500 &
@@
 0.281 & 0.356 & 0.212 & 0.622 & 0.627 \\
-\bottomrule()
 \end{longtable}
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Cornault2022}{}}%
-Cornuault, J. 2022. {``Bayesian Analyses of Comparative Data with the
-Ornstein--Uhlenbeck Model: Potential Pitfalls.''} \emph{Systematic
-Biology} 71 (6): 1524--40. \url{https://doi.org/10.1093/sysbio/syac036}.
-
 \leavevmode\vadjust pre{\hypertarget{ref-DeLisleetal2018}{}}%
@@
 Lineage.''} \emph{Evolution} 62 (3): 700--710.
+
+\leavevmode\vadjust pre{\hypertarget{ref-KendallsTau}{}}%
+KENDALL, M. G. 1938. {``{A NEW MEASURE OF RANK CORRELATION}.''}
+\emph{Biometrika} 30 (1-2): 81--93.
+\url{https://doi.org/10.1093/biomet/30.1-2.81}.
 

### 2024-10-21 manuscript/paper.Rmd
Update Manuscript
YS. editing, revising, and adding to the manuscript.

@@
 
-THIS FIGURE DOESN'T SHOW MUCH. LET'S CUT FOR SPACE. 
-```{r fig.cap = "Fossil sample size for each time period"}
-ggplot(aes(x = time, y = count), data = fortab) + geom_col() + xlab("Sample")  + scale_x_continuous(breaks = 1:18, labels = 1:18)
-```
-
 | Trait Name  | Trait Code | Trait Description           |
-|----------------------|------------|-----------------------------|
+|---------|-----|------------------------------------------|
 |Standard Length|	`stl` |	Distance from anterior tip of premaxilla to posterior end of last vertebra (hypural plate)|
@@
 |Premaxilla|`pmx.sc`|	Length from the anterior tip of the premaxilla to the distal tip of the ascending process of the premaxilla |
-|Dorsal Spine|`Ds#.sc# = 1,2,or 3`|	Length from the base of a dorsal spine above the pterygiophore to its distal tip along the anterior edge|
+|Dorsal Spine|`Ds#.sc #=1/2/3`|	Length from the base of a dorsal spine above the pterygiophore to its distal tip along the anterior edge|
 |Pterygiophore|`lpt.sc`|	Distance between the anterior to posterior tips of the pterygiophore immediately preceding the 3rd dorsal spine (when present)|
 
+\textcolor{red}{This table caption should be part of the table. Top row. at the very least, it should be the width of the table. -YS}
 Table: Traits and trait descriptions. ‘sc’ denotes size correction of trait against standard length. Names of bones follow @Bowne1994 unless otherwise noted.
-
-<!-- I removed the table and replaced it with a figure -->
-<!-- % latex table generated in R 4.2.2 by xtable 1.8-4 package -->
-<!-- % Mon May  1 11:20:28 2023 -->
-<!-- \begin{table}[ht] -->
-<!-- \centering -->
-<!-- \caption{The number of stickleback fossils at each time point.  Sample size at each time point ranges from a low of 22 to a high of 67.} -->
-<!-- \begin{tabular}{rl} -->
-<!--   \hline -->
-<!-- time & count \\  -->
-<!--   \hline -->
-<!--   1 & 43 \\  -->
-<!--     2 & 41 \\  -->
-<!--     3 & 51 \\  -->
-<!--     4 & 41 \\  -->
-<!--     5 & 46 \\  -->
-<!--     6 & 48 \\  -->
-<!--     7 & 67 \\  -->
-<!--     8 & 55 \\  -->
-<!--     9 & 42 \\  -->
-<!--    10 & 33 \\  -->
-<!--    11 & 37 \\  -->
-<!--    12 & 22 \\  -->
-<!--    13 & 41 \\  -->
-<!--    14 & 43 \\  -->
-<!--    15 & 46 \\  -->
-<!--    16 & 47 \\  -->
-<!--    17 & 56 \\  -->
-<!--    18 & 55 \\  -->
-<!--    \hline -->
-<!-- \end{tabular} -->
-<!-- \label{tabn} -->
-
-<!-- \end{table} -->
 
@@
 Measurements from photographs and X-rays were taken with FIJI (@Schindelinetal2012) and its plugin ObjectJ (https://sils.fnwi.uva.nl/bcb/objectj/).
-
 We dissected individuals from the generalist populations (Table S1) to determine sex from the gonads.
-Individuals from the species-pair lakes (TableS1) were previously sexed by the Schluter lab, using either dissection or a genotyping protocol (Whom, personal communication).
-
-The extant data used here consists of a total of `r sum(!is.na(stick$gender))` specimens all with known sex (Table S1).
-Of these, there are `r sum(stick$gender == "Female",na.rm = TRUE)` and `r sum(stick$gender == "Male",na.rm = TRUE)` female and male specimens, respectively.
+Individuals from the species-pair lakes (TableS1) were previously sexed by the Schluter lab, using either dissection or a genotyping protocol \textcolor{red}{(Whom, personal communication)}.
 
 ## Outlier analysis and size correction. 
-
 To check for outliers, we calculated within-group means and standard deviations for each trait separately for K series fossil specimens (pooled across samples) and for extant specimens (within generalist, benthic, or limnetic categories).
@@
 ## Missing data imputation, including fossil sex
-Briefly, We used multiple imputation (@little2002statistical) to impute the sex of the fossils.
+Briefly, we used multiple imputation (@little2002statistical) to impute the sex of the fossils.
 Let $\boldsymbol{W}$ be an $(n_{extant} + n_{fossil}) \times 1$ vector of the covariate sex of the stickleback fish and $\boldsymbol{Y}$ be an $(n_{extant} + n_{fossil}) \times K$ matrix of the $K$ phenotypes of interest.
 Because the sex of the fossilized stickleback fish is unobservable, we further define $\boldsymbol{W} = (\boldsymbol{W}_{extant}^T,\boldsymbol{W}_{fossil}^T)^T$ where $\boldsymbol{W}_{extant}$ and $\boldsymbol{W}_{fossil}$ are the $n_{extant} \times 1$ and $n_{fossil} \times 1$ vectors of the observed extant sex and missing fossil sex, respectively.
-We imputeD missing sex for the fossil data by sampling from the posterior predictive distribution $P(\boldsymbol{W}_{fossil}|\boldsymbol{W}_{extant},\boldsymbol{Y})$ using the multiple imputation by chained equations (MICE) algorithm (@MICE) with predictive mean matching, implemented in R (CITATION).
+We imputed missing sex for the fossil data by sampling from the posterior predictive distribution $P(\boldsymbol{W}_{fossil}|\boldsymbol{W}_{extant},\boldsymbol{Y})$ using the multiple imputation by chained equations (MICE) algorithm (@MICE) with predictive mean matching, implemented in R (CITATION).
 Traditionally, the choice for the number of completed dats sets is a relatively small number such as $M = 5$ or $M = 10$.
@@
 Therefore, we imputed $M = 100$ complete datasets.
-
-\textcolor{red}{presumably this also imputes missing data in the other traits? We should probably say that explicitly. -YS}
-\textcolor{red}{AKHIL'S METHODS FOR VALIDATION OF THE IMPUTATION MODEL GOES HERE}
-\textcolor{red}{THE POOLING DESCRIBED IN THE NEXT SENTENCE. WAS IT DONE, AND IF SO, HOW WAS IT USED IN DOWNSTREAM ANALYSES. MATT SAYS THAT HE RAN IS MODELS ON EACH IMPUTED DATA SET. -YS}
-We then pooled all of the draws from the posterior distribution across all the imputed data sets to estimate the posterior distributions of our parameters of interest (i.e., the trait values), rather than using Rubin's combining rules, following @ZhouReiter2010.
-We used these posterior distributions in our Bayesian analysis.
-
-## Modeling trait means through time for males and females
-The goal of our analysis was to estimate fossil trait means through time, by sex, while incorporating uncertainty in the imputed data, especially uncertainty sex classification.
-We took a Bayesian approach because imputation is the process of sampling from a posterior predictive distribution of missing data, given observed data.
-\textcolor{red}{THIS NEEDS MORE WORK TO JUSTIFY THE BAYESIAN APPROACH.
+\textcolor{red}{PRESUMABLY this also imputes missing data in the other traits?
+We should probably say that explicitly.
 -YS}
 
-\textcolor{red}{DOES THE NEXT SENTENCE CONTRADICT THE PREVIOUS ONE? BELOW, WE SAY THAT WE USED EACH IMPUTED DATA SET. ABOVE, WE SAY THAT WE POOLED ACROSS IMPUTED DATA SETS.-YS}
-To estimate fossil sample means through time, we first dropped the extant individuals from our imputed data sets.
-Then, for a given imputed dataset, we let $W_{ij}$ be the imputed sex and $\boldsymbol{Y}_{ij}$ be the $K \times 1$ vector of phenotypes for stickleback fossil $j$ at time $t_i$ where $i = 1, \ldots, T$ and $j = 1,\ldots,n_{t}$.
+\textcolor{red}{AKHIL'S METHODS FOR VALIDATION OF THE IMPUTATION MODEL GOES HERE}
+
+We then pooled all of the draws from the posterior distribution across all the imputed data sets to estimate the posterior distributions of our parameters of interest (i.e., the trait values), rather than using Rubin's combining rules, following @ZhouReiter2010.
+We then dropped extant fish from our imputed data sets and used these posterior distributions in our downstream analyses of fossil sexual dimorphism.
+
+## Modeling fossil stickleback trait means through time for males and females
+The goal of our analysis was to estimate fossil trait means through time, by sex, while incorporating uncertainty across imputations in the classification of sex.
+We used a Bayesian approach to estimate trait means in part because imputation is the process of sampling from a posterior predictive distribution of missing data, given observed data.
+\textcolor{red}{THIS NEEDS MORE WORK TO JUSTIFY THE BAYESIAN APPROACH, EXPLAINING IT TO MY AUDIENCE.
+-YS}
+
+To estimate fossil sample means through time, for a given imputed dataset, we let $W_{ij}$ be the imputed sex and $\boldsymbol{Y}_{ij}$ be the $K \times 1$ vector of phenotypes for stickleback fossil $j$ at time $t_i$ where $i = 1, \ldots, T$ and $j = 1,\ldots,n_{t}$.
 In addition, we denote $Y_{K,ij}$, the last variable in $\boldsymbol{Y}_{ij}$, to be the standard length of each fish individual, a measure of body size.
@@
 
-It is reasonable to assume that the other continuous traits will have some correlation with standard length \textbf{(CITATION)}.
+Continuous traits will likely have some correlation with standard length (i.e., allometry) (@Huxley1932)(@Vojeetal2022)(@Vojeetal2014).
 We account for this by adding an additional parameter, $\gamma_k$, If $Y_{k,ij}$ is a continuous trait, then
@@
 If $Y_{k,ij}$ is a discrete trait, the conventional modelling method is to fit a Poisson distribution.
-However, because the empirical fossil data violate the assumption that the Poisson variance is equal to its mean, we fit discrete traits to a generalized Poisson model as defined in @GeneralizedPoisson.
+However, the empirical fossil data violate the assumption that the Poisson variance is equal to its mean, so we fitted discrete traits to a generalized Poisson model (@GeneralizedPoisson).
 Specifically, if $X \sim GP(\lambda,\alpha)$, then
@@
 
-In addition, we allowed that two discrete traits, abdominal vertebrae number (mav) and caudal vertebrae number (mcv) could also with  standard length (\textbf{CITATION}).
-This is because these are vertebral counts that run in parallel with the standard length measurement, along the anterior-posterior access of the fish; these vertebrae thus contribute to the length of the fish directly, and may evolve in a correlated fashion.
+In addition, we modeled two discrete traits, abdominal vertebrae number (mav) and caudal vertebrae number (mcv) to vary with standard length.
+This is because these are vertebral counts that made along the anterior-posterior access of the fish; these vertebrae contribute to the length of the fish directly and may evolve in a correlated fashion.
 If $Y_{k,ij}$ is one of the above traits, then
@@
 
-For the other discrete traits, we set $\gamma_k = 0$ because we expect that trait counts are set during early ontogeny, and should not change with size in adult fish. Thus,
+For the other discrete counting traits, we set $\gamma_k = 0$ because we expect that trait counts are set during early ontogeny and should not change with size in adult fish. Thus,
 \begin{align}
@@
 \end{align}
-We point out that, for the discrete traits, the means are represented by $\exp\{\mu_{k,ft_i}\}$ and $\exp\{\mu_{k,mt_i}\}$ for ease of modeling.
+We point out that counting trait means are represented by $\exp\{\mu_{k,ft_i}\}$ and $\exp\{\mu_{k,mt_i}\}$ for ease of modeling.
 
 In the above model descriptions, $\mu_{k,ft_i}$ and $\mu_{k,mt_i}$ model the time-$t_i$, sample-specific trait mean $k$ for females and males, respectively.
-To account for the possibility that there is a time -dependent trend in $\mu_{k,ft_i}$ and $\mu_{k,mt_i}$, we further set
+To account for the possibility that there is a time-dependent (i.e., evolutionary trend) trend in $\mu_{k,ft_i}$ and $\mu_{k,mt_i}$, we further set
 \begin{align}
@@
 In a traditional OU process, the initial value $u_{k,gt_1}$ is assumed to be a (potentially unknown) constant.
-This unknown constant can therefore be considered a latent variable, further justifying our Bayesian approach, which better incorporates prior information and uncertainty from missing data imputation in the estimate of the variable.
+This unknown constant can therefore be considered a latent variable, further justifying our Bayesian approach, which better incorporates prior information and uncertainty from missing data imputation in the estimate of latent variables.
 
 We discuss the procedure for estimating this value below. \textcolor{red}{DID THIS GET DISCUSSED?-YS}
-
-<!-- In our empirical study, we analyze the following four nested models for the mean process outlined in (\ref{eq:mu}): -->
-<!-- \begin{itemize} -->
-<!-- \item OU with Trend: The mean process for each phenotype $k=1,\cdots,K$ and $g \in \{f,m\}$ as previously defined. -->
-<!-- \item OU with No Trend: No linear trend on the mean processes and the overall mean of the phenotype is constant for each sex.
-This is achieved by setting $\beta_{1,kg} = 0$ in (\ref{eq:mu}) for $k = 1,\cdots,K$ and $g \in \{f,m\}$ -->
-<!-- \item No OU with Trend: No random fluctuations on the mean processes, i.e.
-the OU process assumption is not needed.
-This is achieved by setting $u_{k,gt_i} = 0$ in (\ref{eq:mu}) for $k = 1,\cdots,K$, $g \in \{f,m\}$, and $i = 1,\cdots,T$.
--->
-<!-- \item No OU with No Trend: The mean for a specific trait and specific sex at each time point is a constant value.
-This is achieved by setting $\beta_{1,kg} = 0$ and $u_{k,gt_i} = 0$ in (\ref{eq:mu}) for $k = 1,\cdots,K$, $g \in \{f,m\}$, and $i = 1,\cdots,T$.
--->
-<!-- \end{itemize} -->
 
@@
 
-All models were built using @R2022language and (\textbf STAN)
-
 After running these models, we have posterior distributions of all the model parameters of interest.
@@
 \textcolor{red}{WE NEED RESULTS SECTION DESCRIBING THE RESULTS OF AKHIL'S MULTIPLE IMPUTATION, BOTH THE TESTS OF HOW WELL THE MODEL WORKED CLASSIFYING MODEL SAMPLES, AND SUMMARY STATISTICS OF FOSSIL IMPUTATION RESULTS -YS}
-
-<!-- Within each trait, the model with the lowest DIC was chosen for that trait <THIS IS METHODS--WHERE IS THIS DESCRIBED IN THE METHODS>.
-All subsequent results presented here use the model with the lowest DIC individually by trait <THIS IS METHODS--WHERE IS THIS DESCRIBED IN THE METHODS>.
--->
-
-<!-- Table \ref{whichmodel} reports the results of model fitting.
-X of the Y traits were best fit by a trend model.
-Best models included an OU component for caudal and abdominal vertebra number (mcv, mav), pelvic spine length (lpt <LPT IS PTERYGIOPHORE LENGTH>) and standard length (stl).
-Caudal vertebra number was the only trait best fit by a model including both a trend and an OU component.
-This implies that not only does caudal vertebra size have a linear trend across sex <WHAT DO YOU MEAN BY LINEAR TREND ACROSS SEX?>, but the average size at one time period is highly correlated with the average size at the next time period <I'VE ALWAYS FOUND THAT THIS AID TO MODEL INTEPRETATION IS BETTER IN THE METHODS, JUST AFTER THE MODELS ARE DESCRIBED>.
--->
-
-<!-- \textcolor{red} { -->
-<!-- Yoel's interpretation of the above paragraph.
-Taking the model with the lowest DIC score as the best model for each trait, we found that cle, ds1, lps, and pmx did not fit either the trend model or the OU model for male-female trait mean differences.
-In contrast, we found a trend through time in male-female trait mean differences (i.e., sexual dimorphism) for mcv, ds2, ds3, ect, maf, and mdf.
-This suggests that these traits changed their dimorphism from one time period to the next, with an overall tendency toward ???GREATER/LESS???
-dimorphism.
-IS THIS CORRECT?
-- YS -->
-<!-- } -->
-<!-- \textcolor{red} { -->
-<!-- We found evidence for an OU process but no trend for lpt, mav, and stl, suggesting that these traits varied through time, about a fixed value for dimorphism.
-Finally, we found evidence that change in mcv fitted both a trend an an OU process, meaning ???
-WHAT BIOLOGICALLY ???.
--->
-<!-- } -->
-
-<!-- \textcolor{red} {HOW DO WE RECONCILE THE PRESENCE OF TRENDS IN THESE MODELS WITH AN ABSENCE OF VISIBLE TRENDS IN FIGURES 2 THROUGH 4?
-I WONDER IF THIS RESULT PARAGRAPH BELONGS BEST AT THE END OF THE RESULTS, RATHER THAN THE BEGINNING.
-CONSIDER THE STATISTICAL EVIDENCE FOR TRENDS AND OU AFTER WE'VE WALKED THE READERS THROUGH VISUAL EVIDENCE FOR CHANGE.
--->
-<!-- } -->
-
-<!-- ## Best fitting models  -->
-<!-- \begin{table}[ht] -->
-<!-- \centering -->
-<!-- \begin{tabular}{|c||c|c|} -->
-<!--   \hline -->
-<!--  & OU & No OU  \\  -->
-<!--   \hline -->
-<!--   \hline -->
-<!-- Trend  &  mcv  & ds2, ds3, ect, maf, mdf \\  -->
-<!--   \hline -->
-<!--   No Trend & lpt, mav, stl  &  cle, ds1, lps, pmx\\ -->
-<!--    \hline -->
-<!-- \end{tabular} -->
-<!-- \label{whichmodel} -->
-<!-- \end{table} -->
-
-<!-- Trends terms included in were observed in ds2, ds3, ect, maf, mdf, mcv,  -->
-<!-- OU term included in model: mcv, lpt, stl -->
-<!-- model_list[["cle"]] <- 3 -->
-<!-- model_list[["ds1"]] <- 3 -->
-<!-- model_list[["ds2"]] <- 1 -->
-<!-- model_list[["ds3"]] <- 1 -->
-<!-- model_list[["ect"]] <- 1 -->
-<!-- model_list[["lps"]] <- 3 -->
-<!-- model_list[["lpt"]] <- 2 -->
-<!-- model_list[["maf"]] <- 1 -->
-<!-- model_list[["mav"]] <- 2 -->
-<!-- model_list[["mcv"]] <- 0 -->
-<!-- model_list[["mdf"]] <- 1 -->
-<!-- #model_list[["mds"]] <- 3 -->
-<!-- #model_list[["mpt"]] <- 0 -->
-<!-- model_list[["pmx"]] <- 3 -->
-<!-- model_list[["stl"]] <- 2 -->
-<!-- #model_list[["tpg"]] <- 1 -->
-<!-- Greg notes:  -->
-<!-- - mdf and maf have interesting trends.   -->
-<!-- - ds3 flips over time.  Females have larger values to start and over the course of the time frame wer are looking at males end up with larger mean values.   -->
-<!-- - Most of the other traits have mean differences that are relatively constant over time.   -->
+The extant data used here consists of a total of `r sum(!is.na(stick$gender))` specimens all with known sex (Table S1).
+Of these, there are `r sum(stick$gender == "Female",na.rm = TRUE)` and `r sum(stick$gender == "Male",na.rm = TRUE)` female and male specimens, respectively.
 
@@
 # References
+
+<!-- SCRAP CODE and TEXT. YS moved it down here on 21 October 2024 because it was making it hard for me to see the manuscript.  -->
+<!-- THIS FIGURE DOESN'T SHOW MUCH. LET'S CUT FOR SPACE.  -->
+```{r fig.cap = "Fossil sample size for each time period",include=FALSE}
+ggplot(aes(x = time, y = count), data = fortab) + geom_col() + xlab("Sample")  + scale_x_continuous(breaks = 1:18, labels = 1:18)
+```
+<!-- I removed the table and replaced it with a figure -->
+<!-- % latex table generated in R 4.2.2 by xtable 1.8-4 package -->
+<!-- % Mon May  1 11:20:28 2023 -->
+<!-- \begin{table}[ht] -->
+<!-- \centering -->
+<!-- \caption{The number of stickleback fossils at each time point.  Sample size at each time point ranges from a low of 22 to a high of 67.} -->
+<!-- \begin{tabular}{rl} -->
+<!--   \hline -->
+<!-- time & count \\  -->
+<!--   \hline -->
+<!--   1 & 43 \\  -->
+<!--     2 & 41 \\  -->
+<!--     3 & 51 \\  -->
+<!--     4 & 41 \\  -->
+<!--     5 & 46 \\  -->
+<!--     6 & 48 \\  -->
+<!--     7 & 67 \\  -->
+<!--     8 & 55 \\  -->
+<!--     9 & 42 \\  -->
+<!--    10 & 33 \\  -->
+<!--    11 & 37 \\  -->
+<!--    12 & 22 \\  -->
+<!--    13 & 41 \\  -->
+<!--    14 & 43 \\  -->
+<!--    15 & 46 \\  -->
+<!--    16 & 47 \\  -->
+<!--    17 & 56 \\  -->
+<!--    18 & 55 \\  -->
+<!--    \hline -->
+<!-- \end{tabular} -->
+<!-- \label{tabn} -->
+
+<!-- \end{table} -->
+
+<!-- Within each trait, the model with the lowest DIC was chosen for that trait <THIS IS METHODS--WHERE IS THIS DESCRIBED IN THE METHODS>.
+All subsequent results presented here use the model with the lowest DIC individually by trait <THIS IS METHODS--WHERE IS THIS DESCRIBED IN THE METHODS>.
+-->
+
+<!-- Table \ref{whichmodel} reports the results of model fitting.
+X of the Y traits were best fit by a trend model.
+Best models included an OU component for caudal and abdominal vertebra number (mcv, mav), pelvic spine length (lpt <LPT IS PTERYGIOPHORE LENGTH>) and standard length (stl).
+Caudal vertebra number was the only trait best fit by a model including both a trend and an OU component.
+This implies that not only does caudal vertebra size have a linear trend across sex <WHAT DO YOU MEAN BY LINEAR TREND ACROSS SEX?>, but the average size at one time period is highly correlated with the average size at the next time period <I'VE ALWAYS FOUND THAT THIS AID TO MODEL INTEPRETATION IS BETTER IN THE METHODS, JUST AFTER THE MODELS ARE DESCRIBED>.
+-->
+
+<!-- \textcolor{red} { -->
+<!-- Yoel's interpretation of the above paragraph.
+Taking the model with the lowest DIC score as the best model for each trait, we found that cle, ds1, lps, and pmx did not fit either the trend model or the OU model for male-female trait mean differences.
+In contrast, we found a trend through time in male-female trait mean differences (i.e., sexual dimorphism) for mcv, ds2, ds3, ect, maf, and mdf.
+This suggests that these traits changed their dimorphism from one time period to the next, with an overall tendency toward ???GREATER/LESS???
+dimorphism.
+IS THIS CORRECT?
+- YS -->
+<!-- } -->
+<!-- \textcolor{red} { -->
+<!-- We found evidence for an OU process but no trend for lpt, mav, and stl, suggesting that these traits varied through time, about a fixed value for dimorphism.
+Finally, we found evidence that change in mcv fitted both a trend an an OU process, meaning ???
+WHAT BIOLOGICALLY ???.
+-->
+<!-- } -->
+
+<!-- \textcolor{red} {HOW DO WE RECONCILE THE PRESENCE OF TRENDS IN THESE MODELS WITH AN ABSENCE OF VISIBLE TRENDS IN FIGURES 2 THROUGH 4?
+I WONDER IF THIS RESULT PARAGRAPH BELONGS BEST AT THE END OF THE RESULTS, RATHER THAN THE BEGINNING.
+CONSIDER THE STATISTICAL EVIDENCE FOR TRENDS AND OU AFTER WE'VE WALKED THE READERS THROUGH VISUAL EVIDENCE FOR CHANGE.
+-->
+<!-- } -->
+
+<!-- ## Best fitting models  -->
+<!-- \begin{table}[ht] -->
+<!-- \centering -->
+<!-- \begin{tabular}{|c||c|c|} -->
+<!--   \hline -->
+<!--  & OU & No OU  \\  -->
+<!--   \hline -->
+<!--   \hline -->
+<!-- Trend  &  mcv  & ds2, ds3, ect, maf, mdf \\  -->
+<!--   \hline -->
+<!--   No Trend & lpt, mav, stl  &  cle, ds1, lps, pmx\\ -->
+<!--    \hline -->
+<!-- \end{tabular} -->
+<!-- \label{whichmodel} -->
+<!-- \end{table} -->
+
+<!-- Trends terms included in were observed in ds2, ds3, ect, maf, mdf, mcv,  -->
+<!-- OU term included in model: mcv, lpt, stl -->
+<!-- model_list[["cle"]] <- 3 -->
+<!-- model_list[["ds1"]] <- 3 -->
+<!-- model_list[["ds2"]] <- 1 -->
+<!-- model_list[["ds3"]] <- 1 -->
+<!-- model_list[["ect"]] <- 1 -->
+<!-- model_list[["lps"]] <- 3 -->
+<!-- model_list[["lpt"]] <- 2 -->
+<!-- model_list[["maf"]] <- 1 -->
+<!-- model_list[["mav"]] <- 2 -->
+<!-- model_list[["mcv"]] <- 0 -->
+<!-- model_list[["mdf"]] <- 1 -->
+<!-- #model_list[["mds"]] <- 3 -->
+<!-- #model_list[["mpt"]] <- 0 -->
+<!-- model_list[["pmx"]] <- 3 -->
+<!-- model_list[["stl"]] <- 2 -->
+<!-- #model_list[["tpg"]] <- 1 -->
+<!-- Greg notes:  -->
+<!-- - mdf and maf have interesting trends.   -->
+<!-- - ds3 flips over time.  Females have larger values to start and over the course of the time frame wer are looking at males end up with larger mean values.   -->
+<!-- - Most of the other traits have mean differences that are relatively constant over time.   -->
+
+<!-- In our empirical study, we analyze the following four nested models for the mean process outlined in (\ref{eq:mu}): -->
+<!-- \begin{itemize} -->
+<!-- \item OU with Trend: The mean process for each phenotype $k=1,\cdots,K$ and $g \in \{f,m\}$ as previously defined. -->
+<!-- \item OU with No Trend: No linear trend on the mean processes and the overall mean of the phenotype is constant for each sex.
+This is achieved by setting $\beta_{1,kg} = 0$ in (\ref{eq:mu}) for $k = 1,\cdots,K$ and $g \in \{f,m\}$ -->
+<!-- \item No OU with Trend: No random fluctuations on the mean processes, i.e.
+the OU process assumption is not needed.
+This is achieved by setting $u_{k,gt_i} = 0$ in (\ref{eq:mu}) for $k = 1,\cdots,K$, $g \in \{f,m\}$, and $i = 1,\cdots,T$.
+-->
+<!-- \item No OU with No Trend: The mean for a specific trait and specific sex at each time point is a constant value.
+This is achieved by setting $\beta_{1,kg} = 0$ and $u_{k,gt_i} = 0$ in (\ref{eq:mu}) for $k = 1,\cdots,K$, $g \in \{f,m\}$, and $i = 1,\cdots,T$.
+-->
+<!-- \end{itemize} -->

### 2024-10-21 manuscript/paper.tex
YS. editing, revising, and adding to the manuscript.

@@
 \usepackage{amsmath,amssymb}
+\usepackage{lmodern}
 \usepackage{iftex}
@@
 \else % if luatex or xetex
-  \usepackage{unicode-math} % this also loads fontspec
+  \usepackage{unicode-math}
   \defaultfontfeatures{Scale=MatchLowercase}
   \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
-\fi
-\usepackage{lmodern}
-\ifPDFTeX\else
-  % xetex/luatex font selection
 \fi
@@
 \IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\urlstyle{same}
+\urlstyle{same} % disable monospaced font for URLs
 \hypersetup{
@@
 
-THIS FIGURE DOESN'T SHOW MUCH. LET'S CUT FOR SPACE.
-
-\begin{figure}
-
-{\centering \includegraphics{paper_files/figure-latex/unnamed-chunk-2-1}
-
-}
-
-\caption{Fossil sample size for each time period}\label{fig:unnamed-chunk-2}
-\end{figure}
-
 \begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.3492}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.1905}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4603}}@{}}
-\caption{Traits and trait descriptions. `sc' denotes size correction of
-trait against standard length. Names of bones follow Bowne
-(\protect\hyperlink{ref-Bowne1994}{1994}) unless otherwise
-noted.}\tabularnewline
-\toprule\noalign{}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.1607}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.0893}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.7500}}@{}}
+\toprule()
 \begin{minipage}[b]{\linewidth}\raggedright
@@
 \end{minipage} \\
-\midrule\noalign{}
-\endfirsthead
-\toprule\noalign{}
-\begin{minipage}[b]{\linewidth}\raggedright
-Trait Name
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Trait Code
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Trait Description
-\end{minipage} \\
-\midrule\noalign{}
+\midrule()
 \endhead
-\bottomrule\noalign{}
-\endlastfoot
 Standard Length & \texttt{stl} & Distance from anterior tip of
@@
 premaxilla \\
-Dorsal Spine & \texttt{Ds\#.sc\#\ =\ 1,2,or\ 3} & Length from the base
-of a dorsal spine above the pterygiophore to its distal tip along the
+Dorsal Spine & \texttt{Ds\#.sc\ \#=1/2/3} & Length from the base of a
+dorsal spine above the pterygiophore to its distal tip along the
 anterior edge \\
@@
 spine (when present) \\
+\bottomrule()
 \end{longtable}
+
+\textcolor{red}{This table caption should be part of the table. Top row.. -YS}
+Table: Traits and trait descriptions. `sc' denotes size correction of
+trait against standard length. Names of bones follow Bowne
+(\protect\hyperlink{ref-Bowne1994}{1994}) unless otherwise noted.
 
@@
 
-All models were built using Team
-(\protect\hyperlink{ref-R2022language}{2022}) and (\textbf STAN)
-
 After running these models, we have posterior distributions of all the
@@
   >{\raggedleft\arraybackslash}p{(\columnwidth - 26\tabcolsep) * \real{0.0723}}@{}}
-\toprule\noalign{}
+\toprule()
 \begin{minipage}[b]{\linewidth}\raggedleft
@@
 \end{minipage} \\
-\midrule\noalign{}
+\midrule()
 \endhead
-\bottomrule\noalign{}
-\endlastfoot
 1 & 0.913 & 0.215 & 0.359 & 0.203 & 0.865 & 0.428 & 0.853 & 0.944 &
@@
 0.705 & 0.722 & 0.999 & 0.859 & 0.717 \\
+\bottomrule()
 \end{longtable}
@@
 \end{minipage} \\
-\midrule\noalign{}
+\midrule()
 \endhead
-\bottomrule\noalign{}
-\endlastfoot
 1 & 0.500 & 0.500 & 0.500 & 0.500 & 0.500 & 0.500 & 0.500 & 0.500 &
@@
 0.281 & 0.356 & 0.212 & 0.622 & 0.627 \\
+\bottomrule()
 \end{longtable}
@@
 
-\caption{cle}\label{fig:unnamed-chunk-7}
+\caption{cle}\label{fig:unnamed-chunk-6}
 \end{figure}
@@
 
-\caption{ds1}\label{fig:unnamed-chunk-8}
+\caption{ds1}\label{fig:unnamed-chunk-7}
 \end{figure}
@@
 
-\caption{ds2}\label{fig:unnamed-chunk-9}
+\caption{ds2}\label{fig:unnamed-chunk-8}
 \end{figure}
@@
 
-\caption{ds3}\label{fig:unnamed-chunk-10}
+\caption{ds3}\label{fig:unnamed-chunk-9}
 \end{figure}
@@
 
-\caption{ect}\label{fig:unnamed-chunk-11}
+\caption{ect}\label{fig:unnamed-chunk-10}
 \end{figure}
@@
 
-\caption{lps}\label{fig:unnamed-chunk-12}
+\caption{lps}\label{fig:unnamed-chunk-11}
 \end{figure}
@@
 
-\caption{lpt}\label{fig:unnamed-chunk-13}
+\caption{lpt}\label{fig:unnamed-chunk-12}
 \end{figure}
@@
 
-\caption{maf}\label{fig:unnamed-chunk-14}
+\caption{maf}\label{fig:unnamed-chunk-13}
 \end{figure}
@@
 
-\caption{mav}\label{fig:unnamed-chunk-15}
+\caption{mav}\label{fig:unnamed-chunk-14}
 \end{figure}
@@
 
-\caption{mcv}\label{fig:unnamed-chunk-16}
+\caption{mcv}\label{fig:unnamed-chunk-15}
 \end{figure}
@@
 
-\caption{mdf}\label{fig:unnamed-chunk-17}
+\caption{mdf}\label{fig:unnamed-chunk-16}
 \end{figure}
@@
 
-\caption{mds}\label{fig:unnamed-chunk-18}
+\caption{mds}\label{fig:unnamed-chunk-17}
 \end{figure}
@@
 
-\caption{mpt}\label{fig:unnamed-chunk-19}
+\caption{mpt}\label{fig:unnamed-chunk-18}
 \end{figure}
@@
 
-\caption{pmx}\label{fig:unnamed-chunk-20}
+\caption{pmx}\label{fig:unnamed-chunk-19}
 \end{figure}
@@
 
-\caption{stl}\label{fig:unnamed-chunk-21}
+\caption{stl}\label{fig:unnamed-chunk-20}
 \end{figure}
@@
 \hypertarget{references}{%
-\section*{References}\label{references}}
-\addcontentsline{toc}{section}{References}
+\section{References}\label{references}}
 
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-R2022language}{}}%
-Team, R Core. 2022. \emph{R: A Language and Environment for Statistical
-Computing}. Vienna, Austria: R Foundation for Statistical Computing.
-\url{https://www.R-project.org/}.
-
 \leavevmode\vadjust pre{\hypertarget{ref-OUProcess}{}}%

### 2024-10-21 manuscript/paper.Rmd
manuscript improvements, text, revisions, edits.

@@
 title: |
-  |  \Large Increased sexual dimorphism evolves in a fossil stickleback following ecological release from fish piscivores
+  |  \Large Increased sexual dimorphism does not evolve in a fossil stickleback following ecological release from fish piscivores
 author: | 
@@
 abstract: |
-  | Everyone loves the stickle \vspace{2mm}
-  | *Keywords*: Stickle
+  | Theory suggests that populations should expand their habitat- and resource-use niches when they are freed from interactions with competitors and predators--so-called ecological release.
+One way that this ecological release can manifest is through increased differences between males and females of the same species as they diverge into different niches.
+If divergent niche use drives corresponding divergent adaptation in in the traits used to exploit the divergent niches, then theory predicts an increase in sexual dimorphism for those traits, sometimes called character displacement between the sexes.
+Using a dataset of 16 traits collected from 18 temporal samples spread across 16,500 years of evolution of the fossil stickleback fish, *Gasterosteus doryssus*, we tested the prediction that release from predation inferred in this system would result in character displacement between the sexes.
+We used data from populations of extant stickleback (*Gasterosteus aculeatus*) to build an imputation model to classify individual sex in fossil specimens.
+Then, we quantified the extent of sexual dimorphism at each temporal sample, and tracked how it changed through time.
+NEED TO FINISH THIS   \vspace{2mm}
+  | *Keywords*: character displacement, character release, evolutionary time series, *Gasterosteus doryssus*, missing data imputation
 bibliography: references.bib
@@
 | Trait Name  | Trait Code | Trait Description           |
-|---------|-----|------------------------------------------|
-|Standard Length|	`stl` |	Distance from anterior tip of premaxilla to posterior end of last vertebra (hypural plate)|
-|Dorsal Spine | `mds` |	Number of dorsal spines from 0 to 3|
-|Dorsal Fin Ray|`mdf`|	Number of bones in the dorsal fin posterior to the third dorsal spine (i.e., soft dorsal fin rays)
-|Anal Fin Ray |`maf`|	Number of bones in the anal fin posterior to the anal spine (i.e., soft dorsal fin rays)|
-|Abdominal Vertebra|mav|	Number of vertebrae anterior to the first vertebra contacting an anal fin pterygiophore (Aguirre et al. 2014)|
-|Caudal Vertebra|`mcv`	| Number of vertebrae posterior to and including the first vertebra contacting an anal fin pterygiophore (Aguirre et al. 2014)|
-|Pterygiophore number|`mpt`|	Number of pterygiophores anterior to but excluding the pterygiophore under the third dorsal spine, which is immediately anterior to and contiguous with the dorsal fin|
-|Pelvic Spine length|`lps.sc`	|Length from the base of one pelvic spine above its articulation with the pelvic girdle to its distal tip|
-|Ectocoracoid|`ect.sc`|	Length between the anterior and posterior tips of the shoulder girdle base (i.e., ectocoracoid)|
-|Pelvic Girdle|`tpg.sc`|	Length between the anterior to posterior tips along midline. If vestigial, the sum of longest anterior-posterior axis for the vestiges|
-|Cleithrum length|`cle.sc`|	Length from free dorsal tip to ventral tip of the cleithrum on the anterior margin of the shoulder girdle (i.e., cleithrum)|
-|Premaxilla|`pmx.sc`|	Length from the anterior tip of the premaxilla to the distal tip of the ascending process of the premaxilla |
-|Dorsal Spine|`Ds#.sc #=1/2/3`|	Length from the base of a dorsal spine above the pterygiophore to its distal tip along the anterior edge|
-|Pterygiophore|`lpt.sc`|	Distance between the anterior to posterior tips of the pterygiophore immediately preceding the 3rd dorsal spine (when present)|
+|------------|-----|------------------------------------------|
+|Standard Length|	`stl` |	Distance from anterior tip of premaxilla to posterior end of last vertebra (i.e., the llhypural plate)|
+|Dorsal Spine Count| `mds` |	Number of dorsal spines|
+|Dorsal Fin Ray Count|`mdf`|	Number of bones in the dorsal fin posterior to the third dorsal spine (i.e., soft dorsal fin rays)
+|Anal Fin Ray Count|`maf`|	Number of bones in the anal fin posterior to the anal spine (i.e., soft anal fin rays)|
+|Abdominal Vertebra Count|mav|	Number of vertebrae anterior to the first vertebra contacting an anal fin pterygiophore (Aguirre et al. 2014)|
+|Caudal Vertebra Count|`mcv`	| Number of vertebrae posterior to and including the first vertebra contacting an anal fin pterygiophore (Aguirre et al. 2014)|
+|Pterygiophore Count|`mpt`|	Number of pterygiophores anterior to but excluding the pterygiophore under the third dorsal spine|
+|Pelvic Spine Length|`lps.sc`	|Length from the base of one pelvic spine at its pelvic girdle articulation to its distal tip|
+|Ectocoracoid Length|`ect.sc`|	Length from the anterior to posterior tips of the shoulder girdle base|
+|Pelvic Girdle Length|`tpg.sc`|	Length from the anterior to posterior tips along midline. If vestigial, the sum of longest axis along vestiges|
+|Cleithrum Length|`cle.sc`|	Length from dorsal tip to ventral tip on the anterior margin of the shoulder girdle|
+|Premaxilla Length|`pmx.sc`|	Length from the anterior tip of the premaxilla to the distal tip of its ascending process|
+|Dorsal Spine Length|`Ds#.sc #=1/2/3`|	Length from the dorsal spine base at the pterygiophore to its distal tip|
+|Pterygiophore Length|`lpt.sc`|	Distance from anterior to posterior edges of the pterygiophore immediately before ds3 (when present)|
 
@@
 In 2019, we stained these specimens for bone using Alizarin Red.
+
 We next replicated fossil data collection (Table 1) on these extant specimens.
@@
 -YS}
+
+\textcolor{red}{MATT, somewhere we need to describe how you handled traits with lots of evolved zeros (ds1, ds2, ds3, lps, tpg) -YS}
 
@@
 
-We discuss the procedure for estimating this value below. \textcolor{red}{DID THIS GET DISCUSSED?-YS}
-
-Priors: To aid in the sampling procedure for the discrete phenotypes, we first had to transform the parameters of the generalized Poisson distribution onto the real number line in order to ensure our algorithm can properly sample from the posterior distribution of interest using STAN software.
+\textcolor{red}{Q FOR MATT.
+Is transforming the GP onto the real number line separate from setting the priors?
+If so, should the statement of transformation go in the section where the GP and discrete trait modeling is discussed.
+Then priors get their own section here?
+-YS}
+To aid in the sampling procedure for the discrete phenotypes, we first had to transform the parameters of the generalized Poisson distribution onto the real number line in order to ensure our algorithm can properly sample from the posterior distribution of interest using STAN software.
 This was done by declaring $\phi_k = \log\left(\frac{\alpha_k - \max_{i,j}(-\lambda_{k,ij}/y_{k,ij})}{1 - \alpha_k}\right)$ where $\lambda_{k,ij} = \left\{\begin{array}{cc} \exp(\mu_{k,ft_i} + \gamma_k Y_{K,ij}) & W_{ij} = \text{Female} \\ \exp(\mu_{k,mt_i} + \gamma_k Y_{K,ij}) & W_{ij} = \text{Male} \end{array}\right.$.
@@
 
-We follow the method outlined in @ZhouandReiter to obtain $R$ draws from the posterior distribution of the outlined model parameters given our observed data and imputed values.
-More specifically, for the $j^{th}$ imputed data set where $j = 1,\cdots,M$, we obtain $\frac{R}{M}$ draws from the posterior distribution using an MCMC algorithm with a sufficient burnin period, and then we combine these $M$ MCMC draws to say that we have a total of $R$ draws from the overall posterior distribution of interest, which can then be used to assess the presence of sexual dimorphism for a given trait.
-
-After running these models, we have posterior distributions of all the model parameters of interest.
-From these, we can obtain samples from the posterior distribution of mean trait values for each trait for males and females, using equation 6.
-We use these values for the following analyses.
+We follow the method outlined in @ZhouReiter2010 to obtain $R$ draws from the posterior distributions of the model parameters given our observed data and imputed values.
+More specifically, for the $j^{th}$ imputed data set where $j = 1,\cdots,M$, we obtain $\frac{R}{M}$ draws from the posterior distribution using an MCMC algorithm with a sufficient burnin period, and then we combine these $M$ MCMC draws to say that we have a total of $R$ draws from the overall posterior distribution of interest.
+
+For the continuous phenotypes, the mean trait sizes for each sex $\mu_{k,m,t_i}$ and $\mu_{k,f,t_i}$ are calculated using (\ref{eq:mu}).For the discrete phenotypes, equations (\ref{eq:disc_corr}) and (\ref{eq:disc_ind}) calculate the mean trait size is $\lambda_{k,m,t_i} = \exp(\mu_{k,m,t_i})$ and $\lambda_{k,f,t_i} = \exp(\mu_{k,f,t_i})$ for males and females.
+For simplicity, we will use $\mu_{k,m,t_i}$ and $\mu_{k,f,t_i}$ as the notation for mean size for both discrete and continuous traits for the rest of this discussion.
+From our MCMC algorithm, we obtain $M$ samples from the posterior distribution of $\mu_{k,m,t_i} - \mu_{k,f,t_i}$.
+That is, with posterior distributions of the model parameters described above, we obtained $M$ samples from the posterior distribution of mean trait values by trait for males and females at each time step.
+We then used these values to quantify the presence or absence of sexual dimorphism, and describe how sexual dimorphism evolved through time.
 
 ## Quantifying sexual dimorphism, and testing for increased sexual dimorphism through time.
-We identify sexual dimorphism first by calculating the difference in mean trait size for males and females at each time for phenotype $k$ ($\mu_{k,m,t_i} - \mu_{k,m,t_i}$) for $i = 1,\cdots,T$ where, for the continuous phenotypes, the mean trait sizes for each sex $\mu_{k,m,t_i}$ and $\mu_{k,f,t_i}$ are calculated using (\ref{eq:mu}).
-For the discrete phenotypes, we note that in (\ref{eq:disc_corr}) and (\ref{eq:disc_ind}) the mean trait size is $\lambda_{k,m,t_i} = \exp(\mu_{k,m,t_i})$ and $\lambda_{k,f,t_i} = \exp(\mu_{k,f,t_i})$ for males and females, respectively, which is the value we use in our analysis.
-For simplicity, we will use $\mu_{k,m,t_i}$ and $\mu_{k,f,t_i}$ as the notation for mean size for both discrete and continuous traits for the rest of this discussion.
-From our MCMC algorithm, we obtain $M$ samples from the posterior distribution of $\mu_{k,m,t_i} - \mu_{k,f,t_i}$ to use in our analysis.
-
-Recall that we predicted that sexual dimorphism should increase through time in Lineage II, following ecological release from predators.
-However, this change in dimorphism make be increasing at a non-constant rate, as pointed out by the inclusion of the OU process.
-We assess for dimorphism by calculating a Kendall's $\tau$ coefficient (@KendallsTau).
-Kendall's $\tau$ assesses the ordinal association between a set of bivariate observations, where $\tau > 0$ indicates concordance between the observations (i.e., as $x$ increases, $y$ also tends to increase.)
-
-For our analysis, we calculate $\boldsymbol{\tau}_k = \{\tau_{k,r}: r = 1,\cdots,R\}$ for trait $k$ where $\tau_r$ is the estimate of the Kendall's $\tau$ correlation for $\mu_{k,m,t_i} - \mu_{k,f,t_i}$ versus time $t_i$ for the $r^{th}$ MCMC sample from the posterior distribution.
-The posterior probability of sexual dimorphism for trait $k$ can be estimated as the proportion of values in $\boldsymbol{\tau}_k$ that are greater than 0 ($\frac{1}{R} \sum_{r=1}^R I(\tau_{k,r} > 0)$).
-We can also assess the strength of the dimorphism by calculating the proportion of posterior samples that are greater than a prespecified threshold \textcolor{red}{(0.5??, idk what this could be.) -MS}
+We quantified sexual dimorphism at each time step by calculating the difference in mean trait size for males and females at each time for phenotype $k$, $\mu_{k,m,t_i} - \mu_{k,f,t_i}$, for $i = 1,\cdots,T$.
+The presence of sexual dimorphism was indicated by...
+\textcolor{red}{Matt, can you help me with Bayesian the language here?
+I think this is the place to reference the two post probs plots, yes?
+Values far from fifty fifty indicate that the mass of the posterior distribution is away from a male female difference of zero.
+YS}
+
+Then, to test our prediction that ecological release should result in an increase in sexual dimorphism through time, we calculated Kendall's $\tau$ coefficient for each trait (@KendallsTau).
+Kendall's $\tau$ assesses the ordinal association between a set of bivariate observations, where $\tau > 0$ indicates concordance between the observations (i.e., as $x$ increases, $y$ also tends to increase; here,$x$ is time and $y$ is sexual dimorphism.) Kendall's $\tau$ varies between -1 and 1, and can be thought of as behaving like a correlation.
+Values farther from zero indicate a stronger relationship.
+
+$\boldsymbol{\tau}_k = \{\tau_{k,r}: r = 1,\cdots,R\}$ for trait $k$ where $\tau_r$ is the estimate of the Kendall's $\tau$ correlation for $\mu_{k,m,t_i} - \mu_{k,f,t_i}$ versus time $t_i$ for the $r^{th}$ MCMC sample from the posterior distribution.
+\textcolor{red}{Matt, I think there's an absolute value problem here.
+Dimorphism can increase of M-F gets more positive or more negative.
+I've indicated that in the text below, but I could use some help refininig the explanation -YS}
+For each trait, the proportion of values in $\boldsymbol{\tau}_k$ that are greater than 0 ($\frac{1}{R} \sum_{r=1}^R I(\tau_{k,r} > 0)$) may indicate that males have gotten relatively larger than females, indicating more pronounced dimorphism.
+For each trait, the proportion of values in $\boldsymbol{\tau}_k$ that are less than 0 ($\frac{1}{R} \sum_{r=1}^R I(\tau_{k,r} < 0)$) may indicate that females have gotten relatively larger than males, indicating more pronounced dimormphism.
 
 # Results {#sec:results}
+The extant data used here consist of a total of `r sum(!is.na(stick$gender))` specimens all with known sex (Table S1).
+Of these, there are `r sum(stick$gender == "Female",na.rm = TRUE)` and `r sum(stick$gender == "Male",na.rm = TRUE)` female and male specimens, respectively.
+
+## Missing Data Imputation
 \textcolor{red}{WE NEED RESULTS SECTION DESCRIBING THE RESULTS OF AKHIL'S MULTIPLE IMPUTATION, BOTH THE TESTS OF HOW WELL THE MODEL WORKED CLASSIFYING MODEL SAMPLES, AND SUMMARY STATISTICS OF FOSSIL IMPUTATION RESULTS -YS}
-The extant data used here consists of a total of `r sum(!is.na(stick$gender))` specimens all with known sex (Table S1).
-Of these, there are `r sum(stick$gender == "Female",na.rm = TRUE)` and `r sum(stick$gender == "Male",na.rm = TRUE)` female and male specimens, respectively.
 
@@
 
-Traits such as standard length (stl), anal fin ray (maf), dorsal fin ray (mdf), premaxilla (pmx), and dorsal spine 1 (ds1) demonstrate constant sexual dimorphism with all but the last having learger mean values in male sepcimen.
+## Changes over time
+We can also assess the strength of the dimorphism by calculating the proportion of posterior samples that are greater than a prespecified threshold \textcolor{red}{(0.5??, idk what this could be.) -MS}
+
+Table \ref{tab:auc-table} shows the probability that the posterior distribution of the differences in the mean is greater than the posterior distribution of the differences at the first time point.
+Values near 0.5 indicate very little difference between the distributions whereas values closer to 0 and 1 indicate changes in the distributions with values of 1 meaning the difference in means has shifted towards males having a larger mean and values of 0 indicating the difference in means has shifted towards larger means in females.
+These results are shown graphically in figure \ref{fig:AUC-plot}.
+
+# Discussion, Future work and conclusions {#sec:conclusions}
+We predicted that release from predators would result in niche expansion and increased sexual dimorphism, based on several studies of modern stickleback.
+For example, in lakes where sculpin competitors are absent and stickleback (Roesti et al.
+2023)
+See Spoljaric and Reimchen 2008, page 512 right column for references and discussion of differences between benthic males and limnetic females.
+Male stickleback are benthic and littoral (Wootton 1976)….
+Reimchen papers in general good for this section.
+Moreover, tooth wear data from the lineage II sequence suggest that individuals in this lineage began eating more planktonic prey over time, expanding toward an open-water niche from the benthic, bottom-feeding niche they started with (Purnell et al.
+2007).
+That this expansion into open water by lineage II coincided with armor reduction is further indicative of a limnetic system with fewer salmonid piscivores (@SchluterandMcPhail1992; @VamosiandSchluter2004; @Roestietal2023).
+
+# Acknowledgements {#sec:acknowl}
+
+We thank O.
+Abughoush, S.
+Blain, A.
+Chaudhary, M.
+Islam, F.
+Joaquin, C.
+Lawson-Weinert, R.
+Sullivan, J.
+Tien, M.P.
+Travis, and W.
+Shim for help with data collection.
+We thank D.
+Schluter and S.
+Blain for loaning specimens and sharing data.
+We thank K.
+Swagel and C.
+McMahan of the Field Museum for assistance with specimen x-rays.
+This research was supported by NSF grants BSR-8111013, EAR-9870337, and DEB-0322818, the Center for Field Research (Earthwatch), and the National Geographic Society (2869-84) to MAB.
+It was also supported by NSF grants DEB-1456462 and EAR-2145830 to YES.
+And NSF DMS-2015374 (GJM)
+
+# Data and Code Statement {#sec:datastatement}
+All code for reproducing the analyses in this paper is publicly available at https://github.com/Akhil-Ghosh/SticklebackProject
+\textcolor{red}{Is this true? Matt's code is at this GitHub page? -YS}
+
+# References {#sec:references}
+
+# Supplementary Material {#sec:supplementary}
+<!-- SCRAP CODE and TEXT. YS moved it down here on 21 October 2024 because it was making it hard for me to see the manuscript.  -->
+<!-- THIS FIGURE DOESN'T SHOW MUCH. LET'S CUT FOR SPACE.  -->
+```{r fig.cap = "Fossil sample size for each time period",include=FALSE}
+ggplot(aes(x = time, y = count), data = fortab) + geom_col() + xlab("Sample")  + scale_x_continuous(breaks = 1:18, labels = 1:18)
+```
+<!-- I removed the table and replaced it with a figure -->
+<!-- % latex table generated in R 4.2.2 by xtable 1.8-4 package -->
+<!-- % Mon May  1 11:20:28 2023 -->
+<!-- \begin{table}[ht] -->
+<!-- \centering -->
+<!-- \caption{The number of stickleback fossils at each time point.  Sample size at each time point ranges from a low of 22 to a high of 67.} -->
+<!-- \begin{tabular}{rl} -->
+<!--   \hline -->
+<!-- time & count \\  -->
+<!--   \hline -->
+<!--   1 & 43 \\  -->
+<!--     2 & 41 \\  -->
+<!--     3 & 51 \\  -->
+<!--     4 & 41 \\  -->
+<!--     5 & 46 \\  -->
+<!--     6 & 48 \\  -->
+<!--     7 & 67 \\  -->
+<!--     8 & 55 \\  -->
+<!--     9 & 42 \\  -->
+<!--    10 & 33 \\  -->
+<!--    11 & 37 \\  -->
+<!--    12 & 22 \\  -->
+<!--    13 & 41 \\  -->
+<!--    14 & 43 \\  -->
+<!--    15 & 46 \\  -->
+<!--    16 & 47 \\  -->
+<!--    17 & 56 \\  -->
+<!--    18 & 55 \\  -->
+<!--    \hline -->
+<!-- \end{tabular} -->
+<!-- \label{tabn} -->
+
+<!-- \end{table} -->
+
+<!-- Within each trait, the model with the lowest DIC was chosen for that trait <THIS IS METHODS--WHERE IS THIS DESCRIBED IN THE METHODS>.
+All subsequent results presented here use the model with the lowest DIC individually by trait <THIS IS METHODS--WHERE IS THIS DESCRIBED IN THE METHODS>.
+-->
+
+<!-- Table \ref{whichmodel} reports the results of model fitting.
+X of the Y traits were best fit by a trend model.
+Best models included an OU component for caudal and abdominal vertebra number (mcv, mav), pelvic spine length (lpt <LPT IS PTERYGIOPHORE LENGTH>) and standard length (stl).
+Caudal vertebra number was the only trait best fit by a model including both a trend and an OU component.
+This implies that not only does caudal vertebra size have a linear trend across sex <WHAT DO YOU MEAN BY LINEAR TREND ACROSS SEX?>, but the average size at one time period is highly correlated with the average size at the next time period <I'VE ALWAYS FOUND THAT THIS AID TO MODEL INTEPRETATION IS BETTER IN THE METHODS, JUST AFTER THE MODELS ARE DESCRIBED>.
+-->
+
+<!-- \textcolor{red} { -->
+<!-- Yoel's interpretation of the above paragraph.
+Taking the model with the lowest DIC score as the best model for each trait, we found that cle, ds1, lps, and pmx did not fit either the trend model or the OU model for male-female trait mean differences.
+In contrast, we found a trend through time in male-female trait mean differences (i.e., sexual dimorphism) for mcv, ds2, ds3, ect, maf, and mdf.
+This suggests that these traits changed their dimorphism from one time period to the next, with an overall tendency toward ???GREATER/LESS???
+dimorphism.
+IS THIS CORRECT?
+- YS -->
+<!-- } -->
+<!-- \textcolor{red} { -->
+<!-- We found evidence for an OU process but no trend for lpt, mav, and stl, suggesting that these traits varied through time, about a fixed value for dimorphism.
+Finally, we found evidence that change in mcv fitted both a trend an an OU process, meaning ???
+WHAT BIOLOGICALLY ???.
+-->
+<!-- } -->
+
+<!-- \textcolor{red} {HOW DO WE RECONCILE THE PRESENCE OF TRENDS IN THESE MODELS WITH AN ABSENCE OF VISIBLE TRENDS IN FIGURES 2 THROUGH 4?
+I WONDER IF THIS RESULT PARAGRAPH BELONGS BEST AT THE END OF THE RESULTS, RATHER THAN THE BEGINNING.
+CONSIDER THE STATISTICAL EVIDENCE FOR TRENDS AND OU AFTER WE'VE WALKED THE READERS THROUGH VISUAL EVIDENCE FOR CHANGE.
+-->
+<!-- } -->
+
+<!-- ## Best fitting models  -->
+<!-- \begin{table}[ht] -->
+<!-- \centering -->
+<!-- \begin{tabular}{|c||c|c|} -->
+<!--   \hline -->
+<!--  & OU & No OU  \\  -->
+<!--   \hline -->
+<!--   \hline -->
+<!-- Trend  &  mcv  & ds2, ds3, ect, maf, mdf \\  -->
+<!--   \hline -->
+<!--   No Trend & lpt, mav, stl  &  cle, ds1, lps, pmx\\ -->
+<!--    \hline -->
+<!-- \end{tabular} -->
+<!-- \label{whichmodel} -->
+<!-- \end{table} -->
+
+<!-- Trends terms included in were observed in ds2, ds3, ect, maf, mdf, mcv,  -->
+<!-- OU term included in model: mcv, lpt, stl -->
+<!-- model_list[["cle"]] <- 3 -->
+<!-- model_list[["ds1"]] <- 3 -->
+<!-- model_list[["ds2"]] <- 1 -->
+<!-- model_list[["ds3"]] <- 1 -->
+<!-- model_list[["ect"]] <- 1 -->
+<!-- model_list[["lps"]] <- 3 -->
+<!-- model_list[["lpt"]] <- 2 -->
+<!-- model_list[["maf"]] <- 1 -->
+<!-- model_list[["mav"]] <- 2 -->
+<!-- model_list[["mcv"]] <- 0 -->
+<!-- model_list[["mdf"]] <- 1 -->
+<!-- #model_list[["mds"]] <- 3 -->
+<!-- #model_list[["mpt"]] <- 0 -->
+<!-- model_list[["pmx"]] <- 3 -->
+<!-- model_list[["stl"]] <- 2 -->
+<!-- #model_list[["tpg"]] <- 1 -->
+<!-- Greg notes:  -->
+<!-- - mdf and maf have interesting trends.   -->
+<!-- - ds3 flips over time.  Females have larger values to start and over the course of the time frame wer are looking at males end up with larger mean values.   -->
+<!-- - Most of the other traits have mean differences that are relatively constant over time.   -->
+
+<!-- In our empirical study, we analyze the following four nested models for the mean process outlined in (\ref{eq:mu}): -->
+<!-- \begin{itemize} -->
+<!-- \item OU with Trend: The mean process for each phenotype $k=1,\cdots,K$ and $g \in \{f,m\}$ as previously defined. -->
+<!-- \item OU with No Trend: No linear trend on the mean processes and the overall mean of the phenotype is constant for each sex.
+This is achieved by setting $\beta_{1,kg} = 0$ in (\ref{eq:mu}) for $k = 1,\cdots,K$ and $g \in \{f,m\}$ -->
+<!-- \item No OU with Trend: No random fluctuations on the mean processes, i.e.
+the OU process assumption is not needed.
+This is achieved by setting $u_{k,gt_i} = 0$ in (\ref{eq:mu}) for $k = 1,\cdots,K$, $g \in \{f,m\}$, and $i = 1,\cdots,T$.
+-->
+<!-- \item No OU with No Trend: The mean for a specific trait and specific sex at each time point is a constant value.
+This is achieved by setting $\beta_{1,kg} = 0$ and $u_{k,gt_i} = 0$ in (\ref{eq:mu}) for $k = 1,\cdots,K$, $g \in \{f,m\}$, and $i = 1,\cdots,T$.
+-->
+<!-- \end{itemize} -->
+
+<!-- Traits such as standard length (stl), anal fin ray (maf), dorsal fin ray (mdf), premaxilla (pmx), and dorsal spine 1 (ds1) demonstrate constant sexual dimorphism with all but the last having learger mean values in male sepcimen.
 Other traits such as abdominal vertebra (mav), caudal vertebra (mcv), and pelvic spine length (lps) don't appear to have any discernable sexual dimorphism in the means.
@@
 Over time, however, this switches and the male specimens exhibit larger mean values for the latest observed time periods.
-
-\textcolor{red} {
-Yoel's interpretation of the three paragraphs above.
+-->
+
+<!-- \textcolor{red} { -->
+<!-- Yoel's interpretation of the three paragraphs above.
 Male and female G.
@@
 Table \ref{tab:prob0-table} in the Supplementary Material summarizes the posterior probabilities that male means differ from female means by trait and sample.
-}
-
-\textcolor{red}{can we move mu-diff-plot and prob0-plot here?}
-
-## Changes over time
-Table \ref{tab:auc-table} shows the probability that the posterior distribution of the differences in the mean is greater than the posterior distribution of the differences at the first time point.
-Values near 0.5 indicate very little difference between the distributions whereas values closer to 0 and 1 indicate changes in the distributions with values of 1 meaning the difference in means has shifted towards males having a larger mean and values of 0 indicating the difference in means has shifted towards larger means in females.
-These results are shown graphically in figure \ref{fig:AUC-plot}.
-
-The amount of sexual dimporphism over time does not change substantially for a majority of the trait observed here (i.e.
+-->
+<!-- } -->
+<!-- The amount of sexual dimporphism over time does not change substantially for a majority of the trait observed here (i.e.
 cle, ds1, ds2, ect, lps, lpt, mav, mcv, pmx, stl).
 Traits that trend away from the mean difference at the first observed time point include dorsal spine 3 (ds3), anal fin ray (maf), and dorsal fin ray (mdf).
+-->
 
@@
 <!-- apply(mu_f,2,mean)[18] - apply(mu_f,2,mean)[1] #-1.263581  -->
-The posterior mean of the mean number of dorsal spine rays at the first observed time point is 2.29 for males and for females it is 2.45.
+<!-- The posterior mean of the mean number of dorsal spine rays at the first observed time point is 2.29 for males and for females it is 2.45.
 At the last observed time point, the average number for males and female both dropped to 1.265 and 1.185 for males and females, respectively.
@@
 Basically, males and females both lowered the mean number of dorsal spine 3, but females lowered their mean more than the males.
+-->
 
@@
 
-We also observed a change in the differences between males and females for the number of bones int he anal fin ray (maf) and the number of bones in the dorsal fin ray.
+<!-- We also observed a change in the differences between males and females for the number of bones int he anal fin ray (maf) and the number of bones in the dorsal fin ray.
 The difference in the means of males versus females for anal fin ray bones changed by 0.152 over the course of the observed time periods.
@@
 Whereas for males the change in means rose only from  8.518 to 8.549, about 0.0316 bones on average.
-
-A similar change in the difference of the means over time was observed for the number of bones in the dorsal fin (mdf).
+-->
+
+<!-- A similar change in the difference of the means over time was observed for the number of bones in the dorsal fin (mdf).
 However, unlike maf, this change in sexual dimporhism was driven entirely by the males.
@@
 For males, the average number of dorsal fin bones at the first time point was about 9.568 dropping to 9.734 over the course of the observed time periods for a gain of 0.166 bones on average.
+-->
 
@@
 
-\textcolor{red} {
-Yoel's interpretation of the five paragraphs above. 
-Figure \ref{fig:AUC-plot} shows the posterior probability that the mean male-female difference at time t is different from the male-female difference at time zero.
+<!-- \textcolor{red} { -->
+<!-- Yoel's interpretation of the five paragraphs above.  -->
+<!-- Figure \ref{fig:AUC-plot} shows the posterior probability that the mean male-female difference at time t is different from the male-female difference at time zero.
 In other words, Figure \ref{fig:AUC-plot} plots whether the sexual dimorphism changes in subsequent samples, always relative to the first sample.
@@
 cle, ds1, ds2, ect, lps, lpt, mav, mcv, pmx, stl; Figure \ref{fig:AUC-plot}).
-Traits that change in dimorphism are anal fin ray (maf), and dorsal fin ray (mdf), and dorsal spine 3 (ds3).}
-
-\textcolor{red} {
-For example, females at time zero averaged 8.29 rays in the anal fin, dropping to 8.17 rays in the final sample.
+Traits that change in dimorphism are anal fin ray (maf), and dorsal fin ray (mdf), and dorsal spine 3 (ds3).} -->
+
+<!-- \textcolor{red} { -->
+<!-- For example, females at time zero averaged 8.29 rays in the anal fin, dropping to 8.17 rays in the final sample.
 Males averaged 8.52 anal fin rays in the first sample and 8.55 in the last sample.
@@
 BUT DIMORPHISM SHRUNK THEN, AS THE DIFFERENCE BETWEEN MALES AND FEMALES IS NOW SMALLER??
-}
+-->
+<!-- } -->
 
@@
 
-```{r prob0-table, echo = FALSE, tab.cap= "Posterior probability that the mean for males is larger than the mean for females"}
+```{r prob0-table, echo = FALSE, tab.cap= "Posterior probability that the mean for males is larger than the mean for females",include=FALSE}
 #Which model was used
@@
 
-```{r auc-table, echo = FALSE, tab.cap= "AUC comparing time x versus time 0."}
+```{r auc-table, echo = FALSE, tab.cap= "AUC comparing time x versus time 0.",include=FALSE}
 results_auc_mu_diff_table <- results_auc_mu_diff %>%  mutate(auc_mu_diff = round(auc_mu_diff,3)) %>% pivot_wider(names_from = "phenotype", values_from = "auc_mu_diff")
@@
 
-posterior mean difference for males versus females (values above 0 indicate means that were greater for males versus females)
-
-```{r mu-diff-plot, echo = FALSE, fig.cap = "Posterior mean difference between males and females, in UNITS?. Values above 0 indicate larger male means"}
+<!-- posterior mean difference for males versus females (values above 0 indicate means that were greater for males versus females) -->
+
+```{r mu-diff-plot, echo = FALSE, fig.cap = "Posterior mean difference between males and females, in UNITS?. Values above 0 indicate larger male means",include=FALSE}
 ggplot(aes(x = time, y = mu_diff), data = results_posterior_mean_diff) + facet_wrap(~phenotype, scale = "free_y") + geom_path() + geom_hline(yintercept = 0, lty = 3) + ylab("Posterior Mean Difference")
@@
 
-```{r prob0-plot, echo = FALSE, fig.cap = "Posterior probabilities that the male posterior mean is larger than the female posterior mean"}
+```{r prob0-plot, echo = FALSE, fig.cap = "Posterior probabilities that the male posterior mean is larger than the female posterior mean",include=FALSE}
 ggplot(aes(x = time, y = prob_mu_diff_gt_0), data = results_prob_mu_diff_gt_0) + facet_wrap(~phenotype) + geom_path() + #geom_hline(yintercept = 0, lty = 3) + ylim(0,1) + ylab("Posterior Probability") Why yintercept in 0 and not 0.5?
@@
 Values closer to 0 indicate that the difference in means has shifted towards larger means in females.
-Values closer to 1 indicate that the difference in means has shifted toward larger means in males."}
+Values closer to 1 indicate that the difference in means has shifted toward larger means in males.",include=FALSE}
 ggplot(aes(x = time, y = auc_mu_diff), data = results_auc_mu_diff) + facet_wrap(~phenotype) + geom_path() +  geom_path() + geom_hline(yintercept = 0.5, lty = 3) + ylim(0,1) + ylab("Posterior Probability")
@@
 
-# Discussion, Future work and conclusions {#sec:conclusions}
-We predicted that release from predators would result in niche expansion and increased sexual dimorphism, based on several studies of modern stickleback.
-For example, in lakes where sculpin competitors are absent and stickleback (Roesti et al.
-2023)
-See Spoljaric and Reimchen 2008, page 512 right column for references and discussion of differences between benthic males and limnetic females.
-Male stickleback are benthic and littoral (Wootton 1976)….
-Reimchen papers in general good for this section.
-Moreover, tooth wear data from the lineage II sequence suggest that individuals in this lineage began eating more planktonic prey over time, expanding toward an open-water niche from the benthic, bottom-feeding niche they started with (Purnell et al.
-2007).
-That this expansion into open water by lineage II coincided with armor reduction is further indicative of a limnetic system with fewer salmonid piscivores (@SchluterandMcPhail1992; @VamosiandSchluter2004; @Roestietal2023).
-
-# Acknowledgements {-}
-
-We thank O.
-Abughoush, S.
-Blain, A.
-Chaudhary, M.
-Islam, F.
-Joaquin, C.
-Lawson-Weinert, R.
-Sullivan, J.
-Tien, M.P.
-Travis, and W.
-Shim for help with data collection.
-We thank D.
-Schluter and S.
-Blain for loaning specimens and sharing data.
-We thank K.
-Swagel and C.
-McMahan of the Field Museum for assistance with specimen x-rays.
-This research was supported by NSF grants BSR-8111013, EAR-9870337, and DEB-0322818, the Center for Field Research (Earthwatch), and the National Geographic Society (2869-84) to MAB.
-It was also supported by NSF grants DEB-1456462 and EAR-2145830 to YES.
-And NSF DMS-2015374 (GJM)
-
-# Supplementary Material {-}
-
-All code for reproducing the analyses in this paper is publicly available at https://github.com/Akhil-Ghosh/SticklebackProject
-
-DIC results here.
-
-## Figures
-```{r, out.width="0.9\\linewidth", include=TRUE, fig.align="center", fig.cap=c("cle"), echo=FALSE}
+<!-- # Figures -->
+```{r, out.width="0.9\\linewidth", include=TRUE, fig.align="center", fig.cap=c("cle"), echo=FALSE, include = FALSE}
 knitr::include_graphics(here("Figures/cle","mu_diff.pdf"))
 ```
-```{r, out.width="0.9\\linewidth", include=TRUE, fig.align="center", fig.cap=c("ds1"), echo=FALSE}
+```{r, out.width="0.9\\linewidth", include=TRUE, fig.align="center", fig.cap=c("ds1"), echo=FALSE, include = FALSE}
 knitr::include_graphics(here("Figures/ds1","mu_diff.pdf"))
 ```
-```{r, out.width="0.9\\linewidth", include=TRUE, fig.align="center", fig.cap=c("ds2"), echo=FALSE}
+```{r, out.width="0.9\\linewidth", include=TRUE, fig.align="center", fig.cap=c("ds2"), echo=FALSE, include = FALSE}
 knitr::include_graphics(here("Figures/ds2","mu_diff.pdf"))
 ```
-```{r, out.width="0.9\\linewidth", include=TRUE, fig.align="center", fig.cap=c("ds3"), echo=FALSE}
+```{r, out.width="0.9\\linewidth", include=TRUE, fig.align="center", fig.cap=c("ds3"), echo=FALSE, include = FALSE}
 knitr::include_graphics(here("Figures/ds3","mu_diff.pdf"))
 ```
-```{r, out.width="0.9\\linewidth", include=TRUE, fig.align="center", fig.cap=c("ect"), echo=FALSE}
+```{r, out.width="0.9\\linewidth", include=TRUE, fig.align="center", fig.cap=c("ect"), echo=FALSE, include = FALSE}
 knitr::include_graphics(here("Figures/ect","mu_diff.pdf"))
 ```
-```{r, out.width="0.9\\linewidth", include=TRUE, fig.align="center", fig.cap=c("lps"), echo=FALSE}
+```{r, out.width="0.9\\linewidth", include=TRUE, fig.align="center", fig.cap=c("lps"), echo=FALSE, include = FALSE}
 knitr::include_graphics(here("Figures/lps","mu_diff.pdf"))
 ```
-```{r, out.width="0.9\\linewidth", include=TRUE, fig.align="center", fig.cap=c("lpt"), echo=FALSE}
+```{r, out.width="0.9\\linewidth", include=TRUE, fig.align="center", fig.cap=c("lpt"), echo=FALSE, include = FALSE}
 knitr::include_graphics(here("Figures/lpt","mu_diff.pdf"))
 ```
-```{r, out.width="0.9\\linewidth", include=TRUE, fig.align="center", fig.cap=c("maf"), echo=FALSE}
+```{r, out.width="0.9\\linewidth", include=TRUE, fig.align="center", fig.cap=c("maf"), echo=FALSE, include = FALSE}
 knitr::include_graphics(here("Figures/maf","lambda_diff.pdf"))
 ```
-```{r, out.width="0.9\\linewidth", include=TRUE, fig.align="center", fig.cap=c("mav"), echo=FALSE}
+```{r, out.width="0.9\\linewidth", include=TRUE, fig.align="center", fig.cap=c("mav"), echo=FALSE, include = FALSE}
 knitr::include_graphics(here("Figures/mav","lambda_diff.pdf"))
 ```
-```{r, out.width="0.9\\linewidth", include=TRUE, fig.align="center", fig.cap=c("mcv"), echo=FALSE}
+```{r, out.width="0.9\\linewidth", include=TRUE, fig.align="center", fig.cap=c("mcv"), echo=FALSE, include = FALSE}
 knitr::include_graphics(here("Figures/mcv","lambda_diff.pdf"))
 ```
-```{r, out.width="0.9\\linewidth", include=TRUE, fig.align="center", fig.cap=c("mdf"), echo=FALSE}
+```{r, out.width="0.9\\linewidth", include=TRUE, fig.align="center", fig.cap=c("mdf"), echo=FALSE, include = FALSE}
 knitr::include_graphics(here("Figures/mdf","lambda_diff.pdf"))
 ```
-```{r, out.width="0.9\\linewidth", include=TRUE, fig.align="center", fig.cap=c("mds"), echo=FALSE}
+```{r, out.width="0.9\\linewidth", include=TRUE, fig.align="center", fig.cap=c("mds"), echo=FALSE, include = FALSE}
 knitr::include_graphics(here("Figures/mds","lambda_diff.pdf"))
 ```
-```{r, out.width="0.9\\linewidth", include=TRUE, fig.align="center", fig.cap=c("mpt"), echo=FALSE}
+```{r, out.width="0.9\\linewidth", include=TRUE, fig.align="center", fig.cap=c("mpt"), echo=FALSE, include = FALSE}
 knitr::include_graphics(here("Figures/mpt","lambda_diff.pdf"))
 ```
-```{r, out.width="0.9\\linewidth", include=TRUE, fig.align="center", fig.cap=c("pmx"), echo=FALSE}
+```{r, out.width="0.9\\linewidth", include=TRUE, fig.align="center", fig.cap=c("pmx"), echo=FALSE, include = FALSE}
 knitr::include_graphics(here("Figures/pmx","mu_diff.pdf"))
 ```
-
-```{r, out.width="0.9\\linewidth", include=TRUE, fig.align="center", fig.cap=c("stl"), echo=FALSE}
+```{r, out.width="0.9\\linewidth", include=TRUE, fig.align="center", fig.cap=c("stl"), echo=FALSE, include = FALSE}
 knitr::include_graphics(here("Figures/stl","mu_diff.pdf"))
 ```
-
-# References
-
-<!-- SCRAP CODE and TEXT. YS moved it down here on 21 October 2024 because it was making it hard for me to see the manuscript.  -->
-<!-- THIS FIGURE DOESN'T SHOW MUCH. LET'S CUT FOR SPACE.  -->
-```{r fig.cap = "Fossil sample size for each time period",include=FALSE}
-ggplot(aes(x = time, y = count), data = fortab) + geom_col() + xlab("Sample")  + scale_x_continuous(breaks = 1:18, labels = 1:18)
-```
-<!-- I removed the table and replaced it with a figure -->
-<!-- % latex table generated in R 4.2.2 by xtable 1.8-4 package -->
-<!-- % Mon May  1 11:20:28 2023 -->
-<!-- \begin{table}[ht] -->
-<!-- \centering -->
-<!-- \caption{The number of stickleback fossils at each time point.  Sample size at each time point ranges from a low of 22 to a high of 67.} -->
-<!-- \begin{tabular}{rl} -->
-<!--   \hline -->
-<!-- time & count \\  -->
-<!--   \hline -->
-<!--   1 & 43 \\  -->
-<!--     2 & 41 \\  -->
-<!--     3 & 51 \\  -->
-<!--     4 & 41 \\  -->
-<!--     5 & 46 \\  -->
-<!--     6 & 48 \\  -->
-<!--     7 & 67 \\  -->
-<!--     8 & 55 \\  -->
-<!--     9 & 42 \\  -->
-<!--    10 & 33 \\  -->
-<!--    11 & 37 \\  -->
-<!--    12 & 22 \\  -->
-<!--    13 & 41 \\  -->
-<!--    14 & 43 \\  -->
-<!--    15 & 46 \\  -->
-<!--    16 & 47 \\  -->
-<!--    17 & 56 \\  -->
-<!--    18 & 55 \\  -->
-<!--    \hline -->
-<!-- \end{tabular} -->
-<!-- \label{tabn} -->
-
-<!-- \end{table} -->
-
-<!-- Within each trait, the model with the lowest DIC was chosen for that trait <THIS IS METHODS--WHERE IS THIS DESCRIBED IN THE METHODS>.
-All subsequent results presented here use the model with the lowest DIC individually by trait <THIS IS METHODS--WHERE IS THIS DESCRIBED IN THE METHODS>.
--->
-
-<!-- Table \ref{whichmodel} reports the results of model fitting.
-X of the Y traits were best fit by a trend model.
-Best models included an OU component for caudal and abdominal vertebra number (mcv, mav), pelvic spine length (lpt <LPT IS PTERYGIOPHORE LENGTH>) and standard length (stl).
-Caudal vertebra number was the only trait best fit by a model including both a trend and an OU component.
-This implies that not only does caudal vertebra size have a linear trend across sex <WHAT DO YOU MEAN BY LINEAR TREND ACROSS SEX?>, but the average size at one time period is highly correlated with the average size at the next time period <I'VE ALWAYS FOUND THAT THIS AID TO MODEL INTEPRETATION IS BETTER IN THE METHODS, JUST AFTER THE MODELS ARE DESCRIBED>.
--->
-
-<!-- \textcolor{red} { -->
-<!-- Yoel's interpretation of the above paragraph.
-Taking the model with the lowest DIC score as the best model for each trait, we found that cle, ds1, lps, and pmx did not fit either the trend model or the OU model for male-female trait mean differences.
-In contrast, we found a trend through time in male-female trait mean differences (i.e., sexual dimorphism) for mcv, ds2, ds3, ect, maf, and mdf.
-This suggests that these traits changed their dimorphism from one time period to the next, with an overall tendency toward ???GREATER/LESS???
-dimorphism.
-IS THIS CORRECT?
-- YS -->
-<!-- } -->
-<!-- \textcolor{red} { -->
-<!-- We found evidence for an OU process but no trend for lpt, mav, and stl, suggesting that these traits varied through time, about a fixed value for dimorphism.
-Finally, we found evidence that change in mcv fitted both a trend an an OU process, meaning ???
-WHAT BIOLOGICALLY ???.
--->
-<!-- } -->
-
-<!-- \textcolor{red} {HOW DO WE RECONCILE THE PRESENCE OF TRENDS IN THESE MODELS WITH AN ABSENCE OF VISIBLE TRENDS IN FIGURES 2 THROUGH 4?
-I WONDER IF THIS RESULT PARAGRAPH BELONGS BEST AT THE END OF THE RESULTS, RATHER THAN THE BEGINNING.
-CONSIDER THE STATISTICAL EVIDENCE FOR TRENDS AND OU AFTER WE'VE WALKED THE READERS THROUGH VISUAL EVIDENCE FOR CHANGE.
--->
-<!-- } -->
-
-<!-- ## Best fitting models  -->
-<!-- \begin{table}[ht] -->
-<!-- \centering -->
-<!-- \begin{tabular}{|c||c|c|} -->
-<!--   \hline -->
-<!--  & OU & No OU  \\  -->
-<!--   \hline -->
-<!--   \hline -->
-<!-- Trend  &  mcv  & ds2, ds3, ect, maf, mdf \\  -->
-<!--   \hline -->
-<!--   No Trend & lpt, mav, stl  &  cle, ds1, lps, pmx\\ -->
-<!--    \hline -->
-<!-- \end{tabular} -->
-<!-- \label{whichmodel} -->
-<!-- \end{table} -->
-
-<!-- Trends terms included in were observed in ds2, ds3, ect, maf, mdf, mcv,  -->
-<!-- OU term included in model: mcv, lpt, stl -->
-<!-- model_list[["cle"]] <- 3 -->
-<!-- model_list[["ds1"]] <- 3 -->
-<!-- model_list[["ds2"]] <- 1 -->
-<!-- model_list[["ds3"]] <- 1 -->
-<!-- model_list[["ect"]] <- 1 -->
-<!-- model_list[["lps"]] <- 3 -->
-<!-- model_list[["lpt"]] <- 2 -->
-<!-- model_list[["maf"]] <- 1 -->
-<!-- model_list[["mav"]] <- 2 -->
-<!-- model_list[["mcv"]] <- 0 -->
-<!-- model_list[["mdf"]] <- 1 -->
-<!-- #model_list[["mds"]] <- 3 -->
-<!-- #model_list[["mpt"]] <- 0 -->
-<!-- model_list[["pmx"]] <- 3 -->
-<!-- model_list[["stl"]] <- 2 -->
-<!-- #model_list[["tpg"]] <- 1 -->
-<!-- Greg notes:  -->
-<!-- - mdf and maf have interesting trends.   -->
-<!-- - ds3 flips over time.  Females have larger values to start and over the course of the time frame wer are looking at males end up with larger mean values.   -->
-<!-- - Most of the other traits have mean differences that are relatively constant over time.   -->
-
-<!-- In our empirical study, we analyze the following four nested models for the mean process outlined in (\ref{eq:mu}): -->
-<!-- \begin{itemize} -->
-<!-- \item OU with Trend: The mean process for each phenotype $k=1,\cdots,K$ and $g \in \{f,m\}$ as previously defined. -->
-<!-- \item OU with No Trend: No linear trend on the mean processes and the overall mean of the phenotype is constant for each sex.
-This is achieved by setting $\beta_{1,kg} = 0$ in (\ref{eq:mu}) for $k = 1,\cdots,K$ and $g \in \{f,m\}$ -->
-<!-- \item No OU with Trend: No random fluctuations on the mean processes, i.e.
-the OU process assumption is not needed.
-This is achieved by setting $u_{k,gt_i} = 0$ in (\ref{eq:mu}) for $k = 1,\cdots,K$, $g \in \{f,m\}$, and $i = 1,\cdots,T$.
--->
-<!-- \item No OU with No Trend: The mean for a specific trait and specific sex at each time point is a constant value.
-This is achieved by setting $\beta_{1,kg} = 0$ and $u_{k,gt_i} = 0$ in (\ref{eq:mu}) for $k = 1,\cdots,K$, $g \in \{f,m\}$, and $i = 1,\cdots,T$.
--->
-<!-- \end{itemize} -->

### 2024-10-21 manuscript/paper.tex
manuscript improvements, text, revisions, edits.

@@
 
-\title{~\Large Increased sexual dimorphism evolves in a fossil
+\title{~\Medium Increased sexual dimorphism does not evolve in a fossil
 stickleback following ecological release from fish piscivores}
@@
 \begin{abstract}
-Everyone loves the stickle \vspace{2mm}\\
-\emph{Keywords}: Stickle
+Theory suggests that populations should expand their habitat- and
+resource-use niches when they are freed from interactions with
+competitors and predators--so-called ecological release. One way that
+this ecological release can manifest is through increased differences
+between males and females of the same species as they diverge into
+different niches. If divergent niche use drives corresponding divergent
+adaptation in in the traits used to exploit the divergent niches, then
+theory predicts an increase in sexual dimorphism for those traits,
+sometimes called character displacement between the sexes. Using a
+dataset of 16 traits collected from 18 temporal samples spread across
+16,500 years of evolution of the fossil stickleback fish,
+\emph{Gasterosteus doryssus}, we tested the prediction that release from
+predation inferred in this system would result in character displacement
+between the sexes. We used data from populations of extant stickleback
+(\emph{Gasterosteus aculeatus}) to build an imputation model to classify
+individual sex in fossil specimens. Then, we quantified the extent of
+sexual dimorphism at each temporal sample, and tracked how it changed
+through time. \vspace{2mm}\\
+\emph{Keywords}: character displacement, character release, evolutionary
+time series, \emph{Gasterosteus doryssus}, missing data imputation
 \end{abstract}
@@
 \begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.1607}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.0893}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.7500}}@{}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.2034}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.0847}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.7119}}@{}}
 \toprule()
@@
 Standard Length & \texttt{stl} & Distance from anterior tip of
-premaxilla to posterior end of last vertebra (hypural plate) \\
-Dorsal Spine & \texttt{mds} & Number of dorsal spines from 0 to 3 \\
-Dorsal Fin Ray & \texttt{mdf} & Number of bones in the dorsal fin
+premaxilla to posterior end of last vertebra (i.e., the llhypural
+plate) \\
+Dorsal Spine Count & \texttt{mds} & Number of dorsal spines \\
+Dorsal Fin Ray Count & \texttt{mdf} & Number of bones in the dorsal fin
 posterior to the third dorsal spine (i.e., soft dorsal fin rays) \\
-Anal Fin Ray & \texttt{maf} & Number of bones in the anal fin posterior
-to the anal spine (i.e., soft dorsal fin rays) \\
-Abdominal Vertebra & mav & Number of vertebrae anterior to the first
-vertebra contacting an anal fin pterygiophore (Aguirre et al.~2014) \\
-Caudal Vertebra & \texttt{mcv} & Number of vertebrae posterior to and
-including the first vertebra contacting an anal fin pterygiophore
+Anal Fin Ray Count & \texttt{maf} & Number of bones in the anal fin
+posterior to the anal spine (i.e., soft anal fin rays) \\
+Abdominal Vertebra Count & mav & Number of vertebrae anterior to the
+first vertebra contacting an anal fin pterygiophore (Aguirre et
+al.~2014) \\
+Caudal Vertebra Count & \texttt{mcv} & Number of vertebrae posterior to
+and including the first vertebra contacting an anal fin pterygiophore
 (Aguirre et al.~2014) \\
-Pterygiophore number & \texttt{mpt} & Number of pterygiophores anterior
-to but excluding the pterygiophore under the third dorsal spine, which
-is immediately anterior to and contiguous with the dorsal fin \\
-Pelvic Spine length & \texttt{lps.sc} & Length from the base of one
-pelvic spine above its articulation with the pelvic girdle to its distal
-tip \\
-Ectocoracoid & \texttt{ect.sc} & Length between the anterior and
-posterior tips of the shoulder girdle base (i.e., ectocoracoid) \\
-Pelvic Girdle & \texttt{tpg.sc} & Length between the anterior to
-posterior tips along midline. If vestigial, the sum of longest
-anterior-posterior axis for the vestiges \\
-Cleithrum length & \texttt{cle.sc} & Length from free dorsal tip to
-ventral tip of the cleithrum on the anterior margin of the shoulder
-girdle (i.e., cleithrum) \\
-Premaxilla & \texttt{pmx.sc} & Length from the anterior tip of the
-premaxilla to the distal tip of the ascending process of the
-premaxilla \\
-Dorsal Spine & \texttt{Ds\#.sc\ \#=1/2/3} & Length from the base of a
-dorsal spine above the pterygiophore to its distal tip along the
-anterior edge \\
-Pterygiophore & \texttt{lpt.sc} & Distance between the anterior to
-posterior tips of the pterygiophore immediately preceding the 3rd dorsal
-spine (when present) \\
+Pterygiophore Count & \texttt{mpt} & Number of pterygiophores anterior
+to but excluding the pterygiophore under the third dorsal spine \\
+Pelvic Spine Length & \texttt{lps.sc} & Length from the base of one
+pelvic spine at its pelvic girdle articulation to its distal tip \\
+Ectocoracoid Length & \texttt{ect.sc} & Length from the anterior to
+posterior tips of the shoulder girdle base \\
+Pelvic Girdle Length & \texttt{tpg.sc} & Length from the anterior to
+posterior tips along midline. If vestigial, the sum of longest axis
+along vestiges \\
+Cleithrum Length & \texttt{cle.sc} & Length from dorsal tip to ventral
+tip on the anterior margin of the shoulder girdle \\
+Premaxilla Length & \texttt{pmx.sc} & Length from the anterior tip of
+the premaxilla to the distal tip of its ascending process \\
+Dorsal Spine Length & \texttt{Ds\#.sc\ \#=1/2/3} & Length from the
+dorsal spine base at the pterygiophore to its distal tip \\
+Pterygiophore Length & \texttt{lpt.sc} & Distance from anterior to
+posterior edges of the pterygiophore immediately before ds3 (when
+present) \\
 \bottomrule()
@@
 
-\textcolor{red}{This table caption should be part of the table. Top row.. -YS}
+\textcolor{red}{This table caption should be part of the table. Top row. at the very least, it should be the width of the table. -YS}
 Table: Traits and trait descriptions. `sc' denotes size correction of
@@
 40\% isopropanol. In 2019, we stained these specimens for bone using
-Alizarin Red. We next replicated fossil data collection (Table 1) on
-these extant specimens. Standard length as well as pelvic-spine length
-on each side were measured with calipers. We used a dissection
-microscope to count dorsal spines, pelvic spines, dorsal-fin rays, and
-anal-fin rays. Right and left-side pelvic girdle lengths and
-ectocoracoid lengths were measured from ventral photographs taken using
-a Canon EOS Rebel T7 with a Tamron 16-300 mm MACRO lens mounted on a
-leveled Kaiser RS1 copy stand. Specimens were held in place for ventral
-photographs using a small tabletop vise with an attached scale bar.
-Lateral X-rays were used to measure dorsal spine length, number of
-pterygiophores anterior to the pterygiophore holding the third spine,
-length of the pterygiophore just anterior to the third spine, cleithrum
-length, and pre-maxilla ascending branch length. We also counted
-vertebrae from the X-rays: abdominal vertebrae were counted anterior to
-the first vertebra with a haemal spine contacting an anal fin
-pterygiophore. Caudal vertebrae were posterior, including the first
-vertebra with the haemal spine contacting the anal fin pterygiophore
-(following Aguirre, Walker, and Gideon
+Alizarin Red.
+
+We next replicated fossil data collection (Table 1) on these extant
+specimens. Standard length as well as pelvic-spine length on each side
+were measured with calipers. We used a dissection microscope to count
+dorsal spines, pelvic spines, dorsal-fin rays, and anal-fin rays. Right
+and left-side pelvic girdle lengths and ectocoracoid lengths were
+measured from ventral photographs taken using a Canon EOS Rebel T7 with
+a Tamron 16-300 mm MACRO lens mounted on a leveled Kaiser RS1 copy
+stand. Specimens were held in place for ventral photographs using a
+small tabletop vise with an attached scale bar. Lateral X-rays were used
+to measure dorsal spine length, number of pterygiophores anterior to the
+pterygiophore holding the third spine, length of the pterygiophore just
+anterior to the third spine, cleithrum length, and pre-maxilla ascending
+branch length. We also counted vertebrae from the X-rays: abdominal
+vertebrae were counted anterior to the first vertebra with a haemal
+spine contacting an anal fin pterygiophore. Caudal vertebrae were
+posterior, including the first vertebra with the haemal spine contacting
+the anal fin pterygiophore (following Aguirre, Walker, and Gideon
 (\protect\hyperlink{ref-Aguirreetal2014}{2014})). X-rays were taken with
@@
 (\protect\hyperlink{ref-Schindelinetal2012}{2012})) and its plugin
-ObjectJ (\url{https://sils.fnwi.uva.nl/bcb/objectj/}).
-
-We dissected individuals from the generalist populations (Table S1) to
-determine sex from the gonads. Individuals from the species-pair lakes
-(TableS1) were previously sexed by the Schluter lab, using either
-dissection or a genotyping protocol (Whom, personal communication).
-
-The extant data used here consists of a total of 367 specimens all with
-known sex (Table S1). Of these, there are 202 and 165 female and male
-specimens, respectively.
+ObjectJ (\url{https://sils.fnwi.uva.nl/bcb/objectj/}). We dissected
+individuals from the generalist populations (Table S1) to determine sex
+from the gonads. Individuals from the species-pair lakes (TableS1) were
+previously sexed by the Schluter lab, using either dissection or a
+genotyping protocol \textcolor{red}{(Whom, personal communication)}.
 
@@
 
-Briefly, We used multiple imputation (Little and Rubin
+Briefly, we used multiple imputation (Little and Rubin
 (\protect\hyperlink{ref-little2002statistical}{2002})) to impute the sex
@@
 the \(n_{extant} \times 1\) and \(n_{fossil} \times 1\) vectors of the
-observed extant sex and missing fossil sex, respectively. We imputeD
+observed extant sex and missing fossil sex, respectively. We imputed
 missing sex for the fossil data by sampling from the posterior
@@
 complete datasets.
-
-\textcolor{red}{presumably this also imputes missing data in the other traits? We should probably say that explicitly. -YS}
+\textcolor{red}{PRESUMABLY this also imputes missing data in the other traits? We should probably say that explicitly. -YS}
+
 \textcolor{red}{AKHIL'S METHODS FOR VALIDATION OF THE IMPUTATION MODEL GOES HERE}
-\textcolor{red}{THE POOLING DESCRIBED IN THE NEXT SENTENCE. WAS IT DONE, AND IF SO, HOW WAS IT USED IN DOWNSTREAM ANALYSES. MATT SAYS THAT HE RAN IS MODELS ON EACH IMPUTED DATA SET. -YS}
+
 We then pooled all of the draws from the posterior distribution across
@@
 Rubin's combining rules, following Zhou and Reiter
-(\protect\hyperlink{ref-ZhouReiter2010}{2010}). We used these posterior
-distributions in our Bayesian analysis.
-
-\hypertarget{modeling-trait-means-through-time-for-males-and-females}{%
-\subsection{Modeling trait means through time for males and
-females}\label{modeling-trait-means-through-time-for-males-and-females}}
+(\protect\hyperlink{ref-ZhouReiter2010}{2010}). We then dropped extant
+fish from our imputed data sets and used these posterior distributions
+in our downstream analyses of fossil sexual dimorphism.
+
+\hypertarget{modeling-fossil-stickleback-trait-means-through-time-for-males-and-females}{%
+\subsection{Modeling fossil stickleback trait means through time for
+males and
+females}\label{modeling-fossil-stickleback-trait-means-through-time-for-males-and-females}}
 
 The goal of our analysis was to estimate fossil trait means through
-time, by sex, while incorporating uncertainty in the imputed data,
-especially uncertainty sex classification. We took a Bayesian approach
-because imputation is the process of sampling from a posterior
-predictive distribution of missing data, given observed data.
-\textcolor{red}{THIS NEEDS MORE WORK TO JUSTIFY THE BAYESIAN APPROACH. -YS}
-
-\textcolor{red}{DOES THE NEXT SENTENCE CONTRADICT THE PREVIOUS ONE? BELOW, WE SAY THAT WE USED EACH IMPUTED DATA SET. ABOVE, WE SAY THAT WE POOLED ACROSS IMPUTED DATA SETS.-YS}
-To estimate fossil sample means through time, we first dropped the
-extant individuals from our imputed data sets. Then, for a given imputed
+time, by sex, while incorporating uncertainty across imputations in the
+classification of sex. We used a Bayesian approach to estimate trait
+means in part because imputation is the process of sampling from a
+posterior predictive distribution of missing data, given observed data.
+\textcolor{red}{THIS NEEDS MORE WORK TO JUSTIFY THE BAYESIAN APPROACH, EXPLAINING IT TO MY AUDIENCE. -YS}
+
+\textcolor{red}{MATT, somewhere we need to describe how you handled traits with lots of evolved zeros (ds1, ds2, ds3, lps, tpg) -YS}
+
+To estimate fossil sample means through time, for a given imputed
 dataset, we let \(W_{ij}\) be the imputed sex and
@@
 
-It is reasonable to assume that the other continuous traits will have
-some correlation with standard length \textbf{(CITATION)}. We account
-for this by adding an additional parameter, \(\gamma_k\), If
-\(Y_{k,ij}\) is a continuous trait, then \begin{align}
+Continuous traits will likely have some correlation with standard length
+(i.e., allometry) (Huxley
+(\protect\hyperlink{ref-Huxley1932}{1932}))(Voje, Bell, and Stuart
+(\protect\hyperlink{ref-Vojeetal2022}{2022}))(Voje et al.
+(\protect\hyperlink{ref-Vojeetal2014}{2014})). We account for this by
+adding an additional parameter, \(\gamma_k\), If \(Y_{k,ij}\) is a
+continuous trait, then \begin{align}
 {Y}_{k,ij} & \overset{iid}{\sim}\left\{\begin{array}{llll} \mathcal{N}(\mu_{k,ft_i} + \gamma_kY_{K,ij},\sigma_k^2), & W_{ij} = \text{Female} \\ \mathcal{N}(\mu_{k,mt_i} + \gamma_kY_{K,ij},\sigma_k^2), & W_{ij} = \text{Male} \end{array}\right..
@@
 If \(Y_{k,ij}\) is a discrete trait, the conventional modelling method
-is to fit a Poisson distribution. However, because the empirical fossil
-data violate the assumption that the Poisson variance is equal to its
-mean, we fit discrete traits to a generalized Poisson model as defined
-in
-(\protect\hyperlink{ref-GeneralizedPoisson}{\textbf{GeneralizedPoisson?}}).
+is to fit a Poisson distribution. However, the empirical fossil data
+violate the assumption that the Poisson variance is equal to its mean,
+so we fitted discrete traits to a generalized Poisson model
+((\protect\hyperlink{ref-GeneralizedPoisson}{\textbf{GeneralizedPoisson?}})).
 Specifically, if \(X \sim GP(\lambda,\alpha)\), then \begin{align}
@@
 
-In addition, we allowed that two discrete traits, abdominal vertebrae
-number (mav) and caudal vertebrae number (mcv) could also with standard
-length (\textbf{CITATION}). This is because these are vertebral counts
-that run in parallel with the standard length measurement, along the
-anterior-posterior access of the fish; these vertebrae thus contribute
-to the length of the fish directly, and may evolve in a correlated
-fashion. If \(Y_{k,ij}\) is one of the above traits, then \begin{align}
+In addition, we modeled two discrete traits, abdominal vertebrae number
+(mav) and caudal vertebrae number (mcv) to vary with standard length.
+This is because these are vertebral counts that made along the
+anterior-posterior access of the fish; these vertebrae contribute to the
+length of the fish directly and may evolve in a correlated fashion. If
+\(Y_{k,ij}\) is one of the above traits, then \begin{align}
 {Y}_{k,ij} & \sim \left\{\begin{array}{ll} GP(\exp\{\mu_{k,ft_i} + \gamma_kY_{K,ij}\},\alpha_k), & W_{ij} = \text{Female} \\ GP(\exp\{\mu_{k,mt_i} + \gamma_kY_{K,ij}\},\alpha_k), & W_{ij} = \text{Male} \end{array}\right..
@@
 
-For the other discrete traits, we set \(\gamma_k = 0\) because we expect
-that trait counts are set during early ontogeny, and should not change
-with size in adult fish. Thus, \begin{align}
+For the other discrete counting traits, we set \(\gamma_k = 0\) because
+we expect that trait counts are set during early ontogeny and should not
+change with size in adult fish. Thus, \begin{align}
 {Y}_{k,ij} & \sim \left\{\begin{array}{ll} GP(\exp\{\mu_{k,ft_i}\},\alpha_k), & W_{ij} = \text{Female} \\ GP(\exp\{\mu_{k,mt_i}\},\alpha_k), & W_{ij} = \text{Male} \end{array}\right..
 \label{eq:disc_ind}
-\end{align} We point out that, for the discrete traits, the means are
-represented by \(\exp\{\mu_{k,ft_i}\}\) and \(\exp\{\mu_{k,mt_i}\}\) for
-ease of modeling.
+\end{align} We point out that counting trait means are represented by
+\(\exp\{\mu_{k,ft_i}\}\) and \(\exp\{\mu_{k,mt_i}\}\) for ease of
+modeling.
 
@@
 model the time-\(t_i\), sample-specific trait mean \(k\) for females and
-males, respectively. To account for the possibility that there is a time
--dependent trend in \(\mu_{k,ft_i}\) and \(\mu_{k,mt_i}\), we further
-set \begin{align}
+males, respectively. To account for the possibility that there is a
+time-dependent (i.e., evolutionary trend) trend in \(\mu_{k,ft_i}\) and
+\(\mu_{k,mt_i}\), we further set \begin{align}
 \mu_{k,gt_i} = \beta_{0,kg} + \beta_{1,kg}t_i + u_{k,gt_i},
@@
 incorporates prior information and uncertainty from missing data
-imputation in the estimate of the variable.
-
-We discuss the procedure for estimating this value below.
-\textcolor{red}{DID THIS GET DISCUSSED?-YS}
-
-Priors: To aid in the sampling procedure for the discrete phenotypes, we
-first had to transform the parameters of the generalized Poisson
-distribution onto the real number line in order to ensure our algorithm
-can properly sample from the posterior distribution of interest using
-STAN software. This was done by declaring
+imputation in the estimate of latent variables.
+
+\textcolor{red}{Q FOR MATT.
+Is transforming the GP onto the real number line separate from setting the priors?
+If so, should the statement of transformation go in the section where the GP and discrete trait modeling is discussed.
+Then priors get their own section here?
+-YS}
+To aid in the sampling procedure for the discrete phenotypes, we first
+had to transform the parameters of the generalized Poisson distribution
+onto the real number line in order to ensure our algorithm can properly
+sample from the posterior distribution of interest using STAN software.
+This was done by declaring
 \(\phi_k = \log\left(\frac{\alpha_k - \max_{i,j}(-\lambda_{k,ij}/y_{k,ij})}{1 - \alpha_k}\right)\)
@@
 
-We follow the method outlined in
-(\protect\hyperlink{ref-ZhouandReiter}{\textbf{ZhouandReiter?}}) to
-obtain \(R\) draws from the posterior distribution of the outlined model
-parameters given our observed data and imputed values. More
-specifically, for the \(j^{th}\) imputed data set where
-\(j = 1,\cdots,M\), we obtain \(\frac{R}{M}\) draws from the posterior
-distribution using an MCMC algorithm with a sufficient burnin period,
-and then we combine these \(M\) MCMC draws to say that we have a total
-of \(R\) draws from the overall posterior distribution of interest,
-which can then be used to assess the presence of sexual dimorphism for a
-given trait.
-
-After running these models, we have posterior distributions of all the
-model parameters of interest. From these, we can obtain samples from the
-posterior distribution of mean trait values for each trait for males and
-females, using equation 6. We use these values for the following
-analyses.
+We follow the method outlined in Zhou and Reiter
+(\protect\hyperlink{ref-ZhouReiter2010}{2010}) to obtain \(R\) draws
+from the posterior distributions of the model parameters given our
+observed data and imputed values. More specifically, for the \(j^{th}\)
+imputed data set where \(j = 1,\cdots,M\), we obtain \(\frac{R}{M}\)
+draws from the posterior distribution using an MCMC algorithm with a
+sufficient burnin period, and then we combine these \(M\) MCMC draws to
+say that we have a total of \(R\) draws from the overall posterior
+distribution of interest.
+
+For the continuous phenotypes, the mean trait sizes for each sex
+\(\mu_{k,m,t_i}\) and \(\mu_{k,f,t_i}\) are calculated using
+(\ref{eq:mu}).For the discrete phenotypes, equations
+(\ref{eq:disc_corr}) and (\ref{eq:disc_ind}) calculate the mean trait
+size is \(\lambda_{k,m,t_i} = \exp(\mu_{k,m,t_i})\) and
+\(\lambda_{k,f,t_i} = \exp(\mu_{k,f,t_i})\) for males and females. For
+simplicity, we will use \(\mu_{k,m,t_i}\) and \(\mu_{k,f,t_i}\) as the
+notation for mean size for both discrete and continuous traits for the
+rest of this discussion. From our MCMC algorithm, we obtain \(M\)
+samples from the posterior distribution of
+\(\mu_{k,m,t_i} - \mu_{k,f,t_i}\). That is, with posterior distributions
+of the model parameters described above, we obtained \(M\) samples from
+the posterior distribution of mean trait values by trait for males and
+females at each time step. We then used these values to quantify the
+presence or absence of sexual dimorphism, and describe how sexual
+dimorphism evolved through time.
 
@@
 
-We identify sexual dimorphism first by calculating the difference in
-mean trait size for males and females at each time for phenotype \(k\)
-(\(\mu_{k,m,t_i} - \mu_{k,m,t_i}\)) for \(i = 1,\cdots,T\) where, for
-the continuous phenotypes, the mean trait sizes for each sex
-\(\mu_{k,m,t_i}\) and \(\mu_{k,f,t_i}\) are calculated using
-(\ref{eq:mu}). For the discrete phenotypes, we note that in
-(\ref{eq:disc_corr}) and (\ref{eq:disc_ind}) the mean trait size is
-\(\lambda_{k,m,t_i} = \exp(\mu_{k,m,t_i})\) and
-\(\lambda_{k,f,t_i} = \exp(\mu_{k,f,t_i})\) for males and females,
-respectively, which is the value we use in our analysis. For simplicity,
-we will use \(\mu_{k,m,t_i}\) and \(\mu_{k,f,t_i}\) as the notation for
-mean size for both discrete and continuous traits for the rest of this
-discussion. From our MCMC algorithm, we obtain \(M\) samples from the
-posterior distribution of \(\mu_{k,m,t_i} - \mu_{k,f,t_i}\) to use in
-our analysis.
-
-Recall that we predicted that sexual dimorphism should increase through
-time in Lineage II, following ecological release from predators.
-However, this change in dimorphism make be increasing at a non-constant
-rate, as pointed out by the inclusion of the OU process. We assess for
-dimorphism by calculating a Kendall's \(\tau\) coefficient (KENDALL
+We quantified sexual dimorphism at each time step by calculating the
+difference in mean trait size for males and females at each time for
+phenotype \(k\), \(\mu_{k,m,t_i} - \mu_{k,f,t_i}\), for
+\(i = 1,\cdots,T\). The presence of sexual dimorphism was indicated
+by\ldots{}
+\textcolor{red}{Matt, can you help me with Bayesian the language here?
+I think this is the place to reference the two post probs plots, yes?
+Values far from fifty fifty indicate that the mass of the posterior distribution is away from a male female difference of zero.
+YS}
+
+Then, to test our prediction that ecological release should result in an
+increase in sexual dimorphism through time, we calculated Kendall's
+\(\tau\) coefficient for each trait (Kendal
 (\protect\hyperlink{ref-KendallsTau}{1938})). Kendall's \(\tau\)
@@
 observations, where \(\tau > 0\) indicates concordance between the
-observations (i.e., as \(x\) increases, \(y\) also tends to increase.)
-
-For our analysis, we calculate
+observations (i.e., as \(x\) increases, \(y\) also tends to increase;
+here,\(x\) is time and \(y\) is sexual dimorphism.) Kendall's \(\tau\)
+varies between -1 and 1, and can be thought of as behaving like a
+correlation. Values farther from zero indicate a stronger relationship.
+
 \(\boldsymbol{\tau}_k = \{\tau_{k,r}: r = 1,\cdots,R\}\) for trait \(k\)
@@
 for \(\mu_{k,m,t_i} - \mu_{k,f,t_i}\) versus time \(t_i\) for the
-\(r^{th}\) MCMC sample from the posterior distribution. The posterior
-probability of sexual dimorphism for trait \(k\) can be estimated as the
-proportion of values in \(\boldsymbol{\tau}_k\) that are greater than 0
-(\(\frac{1}{R} \sum_{r=1}^R I(\tau_{k,r} > 0)\)). We can also assess the
-strength of the dimorphism by calculating the proportion of posterior
-samples that are greater than a prespecified threshold
-\textcolor{red}{(0.5??, idk what this could be.) -MS}
+\(r^{th}\) MCMC sample from the posterior distribution.
+\textcolor{red}{Matt, I think there's an absolute value problem here.
+Dimorphism can increase of M-F gets more positive or more negative.
+I've indicated that in the text below, but I could use some help refininig the explanation -YS}
+For each trait, the proportion of values in \(\boldsymbol{\tau}_k\) that
+are greater than 0 (\(\frac{1}{R} \sum_{r=1}^R I(\tau_{k,r} > 0)\)) may
+indicate that males have gotten relatively larger than females,
+indicating more pronounced dimorphism. For each trait, the proportion of
+values in \(\boldsymbol{\tau}_k\) that are less than 0
+(\(\frac{1}{R} \sum_{r=1}^R I(\tau_{k,r} < 0)\)) may indicate that
+females have gotten relatively larger than males, indicating more
+pronounced dimormphism.
 
@@
 \section{Results}\label{sec:results}}
+
+The extant data used here consist of a total of 367 specimens all with
+known sex (Table S1). Of these, there are 202 and 165 female and male
+specimens, respectively.
+
+\hypertarget{missing-data-imputation}{%
+\subsection{Missing Data Imputation}\label{missing-data-imputation}}
 
@@
 
-Traits such as standard length (stl), anal fin ray (maf), dorsal fin ray
-(mdf), premaxilla (pmx), and dorsal spine 1 (ds1) demonstrate constant
-sexual dimorphism with all but the last having learger mean values in
-male sepcimen. Other traits such as abdominal vertebra (mav), caudal
-vertebra (mcv), and pelvic spine length (lps) don't appear to have any
-discernable sexual dimorphism in the means. Dorsal spine 3 (ds3)
-exhibits sexual dimorphism that changes over time. The mean number of
-dorsal spine 3 begins larger for females for the earliest observed time
-periods. Over time, however, this switches and the male specimens
-exhibit larger mean values for the latest observed time periods.
-
-\textcolor{red} {
-Yoel's interpretation of the three paragraphs above.
-Male and female G.
-doryssus differ for some traits, indicating detectable sexual dimorphism.
-This is evidenced by traits with substantial differences in male and female posterior means (i.e., deviation from 0 in Figure \ref{fig:mu-diff-plot}), supported by posterior probabilities that deviate from 0.5 (Figure \ref{fig:prob0-plot}).
-For example, males are larger than females by 3-8mm on average through time (stl panel in Figure \ref{fig:mu-diff-plot}), with a posterior probability for that mean difference above 70 percent through time (Figure \ref{fig:prob0-plot}).
-In contrast, the number of abdominal vertebrae (mav) fluctuates about 0 (Figure \ref{fig:mu-diff-plot}), with posterior probabilities fluctuating about 50 percent through time (Figure \ref{fig:prob0-plot}).
-Table \ref{tab:prob0-table} in the Supplementary Material summarizes the posterior probabilities that male means differ from female means by trait and sample.
-}
-
-\textcolor{red}{can we move mu-diff-plot and prob0-plot here?}
-
 \hypertarget{changes-over-time}{%
 \subsection{Changes over time}\label{changes-over-time}}
+
+We can also assess the strength of the dimorphism by calculating the
+proportion of posterior samples that are greater than a prespecified
+threshold \textcolor{red}{(0.5??, idk what this could be.) -MS}
 
@@
 These results are shown graphically in figure \ref{fig:AUC-plot}.
-
-The amount of sexual dimporphism over time does not change substantially
-for a majority of the trait observed here (i.e.~cle, ds1, ds2, ect, lps,
-lpt, mav, mcv, pmx, stl). Traits that trend away from the mean
-difference at the first observed time point include dorsal spine 3
-(ds3), anal fin ray (maf), and dorsal fin ray (mdf).
-
-The posterior mean of the mean number of dorsal spine rays at the first
-observed time point is 2.29 for males and for females it is 2.45. At the
-last observed time point, the average number for males and female both
-dropped to 1.265 and 1.185 for males and females, respectively. This
-resulted in a mean number of dorsal spine 3 lost by males of 1.027 and
-for females it was 1.264. This resulted in a change of the mean
-difference from time point 1 to time point 18 of 0.236. Basically, males
-and females both lowered the mean number of dorsal spine 3, but females
-lowered their mean more than the males.
-
-We also observed a change in the differences between males and females
-for the number of bones int he anal fin ray (maf) and the number of
-bones in the dorsal fin ray. The difference in the means of males versus
-females for anal fin ray bones changed by 0.152 over the course of the
-observed time periods. This change was driven mostly by the change in
-females. At the first observed time period the had on average 8.291
-bones dropping to 8.170 over the 18 times periods observed. Whereas for
-males the change in means rose only from 8.518 to 8.549, about 0.0316
-bones on average.
-
-A similar change in the difference of the means over time was observed
-for the number of bones in the dorsal fin (mdf). However, unlike maf,
-this change in sexual dimporhism was driven entirely by the males. The
-mean change for females over the 18 observed time periods was only
-0.0055 with the average number of dorsal fin bones for females starting
-at 9.287 and the final observed mean of 9.281 resulting in almost no
-change in the mean. For males, the average number of dorsal fin bones at
-the first time point was about 9.568 dropping to 9.734 over the course
-of the observed time periods for a gain of 0.166 bones on average.
-
-\textcolor{red} {
-Yoel's interpretation of the five paragraphs above. 
-Figure \ref{fig:AUC-plot} shows the posterior probability that the mean male-female difference at time t is different from the male-female difference at time zero.
-In other words, Figure \ref{fig:AUC-plot} plots whether the sexual dimorphism changes in subsequent samples, always relative to the first sample.
-The amount of sexual dimporphism over time does not change substantially for a majority of the traits (i.e.
-cle, ds1, ds2, ect, lps, lpt, mav, mcv, pmx, stl; Figure \ref{fig:AUC-plot}).
-Traits that change in dimorphism are anal fin ray (maf), and dorsal fin ray (mdf), and dorsal spine 3 (ds3).}
-
-\textcolor{red} {
-For example, females at time zero averaged 8.29 rays in the anal fin, dropping to 8.17 rays in the final sample.
-Males averaged 8.52 anal fin rays in the first sample and 8.55 in the last sample.
-Thus, the difference between males and females started at 0.23 and ended at 0.38, an increase in sexual dimorphism.
-In contrast, the sign on dimorphism flipped for ds3.
-The posterior mean length of the third dorsal spine is 2.29mm for males and 2.45mm for females.
-By the end of the sequence, mean ds3 for males was 1.27 for males and 1.19 for females.
-Both sexes evolved reduction of ds3, but females lowered their mean more than the males.
-This resulted in a flip in dimorphism, from larger females to larger males, though both sexes evolved trait reduction.
-BUT DIMORPHISM SHRUNK THEN, AS THE DIFFERENCE BETWEEN MALES AND FEMALES IS NOW SMALLER??
-}
-
-\begin{longtable}[]{@{}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 26\tabcolsep) * \real{0.0602}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 26\tabcolsep) * \real{0.0723}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 26\tabcolsep) * \real{0.0723}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 26\tabcolsep) * \real{0.0723}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 26\tabcolsep) * \real{0.0723}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 26\tabcolsep) * \real{0.0723}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 26\tabcolsep) * \real{0.0723}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 26\tabcolsep) * \real{0.0723}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 26\tabcolsep) * \real{0.0723}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 26\tabcolsep) * \real{0.0723}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 26\tabcolsep) * \real{0.0723}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 26\tabcolsep) * \real{0.0723}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 26\tabcolsep) * \real{0.0723}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 26\tabcolsep) * \real{0.0723}}@{}}
-\toprule()
-\begin{minipage}[b]{\linewidth}\raggedleft
-time
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-cle
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-ds1
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-ds2
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-ds3
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-ect
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-lps
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-lpt
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-maf
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-mav
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-mcv
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-mdf
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-pmx
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-stl
-\end{minipage} \\
-\midrule()
-\endhead
-1 & 0.913 & 0.215 & 0.359 & 0.203 & 0.865 & 0.428 & 0.853 & 0.944 &
-0.391 & 0.507 & 0.959 & 0.948 & 0.883 \\
-2 & 0.942 & 0.212 & 0.325 & 0.180 & 0.891 & 0.404 & 0.853 & 0.948 &
-0.520 & 0.520 & 0.963 & 0.967 & 0.908 \\
-3 & 0.843 & 0.207 & 0.249 & 0.153 & 0.770 & 0.279 & 0.803 & 0.965 &
-0.289 & 0.525 & 0.977 & 0.912 & 0.777 \\
-4 & 0.890 & 0.210 & 0.249 & 0.201 & 0.830 & 0.372 & 0.741 & 0.977 &
-0.435 & 0.589 & 0.986 & 0.935 & 0.831 \\
-5 & 0.954 & 0.211 & 0.211 & 0.195 & 0.909 & 0.377 & 0.756 & 0.985 &
-0.533 & 0.622 & 0.992 & 0.975 & 0.921 \\
-6 & 0.885 & 0.211 & 0.134 & 0.171 & 0.820 & 0.291 & 0.686 & 0.991 &
-0.645 & 0.456 & 0.996 & 0.949 & 0.811 \\
-7 & 0.812 & 0.211 & 0.124 & 0.239 & 0.760 & 0.340 & 0.617 & 0.994 &
-0.224 & 0.717 & 0.999 & 0.879 & 0.747 \\
-8 & 0.928 & 0.210 & 0.135 & 0.258 & 0.885 & 0.387 & 0.706 & 0.994 &
-0.342 & 0.730 & 1.000 & 0.969 & 0.872 \\
-9 & 0.899 & 0.213 & 0.145 & 0.304 & 0.857 & 0.406 & 0.715 & 0.995 &
-0.521 & 0.756 & 1.000 & 0.940 & 0.845 \\
-10 & 0.954 & 0.217 & 0.235 & 0.555 & 0.940 & 0.571 & 0.810 & 0.997 &
-0.454 & 0.719 & 1.000 & 0.976 & 0.935 \\
-11 & 0.913 & 0.211 & 0.162 & 0.476 & 0.875 & 0.388 & 0.703 & 0.997 &
-0.363 & 0.657 & 1.000 & 0.957 & 0.864 \\
-12 & 0.949 & 0.221 & 0.262 & 0.726 & 0.929 & 0.585 & 0.803 & 0.997 &
-0.462 & 0.595 & 1.000 & 0.968 & 0.926 \\
-13 & 0.904 & 0.220 & 0.260 & 0.789 & 0.885 & 0.598 & 0.795 & 0.998 &
-0.470 & 0.370 & 1.000 & 0.929 & 0.872 \\
-14 & 0.925 & 0.220 & 0.280 & 0.883 & 0.906 & 0.659 & 0.854 & 0.997 &
-0.526 & 0.472 & 1.000 & 0.958 & 0.898 \\
-15 & 0.880 & 0.225 & 0.283 & 0.839 & 0.849 & 0.617 & 0.796 & 0.997 &
-0.490 & 0.398 & 1.000 & 0.919 & 0.834 \\
-16 & 0.953 & 0.216 & 0.276 & 0.931 & 0.931 & 0.614 & 0.831 & 0.997 &
-0.414 & 0.541 & 1.000 & 0.975 & 0.925 \\
-17 & 0.803 & 0.219 & 0.232 & 0.832 & 0.765 & 0.383 & 0.655 & 0.996 &
-0.569 & 0.758 & 1.000 & 0.854 & 0.751 \\
-18 & 0.780 & 0.212 & 0.225 & 0.817 & 0.726 & 0.297 & 0.558 & 0.995 &
-0.705 & 0.722 & 0.999 & 0.859 & 0.717 \\
-\bottomrule()
-\end{longtable}
-
-\begin{longtable}[]{@{}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 26\tabcolsep) * \real{0.0602}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 26\tabcolsep) * \real{0.0723}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 26\tabcolsep) * \real{0.0723}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 26\tabcolsep) * \real{0.0723}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 26\tabcolsep) * \real{0.0723}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 26\tabcolsep) * \real{0.0723}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 26\tabcolsep) * \real{0.0723}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 26\tabcolsep) * \real{0.0723}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 26\tabcolsep) * \real{0.0723}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 26\tabcolsep) * \real{0.0723}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 26\tabcolsep) * \real{0.0723}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 26\tabcolsep) * \real{0.0723}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 26\tabcolsep) * \real{0.0723}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 26\tabcolsep) * \real{0.0723}}@{}}
-\toprule()
-\begin{minipage}[b]{\linewidth}\raggedleft
-time
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-cle
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-ds1
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-ds2
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-ds3
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-ect
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-lps
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-lpt
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-maf
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-mav
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-mcv
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-mdf
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-pmx
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-stl
-\end{minipage} \\
-\midrule()
-\endhead
-1 & 0.500 & 0.500 & 0.500 & 0.500 & 0.500 & 0.500 & 0.500 & 0.500 &
-0.500 & 0.500 & 0.500 & 0.500 & 0.500 \\
-2 & 0.514 & 0.504 & 0.509 & 0.495 & 0.512 & 0.508 & 0.517 & 0.495 &
-0.396 & 0.490 & 0.495 & 0.511 & 0.513 \\
-3 & 0.636 & 0.514 & 0.559 & 0.516 & 0.628 & 0.601 & 0.602 & 0.475 &
-0.523 & 0.480 & 0.476 & 0.626 & 0.641 \\
-4 & 0.563 & 0.509 & 0.544 & 0.459 & 0.558 & 0.539 & 0.672 & 0.456 &
-0.436 & 0.450 & 0.458 & 0.556 & 0.562 \\
-5 & 0.543 & 0.507 & 0.551 & 0.417 & 0.539 & 0.522 & 0.687 & 0.436 &
-0.389 & 0.425 & 0.439 & 0.537 & 0.544 \\
-6 & 0.618 & 0.512 & 0.589 & 0.415 & 0.613 & 0.579 & 0.716 & 0.415 &
-0.317 & 0.522 & 0.419 & 0.612 & 0.620 \\
-7 & 0.591 & 0.510 & 0.601 & 0.364 & 0.589 & 0.564 & 0.712 & 0.379 &
-0.546 & 0.382 & 0.383 & 0.588 & 0.592 \\
-8 & 0.539 & 0.507 & 0.579 & 0.324 & 0.536 & 0.514 & 0.683 & 0.373 &
-0.478 & 0.368 & 0.378 & 0.537 & 0.540 \\
-9 & 0.524 & 0.506 & 0.586 & 0.299 & 0.521 & 0.510 & 0.636 & 0.355 &
-0.408 & 0.340 & 0.359 & 0.523 & 0.522 \\
-10 & 0.379 & 0.497 & 0.544 & 0.203 & 0.377 & 0.400 & 0.578 & 0.321 &
-0.438 & 0.359 & 0.324 & 0.385 & 0.373 \\
-11 & 0.531 & 0.508 & 0.605 & 0.226 & 0.528 & 0.517 & 0.685 & 0.301 &
-0.528 & 0.403 & 0.303 & 0.530 & 0.533 \\
-12 & 0.363 & 0.492 & 0.547 & 0.150 & 0.360 & 0.386 & 0.566 & 0.284 &
-0.439 & 0.436 & 0.284 & 0.369 & 0.358 \\
-13 & 0.373 & 0.495 & 0.555 & 0.133 & 0.372 & 0.386 & 0.561 & 0.269 &
-0.432 & 0.575 & 0.268 & 0.378 & 0.371 \\
-14 & 0.326 & 0.496 & 0.548 & 0.102 & 0.326 & 0.349 & 0.514 & 0.255 &
-0.399 & 0.514 & 0.253 & 0.334 & 0.323 \\
-15 & 0.365 & 0.492 & 0.552 & 0.107 & 0.365 & 0.373 & 0.502 & 0.250 &
-0.433 & 0.554 & 0.247 & 0.368 & 0.367 \\
-16 & 0.361 & 0.497 & 0.574 & 0.084 & 0.360 & 0.381 & 0.570 & 0.234 &
-0.462 & 0.463 & 0.229 & 0.369 & 0.356 \\
-17 & 0.561 & 0.504 & 0.629 & 0.116 & 0.559 & 0.542 & 0.669 & 0.226 &
-0.364 & 0.313 & 0.219 & 0.559 & 0.565 \\
-18 & 0.625 & 0.512 & 0.648 & 0.121 & 0.617 & 0.596 & 0.735 & 0.220 &
-0.281 & 0.356 & 0.212 & 0.622 & 0.627 \\
-\bottomrule()
-\end{longtable}
-
-posterior mean difference for males versus females (values above 0
-indicate means that were greater for males versus females)
-
-\begin{figure}
-
-{\centering \includegraphics{paper_files/figure-latex/mu-diff-plot-1}
-
-}
-
-\caption{Posterior mean difference between males and females, in UNITS?. Values above 0 indicate larger male means}\label{fig:mu-diff-plot}
-\end{figure}
-
-\begin{figure}
-
-{\centering \includegraphics{paper_files/figure-latex/prob0-plot-1}
-
-}
-
-\caption{Posterior probabilities that the male posterior mean is larger than the female posterior mean}\label{fig:prob0-plot}
-\end{figure}
-
-\begin{figure}
-
-{\centering \includegraphics{paper_files/figure-latex/AUC-plot-1}
-
-}
-
-\caption{The Posterior Probability that male-female differences in subsequent samples are different from male-female differenceds in the initial sample.
-Values near 0.5 indicate little difference between the distributions.
-Values closer to 0 indicate that the difference in means has shifted towards larger means in females.
-Values closer to 1 indicate that the difference in means has shifted toward larger means in males.}\label{fig:AUC-plot}
-\end{figure}
 
@@
 
-\hypertarget{acknowledgements}{%
-\section*{Acknowledgements}\label{acknowledgements}}
-\addcontentsline{toc}{section}{Acknowledgements}
+\hypertarget{sec:acknowl}{%
+\section{Acknowledgements}\label{sec:acknowl}}
 
@@
 
-\hypertarget{supplementary-material}{%
-\section*{Supplementary Material}\label{supplementary-material}}
-\addcontentsline{toc}{section}{Supplementary Material}
+\hypertarget{sec:datastatement}{%
+\section{Data and Code Statement}\label{sec:datastatement}}
 
@@
 available at \url{https://github.com/Akhil-Ghosh/SticklebackProject}
-
-DIC results here.
-
-\hypertarget{figures}{%
-\subsection{Figures}\label{figures}}
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.9\linewidth]{../Figures/cle/mu_diff}
-
-}
-
-\caption{cle}\label{fig:unnamed-chunk-6}
-\end{figure}
-\begin{figure}
-
-{\centering \includegraphics[width=0.9\linewidth]{../Figures/ds1/mu_diff}
-
-}
-
-\caption{ds1}\label{fig:unnamed-chunk-7}
-\end{figure}
-\begin{figure}
-
-{\centering \includegraphics[width=0.9\linewidth]{../Figures/ds2/mu_diff}
-
-}
-
-\caption{ds2}\label{fig:unnamed-chunk-8}
-\end{figure}
-\begin{figure}
-
-{\centering \includegraphics[width=0.9\linewidth]{../Figures/ds3/mu_diff}
-
-}
-
-\caption{ds3}\label{fig:unnamed-chunk-9}
-\end{figure}
-\begin{figure}
-
-{\centering \includegraphics[width=0.9\linewidth]{../Figures/ect/mu_diff}
-
-}
-
-\caption{ect}\label{fig:unnamed-chunk-10}
-\end{figure}
-\begin{figure}
-
-{\centering \includegraphics[width=0.9\linewidth]{../Figures/lps/mu_diff}
-
-}
-
-\caption{lps}\label{fig:unnamed-chunk-11}
-\end{figure}
-\begin{figure}
-
-{\centering \includegraphics[width=0.9\linewidth]{../Figures/lpt/mu_diff}
-
-}
-
-\caption{lpt}\label{fig:unnamed-chunk-12}
-\end{figure}
-\begin{figure}
-
-{\centering \includegraphics[width=0.9\linewidth]{../Figures/maf/lambda_diff}
-
-}
-
-\caption{maf}\label{fig:unnamed-chunk-13}
-\end{figure}
-\begin{figure}
-
-{\centering \includegraphics[width=0.9\linewidth]{../Figures/mav/lambda_diff}
-
-}
-
-\caption{mav}\label{fig:unnamed-chunk-14}
-\end{figure}
-\begin{figure}
-
-{\centering \includegraphics[width=0.9\linewidth]{../Figures/mcv/lambda_diff}
-
-}
-
-\caption{mcv}\label{fig:unnamed-chunk-15}
-\end{figure}
-\begin{figure}
-
-{\centering \includegraphics[width=0.9\linewidth]{../Figures/mdf/lambda_diff}
-
-}
-
-\caption{mdf}\label{fig:unnamed-chunk-16}
-\end{figure}
-\begin{figure}
-
-{\centering \includegraphics[width=0.9\linewidth]{../Figures/mds/lambda_diff}
-
-}
-
-\caption{mds}\label{fig:unnamed-chunk-17}
-\end{figure}
-\begin{figure}
-
-{\centering \includegraphics[width=0.9\linewidth]{../Figures/mpt/lambda_diff}
-
-}
-
-\caption{mpt}\label{fig:unnamed-chunk-18}
-\end{figure}
-\begin{figure}
-
-{\centering \includegraphics[width=0.9\linewidth]{../Figures/pmx/mu_diff}
-
-}
-
-\caption{pmx}\label{fig:unnamed-chunk-19}
-\end{figure}
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.9\linewidth]{../Figures/stl/mu_diff}
-
-}
-
-\caption{stl}\label{fig:unnamed-chunk-20}
-\end{figure}
-
-\hypertarget{references}{%
-\section{References}\label{references}}
+\textcolor{red}{Is this true? Matt's code is at this GitHub page? -YS}
+
+\hypertarget{sec:references}{%
+\section{References}\label{sec:references}}
+
+\hypertarget{sec:supplementary}{%
+\section{Supplementary Material}\label{sec:supplementary}}
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-Huxley1932}{}}%
+Huxley, J S. 1932. \emph{Problems of Relative Growth}. L. MacVeagh.
+
 \leavevmode\vadjust pre{\hypertarget{ref-KendallsTau}{}}%
-KENDALL, M. G. 1938. {``{A NEW MEASURE OF RANK CORRELATION}.''}
-\emph{Biometrika} 30 (1-2): 81--93.
+Kendal, M G. 1938. {``{A new measure of rank correlation}.''}
+\emph{Biometrika} 30 (June): 81--93.
 \url{https://doi.org/10.1093/biomet/30.1-2.81}.
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-Vojeetal2014}{}}%
+Voje, K L, T F Hansen, C K Egset, G H Bolstad, and C Pelabon. 2014.
+{``Allometric Constraints and the Evolution of Allometry.''}
+\emph{Evolution} 68: 866--85.
+
 \leavevmode\vadjust pre{\hypertarget{ref-ZhouReiter2010}{}}%

### 2024-10-21 manuscript/paper.Rmd
finished for the day.

@@
 |Anal Fin Ray Count|`maf`|	Number of bones in the anal fin posterior to the anal spine (i.e., soft anal fin rays)|
-|Abdominal Vertebra Count|mav|	Number of vertebrae anterior to the first vertebra contacting an anal fin pterygiophore (Aguirre et al. 2014)|
-|Caudal Vertebra Count|`mcv`	| Number of vertebrae posterior to and including the first vertebra contacting an anal fin pterygiophore (Aguirre et al. 2014)|
+|Abdominal Vertebra Count|`mav`|	Number of vertebrae anterior to the first vertebra contacting an anal fin pterygiophore (Aguirre et al. 2014)|
+|Caudal Vertebra Count|`mcv`	| Vertebrae including and posterior to the first that contacts an anal fin pterygiophore (Aguirre et al. 2014)|
 |Pterygiophore Count|`mpt`|	Number of pterygiophores anterior to but excluding the pterygiophore under the third dorsal spine|
@@
 Therefore, we imputed $M = 100$ complete datasets.
-\textcolor{red}{PRESUMABLY this also imputes missing data in the other traits?
-We should probably say that explicitly.
--YS}
-
-\textcolor{red}{AKHIL'S METHODS FOR VALIDATION OF THE IMPUTATION MODEL GOES HERE}
+\textcolor{red}{Greg, PRESUMABLY this also imputes missing data in the other traits? We should probably say that explicitly. -YS}
+\textcolor{red}{Greg, AKHIL'S METHODS FOR VALIDATION OF THE IMPUTATION MODEL GOES HERE}
 
@@
 We used a Bayesian approach to estimate trait means in part because imputation is the process of sampling from a posterior predictive distribution of missing data, given observed data.
-\textcolor{red}{THIS NEEDS MORE WORK TO JUSTIFY THE BAYESIAN APPROACH, EXPLAINING IT TO MY AUDIENCE.
+\textcolor{red}{Matt, THIS NEEDS MORE WORK TO JUSTIFY THE BAYESIAN APPROACH, EXPLAINING IT TO MY AUDIENCE.
 -YS}
@@
 ## Missing Data Imputation
-\textcolor{red}{WE NEED RESULTS SECTION DESCRIBING THE RESULTS OF AKHIL'S MULTIPLE IMPUTATION, BOTH THE TESTS OF HOW WELL THE MODEL WORKED CLASSIFYING MODEL SAMPLES, AND SUMMARY STATISTICS OF FOSSIL IMPUTATION RESULTS -YS}
+\textcolor{red}{Greg, WE NEED RESULTS SECTION DESCRIBING THE RESULTS OF AKHIL'S MULTIPLE IMPUTATION, BOTH THE TESTS OF HOW WELL THE MODEL WORKED CLASSIFYING MODEL SAMPLES, AND SUMMARY STATISTICS OF FOSSIL IMPUTATION RESULTS -YS}
 

### 2024-10-24 manuscript/paper.Rmd
Add Matt's changes to the manuscript.

@@
  The goal of our analysis was to estimate fossil trait means through time, by sex, while incorporating uncertainty across imputations in the classification of sex.
-We used a Bayesian approach to estimate trait means in part because imputation is the process of sampling from a posterior predictive distribution of missing data, given observed data.
+As we explain in this section, we are implementing a complex model structure to account for the fact that the rate of change in the mean of the trait over time may be a non-constant value.
+In addition, we are using multiple imputation procedure to account for the fact that the sex is unobserved in the fossils in addition to other pieces of information that are missing.
+We implement a Bayesian approach because these methods are more effective for analyzing hierarchical model structures as well as incorporating missing data.<!--We used a Bayesian approach to estimate trait means in part because imputation is the process of sampling from a posterior predictive distribution of missing data, given observed data.
 \textcolor{red}{Matt, THIS NEEDS MORE WORK TO JUSTIFY THE BAYESIAN APPROACH, EXPLAINING IT TO MY AUDIENCE.
--YS}
-
-\textcolor{red}{MATT, somewhere we need to describe how you handled traits with lots of evolved zeros (ds1, ds2, ds3, lps, tpg) -YS}
+-YS}  -->
+
+<!-- \textcolor{red}{MATT, somewhere we need to describe how you handled traits with lots of evolved zeros (ds1, ds2, ds3, lps, tpg) -YS} -->
 
@@
 Continuous traits will likely have some correlation with standard length (i.e., allometry) (@Huxley1932)(@Vojeetal2022)(@Vojeetal2014).
-We account for this by adding an additional parameter, $\gamma_k$, If $Y_{k,ij}$ is a continuous trait, then
+We account for this by adding an additional parameter, $\gamma_k$.
+In addition, the continuous traits ds1, ds2, ds3, lps, tpg are unique in that these traits tend to be lost over time, results in a proportion of zeros in our empirical dataset.
+For these random variables, we add binary variable $Z_{ij}$, where $Z_{k,ij} = 1$ means that the trait is lost, and $Z_{k,ij} = 0$ means the trait is observed.
+For all other continuous traits, we assume $Z_{k,ij} = 0$ <!--We also define $Y_{k,ij}$ as -->
+
+<!-- \begin{align} -->
+<!-- {Y}_{k,ij} & = \left\{\begin{array}{llll} Y_{k,ij} & Z_{k,ij} = 0 \\ 0 & Z_{k,ij} = 1 \end{array}\right. -->
+<!-- \end{align} -->
+
+If $Y_{k,ij}$ is a continuous trait, then, potentially conditioning on $Z_{k,ij} = 0$, 
 \begin{align}
@@
 \end{align}
+In other words, (\ref{eq:cont}) only considers the subset of fossils where the given trait is present when performing a data analysis.
+For the traits with potential loss, we model the $Z_{k,ij}$ via a $Bernoulli(p_{k,ij})$ distribution where
+\begin{align}
+\text{logit}({p}_{k,ij}) & = \left\{\begin{array}{llll} \alpha_{k,ft_i} + \beta_kY_{K,ij} & W_{ij} = \text{Female} \\ \alpha_{k,mt_i} + \beta_kY_{K,ij}, & W_{ij} = \text{Male} \end{array}\right.,
+\label{eq:cont_prob}
+\end{align}
+and logit is a function to convert the probability to the log-odds, to allow for easier modelling of linear techniques while still ensuring any probability predictions are guaranteed to be between 0 and 1.
 
@@
 \begin{align}
-{Y}_{k,ij} & \sim \left\{\begin{array}{ll} GP(\exp\{\mu_{k,ft_i} + \gamma_kY_{K,ij}\},\alpha_k), & W_{ij} = \text{Female} \\ GP(\exp\{\mu_{k,mt_i} + \gamma_kY_{K,ij}\},\alpha_k), & W_{ij} = \text{Male} \end{array}\right..
+{Y}_{k,ij} & \sim \left\{\begin{array}{ll} GP(\lambda_{k,ft_i} = \exp\{\mu_{k,ft_i} + \gamma_kY_{K,ij}\},\alpha_k), & W_{ij} = \text{Female} \\ GP(\lambda_{k,mt_i} = \exp\{\mu_{k,mt_i} + \gamma_kY_{K,ij}\},\alpha_k), & W_{ij} = \text{Male} \end{array}\right..
 \label{eq:disc_corr}
@@
 \begin{align}
-{Y}_{k,ij} & \sim \left\{\begin{array}{ll} GP(\exp\{\mu_{k,ft_i}\},\alpha_k), & W_{ij} = \text{Female} \\ GP(\exp\{\mu_{k,mt_i}\},\alpha_k), & W_{ij} = \text{Male} \end{array}\right..
+{Y}_{k,ij} & \sim \left\{\begin{array}{ll} GP(\lambda_{k,ft_i} = \exp\{\mu_{k,ft_i}\},\alpha_k), & W_{ij} = \text{Female} \\ GP(\lambda_{k,mt_i} = \exp\{\mu_{k,mt_i}\},\alpha_k), & W_{ij} = \text{Male} \end{array}\right..
 \label{eq:disc_ind}
 \end{align}
-We point out that counting trait means are represented by $\exp\{\mu_{k,ft_i}\}$ and $\exp\{\mu_{k,mt_i}\}$ for ease of modeling.
-
-In the above model descriptions, $\mu_{k,ft_i}$ and $\mu_{k,mt_i}$ model the time-$t_i$, sample-specific trait mean $k$ for females and males, respectively.
-To account for the possibility that there is a time-dependent (i.e., evolutionary trend) trend in $\mu_{k,ft_i}$ and $\mu_{k,mt_i}$, we further set
+We point out that counting trait means are represented by $\exp\{\mu_{k,ft_i}\}$ and $\exp\{\mu_{k,mt_i}\}$.
+This distinction helps in our modelling procedure because $\mu_{k,ft_i}$ and $\mu_{k,mt_i}$ are represented on the real number line.
+This allows for easier modelling by linear techniques, and that any mean for the discrete traits is strictly positive.
+In addition, to ensure proper sampling from our posterior distribution of interest, we transform $\phi_k = \log\left(\frac{\alpha_k - \max_{i,j}(-\lambda_{k,ij}/y_{k,ij})}{1 - \alpha_k}\right)$ where $\lambda_{k,ij} = \left\{\begin{array}{cc} \exp(\mu_{k,ft_i} + \gamma_k Y_{K,ij}) & W_{ij} = \text{Female} \\ \exp(\mu_{k,mt_i} + \gamma_k Y_{K,ij}) & W_{ij} = \text{Male} \end{array}\right.$.
+
+<!-- In the above model descriptions, $\mu_{k,ft_i}$ and $\mu_{k,mt_i}$ model the time-$t_i$, sample-specific (potentially log-transformed) trait mean $k$ for females and males, respectively.
+-->  To account for the possibility that there is a time-dependent (i.e., evolutionary trend) trend in $\mu_{k,ft_i}$ and $\mu_{k,mt_i}$, we further set
 \begin{align}
@@
 for $g \in \{f,m\}$, where $\beta_{0,kg}$ and $\beta_{1,kg}$ are regression parameters of phenotype $Y_k$ for each sex, and $u_{k,gt_i}$ is the corresponding residual.
+For the continuous phenotypes with potential lost trait values, we model $\alpha{k,gt_i}$ in (\ref{eq:cont_prob}) by
+\begin{align}
+\alpha_{k,gt_i} = \beta_{0,p,kg} + \beta_{1,p,kg}t_i + u_{k,p,gt_i}.
+\label{eq:prob}
+\end{align}
+For convenience, we drop the subscript $p$ for all model descriptions to come, and assume same modelling techniques apply.
 To account for potential correlations between the residuals for a given trait $k$ and sex $g$, we fit an Ornstein-Uhlenback (OU) process (@OUProcess).
@@
 
-\textcolor{red}{Q FOR MATT.
+<!-- \textcolor{red}{Q FOR MATT.
 Is transforming the GP onto the real number line separate from setting the priors?
@@
 Then priors get their own section here?
--YS}
-To aid in the sampling procedure for the discrete phenotypes, we first had to transform the parameters of the generalized Poisson distribution onto the real number line in order to ensure our algorithm can properly sample from the posterior distribution of interest using STAN software.
-This was done by declaring $\phi_k = \log\left(\frac{\alpha_k - \max_{i,j}(-\lambda_{k,ij}/y_{k,ij})}{1 - \alpha_k}\right)$ where $\lambda_{k,ij} = \left\{\begin{array}{cc} \exp(\mu_{k,ft_i} + \gamma_k Y_{K,ij}) & W_{ij} = \text{Female} \\ \exp(\mu_{k,mt_i} + \gamma_k Y_{K,ij}) & W_{ij} = \text{Male} \end{array}\right.$.
+-YS} -->
 
@@
 
-For the continuous phenotypes, the mean trait sizes for each sex $\mu_{k,m,t_i}$ and $\mu_{k,f,t_i}$ are calculated using (\ref{eq:mu}).For the discrete phenotypes, equations (\ref{eq:disc_corr}) and (\ref{eq:disc_ind}) calculate the mean trait size is $\lambda_{k,m,t_i} = \exp(\mu_{k,m,t_i})$ and $\lambda_{k,f,t_i} = \exp(\mu_{k,f,t_i})$ for males and females.
-For simplicity, we will use $\mu_{k,m,t_i}$ and $\mu_{k,f,t_i}$ as the notation for mean size for both discrete and continuous traits for the rest of this discussion.
-From our MCMC algorithm, we obtain $M$ samples from the posterior distribution of $\mu_{k,m,t_i} - \mu_{k,f,t_i}$.
-That is, with posterior distributions of the model parameters described above, we obtained $M$ samples from the posterior distribution of mean trait values by trait for males and females at each time step.
+<!-- For the continuous phenotypes, the mean trait sizes for each sex $\mu_{k,m,t_i}$ and $\mu_{k,f,t_i}$ are calculated using (\ref{eq:mu}).For the discrete phenotypes, equations (\ref{eq:disc_corr}) and (\ref{eq:disc_ind}) calculate the mean trait size is $\lambda_{k,m,t_i} = \exp(\mu_{k,m,t_i})$ and $\lambda_{k,f,t_i} = \exp(\mu_{k,f,t_i})$ for males and females.
+With posterior distributions of the model parameters described above, we obtained $R$ samples from the posterior distribution of mean trait values by trait for males and females at each time step.
 We then used these values to quantify the presence or absence of sexual dimorphism, and describe how sexual dimorphism evolved through time.
+-->
 
 ## Quantifying sexual dimorphism, and testing for increased sexual dimorphism through time.
-We quantified sexual dimorphism at each time step by calculating the difference in mean trait size for males and females at each time for phenotype $k$, $\mu_{k,m,t_i} - \mu_{k,f,t_i}$, for $i = 1,\cdots,T$.
-The presence of sexual dimorphism was indicated by...
+We quantify sexual dimorphism at each time $t_i: i = 1,\cdots,T$ by calculating the difference in mean trait size for males and females at each time for a continuous phenotype, $SD_{k,t_i} = \mu_{k,m,t_i} - \mu_{k,f,t_i}$ and for a discrete phenotype $SD_{k,t_i} = \lambda_{k,m,t_i} - \lambda_{k,f,t_i}$.
+For simplicity, we will define sexual dimorphism for both continuous and discrete traits as $SD_{k,t_i} = \mu_{k,m,t_i} - \mu_{k,f,t_i}$.
+Because we are performing a Bayesian analysis, we do not obtain a single point estimate of sexual dimorphism at a particular time, but rather $R$ samples from the posterior distribution of sexual dimorphism given the observed fossil and extant data.
+Therefore, our analysis will be conducted by calculating $SD_{k,t_i,r} = \mu_{k,m,t_i,r} - \mu_{k,f,t_i,r}$ for $r = 1,\dots,R$ and then perfoming our data analysis on these samples.
+Presence (or absense) of sexual dimorphism for trait $k$ can be indicated by estimating the posterior probability of positive sexual dimorphism $P(SD_{k,t_i} > 0 | \boldsymbol{y}) = \frac{1}{R} \sum_{r=1}^R I(SD_{k,t_i,r} > 0)$.
+Figures \ref{fig:post_probs_armor} and \ref{fig:post_probs_nonarmor}  plot these posterior probabilities across time for each of the armored traits and non-armored traits, respectively.
+The closer the $P(SD_{k,t_i} > 0 | \boldsymbol{y})$ is to 1, the larger the evidence that the average trait size for males is larger than females, while if $P(SD_{k,t_i} > 0 | \boldsymbol{y})$ is close to 0, the larger the evidence that the average trait size for females is larger than males, both indicating strong presence of sexual dimorphism.
+On the contrary, $P(SD_{k,t_i} > 0 | \boldsymbol{y})$ around 0.5 indicates that the average trait size is roughly equivalent for both males and females, providing little evidence of dimorphism.
+
+```{r, echo=FALSE, fig.align='center',fig.cap="Posterior probability of positive sexual dimorphism for armor traits"}
+knitr::include_graphics("../Figures/SD_Figures/post_probs_armor.pdf")
+```
+\label{fig:post_probs_armor}
+
+```{r, echo=FALSE, fig.align='center',fig.cap="Posterior probability of positive sexual dimorphism for non-armor traits"}
+knitr::include_graphics("../Figures/SD_Figures/post_probs_nonarmor.pdf")
+```
+\label{fig:post_probs_nonarmor}
+
+<!-- The presence of sexual dimorphism was indicated by...
 \textcolor{red}{Matt, can you help me with Bayesian the language here?
@@
 Values far from fifty fifty indicate that the mass of the posterior distribution is away from a male female difference of zero.
-YS}
+YS} -->
 

### 2024-10-24 manuscript/paper.tex
finished for the day.
Add Matt's changes to the manuscript.

@@
 
-\title{~\Medium Increased sexual dimorphism does not evolve in a fossil
+\title{~\Large Increased sexual dimorphism does not evolve in a fossil
 stickleback following ecological release from fish piscivores}
@@
 sexual dimorphism at each temporal sample, and tracked how it changed
-through time. \vspace{2mm}\\
+through time. NEED TO FINISH THIS \vspace{2mm}\\
 \emph{Keywords}: character displacement, character release, evolutionary
@@
   >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.7119}}@{}}
-\toprule()
+\toprule\noalign{}
 \begin{minipage}[b]{\linewidth}\raggedright
@@
 posterior to the anal spine (i.e., soft anal fin rays) \\
-Abdominal Vertebra Count & mav & Number of vertebrae anterior to the
-first vertebra contacting an anal fin pterygiophore (Aguirre et
+Abdominal Vertebra Count & \texttt{mav} & Number of vertebrae anterior
+to the first vertebra contacting an anal fin pterygiophore (Aguirre et
 al.~2014) \\
-Caudal Vertebra Count & \texttt{mcv} & Number of vertebrae posterior to
-and including the first vertebra contacting an anal fin pterygiophore
-(Aguirre et al.~2014) \\
+Caudal Vertebra Count & \texttt{mcv} & Vertebrae including and posterior
+to the first that contacts an anal fin pterygiophore (Aguirre et
+al.~2014) \\
 Pterygiophore Count & \texttt{mpt} & Number of pterygiophores anterior
@@
 present) \\
-\bottomrule()
 \end{longtable}
@@
 complete datasets.
-\textcolor{red}{PRESUMABLY this also imputes missing data in the other traits? We should probably say that explicitly. -YS}
-
-\textcolor{red}{AKHIL'S METHODS FOR VALIDATION OF THE IMPUTATION MODEL GOES HERE}
+\textcolor{red}{Greg, PRESUMABLY this also imputes missing data in the other traits? We should probably say that explicitly. -YS}
+\textcolor{red}{Greg, AKHIL'S METHODS FOR VALIDATION OF THE IMPUTATION MODEL GOES HERE}
 
@@
 time, by sex, while incorporating uncertainty across imputations in the
-classification of sex. We used a Bayesian approach to estimate trait
-means in part because imputation is the process of sampling from a
-posterior predictive distribution of missing data, given observed data.
-\textcolor{red}{THIS NEEDS MORE WORK TO JUSTIFY THE BAYESIAN APPROACH, EXPLAINING IT TO MY AUDIENCE. -YS}
-
-\textcolor{red}{MATT, somewhere we need to describe how you handled traits with lots of evolved zeros (ds1, ds2, ds3, lps, tpg) -YS}
+classification of sex. As we explain in this section, we are
+implementing a complex model structure to account for the fact that the
+rate of change in the mean of the trait over time may be a non-constant
+value. In addition, we are using multiple imputation procedure to
+account for the fact that the sex is unobserved in the fossils in
+addition to other pieces of information that are missing. We implement a
+Bayesian approach because these methods are more effective for analyzing
+hierarchical model structures as well as incorporating missing data.
 
@@
 (\protect\hyperlink{ref-Vojeetal2014}{2014})). We account for this by
-adding an additional parameter, \(\gamma_k\), If \(Y_{k,ij}\) is a
-continuous trait, then \begin{align}
+adding an additional parameter, \(\gamma_k\). In addition, the
+continuous traits ds1, ds2, ds3, lps, tpg are unique in that these
+traits tend to be lost over time, results in a proportion of zeros in
+our empirical dataset. For these random variables, we add binary
+variable \(Z_{ij}\), where \(Z_{k,ij} = 1\) means that the trait is
+lost, and \(Z_{k,ij} = 0\) means the trait is observed. For all other
+continuous traits, we assume \(Z_{k,ij} = 0\)
+
+If \(Y_{k,ij}\) is a continuous trait, then, potentially conditioning on
+\(Z_{k,ij} = 0\), \begin{align}
 {Y}_{k,ij} & \overset{iid}{\sim}\left\{\begin{array}{llll} \mathcal{N}(\mu_{k,ft_i} + \gamma_kY_{K,ij},\sigma_k^2), & W_{ij} = \text{Female} \\ \mathcal{N}(\mu_{k,mt_i} + \gamma_kY_{K,ij},\sigma_k^2), & W_{ij} = \text{Male} \end{array}\right..
 \label{eq:cont}
-\end{align}
+\end{align} In other words, (\ref{eq:cont}) only considers the subset of
+fossils where the given trait is present when performing a data
+analysis. For the traits with potential loss, we model the \(Z_{k,ij}\)
+via a \(Bernoulli(p_{k,ij})\) distribution where \begin{align}
+\text{logit}({p}_{k,ij}) & = \left\{\begin{array}{llll} \alpha_{k,ft_i} + \beta_kY_{K,ij} & W_{ij} = \text{Female} \\ \alpha_{k,mt_i} + \beta_kY_{K,ij}, & W_{ij} = \text{Male} \end{array}\right.,
+\label{eq:cont_prob}
+\end{align} and logit is a function to convert the probability to the
+log-odds, to allow for easier modelling of linear techniques while still
+ensuring any probability predictions are guaranteed to be between 0 and
+1.
 
@@
 \(Y_{k,ij}\) is one of the above traits, then \begin{align}
-{Y}_{k,ij} & \sim \left\{\begin{array}{ll} GP(\exp\{\mu_{k,ft_i} + \gamma_kY_{K,ij}\},\alpha_k), & W_{ij} = \text{Female} \\ GP(\exp\{\mu_{k,mt_i} + \gamma_kY_{K,ij}\},\alpha_k), & W_{ij} = \text{Male} \end{array}\right..
+{Y}_{k,ij} & \sim \left\{\begin{array}{ll} GP(\lambda_{k,ft_i} = \exp\{\mu_{k,ft_i} + \gamma_kY_{K,ij}\},\alpha_k), & W_{ij} = \text{Female} \\ GP(\lambda_{k,mt_i} = \exp\{\mu_{k,mt_i} + \gamma_kY_{K,ij}\},\alpha_k), & W_{ij} = \text{Male} \end{array}\right..
 \label{eq:disc_corr}
@@
 change with size in adult fish. Thus, \begin{align}
-{Y}_{k,ij} & \sim \left\{\begin{array}{ll} GP(\exp\{\mu_{k,ft_i}\},\alpha_k), & W_{ij} = \text{Female} \\ GP(\exp\{\mu_{k,mt_i}\},\alpha_k), & W_{ij} = \text{Male} \end{array}\right..
+{Y}_{k,ij} & \sim \left\{\begin{array}{ll} GP(\lambda_{k,ft_i} = \exp\{\mu_{k,ft_i}\},\alpha_k), & W_{ij} = \text{Female} \\ GP(\lambda_{k,mt_i} = \exp\{\mu_{k,mt_i}\},\alpha_k), & W_{ij} = \text{Male} \end{array}\right..
 \label{eq:disc_ind}
 \end{align} We point out that counting trait means are represented by
-\(\exp\{\mu_{k,ft_i}\}\) and \(\exp\{\mu_{k,mt_i}\}\) for ease of
-modeling.
-
-In the above model descriptions, \(\mu_{k,ft_i}\) and \(\mu_{k,mt_i}\)
-model the time-\(t_i\), sample-specific trait mean \(k\) for females and
-males, respectively. To account for the possibility that there is a
-time-dependent (i.e., evolutionary trend) trend in \(\mu_{k,ft_i}\) and
-\(\mu_{k,mt_i}\), we further set \begin{align}
+\(\exp\{\mu_{k,ft_i}\}\) and \(\exp\{\mu_{k,mt_i}\}\). This distinction
+helps in our modelling procedure because \(\mu_{k,ft_i}\) and
+\(\mu_{k,mt_i}\) are represented on the real number line. This allows
+for easier modelling by linear techniques, and that any mean for the
+discrete traits is strictly positive. In addition, to ensure proper
+sampling from our posterior distribution of interest, we transform
+\(\phi_k = \log\left(\frac{\alpha_k - \max_{i,j}(-\lambda_{k,ij}/y_{k,ij})}{1 - \alpha_k}\right)\)
+where
+\(\lambda_{k,ij} = \left\{\begin{array}{cc} \exp(\mu_{k,ft_i} + \gamma_k Y_{K,ij}) & W_{ij} = \text{Female} \\ \exp(\mu_{k,mt_i} + \gamma_k Y_{K,ij}) & W_{ij} = \text{Male} \end{array}\right.\).
+
+To account for the possibility that there is a time-dependent (i.e.,
+evolutionary trend) trend in \(\mu_{k,ft_i}\) and \(\mu_{k,mt_i}\), we
+further set \begin{align}
 \mu_{k,gt_i} = \beta_{0,kg} + \beta_{1,kg}t_i + u_{k,gt_i},
@@
 \(\beta_{1,kg}\) are regression parameters of phenotype \(Y_k\) for each
-sex, and \(u_{k,gt_i}\) is the corresponding residual. To account for
-potential correlations between the residuals for a given trait \(k\) and
-sex \(g\), we fit an Ornstein-Uhlenback (OU) process (Uhlenbeck and
-Ornstein (\protect\hyperlink{ref-OUProcess}{1930})). More specifically,
-we defined \(du_{k,gt} = u_{k,g(t + dt)} - u_{k,gt}\), the change in
-\(u_k,gt\) for a given trait \(k\) and sex \(g\) over a miniscule time
-period \(dt\). The OU process is defined as \begin{align}
+sex, and \(u_{k,gt_i}\) is the corresponding residual. For the
+continuous phenotypes with potential lost trait values, we model
+\(\alpha{k,gt_i}\) in (\ref{eq:cont_prob}) by \begin{align}
+\alpha_{k,gt_i} = \beta_{0,p,kg} + \beta_{1,p,kg}t_i + u_{k,p,gt_i}.
+\label{eq:prob}
+\end{align} For convenience, we drop the subscript \(p\) for all model
+descriptions to come, and assume same modelling techniques apply. To
+account for potential correlations between the residuals for a given
+trait \(k\) and sex \(g\), we fit an Ornstein-Uhlenback (OU) process
+(Uhlenbeck and Ornstein (\protect\hyperlink{ref-OUProcess}{1930})). More
+specifically, we defined \(du_{k,gt} = u_{k,g(t + dt)} - u_{k,gt}\), the
+change in \(u_k,gt\) for a given trait \(k\) and sex \(g\) over a
+miniscule time period \(dt\). The OU process is defined as \begin{align}
 du_{k,gt} = -\kappa_k u_{k,gt} dt + \tau_k dW_t,
@@
 
-\textcolor{red}{Q FOR MATT.
-Is transforming the GP onto the real number line separate from setting the priors?
-If so, should the statement of transformation go in the section where the GP and discrete trait modeling is discussed.
-Then priors get their own section here?
--YS}
-To aid in the sampling procedure for the discrete phenotypes, we first
-had to transform the parameters of the generalized Poisson distribution
-onto the real number line in order to ensure our algorithm can properly
-sample from the posterior distribution of interest using STAN software.
-This was done by declaring
-\(\phi_k = \log\left(\frac{\alpha_k - \max_{i,j}(-\lambda_{k,ij}/y_{k,ij})}{1 - \alpha_k}\right)\)
-where
-\(\lambda_{k,ij} = \left\{\begin{array}{cc} \exp(\mu_{k,ft_i} + \gamma_k Y_{K,ij}) & W_{ij} = \text{Female} \\ \exp(\mu_{k,mt_i} + \gamma_k Y_{K,ij}) & W_{ij} = \text{Male} \end{array}\right.\).
-
 We chose the following priors, designing them to be weakly informative
@@
 
-For the continuous phenotypes, the mean trait sizes for each sex
-\(\mu_{k,m,t_i}\) and \(\mu_{k,f,t_i}\) are calculated using
-(\ref{eq:mu}).For the discrete phenotypes, equations
-(\ref{eq:disc_corr}) and (\ref{eq:disc_ind}) calculate the mean trait
-size is \(\lambda_{k,m,t_i} = \exp(\mu_{k,m,t_i})\) and
-\(\lambda_{k,f,t_i} = \exp(\mu_{k,f,t_i})\) for males and females. For
-simplicity, we will use \(\mu_{k,m,t_i}\) and \(\mu_{k,f,t_i}\) as the
-notation for mean size for both discrete and continuous traits for the
-rest of this discussion. From our MCMC algorithm, we obtain \(M\)
-samples from the posterior distribution of
-\(\mu_{k,m,t_i} - \mu_{k,f,t_i}\). That is, with posterior distributions
-of the model parameters described above, we obtained \(M\) samples from
-the posterior distribution of mean trait values by trait for males and
-females at each time step. We then used these values to quantify the
-presence or absence of sexual dimorphism, and describe how sexual
-dimorphism evolved through time.
-
 \hypertarget{quantifying-sexual-dimorphism-and-testing-for-increased-sexual-dimorphism-through-time.}{%
@@
 
-We quantified sexual dimorphism at each time step by calculating the
-difference in mean trait size for males and females at each time for
-phenotype \(k\), \(\mu_{k,m,t_i} - \mu_{k,f,t_i}\), for
-\(i = 1,\cdots,T\). The presence of sexual dimorphism was indicated
-by\ldots{}
-\textcolor{red}{Matt, can you help me with Bayesian the language here?
-I think this is the place to reference the two post probs plots, yes?
-Values far from fifty fifty indicate that the mass of the posterior distribution is away from a male female difference of zero.
-YS}
+We quantify sexual dimorphism at each time \(t_i: i = 1,\cdots,T\) by
+calculating the difference in mean trait size for males and females at
+each time for a continuous phenotype,
+\(SD_{k,t_i} = \mu_{k,m,t_i} - \mu_{k,f,t_i}\) and for a discrete
+phenotype \(SD_{k,t_i} = \lambda_{k,m,t_i} - \lambda_{k,f,t_i}\). For
+simplicity, we will define sexual dimorphism for both continuous and
+discrete traits as \(SD_{k,t_i} = \mu_{k,m,t_i} - \mu_{k,f,t_i}\).
+Because we are performing a Bayesian analysis, we do not obtain a single
+point estimate of sexual dimorphism at a particular time, but rather
+\(R\) samples from the posterior distribution of sexual dimorphism given
+the observed fossil and extant data. Therefore, our analysis will be
+conducted by calculating
+\(SD_{k,t_i,r} = \mu_{k,m,t_i,r} - \mu_{k,f,t_i,r}\) for
+\(r = 1,\dots,R\) and then perfoming our data analysis on these samples.
+Presence (or absense) of sexual dimorphism for trait \(k\) can be
+indicated by estimating the posterior probability of positive sexual
+dimorphism
+\(P(SD_{k,t_i} > 0 | \boldsymbol{y}) = \frac{1}{R} \sum_{r=1}^R I(SD_{k,t_i,r} > 0)\).
+Figures \ref{fig:post_probs_armor} and \ref{fig:post_probs_nonarmor}
+plot these posterior probabilities across time for each of the armored
+traits and non-armored traits, respectively. The closer the
+\(P(SD_{k,t_i} > 0 | \boldsymbol{y})\) is to 1, the larger the evidence
+that the average trait size for males is larger than females, while if
+\(P(SD_{k,t_i} > 0 | \boldsymbol{y})\) is close to 0, the larger the
+evidence that the average trait size for females is larger than males,
+both indicating strong presence of sexual dimorphism. On the contrary,
+\(P(SD_{k,t_i} > 0 | \boldsymbol{y})\) around 0.5 indicates that the
+average trait size is roughly equivalent for both males and females,
+providing little evidence of dimorphism.
+
+\begin{figure}
+
+{\centering \includegraphics{../Figures/SD_Figures/post_probs_armor}
+
+}
+
+\caption{Posterior probability of positive sexual dimorphism for armor traits}\label{fig:unnamed-chunk-2}
+\end{figure}
+
+\label{fig:post_probs_armor}
+
+\begin{figure}
+
+{\centering \includegraphics{../Figures/SD_Figures/post_probs_nonarmor}
+
+}
+
+\caption{Posterior probability of positive sexual dimorphism for non-armor traits}\label{fig:unnamed-chunk-3}
+\end{figure}
+
+\label{fig:post_probs_nonarmor}
 
@@
 
-\textcolor{red}{WE NEED RESULTS SECTION DESCRIBING THE RESULTS OF AKHIL'S MULTIPLE IMPUTATION, BOTH THE TESTS OF HOW WELL THE MODEL WORKED CLASSIFYING MODEL SAMPLES, AND SUMMARY STATISTICS OF FOSSIL IMPUTATION RESULTS -YS}
+\textcolor{red}{Greg, WE NEED RESULTS SECTION DESCRIBING THE RESULTS OF AKHIL'S MULTIPLE IMPUTATION, BOTH THE TESTS OF HOW WELL THE MODEL WORKED CLASSIFYING MODEL SAMPLES, AND SUMMARY STATISTICS OF FOSSIL IMPUTATION RESULTS -YS}
 

### 2024-10-30 manuscript/paper.tex
not sure what these are

--- 2024-10-21 manuscript/paper.tex
@@
 posterior predictive distribution of missing data, given observed data.
-\textcolor{red}{THIS NEEDS MORE WORK TO JUSTIFY THE BAYESIAN APPROACH, EXPLAINING IT TO MY AUDIENCE. -YS}
+\textcolor{red}{Matt, THIS NEEDS MORE WORK TO JUSTIFY THE BAYESIAN APPROACH, EXPLAINING IT TO MY AUDIENCE. -YS}
 

### 2024-11-01 manuscript/paper.tex
knit productions

--- 2024-10-24 manuscript/paper.tex
@@
   >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.7119}}@{}}
-\toprule\noalign{}
+\toprule()
 \begin{minipage}[b]{\linewidth}\raggedright
@@
 \end{minipage} \\
-\midrule\noalign{}
+\midrule()
 \endhead
-\bottomrule\noalign{}
-\endlastfoot
 Standard Length & \texttt{stl} & Distance from anterior tip of
@@
 present) \\
+\bottomrule()
 \end{longtable}

### 2024-11-01 manuscript/paper.Rmd
manuscript edits

@@
 As we explain in this section, we are implementing a complex model structure to account for the fact that the rate of change in the mean of the trait over time may be a non-constant value.
-In addition, we are using multiple imputation procedure to account for the fact that the sex is unobserved in the fossils in addition to other pieces of information that are missing.
-We implement a Bayesian approach because these methods are more effective for analyzing hierarchical model structures as well as incorporating missing data.<!--We used a Bayesian approach to estimate trait means in part because imputation is the process of sampling from a posterior predictive distribution of missing data, given observed data.
+We implement a Bayesian approach because these methods are more effective for analyzing the hierarchical model structures we implement.
+<!--We used a Bayesian approach to estimate trait means in part because imputation is the process of sampling from a posterior predictive distribution of missing data, given observed data.
 \textcolor{red}{Matt, THIS NEEDS MORE WORK TO JUSTIFY THE BAYESIAN APPROACH, EXPLAINING IT TO MY AUDIENCE.
@@
 We account for this by adding an additional parameter, $\gamma_k$.
-In addition, the continuous traits ds1, ds2, ds3, lps, tpg are unique in that these traits tend to be lost over time, results in a proportion of zeros in our empirical dataset.
-For these random variables, we add binary variable $Z_{ij}$, where $Z_{k,ij} = 1$ means that the trait is lost, and $Z_{k,ij} = 0$ means the trait is observed.
+In addition, the continuous traits ds1, ds2, ds3, lps, tpg are notable because they tend to be lost during the evolution of armor reduction in this system (CITATIONS).
+This generates meaningful zeros for trait lengths in our empirical dataset that \textcolor{red}{MATT, help describing why we wanted to analyze with and without 0s?
+-YS}.
+For these random variables, we added the binary variable $Z_{ij}$, where $Z_{k,ij} = 1$ means that the trait is lost, and $Z_{k,ij} = 0$ means the trait is observed.
 For all other continuous traits, we assume $Z_{k,ij} = 0$ <!--We also define $Y_{k,ij}$ as -->

### 2024-11-01 manuscript/paper.Rmd
results text

@@
 ## Modeling fossil stickleback trait means through time for males and females
- The goal of our analysis was to estimate fossil trait means through time, by sex, while incorporating uncertainty across imputations in the classification of sex.
-As we explain in this section, we are implementing a complex model structure to account for the fact that the rate of change in the mean of the trait over time may be a non-constant value.
-We implement a Bayesian approach because these methods are more effective for analyzing the hierarchical model structures we implement.
-<!--We used a Bayesian approach to estimate trait means in part because imputation is the process of sampling from a posterior predictive distribution of missing data, given observed data.
-\textcolor{red}{Matt, THIS NEEDS MORE WORK TO JUSTIFY THE BAYESIAN APPROACH, EXPLAINING IT TO MY AUDIENCE.
--YS}  -->
-
-<!-- \textcolor{red}{MATT, somewhere we need to describe how you handled traits with lots of evolved zeros (ds1, ds2, ds3, lps, tpg) -YS} -->
+ The goal of our analysis was to model fossil trait means through time, by sex, while incorporating uncertainty across imputations in the classification of sex.
+We therefore implemented a Bayesian hierarchical model structure (described below) to account for classfication uncertainty and the fact that the rate of change of any one trait mean over time may be a non-constant value.
 
@@
 We account for this by adding an additional parameter, $\gamma_k$.
-In addition, the continuous traits ds1, ds2, ds3, lps, tpg are notable because they tend to be lost during the evolution of armor reduction in this system (CITATIONS).
-This generates meaningful zeros for trait lengths in our empirical dataset that \textcolor{red}{MATT, help describing why we wanted to analyze with and without 0s?
--YS}.
-For these random variables, we added the binary variable $Z_{ij}$, where $Z_{k,ij} = 1$ means that the trait is lost, and $Z_{k,ij} = 0$ means the trait is observed.
+In addition, the continuous traits ds1, ds2, ds3, lps, tpg are notable because they tend to be lost during the evolution of armor reduction in this system (@Belletal1985, @Stuartetal2020).
+This generates meaningful zeros for trait lengths in our empirical dataset that violate model assumptions.
+Therefore, for these random variables, we added the binary variable $Z_{ij}$, where $Z_{k,ij} = 1$ means that the trait is lost, and $Z_{k,ij} = 0$ means the trait is observed.
 For all other continuous traits, we assume $Z_{k,ij} = 0$ <!--We also define $Y_{k,ij}$ as -->
@@
 and logit is a function to convert the probability to the log-odds, to allow for easier modelling of linear techniques while still ensuring any probability predictions are guaranteed to be between 0 and 1.
+Modeling traits this way allows us to first estimate the probability of trait loss for an individual at a given time, and ask if the sexes differ in their tendency toward loss through time.
+Then, for individuals that have not lost the trait, we can estimate trait values and calculate sexual dimorphism through time.
 
@@
 
-<!-- \textcolor{red}{Q FOR MATT.
-Is transforming the GP onto the real number line separate from setting the priors?
-If so, should the statement of transformation go in the section where the GP and discrete trait modeling is discussed.
-Then priors get their own section here?
--YS} -->
-
 We chose the following priors, designing them to be weakly informative such that the posterior distributions could be influenced by the data, if warranted. For $k = 1,\cdots,K$,
@@
 $\boldsymbol{\tau}_k = \{\tau_{k,r}: r = 1,\cdots,R\}$ for trait $k$ where $\tau_r$ is the estimate of the Kendall's $\tau$ correlation for $\mu_{k,m,t_i} - \mu_{k,f,t_i}$ versus time $t_i$ for the $r^{th}$ MCMC sample from the posterior distribution.
-\textcolor{red}{Matt, I think there's an absolute value problem here.
-Dimorphism can increase of M-F gets more positive or more negative.
-I've indicated that in the text below, but I could use some help refininig the explanation -YS}
-For each trait, the proportion of values in $\boldsymbol{\tau}_k$ that are greater than 0 ($\frac{1}{R} \sum_{r=1}^R I(\tau_{k,r} > 0)$) may indicate that males have gotten relatively larger than females, indicating more pronounced dimorphism.
-For each trait, the proportion of values in $\boldsymbol{\tau}_k$ that are less than 0 ($\frac{1}{R} \sum_{r=1}^R I(\tau_{k,r} < 0)$) may indicate that females have gotten relatively larger than males, indicating more pronounced dimormphism.
+For each trait, a large proportion of values in $\boldsymbol{\tau}_k$ that are greater than 0 ($\frac{1}{R} \sum_{r=1}^R I(\tau_{k,r} > 0)$) may indicate that males have gotten relatively larger than females through time, indicating an increase in male-biased dimorphism through time.
+A large proportion of values in $\boldsymbol{\tau}_k$ that are less than 0 ($\frac{1}{R} \sum_{r=1}^R I(\tau_{k,r} < 0)$) may indicate that females have gotten relatively larger than males, indicating an increase in female-biased dimorphism through time.
 
@@
 ## Sexual Dimorphism
+
 Table \ref{tab:prob0-table} shows the posterior probability of the differences in the means between male and female specimens at each time point.
@@
 ## Changes over time
+We estimated the change in sexual dimorphism over time by calculating the distribution of Kendall's Tau for each trait.
+Here, positive values of Tau indicate that sexual dimorphism (defined as male means minus female means) increased through time.
+Average Tau values ranged from X to Y across traits and the proportion of positive area under the curve ranged from Z to W (Table Tau).
+Distributions are available for each trait in the Supplementary Material (Figures SX-SX+15).
+In Figure (Tau), we show representative distributions.
+For example,
+
+Something about the probability of dimorphim in the probability of loss?
+I.e., the prob_zero plots.
+For example, ds2_prob_zero has lots of positive area under the curve indicating that the rate of loss in males and females diverged, such that males were more likely to lose ds2 through time.
+Interestingly, for the individuals that kept their spines, dimorphism increased, with females getting larger than males (or males getting smaller than females).
+
 We can also assess the strength of the dimorphism by calculating the proportion of posterior samples that are greater than a prespecified threshold \textcolor{red}{(0.5??, idk what this could be.) -MS}
 
-Table \ref{tab:auc-table} shows the probability that the posterior distribution of the differences in the mean is greater than the posterior distribution of the differences at the first time point.
+<!-- Table \ref{tab:auc-table} shows the probability that the posterior distribution of the differences in the mean is greater than the posterior distribution of the differences at the first time point.
 Values near 0.5 indicate very little difference between the distributions whereas values closer to 0 and 1 indicate changes in the distributions with values of 1 meaning the difference in means has shifted towards males having a larger mean and values of 0 indicating the difference in means has shifted towards larger means in females.
 These results are shown graphically in figure \ref{fig:AUC-plot}.
+-->
 

### 2024-11-08 manuscript/paper.Rmd
updated.

@@
 ## Missing Data Imputation
+```{r echo = FALSE}
+#Summaries
+dat <- read.csv("Data/updated_stickle_imputations_100.csv")
+
+```
+
+```{r eval = FALSE}
+library(mixgb)
+library(mice)
+library(tidyverse)
+library(janitor)
+library(missMethods)
+library(ROCR)
+
+stickle <- read_csv('./Data/Stickle_revamped.csv')
+
+stick <- stickle %>% 
+  mutate(gender = as.factor(case_when(
+    grepl(".M.",univID,fixed=TRUE) ~ "Male",
+    grepl(".F.",univID,fixed=TRUE) ~ "Female",
+    grepl("Fos",univID,fixed=TRUE) ~ NA_character_
+    )), ips = as.factor(ips)
+  ) %>% 
+  select(gender, everything())
+
+stick
+
+extant_stickle <- stick %>% filter(gender != "Fossil") %>% select(!univID)
+extant_stickle
+summary(extant_stickle)
+
+M = 10
+k = 5
+stick_matrix <- matrix(NA, nrow=nrow(extant_stickle), ncol = M)
+set.seed(12)
+fold <- sample(1:k, size = nrow(extant_stickle), replace=TRUE)
+for(i in 1:k){
+extant_stickle_temp <- extant_stickle
+extant_stickle_temp$gender[fold==i] <- NA
+imputed_stick <- mice(extant_stickle_temp,m=M)
+for(n in 1:M){
+
+  complete_stick <- complete(imputed_stick, n)
+
+  complete_stick$gender[fold==i]
+  stick_matrix[fold==i, n] <- as.character(complete_stick$gender[fold==i])
+}
+
+}
+pct_male <- apply(stick_matrix,1,function(x){
+  mean(x == "Male")
+})
+
+imptest <- data_frame(truth = as.factor(extant_stickle$gender), pct_male)
+save(imptest, file = "./Data/imptest.RData")
+roc_pred <- prediction(imptest$pct_male, imptest$truth)
+
+auc_rocr <- performance(roc_pred, measure="auc")
+plot(performance(roc_pred, measure="tpr", x.measure="fpr"))
+
+auc_rocr@y.values[[1]]
+
+```
+
+```{r}
+load("./Data/imptest.RData")
+roc_pred <- prediction(imptest$pct_male, imptest$truth)
+
+auc_rocr <- performance(roc_pred, measure="auc")
+perf <- performance(roc_pred, measure="tpr", x.measure="fpr")
+plot(perf)
+dat <- data.frame(fpr = perf@x.values[[1]], tpr = perf@y.values[[1]])
+ggplot(aes(x = fpr, y = tpr), data = dat) + geom_path() + theme_bw() + xlab("False Positive Rate") + ylab("True Positive Rate") + geom_abline(slope = 1, lty = 2, color = rgb(0,0,0,0.2)) + coord_fixed()
+
+```
+
+total sampel size1181
+fossils 814
+extant 367
+`r auc_rocr@y.values[[1]]`
+
 \textcolor{red}{Greg, WE NEED RESULTS SECTION DESCRIBING THE RESULTS OF AKHIL'S MULTIPLE IMPUTATION, BOTH THE TESTS OF HOW WELL THE MODEL WORKED CLASSIFYING MODEL SAMPLES, AND SUMMARY STATISTICS OF FOSSIL IMPUTATION RESULTS -YS}

### 2024-11-08 manuscript/paper.Rmd
FMI stuff.  AUC stuff.  Some other updates

@@
 ## Missing Data Imputation
-```{r echo = FALSE}
-#Summaries
-dat <- read.csv("Data/updated_stickle_imputations_100.csv")
-
-```
 
@@
 
-stickle <- read_csv('./Data/Stickle_revamped.csv')
+stickle <- read_csv('../Data/Stickle_revamped.csv')
 
@@
 load("./Data/imptest.RData")
+library(ROCR)
+library(tidyverse)
 roc_pred <- prediction(imptest$pct_male, imptest$truth)
@@
 perf <- performance(roc_pred, measure="tpr", x.measure="fpr")
-plot(perf)
 dat <- data.frame(fpr = perf@x.values[[1]], tpr = perf@y.values[[1]])
+```
+
+```{r fig.cap = "Reciver-Operating Characteristic Curve comparing imputed sex to actual sex for the extant cases using 5-fold cross validation (AUC = 0.831)"}
 ggplot(aes(x = fpr, y = tpr), data = dat) + geom_path() + theme_bw() + xlab("False Positive Rate") + ylab("True Positive Rate") + geom_abline(slope = 1, lty = 2, color = rgb(0,0,0,0.2)) + coord_fixed()
-
-```
-
-total sampel size1181
-fossils 814
-extant 367
-`r auc_rocr@y.values[[1]]`
+```
+
+<!-- While sex is observed for extant fish, the sex for fossil specimen cannot be observed.
+Therefore, we turn to multiple imputation to fill in missing values of sex of fossil specimen using the observed extant fish cases.
+In addition to missing sex for all of the fossil cases, there are several other cases of missing data in some variables in the fossil data, and those are imputed at the same time as sex is imputed.
+-->
+
+In order to test the efficacy of imputing sex in the fossil data, we performed 5-fold cross validation using only the extant cases.
+For each fold, the missing values, most notably sex, were imputed $M = 5$ times.
+These imputations were then combined to estimate the probability of male vs female and an ROC curve was computed to evaluate the accuracy of the imputations.
+The AUC was computed to be `r auc_rocr@y.values[[1]]`, indicating a good performance in estimating the probability of a fish being male/female.
+
+<!-- total sampel size1181 -->
+<!-- fossils 814 -->
+<!-- extant 367 -->
+
+```{r echo = FALSE, fig.cap = "Density estimate of the percentage of specimen that were imputted to be male across the M = 100 imputations"}
+#Summaries
+dat <- read.csv("../Data/updated_stickle_imputations_100.csv")
+dat %>% group_by(imp) %>% summarize(pct = 100*mean(gender == "Male")) %>% ggplot(aes(x = pct)) + geom_density() + theme_bw()
+```
+
+```{r echo = FALSE}
+meanpct <- dat %>% group_by(imp) %>% summarize(pct = 100*mean(gender == "Male")) %>% ungroup() %>% summarize(pct = mean(pct))
+sdpct <- dat %>% group_by(imp) %>% summarize(pct = 100*mean(gender == "Male")) %>% ungroup() %>% summarize(sd = sd(pct))
+
+min <- dat %>% group_by(imp) %>% summarize(pct = 100*mean(gender == "Male")) %>% ungroup() %>% summarize(sd = min(pct))
+max <- dat %>% group_by(imp) %>% summarize(pct = 100*mean(gender == "Male")) %>% ungroup() %>% summarize(sd = min(pct))
+
+dat %>% group_by(imp) %>% summarize(pct = sum(gender == "Male") == sum(gender == "Female")) %>% ungroup() %>% summarize(pct = mean(pct))
+
+```
+
+Across the 100 imputations of the 814 fossil specimen, the average percentage of males in an imputation was `r paste0(round(meanpct$pct,2),"%")` with a standard deviation of `r paste0(round(sdpct$sd,2),"%")` ranging from `r paste0(round(min$sd,2),"%")` to `r paste0(round(max$sd,2),"%")`.
+In only 31 of the 100 imputations were there more males than females with the remaining 69 imputations having exactly the same number or more females.
+
+\textcolor{red}{Yoel, I think I remember you talking about a reason why it made more sense that there were fewer males in the sample tham the females, but I don't remember the details.
+Am I remembering this correctly?
+Would be nice to add something here about that and say why these imputations are consistent with that.
+-GJM}
+
+<!-- If we want to compute fraction of missing information.  This also computes imputed means.  -->
+```{r echo = FALSE}
+greg <- function(x){
+dat %>% group_by(imp) %>% summarize(mean = mean(.data[[x]]), var = var(.data[[x]]))  %>% summarize(Q = mean(mean),  W = mean(var), B = var(var)) %>% mutate(Tm = (1+1/100)*B + W, lambda = (1+1/100)*B/Tm, se= sqrt(Tm))
+}
+data.frame(var = names(dat)[4:28],do.call(rbind,lapply(as.list(names(dat)[4:28]),greg))) %>% select(var, Q, se, lambda)
+
+#Fraction of missing information
+# Raghunathan, T. 2016. Missing Data Analysis in Practice. Boca Raton, FL: Boca Raton: CRC Press.
+# Van Buuren, S. 2018. Flexible Imputation of Missing Data. Second Edition. Boca Raton, FL: Chapman & Hall/CRC.
+```
 
@@
 ## Sexual Dimorphism 
-
 Table \ref{tab:prob0-table} shows the posterior probability of the differences in the means between male and female specimens at each time point.

### 2024-11-11 manuscript/paper.Rmd
Paper and bib updated.

@@
 ## Missing data imputation, including fossil sex
-Briefly, we used multiple imputation (@little2002statistical) to impute the sex of the fossils.
+Fossil and extant data were combined to create a single data set which contained small amounts on missingness on the observed traits with the exception of sex, which was completely missing for fossil specimens as this was unobservable.
+The sex of the fossil data was considered to be missing data and was imputed in the same step of the rest of the missing data using multiple imputation (@little2002statistical).
+Imputation was performed using the multiple imputation by chained equations (MICE) algorithm (@MICE) via predictive mean matching (PMM) (@hotdeck) and implemented using the R programming language (@R2022language).
+Traditionally, the choice for the number of completed data sets is a relatively small number such as $M = 5$ or $M = 10$.
+However, @ZhouReiter2010 recommend a larger number of imputed data sets if users intend on performing Bayesian analysis, as was performed here, after imputation.
+Therefore, we imputed $M = 100$ complete data sets.
+Extant fish records were then dropped from each of the imputed data sets, and only the fossil data was used in downstream analyses of fossil sexual dimorphism.
+
+After performing the Bayesian analysis on each of the $M = 100$ imputed data sets, draws from the posterior distribution based on each of the imputed data sets were pooled together to combine posterior draws across imputations as recommended in @ZhouReiter2010 rather than using Rubin's traditional combining rules.
+
+<!-- Briefly, we used multiple imputation (@little2002statistical) to impute the sex of the fossils.
 Let $\boldsymbol{W}$ be an $(n_{extant} + n_{fossil}) \times 1$ vector of the covariate sex of the stickleback fish and $\boldsymbol{Y}$ be an $(n_{extant} + n_{fossil}) \times K$ matrix of the $K$ phenotypes of interest.
@@
 Therefore, we imputed $M = 100$ complete datasets.
+-->
 \textcolor{red}{Greg, PRESUMABLY this also imputes missing data in the other traits? We should probably say that explicitly. -YS}
+\textcolor{orange}{Updated - GReg}
 \textcolor{red}{Greg, AKHIL'S METHODS FOR VALIDATION OF THE IMPUTATION MODEL GOES HERE}
-
-We then pooled all of the draws from the posterior distribution across all the imputed data sets to estimate the posterior distributions of our parameters of interest (i.e., the trait values), rather than using Rubin's combining rules, following @ZhouReiter2010.
+\textcolor{orange}{Are we sure we want that here?  I was going to add it in later}
+
+<!-- We then pooled all of the draws from the posterior distribution across all the imputed data sets to estimate the posterior distributions of our parameters of interest (i.e., the trait values), rather than using Rubin's combining rules, following @ZhouReiter2010.
 We then dropped extant fish from our imputed data sets and used these posterior distributions in our downstream analyses of fossil sexual dimorphism.
+-->
 
@@
 # Results {#sec:results}
-The extant data used here consist of a total of `r sum(!is.na(stick$gender))` specimens all with known sex (Table S1).
+
+## Missing Data Imputation
+<!-- The extant data used here consist of a total of `r sum(!is.na(stick$gender))` specimens all with known sex (Table S1).
 Of these, there are `r sum(stick$gender == "Female",na.rm = TRUE)` and `r sum(stick$gender == "Male",na.rm = TRUE)` female and male specimens, respectively.
-
-## Missing Data Imputation
-
+-->
+
+### Validating Imputation of Sex
 ```{r eval = FALSE}
@@
 ```{r}
-load("./Data/imptest.RData")
+load("../Data/imptest.RData")
 library(ROCR)
@@
 
-```{r fig.cap = "Reciver-Operating Characteristic Curve comparing imputed sex to actual sex for the extant cases using 5-fold cross validation (AUC = 0.831)"}
+```{r roc, fig.cap = "Reciver-Operating Characteristic Curve comparing imputed sex to actual sex for the extant cases using 5-fold cross validation (AUC = 0.831)"}
 ggplot(aes(x = fpr, y = tpr), data = dat) + geom_path() + theme_bw() + xlab("False Positive Rate") + ylab("True Positive Rate") + geom_abline(slope = 1, lty = 2, color = rgb(0,0,0,0.2)) + coord_fixed()
@@
 
-In order to test the efficacy of imputing sex in the fossil data, we performed 5-fold cross validation using only the extant cases.
-For each fold, the missing values, most notably sex, were imputed $M = 5$ times.
-These imputations were then combined to estimate the probability of male vs female and an ROC curve was computed to evaluate the accuracy of the imputations.
-The AUC was computed to be `r auc_rocr@y.values[[1]]`, indicating a good performance in estimating the probability of a fish being male/female.
-
-<!-- total sampel size1181 -->
-<!-- fossils 814 -->
-<!-- extant 367 -->
-
-```{r echo = FALSE, fig.cap = "Density estimate of the percentage of specimen that were imputted to be male across the M = 100 imputations"}
+In order to validate the efficacy of imputing sex in the fossil data, we performed 5-fold cross validation using only the extant cases.
+For a given fold, we took the true, observed values of sex and removed them to create missing data similar to what we observe with the fossil data.
+Then with one fold held out, we performed multiple imputation with $M$ = 5 (using the same model that we use in the actual imputation (i.e.
+MICE with PMM)) using the remaining folds with sex observed to build a model used to imputed the sex of each fish in the held out fold (along with any other variables that needed to be imputed).
+We can then compare the imputed values of sex to the true values of sex to evaluate the imputation accuracy of the imputation model.
+These imputations on the hold out samples were then combined to estimate the probability of male vs female (i.e.
+the probability of being male was estimated as the proportion of imputations that predicted male.
+Using these probabilities, an ROC curve (Figure \ref{fig:roc}) was constructed to evaluate the accuracy of the imputations.
+The AUC  was `r round(auc_rocr@y.values[[1]],3)`, indicating excellent discrimination (@HosmerLemeshow, Page 177) between a fish being male or female.
+
+<!-- So, what area under the ROC curve describes good discrimination? Unfortunately there is no “magic” number, only general guidelines. In general, we use the -->
+<!-- following rule of thumb: -->
+<!-- If -->
+<!-- ⎧ -->
+<!-- ⎪⎪⎪⎪⎪⎪⎪⎪⎨ -->
+<!-- ⎪⎪⎪⎪⎪⎪⎪⎪⎩ -->
+<!-- ROC = 0.5 This suggests no discrimination, so we might as well -->
+<!-- flip a coin. -->
+<!-- 0.5 < ROC < 0.7 We consider this poor discrimination, not much better -->
+<!-- than a coin toss. -->
+<!-- 0.7 ≤ ROC < 0.8 We consider this acceptable discrimination. -->
+<!-- 0.8 ≤ ROC < 0.9 We consider this excellent discrimination. -->
+<!-- ROC ≥ 0.9 We consider this outstanding discrimination. -->
+
+### Summary of Imputed Data 
+```{r est-p, echo = FALSE, fig.cap = "Density estimate of the percentage of specimen that were imputted to be male across the M = 100 imputations"}
 #Summaries
@@
 min <- dat %>% group_by(imp) %>% summarize(pct = 100*mean(gender == "Male")) %>% ungroup() %>% summarize(sd = min(pct))
-max <- dat %>% group_by(imp) %>% summarize(pct = 100*mean(gender == "Male")) %>% ungroup() %>% summarize(sd = min(pct))
-
-dat %>% group_by(imp) %>% summarize(pct = sum(gender == "Male") == sum(gender == "Female")) %>% ungroup() %>% summarize(pct = mean(pct))
-
-```
-
-Across the 100 imputations of the 814 fossil specimen, the average percentage of males in an imputation was `r paste0(round(meanpct$pct,2),"%")` with a standard deviation of `r paste0(round(sdpct$sd,2),"%")` ranging from `r paste0(round(min$sd,2),"%")` to `r paste0(round(max$sd,2),"%")`.
+max <- dat %>% group_by(imp) %>% summarize(pct = 100*mean(gender == "Male")) %>% ungroup() %>% summarize(sd = max(pct))
+
+#dat %>% group_by(imp) %>% summarize(pct = sum(gender == "Male") == sum(gender == "Female")) %>% ungroup() %>% summarize(pct = mean(pct))
+
+```
+
+```{r echo = FALSE}
+#Ignoring time compute lambda 
+library(tidyverse)
+sum_p <- dat %>% group_by(imp) %>% summarize(pctmale = mean(gender == "Male"), n = n()) %>% mutate(W = pctmale*(1-pctmale)/n) %>% ungroup() %>% summarize(q = mean(pctmale), W = mean(W), B = var(pctmale)) %>% mutate(Tm = (1+1/100)*B,lambda = (1+1/100)*B/Tm, se= sqrt(Tm))
+
+```
+
+Across the 100 imputations of the 814 fossil specimen, the average percentage of males in an imputation was `r paste0(round(meanpct$pct,2),"%")` with a standard deviation across imputations of `r paste0(round(sdpct$sd,2),"%")` ranging from `r paste0(round(min$sd,2),"%")` to `r paste0(round(max$sd,2),"%")`.
 In only 31 of the 100 imputations were there more males than females with the remaining 69 imputations having exactly the same number or more females.
+Figure \ref{fig:est-p} shows the distribution of the proportion of males across the $M$ = 100 imputations across all time periods.
+
+```{r male-prop, fig.cap = "Density estimates of the distribution of the proportions of male fish at each time period across 100 imputations"}
+#Now do this across time. 
+library(ggridges)
+dat %>% group_by(imp, time) %>% summarize(pct_male = mean(gender == "Male")) %>% ggplot(aes(x = pct_male, y = time, group = time)) +  geom_density_ridges(fill = "#00AFBB", rel_min_height = 0.05, quantile_lines = TRUE, quantiles = 0.5) + theme_bw() + ylab("Time Period") + xlab("Male Proportion")
+
+#Means of male proportion at each time.
+#dat %>% group_by(imp, time) %>% summarize(mn = mean(gender == "Male")) %>% group_by(time) %>% summarize(pct_male = mean(mn)) %>% ggplot(aes(x = time, y = pct_male)) + geom_point() + geom_path() + theme_bw()
+```
+
+```{r}
+pct_male_medians <- dat %>% group_by(imp, time) %>% summarize(mn = mean(gender == "Male")) %>% group_by(time) %>% summarize(pct_male = median(mn))
+```
+
+Figure \ref{fig:male-prop} shows density estimates of the distribution of the proportions of male fish at each time period across 100 imputations along with the median at each time point.
+We can see that the first three time points have proportions of males centered around about 25\% followed by a shift up in proportion of males between time periods 5 and 9 with median values ranging from 44\% to 66.7\%.
+Time periods 10 through 13 show a shift back towards more females with the median proportions of males ranging from 30.3\% to 41.5\%.
+The last several time points, from 14 through 18, all have median male proportions just above 50\% with the exception of time period 17 with a median male proportion of 39.3\%.
+
+<!-- Finally, the fraction of missing information, $\lambda$, for the mean of all traits other than sex were all below 0.003. -->
 
@@
 <!-- If we want to compute fraction of missing information.  This also computes imputed means.  -->
-```{r echo = FALSE}
+```{r echo = FALSE, eval = FALSE}
 greg <- function(x){
@@
 }
-data.frame(var = names(dat)[4:28],do.call(rbind,lapply(as.list(names(dat)[4:28]),greg))) %>% select(var, Q, se, lambda)
+temp <- data.frame(var = names(dat)[4:28],do.call(rbind,lapply(as.list(names(dat)[4:28]),greg))) %>% select(var, Q, se, lambda)
 
@@
 
-\textcolor{red}{Greg, WE NEED RESULTS SECTION DESCRIBING THE RESULTS OF AKHIL'S MULTIPLE IMPUTATION, BOTH THE TESTS OF HOW WELL THE MODEL WORKED CLASSIFYING MODEL SAMPLES, AND SUMMARY STATISTICS OF FOSSIL IMPUTATION RESULTS -YS}
+<!-- total sampel size1181 -->
+<!-- fossils 814 -->
+<!-- extant 367 -->
+
+<!-- \textcolor{red}{Greg, WE NEED RESULTS SECTION DESCRIBING THE RESULTS OF AKHIL'S MULTIPLE IMPUTATION, BOTH THE TESTS OF HOW WELL THE MODEL WORKED CLASSIFYING MODEL SAMPLES, AND SUMMARY STATISTICS OF FOSSIL IMPUTATION RESULTS -YS} -->
 

### 2024-11-25 manuscript/paper.Rmd
addressing Greg's changes and additions. Some questions back to you.

@@
 Ecological release theory suggests that a population’s niche should change when important species interactions like resource competition or predation are relaxed or removed (reviewed in @Herrmannetal2021).
-Removal is posited to create ecological opportunity—i.e., aspects of the niche become newly accessible, and the focal population shifts and/or expands its resource use new resources (@ParentandCrespi2009; @Herrmannetal2021).
+Removal is posited to create ecological opportunity—i.e., aspects of the niche become newly accessible, and the focal population shifts and/or expands its resource use (@ParentandCrespi2009; @Herrmannetal2021).
 Ecological release may then be followed by adaptive morphological evolution as traits change to reflect the new niche (@ParentandCrespi2009; @Herrmannetal2021).
 
-For example, a population undergoing ecological release can experience disruptive selection on males and females stemming from intraspecific competition over newly accessible resources (@BolnickandDoebeli2003; @BolnickandLau2008; @Cooperetal2011).
+For example, a population undergoing ecological release can experience novel disruptive selection on males and females stemming from intraspecific competition over newly accessible resources (@BolnickandDoebeli2003; @BolnickandLau2008; @Cooperetal2011).
 This is predicted to result in intersexual divergence in habitat use and associated phenotypes (@Schoener1968; @Shine1989; @BolnickandDoebeli2003; @Butleretal2007; @BolnickandLau2008; @Cooperetal2011; but see @Stuartetal2021; @Blain2022).
@@
 Much of the theory and empirical data for character displacement between the sexes is based on release from resource competition specifically (e.g., @BolnickandDoebeli2003; @Cooperetal2011; @PfennigandPfennig2012; @DeLisleandRowe2015; @DeLisleetal2018).
-However, release from predation might also drive the evolution of increased sexual dimorphism because an absence of predators should generate ecological opportunity (@ReimchenandNosil2004; @ParentandCrespi2009; @Herrmannetal2021).
-
-Here, we tested the prediction that a release from predation results in the evolution of increased sexual dimorphism using a well-preserved, finely-resolved sequence of a fossil threespine stickleback fish (Gasterosteus doryssus).
+However, release from predation might also drive the evolution of increased sexual dimorphism because an absence of predators should also generate ecological opportunity (@ReimchenandNosil2004; @ParentandCrespi2009; @Herrmannetal2021).
+
+Here, we tested this prediction that a release from predation results in the evolution of increased sexual dimorphism using a well-preserved, finely-resolved sequence of a fossil threespine stickleback fish (*Gasterosteus doryssus*).
 The sequence in this depositional environment is comprised of two lineages (@Bell2009; @Cerasonietal2024).
@@
 
-The observation of 93,000 years of low-armor stasis in lineage I (@Belletal1985) and the observation of rapid evolution of armor loss by lineage II, both suggest that this depositional environment lacked piscivorous fish like trout and other salmonids known to prey on modern threespine stickleback.
+The observation of 93,000 years of low-armor stasis in lineage I and the observation of rapid evolution of armor loss by lineage II, both suggest that this depositional environment lacked piscivorous fish like trout and other salmonids known to prey on modern threespine stickleback.
 Armor presence in extant threespine stickleback (Gasterosteus aculeatus) correlates strongly with the presence of vertebrate piscivores; populations with less predation pressure typically have less armor (@Reimchen1994, @ReimchenandNosil2004; @Belletal1993; @Roestietal2023).
 In our paleolake basin, only three fossil trout have been found in the same section of rock that has revealed >20,000 threespine stickleback fossils as well as occasional killifish (Fundulus nevadensis) (@Bell2009; @Cerasonietal2024).
-Thus, we reconstruct an evolutionar history in which it is likely that lineage II migrated from a nearby paleolake basin that had predators since it was armored when it arrived (@Bell2009; @Cerasonietal2024).
-Lineage II then experienced release from predators in the focal paleolake basin, generating the initial conditions of evolutionary models predicting the evolution of increased sexual dimorphism following ecological release.
-We test this prediction here.
+Thus, we reconstruct an evolutionary history in which it is likely that lineage II migrated from a nearby paleolake basin that had predators since it was armored when it arrived (@Bell2009; @Cerasonietal2024).
+Lineage II then experienced release from predators in the focal paleolake basin studied here, generating the initial conditions of evolutionary models predicting the evolution of increased sexual dimorphism following ecological release.
 
@@
 
-For our study, we inferred sex in G.
-doryssus fossils by comparing multivariate morphological trait data from lineage II samples to the same multivariate trait set collected from multiple populations in the closely-related, extant Threespine Stickleback species complex (Gasterosteus aculeatus).
-Crucially, we determined G.
-aculeatus sex directly via dissection and/or PCR genotyping.
-This enabled us to use Multiple Imputation by Chained Equations (MICE; @MICE) to build a predictive multiple imputation algorithm (@little2002statistical) based on the multivariate morphology of G.
-aculeatus individuals of known sex.
+For our study, we inferred sex in *G.
+doryssus* fossils by comparing multivariate morphological trait data from lineage II samples to the same multivariate trait set collected from multiple populations in the closely-related, extant Threespine Stickleback species complex (Gasterosteus aculeatus).
+Crucially, we determined extant *Gasterosteus aculeatus* sex directly via dissection and/or PCR genotyping.
+This enabled us to use Multiple Imputation by Chained Equations (MICE; @MICE) to build a predictive multiple imputation algorithm (@little2002statistical) based on the multivariate morphology of *G.
+aculeatus* individuals of known sex.
 We applied this algorithm to the fossil data to impute individual sex based on morphology, treating fossil sex as a missing variable.
-With sex assigned to fossil specimens, we fit a modified Ornstein-Uhlenbeck (OU) (@OUProcess) model using a Bayesian framework that accounts for uncertainty in sex classification to test for evolution of sexual dimorphism in each trait over ~16,000 years of lineage II.
+With sex assigned to fossil specimens, we fit a modified Ornstein-Uhlenbeck (OU) (@OUProcess) model using a Bayesian framework that accounts for uncertainty in sex classification to estimate trait means through time.
+Then we quantified the evolution of sexual dimorphism in each trait over ~16,000 years of lineage II.
 
@@
 Series K consisted of 18 samples taken at ~1000-year intervals, and mean sample times span ~16,363 years.
-Fish from series K were measured for 16 ecomorphological traits related to armor, swimming, and feeding (Table 1).
+Fish from series K were previously measured @Stuartetal2020 for 16 ecomorphological traits related to armor, swimming, and feeding (Table 1).
 Series K started at the previously documented horizon when lineage I was replaced by lineage II.
@@
 
-We next replicated fossil data collection (Table 1) on these extant specimens.
+In 2020, we replicated fossil data collection (Table 1) on these extant specimens.
 Standard length as well as pelvic-spine length on each side were measured with calipers.
@@
 
-## Missing data imputation, including fossil sex
-Fossil and extant data were combined to create a single data set which contained small amounts on missingness on the observed traits with the exception of sex, which was completely missing for fossil specimens as this was unobservable.
-The sex of the fossil data was considered to be missing data and was imputed in the same step of the rest of the missing data using multiple imputation (@little2002statistical).
-Imputation was performed using the multiple imputation by chained equations (MICE) algorithm (@MICE) via predictive mean matching (PMM) (@hotdeck) and implemented using the R programming language (@R2022language).
+## Missing data imputation to classify fossil sex
+Fossil and extant data were combined to create a single data set which contained small amounts of missingness on the observed traits \textcolor{red}{Greg, THIS MAY NOT BE TRUE.
+SOME FOSSIL TRAITS HAD LOTS OF MISSINGNESS.
+IS THIS A PROBLEM?} with the exception of sex, which was completely missing for fossil specimens as this was unobservable.
+WE imputed sex and the rest of the missing data in the same step using multiple imputation (@little2002statistical), performed via the "multiple imputation by chained equations" (MICE) algorithm (@MICE) with predictive mean matching (@hotdeck), and implemented using the R programming language (@R2022language).
 Traditionally, the choice for the number of completed data sets is a relatively small number such as $M = 5$ or $M = 10$.
@@
 Therefore, we imputed $M = 100$ complete data sets.
-Extant fish records were then dropped from each of the imputed data sets, and only the fossil data was used in downstream analyses of fossil sexual dimorphism.
-
+Extant fish records were then dropped from each of the imputed data sets, and only the fossil data were used in downstream analyses of fossil sexual dimorphism.
+To test the accuracy of our multiple imputation model in estimating classifying sex,wWe performed 5-fold cross validation using the extant cases only.
+For a given fold, we took the observed values of sex and removed them \textcolor{red}{Greg, HOW?
+I GUESS WE REMOVED SEX FOR EVERY INDIVIDUAL?} to reproduce the missingness observed in the fossil data.
+Then with one fold held out, we used the remaining folds with observed sex to perform multiple imputation with predictive mean matching, as before, with $M$ = 5,  to build a model that we then used to impute the sex of each fish in the held-out fold (along with any other variables that needed to be imputed).
+We then compared the imputed values of sex in each held-out fold to the true values of sex to evaluate the accuracy of the imputation model, and combined the output to estimate the probability of being correctly called male versus female (i.e.
+the probability of being male was estimated as the proportion of imputations that predicted male \textcolor{red}{Greg, THE PROPORTION OF IMPUTATIONS THAT PREDICTED MALE ***CORRECTLY***?}).
+We constructed an ROC curve from these probabilities (Figure \ref{fig:roc}) to visualize imputation accuracy.
+
+\textcolor{red}{THIS STATEMENT ABOUT POOLING DRAWS SEEMS OUT OF PLACE. SHOULDN'T IT GO JUST BEFORE WE TALK ABOUT USING THE POOLED DRAWS?}
 After performing the Bayesian analysis on each of the $M = 100$ imputed data sets, draws from the posterior distribution based on each of the imputed data sets were pooled together to combine posterior draws across imputations as recommended in @ZhouReiter2010 rather than using Rubin's traditional combining rules.
@@
 -->
-\textcolor{red}{Greg, PRESUMABLY this also imputes missing data in the other traits? We should probably say that explicitly. -YS}
-\textcolor{orange}{Updated - GReg}
-\textcolor{red}{Greg, AKHIL'S METHODS FOR VALIDATION OF THE IMPUTATION MODEL GOES HERE}
-\textcolor{orange}{Are we sure we want that here?  I was going to add it in later}
 
@@
  The goal of our analysis was to model fossil trait means through time, by sex, while incorporating uncertainty across imputations in the classification of sex.
-We therefore implemented a Bayesian hierarchical model structure (described below) to account for classfication uncertainty and the fact that the rate of change of any one trait mean over time may be a non-constant value.
-
-To estimate fossil sample means through time, for a given imputed dataset, we let $W_{ij}$ be the imputed sex and $\boldsymbol{Y}_{ij}$ be the $K \times 1$ vector of phenotypes for stickleback fossil $j$ at time $t_i$ where $i = 1, \ldots, T$ and $j = 1,\ldots,n_{t}$.
+We therefore implemented a Bayesian hierarchical model structure (described nesxt) to account for classfication uncertainty and the fact that the rate of change of any one trait mean over time may be a non-constant value.
+
+\textcolor{red}We used each for the M = 100 imputed data sets to estimate fossil sample means through time.} For a given imputed dataset, we let $W_{ij}$ be the imputed sex and $\boldsymbol{Y}_{ij}$ be the $K \times 1$ vector of phenotypes for stickleback fossil $j$ at time $t_i$ where $i = 1, \ldots, T$ and $j = 1,\ldots,n_{t}$.
 In addition, we denote $Y_{K,ij}$, the last variable in $\boldsymbol{Y}_{ij}$, to be the standard length of each fish individual, a measure of body size.
+\textcolor{red}I can't remember. Do we analyze the size-corrected data at all? If not, we should say we left those behind, because we did use those to build the MICE model}
 \begin{align}
@@
 \end{align}
-and logit is a function to convert the probability to the log-odds, to allow for easier modelling of linear techniques while still ensuring any probability predictions are guaranteed to be between 0 and 1.
-Modeling traits this way allows us to first estimate the probability of trait loss for an individual at a given time, and ask if the sexes differ in their tendency toward loss through time.
+and logit is a function to convert the probability to the log-odds, to allow for easier modelling of linear techniques while still ensuring any probability predictions are between 0 and 1.
+Modeling traits this way allows us to first estimate the probability of trait loss for an individual at a given time and ask if the sexes differ in their tendency toward loss through time.
 Then, for individuals that have not lost the trait, we can estimate trait values and calculate sexual dimorphism through time.
@@
 \end{align}
-where GP denotes a generalized Poisson distribution with $E(X) = \lambda$ and $Var(X) = \frac{\lambda}{(1 - \alpha)^2}$.
+where GP denotes a Generalized Poisson distribution with $E(X) = \lambda$ and $Var(X) = \frac{\lambda}{(1 - \alpha)^2}$.
 
 In addition, we modeled two discrete traits, abdominal vertebrae number (mav) and caudal vertebrae number (mcv) to vary with standard length.
-This is because these are vertebral counts that made along the anterior-posterior access of the fish; these vertebrae contribute to the length of the fish directly and may evolve in a correlated fashion.
+This is because these are vertebral counts that made along the anterior-posterior access of the fish; these vertebrae contribute to the length of the fish directly and may be correlated with standard length.
 If $Y_{k,ij}$ is one of the above traits, then
@@
 <!-- In the above model descriptions, $\mu_{k,ft_i}$ and $\mu_{k,mt_i}$ model the time-$t_i$, sample-specific (potentially log-transformed) trait mean $k$ for females and males, respectively.
--->  To account for the possibility that there is a time-dependent (i.e., evolutionary trend) trend in $\mu_{k,ft_i}$ and $\mu_{k,mt_i}$, we further set
+-->  To account for the possibility that there is a time-dependent (i.e., evolutionary) trend in $\mu_{k,ft_i}$ and $\mu_{k,mt_i}$, we further set
 \begin{align}
@@
 More specifically, for the $j^{th}$ imputed data set where $j = 1,\cdots,M$, we obtain $\frac{R}{M}$ draws from the posterior distribution using an MCMC algorithm with a sufficient burnin period, and then we combine these $M$ MCMC draws to say that we have a total of $R$ draws from the overall posterior distribution of interest.
+\textcolor{red}THESE DRAWS ARE USED TO ...
+Is this similar to that statement/paragraph above about pooling draws?
+Should that statement be condensed revised into this section?}
 
@@
 ## Quantifying sexual dimorphism, and testing for increased sexual dimorphism through time.
-We quantify sexual dimorphism at each time $t_i: i = 1,\cdots,T$ by calculating the difference in mean trait size for males and females at each time for a continuous phenotype, $SD_{k,t_i} = \mu_{k,m,t_i} - \mu_{k,f,t_i}$ and for a discrete phenotype $SD_{k,t_i} = \lambda_{k,m,t_i} - \lambda_{k,f,t_i}$.
+We quantify sexual dimorphism at each of the 18 temporal samples $t_i: i = 1,\cdots,T$ by calculating the difference in mean trait size for males and females at each time for a continuous phenotype, $SD_{k,t_i} = \mu_{k,m,t_i} - \mu_{k,f,t_i}$ and for a discrete phenotype $SD_{k,t_i} = \lambda_{k,m,t_i} - \lambda_{k,f,t_i}$.
+That is, we subtract female means from male means.
 For simplicity, we will define sexual dimorphism for both continuous and discrete traits as $SD_{k,t_i} = \mu_{k,m,t_i} - \mu_{k,f,t_i}$.
-Because we are performing a Bayesian analysis, we do not obtain a single point estimate of sexual dimorphism at a particular time, but rather $R$ samples from the posterior distribution of sexual dimorphism given the observed fossil and extant data.
+Because we are performing a Bayesian analysis, we do not obtain a single point estimate of sexual dimorphism at a particular time, but rather $R$ samples from the posterior distribution of sexual dimorphism given the observed fossil and extant data \textcolor{red}OKAY, HERE AGAIN WE TALK ABOUT R SAMPLES DRAWN FROM THE IMPUTATIONS.
+THIS IS GETTING CONFUSING.}.
 Therefore, our analysis will be conducted by calculating $SD_{k,t_i,r} = \mu_{k,m,t_i,r} - \mu_{k,f,t_i,r}$ for $r = 1,\dots,R$ and then perfoming our data analysis on these samples.
 Presence (or absense) of sexual dimorphism for trait $k$ can be indicated by estimating the posterior probability of positive sexual dimorphism $P(SD_{k,t_i} > 0 | \boldsymbol{y}) = \frac{1}{R} \sum_{r=1}^R I(SD_{k,t_i,r} > 0)$.
-Figures \ref{fig:post_probs_armor} and \ref{fig:post_probs_nonarmor}  plot these posterior probabilities across time for each of the armored traits and non-armored traits, respectively.
-The closer the $P(SD_{k,t_i} > 0 | \boldsymbol{y})$ is to 1, the larger the evidence that the average trait size for males is larger than females, while if $P(SD_{k,t_i} > 0 | \boldsymbol{y})$ is close to 0, the larger the evidence that the average trait size for females is larger than males, both indicating strong presence of sexual dimorphism.
+Figures \ref{fig:post_probs_armor} and \ref{fig:post_probs_nonarmor}  plot these posterior probabilities across time for armored traits and non-armored traits, respectively.
+The closer the $P(SD_{k,t_i} > 0 | \boldsymbol{y})$ is to 1, the larger the evidence that the average trait size for males is larger than females, while if $P(SD_{k,t_i} > 0 | \boldsymbol{y})$ is close to 0, the larger the evidence that the average trait size for females is larger than males, both indicating sexual dimorphism.
 On the contrary, $P(SD_{k,t_i} > 0 | \boldsymbol{y})$ around 0.5 indicates that the average trait size is roughly equivalent for both males and females, providing little evidence of dimorphism.
@@
 Then, to test our prediction that ecological release should result in an increase in sexual dimorphism through time, we calculated Kendall's $\tau$ coefficient for each trait (@KendallsTau).
-Kendall's $\tau$ assesses the ordinal association between a set of bivariate observations, where $\tau > 0$ indicates concordance between the observations (i.e., as $x$ increases, $y$ also tends to increase; here,$x$ is time and $y$ is sexual dimorphism.) Kendall's $\tau$ varies between -1 and 1, and can be thought of as behaving like a correlation.
-Values farther from zero indicate a stronger relationship.
+Kendall's $\tau$ assesses the ordinal association between a set of bivariate observations, where $\tau > 0$ indicates concordance between the observations (i.e., as $x$ increases, $y$ also tends to increase; here,$x$ is time and $y$ is sexual dimorphism.) Kendall's $\tau$ varies between -1 and 1, and behaves like a correlation.
+Values farther from zero indicate a stronger relationship, which in this case is evidence of a directional change in sexual dimorphism through time.
 
@@
 
-### Validating Imputation of Sex
+### Sex Imputation
 ```{r eval = FALSE}
@@
 ```
-
-<!-- While sex is observed for extant fish, the sex for fossil specimen cannot be observed.
-Therefore, we turn to multiple imputation to fill in missing values of sex of fossil specimen using the observed extant fish cases.
-In addition to missing sex for all of the fossil cases, there are several other cases of missing data in some variables in the fossil data, and those are imputed at the same time as sex is imputed.
--->
-
-In order to validate the efficacy of imputing sex in the fossil data, we performed 5-fold cross validation using only the extant cases.
-For a given fold, we took the true, observed values of sex and removed them to create missing data similar to what we observe with the fossil data.
-Then with one fold held out, we performed multiple imputation with $M$ = 5 (using the same model that we use in the actual imputation (i.e.
-MICE with PMM)) using the remaining folds with sex observed to build a model used to imputed the sex of each fish in the held out fold (along with any other variables that needed to be imputed).
-We can then compare the imputed values of sex to the true values of sex to evaluate the imputation accuracy of the imputation model.
-These imputations on the hold out samples were then combined to estimate the probability of male vs female (i.e.
-the probability of being male was estimated as the proportion of imputations that predicted male.
-Using these probabilities, an ROC curve (Figure \ref{fig:roc}) was constructed to evaluate the accuracy of the imputations.
-The AUC  was `r round(auc_rocr@y.values[[1]],3)`, indicating excellent discrimination (@HosmerLemeshow, Page 177) between a fish being male or female.
-
 <!-- So, what area under the ROC curve describes good discrimination? Unfortunately there is no “magic” number, only general guidelines. In general, we use the -->
@@
 <!-- ROC ≥ 0.9 We consider this outstanding discrimination. -->
-
-### Summary of Imputed Data 
+We found that the AUC  was `r round(auc_rocr@y.values[[1]],3)`, indicating that our imputation model had excellent discrimination (@HosmerLemeshow, Page 177) in categorizing fish as male or female in extant *G.
+aculeatus* with known sex.
+
 ```{r est-p, echo = FALSE, fig.cap = "Density estimate of the percentage of specimen that were imputted to be male across the M = 100 imputations"}
@@
 
-Across the 100 imputations of the 814 fossil specimen, the average percentage of males in an imputation was `r paste0(round(meanpct$pct,2),"%")` with a standard deviation across imputations of `r paste0(round(sdpct$sd,2),"%")` ranging from `r paste0(round(min$sd,2),"%")` to `r paste0(round(max$sd,2),"%")`.
+Across the 100 missing data imputations of the 814 fossil specimens, the average percentage of males in an imputation was `r paste0(round(meanpct$pct,1),"%")` with a standard deviation across imputations of `r paste0(round(sdpct$sd,1),"%")`, ranging from `r paste0(round(min$sd,1),"%")` to `r paste0(round(max$sd,1),"%")` Figure \ref{fig:est-p}.
 In only 31 of the 100 imputations were there more males than females with the remaining 69 imputations having exactly the same number or more females.
-Figure \ref{fig:est-p} shows the distribution of the proportion of males across the $M$ = 100 imputations across all time periods.
 

### 2024-11-25 manuscript/paper.tex
addressing Greg's changes and additions. Some questions back to you.

@@
 
-Briefly, we used multiple imputation (Little and Rubin
-(\protect\hyperlink{ref-little2002statistical}{2002})) to impute the sex
-of the fossils. Let \(\boldsymbol{W}\) be an
-\((n_{extant} + n_{fossil}) \times 1\) vector of the covariate sex of
-the stickleback fish and \(\boldsymbol{Y}\) be an
-\((n_{extant} + n_{fossil}) \times K\) matrix of the \(K\) phenotypes of
-interest. Because the sex of the fossilized stickleback fish is
-unobservable, we further define
-\(\boldsymbol{W} = (\boldsymbol{W}_{extant}^T,\boldsymbol{W}_{fossil}^T)^T\)
-where \(\boldsymbol{W}_{extant}\) and \(\boldsymbol{W}_{fossil}\) are
-the \(n_{extant} \times 1\) and \(n_{fossil} \times 1\) vectors of the
-observed extant sex and missing fossil sex, respectively. We imputed
-missing sex for the fossil data by sampling from the posterior
-predictive distribution
-\(P(\boldsymbol{W}_{fossil}|\boldsymbol{W}_{extant},\boldsymbol{Y})\)
-using the multiple imputation by chained equations (MICE) algorithm
-(Buuren and Groothuis-Oudshoorn (\protect\hyperlink{ref-MICE}{2011}))
-with predictive mean matching, implemented in R (CITATION).
-Traditionally, the choice for the number of completed dats sets is a
-relatively small number such as \(M = 5\) or \(M = 10\). However, Zhou
-and Reiter (\protect\hyperlink{ref-ZhouReiter2010}{2010}) recommend a
-larger number of imputed data sets if users intend on performing
-Bayesian analysis after imputation. Therefore, we imputed \(M = 100\)
-complete datasets.
+Fossil and extant data were combined to create a single data set which
+contained small amounts on missingness on the observed traits with the
+exception of sex, which was completely missing for fossil specimens as
+this was unobservable. The sex of the fossil data was considered to be
+missing data and was imputed in the same step of the rest of the missing
+data using multiple imputation (Little and Rubin
+(\protect\hyperlink{ref-little2002statistical}{2002})). Imputation was
+performed using the multiple imputation by chained equations (MICE)
+algorithm (Buuren and Groothuis-Oudshoorn
+(\protect\hyperlink{ref-MICE}{2011})) via predictive mean matching (PMM)
+(Andridge and Little (\protect\hyperlink{ref-hotdeck}{2010})) and
+implemented using the R programming language (R Core Team
+(\protect\hyperlink{ref-R2022language}{2022})). Traditionally, the
+choice for the number of completed data sets is a relatively small
+number such as \(M = 5\) or \(M = 10\). However, Zhou and Reiter
+(\protect\hyperlink{ref-ZhouReiter2010}{2010}) recommend a larger number
+of imputed data sets if users intend on performing Bayesian analysis, as
+was performed here, after imputation. Therefore, we imputed \(M = 100\)
+complete data sets. Extant fish records were then dropped from each of
+the imputed data sets, and only the fossil data was used in downstream
+analyses of fossil sexual dimorphism.
+
+After performing the Bayesian analysis on each of the \(M = 100\)
+imputed data sets, draws from the posterior distribution based on each
+of the imputed data sets were pooled together to combine posterior draws
+across imputations as recommended in Zhou and Reiter
+(\protect\hyperlink{ref-ZhouReiter2010}{2010}) rather than using Rubin's
+traditional combining rules.
+
 \textcolor{red}{Greg, PRESUMABLY this also imputes missing data in the other traits? We should probably say that explicitly. -YS}
+\textcolor{orange}{Updated - GReg}
 \textcolor{red}{Greg, AKHIL'S METHODS FOR VALIDATION OF THE IMPUTATION MODEL GOES HERE}
-
-We then pooled all of the draws from the posterior distribution across
-all the imputed data sets to estimate the posterior distributions of our
-parameters of interest (i.e., the trait values), rather than using
-Rubin's combining rules, following Zhou and Reiter
-(\protect\hyperlink{ref-ZhouReiter2010}{2010}). We then dropped extant
-fish from our imputed data sets and used these posterior distributions
-in our downstream analyses of fossil sexual dimorphism.
+\textcolor{orange}{Are we sure we want that here?  I was going to add it in later}
 
@@
 
-The goal of our analysis was to estimate fossil trait means through
-time, by sex, while incorporating uncertainty across imputations in the
-classification of sex. As we explain in this section, we are
-implementing a complex model structure to account for the fact that the
-rate of change in the mean of the trait over time may be a non-constant
-value. In addition, we are using multiple imputation procedure to
-account for the fact that the sex is unobserved in the fossils in
-addition to other pieces of information that are missing. We implement a
-Bayesian approach because these methods are more effective for analyzing
-hierarchical model structures as well as incorporating missing data.
+The goal of our analysis was to model fossil trait means through time,
+by sex, while incorporating uncertainty across imputations in the
+classification of sex. We therefore implemented a Bayesian hierarchical
+model structure (described below) to account for classfication
+uncertainty and the fact that the rate of change of any one trait mean
+over time may be a non-constant value.
 
@@
 adding an additional parameter, \(\gamma_k\). In addition, the
-continuous traits ds1, ds2, ds3, lps, tpg are unique in that these
-traits tend to be lost over time, results in a proportion of zeros in
-our empirical dataset. For these random variables, we add binary
-variable \(Z_{ij}\), where \(Z_{k,ij} = 1\) means that the trait is
-lost, and \(Z_{k,ij} = 0\) means the trait is observed. For all other
+continuous traits ds1, ds2, ds3, lps, tpg are notable because they tend
+to be lost during the evolution of armor reduction in this system (Bell,
+Baumgartner, and Oslon (\protect\hyperlink{ref-Belletal1985}{1985}),
+Stuart, Travis, and Bell
+(\protect\hyperlink{ref-Stuartetal2020}{2020})). This generates
+meaningful zeros for trait lengths in our empirical dataset that violate
+model assumptions. Therefore, for these random variables, we added the
+binary variable \(Z_{ij}\), where \(Z_{k,ij} = 1\) means that the trait
+is lost, and \(Z_{k,ij} = 0\) means the trait is observed. For all other
 continuous traits, we assume \(Z_{k,ij} = 0\)
@@
 ensuring any probability predictions are guaranteed to be between 0 and
-1.
+1. Modeling traits this way allows us to first estimate the probability
+of trait loss for an individual at a given time, and ask if the sexes
+differ in their tendency toward loss through time. Then, for individuals
+that have not lost the trait, we can estimate trait values and calculate
+sexual dimorphism through time.
 
@@
 for \(\mu_{k,m,t_i} - \mu_{k,f,t_i}\) versus time \(t_i\) for the
-\(r^{th}\) MCMC sample from the posterior distribution.
-\textcolor{red}{Matt, I think there's an absolute value problem here.
-Dimorphism can increase of M-F gets more positive or more negative.
-I've indicated that in the text below, but I could use some help refininig the explanation -YS}
-For each trait, the proportion of values in \(\boldsymbol{\tau}_k\) that
-are greater than 0 (\(\frac{1}{R} \sum_{r=1}^R I(\tau_{k,r} > 0)\)) may
-indicate that males have gotten relatively larger than females,
-indicating more pronounced dimorphism. For each trait, the proportion of
-values in \(\boldsymbol{\tau}_k\) that are less than 0
+\(r^{th}\) MCMC sample from the posterior distribution. For each trait,
+a large proportion of values in \(\boldsymbol{\tau}_k\) that are greater
+than 0 (\(\frac{1}{R} \sum_{r=1}^R I(\tau_{k,r} > 0)\)) may indicate
+that males have gotten relatively larger than females through time,
+indicating an increase in male-biased dimorphism through time. A large
+proportion of values in \(\boldsymbol{\tau}_k\) that are less than 0
 (\(\frac{1}{R} \sum_{r=1}^R I(\tau_{k,r} < 0)\)) may indicate that
-females have gotten relatively larger than males, indicating more
-pronounced dimormphism.
+females have gotten relatively larger than males, indicating an increase
+in female-biased dimorphism through time.
 
@@
 
-The extant data used here consist of a total of 367 specimens all with
-known sex (Table S1). Of these, there are 202 and 165 female and male
-specimens, respectively.
-
 \hypertarget{missing-data-imputation}{%
@@
 
-\textcolor{red}{Greg, WE NEED RESULTS SECTION DESCRIBING THE RESULTS OF AKHIL'S MULTIPLE IMPUTATION, BOTH THE TESTS OF HOW WELL THE MODEL WORKED CLASSIFYING MODEL SAMPLES, AND SUMMARY STATISTICS OF FOSSIL IMPUTATION RESULTS -YS}
+\hypertarget{validating-imputation-of-sex}{%
+\subsubsection{Validating Imputation of
+Sex}\label{validating-imputation-of-sex}}
+
+\begin{figure}
+
+{\centering \includegraphics{paper_files/figure-latex/roc-1}
+
+}
+
+\caption{Reciver-Operating Characteristic Curve comparing imputed sex to actual sex for the extant cases using 5-fold cross validation (AUC = 0.831)}\label{fig:roc}
+\end{figure}
+
+In order to validate the efficacy of imputing sex in the fossil data, we
+performed 5-fold cross validation using only the extant cases. For a
+given fold, we took the true, observed values of sex and removed them to
+create missing data similar to what we observe with the fossil data.
+Then with one fold held out, we performed multiple imputation with \(M\)
+= 5 (using the same model that we use in the actual imputation
+(i.e.~MICE with PMM)) using the remaining folds with sex observed to
+build a model used to imputed the sex of each fish in the held out fold
+(along with any other variables that needed to be imputed). We can then
+compare the imputed values of sex to the true values of sex to evaluate
+the imputation accuracy of the imputation model. These imputations on
+the hold out samples were then combined to estimate the probability of
+male vs female (i.e.~the probability of being male was estimated as the
+proportion of imputations that predicted male. Using these
+probabilities, an ROC curve (Figure \ref{fig:roc}) was constructed to
+evaluate the accuracy of the imputations. The AUC was 0.831, indicating
+excellent discrimination (Hosmer, Lemeshow, and Sturdivant
+(\protect\hyperlink{ref-HosmerLemeshow}{2013}), Page 177) between a fish
+being male or female.
+
+\hypertarget{summary-of-imputed-data}{%
+\subsubsection{Summary of Imputed Data}\label{summary-of-imputed-data}}
+
+\begin{figure}
+
+{\centering \includegraphics{paper_files/figure-latex/est-p-1}
+
+}
+
+\caption{Density estimate of the percentage of specimen that were imputted to be male across the M = 100 imputations}\label{fig:est-p}
+\end{figure}
+
+Across the 100 imputations of the 814 fossil specimen, the average
+percentage of males in an imputation was 46.64\% with a standard
+deviation across imputations of 6.79\% ranging from 30.1\% to 62.29\%.
+In only 31 of the 100 imputations were there more males than females
+with the remaining 69 imputations having exactly the same number or more
+females. Figure \ref{fig:est-p} shows the distribution of the proportion
+of males across the \(M\) = 100 imputations across all time periods.
+
+\begin{figure}
+
+{\centering \includegraphics{paper_files/figure-latex/male-prop-1}
+
+}
+
+\caption{Density estimates of the distribution of the proportions of male fish at each time period across 100 imputations}\label{fig:male-prop}
+\end{figure}
+
+Figure \ref{fig:male-prop} shows density estimates of the distribution
+of the proportions of male fish at each time period across 100
+imputations along with the median at each time point. We can see that
+the first three time points have proportions of males centered around
+about 25\% followed by a shift up in proportion of males between time
+periods 5 and 9 with median values ranging from 44\% to 66.7\%. Time
+periods 10 through 13 show a shift back towards more females with the
+median proportions of males ranging from 30.3\% to 41.5\%. The last
+several time points, from 14 through 18, all have median male
+proportions just above 50\% with the exception of time period 17 with a
+median male proportion of 39.3\%.
+
+\textcolor{red}{Yoel, I think I remember you talking about a reason why it made more sense that there were fewer males in the sample tham the females, but I don't remember the details.
+Am I remembering this correctly?
+Would be nice to add something here about that and say why these imputations are consistent with that.
+-GJM}
 
@@
 
+We estimated the change in sexual dimorphism over time by calculating
+the distribution of Kendall's Tau for each trait. Here, positive values
+of Tau indicate that sexual dimorphism (defined as male means minus
+female means) increased through time. Average Tau values ranged from X
+to Y across traits and the proportion of positive area under the curve
+ranged from Z to W (Table Tau). Distributions are available for each
+trait in the Supplementary Material (Figures SX-SX+15). In Figure (Tau),
+we show representative distributions. For example,
+
+Something about the probability of dimorphim in the probability of loss?
+I.e., the prob\_zero plots. For example, ds2\_prob\_zero has lots of
+positive area under the curve indicating that the rate of loss in males
+and females diverged, such that males were more likely to lose ds2
+through time. Interestingly, for the individuals that kept their spines,
+dimorphism increased, with females getting larger than males (or males
+getting smaller than females).
+
 We can also assess the strength of the dimorphism by calculating the
@@
 threshold \textcolor{red}{(0.5??, idk what this could be.) -MS}
-
-Table \ref{tab:auc-table} shows the probability that the posterior
-distribution of the differences in the mean is greater than the
-posterior distribution of the differences at the first time point.
-Values near 0.5 indicate very little difference between the
-distributions whereas values closer to 0 and 1 indicate changes in the
-distributions with values of 1 meaning the difference in means has
-shifted towards males having a larger mean and values of 0 indicating
-the difference in means has shifted towards larger means in females.
-These results are shown graphically in figure \ref{fig:AUC-plot}.
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-hotdeck}{}}%
+Andridge, R. R., and R. J. A. Little. 2010. {``A Review of Hot Deck
+Imputation for Survey Non-Response.''} \emph{International Statistical
+Review} 78 (1): 40--64.
+
 \leavevmode\vadjust pre{\hypertarget{ref-Baumgartneretal1988}{}}%
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-HosmerLemeshow}{}}%
+Hosmer, D. W., Jr., S. Lemeshow, and R. X. Sturdivant. 2013.
+\emph{Applied Logistic Regression, 3rd Edition}. Wiley.
+
 \leavevmode\vadjust pre{\hypertarget{ref-Huntetal2008}{}}%
@@
 Press, Berkeley, USA.
+
+\leavevmode\vadjust pre{\hypertarget{ref-R2022language}{}}%
+R Core Team. 2022. \emph{R: A Language and Environment for Statistical
+Computing}. Vienna, Austria: R Foundation for Statistical Computing.
+\url{https://www.R-project.org/}.
 

### 2025-09-04 R/StickleImputationCode.Rmd
updated imputations code.

@@
 ```{r}
-setwd('/Users/akhilghosh/Desktop/RStudioProjects2022Fall/SticklebackProject/RawData_ReadOnly')
-stickle <- read_csv("AO.sexualdimorphismPruned.sc.forAnalysis_forMICE_Modeling_20June2023.csv")
+#setwd('/Users/akhilghosh/Desktop/RStudioProjects2022Fall/SticklebackProject/RawData_ReadOnly')
+stickle <- read_csv("../RawData_ReadOnly/AO.sexualdimorphismPruned.sc.forAnalysis_forMICE_Modeling_20June2023.csv")
 
@@
 
+#Manual Fixes
+ # <!-- LQx.B.M.139 was called male in hand (and in our database) but is female by dissection. -->
+ # <!-- LQx.B.M.155 was called male in hand (and in our database) but is female by dissection. -->
+ # <!--     Pax.B.M.081 was called male in hand (and in our database) but is female by dissection. -->
+ # <!--     Pax.L.M.113 was called male in hand (and in our database) but is female by dissection. -->
+ # <!--     Pax.L.M.115 was called male in hand (and in our database) but is female by dissection. -->
+stick$gender[stick$univID == "LQx.B.M.139"] <- "Female"
+#stick$gender[stick$univID == "LQx.B.M.155"] <- "Female"
+stick$gender[stick$univID == "Pax.B.M.081"] <- "Female"
+stick$gender[stick$univID == "Pax.L.M.113"]<- "Female"
+stick$gender[stick$univID == "Pax.L.M.115"] <- "Female"
 ```
@@
 combined_csv<- do.call(rbind,csv_list) %>% select(imp, time,gender, everything())
+write_csv(combined_csv, file = "../Data/updated_stickle_imputations_100_20250904.csv")
 ```

### 2025-09-29 R/StickleImputationCode.Rmd
updated imputation code.

@@
 ```
+
+```{r}
+combined_csv <- read.csv("../Data/updated_stickle_imputations_100_20250904.csv")
+combined_csv %>% group_by(imp) %>% summarize(percent_male = 100*mean(gender == "Male")) %>% ggplot(aes(x = percent_male)) + geom_density() + theme_bw() + xlab("Percent Male")
+```
+
+# Cross validation
+
+```{r}
+set.seed(2025)
+stick_extant <- stick %>% filter(substring(univID,1,3) != "Fos") 
+stick_extant$ind <- sample(rep(1:5,75), nrow(stick_extant),replace = FALSE)
+
+M <- 100
+preds <- list()
+for (i in 1:5){
+temp <- stick_extant 
+temp$gender[temp$ind == i] <- NA
+imp_temp <- mice(temp %>% select(-ind),m = M)
+comp_temp <- complete(imp_temp, "long")
+comp_temp$ind <- rep(temp$ind,M)
+comp_temp <- comp_temp %>% mutate(gender_true = rep(stick_extant$gender,M)) %>% filter(ind == i) 
+preds[[i]] <- comp_temp %>% group_by(.id, gender_true) %>% summarize(pct_male = mean(gender == "Male"))
+}
+
+stack <- do.call(rbind,preds)
+
+library(ROCR)
+roc_pred <- prediction(stack$pct_male, stack$gender_true)
+auc_rocr <- performance(roc_pred, measure="auc")
+perf <- performance(roc_pred, measure="tpr", x.measure="fpr")
+data.frame(FPR = perf@x.values[[1]], TPR = perf@y.values[[1]]) %>% ggplot(aes(x = FPR,y = TPR)) + geom_line() + theme_bw() + geom_abline(slope = 1, linetype = 3)
+```

### 2025-09-29 R/stickle_imputation.Rmd
updated imputation code.

@@
 
+ <!-- LQx.B.M.139 was called male in hand (and in our database) but is female by dissection. -->
+ <!-- LQx.B.M.155 was called male in hand (and in our database) but is female by dissection. -->
+ <!--     Pax.B.M.081 was called male in hand (and in our database) but is female by dissection. -->
+ <!--     Pax.L.M.113 was called male in hand (and in our database) but is female by dissection. -->
+ <!--     Pax.L.M.115 was called male in hand (and in our database) but is female by dissection. -->
+
 ```{r}
-stickle <- read_csv('stickle_revamped.csv')
+stickle <- read_csv('../Data/Stickle_revamped.csv')
 
@@
 
-stick
+#Manual Fixes
+ # <!-- LQx.B.M.139 was called male in hand (and in our database) but is female by dissection. -->
+ # <!-- LQx.B.M.155 was called male in hand (and in our database) but is female by dissection. -->
+ # <!--     Pax.B.M.081 was called male in hand (and in our database) but is female by dissection. -->
+ # <!--     Pax.L.M.113 was called male in hand (and in our database) but is female by dissection. -->
+ # <!--     Pax.L.M.115 was called male in hand (and in our database) but is female by dissection. -->
+stick$gender[stick$univID == "LQx.B.M.139"] <- "Female"
+#stick$gender[stick$univID == "LQx.B.M.155"] <- "Female"
+stick$gender[stick$univID == "Pax.B.M.081"] <- "Female"
+stick$gender[stick$univID == "Pax.L.M.113"]<- "Female"
+stick$gender[stick$univID == "Pax.L.M.115"] <- "Female"
+
+#Exploring the ones that were changed
+stick %>% filter(univID %in% c("LQx.B.M.139","Pax.B.M.081","Pax.L.M.113","Pax.L.M.115"))
+
+stick %>% mutate(changed = univID %in% c("LQx.B.M.139","Pax.B.M.081","Pax.L.M.113","Pax.L.M.115")) %>% filter(changed) %>% ggplot(aes(x = ds3, y = lps, col = changed)) + geom_point(alpha = 0.5)
+
+stick %>% mutate(changed = univID %in% c("LQx.B.M.139","Pax.B.M.081","Pax.L.M.113","Pax.L.M.115"))  %>% ggplot(aes(x = changed, y = lps, color = changed)) + geom_boxplot() + geom_jitter()
+
+stick %>% mutate(changed = univID %in% c("LQx.B.M.139","Pax.B.M.081","Pax.L.M.113","Pax.L.M.115"))  %>% ggplot(aes(x = changed, y = ds3, color = changed)) + geom_boxplot() + geom_jitter()
 
